---
title: "Join data sources (daily)"
author: "Fintan Mooney"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

```{r initiate}
rm(list = ls())
# 0a.Declare root directory, folder location and load essential stuff
project.folder = paste0(print(here::here()), '/')
source(paste0(project.folder, 'create_folder_structure.R'))
source(paste0(functions.folder, 'script_initiate.R'))
```

# 1. Load 10 km fishnet geometry

```{r fishnet}
fishnet_geom <- st_read(paste0(spatial_data, "/fishnet_10km_timeseries.gpkg")) %>%
  distinct(zone_id, geom)
```

# 2. Build the zone × date scaffold

```{r scaffold}
study_start <- as.Date("2021-11-15")
study_end   <- as.Date("2022-12-31")
date_seq    <- seq(study_start, study_end, by = "day")
zone_ids    <- fishnet_geom %>% st_drop_geometry() %>% pull(zone_id) %>% unique()

daily_fishnet <- expand.grid(
  zone_id = zone_ids,
  date    = date_seq,
  stringsAsFactors = FALSE
) %>%
  mutate(
    year     = lubridate::year(date),
    month    = lubridate::month(date),
    iso_week = lubridate::isoweek(date),
    iso_year = lubridate::isoyear(date),
    dow      = lubridate::wday(date, label = TRUE, week_start = 1),
    doy      = lubridate::yday(date)
  ) %>%
  as.data.table()

setkey(daily_fishnet, zone_id, date)
```

# 3. Load and join spillover events

```{r spillover}
spillover_file <- file.path(spillover_data, "hpai-outbreaks-clean-placeholder-032124.csv")
if (!file.exists(spillover_file)) {
  stop("Missing ", spillover_file)
}
spillover_raw <- read_csv(spillover_file)

# 3a.Keep IND (independent introduction) and INR; exclude CSLT (common source lateral transfer)
spillover_raw <- spillover_raw %>% filter(link %in% c("IND", "INR"))

# 3b.Use date_confirmed as the case-defining date (USDA surveillance convention;
#    median 2 d after presumptive onset). A date_presumptive sensitivity is
#    reported separately.
spillover_raw <- spillover_raw %>% mutate(case_date = date_confirmed)

spillover_mn <- spillover_raw %>%
  filter(state == "MN",
         dplyr::between(case_date, as.Date("2022-01-01"), as.Date("2022-12-31")))

spillover_sf <- spillover_mn %>%
  st_as_sf(coords = c("lon", "lat"), crs = 4326) %>%
  st_transform(crs = st_crs(fishnet_geom))

outbreaks_in_cells <- st_join(spillover_sf, fishnet_geom, join = st_within) %>%
  st_drop_geometry()

outbreak_daily <- outbreaks_in_cells %>%
  group_by(zone_id, date = case_date) %>%
  summarise(
    outbreak_count = n(),
    outbreak_id    = paste(outbreak_id, collapse = ", "),
    .groups = "drop"
  ) %>%
  mutate(outbreak_binary = 1L) %>%
  as.data.table()

any_outbreak_zones <- outbreak_daily[, .(any_outbreak = 1L), by = zone_id]
outbreak_dates_by_zone <- outbreak_daily[
  , .(outbreak_date = paste(sort(unique(as.character(date))), collapse = ", ")),
  by = zone_id
]

setkey(outbreak_daily, zone_id, date)
daily_fishnet <- merge(daily_fishnet, outbreak_daily,
                       by = c("zone_id", "date"), all.x = TRUE)
daily_fishnet <- merge(daily_fishnet, any_outbreak_zones, by = "zone_id", all.x = TRUE)
daily_fishnet <- merge(daily_fishnet, outbreak_dates_by_zone, by = "zone_id", all.x = TRUE)

daily_fishnet[, `:=`(
  outbreak_count  = fcoalesce(outbreak_count, 0L),
  outbreak_binary = fcoalesce(outbreak_binary, 0L),
  any_outbreak    = fcoalesce(any_outbreak, 0L),
  outbreak_id     = fcoalesce(outbreak_id, NA_character_),
  outbreak_date   = fcoalesce(outbreak_date, NA_character_)
)]
```

# 4. Load daily ERA5-Land met
#
# Two supported sources, selected by USE_MULTISTATE_ERA5:
#
# (A) Multistate (default): bi-weekly CSVs from the new GEE pipeline
#     (a_00_era5_land_gee_multistate.js), one CSV per (state, window). These
#     carry raw ERA5 band names and ~2 000 MN cells with their own zone_ids.
#     We map those onto the existing MN fishnet via a centroid-contains spatial
#     join so the rest of the pipeline (cases, birds) stays in the legacy
#     zone_id space. The multistate CSVs expose the new variables we added in
#     the Nov 2025 refresh: snow_depth, snow_cover, skin_temperature,
#     soil_temperature, soil_moisture_2, snowfall.
#
# (B) Legacy single-state: the older mn_daily_era5land_*.csv files from the
#     rgee pipeline, which already use friendly column names but lack the new
#     variables. Kept as a fallback if you ever want to re-run the previous
#     analysis exactly.

```{r met}
USE_MULTISTATE_ERA5 <- TRUE   # set FALSE to revert to legacy single-state CSVs

if (USE_MULTISTATE_ERA5) {

  multistate_dir <- file.path(env_data, "hpai_multistate_era5")
  state_filter   <- "Minnesota"
  state_tag      <- "MN"

  # ---- 4a. Load multistate fishnet (5 states) and filter to MN ----
  # GEE exports GeoJSON with EPSG:5070 (NAD83 Albers) projected coordinates but
  # tags the metadata as WGS84 — we must override the CRS rather than transform.
  multistate_geom <- st_read(file.path(multistate_dir, "fishnet_8state.geojson"),
                             quiet = TRUE)
  multistate_geom <- multistate_geom[!st_is_empty(multistate_geom), ]
  multistate_geom <- st_set_crs(multistate_geom, 5070)
  if (!"state" %in% names(multistate_geom)) {
    stop("Multistate fishnet missing `state` column")
  }
  multistate_mn <- multistate_geom %>% filter(state == state_filter)
  multistate_mn <- st_transform(multistate_mn, st_crs(fishnet_geom))
  message("Multistate MN cells: ", nrow(multistate_mn),
          "  |  legacy MN fishnet cells: ", nrow(fishnet_geom))

  # ---- 4b. Build legacy_zone → multistate_zone mapping via centroid-contains ----
  legacy_centroids <- st_centroid(fishnet_geom)
  legacy_centroids$legacy_zone_id <- fishnet_geom$zone_id
  legacy_centroids <- legacy_centroids[, "legacy_zone_id"]

  zone_map <- st_join(legacy_centroids,
                      multistate_mn %>% select(multistate_zone_id = zone_id),
                      join = st_within) %>%
    st_drop_geometry() %>%
    as.data.table()
  # Some legacy cells sit on the multistate-fishnet boundary or just outside MN
  # (e.g. small island cells) — assign via nearest-feature as a fallback.
  if (any(is.na(zone_map$multistate_zone_id))) {
    missing_idx <- which(is.na(zone_map$multistate_zone_id))
    nn <- st_nearest_feature(legacy_centroids[missing_idx, ], multistate_mn)
    zone_map[missing_idx, multistate_zone_id := multistate_mn$zone_id[nn]]
    message("Centroid-contain missed ", length(missing_idx),
            " legacy cells; filled via nearest-feature.")
  }
  setkey(zone_map, multistate_zone_id)

  # ---- 4c. Load all bi-weekly MN CSVs and concatenate ----
  mn_csvs <- list.files(multistate_dir,
                        pattern = paste0("^era5_", state_tag, "_.*\\.csv$"),
                        full.names = TRUE)
  if (length(mn_csvs) == 0) stop("No multistate MN CSVs found in ", multistate_dir)
  message("Loading ", length(mn_csvs), " bi-weekly MN CSVs...")
  met_daily <- rbindlist(lapply(mn_csvs, fread), use.names = TRUE, fill = TRUE)
  setnames(met_daily, "zone_id", "multistate_zone_id")

  # Multistate CSV date is "YYYY-MM-DD" (GEE format), not "YYYYMMDD"
  met_daily[, date := as.Date(as.character(date))]

  # ---- 4d. Rename raw ERA5 bands to friendly column names ----
  raw_to_friendly <- c(
    "temperature_2m"                 = "temperature",
    "dewpoint_temperature_2m"        = "dewpoint",
    "skin_temperature"               = "skin_temperature",      # already friendly
    "soil_temperature_level_1"       = "soil_temperature",
    "volumetric_soil_water_layer_1"  = "soil_moisture",
    "volumetric_soil_water_layer_2"  = "soil_moisture_2",
    "total_precipitation_hourly"     = "precipitation",
    "snowfall_hourly"                = "snowfall",
    "surface_runoff_hourly"          = "runoff",
    "total_evaporation_hourly"       = "evapotranspiration",
    "surface_solar_radiation_downwards_hourly"   = "shortwave_radiation",
    "surface_thermal_radiation_downwards_hourly" = "longwave_radiation",
    "leaf_area_index_low_vegetation" = "leaf_area_index"
  )
  for (raw in names(raw_to_friendly)) {
    if (raw %in% names(met_daily) && raw_to_friendly[[raw]] != raw) {
      setnames(met_daily, raw, raw_to_friendly[[raw]])
    }
  }

  # ---- 4e. Derive wind_speed and relative_humidity ----
  met_daily[, wind_speed := sqrt(u_component_of_wind_10m^2 +
                                  v_component_of_wind_10m^2)]
  # Magnus formula (with T, Td in °C). ERA5-Land gives K.
  T_C  <- met_daily$temperature - 273.15
  Td_C <- met_daily$dewpoint   - 273.15
  met_daily[, relative_humidity := 100 *
              exp((17.625 * Td_C) / (243.04 + Td_C)) /
              exp((17.625 * T_C ) / (243.04 + T_C ))]

  # snow_density fill (~100 = no-snow sentinel)
  met_daily[, snow_density := fifelse(snow_density <= 100, 0, snow_density)]

  # ---- 4f. Aggregate multistate cells → legacy zone_id ----
  # legacy zone_id ↔ exactly one multistate zone_id (centroid-contained).
  # If multiple legacy cells map to the same multistate cell, they share the
  # multistate value (no aggregation needed). Mean is a no-op in that case.
  met_daily <- merge(met_daily, zone_map,
                     by = "multistate_zone_id", all.x = FALSE,
                     allow.cartesian = TRUE)
  setnames(met_daily, "legacy_zone_id", "zone_id")

} else {

  # ---- Legacy single-state path (kept for reproducibility) ----
  met_files <- c(
    file.path(env_data, "mn_daily_era5land_2021_novdec.csv"),
    file.path(env_data, "mn_daily_era5land_2022.csv")
  )
  missing <- met_files[!file.exists(met_files)]
  if (length(missing) > 0) stop("Missing daily met file(s):\n  ",
                                paste(missing, collapse = "\n  "))
  met_daily <- rbindlist(lapply(met_files, fread))
  met_daily[, date := as.Date(as.character(date), format = "%Y%m%d")]
  # `runoff` means SURFACE runoff, same as the multistate branch up top and the
  # 8-state panel. the mn_daily_era5land_*.csv files on disk predate that and only
  # carry runoff_sum (surface + sub-surface), so this branch falls back and warns.
  # that's the mismatch the 8-state move was meant to kill, which is why this path
  # is a fallback and not the one you want.
  if ("surface_runoff" %in% names(met_daily)) {
    setnames(met_daily, "surface_runoff", "runoff")
  } else if ("runoff_sum" %in% names(met_daily)) {
    warning("No surface_runoff column: falling back to TOTAL runoff. Re-export ",
            "on the 8-state grid to match the flyway panel.")
    setnames(met_daily, "runoff_sum", "runoff")
  }
  # single-state CSV columns are already friendly. stub the new ones with NA so
  # a pre-refresh export doesn't blow up downstream.
  for (col in c("snow_depth", "snow_cover", "skin_temperature", "soil_temperature",
                "soil_moisture_2", "snowfall", "sub_surface_runoff", "runoff_sum",
                "temperature_C", "leaf_area_index_low", "shortwave_radiation_down")) {
    if (!col %in% names(met_daily)) met_daily[, (col) := NA_real_]
  }
  met_daily[, snow_density := fifelse(snow_density <= 100, 0, snow_density)]
}

# ---- 4g. Common post-load steps for both paths ----
# whitelist, and the intersect() below drops silently, so anything new in the
# CSVs has to be named here or it never reaches the models.
met_cols <- c("temperature", "temperature_C", "dewpoint",
              "relative_humidity", "wind_speed",
              "skin_temperature", "soil_temperature",
              "soil_moisture", "soil_moisture_2",
              "snow_depth", "snow_cover", "snow_density",
              "precipitation", "snowfall",
              "runoff",              # SURFACE runoff, matches the 8-state panel
              "sub_surface_runoff",  # the other half, for the comparison figure
              "runoff_sum",          # surface + sub-surface, what MN used to use
              "evapotranspiration", "leaf_area_index", "leaf_area_index_low",
              "shortwave_radiation", "shortwave_radiation_down",
              "longwave_radiation", "surface_pressure")
# Drop any met_cols not present (single-state legacy path may be missing some)
met_cols <- intersect(met_cols, names(met_daily))
met_daily <- met_daily[, c("zone_id", "date", ..met_cols), with = FALSE]
setkey(met_daily, zone_id, date)

daily_fishnet <- merge(daily_fishnet, met_daily, by = c("zone_id", "date"), all.x = TRUE)
```

# 5. Load weekly eBird and interpolate to daily within each zone

```{r bird}
bird_outbreak <- read_csv(file.path(bird_abundance_data,
                                    "mn-outbreak-weekly-bird-estimate-2022-012926.csv"))

anseriformes <- c(
  "mallar3", "cangoo", "wooduc", "rinduc", "comgol", "norsho",
  "buffle", "lessca", "rebmer", "gnwtea", "gadwal", "ambduc",
  "rudduc", "rethaw", "truswa", "commer"
)
predators <- c("buwtea", "grhowl", "amwpel", "baleag", "brdowl")

bird_wide <- bird_outbreak %>%
  pivot_wider(
    id_cols = c(outbreak_id, mmwr_week, week, week_outbreak),
    names_from = species_name,
    values_from = c(abd_avg, abd_sum),
    names_glue = "{.value}_{species_name}"
  ) %>%
  mutate(
    total_bird_abundance = rowSums(across(starts_with("abd_avg_") &
                                          !matches("anseriformes|predators")), na.rm = TRUE),
    anseriformes = rowSums(across(all_of(paste0("abd_avg_", anseriformes))), na.rm = TRUE),
    predators    = rowSums(across(all_of(paste0("abd_avg_", predators))),    na.rm = TRUE)
  ) %>%
  select(outbreak_id, mmwr_week, total_bird_abundance, anseriformes, predators)

zone_outbreak_map <- outbreaks_in_cells %>%
  select(outbreak_id, zone_id) %>%
  distinct()

bird_zone_week <- bird_wide %>%
  inner_join(zone_outbreak_map, by = "outbreak_id") %>%
  group_by(zone_id, mmwr_week) %>%
  summarise(
    total_bird_abundance = mean(total_bird_abundance, na.rm = TRUE),
    anseriformes         = mean(anseriformes,         na.rm = TRUE),
    predators            = mean(predators,            na.rm = TRUE),
    .groups = "drop"
  ) %>%
  as.data.table()

bird_neg_path <- file.path(bird_abundance_data, "all-mn-feedlot-species-abundance-070125.csv")
bird_neg_raw  <- fread(bird_neg_path)

week_cols <- setdiff(names(bird_neg_raw), c("zone_id", "species"))
bird_neg_long <- melt(
  bird_neg_raw,
  id.vars = c("zone_id", "species"),
  measure.vars = week_cols,
  variable.name = "mmwr_week",
  value.name = "abundance"
)
bird_neg_long[, mmwr_week := as.integer(gsub("week_", "", mmwr_week))]
bird_neg_long[, group := fifelse(species %in% anseriformes, "anseriformes",
                          fifelse(species %in% predators, "predators", "other"))]

bird_neg_zone_week <- bird_neg_long[
  , .(abundance = mean(abundance, na.rm = TRUE)),
  by = .(zone_id, mmwr_week, species, group)
]
bird_neg_group <- bird_neg_zone_week[
  , .(value = sum(abundance, na.rm = TRUE)),
  by = .(zone_id, mmwr_week, group)
]
bird_neg_wide <- dcast(bird_neg_group, zone_id + mmwr_week ~ group,
                      value.var = "value", fill = 0)
if (!"anseriformes" %in% names(bird_neg_wide)) bird_neg_wide[, anseriformes := 0]
if (!"predators"    %in% names(bird_neg_wide)) bird_neg_wide[, predators    := 0]

bird_neg_total <- bird_neg_zone_week[
  , .(total_bird_abundance = sum(abundance, na.rm = TRUE)),
  by = .(zone_id, mmwr_week)
]
bird_neg_wide <- merge(bird_neg_wide, bird_neg_total, by = c("zone_id", "mmwr_week"), all = TRUE)
bird_neg_wide[is.na(total_bird_abundance), total_bird_abundance := 0]
bird_neg_wide[is.na(anseriformes), anseriformes := 0]
bird_neg_wide[is.na(predators),    predators    := 0]

case_zone_ids <- unique(bird_zone_week$zone_id)
bird_zone_week_all <- rbind(
  bird_zone_week[, .(zone_id, mmwr_week, total_bird_abundance, anseriformes, predators)],
  bird_neg_wide[!zone_id %in% case_zone_ids,
                .(zone_id, mmwr_week, total_bird_abundance, anseriformes, predators)]
)
setkey(bird_zone_week_all, zone_id, mmwr_week)

mmwr_anchor <- function(week, year = 2022) {
  jan1       <- as.Date(paste0(year, "-01-01"))
  wday1      <- as.integer(format(jan1, "%w"))
  mmwr_start <- jan1 - wday1
  mmwr_start + 7 * (week - 1) + 3
}
bird_zone_week_all[, anchor_date := mmwr_anchor(mmwr_week, 2022)]

interpolate_bird_daily <- function(anchors_df, target_dates) {
  vars <- c("total_bird_abundance", "anseriformes", "predators")
  out <- data.frame(date = target_dates)
  for (v in vars) {
    if (nrow(anchors_df) >= 2) {
      f <- stats::approxfun(anchors_df$anchor_date, anchors_df[[v]],
                            method = "linear", rule = 2)
      out[[v]] <- f(target_dates)
    } else if (nrow(anchors_df) == 1) {
      out[[v]] <- rep(anchors_df[[v]][1], length(target_dates))
    } else {
      out[[v]] <- NA_real_
    }
  }
  out
}

bird_daily_list <- lapply(split(bird_zone_week_all, bird_zone_week_all$zone_id), function(g) {
  anchors <- as.data.frame(g[order(anchor_date)])
  daily   <- interpolate_bird_daily(anchors, date_seq)
  daily$zone_id <- g$zone_id[1]
  daily
})
bird_daily <- rbindlist(bird_daily_list)
setcolorder(bird_daily, c("zone_id", "date", "total_bird_abundance",
                          "anseriformes", "predators"))
setkey(bird_daily, zone_id, date)

daily_fishnet <- merge(daily_fishnet, bird_daily, by = c("zone_id", "date"), all.x = TRUE)
daily_fishnet[is.na(total_bird_abundance), total_bird_abundance := 0]
daily_fishnet[is.na(anseriformes),         anseriformes         := 0]
daily_fishnet[is.na(predators),            predators            := 0]

zone_zero_summary <- daily_fishnet[, .(total_abundance = sum(total_bird_abundance, na.rm = TRUE)),
                                   by = zone_id]
zone_ids_all_zero <- zone_zero_summary[total_abundance == 0, zone_id]
daily_fishnet[, bird_zero := fifelse(zone_id %in% zone_ids_all_zero, "No Birds", "Has Birds")]
```

# 6. Feedlot covariates

```{r feedlots}
feedlots <- st_read(file.path(feedlot_data, "feedlots.shp")) %>%
  st_transform(crs = st_crs(fishnet_geom))

feedlots_clean <- feedlots %>%
  mutate(
    cafo_flag  = ifelse(!is.na(cafo_flag) & cafo_flag != "N", 1L, 0L),
    near_river = ifelse(!is.na(river_stre) & river_stre != "N", 1L, 0L)
  )

feedlots_in_cells <- st_join(feedlots_clean, fishnet_geom, join = st_within) %>%
  st_drop_geometry()

feedlot_summary <- feedlots_in_cells %>%
  group_by(zone_id) %>%
  summarise(
    feedlot_count = n(),
    au_count      = sum(au_count, na.rm = TRUE),
    birds_au      = sum(birds_au, na.rm = TRUE),
    cafo_flag     = max(cafo_flag, na.rm = TRUE),
    near_river    = max(near_river, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  as.data.table()

daily_fishnet <- merge(daily_fishnet, feedlot_summary, by = "zone_id", all.x = TRUE)
daily_fishnet[, `:=`(
  feedlot_count = fcoalesce(feedlot_count, 0L),
  au_count      = fcoalesce(au_count,      0),
  birds_au      = fcoalesce(birds_au,      0),
  cafo_flag     = fcoalesce(cafo_flag,     0L),
  near_river    = fcoalesce(near_river,    0L)
)]
```

# 7. Land cover

```{r landcover}
land_cover <- st_read(file.path(spatial_data, "fishnet_landcover.gpkg"))

land_cover_clean <- land_cover %>%
  st_drop_geometry() %>%
  select(zone_id, land_cover_mode, land_cover_label, lat, lon)

daily_fishnet <- merge(daily_fishnet, land_cover_clean, by = "zone_id", all.x = TRUE)
daily_fishnet[, land_cover_label := relevel(
  factor(land_cover_label), ref = "Developed, Medium Intensity"
)]
daily_fishnet[, farm_possible := fifelse(
  land_cover_label %in% c("Cultivated Crops", "Pasture/Hay", "Emergent Herbaceous Wetlands"),
  1L, 0L
)]
```

# 8. Export

```{r export}
setkey(daily_fishnet, zone_id, date)

saveRDS(daily_fishnet, paste0(objects_folder, "mn_fishnet_daily_processed.rds"))
fwrite(daily_fishnet,  paste0(processed_data, "mn_fishnet_daily_processed.csv"))

fishnet_spatial <- fishnet_geom %>%
  left_join(
    daily_fishnet[, .(any_outbreak     = max(any_outbreak),
                      feedlot_count    = first(feedlot_count),
                      bird_zero        = first(bird_zero),
                      land_cover_label = first(as.character(land_cover_label))),
                  by = zone_id],
    by = "zone_id"
  )
saveRDS(fishnet_spatial, paste0(objects_folder, "mn_fishnet_daily_processed_sf.rds"))
st_write(fishnet_spatial, paste0(processed_data, "mn_fishnet_daily_processed.gpkg"),
         delete_dsn = TRUE)
```
