---
title: "Within-window effect timing (weekly non-overlapping) and power analysis"
author: "Fintan Mooney"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

```{r initiate}
rm(list = ls())
# 0a.Declare root directory, folder location and load essential stuff
project.folder = paste0(print(here::here()), '/')
source(paste0(project.folder, 'create_folder_structure.R'))
source(paste0(functions.folder, 'script_initiate.R'))
library(MASS)
```

# 1. Load case-crossover dataframe

```{r load}
case_crossover_df <- readRDS(paste0(objects_folder, "case_crossover_df.RDS"))
setDT(case_crossover_df)
N_cases <- sum(case_crossover_df$outbreak_binary == 1)
cat("Case-zone-days in primary IND+INR analysis: ", N_cases, "\n")
```

# 2. Build crossbases (primary spec)

```{r crossbases}
cb_runoff <- crossbasis(
  as.matrix(case_crossover_df[, paste0("runoff_Lag", 0:28), with = FALSE]),
  lag = c(0, 28), argvar = list(fun = "lin"), arglag = list(fun = "ns", df = 3)
)
cb_soil_moisture <- crossbasis(
  as.matrix(case_crossover_df[, paste0("soil_moisture_Lag", 0:28), with = FALSE]),
  lag = c(0, 28), argvar = list(fun = "lin"), arglag = list(fun = "ns", df = 3)
)
cb_temperature <- crossbasis(
  as.matrix(case_crossover_df[, paste0("temperature_Lag", 0:28), with = FALSE]),
  lag = c(0, 28), argvar = list(fun = "lin"), arglag = list(fun = "ns", df = 3)
)
cb_precipitation <- crossbasis(
  as.matrix(case_crossover_df[, paste0("precipitation_Lag", 0:28), with = FALSE]),
  lag = c(0, 28), argvar = list(fun = "lin"), arglag = list(fun = "ns", df = 3)
)
cb_wind_speed <- crossbasis(
  as.matrix(case_crossover_df[, paste0("wind_speed_Lag", 0:28), with = FALSE]),
  lag = c(0, 28), argvar = list(fun = "lin"), arglag = list(fun = "ns", df = 3)
)
cb_anseriformes <- crossbasis(
  as.matrix(case_crossover_df[, paste0("anseriformes_Lag", 0:28), with = FALSE]),
  lag = c(0, 28), argvar = list(fun = "lin"),
  arglag = list(fun = "strata", breaks = c(7, 14, 21))
)
```

# 3. Frequentist clogit and Bayesian posterior samples

```{r fit-and-posterior}
freq_model <- clogit(
  outbreak_binary ~
    cb_runoff + cb_soil_moisture + cb_temperature +
    cb_precipitation + cb_wind_speed + cb_anseriformes +
    strata(stratum_id),
  method = "efron", data = case_crossover_df
)

beta_hat <- coef(freq_model)
V_hat    <- vcov(freq_model)

set.seed(2026)
post_flat <- MASS::mvrnorm(10000, beta_hat, V_hat)
colnames(post_flat) <- names(beta_hat)

# Weakly informative N(0, 0.3) prior
prior_sd <- 0.3
V_prior  <- diag(prior_sd^2, nrow = length(beta_hat))
V_post   <- solve(solve(V_prior) + solve(V_hat))
mu_post  <- as.vector(V_post %*% solve(V_hat) %*% beta_hat)

set.seed(2026)
post_info <- MASS::mvrnorm(10000, mu_post, V_post)
colnames(post_info) <- names(beta_hat)
```

# 4. Non-overlapping weekly-window contrast vectors

```{r contrasts}
lag_seq        <- 0:28
lag_basis_met  <- do.call(dlnm::onebasis,
                          c(list(x = lag_seq), attr(cb_runoff,       "arglag")))
lag_basis_bird <- do.call(dlnm::onebasis,
                          c(list(x = lag_seq), attr(cb_anseriformes, "arglag")))

# Non-overlapping windows — partition the 0-28 d range into four week-long bins
# (matching the existing weekly-window convention in c_00).
build_contrast_window <- function(basis, lo, hi) {
  0.5 * colSums(basis[lag_seq >= lo & lag_seq <= hi, , drop = FALSE])
}

windows <- list(
  "0–7 d"    = c(0,  7),
  "8–14 d"   = c(8,  14),
  "15–21 d"  = c(15, 21),
  "22–28 d"  = c(22, 28)
)

contrasts_met  <- lapply(windows, function(w) build_contrast_window(lag_basis_met,  w[1], w[2]))
contrasts_bird <- lapply(windows, function(w) build_contrast_window(lag_basis_bird, w[1], w[2]))
```

# 5. Within-window OR table (frequentist + Bayesian flat + Bayesian informative)

```{r weekly-window-table}
predictor_specs <- list(
  "Runoff"        = list(prefix = "cb_runoff",        is_bird = FALSE),
  "Soil Moisture" = list(prefix = "cb_soil_moisture", is_bird = FALSE),
  "Temperature"   = list(prefix = "cb_temperature",   is_bird = FALSE),
  "Precipitation" = list(prefix = "cb_precipitation", is_bird = FALSE),
  "Wind Speed"    = list(prefix = "cb_wind_speed",    is_bird = FALSE),
  "Anseriformes"  = list(prefix = "cb_anseriformes",  is_bird = TRUE)
)

summarize_window <- function(prefix, contrast) {
  cols <- grep(paste0("^", prefix, "v"), names(beta_hat))
  b    <- beta_hat[cols]
  V    <- V_hat[cols, cols]

  log_or_freq <- as.numeric(contrast %*% b)
  se          <- sqrt(as.numeric(t(contrast) %*% V %*% contrast))

  log_or_flat <- as.numeric(post_flat[, cols, drop = FALSE] %*% contrast)
  log_or_info <- as.numeric(post_info[, cols, drop = FALSE] %*% contrast)

  list(
    freq_or   = exp(log_or_freq),
    freq_lo   = exp(log_or_freq - 1.96 * se),
    freq_hi   = exp(log_or_freq + 1.96 * se),
    se        = se,
    flat_med  = median(exp(log_or_flat)),
    flat_lo   = unname(quantile(exp(log_or_flat), 0.025)),
    flat_hi   = unname(quantile(exp(log_or_flat), 0.975)),
    flat_ppos = mean(log_or_flat > 0),
    info_med  = median(exp(log_or_info)),
    info_lo   = unname(quantile(exp(log_or_info), 0.025)),
    info_hi   = unname(quantile(exp(log_or_info), 0.975)),
    info_ppos = mean(log_or_info > 0)
  )
}

weekly_tbl <- rbindlist(lapply(names(predictor_specs), function(name) {
  spec <- predictor_specs[[name]]
  rbindlist(lapply(seq_along(windows), function(i) {
    contrast <- if (spec$is_bird) contrasts_bird[[i]] else contrasts_met[[i]]
    s <- summarize_window(spec$prefix, contrast)
    data.table(
      Predictor                = name,
      Window                   = names(windows)[i],
      `Freq OR (95% CI)`       = sprintf("%.2f (%.2f, %.2f)", s$freq_or, s$freq_lo, s$freq_hi),
      `Bayes flat OR (95% CrI)`= sprintf("%.2f (%.2f, %.2f)", s$flat_med, s$flat_lo, s$flat_hi),
      `P(>0) flat`             = sprintf("%.3f", s$flat_ppos),
      `Bayes info OR (95% CrI)`= sprintf("%.2f (%.2f, %.2f)", s$info_med, s$info_lo, s$info_hi),
      `P(>0) info`             = sprintf("%.3f", s$info_ppos)
    )
  }))
}))

print(weekly_tbl)

ft_wk <- weekly_tbl %>%
  flextable() %>%
  fontsize(size = 9, part = "all") %>%
  autofit() %>%
  align(align = "center", part = "all") %>%
  theme_booktabs()

doc <- read_docx() %>%
  body_add_flextable(ft_wk) %>%
  body_add_par(
    "Within-window cumulative OR estimates for the primary 6-variable case-crossover DLNM. Windows are non-overlapping weekly bins (0-7, 8-14, 15-21, 22-28 d before the case day). For each predictor and window, the cumulative effect is computed by integrating the lag-response over the window at a constant +0.5 SD exposure. Frequentist OR with 95% Wald CI; Bayesian medians with 95% credible intervals under (a) flat prior (asymptotic MVN approximation) and (b) weakly informative N(0, 0.3) prior on each crossbasis coefficient. P(>0) is the posterior probability that the cumulative log-OR within the window exceeds zero.",
    style = "Normal"
  )
print(doc, target = paste0(tables_sensitivity_folder, "Table_Weekly_Window_Bayesian.docx"))
```

# 5b. Wide-format Bayesian info-prior table (windows as columns)

```{r weekly-window-wide-info}
# Combine OR (95% CrI) with P(>0) into a single cell per window, then pivot wide.
weekly_wide_info <- copy(weekly_tbl)
weekly_wide_info[, cell := paste0(`Bayes info OR (95% CrI)`, ", P=", `P(>0) info`)]
weekly_wide_info <- dcast(weekly_wide_info, Predictor ~ Window, value.var = "cell")
setcolorder(weekly_wide_info, c("Predictor", "0–7 d", "8–14 d", "15–21 d", "22–28 d"))

print(weekly_wide_info)

ft_wk_info <- weekly_wide_info %>%
  flextable() %>%
  fontsize(size = 10, part = "all") %>%
  autofit() %>%
  align(align = "center", part = "all") %>%
  theme_booktabs()

doc <- read_docx() %>%
  body_add_flextable(ft_wk_info) %>%
  body_add_par(
    "Bayesian posterior summaries for within-window cumulative effects (per +0.5 SD) from the time-stratified daily case-crossover DLNM under a weakly informative N(0, 0.3) prior on each crossbasis coefficient. Cells report posterior median odds ratio with 95% credible interval and the posterior probability that the cumulative log-OR within the window exceeds zero. P > 0.95 or P < 0.05 indicates strong directional evidence; 0.90 < P < 0.95 or 0.05 < P < 0.10 indicates moderate directional evidence.",
    style = "Normal"
  )
print(doc, target = paste0(tables_sensitivity_folder, "Table_Weekly_Window_Info_Wide.docx"))
```

# 6. Lag-response slice plot (per-day OR at +0.5 SD with credible band)

```{r slice-plot, fig.width = 12, fig.height = 8}
# Per-day cumulative log-OR at +0.5 SD: each lag day k contributes 0.5 × basis(k) %*% beta
# Use posterior samples to get a 95% credible band on the lag-response curve.
draw_slice <- function(prefix, basis, post_samples, ylim_fix = NULL) {
  cols <- grep(paste0("^", prefix, "v"), colnames(post_samples))
  # For each lag k, log-OR contribution = 0.5 × basis(k) %*% beta
  log_or_by_lag <- 0.5 * basis %*% t(post_samples[, cols, drop = FALSE])
  or_med <- exp(apply(log_or_by_lag, 1, median))
  or_lo  <- exp(apply(log_or_by_lag, 1, quantile, 0.025))
  or_hi  <- exp(apply(log_or_by_lag, 1, quantile, 0.975))
  list(med = or_med, lo = or_lo, hi = or_hi)
}

png(paste0(figures_sensitivity_folder, "figure_lag_response_bayesian.png"),
    width = 12, height = 8, units = "in", res = 300)
par(mfrow = c(2, 3), mar = c(4, 4, 2.5, 1), oma = c(2, 2, 0, 1),
    cex.main = 1.1, cex.lab = 1.0, cex.axis = 0.9)

for (name in names(predictor_specs)) {
  spec  <- predictor_specs[[name]]
  basis <- if (spec$is_bird) lag_basis_bird else lag_basis_met
  s     <- draw_slice(spec$prefix, basis, post_flat)
  plot(lag_seq, s$med, type = "l", lwd = 1.5, col = "black",
       ylim = range(c(s$lo, s$hi)),
       main = name, xlab = "Lag (days)", ylab = "OR (95% CrI) per +0.5 SD",
       las = 1)
  polygon(c(lag_seq, rev(lag_seq)), c(s$lo, rev(s$hi)),
          col = rgb(0, 0, 0, 0.15), border = NA)
  abline(h = 1, lty = 2, col = "gray40")
  for (xv in c(7, 14, 21)) abline(v = xv, lty = 3, col = "gray70")
}

mtext("Lag-response slices with Bayesian flat-prior 95% credible bands",
      side = 3, outer = TRUE, line = -1.5, cex = 1.1)
dev.off()
```

# 7. Power analysis at the cumulative ≤28 d window

```{r power-analysis}
# Detectability at α = 0.05 (two-sided) and 80% power requires:
#   |log-OR| ≥ (z_{α/2} + z_β) × SE = (1.96 + 0.84) × SE = 2.80 × SE
# MDE at current N: exp(2.80 × SE_observed)
# N required to detect target OR_t (assuming SE scales as 1/sqrt(N)):
#   N_target = N_current × (2.80 × SE_current / log(OR_t))^2
zsum <- qnorm(0.975) + qnorm(0.80)

# Contrast for the cumulative ≤28 d window
contrast_le28_met  <- 0.5 * colSums(lag_basis_met)
contrast_le28_bird <- 0.5 * colSums(lag_basis_bird)

power_tbl <- rbindlist(lapply(names(predictor_specs), function(name) {
  spec <- predictor_specs[[name]]
  contrast <- if (spec$is_bird) contrast_le28_bird else contrast_le28_met
  cols <- grep(paste0("^", spec$prefix, "v"), names(beta_hat))
  b    <- beta_hat[cols]
  V    <- V_hat[cols, cols]
  log_or_obs <- as.numeric(contrast %*% b)
  se         <- sqrt(as.numeric(t(contrast) %*% V %*% contrast))

  mde_or <- exp(zsum * se)

  n_for_or <- function(target_or) {
    if (target_or <= 1) return(NA_real_)
    N_cases * (zsum * se / log(target_or))^2
  }

  data.table(
    Predictor                          = name,
    `Observed OR (95% CI), ≤28 d`      = sprintf("%.2f (%.2f, %.2f)",
                                                exp(log_or_obs),
                                                exp(log_or_obs - 1.96 * se),
                                                exp(log_or_obs + 1.96 * se)),
    `MDE at current N (α=0.05, 80%)`   = sprintf("%.2f", mde_or),
    `N to detect OR=1.5`               = sprintf("%.0f", n_for_or(1.5)),
    `N to detect OR=2.0`               = sprintf("%.0f", n_for_or(2.0)),
    `N to detect OR=2.5`               = sprintf("%.0f", n_for_or(2.5)),
    `N to detect OR=3.0`               = sprintf("%.0f", n_for_or(3.0))
  )
}))

print(power_tbl)

ft_pwr <- power_tbl %>%
  flextable() %>%
  fontsize(size = 10, part = "all") %>%
  autofit() %>%
  align(align = "center", part = "all") %>%
  theme_booktabs()

doc <- read_docx() %>%
  body_add_flextable(ft_pwr) %>%
  body_add_par(
    sprintf("Post-hoc power analysis at the cumulative ≤28 d window (current N = %d case-zone-days). MDE: minimum detectable cumulative OR at α = 0.05 (two-sided) with 80%% power. N columns: number of case-zone-days required to detect the indicated target OR with 80%% power, assuming the predictor SE scales as 1/sqrt(N). A standard-error inflation factor of (1.96 + 0.84) = 2.80 is used for the α = 0.05, 80%% power criterion.",
            N_cases),
    style = "Normal"
  )
print(doc, target = paste0(tables_sensitivity_folder, "Table_Power_Analysis.docx"))
```

# 8. Power scenarios across plausible multi-state sample sizes

```{r power-scenarios}
scenarios <- c(
  "MN only (current)"     = N_cases,
  "MN + IA"               = round(N_cases * 1.8),
  "Miss Flyway (MN+IA+WI)"= round(N_cases * 2.9),
  "Top 5 HPAI states"     = round(N_cases * 4.0),
  "All CONUS 2022"        = round(N_cases * 6.0)
)

target_ors <- c(1.5, 2.0, 2.5, 3.0)

power_curves <- rbindlist(lapply(names(predictor_specs), function(name) {
  spec <- predictor_specs[[name]]
  contrast <- if (spec$is_bird) contrast_le28_bird else contrast_le28_met
  cols <- grep(paste0("^", spec$prefix, "v"), names(beta_hat))
  V    <- V_hat[cols, cols]
  se_current <- sqrt(as.numeric(t(contrast) %*% V %*% contrast))

  rbindlist(lapply(names(scenarios), function(sc) {
    n_target <- scenarios[sc]
    se_target <- se_current * sqrt(N_cases / n_target)
    powers <- sapply(target_ors, function(or_t) {
      delta <- log(or_t)
      pnorm(delta / se_target - qnorm(0.975))
    })
    data.table(
      Predictor      = name,
      Scenario       = sprintf("%s (N=%d)", sc, n_target),
      `Power OR=1.5` = sprintf("%.2f", powers[1]),
      `Power OR=2.0` = sprintf("%.2f", powers[2]),
      `Power OR=2.5` = sprintf("%.2f", powers[3]),
      `Power OR=3.0` = sprintf("%.2f", powers[4])
    )
  }))
}))

print(power_curves)

ft_pc <- power_curves %>%
  flextable() %>%
  fontsize(size = 9, part = "all") %>%
  autofit() %>%
  align(align = "center", part = "all") %>%
  theme_booktabs()

doc <- read_docx() %>%
  body_add_flextable(ft_pc) %>%
  body_add_par(
    "Power projections across plausible multi-state sample-size expansions. Power computed at α = 0.05 (two-sided) for cumulative ≤28 d OR detection, assuming the per-predictor SE scales as 1/sqrt(N) and the underlying effect size matches the target OR. Multi-state event counts are projections based on USDA APHIS 2022 commercial poultry HPAI outbreak data.",
    style = "Normal"
  )
print(doc, target = paste0(tables_sensitivity_folder, "Table_Power_Scenarios.docx"))
```

# 9. Save bundle

```{r save}
saveRDS(list(
  weekly_tbl   = weekly_tbl,
  power_tbl    = power_tbl,
  power_curves = power_curves,
  N_cases      = N_cases,
  scenarios    = scenarios
), paste0(objects_folder, "weekly_window_and_power.RDS"))
```
