---
title: "Sensitivity: new ERA5 variables — testing strategy for the daily primary"
author: "Fintan Mooney"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

```{r initiate}
rm(list = ls())
project.folder = paste0(print(here::here()), '/')
source(paste0(project.folder, 'create_folder_structure.R'))
source(paste0(functions.folder, 'script_initiate.R'))
```

# Testing strategy

The Nov 2025 ERA5-Land refresh added five candidate met variables to the
case-crossover dataframe alongside `snow_depth` (which is already in the
primary model after replacing `snow_density`):

- `snow_cover` (%)            — threshold version of snow_depth
- `skin_temperature` (K)      — land-surface temp, frozen-water proxy
- `soil_temperature` (K)      — 0–7 cm soil temp, frost vs thaw
- `soil_moisture_2` (m³/m³)   — 7–28 cm soil moisture (depth sensitivity)
- `snowfall` (m water-eq/day) — daily fresh-snow accumulation

We evaluate each in a 4-tier sequence:

- **Tier 0**: descriptive — pairwise correlations with existing predictors.
- **Tier 1**: univariate screen — does the variable carry signal at all when paired only with anseriformes + stratum?
- **Tier 2**: swap tests — does the new variable improve the primary when it
  *replaces* a correlated existing predictor (skin↔temp, soil_moisture_2↔soil_moisture, snow_cover↔snow_depth)?
- **Tier 3**: addition tests — does adding the new variable to the primary improve fit (LRT + ΔAIC)? Reports change in *other* predictors' coefficients.
- **Tier 4**: pre-registered biological hypotheses, Bonferroni-corrected across 3 tests (α/3 = 0.0167):
    1. Surface-freezing concentrates waterfowl on remaining open water → `skin_temperature` cumulative OR at lag 7–14 d ≠ 1
    2. Rain-on-snow drives the runoff signal → primary + `snowfall` shifts the precipitation OR away from primary-only's precipitation OR
    3. Spring soil thaw shifts case timing → `soil_temperature` cumulative OR at lag 14–28 d ≠ 1

# 1. Load case-crossover dataframe

```{r load-data}
cc <- readRDS(paste0(objects_folder, "case_crossover_df.RDS"))
setDT(cc)

MAX_LAG  <- 28
ARGLAG   <- list(fun = "ns", df = 3)
ARGVAR   <- list(fun = "lin")

# Sanity: all new variables present?
new_vars <- c("snow_cover", "skin_temperature", "soil_temperature",
              "soil_moisture_2", "snowfall")
missing  <- new_vars[!sapply(new_vars, function(v)
              all(paste0(v, "_Lag", 0:MAX_LAG) %in% names(cc)))]
if (length(missing) > 0) {
  stop("Missing lag columns for: ", paste(missing, collapse = ", "),
       "\nRe-run a_01 → a_02 → a_03 with the multistate ERA5 path first.")
}

cat("Rows:", nrow(cc), " | cases:", sum(cc$outbreak_binary == 1),
    " | strata:", uniqueN(cc$stratum_id), "\n")
```

# 2. Crossbasis builders (shared across all tiers)

```{r cb-builders}
mk_cb <- function(varname, argvar = ARGVAR, arglag = ARGLAG) {
  X <- as.matrix(cc[, paste0(varname, "_Lag", 0:MAX_LAG), with = FALSE])
  crossbasis(X, lag = c(0, MAX_LAG), argvar = argvar, arglag = arglag)
}
mk_cb_bird <- function() {
  X <- as.matrix(cc[, paste0("anseriformes_Lag", 0:MAX_LAG), with = FALSE])
  crossbasis(X, lag = c(0, MAX_LAG),
             argvar = list(fun = "lin"),
             arglag = list(fun = "strata", breaks = c(7, 14, 21)))
}

# Pre-build the standard crossbases used in primary
cb_bird          <- mk_cb_bird()
cb_runoff        <- mk_cb("runoff")
cb_soil_moisture <- mk_cb("soil_moisture")
cb_temperature   <- mk_cb("temperature")
cb_precipitation <- mk_cb("precipitation")
cb_wind_speed    <- mk_cb("wind_speed")
cb_snow_depth    <- mk_cb("snow_depth")

# New-variable crossbases
cb_snow_cover       <- mk_cb("snow_cover")
cb_skin_temperature <- mk_cb("skin_temperature")
cb_soil_temperature <- mk_cb("soil_temperature")
cb_soil_moisture_2  <- mk_cb("soil_moisture_2")
cb_snowfall         <- mk_cb("snowfall")
```

# 3. Cumulative-OR extractor (manual contrast — robust across model variants)

```{r extractor}
# +0.5 SD cumulative over a contiguous lag window
cum_OR <- function(model, prefix, lag_from = 0, lag_to = MAX_LAG, df_lag = 3) {
  beta <- coef(model); V <- vcov(model)
  cols <- grep(paste0("^", prefix, "v"), names(beta))
  if (length(cols) == 0) return(c(OR = NA, lo = NA, hi = NA))
  b <- beta[cols]; Vsub <- V[cols, cols]
  lag_seq <- 0:MAX_LAG
  lag_basis <- do.call(dlnm::onebasis,
                       c(list(x = lag_seq), list(fun = "ns", df = df_lag)))
  contrast <- 0.5 * colSums(lag_basis[lag_seq >= lag_from & lag_seq <= lag_to, ,
                                       drop = FALSE])
  lo  <- as.numeric(contrast %*% b)
  se  <- sqrt(as.numeric(t(contrast) %*% Vsub %*% contrast))
  c(OR = exp(lo), lo = exp(lo - 1.96 * se), hi = exp(lo + 1.96 * se))
}
fmt_OR <- function(x) sprintf("%.2f (%.2f, %.2f)", x["OR"], x["lo"], x["hi"])
```

# Tier 0 — Correlation with existing predictors

```{r tier-0-correlation}
existing <- c("temperature", "soil_moisture", "snow_depth",
              "precipitation", "wind_speed", "runoff")
all_vars <- c(existing, new_vars)
# Use Lag 0 of each (instantaneous correlations across all zone-days)
m <- as.matrix(cc[, paste0(all_vars, "_Lag0"), with = FALSE])
colnames(m) <- all_vars
corr_mat <- round(cor(m, use = "pairwise.complete.obs"), 2)
print(corr_mat)

# Flag |r| > 0.7 with existing predictors
threshold <- 0.7
flags <- vapply(new_vars, function(nv) {
  paste(existing[abs(corr_mat[nv, existing]) > threshold], collapse = ", ")
}, character(1))
cat("\nNew vars correlated (|r|>", threshold, ") with existing:\n", sep = "")
for (v in new_vars) cat(sprintf("  %-18s : %s\n", v, flags[v]))
```

# Tier 1 — Univariate screen

```{r tier-1-univariate}
univariate_results <- rbindlist(lapply(new_vars, function(v) {
  cb_new <- get(paste0("cb_", v))
  # Build formula on the fly so the prefix matches the cb object name
  assign("cb_target", cb_new, envir = .GlobalEnv)
  m <- clogit(outbreak_binary ~ cb_target + cb_bird + strata(stratum_id),
              data = cc, method = "efron")
  cu <- cum_OR(m, "cb_target")
  data.table(Variable = v,
             AIC      = round(AIC(m), 2),
             Cumulative_OR = fmt_OR(cu))
}))

# Also include a bird-only baseline for AIC comparison
m_baseline <- clogit(outbreak_binary ~ cb_bird + strata(stratum_id),
                     data = cc, method = "efron")
baseline_row <- data.table(Variable = "(bird only — baseline)",
                           AIC = round(AIC(m_baseline), 2),
                           Cumulative_OR = "—")
univariate_results <- rbind(baseline_row, univariate_results)
univariate_results[, dAIC := round(AIC - AIC[Variable == "(bird only — baseline)"], 2)]

print(univariate_results)
```

# Tier 2 — Swap tests

Three biologically-motivated swaps. Each refits the 7-predictor primary with one
predictor replaced and compares AIC.

```{r tier-2-swaps}
# Reference: current primary (snow_depth + temperature + soil_moisture)
m_primary <- clogit(
  outbreak_binary ~ cb_runoff + cb_soil_moisture + cb_temperature +
                    cb_precipitation + cb_wind_speed + cb_snow_depth +
                    cb_bird + strata(stratum_id),
  data = cc, method = "efron")

# Swap 1: skin_temperature replaces temperature
m_swap_skin <- clogit(
  outbreak_binary ~ cb_runoff + cb_soil_moisture + cb_skin_temperature +
                    cb_precipitation + cb_wind_speed + cb_snow_depth +
                    cb_bird + strata(stratum_id),
  data = cc, method = "efron")

# Swap 2: soil_moisture_2 replaces soil_moisture
m_swap_soil <- clogit(
  outbreak_binary ~ cb_runoff + cb_soil_moisture_2 + cb_temperature +
                    cb_precipitation + cb_wind_speed + cb_snow_depth +
                    cb_bird + strata(stratum_id),
  data = cc, method = "efron")

# Swap 3: snow_cover replaces snow_depth
m_swap_cover <- clogit(
  outbreak_binary ~ cb_runoff + cb_soil_moisture + cb_temperature +
                    cb_precipitation + cb_wind_speed + cb_snow_cover +
                    cb_bird + strata(stratum_id),
  data = cc, method = "efron")

swap_tbl <- data.table(
  Spec = c("Primary (snow_depth + temp + soil_moisture)",
           "Swap: skin_temperature ↔ temperature",
           "Swap: soil_moisture_2 ↔ soil_moisture",
           "Swap: snow_cover ↔ snow_depth"),
  AIC  = c(AIC(m_primary), AIC(m_swap_skin), AIC(m_swap_soil), AIC(m_swap_cover)),
  Concordance = c(summary(m_primary)$concordance[1],
                  summary(m_swap_skin)$concordance[1],
                  summary(m_swap_soil)$concordance[1],
                  summary(m_swap_cover)$concordance[1])
)
swap_tbl[, dAIC := round(AIC - AIC[1], 2)]
swap_tbl[, `:=`(AIC = round(AIC, 2), Concordance = round(Concordance, 3))]
print(swap_tbl)

# How does each swap change the cumulative ORs of the other predictors?
swap_or_tbl <- data.table(
  Predictor   = c("Runoff", "Soil Moisture", "Temperature/Skin temp",
                  "Precipitation", "Wind Speed", "Snow Depth/Snow Cover"),
  Primary     = c(fmt_OR(cum_OR(m_primary, "cb_runoff")),
                  fmt_OR(cum_OR(m_primary, "cb_soil_moisture")),
                  fmt_OR(cum_OR(m_primary, "cb_temperature")),
                  fmt_OR(cum_OR(m_primary, "cb_precipitation")),
                  fmt_OR(cum_OR(m_primary, "cb_wind_speed")),
                  fmt_OR(cum_OR(m_primary, "cb_snow_depth"))),
  Swap_skin   = c(fmt_OR(cum_OR(m_swap_skin, "cb_runoff")),
                  fmt_OR(cum_OR(m_swap_skin, "cb_soil_moisture")),
                  fmt_OR(cum_OR(m_swap_skin, "cb_skin_temperature")),
                  fmt_OR(cum_OR(m_swap_skin, "cb_precipitation")),
                  fmt_OR(cum_OR(m_swap_skin, "cb_wind_speed")),
                  fmt_OR(cum_OR(m_swap_skin, "cb_snow_depth"))),
  Swap_soil   = c(fmt_OR(cum_OR(m_swap_soil, "cb_runoff")),
                  fmt_OR(cum_OR(m_swap_soil, "cb_soil_moisture_2")),
                  fmt_OR(cum_OR(m_swap_soil, "cb_temperature")),
                  fmt_OR(cum_OR(m_swap_soil, "cb_precipitation")),
                  fmt_OR(cum_OR(m_swap_soil, "cb_wind_speed")),
                  fmt_OR(cum_OR(m_swap_soil, "cb_snow_depth"))),
  Swap_cover  = c(fmt_OR(cum_OR(m_swap_cover, "cb_runoff")),
                  fmt_OR(cum_OR(m_swap_cover, "cb_soil_moisture")),
                  fmt_OR(cum_OR(m_swap_cover, "cb_temperature")),
                  fmt_OR(cum_OR(m_swap_cover, "cb_precipitation")),
                  fmt_OR(cum_OR(m_swap_cover, "cb_wind_speed")),
                  fmt_OR(cum_OR(m_swap_cover, "cb_snow_cover")))
)
print(swap_or_tbl)
```

# Tier 3 — Addition tests

Two non-substitute additions: `snowfall` (decomposes precip into snow vs rain) and `soil_temperature` (orthogonal info to air temp at the freezing edge).

```{r tier-3-additions}
m_add_snowfall <- clogit(
  outbreak_binary ~ cb_runoff + cb_soil_moisture + cb_temperature +
                    cb_precipitation + cb_wind_speed + cb_snow_depth +
                    cb_snowfall + cb_bird + strata(stratum_id),
  data = cc, method = "efron")

m_add_soil_t <- clogit(
  outbreak_binary ~ cb_runoff + cb_soil_moisture + cb_temperature +
                    cb_precipitation + cb_wind_speed + cb_snow_depth +
                    cb_soil_temperature + cb_bird + strata(stratum_id),
  data = cc, method = "efron")

# Likelihood-ratio tests vs primary
lrt <- function(m_full, m_red) {
  chi <- 2 * (as.numeric(logLik(m_full)) - as.numeric(logLik(m_red)))
  ddf <- attr(logLik(m_full), "df") - attr(logLik(m_red), "df")
  list(chi = chi, df = ddf,
       p = pchisq(chi, df = ddf, lower.tail = FALSE))
}
lrt_snowfall <- lrt(m_add_snowfall, m_primary)
lrt_soil_t   <- lrt(m_add_soil_t,   m_primary)

add_tbl <- data.table(
  Spec        = c("Primary",
                  "+ snowfall (decompose precip)",
                  "+ soil_temperature"),
  AIC         = round(c(AIC(m_primary), AIC(m_add_snowfall), AIC(m_add_soil_t)), 2),
  dAIC        = round(c(0,
                        AIC(m_add_snowfall) - AIC(m_primary),
                        AIC(m_add_soil_t)   - AIC(m_primary)), 2),
  LRT_chisq   = c("—", sprintf("%.2f", lrt_snowfall$chi), sprintf("%.2f", lrt_soil_t$chi)),
  LRT_df      = c("—", lrt_snowfall$df, lrt_soil_t$df),
  LRT_p       = c("—", sprintf("%.4f", lrt_snowfall$p), sprintf("%.4f", lrt_soil_t$p))
)
print(add_tbl)

# How does each addition change the precipitation OR (key for the rain-on-snow story)?
add_or_tbl <- data.table(
  Predictor     = c("Runoff", "Temperature", "Precipitation", "Snow depth",
                    "Snowfall (new)", "Soil temperature (new)"),
  Primary       = c(fmt_OR(cum_OR(m_primary, "cb_runoff")),
                    fmt_OR(cum_OR(m_primary, "cb_temperature")),
                    fmt_OR(cum_OR(m_primary, "cb_precipitation")),
                    fmt_OR(cum_OR(m_primary, "cb_snow_depth")),
                    "—", "—"),
  Plus_snowfall = c(fmt_OR(cum_OR(m_add_snowfall, "cb_runoff")),
                    fmt_OR(cum_OR(m_add_snowfall, "cb_temperature")),
                    fmt_OR(cum_OR(m_add_snowfall, "cb_precipitation")),
                    fmt_OR(cum_OR(m_add_snowfall, "cb_snow_depth")),
                    fmt_OR(cum_OR(m_add_snowfall, "cb_snowfall")),
                    "—"),
  Plus_soil_t   = c(fmt_OR(cum_OR(m_add_soil_t, "cb_runoff")),
                    fmt_OR(cum_OR(m_add_soil_t, "cb_temperature")),
                    fmt_OR(cum_OR(m_add_soil_t, "cb_precipitation")),
                    fmt_OR(cum_OR(m_add_soil_t, "cb_snow_depth")),
                    "—",
                    fmt_OR(cum_OR(m_add_soil_t, "cb_soil_temperature")))
)
print(add_or_tbl)
```

# Tier 4 — Pre-registered biological hypotheses

Bonferroni: α/3 = 0.0167 → 98.3% CIs. We declare reject only if CI excludes 1
at the 98.3% level.

```{r tier-4-pre-registered}
# Wider CI helper for 98.3% intervals
cum_OR_98 <- function(model, prefix, lag_from = 0, lag_to = MAX_LAG, df_lag = 3) {
  beta <- coef(model); V <- vcov(model)
  cols <- grep(paste0("^", prefix, "v"), names(beta))
  b <- beta[cols]; Vsub <- V[cols, cols]
  lag_seq <- 0:MAX_LAG
  lag_basis <- do.call(dlnm::onebasis,
                       c(list(x = lag_seq), list(fun = "ns", df = df_lag)))
  contrast <- 0.5 * colSums(lag_basis[lag_seq >= lag_from & lag_seq <= lag_to, ,
                                       drop = FALSE])
  lo  <- as.numeric(contrast %*% b)
  se  <- sqrt(as.numeric(t(contrast) %*% Vsub %*% contrast))
  z_983 <- qnorm(1 - 0.0167 / 2)   # 98.3% z-multiplier ≈ 2.39
  c(OR = exp(lo), lo = exp(lo - z_983 * se), hi = exp(lo + z_983 * se))
}

# Hypothesis 1: skin_temperature lag 7-14 in a primary-style model (swap variant)
h1 <- cum_OR_98(m_swap_skin, "cb_skin_temperature", lag_from = 7, lag_to = 14)
# Hypothesis 2: snowfall in primary + snowfall model — does adding it shift precip OR?
#   Operationalized as: snowfall's own cumulative OR ≠ 1 (i.e. it carries info)
h2 <- cum_OR_98(m_add_snowfall, "cb_snowfall", lag_from = 0, lag_to = MAX_LAG)
# Hypothesis 3: soil_temperature lag 14-28 in primary + soil_t model
h3 <- cum_OR_98(m_add_soil_t, "cb_soil_temperature", lag_from = 14, lag_to = 28)

prereg_tbl <- data.table(
  Hypothesis = c(
    "H1: skin_temperature lag 7-14 d (open-water concentration)",
    "H2: snowfall cumulative lag 0-28 d (rain-on-snow decomp)",
    "H3: soil_temperature lag 14-28 d (spring thaw)"
  ),
  `Cumulative OR (98.3% CI)` = c(fmt_OR(h1), fmt_OR(h2), fmt_OR(h3)),
  Reject_H0 = c(
    h1["lo"] > 1 | h1["hi"] < 1,
    h2["lo"] > 1 | h2["hi"] < 1,
    h3["lo"] > 1 | h3["hi"] < 1
  )
)
print(prereg_tbl)
```

# Verdict

```{r verdict, echo = FALSE, results = "asis"}
cat("Decision rules (deterministic, declared before running):\n\n")
cat("- **Tier 2 swap adopted** if ΔAIC < -2 (favors new variable) AND concordance not worse by >0.005\n")
cat("- **Tier 3 addition adopted** if LRT p < 0.01 AND ΔAIC < -4\n")
cat("- **Tier 4 hypothesis supported** if 98.3% CI excludes 1\n\n")

# Apply rules
swap_winner <- swap_tbl[dAIC < -2 & Concordance >= max(swap_tbl$Concordance) - 0.005]
adds <- add_tbl[Spec != "Primary" & dAIC < -4 & as.numeric(LRT_p) < 0.01]

cat("**Swap test verdicts:** ",
    if (nrow(swap_winner) == 0) "no swap meets criteria — primary unchanged" else
    paste("adopt", paste(swap_winner$Spec, collapse = "; ")), "\n\n")
cat("**Addition test verdicts:** ",
    if (nrow(adds) == 0) "no addition meets criteria" else
    paste("add", paste(adds$Spec, collapse = "; ")), "\n\n")
cat("**Pre-registered hypothesis verdicts:**\n")
for (i in seq_len(nrow(prereg_tbl))) {
  cat("- ", prereg_tbl$Hypothesis[i], ": ",
      if (prereg_tbl$Reject_H0[i]) "**supported**" else "not supported", "\n")
}
```

# Export supplementary tables

```{r export-tables}
# Save all four output tables to a single docx for supplementary material
doc <- read_docx() %>%
  body_add_par("Sensitivity Analysis: New ERA5 Variables", style = "heading 1") %>%
  body_add_par(paste0("Sample: ", sum(cc$outbreak_binary == 1),
                      " cases across ", uniqueN(cc$stratum_id),
                      " strata. Daily 4-week unidirectional case-crossover, ",
                      "lag 0-28 d, lin × ns(df=3); anseriformes uses strata(7,14,21)."),
               style = "Normal") %>%
  body_add_par("Tier 1: Univariate screen", style = "heading 2") %>%
  body_add_flextable(flextable(univariate_results) %>% autofit() %>%
                       theme_booktabs()) %>%
  body_add_par("Tier 2: Swap-test fit", style = "heading 2") %>%
  body_add_flextable(flextable(swap_tbl) %>% autofit() %>% theme_booktabs()) %>%
  body_add_par("Tier 2: Swap-test cumulative ORs", style = "heading 2") %>%
  body_add_flextable(flextable(swap_or_tbl) %>% autofit() %>% theme_booktabs()) %>%
  body_add_par("Tier 3: Addition-test fit + LRT", style = "heading 2") %>%
  body_add_flextable(flextable(add_tbl) %>% autofit() %>% theme_booktabs()) %>%
  body_add_par("Tier 3: Addition-test cumulative ORs", style = "heading 2") %>%
  body_add_flextable(flextable(add_or_tbl) %>% autofit() %>% theme_booktabs()) %>%
  body_add_par("Tier 4: Pre-registered hypothesis tests (Bonferroni-corrected)",
               style = "heading 2") %>%
  body_add_flextable(flextable(prereg_tbl) %>% autofit() %>% theme_booktabs())

print(doc, target = paste0(tables_sensitivity_folder, "Table_New_Variable_Sensitivity.docx"))
cat("Saved: ", tables_sensitivity_folder, "Table_New_Variable_Sensitivity.docx\n", sep = "")
```

# Save full bundle

```{r save-bundle}
saveRDS(list(
  univariate_results = univariate_results,
  swap_tbl           = swap_tbl,
  swap_or_tbl        = swap_or_tbl,
  add_tbl            = add_tbl,
  add_or_tbl         = add_or_tbl,
  prereg_tbl         = prereg_tbl,
  corr_mat           = corr_mat,
  models = list(primary      = m_primary,
                swap_skin    = m_swap_skin,
                swap_soil    = m_swap_soil,
                swap_cover   = m_swap_cover,
                add_snowfall = m_add_snowfall,
                add_soil_t   = m_add_soil_t)
), paste0(objects_folder, "new_variable_sensitivity.RDS"))
```
