---
title: "Sensitivity: poultry-density × runoff interaction (1:4 unidirectional weekly)"
author: "Fintan Mooney"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

```{r initiate}
rm(list = ls())
project.folder = paste0(print(here::here()), '/')
source(paste0(project.folder, 'create_folder_structure.R'))
source(paste0(functions.folder, 'script_initiate.R'))
```

# Hypothesis (same as c_26)

If runoff is the spillover mechanism, the runoff effect should amplify in
zones with higher poultry density. We replicate the daily interaction test
on the **1:4 unidirectional weekly** model (1:4 unidirectional, CSLT excluded) to
confirm the pattern holds at the weekly resolution. The 1:4 unidirectional case-crossover has
the runoff cumulative HR finding that the manuscript's discussion section
references (HR 4.59 [1.28, 16.46]), so this is the design where the
interaction matters most.

# 1. Load weekly bundle and daily case-crossover (for birds_au)

```{r load}
ab        <- readRDS(paste0(objects_folder, "weekly_primary_ab.RDS"))
cc_A      <- copy(ab$cc_A); setDT(cc_A)
m_primary <- ab$model_A

# Detect snow variable (cc_A may have snow_depth_* or snow_cover_* depending on
# whether c_20 has been re-run with the snow_cover swap). Prefer snow_cover.
snow_var <- if (any(grepl("^snow_cover_Lag", names(cc_A)))) "snow_cover" else
            if (any(grepl("^snow_depth_Lag", names(cc_A)))) "snow_depth" else
            stop("Neither snow_cover_Lag* nor snow_depth_Lag* found in cc_A — re-run c_20 first")
cat("Snow variable in cc_A: ", snow_var, "\n", sep = "")

# birds_au is zone-fixed; load from the full daily fishnet (has all zones,
# not just case-crossover cells). Falls back to the daily case-crossover df if
# the processed fishnet is unavailable.
if (file.exists("03_output/3b_model_output/dataframes/mn_fishnet_daily_processed.rds")) {
  full <- readRDS("03_output/3b_model_output/dataframes/mn_fishnet_daily_processed.rds")
  setDT(full)
  zone_birds <- unique(full[, .(zone_id, birds_au)])
} else {
  daily_cc <- readRDS(paste0(objects_folder, "case_crossover_df.RDS"))
  setDT(daily_cc)
  zone_birds <- unique(daily_cc[, .(zone_id, birds_au)])
}

cc_A <- merge(cc_A, zone_birds, by = "zone_id", all.x = TRUE)
# Any zones still missing birds_au get 0 (these are typically cells with no
# registered feedlots — biologically the "no poultry density" baseline)
n_na <- sum(is.na(cc_A$birds_au))
if (n_na > 0) {
  cc_A[is.na(birds_au), birds_au := 0]
  cat("Filled ", n_na, " NA birds_au cells with 0 (no registered feedlots)\n", sep = "")
}
cat("Weekly 1:4 unidirectional rows: ", nrow(cc_A),
    " | cases: ", sum(cc_A$outbreak_binary == 1),
    " | strata: ", uniqueN(cc_A$stratum_id), "\n", sep = "")
cat("birds_au range: [", paste(round(range(cc_A$birds_au)), collapse=", "), "]\n", sep = "")
```

# 2. Standardise poultry density (log+z, weekly scale)

```{r standardise-birds-au}
cc_A[, birds_au_z := as.numeric(scale(log1p(birds_au)))]
print(summary(cc_A$birds_au_z))
```

# 3. Rebuild weekly crossbases (matching primary spec)

```{r weekly-cb}
MAX_LAG_WK <- 4
ARGLAG_WK  <- list(fun = "ns", df = 2)
ARGVAR_WK  <- list(fun = "lin")

mk_cb_wk <- function(varname) {
  X <- as.matrix(cc_A[, paste0(varname, "_Lag", 0:MAX_LAG_WK), with = FALSE])
  crossbasis(X, lag = c(0, MAX_LAG_WK), argvar = ARGVAR_WK, arglag = ARGLAG_WK)
}
cb_runoff_A         <- mk_cb_wk("runoff")
cb_soil_moisture_A  <- mk_cb_wk("soil_moisture")
cb_temperature_A    <- mk_cb_wk("temperature")
cb_precipitation_A  <- mk_cb_wk("precipitation")
cb_wind_speed_A     <- mk_cb_wk("wind_speed")
cb_snow_A           <- mk_cb_wk(snow_var)              # snow_cover or snow_depth, whichever is present
cb_anseriformes_A   <- mk_cb_wk("anseriformes")
```

# 4. Build interaction crossbasis (weekly runoff × birds_au_z)

```{r interaction-cb-weekly}
runoff_mat_wk    <- as.matrix(cc_A[, paste0("runoff_Lag", 0:MAX_LAG_WK), with = FALSE])
runoff_x_birds   <- runoff_mat_wk * cc_A$birds_au_z
cb_runoff_x_birds_A <- crossbasis(runoff_x_birds,
                                   lag    = c(0, MAX_LAG_WK),
                                   argvar = ARGVAR_WK,
                                   arglag = ARGLAG_WK)
```

# 5. Refit primary + interaction

```{r fit-weekly}
# Refit primary on the freshly-rebuilt crossbases (so coef names match below)
m_primary_wk <- clogit(
  outbreak_binary ~ cb_runoff_A + cb_soil_moisture_A + cb_temperature_A +
                    cb_precipitation_A + cb_wind_speed_A + cb_snow_A +
                    cb_anseriformes_A + strata(stratum_id),
  data = cc_A, method = "efron")

m_inter_wk <- clogit(
  outbreak_binary ~ cb_runoff_A + cb_runoff_x_birds_A +
                    cb_soil_moisture_A + cb_temperature_A +
                    cb_precipitation_A + cb_wind_speed_A + cb_snow_A +
                    cb_anseriformes_A + strata(stratum_id),
  data = cc_A, method = "efron")

cat("\n--- Fit comparison (1:4 unidirectional weekly) ---\n")
cat(sprintf("Primary       AIC: %.2f\n", AIC(m_primary_wk)))
cat(sprintf("+ Interaction AIC: %.2f  (ΔAIC = %+.2f)\n",
            AIC(m_inter_wk), AIC(m_inter_wk) - AIC(m_primary_wk)))
chi <- 2 * (as.numeric(logLik(m_inter_wk)) - as.numeric(logLik(m_primary_wk)))
ddf <- attr(logLik(m_inter_wk), "df") - attr(logLik(m_primary_wk), "df")
p_lrt <- pchisq(chi, df = ddf, lower.tail = FALSE)
cat(sprintf("LRT (interaction): chi^2 = %.2f, df = %d, p = %.4f\n",
            chi, ddf, p_lrt))
```

# 6. Cumulative + lag-specific runoff OR by density

```{r weekly-or-by-density}
beta <- coef(m_inter_wk); V <- vcov(m_inter_wk)
i_main <- grep("^cb_runoff_Av",            names(beta))
i_intx <- grep("^cb_runoff_x_birds_Av",    names(beta))
b_main <- beta[i_main]; b_intx <- beta[i_intx]
V_mm   <- V[i_main, i_main]; V_ii <- V[i_intx, i_intx]; V_mi <- V[i_main, i_intx]

lag_basis_wk <- do.call(dlnm::onebasis,
                        c(list(x = 0:MAX_LAG_WK), ARGLAG_WK))

or_at_wk <- function(contrast, ba_z) {
  lo  <- as.numeric(contrast %*% (b_main + ba_z * b_intx))
  var <- as.numeric(t(contrast) %*% (V_mm + 2 * ba_z * V_mi + ba_z^2 * V_ii) %*% contrast)
  se  <- sqrt(var)
  c(OR = exp(lo), lo = exp(lo - 1.96*se), hi = exp(lo + 1.96*se))
}

ba_low    <- quantile(cc_A$birds_au_z, 0.10)
ba_median <- quantile(cc_A$birds_au_z, 0.50)
ba_high   <- quantile(cc_A$birds_au_z, 0.90)
ba_grid_raw <- exp(c(ba_low, ba_median, ba_high) * sd(log1p(cc_A$birds_au), na.rm=TRUE) +
                   mean(log1p(cc_A$birds_au), na.rm=TRUE)) - 1

# 6a. Cumulative OR (weekly lag 0-4 = days 0-28)
contrast_cum_wk <- 0.5 * colSums(lag_basis_wk)
cum_tbl_wk <- data.table(
  Density_label = c("Low (10th pctl)", "Median (50th pctl)", "High (90th pctl)"),
  birds_au_raw  = round(ba_grid_raw, 1),
  Cumulative_runoff_OR_95CI = c(
    sprintf("%.2f (%.2f, %.2f)", or_at_wk(contrast_cum_wk, ba_low)["OR"],
            or_at_wk(contrast_cum_wk, ba_low)["lo"], or_at_wk(contrast_cum_wk, ba_low)["hi"]),
    sprintf("%.2f (%.2f, %.2f)", or_at_wk(contrast_cum_wk, ba_median)["OR"],
            or_at_wk(contrast_cum_wk, ba_median)["lo"], or_at_wk(contrast_cum_wk, ba_median)["hi"]),
    sprintf("%.2f (%.2f, %.2f)", or_at_wk(contrast_cum_wk, ba_high)["OR"],
            or_at_wk(contrast_cum_wk, ba_high)["lo"], or_at_wk(contrast_cum_wk, ba_high)["hi"])
  )
)
cat("\n--- Cumulative weekly runoff OR over lag 0-4 wk by poultry density ---\n")
print(cum_tbl_wk)

# 6b. Lag-specific OR at each weekly lag point (5 points = 0, 7, 14, 21, 28 d)
lag_specific_wk <- rbindlist(lapply(0:MAX_LAG_WK, function(k) {
  c_k <- 0.5 * lag_basis_wk[k + 1, ]
  data.table(
    lag_days       = k * 7,
    Low_OR_95CI    = sprintf("%.2f (%.2f, %.2f)", or_at_wk(c_k, ba_low)["OR"],
                              or_at_wk(c_k, ba_low)["lo"], or_at_wk(c_k, ba_low)["hi"]),
    Median_OR_95CI = sprintf("%.2f (%.2f, %.2f)", or_at_wk(c_k, ba_median)["OR"],
                              or_at_wk(c_k, ba_median)["lo"], or_at_wk(c_k, ba_median)["hi"]),
    High_OR_95CI   = sprintf("%.2f (%.2f, %.2f)", or_at_wk(c_k, ba_high)["OR"],
                              or_at_wk(c_k, ba_high)["lo"], or_at_wk(c_k, ba_high)["hi"])
  )
}))
cat("\n--- Lag-specific weekly runoff OR (per +0.5 SD on that week) ---\n")
print(lag_specific_wk)

# Numeric long form for plotting
lag_specific_wk_long <- rbindlist(lapply(0:MAX_LAG_WK, function(k) {
  c_k <- 0.5 * lag_basis_wk[k + 1, ]
  rbindlist(lapply(list(Low = ba_low, Median = ba_median, High = ba_high), function(z) {
    o <- or_at_wk(c_k, z)
    data.table(lag_days = k * 7, OR = o["OR"], OR_lo = o["lo"], OR_hi = o["hi"])
  }), idcol = "Density")
}))
lag_specific_wk_long[, Density := factor(Density, levels = c("Low", "Median", "High"))]
```

# 7. Verdict

```{r verdict-weekly}
verdict <- if (p_lrt < 0.05 && AIC(m_inter_wk) - AIC(m_primary_wk) < -2) {
  "SUPPORTED — interaction is jointly significant (LRT p<0.05) and improves AIC by >2"
} else if (p_lrt < 0.05) {
  "PARTIAL — LRT significant but ΔAIC does not exceed -2"
} else {
  "NOT SUPPORTED — interaction is not jointly significant at α=0.05"
}
cat("Decision rule: LRT p < 0.05 AND ΔAIC < -2 → adopt interaction\n")
cat("Verdict: ", verdict, "\n", sep = "")
```

# 8. Lag-specific figure (weekly)

```{r lag-specific-figure-wk, fig.width = 10, fig.height = 5}
p_wk <- ggplot(lag_specific_wk_long, aes(x = lag_days, y = OR,
                                          color = Density, fill = Density)) +
  geom_hline(yintercept = 1, linetype = "dashed", color = "gray40") +
  geom_ribbon(aes(ymin = OR_lo, ymax = OR_hi), alpha = 0.18, color = NA) +
  geom_line(linewidth = 0.9) +
  geom_point(size = 2.2) +
  scale_y_log10() +
  scale_x_continuous(breaks = c(0, 7, 14, 21, 28)) +
  scale_color_manual(values = c("Low" = "steelblue",
                                "Median" = "gray40",
                                "High" = "firebrick")) +
  scale_fill_manual( values = c("Low" = "steelblue",
                                "Median" = "gray40",
                                "High" = "firebrick")) +
  labs(
    x = "Lag (days)",
    y = "Weekly runoff OR (95% CI), per +0.5 SD on that lag week",
    color = "Poultry density", fill = "Poultry density",
    title    = "Lag-specific runoff OR by poultry density — 1:4 unidirectional weekly",
    subtitle = "Weekly lin × ns(df=2); density = standardised log birds_au at 10th/50th/90th pctl"
  ) +
  theme_minimal(base_size = 12) +
  theme(panel.grid.minor = element_blank(), legend.position = "top",
        strip.text = element_text(face = "bold"))
ggsave(paste0(figures_sensitivity_folder, "figure_feedlot_runoff_lag_specific_weekly.png"),
       p_wk, width = 10, height = 5, units = "in", dpi = 300)
print(p_wk)
```

# 9. Combined daily + weekly lag-specific figure (if daily bundle is on disk)

```{r combined-figure, fig.width = 12, fig.height = 6}
daily_bundle_path <- paste0(objects_folder, "feedlot_interaction_daily.RDS")
if (file.exists(daily_bundle_path)) {
  d_bundle <- readRDS(daily_bundle_path)
  d_long   <- d_bundle$lag_specific_numeric
  d_long[, Resolution := "Daily"]
  w_long <- copy(lag_specific_wk_long); w_long[, Resolution := "Weekly"]
  combined <- rbind(d_long, w_long, fill = TRUE)
  combined[, Density := factor(Density, levels = c("Low", "Median", "High"))]

  p_combo <- ggplot(combined, aes(x = lag_days, y = OR,
                                  color = Density, fill = Density)) +
    geom_hline(yintercept = 1, linetype = "dashed", color = "gray40") +
    geom_ribbon(aes(ymin = OR_lo, ymax = OR_hi), alpha = 0.15, color = NA) +
    geom_line(linewidth = 0.9) +
    facet_wrap(~ Resolution, scales = "free_y") +
    scale_y_log10() +
    scale_x_continuous(breaks = c(0, 7, 14, 21, 28)) +
    scale_color_manual(values = c("Low" = "steelblue",
                                  "Median" = "gray40",
                                  "High" = "firebrick")) +
    scale_fill_manual( values = c("Low" = "steelblue",
                                  "Median" = "gray40",
                                  "High" = "firebrick")) +
    labs(
      x = "Lag (days)",
      y = "Runoff OR (95% CI), per +0.5 SD",
      color = "Poultry density", fill = "Poultry density",
      title    = "Runoff × poultry-density interaction — daily vs weekly",
      subtitle = "Lag-specific OR at three poultry-density levels; lines = point estimate, ribbons = 95% CI"
    ) +
    theme_minimal(base_size = 12) +
    theme(panel.grid.minor = element_blank(), legend.position = "top",
          strip.text = element_text(face = "bold"))
  ggsave(paste0(figures_sensitivity_folder, "figure_feedlot_runoff_lag_specific_combined.png"),
         p_combo, width = 12, height = 6, units = "in", dpi = 300)
  print(p_combo)
} else {
  cat("Daily bundle not found at ", daily_bundle_path,
      " — run c_26 first to enable the combined figure.\n", sep = "")
}
```

# 10. Export tables

```{r export-wk}
fit_tbl_wk <- data.table(
  Spec    = c("Primary (snow_cover, 1:4 unidirectional weekly)",
              "+ runoff × poultry-density interaction"),
  AIC     = c(round(AIC(m_primary_wk), 2), round(AIC(m_inter_wk), 2)),
  dAIC    = c(0, round(AIC(m_inter_wk) - AIC(m_primary_wk), 2)),
  LRT_chi = c("—", sprintf("%.2f", chi)),
  LRT_df  = c("—", as.character(ddf)),
  LRT_p   = c("—", sprintf("%.4f", p_lrt))
)

doc <- read_docx() %>%
  body_add_par("Sensitivity Analysis: Poultry-density × Runoff Interaction (Weekly 1:4 unidirectional)",
               style = "heading 1") %>%
  body_add_par(paste0("Sample: ", sum(cc_A$outbreak_binary == 1),
                      " case-weeks across ", uniqueN(cc_A$stratum_id),
                      " strata. Interaction tested via a 2-df runoff × birds_au_z ",
                      "crossbasis on the 1:4 unidirectional weekly primary spec."),
               style = "Normal") %>%
  body_add_par("Model fit", style = "heading 2") %>%
  body_add_flextable(flextable(fit_tbl_wk) %>% autofit() %>% theme_booktabs()) %>%
  body_add_par("Cumulative weekly runoff OR (per +0.5 SD across lag 0–4 wk)",
               style = "heading 2") %>%
  body_add_flextable(flextable(cum_tbl_wk) %>% autofit() %>% theme_booktabs()) %>%
  body_add_par("Lag-specific weekly runoff OR (per +0.5 SD on that lag week)",
               style = "heading 2") %>%
  body_add_flextable(flextable(lag_specific_wk) %>% autofit() %>% theme_booktabs())

print(doc, target = paste0(tables_sensitivity_folder, "Table_Feedlot_Interaction_Weekly.docx"))
cat("Saved: ", tables_sensitivity_folder, "Table_Feedlot_Interaction_Weekly.docx\n", sep = "")

saveRDS(list(
  m_primary_wk    = m_primary_wk,
  m_inter_wk      = m_inter_wk,
  fit_tbl_wk      = fit_tbl_wk,
  cum_tbl_wk      = cum_tbl_wk,
  lag_specific_wk = lag_specific_wk,
  lag_specific_wk_long = lag_specific_wk_long,
  LRT_p           = p_lrt
), paste0(objects_folder, "feedlot_interaction_weekly.RDS"))
```
