---
title: "Lag Window Sensitivity Analysis"
author: "Fintan Mooney"
date: "2025-10-15"
output: html_document
---

# Sensitivity analysis testing alternative lag windows (2, 6, 8 weeks) for the case-crossover DLNM... checking whether results are robust to different exposure lag assumptions!

```{r include=FALSE}
rm(list=ls())
project.folder = paste0(print(here::here()),'/')
source(paste0(project.folder,'create_folder_structure.R'))
source(paste0(functions.folder,'script_initiate.R'))
```


## Read in the timeseries dataset

```{r, read in timeseries dataset}

# Timeseries
fishnet_timeseries_clean <- readRDS(file.path(objects_folder, "fishnet_timeseries.RDS"))

fishnet_timeseries_clean <- fishnet_timeseries_clean %>%
  arrange(zone_id, week) %>%
  # 1. Drop all columns that end with ".x"
  select(-ends_with(".x")) %>%
  # 2. Rename columns ending with ".y" to remove the suffix
  rename_with(~ str_remove(.x, "\\.y$"), ends_with(".y"))

# Spatial GPKG
gpkg_path <- file.path(spatial_data, "fishnet_10km_timeseries.gpkg")

fishnet_weeks <- st_read(gpkg_path)

## Simplify
fishnet_weeks <- fishnet_weeks %>%
  dplyr::select(week, shape_length, shape_area, zone_id, geom)

```

## Map Cases in Casecrossover Model

```{r, casecrossover map}


# Map 
# Identify case zones
case_zone_ids <- fishnet_timeseries_clean %>%
  filter(any_outbreak == 1) %>%
  distinct(zone_id)

# Subset spatial features
case_zones_sf <- fishnet_weeks %>% filter(zone_id %in% case_zone_ids$zone_id)
control_zones_sf <- fishnet_weeks %>% filter(!(zone_id %in% case_zone_ids$zone_id))

# Define output paths
case_gpkg_out    <- file.path(processed_data, "Case_Zones.gpkg")
control_gpkg_out <- file.path(processed_data, "Control_Zones.gpkg")

# Export as GPKG
st_write(case_zones_sf, case_gpkg_out,    delete_layer = TRUE)
st_write(control_zones_sf, control_gpkg_out, delete_layer = TRUE)

# Reproject to EPSG:3857 (Web Mercator) for basemap compatibility
case_zones_sf_proj <- st_transform(case_zones_sf, 3857)
control_zones_sf_proj <- st_transform(control_zones_sf, 3857)

# Plot with basemap
ggplot() +
  # Add simple basemap
  annotation_map_tile(type = "cartolight", zoomin = 0) +

  # Control zones: outlines
  geom_sf(data = control_zones_sf_proj, fill = NA, color = "lightblue", linewidth = 0.05) +

  # Case zones: bold fill
  geom_sf(data = case_zones_sf_proj, fill = "#d73027", color = "red", linewidth = 0.1, alpha = 0.9) +

  # Scale bar
  annotation_scale(location = "bl", width_hint = 0.3, text_cex = 0.75, line_width = 0.5, style = "ticks") +

  # Theme and layout
  theme_void(base_size = 12) +
  theme(
    plot.title = element_text(size = 12, face = "bold", hjust = 0.5),
    plot.caption = element_text(size = 10, hjust = 0.5, face = "italic"),
    plot.background = element_rect(fill = "white", color = NA)
  ) +
  coord_sf(expand = FALSE) +
  labs(
    title = "HPAI Outbreak Case Zones in Minnesota (2022)"
  )
```

## Define Controls and Create Case-Crossover Dataset
```{r, define controls}

# DEFINE CASES + CONTROLS (REMOVE OVERLAPPING CONTROLS)

# Match
case_zones <- fishnet_timeseries_clean %>%
  dplyr::filter(any_outbreak == 1)

# Get binary cases: one row per zone_id and week with outbreak
cases <- case_zones %>%
  dplyr::filter(outbreak_count > 0) %>%
  dplyr::distinct(zone_id, case_week = week)

# Deduplicate case_zones to avoid many-to-many join issues
case_zones_unique <- case_zones %>%
  dplyr::distinct(zone_id, week, .keep_all = TRUE)

# Create full lag grid (1 to 4 weeks before each case)
case_lags <- cases %>%
  tidyr::crossing(lag = 1:4) %>%
  dplyr::mutate(control_week = case_week - lag) %>%
  dplyr::filter(control_week >= 1)

# --- REMOVE OVERLAPPING CONTROLS (control_week is also a case_week in same zone) ---
case_zone_weeks <- cases %>% dplyr::select(zone_id, case_week)

case_lags_clean <- case_lags %>%
  dplyr::anti_join(case_zone_weeks, by = c("zone_id", "control_week" = "case_week"))

cat("Controls removed (overlap with case weeks):",
    nrow(case_lags) - nrow(case_lags_clean), "\n")

# Controls (week-lagged, outbreak_binary = 0)
controls <- case_lags_clean %>%
  dplyr::left_join(case_zones_unique, by = c("zone_id", "control_week" = "week")) %>%
  dplyr::mutate(
    case_control    = "control",
    outbreak_binary = 0,
    stratum_id      = paste(zone_id, case_week, sep = "_"),
    week            = control_week
  ) %>%
  dplyr::select(-outbreak_count)

# Cases (outbreak_binary = 1)
case_data <- cases %>%
  dplyr::left_join(case_zones_unique, by = c("zone_id", "case_week" = "week")) %>%
  dplyr::mutate(
    case_control    = "case",
    outbreak_binary = 1,
    stratum_id      = paste(zone_id, case_week, sep = "_"),
    week            = case_week,
    control_week    = NA_integer_
  ) %>%
  dplyr::select(-outbreak_count)

# Combine cases and controls
case_crossover_df <- dplyr::bind_rows(case_data, controls)

# Summary check
table(case_crossover_df$case_control)
sum(case_crossover_df$outbreak_binary)
colnames(case_crossover_df)

```


# Check distribution of casecrossover variables

```{r, distribution caseccrossover}

# Move from m to mm

case_crossover_df <- case_crossover_df %>%
  mutate(
    runoff = runoff * 1000,
    precipitation = precipitation * 1000
  )


# Variables to inspect
vars_to_check <- c(
  "total_bird_abundance",   
  "leaf_area_index", "longwave_radiation", "runoff", "shortwave_radiation",
  "snow_density", "soil_moisture", "temperature",
  "evapotranspiration", "precipitation", "wind_speed", "relative_humidity", "anseriformes", "predators"
)

# Create histograms
plots <- lapply(vars_to_check, function(var) {
  ggplot(case_crossover_df, aes_string(x = var)) +
    geom_histogram(bins = 50, fill = "steelblue", color = "white") +
    labs(title = var) +
    theme_minimal()
})

# Display plots in a grid (adjust number of columns if needed)
wrap_plots(plots, ncol = )

# Subset to variables
numeric_vars <- case_crossover_df %>%
  select(all_of(vars_to_check))

# Initialize result storage
shapiro_results <- lapply(names(numeric_vars), function(var) {
  x <- numeric_vars[[var]]
  x <- na.omit(x)
  
  # Only apply Shapiro if <= 5000 non-NA values
  shapiro_p <- if (length(x) <= 5000) {
    shapiro.test(x)$p.value
  } else {
    NA
  }
  
  data.frame(
    variable = var,
    skewness = skewness(x),
    shapiro_p = shapiro_p,
    n = length(x)
  )
})

# Combine into summary table
shapiro_df <- bind_rows(shapiro_results)

# Flag non-normality
shapiro_df <- shapiro_df %>%
  mutate(non_normal = ifelse(!is.na(shapiro_p) & shapiro_p < 0.05, TRUE, FALSE)) %>%
  arrange(desc(abs(skewness)))

# View top
print(shapiro_df)

### Need to transform runoff, total_bird_abundance, specific humidity, and precipitation, anseriformes and predators


```

# Generate Case-Crossover Datasets with Different Lag Windows

```{r, different lags}


# LAG SENSITIVITY: BUILD DATASETS (REMOVE OVERLAPPING CONTROLS)

max_lags <- c(2, 6, 8)   # lag windows to test for predictors
n_controls <- 4          # fix the number of control weeks per case (still remove overlaps)
vars_to_lag <- c(
  "total_bird_abundance", "leaf_area_index", "longwave_radiation",
  "runoff", "shortwave_radiation", "snow_density", "soil_moisture",
  "temperature", "evapotranspiration", "precipitation",
  "wind_speed", "relative_humidity", "anseriformes", "predators"
)

case_zones_unique <- case_zones %>%
  dplyr::distinct(zone_id, week, .keep_all = TRUE)

# Function to build case-crossover dataset for a given lag window
build_casecrossover <- function(max_lag, n_controls = 4) {
  message("Building case-crossover dataset with ", max_lag, " lags (", n_controls, " controls per case)...")

  # ---- Create controls ----
  case_lags <- cases %>%
    tidyr::crossing(lag = 1:n_controls) %>%
    dplyr::mutate(control_week = case_week - lag) %>%
    dplyr::filter(control_week >= 1)

  case_zone_weeks <- cases %>% dplyr::select(zone_id, case_week)

  case_lags_clean <- case_lags %>%
    dplyr::anti_join(case_zone_weeks, by = c("zone_id", "control_week" = "case_week"))

  message("Controls removed (overlap with case weeks): ", nrow(case_lags) - nrow(case_lags_clean))

  controls <- case_lags_clean %>%
    dplyr::left_join(case_zones_unique, by = c("zone_id", "control_week" = "week")) %>%
    dplyr::mutate(
      case_control    = "control",
      outbreak_binary = 0,
      stratum_id      = paste(zone_id, case_week, sep = "_"),
      week            = control_week
    ) %>%
    dplyr::select(-outbreak_count)

  case_data <- cases %>%
    dplyr::left_join(case_zones_unique, by = c("zone_id", "case_week" = "week")) %>%
    dplyr::mutate(
      case_control    = "case",
      outbreak_binary = 1,
      stratum_id      = paste(zone_id, case_week, sep = "_"),
      week            = case_week,
      control_week    = NA_integer_
    ) %>%
    dplyr::select(-outbreak_count)

  case_crossover_df <- dplyr::bind_rows(case_data, controls)

  # ---- Remove non-independent strata (500m threshold) ----
  # Only remove strata where the sole outbreak is non-independent
  flagged_strata <- case_crossover_df %>%
    filter(flag_nearby_500m == TRUE, case_control == "case") %>%
    distinct(stratum_id, zone_id, week)
  
  if (nrow(flagged_strata) > 0) {
    flagged_with_count <- flagged_strata %>%
      left_join(
        case_zones_unique %>% select(zone_id, week, outbreak_count),
        by = c("zone_id", "week")
      )
    
    strata_to_remove <- flagged_with_count %>%
      filter(outbreak_count == 1) %>%
      distinct(stratum_id)
    
    case_crossover_df <- case_crossover_df %>%
      filter(!(stratum_id %in% strata_to_remove$stratum_id))
    
    message("Strata removed (non-independent): ", nrow(strata_to_remove))
  }
  
  message("Cases after exclusion: ", sum(case_crossover_df$outbreak_binary == 1))

  # ---- Add lagged predictors (0:max_lag) ----
  add_lagged_predictors <- function(df, original_df, vars, id_col = "zone_id", time_col = "week", max_lag = max_lag) {
    for (varname in vars) {
      for (lag in 0:max_lag) {
        lagged_col <- paste0(varname, "_Lag", lag)
        df <- df %>%
          dplyr::left_join(
            original_df %>%
              dplyr::select(!!rlang::sym(id_col), !!rlang::sym(time_col), !!rlang::sym(varname)) %>%
              dplyr::rename(!!lagged_col := !!rlang::sym(varname)) %>%
              dplyr::mutate(!!rlang::sym(time_col) := !!rlang::sym(time_col) + lag),
            by = c(id_col, time_col)
          )
      }
    }
    df
  }

  case_crossover_df <- add_lagged_predictors(
    df          = case_crossover_df,
    original_df = case_zones_unique,
    vars        = vars_to_lag,
    max_lag     = max_lag
  )

  # ---- Transform + scale ----
  skewed_vars <- c("total_bird_abundance", "runoff", "precipitation", "anseriformes", "predators")

  for (v in skewed_vars) {
    for (lag in 0:max_lag) {
      col <- paste0(v, "_Lag", lag)
      if (col %in% names(case_crossover_df)) {
        case_crossover_df[[col]] <- log1p(case_crossover_df[[col]])
      }
    }
  }

  for (v in vars_to_lag) {
    for (lag in 0:max_lag) {
      col <- paste0(v, "_Lag", lag)
      case_crossover_df[[col]] <- as.numeric(scale(case_crossover_df[[col]]))
    }
  }

  # ---- Restrict to complete strata (case + ≥1 control) ----
  keep_strata <- names(which(tapply(
    case_crossover_df$outbreak_binary,
    case_crossover_df$stratum_id,
    function(z) length(unique(z)) == 2
  )))

  case_crossover_df <- case_crossover_df %>%
    dplyr::filter(stratum_id %in% keep_strata)

  # ---- Save ----
  out_name <- paste0("case_crossover_df_", max_lag, "lag.RDS")
  saveRDS(case_crossover_df, file.path(objects_folder, out_name))
  message("Saved: ", out_name)

  return(case_crossover_df)
}

# Run for all lag windows
case_crossover_df_2lag <- build_casecrossover(2, n_controls = 4)
case_crossover_df_6lag <- build_casecrossover(6, n_controls = 4)
case_crossover_df_8lag <- build_casecrossover(8, n_controls = 4)


```

## Test against the casecrossover model with 4 lags

### Compare AICs

```{r, compare AICs}

# Load  4 lag model

case_crossover_df_4lag <- readRDS(file.path(objects_folder, "case_crossover_df.RDS")) # baseline 4-lag


# Helper to build crossbasis and fit clogit model

fit_casecrossover_model <- function(df, max_lag) {
  message("Fitting model for ", max_lag, " lags ...")
  
  # Build crossbasis objects using correct lag range
  cb_runoff    <- crossbasis(as.matrix(df[, paste0("runoff_Lag", 0:max_lag)]),
                             lag = c(0, max_lag),
                             argvar = list(fun="lin"),
                             arglag = list(fun="ns", df=2))
  cb_snow      <- crossbasis(as.matrix(df[, paste0("snow_density_Lag", 0:max_lag)]),
                             lag = c(0, max_lag),
                             argvar = list(fun="lin"),
                             arglag = list(fun="ns", df=2))
  cb_soil      <- crossbasis(as.matrix(df[, paste0("soil_moisture_Lag", 0:max_lag)]),
                             lag = c(0, max_lag),
                             argvar = list(fun="lin"),
                             arglag = list(fun="ns", df=2))
  cb_temp      <- crossbasis(as.matrix(df[, paste0("temperature_Lag", 0:max_lag)]),
                             lag = c(0, max_lag),
                             argvar = list(fun="lin"),
                             arglag = list(fun="ns", df=2))
  cb_precip    <- crossbasis(as.matrix(df[, paste0("precipitation_Lag", 0:max_lag)]),
                             lag = c(0, max_lag),
                             argvar = list(fun="lin"),
                             arglag = list(fun="ns", df=2))
  cb_wind      <- crossbasis(as.matrix(df[, paste0("wind_speed_Lag", 0:max_lag)]),
                             lag = c(0, max_lag),
                             argvar = list(fun="lin"),
                             arglag = list(fun="ns", df=2))
  cb_bird      <- crossbasis(as.matrix(df[, paste0("total_bird_abundance_Lag", 0:max_lag)]),
                             lag = c(0, max_lag),
                             argvar = list(fun="lin"),
                             arglag = list(fun="ns", df=2))
  
  # Fit clogit model
  fit <- clogit(
    outbreak_binary ~ 
      cb_runoff + cb_snow + cb_soil + cb_temp + 
      cb_precip + cb_wind + cb_bird + 
      strata(stratum_id),
    method = "efron",
    data = df
  )
  
  return(fit)
}


# Fit models and extract AICs

fit_2lag <- fit_casecrossover_model(case_crossover_df_2lag, 2)
fit_4lag <- fit_casecrossover_model(case_crossover_df_4lag, 4)
fit_6lag <- fit_casecrossover_model(case_crossover_df_6lag, 6)
fit_8lag <- fit_casecrossover_model(case_crossover_df_8lag, 8)

aic_table <- tibble(
  Model = c("2-lag", "4-lag", "6-lag", "8-lag"),
  AIC = c(
    AIC(fit_2lag),
    AIC(fit_4lag),
    AIC(fit_6lag),
    AIC(fit_8lag)
  )
)

# SUMMARIZE THE MODELS !
aic_table %>%
  arrange(AIC) %>%
  mutate(
    DeltaAIC = AIC - min(AIC)
  ) %>%
  gt::gt() %>%
  gt::tab_header(
    title = md("**Case-Crossover Model Comparison by Lag Window**"),
    subtitle = "AIC values for models with 2, 4, 6, and 8 week lag structures"
  ) %>%
  gt::fmt_number(columns = c(AIC, DeltaAIC), decimals = 2)



```


# Lets compute results for the model with 8-lags as it has the lowest AIC

## Analyzing cumulative and lag specific odds ratios from case-crossover with 8 lags 

```{r, 8 lag test}

## Crossbasis for 8-lag

# Finalized Crossbasis Terms for Climate and Bird Predictors

# Leaf Area Index
cb_leaf_area_index <- crossbasis(
  x = as.matrix(case_crossover_df_8lag[, paste0("leaf_area_index_Lag", 0:8)]),
  lag = c(0, 8),
  argvar = list(fun = "lin"),
  arglag = list(fun = "ns", df = 2)
)

# Longwave Radiation
cb_longwave_radiation <- crossbasis(
  x = as.matrix(case_crossover_df_8lag[, paste0("longwave_radiation_Lag", 0:8)]),
  lag = c(0, 8),
  argvar = list(fun = "lin"),
  arglag = list(fun = "ns", df = 2)
)

# Precipitation
cb_precipitation <- crossbasis(
  x = as.matrix(case_crossover_df_8lag[, paste0("precipitation_Lag", 0:8)]),
  lag = c(0, 8),
  argvar = list(fun = "lin"),
  arglag = list(fun = "ns", df = 2)
)

# Relative Humidity
cb_relative_humidity <- crossbasis(
  x = as.matrix(case_crossover_df_8lag[, paste0("relative_humidity_Lag", 0:8)]),
  lag = c(0, 8),
  argvar = list(fun = "lin"),
  arglag = list(fun = "ns", df = 2)
)

# Runoff
cb_runoff <- crossbasis(
  x = as.matrix(case_crossover_df_8lag[, paste0("runoff_Lag", 0:8)]),
  lag = c(0, 8),
  argvar = list(fun = "lin"),
  arglag = list(fun = "ns", df = 2)
)

# Shortwave Radiation
cb_shortwave_radiation <- crossbasis(
  x = as.matrix(case_crossover_df_8lag[, paste0("shortwave_radiation_Lag", 0:8)]),
  lag = c(0, 8),
  argvar = list(fun = "lin"),
  arglag = list(fun = "ns", df = 2)
)

# Snow Density
cb_snow_density <- crossbasis(
  x = as.matrix(case_crossover_df_8lag[, paste0("snow_density_Lag", 0:8)]),
  lag = c(0, 8),
  argvar = list(fun = "lin"),
  arglag = list(fun = "ns", df = 2)
)

# Soil Moisture
cb_soil_moisture <- crossbasis(
  x = as.matrix(case_crossover_df_8lag[, paste0("soil_moisture_Lag", 0:8)]),
  lag = c(0, 8),
  argvar = list(fun = "lin"),
  arglag = list(fun = "ns", df = 2)
)

# Temperature
cb_temperature <- crossbasis(
  x = as.matrix(case_crossover_df_8lag[, paste0("temperature_Lag", 0:8)]),
  lag = c(0, 8),
  argvar = list(fun = "lin"),
  arglag = list(fun = "ns", df = 2)
)


# Total Bird Abundance
cb_total_bird_abundance <- crossbasis(
  x = as.matrix(case_crossover_df_8lag[, paste0("total_bird_abundance_Lag", 0:8)]),
  lag = c(0, 8),
  argvar = list(fun = "lin"),
  arglag = list(fun = "ns", df = 2)
)

# Wind Speed
cb_wind_speed <- crossbasis(
  x = as.matrix(case_crossover_df_8lag[, paste0("wind_speed_Lag", 0:8)]),
  lag = c(0, 8),
  argvar = list(fun = "lin"),
  arglag = list(fun = "ns", df = 2)
)

# # # Anseriformes:
#  cb_anseri <- crossbasis(
#    x = as.matrix(case_crossover_df_8lag[, paste0("anseriformes_Lag", 0:8)]),
#    lag = c(0, 8),
#    argvar = list(fun = "lin"),
#    arglag = list(fun = "poly", degree = 2)
#  )
#  
# # # Predator:
#  cb_predators <- crossbasis(
#    x = as.matrix(case_crossover_df_8lag[, paste0("predators_Lag", 0:8)]),
#    lag = c(0, 8),
#    argvar = list(fun = "lin"),
#    arglag = list(fun = "poly", degree = 2)
#  )


# Fit Model

model_8lag <- clogit(
  outbreak_binary ~ 
    cb_runoff + cb_snow_density + cb_soil_moisture + cb_temperature +
    cb_precipitation + cb_wind_speed + cb_total_bird_abundance +
    strata(stratum_id),
  method = "efron",
  data = case_crossover_df_8lag
)


# Crosspred Objects
cp_runoff       <- crosspred(cb_runoff,       model_8lag, cen = 0, at = 0.5, cumul = TRUE)
cp_snow         <- crosspred(cb_snow_density, model_8lag, cen = 0, at = 0.5, cumul = TRUE)
cp_soil         <- crosspred(cb_soil_moisture,model_8lag, cen = 0, at = 0.5, cumul = TRUE)
cp_temp         <- crosspred(cb_temperature,  model_8lag, cen = 0, at = 0.5, cumul = TRUE)
cp_precip       <- crosspred(cb_precipitation,model_8lag, cen = 0, at = 0.5, cumul = TRUE)
cp_wind         <- crosspred(cb_wind_speed,   model_8lag, cen = 0, at = 0.5, cumul = TRUE)
cp_bird         <- crosspred(cb_total_bird_abundance, model_8lag, cen = 0, at = 0.5, cumul = TRUE)

cp_slice_runoff <- crosspred(cb_runoff,       model_8lag, cen = 0, at = 0.5)
cp_slice_snow   <- crosspred(cb_snow_density, model_8lag, cen = 0, at = 0.5)
cp_slice_soil   <- crosspred(cb_soil_moisture,model_8lag, cen = 0, at = 0.5)
cp_slice_temp   <- crosspred(cb_temperature,  model_8lag, cen = 0, at = 0.5)
cp_slice_precip <- crosspred(cb_precipitation,model_8lag, cen = 0, at = 0.5)
cp_slice_wind   <- crosspred(cb_wind_speed,   model_8lag, cen = 0, at = 0.5)
cp_slice_bird   <- crosspred(cb_total_bird_abundance, model_8lag, cen = 0, at = 0.5)


# Cumulative OR Summary Table

cumulative_summary <- tibble(
  Predictor = c("Runoff", "Snow Density", "Soil Moisture", "Temperature",
                "Precipitation", "Wind Speed", "Total Bird Abundance"),
  OR = c(
    cp_runoff$allRRfit["0.5"],
    cp_snow$allRRfit["0.5"],
    cp_soil$allRRfit["0.5"],
    cp_temp$allRRfit["0.5"],
    cp_precip$allRRfit["0.5"],
    cp_wind$allRRfit["0.5"],
    cp_bird$allRRfit["0.5"]
  ),
  CI_low = c(
    cp_runoff$allRRlow["0.5"],
    cp_snow$allRRlow["0.5"],
    cp_soil$allRRlow["0.5"],
    cp_temp$allRRlow["0.5"],
    cp_precip$allRRlow["0.5"],
    cp_wind$allRRlow["0.5"],
    cp_bird$allRRlow["0.5"]
  ),
  CI_high = c(
    cp_runoff$allRRhigh["0.5"],
    cp_snow$allRRhigh["0.5"],
    cp_soil$allRRhigh["0.5"],
    cp_temp$allRRhigh["0.5"],
    cp_precip$allRRhigh["0.5"],
    cp_wind$allRRhigh["0.5"],
    cp_bird$allRRhigh["0.5"]
  )
) %>%
  mutate(`OR (95% CI)` = sprintf("%.2f (%.2f, %.2f)", OR, CI_low, CI_high)) %>%
  select(Predictor, `OR (95% CI)`) %>%
  gt() %>%
  tab_header(
    title = md("**Case-Crossover Model (8-lag): Cumulative Odds Ratios**"),
    subtitle = "Per 0.5 SD Increase"
  ) %>%
  cols_label(`OR (95% CI)` = "Odds Ratio (95% Confidence Interval)") %>%
  tab_options(
    table.font.size = 14,
    heading.title.font.size = 16,
    heading.subtitle.font.size = 14,
    data_row.padding = px(6)
  )

cumulative_summary 


# Lag-Specific Odds Ratio Table (0–8 weeks)


extract_lag_ORs_wide <- function(cp_obj, varname, varval = "0.5", lags = 0:8) {
  lag_labels <- paste0("lag", lags)
  or   <- cp_obj$matRRfit[varval, lag_labels]
  low  <- cp_obj$matRRlow[varval, lag_labels]
  high <- cp_obj$matRRhigh[varval, lag_labels]
  tibble(
    Predictor = varname,
    !!!setNames(as.list(or),   paste0("OR_Lag", lags)),
    !!!setNames(as.list(low),  paste0("Low_Lag", lags)),
    !!!setNames(as.list(high), paste0("High_Lag", lags)),
    !!!setNames(as.list(sprintf("%.2f (%.2f, %.2f)", or, low, high)), paste0("Lag ", lags))
  )
}

# Combine all predictors
table_lag_ORs <- bind_rows(
  extract_lag_ORs_wide(cp_slice_runoff, "Runoff"),
  extract_lag_ORs_wide(cp_slice_snow,   "Snow Density"),
  extract_lag_ORs_wide(cp_slice_soil,   "Soil Moisture"),
  extract_lag_ORs_wide(cp_slice_temp,   "Temperature"),
  extract_lag_ORs_wide(cp_slice_precip, "Precipitation"),
  extract_lag_ORs_wide(cp_slice_wind,   "Wind Speed"),
  extract_lag_ORs_wide(cp_slice_bird,   "Total Bird Abundance")
)

# Identify significant lags (CI entirely above/below 1)
lags <- 0:8
highlight_matrix <- sapply(lags, function(i) {
  table_lag_ORs[[paste0("Low_Lag", i)]] > 1 | table_lag_ORs[[paste0("High_Lag", i)]] < 1
})
colnames(highlight_matrix) <- paste0("Lag ", lags)

# Build flextable for export
ft <- table_lag_ORs %>%
  select(Predictor, paste0("Lag ", lags)) %>%
  flextable() %>%
  set_header_labels(Predictor = "Predictor") %>%
  add_header_row(values = c("Predictor", "Odds Ratio (95% CI)"), colwidths = c(1, length(lags))) %>%
  fontsize(size = 11, part = "all") %>%
  autofit() %>%
  align(align = "center", part = "all") %>%
  theme_booktabs()

for (i in lags) {
  ft <- bold(ft, i = which(highlight_matrix[, paste0("Lag ", i)]), j = paste0("Lag ", i))
}

# Export Word table
doc <- read_docx() %>%
  body_add_flextable(ft) %>%
  body_add_par(
    "Table: Lag-Specific Odds Ratios for the 8-week Case-Crossover Model of Environmental and Bird Predictors of HPAI Outbreaks in Minnesota (2022). Odds ratios are per 0.5 SD increase; statistically significant values are shown in bold.",
    style = "Normal"
  )

print(doc, target = file.path(tables_folder, "Table_LagSpecific_OR_HPAI_Minnesota_8lag.docx"))

ft



```

## Analyzing cumulative and lag specific odds ratios with arglag in crossbasis = poly (df = 2)

```{r, 8 lag test with poly df=2}

## Crossbasis for 8-lag

# Finalized Crossbasis Terms for Climate and Bird Predictors

# Leaf Area Index
cb_leaf_area_index <- crossbasis(
  x = as.matrix(case_crossover_df_8lag[, paste0("leaf_area_index_Lag", 0:8)]),
  lag = c(0, 8),
  argvar = list(fun = "lin"),
  arglag = list(fun="poly", degree=2)
)

# Longwave Radiation
cb_longwave_radiation <- crossbasis(
  x = as.matrix(case_crossover_df_8lag[, paste0("longwave_radiation_Lag", 0:8)]),
  lag = c(0, 8),
  argvar = list(fun = "lin"),
  arglag = list(fun="poly", degree=2)
)

# Precipitation
cb_precipitation <- crossbasis(
  x = as.matrix(case_crossover_df_8lag[, paste0("precipitation_Lag", 0:8)]),
  lag = c(0, 8),
  argvar = list(fun = "lin"),
  arglag = list(fun="poly", degree=2)
)

# Relative Humidity
cb_relative_humidity <- crossbasis(
  x = as.matrix(case_crossover_df_8lag[, paste0("relative_humidity_Lag", 0:8)]),
  lag = c(0, 8),
  argvar = list(fun = "lin"),
  arglag = list(fun="poly", degree=2)
)

# Runoff
cb_runoff <- crossbasis(
  x = as.matrix(case_crossover_df_8lag[, paste0("runoff_Lag", 0:8)]),
  lag = c(0, 8),
  argvar = list(fun = "lin"),
  arglag = list(fun="poly", degree=2)
)

# Shortwave Radiation
cb_shortwave_radiation <- crossbasis(
  x = as.matrix(case_crossover_df_8lag[, paste0("shortwave_radiation_Lag", 0:8)]),
  lag = c(0, 8),
  argvar = list(fun = "lin"),
  arglag = list(fun="poly", degree=2)
)

# Snow Density
cb_snow_density <- crossbasis(
  x = as.matrix(case_crossover_df_8lag[, paste0("snow_density_Lag", 0:8)]),
  lag = c(0, 8),
  argvar = list(fun = "lin"),
  arglag = list(fun="poly", degree=2)
)

# Soil Moisture
cb_soil_moisture <- crossbasis(
  x = as.matrix(case_crossover_df_8lag[, paste0("soil_moisture_Lag", 0:8)]),
  lag = c(0, 8),
  argvar = list(fun = "lin"),
  arglag = list(fun="poly", degree=2)
)

# Temperature
cb_temperature <- crossbasis(
  x = as.matrix(case_crossover_df_8lag[, paste0("temperature_Lag", 0:8)]),
  lag = c(0, 8),
  argvar = list(fun = "lin"),
  arglag = list(fun="poly", degree=2)
)


# Total Bird Abundance
cb_total_bird_abundance <- crossbasis(
  x = as.matrix(case_crossover_df_8lag[, paste0("total_bird_abundance_Lag", 0:8)]),
  lag = c(0, 8),
  argvar = list(fun = "lin"),
  arglag = list(fun="poly", degree=2)
)

# Wind Speed
cb_wind_speed <- crossbasis(
  x = as.matrix(case_crossover_df_8lag[, paste0("wind_speed_Lag", 0:8)]),
  lag = c(0, 8),
  argvar = list(fun = "lin"),
  arglag = list(fun="poly", degree=2)
)

# # # Anseriformes:
#  cb_anseri <- crossbasis(
#    x = as.matrix(case_crossover_df_8lag[, paste0("anseriformes_Lag", 0:8)]),
#    lag = c(0, 8),
#    argvar = list(fun = "lin"),
#    arglag = list(fun = "poly", degree = 2)
#  )
#  
# # # Predator:
#  cb_predators <- crossbasis(
#    x = as.matrix(case_crossover_df_8lag[, paste0("predators_Lag", 0:8)]),
#    lag = c(0, 8),
#    argvar = list(fun = "lin"),
#    arglag = list(fun = "poly", degree = 2)
#  )


# Fit Model

model_8lag <- clogit(
  outbreak_binary ~ 
    cb_runoff + cb_snow_density + cb_soil_moisture + cb_temperature +
    cb_precipitation + cb_wind_speed + cb_total_bird_abundance +
    strata(stratum_id),
  method = "efron",
  data = case_crossover_df_8lag
)


# Crosspred Objects

cp_runoff       <- crosspred(cb_runoff,       model_8lag, cen = 0, at = 0.5, cumul = TRUE)
cp_snow         <- crosspred(cb_snow_density, model_8lag, cen = 0, at = 0.5, cumul = TRUE)
cp_soil         <- crosspred(cb_soil_moisture,model_8lag, cen = 0, at = 0.5, cumul = TRUE)
cp_temp         <- crosspred(cb_temperature,  model_8lag, cen = 0, at = 0.5, cumul = TRUE)
cp_precip       <- crosspred(cb_precipitation,model_8lag, cen = 0, at = 0.5, cumul = TRUE)
cp_wind         <- crosspred(cb_wind_speed,   model_8lag, cen = 0, at = 0.5, cumul = TRUE)
cp_bird         <- crosspred(cb_total_bird_abundance, model_8lag, cen = 0, at = 0.5, cumul = TRUE)

cp_slice_runoff <- crosspred(cb_runoff,       model_8lag, cen = 0, at = 0.5)
cp_slice_snow   <- crosspred(cb_snow_density, model_8lag, cen = 0, at = 0.5)
cp_slice_soil   <- crosspred(cb_soil_moisture,model_8lag, cen = 0, at = 0.5)
cp_slice_temp   <- crosspred(cb_temperature,  model_8lag, cen = 0, at = 0.5)
cp_slice_precip <- crosspred(cb_precipitation,model_8lag, cen = 0, at = 0.5)
cp_slice_wind   <- crosspred(cb_wind_speed,   model_8lag, cen = 0, at = 0.5)
cp_slice_bird   <- crosspred(cb_total_bird_abundance, model_8lag, cen = 0, at = 0.5)


# Cumulative OR Summary Table

cumulative_summary <- tibble(
  Predictor = c("Runoff", "Snow Density", "Soil Moisture", "Temperature",
                "Precipitation", "Wind Speed", "Total Bird Abundance"),
  OR = c(
    cp_runoff$allRRfit["0.5"],
    cp_snow$allRRfit["0.5"],
    cp_soil$allRRfit["0.5"],
    cp_temp$allRRfit["0.5"],
    cp_precip$allRRfit["0.5"],
    cp_wind$allRRfit["0.5"],
    cp_bird$allRRfit["0.5"]
  ),
  CI_low = c(
    cp_runoff$allRRlow["0.5"],
    cp_snow$allRRlow["0.5"],
    cp_soil$allRRlow["0.5"],
    cp_temp$allRRlow["0.5"],
    cp_precip$allRRlow["0.5"],
    cp_wind$allRRlow["0.5"],
    cp_bird$allRRlow["0.5"]
  ),
  CI_high = c(
    cp_runoff$allRRhigh["0.5"],
    cp_snow$allRRhigh["0.5"],
    cp_soil$allRRhigh["0.5"],
    cp_temp$allRRhigh["0.5"],
    cp_precip$allRRhigh["0.5"],
    cp_wind$allRRhigh["0.5"],
    cp_bird$allRRhigh["0.5"]
  )
) %>%
  mutate(`OR (95% CI)` = sprintf("%.2f (%.2f, %.2f)", OR, CI_low, CI_high)) %>%
  select(Predictor, `OR (95% CI)`) %>%
  gt() %>%
  tab_header(
    title = md("**Case-Crossover Model (8-lag): Cumulative Odds Ratios**"),
    subtitle = "Per 0.5 SD Increase"
  ) %>%
  cols_label(`OR (95% CI)` = "Odds Ratio (95% Confidence Interval)") %>%
  tab_options(
    table.font.size = 14,
    heading.title.font.size = 16,
    heading.subtitle.font.size = 14,
    data_row.padding = px(6)
  )

cumulative_summary 


# Lag-Specific Odds Ratio Table (0–8 weeks)

extract_lag_ORs_wide <- function(cp_obj, varname, varval = "0.5", lags = 0:8) {
  lag_labels <- paste0("lag", lags)
  or   <- cp_obj$matRRfit[varval, lag_labels]
  low  <- cp_obj$matRRlow[varval, lag_labels]
  high <- cp_obj$matRRhigh[varval, lag_labels]
  tibble(
    Predictor = varname,
    !!!setNames(as.list(or),   paste0("OR_Lag", lags)),
    !!!setNames(as.list(low),  paste0("Low_Lag", lags)),
    !!!setNames(as.list(high), paste0("High_Lag", lags)),
    !!!setNames(as.list(sprintf("%.2f (%.2f, %.2f)", or, low, high)), paste0("Lag ", lags))
  )
}

# Combine all predictors
table_lag_ORs <- bind_rows(
  extract_lag_ORs_wide(cp_slice_runoff, "Runoff"),
  extract_lag_ORs_wide(cp_slice_snow,   "Snow Density"),
  extract_lag_ORs_wide(cp_slice_soil,   "Soil Moisture"),
  extract_lag_ORs_wide(cp_slice_temp,   "Temperature"),
  extract_lag_ORs_wide(cp_slice_precip, "Precipitation"),
  extract_lag_ORs_wide(cp_slice_wind,   "Wind Speed"),
  extract_lag_ORs_wide(cp_slice_bird,   "Total Bird Abundance")
)

# Identify significant lags (CI entirely above/below 1)
lags <- 0:8
highlight_matrix <- sapply(lags, function(i) {
  table_lag_ORs[[paste0("Low_Lag", i)]] > 1 | table_lag_ORs[[paste0("High_Lag", i)]] < 1
})
colnames(highlight_matrix) <- paste0("Lag ", lags)

# Build flextable for export
ft <- table_lag_ORs %>%
  select(Predictor, paste0("Lag ", lags)) %>%
  flextable() %>%
  set_header_labels(Predictor = "Predictor") %>%
  add_header_row(values = c("Predictor", "Odds Ratio (95% CI)"), colwidths = c(1, length(lags))) %>%
  fontsize(size = 11, part = "all") %>%
  autofit() %>%
  align(align = "center", part = "all") %>%
  theme_booktabs()

for (i in lags) {
  ft <- bold(ft, i = which(highlight_matrix[, paste0("Lag ", i)]), j = paste0("Lag ", i))
}

# Export Word table
doc <- read_docx() %>%
  body_add_flextable(ft) %>%
  body_add_par(
    "Table: Lag-Specific Odds Ratios for the 8-week Case-Crossover Model of Environmental and Bird Predictors of HPAI Outbreaks in Minnesota (2022). Odds ratios are per 0.5 SD increase; statistically significant values are shown in bold.",
    style = "Normal"
  )

print(doc, target = file.path(tables_folder, "Table_LagSpecific_OR_HPAI_Minnesota_8lag_poly_df2.docx"))

ft






```

# Even though the AIC is the lowest for the case-crossover over 8-lags, we can see from the cumlative and lag-specific odds ratios that the model is very unstable (wide confidence intervals). We thus choose the case-crossover over 4-lags that has a moderate AIC and stable results



