---
title: "Seasonal Stratification Sensitivity"
author: "Fintan Mooney"
date: "2025-10-07"
output: html_document
---
# Sensitivity analysis stratifying the case-crossover model by season... testing whether HPAI outbreak risk factors differ across spring, summer, fall, and winter!

```{r include=FALSE}
rm(list=ls())
project.folder = paste0(print(here::here()),'/')
source(paste0(project.folder,'create_folder_structure.R'))
source(paste0(functions.folder,'script_initiate.R'))
```

## Import casecrossover df

```{r, import casecrossover df}

case_crossover_df <- readRDS(file.path(objects_folder, "case_crossover_df.RDS"))

```

## Investigate distribution of outbreaks by season (month)

```{r, investigate distribution of outbreaks by season}

# Outbreak counts by month with adjusted seasonal vertical lines


# Summarize total outbreaks per month
outbreak_by_month <- case_crossover_df |>
  dplyr::group_by(month) |>
  dplyr::summarise(outbreaks = sum(outbreak_binary, na.rm = TRUE)) |>
  # Ensure all 12 months are represented
  dplyr::right_join(tibble(month = 1:12), by = "month") |>
  dplyr::mutate(outbreaks = dplyr::coalesce(outbreaks, 0))

# Define vertical line positions 
season_markers <- data.frame(
  month = c(3, 6, 9, 12) - 0.5,  # shift lines back by 0.5 month
  season = c("Spring", "Summer", "Fall", "Winter")
)

# Build plot
p_outbreaks_month <- ggplot(outbreak_by_month, aes(x = month, y = outbreaks)) +
  # Bars
  geom_col(fill = "grey50", color = "black", width = 0.8) +

  # Numeric labels above bars
  geom_text(aes(label = outbreaks),
            vjust = -0.3, size = 3.3, fontface = "bold") +

  # Red vertical dashed lines for seasons
  geom_vline(
    data = season_markers,
    aes(xintercept = month),
    color = "red3",
    linetype = "dashed",
    linewidth = 0.6
  ) +

  # Vertical rotated season labels
  geom_text(
    data = season_markers,
    aes(x = month, y = max(outbreak_by_month$outbreaks) * 1.05, label = season),
    color = "red4",
    angle = 90,
    vjust = 1.5,
    hjust = 1,
    size = 3.5,
    fontface = "bold"
  ) +

  # Axes and labels
  scale_x_continuous(
    breaks = 1:12,
    labels = month.abb,
    expand = c(0, 0)
  ) +
  scale_y_continuous(
    expand = expansion(mult = c(0, 0.1))
  ) +
  labs(
    title = "Monthly Distribution of HPAI Outbreaks",
    x = "Month",
    y = "Number of Outbreaks"
  ) +
  theme_bw(base_size = 12) +
  theme(
    panel.grid = element_blank(),
    plot.title = element_text(face = "bold", size = 13),
    axis.title = element_text(size = 11),
    axis.text.x = element_text(size = 9)
  )

# Save

# PNG version
ggsave(
  filename = file.path(figures_folder, "figure_outbreaks_by_month.png"),
  plot = p_outbreaks_month,
  width = 8, height = 5, dpi = 600
)


# Manuscript version - by week with calendar season markers
# Summarize total outbreaks per week
outbreak_by_week <- case_crossover_df |>
  dplyr::group_by(week) |>
  dplyr::summarise(outbreaks = sum(outbreak_binary, na.rm = TRUE)) |>
  # Ensure all 52 weeks are represented
  dplyr::right_join(tibble(week = 1:52), by = "week") |>
  dplyr::mutate(outbreaks = dplyr::coalesce(outbreaks, 0))

# Define vertical line positions for calendar seasons (week ~9, ~22, ~35, ~48)
# March 1 ≈ week 9, June 1 ≈ week 22, Sept 1 ≈ week 35, Dec 1 ≈ week 48
season_markers <- data.frame(
  week = c(9, 22, 35, 48),
  season = c("Spring", "Summer", "Fall", "Winter")
)

# Build plot
p_outbreaks_week <- ggplot(outbreak_by_week, aes(x = week, y = outbreaks)) +
  # Bars
  geom_col(fill = "grey50", color = "black", width = 0.8) +
  # Red vertical dashed lines for seasons
  geom_vline(
    data = season_markers,
    aes(xintercept = week),
    color = "red3",
    linetype = "dashed",
    linewidth = 0.6
  ) +
  # Vertical rotated season labels
  geom_text(
    data = season_markers,
    aes(x = week, y = max(outbreak_by_week$outbreaks) * 1.05, label = season),
    color = "red4",
    angle = 90,
    vjust = 1.5,
    hjust = 1,
    size = 3.5,
    fontface = "bold"
  ) +
  # Axes and labels
  scale_x_continuous(
    breaks = seq(2, 52, by = 2),
    expand = c(0, 0)
  ) +
  scale_y_continuous(
    breaks = seq(0, max(outbreak_by_week$outbreaks) + 2, by = 2),
    expand = expansion(mult = c(0, 0.1))
  ) +
  labs(
    title = NULL,
    x = "Week",
    y = "Number of Spillover Events"
  ) +
  theme_bw(base_size = 12) +
  theme(
    panel.grid = element_blank(),
    axis.title = element_text(size = 11),
    axis.text.x = element_text(size = 9)
  )

p_outbreaks_week

# Save
ggsave(
  filename = file.path(figures_folder, "figure_outbreaks_by_week_manuscript.png"),
  plot = p_outbreaks_week,
  width = 10, height = 5, dpi = 600
)
```

## Due to convergence issues (little number of outbreaks when stratifying by season), we will remove the winter and summer data

## We can filter by astronomical or by calander month for season; we chose to filter to calander month in order to pick up the exact migratory periods of wild birds (the fall period would blend into summer if we used astronomical)

```{r, create astronomical seasons}

# Approximate week boundaries for astronomical seasons in 2022:
# Spring: March 20 = ~week 12
# Summer: June 21 = ~week 25
# Fall: Sept 22 = ~week 39
# Winter: Dec 21 = ~week 51

# case_crossover_df <- case_crossover_df %>%
#   mutate(
#     season = case_when(
#       week >= 12 & week < 25 ~ "Spring",
#       week >= 25 & week < 39 ~ "Summer",
#       week >= 39 & week < 51 ~ "Fall",
#       TRUE ~ "Winter"
#     ),
# 
#     spring_astro = as.integer(season == "Spring"),
#     fall_astro = as.integer(season == "Fall"),
#     spring_fall_astro = as.integer(season %in% c("Spring", "Fall"))
#   )


```

# Outbreak counts by week with astronomical season markers
```{r, astronomical distribution}
# # Summarize total outbreaks per week
# outbreak_by_week <- case_crossover_df |>
#   dplyr::group_by(week) |>
#   dplyr::summarise(outbreaks = sum(outbreak_binary, na.rm = TRUE)) |>
#   # Ensure all 52 weeks are represented
#   dplyr::right_join(tibble(week = 1:52), by = "week") |>
#   dplyr::mutate(outbreaks = dplyr::coalesce(outbreaks, 0))
# 
# # Define vertical line positions for astronomical seasons
# season_markers <- data.frame(
#   week = c(12, 25, 39, 51),
#   season = c("Spring", "Summer", "Fall", "Winter")
# )
# 
# # Build plot
# p_outbreaks_week <- ggplot(outbreak_by_week, aes(x = week, y = outbreaks)) +
#   # Bars
#   geom_col(fill = "grey50", color = "black", width = 0.8) +
#   # Red vertical dashed lines for astronomical seasons
#   geom_vline(
#     data = season_markers,
#     aes(xintercept = week),
#     color = "red3",
#     linetype = "dashed",
#     linewidth = 0.6
#   ) +
#   # Vertical rotated season labels
#   geom_text(
#     data = season_markers,
#     aes(x = week, y = max(outbreak_by_week$outbreaks) * 1.05, label = season),
#     color = "red4",
#     angle = 90,
#     vjust = 1.5,
#     hjust = 1,
#     size = 3.5,
#     fontface = "bold"
#   ) +
#   # Axes and labels
#   scale_x_continuous(
#     breaks = seq(2, 52, by = 2),
#     expand = c(0, 0)
#   ) +
#   scale_y_continuous(
#     breaks = seq(0, max(outbreak_by_week$outbreaks) + 2, by = 2),
#     expand = expansion(mult = c(0, 0.1))
#   ) +
#   labs(
#     title = NULL,
#     x = "Week",
#     y = "Number of Outbreaks"
#   ) +
#   theme_bw(base_size = 12) +
#   theme(
#     panel.grid = element_blank(),
#     plot.title = element_text(face = "bold", size = 13),
#     axis.title = element_text(size = 11),
#     axis.text.x = element_text(size = 9)
#   )
# 
# p_outbreaks_week
# 
# # Save
# ggsave(
#   filename = file.path(figures_folder, "figure_outbreaks_by_week_astro.png"),
#   plot = p_outbreaks_week,
#   width = 10, height = 5, dpi = 600
# )
```

## Split the datasets

```{r, split into fall and spring}

# Check case counts per season
case_crossover_df %>%
  filter(outbreak_binary == 1) %>%
  count(season)

# case_spring <- case_crossover_df %>%
#   filter(spring_astro == 1)
# 
# case_springfall <- case_crossover_df %>%
#   filter(spring_fall_astro == 1)

case_spring <- case_crossover_df %>%
  filter(spring== 1)

case_springfall <- case_crossover_df %>%
  filter(spring_fall == 1)

sum(case_spring$outbreak_binary)
sum(case_springfall$outbreak_binary)

# count cells by season

case_crossover_df %>%
count(season) %>%
  print()
```

## Case-Crossover Models on dataframe stratfied to only spring and fall

### Crossbasis 

```{r, crossbasis spring fall}

# Leaf Area Index: ns(df = 3), ns(df = 2)
cb_springfall_leaf_area_index <- crossbasis(
  x = as.matrix(case_springfall[, paste0("leaf_area_index_Lag", 0:4)]),
  lag = c(0, 4),
  argvar = list(fun = "lin"),
  arglag = list(fun = "ns", df = 2)
)

# Precipitation: ns(df = 3), ns(df = 2)
cb_springfall_precipitation <- crossbasis(
  x = as.matrix(case_springfall[, paste0("precipitation_Lag", 0:4)]),
  lag = c(0, 4),
  argvar = list(fun = "lin"),
  arglag = list(fun = "ns", df = 2)
)

# Relative Humidity: ns(df = 3), poly(degree = 2)
cb_springfall_relative_humidity <- crossbasis(
  x = as.matrix(case_springfall[, paste0("relative_humidity_Lag", 0:4)]),
  lag = c(0, 4),
  argvar = list(fun = "lin"),
  arglag = list(fun = "ns", df = 2)
)



# Runoff: ns(df = 2), poly(degree = 2)
cb_springfall_runoff <- crossbasis(
  x = as.matrix(case_springfall[, paste0("runoff_Lag", 0:4)]),
  lag = c(0, 4),
  argvar = list(fun = "lin"),
  arglag = list(fun = "ns", df = 2)
)

# Shortwave Radiation: ns(df = 3), poly(degree = 4)
cb_springfall_shortwave_radiation <- crossbasis(
  x = as.matrix(case_springfall[, paste0("shortwave_radiation_Lag", 0:4)]),
  lag = c(0, 4),
  argvar = list(fun = "lin"),
  arglag = list(fun = "ns", df = 2)
)

# Longwave Radiation: ns(df = 3), ns(df = 2)
cb_springfall_longwave_radiation <- crossbasis(
  x = as.matrix(case_springfall[, paste0("shortwave_radiation_Lag", 0:4)]),
  lag = c(0, 4),
  argvar = list(fun = "lin"),
  arglag = list(fun = "ns", df = 2)
)

# Snow Density: ns(df = 2), poly(degree = 3)
cb_springfall_snow_density <- crossbasis(
  x = as.matrix(case_springfall[, paste0("snow_density_Lag", 0:4)]),
  lag = c(0, 4),
  argvar = list(fun = "lin"),
  arglag = list(fun = "ns", df = 2)
)

# Soil Moisture: linear exposure, ns(df = 2)
cb_springfall_soil_moisture <- crossbasis(
  x = as.matrix(case_springfall[, paste0("soil_moisture_Lag", 0:4)]),
  lag = c(0, 4),
  argvar = list(fun = "lin"),
  arglag = list(fun = "ns", df = 2)
)

# Temperature: ns(df = 3), poly(degree = 3)
cb_springfall_temperature <- crossbasis(
  x = as.matrix(case_springfall[, paste0("temperature_Lag", 0:4)]),
  lag = c(0, 4),
  argvar = list(fun = "lin"),
  arglag = list(fun = "ns", df = 2)
)

# Bird Abundance: ns(df = 3), ns(df = 2)
cb_springfall_total_bird_abundance <- crossbasis(
  x = as.matrix(case_springfall[, paste0("total_bird_abundance_Lag", 0:4)]),
  lag = c(0, 4),
  argvar = list(fun = "lin"),
  arglag = list(fun = "ns", df = 2)
)

# Wind Speed: ns(df = 2), poly(degree = 4)
cb_springfall_wind_speed <- crossbasis(
  x = as.matrix(case_springfall[, paste0("wind_speed_Lag", 0:4)]),
  lag = c(0, 4),
  argvar = list(fun = "lin"),
  arglag = list(fun = "ns", df = 2)
)






```

### Spring-Fall Final Model

```{r, spring fall model}

springfall_model <- clogit(
  outbreak_binary ~ 
    cb_springfall_runoff +            
    cb_springfall_snow_density +           
    cb_springfall_soil_moisture +          
    cb_springfall_temperature +           
    cb_springfall_precipitation +         
    cb_springfall_wind_speed +             
    cb_springfall_total_bird_abundance +   
    strata(stratum_id),
  method = "efron",
  data = case_springfall
)

summary(springfall_model)

```

### Spring-Fall Crosspred Objects

```{r, spring fall crosspred}
center_runoff <- mean(case_springfall$runoff, na.rm = TRUE)
center_snow   <- mean(case_springfall$snow_density, na.rm = TRUE)
center_soil   <- mean(case_springfall$soil_moisture, na.rm = TRUE)
center_temp   <- mean(case_springfall$temperature, na.rm = TRUE)
center_precip <- mean(case_springfall$precipitation, na.rm = TRUE)
center_wind   <- mean(case_springfall$wind_speed, na.rm = TRUE)
center_bird   <- mean(case_springfall$total_bird_abundance, na.rm = TRUE)


cp_springfall_runoff       <- crosspred(cb_springfall_runoff, springfall_model, center_runoff, at = 0.5, cumul = TRUE)
cp_springfall_snow         <- crosspred(cb_springfall_snow_density, springfall_model, center_snow , at = 0.5, cumul = TRUE)
cp_springfall_soil         <- crosspred(cb_springfall_soil_moisture, springfall_model, center_soil , at = 0.5, cumul = TRUE)
cp_springfall_temp         <- crosspred(cb_springfall_temperature, springfall_model, center_temp    , at = 0.5, cumul = TRUE)
cp_springfall_precip       <- crosspred(cb_springfall_precipitation, springfall_model, center_precip, at = 0.5, cumul = TRUE)
cp_springfall_wind         <- crosspred(cb_springfall_wind_speed, springfall_model, center_wind, at = 0.5, cumul = TRUE)
cp_springfall_bird         <- crosspred(cb_springfall_total_bird_abundance, springfall_model, center_bird  , at = 0.5, cumul = TRUE)

cp_springfall_slice_runoff <- crosspred(cb_springfall_runoff, springfall_model, center_runoff, at = 0.5)
cp_springfall_slice_snow   <- crosspred(cb_springfall_snow_density, springfall_model, center_snow, at = 0.5)
cp_springfall_slice_soil   <- crosspred(cb_springfall_soil_moisture, springfall_model, center_soil, at = 0.5)
cp_springfall_slice_temp   <- crosspred(cb_springfall_temperature, springfall_model, center_temp, at = 0.5)
cp_springfall_slice_precip <- crosspred(cb_springfall_precipitation, springfall_model,  center_precip, at = 0.5)
cp_springfall_slice_wind   <- crosspred(cb_springfall_wind_speed, springfall_model, center_wind, at = 0.5)
cp_springfall_slice_bird   <- crosspred(cb_springfall_total_bird_abundance, springfall_model, center_bird, at = 0.5)

```

### Results from Spring-Fall Model

```{r, cumulative odds ratios spring fall}
# ----------------------------
# Cumulative OR Summary Table
# ----------------------------
cumulative_summary <- tibble(
  Predictor = c("Runoff", "Snow Density", "Soil Moisture", "Temperature",
                "Precipitation", "Wind Speed", "Total Bird Abundance"),
      OR = c(
        cp_springfall_runoff$allRRfit["0.5"],
        cp_springfall_snow$allRRfit["0.5"],
        cp_springfall_soil$allRRfit["0.5"],
        cp_springfall_temp$allRRfit["0.5"],
        cp_springfall_precip$allRRfit["0.5"],
        cp_springfall_wind$allRRfit["0.5"],
        cp_springfall_bird$allRRfit["0.5"]
      ),
      CI_low = c(
        cp_springfall_runoff$allRRlow["0.5"],
        cp_springfall_snow$allRRlow["0.5"],
        cp_springfall_soil$allRRlow["0.5"],
        cp_springfall_temp$allRRlow["0.5"],
        cp_springfall_precip$allRRlow["0.5"],
        cp_springfall_wind$allRRlow["0.5"],
        cp_springfall_bird$allRRlow["0.5"]
      ),
      CI_high = c(
        cp_springfall_runoff$allRRhigh["0.5"],
        cp_springfall_snow$allRRhigh["0.5"],
        cp_springfall_soil$allRRhigh["0.5"],
        cp_springfall_temp$allRRhigh["0.5"],
        cp_springfall_precip$allRRhigh["0.5"],
        cp_springfall_wind$allRRhigh["0.5"],
        cp_springfall_bird$allRRhigh["0.5"]
      )
) %>%
  mutate(`OR (95% CI)` = sprintf("%.2f (%.2f, %.2f)", OR, CI_low, CI_high)) %>%
  select(Predictor, `OR (95% CI)`) %>%
  gt() %>%
  tab_header(
    title = md("**Case-Crossover Model: Cumulative Odds Ratios**"),
    subtitle = "Per 0.5 SD Increase"
  ) %>%
  cols_label(`OR (95% CI)` = "Odds Ratio (95% Confidence Interval)") %>%
  tab_options(
    table.font.size = 14,
    heading.title.font.size = 16,
    heading.subtitle.font.size = 14,
    data_row.padding = px(6)
  )

cumulative_summary



# ----------------------------
# Plotting Functions
# ----------------------------
plot_cumulative <- function(cp, title) {
  plot(cp, type = "cumulative", var = 0.5, ci = "area",
       col = "black", ci.arg = list(col = rgb(0,0,0,0.1), border = NA),
       lwd = 1.7, main = title, xlab = "", ylab = "", las = 1)
  abline(h = 1, lty = 2, col = "gray40", lwd = 1)
}

plot_slices <- function(cp, title) {
  plot(cp, type = "slices", var = 0.5, ci = "area",
       col = "black", ci.arg = list(col = rgb(0,0,0,0.1), border = NA),
       lwd = 1.3, main = title, xlab = "", ylab = "", las = 1)
  abline(h = 1, lty = 2, col = "gray40", lwd = 1)
}

# ----------------------------
# Save Cumulative Plot
# ----------------------------
png(file.path(figures_folder,"figure_cumulative_ORs_casecrossover_springfall.png"), width = 11, height = 7, units = "in", res = 300)

par(
  mfrow = c(2, 4),
  mar = c(4.2, 4.2, 2.5, 1),
  oma = c(3, 4, 0, 1),
  cex.main = 1.2,
  cex.lab = 1.2,
  cex.axis = 1.1,
  tcl = -0.3, mgp = c(2.5, 0.7, 0)
)

plot_cumulative(cp_springfall_runoff, "Runoff")
plot_cumulative(cp_springfall_snow, "Snow Density")
plot_cumulative(cp_springfall_soil, "Soil Moisture")
plot_cumulative(cp_springfall_temp, "Temperature")
plot_cumulative(cp_springfall_precip, "Precipitation")
plot_cumulative(cp_springfall_wind, "Wind Speed")
plot_cumulative(cp_springfall_bird, "Total Bird Abundance")

mtext("Lag (weeks)", side = 1, outer = TRUE, line = 1, cex = 1.2)
mtext("Odds Ratio (95% CI)", side = 2, outer = TRUE, line = 1.5, cex = 1.2)

dev.off()


# ----------------------------
# Save Slices Plot
# ----------------------------
png(file.path(figures_folder,"figure_slice_ORs_casecrossover_springfall.png"), width = 11, height = 7, units = "in", res = 300)

par(
  mfrow = c(2, 4),
  mar = c(4.2, 4.2, 2.5, 1),
  oma = c(3, 4, 0, 1),
  cex.main = 1.2,
  cex.lab = 1.2,
  cex.axis = 1.1,
  tcl = -0.3, mgp = c(2.5, 0.7, 0)
)

plot_slices(cp_springfall_slice_runoff, "Runoff")
plot_slices(cp_springfall_slice_snow, "Snow Density")
plot_slices(cp_springfall_slice_soil, "Soil Moisture")
plot_slices(cp_springfall_slice_temp, "Temperature")
plot_slices(cp_springfall_slice_precip, "Precipitation")
plot_slices(cp_springfall_slice_wind, "Wind Speed")
plot_slices(cp_springfall_slice_bird, "Total Bird Abundance")

mtext("Lag (weeks)", side = 1, outer = TRUE, line = 1, cex = 1.2)
mtext("Odds Ratio (95% CI)", side = 2, outer = TRUE, line = 1.5, cex = 1.2)

dev.off()


# Lag-Specific Odds Ratios — Season-Adjusted Timeseries Model

extract_lag_ORs_wide <- function(cp_obj, varname, varval = "0.5", lags = 0:4) {
  lag_labels <- paste0("lag", lags)
  or   <- cp_obj$matRRfit[varval, lag_labels]
  low  <- cp_obj$matRRlow[varval, lag_labels]
  high <- cp_obj$matRRhigh[varval, lag_labels]

  tibble(
    Predictor = varname,
    !!!setNames(as.list(or),   paste0("OR_Lag", lags)),
    !!!setNames(as.list(low),  paste0("Low_Lag", lags)),
    !!!setNames(as.list(high), paste0("High_Lag", lags)),
    !!!setNames(as.list(sprintf("%.2f (%.2f, %.2f)", or, low, high)), paste0("Lag ", lags))
  )
}

table_lag_ORs <- bind_rows(
  extract_lag_ORs_wide(cp_springfall_slice_runoff, "Runoff"),
  extract_lag_ORs_wide(cp_springfall_slice_soil,   "Soil Moisture"),
  extract_lag_ORs_wide(cp_springfall_slice_temp,   "Temperature"),
  extract_lag_ORs_wide(cp_springfall_slice_precip, "Precipitation"),
  extract_lag_ORs_wide(cp_springfall_slice_snow,   "Snow Density"),
  extract_lag_ORs_wide(cp_springfall_slice_wind,   "Wind Speed"),
  extract_lag_ORs_wide(cp_springfall_slice_bird,   "Total Bird Abundance")
)

lags <- 0:4
highlight_matrix <- sapply(lags, function(i) {
  table_lag_ORs[[paste0("Low_Lag", i)]] > 1 | table_lag_ORs[[paste0("High_Lag", i)]] < 1
})
colnames(highlight_matrix) <- paste0("Lag ", lags)

# Build Flextable

ft <- table_lag_ORs %>%
  select(Predictor, paste0("Lag ", lags)) %>%
  flextable() %>%
  set_header_labels(Predictor = "Predictor") %>%
  add_header_row(
    values = c("Predictor", "Odds Ratio (95% CI)"),
    colwidths = c(1, length(lags))
  ) %>%
  fontsize(size = 11, part = "all") %>%
  autofit() %>%
  align(align = "center", part = "all") %>%
  theme_booktabs()

# Bold significant lag cells
for (i in lags) {
  ft <- bold(ft, i = which(highlight_matrix[, paste0("Lag ", i)]),
             j = paste0("Lag ", i))
}

# Export to Word
doc <- read_docx() %>%
  body_add_flextable(ft) %>%
  body_add_par(
    "Table: Lag-specific odds ratios from the case-crossover stratified to Fall and Spring with DLNM crossbasis functions. Estimates represent odds ratios (95% CI) per 0.5 standard deviation increase in each environmental predictor. Statistically significant lags are shown in bold.",
    style = "Normal"
  )

print(doc, target = file.path(tables_folder, "Table_LagSpecific_OR_Casecrossover_Springfall.docx"))

ft


```


## Case-Crossover Models on dataframe stratfied to only spring 

### Crossbasis 

```{r, crossbasis spring}

# Leaf Area Index: ns(df = 3), ns(df = 1)
cb_spring_leaf_area_index <- crossbasis(
  x = as.matrix(case_spring[, paste0("leaf_area_index_Lag", 0:4)]),
  lag = c(0, 4),
  argvar = list(fun = "lin"),
  arglag = list(fun = "poly", degree = 1)
)

# Precipitation: ns(df = 3), ns(df = 1)
cb_spring_precipitation <- crossbasis(
  x = as.matrix(case_spring[, paste0("precipitation_Lag", 0:4)]),
  lag = c(0, 4),
  argvar = list(fun = "lin"),
  arglag = list(fun = "poly", degree = 1)
)

# Relative Humidity: ns(df = 3), poly(degree = 2)
cb_spring_relative_humidity <- crossbasis(
  x = as.matrix(case_spring[, paste0("relative_humidity_Lag", 0:4)]),
  lag = c(0, 4),
  argvar = list(fun = "lin"),
  arglag = list(fun = "poly", degree = 1)
)



# Runoff: ns(df = 1), poly(degree = 2)
cb_spring_runoff <- crossbasis(
  x = as.matrix(case_spring[, paste0("runoff_Lag", 0:4)]),
  lag = c(0, 4),
  argvar = list(fun = "lin"),
  arglag = list(fun = "poly", degree = 1)
)

# Shortwave Radiation: ns(df = 3), poly(degree = 4)
cb_spring_shortwave_radiation <- crossbasis(
  x = as.matrix(case_spring[, paste0("shortwave_radiation_Lag", 0:4)]),
  lag = c(0, 4),
  argvar = list(fun = "lin"),
  arglag = list(fun = "poly", degree = 1)
)

# Longwave Radiation: ns(df = 3), ns(df = 1)
cb_spring_longwave_radiation <- crossbasis(
  x = as.matrix(case_spring[, paste0("shortwave_radiation_Lag", 0:4)]),
  lag = c(0, 4),
  argvar = list(fun = "lin"),
  arglag = list(fun = "poly", degree = 1)
)

# Snow Density: ns(df = 1), poly(degree = 3)
cb_spring_snow_density <- crossbasis(
  x = as.matrix(case_spring[, paste0("snow_density_Lag", 0:4)]),
  lag = c(0, 4),
  argvar = list(fun = "lin"),
  arglag = list(fun = "poly", degree = 1)
)

# Soil Moisture: linear exposure, ns(df = 1)
cb_spring_soil_moisture <- crossbasis(
  x = as.matrix(case_spring[, paste0("soil_moisture_Lag", 0:4)]),
  lag = c(0, 4),
  argvar = list(fun = "lin"),
  arglag = list(fun = "poly", degree = 1)
)

# Temperature: ns(df = 3), poly(degree = 3)
cb_spring_temperature <- crossbasis(
  x = as.matrix(case_spring[, paste0("temperature_Lag", 0:4)]),
  lag = c(0, 4),
  argvar = list(fun = "lin"),
  arglag = list(fun = "poly", degree = 1)
)

# Bird Abundance: ns(df = 3), ns(df = 1)
cb_spring_total_bird_abundance <- crossbasis(
  x = as.matrix(case_spring[, paste0("total_bird_abundance_Lag", 0:4)]),
  lag = c(0, 4),
  argvar = list(fun = "lin"),
  arglag = list(fun = "poly", degree = 1)
)

# Wind Speed: ns(df = 1), poly(degree = 4)
cb_spring_wind_speed <- crossbasis(
  x = as.matrix(case_spring[, paste0("wind_speed_Lag", 0:4)]),
  lag = c(0, 4),
  argvar = list(fun = "lin"),
  arglag = list(fun = "poly", degree = 1)
)






```

### Spring Model (will not converge)

```{r, spring model}

spring_model <- clogit(
  outbreak_binary ~ 
    cb_spring_runoff +            
    cb_spring_snow_density +           
    cb_spring_soil_moisture +          
    cb_spring_temperature +           
    cb_spring_precipitation +         
    cb_spring_wind_speed +             
    cb_spring_total_bird_abundance +   
    strata(stratum_id),
  method = "efron",
  data = case_spring
)

summary(spring_model)

```

```{r, spring crosspred}
center_runoff <- mean(case_spring$runoff, na.rm = TRUE)
center_snow   <- mean(case_spring$snow_density, na.rm = TRUE)
center_soil   <- mean(case_spring$soil_moisture, na.rm = TRUE)
center_temp   <- mean(case_spring$temperature, na.rm = TRUE)
center_precip <- mean(case_spring$precipitation, na.rm = TRUE)
center_wind   <- mean(case_spring$wind_speed, na.rm = TRUE)
center_bird   <- mean(case_spring$total_bird_abundance, na.rm = TRUE)


cp_spring_runoff       <- crosspred(cb_spring_runoff, spring_model, center_runoff, at = 0.5, cumul = TRUE)
cp_spring_snow         <- crosspred(cb_spring_snow_density, spring_model, center_snow , at = 0.5, cumul = TRUE)
cp_spring_soil         <- crosspred(cb_spring_soil_moisture, spring_model, center_soil , at = 0.5, cumul = TRUE)
cp_spring_temp         <- crosspred(cb_spring_temperature, spring_model, center_temp    , at = 0.5, cumul = TRUE)
cp_spring_precip       <- crosspred(cb_spring_precipitation, spring_model, center_precip, at = 0.5, cumul = TRUE)
cp_spring_wind         <- crosspred(cb_spring_wind_speed, spring_model, center_wind, at = 0.5, cumul = TRUE)
cp_spring_bird         <- crosspred(cb_spring_total_bird_abundance, spring_model, center_bird  , at = 0.5, cumul = TRUE)

cp_spring_slice_runoff <- crosspred(cb_spring_runoff, spring_model, center_runoff, at = 0.5)
cp_spring_slice_snow   <- crosspred(cb_spring_snow_density, spring_model, center_snow, at = 0.5)
cp_spring_slice_soil   <- crosspred(cb_spring_soil_moisture, spring_model, center_soil, at = 0.5)
cp_spring_slice_temp   <- crosspred(cb_spring_temperature, spring_model, center_temp, at = 0.5)
cp_spring_slice_precip <- crosspred(cb_spring_precipitation, spring_model,  center_precip, at = 0.5)
cp_spring_slice_wind   <- crosspred(cb_spring_wind_speed, spring_model, center_wind, at = 0.5)
cp_spring_slice_bird   <- crosspred(cb_spring_total_bird_abundance, spring_model, center_bird, at = 0.5)

```

### Results from Spring Model (will not converge)

```{r, cumulative odds ratios spring fall}
# # ----------------------------
# # Cumulative OR Summary Table
# # ----------------------------
# cumulative_summary <- tibble(
#   Predictor = c("Runoff", "Snow Density", "Soil Moisture", "Temperature",
#                 "Precipitation", "Wind Speed", "Total Bird Abundance"),
#       OR = c(
#         cp_spring_runoff$allRRfit["0.5"],
#         cp_spring_snow$allRRfit["0.5"],
#         cp_spring_soil$allRRfit["0.5"],
#         cp_spring_temp$allRRfit["0.5"],
#         cp_spring_precip$allRRfit["0.5"],
#         cp_spring_wind$allRRfit["0.5"],
#         cp_spring_bird$allRRfit["0.5"]
#       ),
#       CI_low = c(
#         cp_spring_runoff$allRRlow["0.5"],
#         cp_spring_snow$allRRlow["0.5"],
#         cp_spring_soil$allRRlow["0.5"],
#         cp_spring_temp$allRRlow["0.5"],
#         cp_spring_precip$allRRlow["0.5"],
#         cp_spring_wind$allRRlow["0.5"],
#         cp_spring_bird$allRRlow["0.5"]
#       ),
#       CI_high = c(
#         cp_spring_runoff$allRRhigh["0.5"],
#         cp_spring_snow$allRRhigh["0.5"],
#         cp_spring_soil$allRRhigh["0.5"],
#         cp_spring_temp$allRRhigh["0.5"],
#         cp_spring_precip$allRRhigh["0.5"],
#         cp_spring_wind$allRRhigh["0.5"],
#         cp_spring_bird$allRRhigh["0.5"]
#       )
# ) %>%
#   mutate(`OR (95% CI)` = sprintf("%.2f (%.2f, %.2f)", OR, CI_low, CI_high)) %>%
#   select(Predictor, `OR (95% CI)`) %>%
#   gt() %>%
#   tab_header(
#     title = md("**Case-Crossover Model: Cumulative Odds Ratios**"),
#     subtitle = "Per 0.5 SD Increase"
#   ) %>%
#   cols_label(`OR (95% CI)` = "Odds Ratio (95% Confidence Interval)") %>%
#   tab_options(
#     table.font.size = 14,
#     heading.title.font.size = 16,
#     heading.subtitle.font.size = 14,
#     data_row.padding = px(6)
#   )
# 
# cumulative_summary
# 
# # ----------------------------
# # Plotting Functions
# # ----------------------------
# plot_cumulative <- function(cp, title) {
#   plot(cp, type = "cumulative", var = 0.5, ci = "area",
#        col = "black", ci.arg = list(col = rgb(0,0,0,0.1), border = NA),
#        lwd = 1.5, main = title, xlab = "", ylab = "", las = 1)
#   abline(h = 1, lty = 2, col = "gray40")
# }
# 
# plot_slices <- function(cp, title) {
#   plot(cp, type = "slices", var = 0.5, ci = "area",
#        col = "black", ci.arg = list(col = rgb(0,0,0,0.1), border = NA),
#        lwd = 1.3, main = title, xlab = "Lag ", ylab = "", las = 1)
#   abline(h = 1, lty = 2, col = "gray40")
# }
# 
# # ----------------------------
# # Save Cumulative Plot
# # ----------------------------
# png(file.path(figures_folder,"figure_cumulative_ORs_casecrossover_spring.png"), width = 11, height = 7, units = "in", res = 300)
# 
# par(
#   mfrow = c(2, 4),
#   mar = c(4.2, 4.2, 2.5, 1),
#   oma = c(3, 4, 0, 1),
#   cex.main = 1.2,
#   cex.lab = 1.2,
#   cex.axis = 1.1,
#   tcl = -0.3, mgp = c(2.5, 0.7, 0)
# )
# 
# plot_cumulative(cp_spring_runoff, "Runoff")
# plot_cumulative(cp_spring_snow, "Snow Density")
# plot_cumulative(cp_spring_soil, "Soil Moisture")
# plot_cumulative(cp_spring_temp, "Temperature")
# plot_cumulative(cp_spring_precip, "Precipitation")
# plot_cumulative(cp_spring_wind, "Wind Speed")
# plot_cumulative(cp_spring_bird, "Total Bird Abundance")
# 
# mtext("Lag (weeks)", side = 1, outer = TRUE, line = 0, cex = 1.2)
# mtext("Odds Ratio (95% CI)", side = 2, outer = TRUE, line = 0, cex = 1.2)
# 
# # ----------------------------
# # Save Slices Plot
# # ----------------------------
# png(file.path(figures_folder,"figure_slice_ORs_casecrossover_spring.png"), width = 11, height = 7, units = "in", res = 300)
# 
# par(
#   mfrow = c(2, 4),
#   mar = c(4.2, 4.2, 2.5, 1),
#   oma = c(3, 4, 0, 1),
#   cex.main = 1.2,
#   cex.lab = 1.2,
#   cex.axis = 1.1,
#   tcl = -0.3, mgp = c(2.5, 0.7, 0)
# )
# 
# plot_slices(cp_spring_slice_runoff, "Runoff")
# plot_slices(cp_spring_slice_snow, "Snow Density")
# plot_slices(cp_spring_slice_soil, "Soil Moisture")
# plot_slices(cp_spring_slice_temp, "Temperature")
# plot_slices(cp_spring_slice_precip, "Precipitation")
# plot_slices(cp_spring_slice_wind, "Wind Speed")
# plot_slices(cp_spring_slice_bird, "Total Bird Abundance")
# 
# dev.off()
# 
# mtext("Lag (weeks)", side = 1, outer = TRUE, line = 0, cex = 1.2)
# mtext("Odds Ratio (95% CI)", side = 2, outer = TRUE, line = 0, cex = 1.2)
# 
# 
# # Lag-Specific Odds Ratios — Season-Adjusted Timeseries Model
# 
# extract_lag_ORs_wide <- function(cp_obj, varname, varval = "0.5", lags = 0:4) {
#   lag_labels <- paste0("lag", lags)
#   or   <- cp_obj$matRRfit[varval, lag_labels]
#   low  <- cp_obj$matRRlow[varval, lag_labels]
#   high <- cp_obj$matRRhigh[varval, lag_labels]
# 
#   tibble(
#     Predictor = varname,
#     !!!setNames(as.list(or),   paste0("OR_Lag", lags)),
#     !!!setNames(as.list(low),  paste0("Low_Lag", lags)),
#     !!!setNames(as.list(high), paste0("High_Lag", lags)),
#     !!!setNames(as.list(sprintf("%.2f (%.2f, %.2f)", or, low, high)), paste0("Lag ", lags))
#   )
# }
# 
# table_lag_ORs <- bind_rows(
#   extract_lag_ORs_wide(cp_spring_slice_runoff, "Runoff"),
#   extract_lag_ORs_wide(cp_spring_slice_soil,   "Soil Moisture"),
#   extract_lag_ORs_wide(cp_spring_slice_temp,   "Temperature"),
#   extract_lag_ORs_wide(cp_spring_slice_precip, "Precipitation"),
#   extract_lag_ORs_wide(cp_spring_slice_snow,   "Snow Density"),
#   extract_lag_ORs_wide(cp_spring_slice_wind,   "Wind Speed"),
#   extract_lag_ORs_wide(cp_spring_slice_bird,   "Total Bird Abundance")
# )
# 
# lags <- 0:4
# highlight_matrix <- sapply(lags, function(i) {
#   table_lag_ORs[[paste0("Low_Lag", i)]] > 1 | table_lag_ORs[[paste0("High_Lag", i)]] < 1
# })
# colnames(highlight_matrix) <- paste0("Lag ", lags)
# 
# # Build Flextable
# 
# ft <- table_lag_ORs %>%
#   select(Predictor, paste0("Lag ", lags)) %>%
#   flextable() %>%
#   set_header_labels(Predictor = "Predictor") %>%
#   add_header_row(
#     values = c("Predictor", "Odds Ratio (95% CI)"),
#     colwidths = c(1, length(lags))
#   ) %>%
#   fontsize(size = 11, part = "all") %>%
#   autofit() %>%
#   align(align = "center", part = "all") %>%
#   theme_booktabs()
# 
# # Bold significant lag cells
# for (i in lags) {
#   ft <- bold(ft, i = which(highlight_matrix[, paste0("Lag ", i)]),
#              j = paste0("Lag ", i))
# }
# 
# # Export to Word
# doc <- read_docx() %>%
#   body_add_flextable(ft) %>%
#   body_add_par(
#     "Table: Lag-specific odds ratios from the case-crossover stratified to Spring with DLNM crossbasis functions. Estimates represent odds ratios (95% CI) per 0.5 standard deviation increase in each environmental predictor. Statistically significant lags are shown in bold.",
#     style = "Normal"
#   )
# 
# print(doc, target = file.path(tables_folder, "Table_LagSpecific_OR_Casecrossover_spring.docx"))
# 
# ft
```


## Timeseries 

### Read in Timeseries 

```{r, import timeseries dataframe}

# Timeseries
fishnet_timeseries_clean <- readRDS(file.path(objects_folder, "fishnet_timeseries_modeling.RDS"))


```

### Prepare for modeling

```{r, prepare}


timeseries_clean <- fishnet_timeseries_clean %>%
  arrange(zone_id, week)

timeseries_clean$season <- factor(timeseries_clean$season,
                                  levels = c("Fall", "Spring", "Winter", "Summer"))

timeseries_clean %>%
  group_by(zone_id) %>%
  summarise(
    has_feedlot = max(feedlot_count, na.rm = TRUE) > 0,
    has_outbreak = max(any_outbreak, na.rm = TRUE) > 0
  ) %>%
  count(has_feedlot, has_outbreak)

# Identify zones to keep: has a feedlot OR had an outbreak
zones_to_keep <- timeseries_clean %>%
  group_by(zone_id) %>%
  summarise(
    has_feedlot = max(feedlot_count, na.rm = TRUE) > 0,
    has_outbreak = max(any_outbreak, na.rm = TRUE) > 0
  ) %>%
  filter(has_feedlot | has_outbreak) %>%
  pull(zone_id)

timeseries_clean <- timeseries_clean %>%
  filter(zone_id %in% zones_to_keep)

cat("Zones retained:", length(zones_to_keep), "\n")
cat("Zones removed:", 3334 - length(zones_to_keep), "\n")

```

### Checks

```{r, timeseries checks}

# count cells by season

timeseries_clean %>%
count(season) %>%
  print()

case_cells <- timeseries_clean %>%
  filter(any_outbreak == 1)

case_cells %>%
count(season) %>%
  print()

```


## Scale

```{r, scale}

# Variables to log1p transform
log_vars <- c("total_bird_abundance", "runoff", "precipitation")

# Apply log1p transformation only if column exists
timeseries_clean <- timeseries_clean %>%
  mutate(across(all_of(log_vars), ~ log1p(.), .names = "{.col}"))

# All variables to scale 
vars_to_scale <- c(
  "total_bird_abundance", "leaf_area_index", "longwave_radiation", "runoff",
  "shortwave_radiation", "snow_density", "soil_moisture", 
  "temperature", "evapotranspiration", "precipitation", "wind_speed",
  "relative_humidity"
)

# Apply z-score scaling
timeseries_clean <- timeseries_clean %>%
  mutate(across(all_of(vars_to_scale), scale, .names = "{.col}"))


```

### Crossbasis terms (timeseries)

```{r, crossbasis ns}


# Crossbasis terms for selected variables using Lag0–Lag4

cb_relative_humidity <- crossbasis(
  timeseries_clean$relative_humidity,
  lag = 4,
  argvar = list(fun = "ns", df = 2),
  arglag = list(fun = "ns", df = 2)
)

cb_evap <- crossbasis(
  timeseries_clean$evapotranspiration,
  lag = 4,
  argvar = list(fun = "ns", df = 2),
  arglag = list(fun = "ns", df = 2)
)

cb_leaf_area_index <- crossbasis(
  timeseries_clean$leaf_area_index,
  lag = 4,
  argvar = list(fun = "ns", df = 2),
  arglag = list(fun = "ns", df = 2)
)

cb_runoff <- crossbasis(
  timeseries_clean$runoff,
  lag = 4,
  argvar = list(fun = "ns", df = 2),
  arglag = list(fun = "ns", df = 2)
)

cb_shortwave_radiation <- crossbasis(
  timeseries_clean$shortwave_radiation,
  lag = 4,
  argvar = list(fun = "ns", df = 2),
  arglag = list(fun = "ns", df = 2)
)

cb_longwave_radiation <- crossbasis(
  timeseries_clean$longwave_radiation,
  lag = 4,
  argvar = list(fun = "ns", df = 2),
  arglag = list(fun = "ns", df = 2)
)

cb_snow_density <- crossbasis(
  timeseries_clean$snow_density,
  lag = 4,
  argvar = list(fun = "ns", df = 2),
  arglag = list(fun = "ns", df = 2)
)

cb_soil_moisture <- crossbasis(
  timeseries_clean$soil_moisture,
  lag = 4,
  argvar = list(fun = "ns", df = 2),
  arglag = list(fun = "ns", df = 2)
)

cb_temperature <- crossbasis(
  timeseries_clean$temperature,
  lag = 4,
  argvar = list(fun = "ns", df = 2),
  arglag = list(fun = "ns", df = 2)
)

cb_precipitation <- crossbasis(
  timeseries_clean$precipitation,
  lag = 4,
  argvar = list(fun = "ns", df = 2),
  arglag = list(fun = "ns", df = 2)
)

cb_wind_speed <- crossbasis(
  timeseries_clean$wind_speed,
  lag = 4,
  argvar = list(fun = "ns", df = 2),
  arglag = list(fun = "ns", df = 2)
)

cb_bird_abundance <- crossbasis(
  timeseries_clean$total_bird_abundance,
  lag = 4,
  argvar = list(fun = "ns", df = 2),
  arglag = list(fun = "ns", df = 2)
)


```


## Model with Seasonal Fixed Effects


```{r, season fixed effect model}

season_timeseries_model <- bam(outbreak_binary ~
                   cb_precipitation +
                   cb_snow_density + 
                   cb_runoff +
                   cb_soil_moisture +
                   cb_bird_abundance +
                   cb_wind_speed +
                   cb_temperature +
                     s(lat, lon, bs ="ds", k =200) +
                    season, # fixed effect
                    family = nb(),
                    data = timeseries_clean,
                    method = "fREML",
                    discrete = TRUE)

summary(season_timeseries_model)

# Check number of cells analyzed 
# model frame = rows actually used (after NA dropping etc.)
mf <- model.frame(season_timeseries_model)

# get the original row indices used by bam()
idx <- as.integer(rownames(mf))

df_used <- timeseries_clean[idx, ]

# outcome
y <- df_used$outbreak_binary

# cell-weeks
pos_cell_weeks <- sum(y == 1, na.rm = TRUE)
neg_cell_weeks <- sum(y == 0, na.rm = TRUE)

# cells
case_cells <- unique(df_used$zone_id[y == 1])
n_case_cells <- length(case_cells)

all_cells <- unique(df_used$zone_id)
control_cells <- setdiff(all_cells, case_cells)
n_control_cells <- length(control_cells)

cat("=== Counts (rows used in bam()) ===\n")
cat("Positive cell-weeks: ", pos_cell_weeks, "\n", sep = "")
cat("Negative cell-weeks: ", neg_cell_weeks, "\n", sep = "")
cat("Case cells (ever positive): ", n_case_cells, "\n", sep = "")
cat("Control cells (never positive): ", n_control_cells, "\n", sep = "")
cat("Total unique cells: ", length(all_cells), "\n", sep = "")

# --- feedlot cells (ever have feedlot_count > 0 in any week) ---
feedlot_cells <- df_used |>
  dplyr::group_by(zone_id) |>
  dplyr::summarise(has_feedlot = any(feedlot_count > 0, na.rm = TRUE), .groups = "drop") |>
  dplyr::filter(has_feedlot) |>
  dplyr::pull(zone_id)

# --- feedlot control cells = feedlot_cells that are also never-positive cells ---
feedlot_control_cells <- intersect(feedlot_cells, control_cells)
n_feedlot_control_cells <- length(feedlot_control_cells)

# Optional: sanity checks
cat("Control cells (never positive): ", n_control_cells, "\n", sep = "")
cat("Feedlot cells (ever feedlot): ", length(feedlot_cells), "\n", sep = "")
cat("Feedlot control cells (feedlot, never positive): ", n_feedlot_control_cells, "\n", sep = "")

```


```{r, crosspred season timeseries model}


## Center values for each variable
center_precip <- mean(timeseries_clean$precipitation, na.rm = TRUE)
center_snow     <- mean(timeseries_clean$snow_density, na.rm = TRUE)
center_runoff <- mean(timeseries_clean$runoff, na.rm = TRUE)
center_soil   <- mean(timeseries_clean$soil_moisture, na.rm = TRUE)
center_bird   <- mean(timeseries_clean$total_bird_abundance, na.rm = TRUE)
center_wind   <- mean(timeseries_clean$wind_speed, na.rm = TRUE)
center_temp   <- mean(timeseries_clean$temperature, na.rm = TRUE)

## Cumulative ORs (per 0.5 SD increase)
cp_runoff   <- crosspred(cb_runoff,   season_timeseries_model, cen = center_runoff, at = 0.5, cumul = TRUE)
cp_soil     <- crosspred(cb_soil_moisture, season_timeseries_model, cen = center_soil, at = 0.5, cumul = TRUE)
cp_temp     <- crosspred(cb_temperature, season_timeseries_model, cen = center_temp, at = 0.5, cumul = TRUE)
cp_precip   <- crosspred(cb_precipitation, season_timeseries_model, cen = center_precip, at = 0.5, cumul = TRUE)
cp_snow       <- crosspred(cb_snow_density, season_timeseries_model, cen = center_snow, at = 0.5, cumul = TRUE)
cp_wind     <- crosspred(cb_wind_speed, season_timeseries_model, cen = center_wind, at = 0.5, cumul = TRUE)
cp_bird     <- crosspred(cb_bird_abundance, season_timeseries_model, cen = center_bird, at = 0.5, cumul = TRUE)

## Slice (lag-specific)
cp_runoff_slice   <- crosspred(cb_runoff,   season_timeseries_model, cen = center_runoff, at = 0.5)
cp_soil_slice     <- crosspred(cb_soil_moisture, season_timeseries_model, cen = center_soil, at = 0.5)
cp_temp_slice     <- crosspred(cb_temperature, season_timeseries_model, cen = center_temp, at = 0.5)
cp_precip_slice   <- crosspred(cb_precipitation, season_timeseries_model, cen = center_precip, at = 0.5)
cp_snow_slice       <- crosspred(cb_snow_density, season_timeseries_model, cen = center_snow, at = 0.5)
cp_wind_slice     <- crosspred(cb_wind_speed, season_timeseries_model, cen = center_wind, at = 0.5)
cp_bird_slice     <- crosspred(cb_bird_abundance, season_timeseries_model, cen = center_bird, at = 0.5)

```

### Results from Timeseries Model with Seasonal Fixed Effects


```{r, results season timeseries model}

cumulative_summary <- tibble(
  Predictor = c("Runoff", "Soil Moisture", "Temperature",
                "Precipitation", "Snow Density", "Wind Speed", "Total Bird Abundance"),
  OR = c(
    cp_runoff$allRRfit["0.5"],
    cp_soil$allRRfit["0.5"],
    cp_temp$allRRfit["0.5"],
    cp_precip$allRRfit["0.5"],
    cp_snow$allRRfit["0.5"],
    cp_wind$allRRfit["0.5"],
    cp_bird$allRRfit["0.5"]
  ),
  CI_low = c(
    cp_runoff$allRRlow["0.5"],
    cp_soil$allRRlow["0.5"],
    cp_temp$allRRlow["0.5"],
    cp_precip$allRRlow["0.5"],
    cp_snow$allRRlow["0.5"],
    cp_wind$allRRlow["0.5"],
    cp_bird$allRRlow["0.5"]
  ),
  CI_high = c(
    cp_runoff$allRRhigh["0.5"],
    cp_soil$allRRhigh["0.5"],
    cp_temp$allRRhigh["0.5"],
    cp_precip$allRRhigh["0.5"],
    cp_snow$allRRhigh["0.5"],
    cp_wind$allRRhigh["0.5"],
    cp_bird$allRRhigh["0.5"]
  )
) %>%
  mutate(`OR (95% CI)` = sprintf("%.2f (%.2f, %.2f)", OR, CI_low, CI_high)) %>%
  select(Predictor, `OR (95% CI)`) %>%
  gt() %>%
  tab_header(
    title = md("**Adjusted Timeseries Model: Cumulative Odds Ratios**"),
    subtitle = "Per 0.5 SD Increase, Adjusted for Season"
  ) %>%
  cols_label(`OR (95% CI)` = "Odds Ratio (95% Confidence Interval)") %>%
  tab_options(
    table.font.size = 14,
    heading.title.font.size = 16,
    heading.subtitle.font.size = 14,
    data_row.padding = px(6)
  )

cumulative_summary

plot_cumulative <- function(cp, title) {
  plot(cp, type = "cumulative", var = 0.5, ci = "area",
       col = "black", ci.arg = list(col = rgb(0,0,0,0.1), border = NA),
       lwd = 1.5, main = title, xlab = "", ylab = "", las = 1)
  abline(h = 1, lty = 2, col = "gray40")
}

plot_slices <- function(cp, title) {
  plot(cp, type = "slices", var = 0.5, ci = "area",
       col = "black", ci.arg = list(col = rgb(0,0,0,0.1), border = NA),
       lwd = 1.3, main = title, xlab = "", ylab = "", las = 1)
  abline(h = 1, lty = 2, col = "gray40")
}

png(file.path(figures_folder, "figure_cumulative_ORs_season_fixed_efffect.png"),
    width = 11, height = 7, units = "in", res = 300)

par(
  mfrow = c(2, 4),
  mar = c(4.2, 4.2, 2.5, 1),
  oma = c(3, 4, 0, 1),
  cex.main = 1.2,
  cex.lab = 1.2,
  cex.axis = 1.1,
  tcl = -0.3, mgp = c(2.5, 0.7, 0)
)


plot_cumulative(cp_runoff, "Runoff")
plot_cumulative(cp_soil,   "Soil Moisture")
plot_cumulative(cp_temp,   "Temperature")
plot_cumulative(cp_precip, "Precipitation")
plot_cumulative(cp_snow,     "Snow Density")
plot_cumulative(cp_wind,   "Wind Speed")
plot_cumulative(cp_bird,   "Total Bird Abundance")

mtext("Lag (weeks)", side = 1, outer = TRUE, line = 0, cex = 1.2)
mtext("Odds Ratio (95% CI)", side = 2, outer = TRUE, line = 0, cex = 1.2)

dev.off()

png(file.path(figures_folder, "figure_slice_ORs_season_fixed_efffect.png"),
    width = 11, height = 7, units = "in", res = 300)

par(
  mfrow = c(2, 4),
  mar = c(4.2, 4.2, 2.5, 1),
  oma = c(3, 4, 0, 1),
  cex.main = 1.2,
  cex.lab = 1.2,
  cex.axis = 1.1,
  tcl = -0.3, mgp = c(2.5, 0.7, 0)
)


plot_slices(cp_runoff_slice, "Runoff")
plot_slices(cp_soil_slice,   "Soil Moisture")
plot_slices(cp_temp_slice,   "Temperature")
plot_slices(cp_precip_slice, "Precipitation")
plot_slices(cp_snow_slice,     "Snow Density")
plot_slices(cp_wind_slice,   "Wind Speed")
plot_slices(cp_bird_slice,   "Total Bird Abundance")

mtext("Lag (weeks)", side = 1, outer = TRUE, line = 0, cex = 1.2)
mtext("Odds Ratio (95% CI)", side = 2, outer = TRUE, line = 0, cex = 1.2)

dev.off()

# Lag-Specific Odds Ratios — Season-Adjusted Timeseries Model

extract_lag_ORs_wide <- function(cp_obj, varname, varval = "0.5", lags = 0:4) {
  lag_labels <- paste0("lag", lags)
  or   <- cp_obj$matRRfit[varval, lag_labels]
  low  <- cp_obj$matRRlow[varval, lag_labels]
  high <- cp_obj$matRRhigh[varval, lag_labels]

  tibble(
    Predictor = varname,
    !!!setNames(as.list(or),   paste0("OR_Lag", lags)),
    !!!setNames(as.list(low),  paste0("Low_Lag", lags)),
    !!!setNames(as.list(high), paste0("High_Lag", lags)),
    !!!setNames(as.list(sprintf("%.2f (%.2f, %.2f)", or, low, high)), paste0("Lag ", lags))
  )
}

table_lag_ORs <- bind_rows(
  extract_lag_ORs_wide(cp_runoff_slice, "Runoff"),
  extract_lag_ORs_wide(cp_soil_slice,   "Soil Moisture"),
  extract_lag_ORs_wide(cp_temp_slice,   "Temperature"),
  extract_lag_ORs_wide(cp_precip_slice, "Precipitation"),
  extract_lag_ORs_wide(cp_snow_slice,   "Snow Density"),
  extract_lag_ORs_wide(cp_wind_slice,   "Wind Speed"),
  extract_lag_ORs_wide(cp_bird_slice,   "Total Bird Abundance")
)

lags <- 0:4
highlight_matrix <- sapply(lags, function(i) {
  table_lag_ORs[[paste0("Low_Lag", i)]] > 1 | table_lag_ORs[[paste0("High_Lag", i)]] < 1
})
colnames(highlight_matrix) <- paste0("Lag ", lags)

# Build Flextable

ft <- table_lag_ORs %>%
  select(Predictor, paste0("Lag ", lags)) %>%
  flextable() %>%
  set_header_labels(Predictor = "Predictor") %>%
  add_header_row(
    values = c("Predictor", "Odds Ratio (95% CI)"),
    colwidths = c(1, length(lags))
  ) %>%
  fontsize(size = 11, part = "all") %>%
  autofit() %>%
  align(align = "center", part = "all") %>%
  theme_booktabs()

# Bold significant lag cells
for (i in lags) {
  ft <- bold(ft, i = which(highlight_matrix[, paste0("Lag ", i)]),
             j = paste0("Lag ", i))
}

# Export to Word
doc <- read_docx() %>%
  body_add_flextable(ft) %>%
  body_add_par(
    "Table: Lag-specific odds ratios from the adjusted time series model with DLNM crossbasis functions and season fixed effects. Estimates represent odds ratios (95% CI) per 0.5 standard deviation increase in each environmental predictor. Statistically significant lags are shown in bold.",
    style = "Normal"
  )

print(doc, target = file.path(tables_folder, "Table_LagSpecific_OR_SeasonAdjusted_Timeseries.docx"))

ft


```

# Astronomical Season Fixed Effect

```{r, astronomical season model}


# Astronomical season variable based on week number
# Spring: March 20 ≈ week 12 to June 20 ≈ week 25
# Summer: June 21 ≈ week 25 to Sept 22 ≈ week 39
# Fall: Sept 22 ≈ week 39 to Dec 21 ≈ week 51
# Winter: Dec 21 ≈ week 51 to March 20 ≈ week 12

timeseries_clean <- timeseries_clean %>%
  mutate(
    season_astro = case_when(
      week >= 12 & week < 25 ~ "Spring",
      week >= 25 & week < 39 ~ "Summer",
      week >= 39 & week < 51 ~ "Fall",
      TRUE ~ "Winter"
    ),
    season_astro = factor(season_astro, levels = c("Winter", "Spring", "Summer", "Fall"))
  )

# Check distribution
timeseries_clean %>%
  count(season_astro)

# Run model with astronomical seasons
season_timeseries_model <- bam(outbreak_binary ~
                   cb_precipitation +
                   cb_snow_density + 
                   cb_runoff +
                   cb_soil_moisture +
                   cb_bird_abundance +
                   cb_wind_speed +
                   cb_temperature +
                   s(lat, lon, bs = "ds", k = 200) +
                   season_astro,
                   family = nb(),
                   data = timeseries_clean,
                   method = "fREML",
                   discrete = TRUE)

summary(season_timeseries_model)

# Check number of cells analyzed 
mf <- model.frame(season_timeseries_model)
idx <- as.integer(rownames(mf))
df_used <- timeseries_clean[idx, ]

y <- df_used$outbreak_binary

pos_cell_weeks <- sum(y == 1, na.rm = TRUE)
neg_cell_weeks <- sum(y == 0, na.rm = TRUE)

case_cells <- unique(df_used$zone_id[y == 1])
n_case_cells <- length(case_cells)
all_cells <- unique(df_used$zone_id)
control_cells <- setdiff(all_cells, case_cells)
n_control_cells <- length(control_cells)

cat("=== Counts (rows used in bam()) ===\n")
cat("Positive cell-weeks: ", pos_cell_weeks, "\n", sep = "")
cat("Negative cell-weeks: ", neg_cell_weeks, "\n", sep = "")
cat("Case cells (ever positive): ", n_case_cells, "\n", sep = "")
cat("Control cells (never positive): ", n_control_cells, "\n", sep = "")
cat("Total unique cells: ", length(all_cells), "\n", sep = "")

```

```{r, astronomical cross predict}


## Center values for each variable
center_precip <- mean(timeseries_clean$precipitation, na.rm = TRUE)
center_snow     <- mean(timeseries_clean$snow_density, na.rm = TRUE)
center_runoff <- mean(timeseries_clean$runoff, na.rm = TRUE)
center_soil   <- mean(timeseries_clean$soil_moisture, na.rm = TRUE)
center_bird   <- mean(timeseries_clean$total_bird_abundance, na.rm = TRUE)
center_wind   <- mean(timeseries_clean$wind_speed, na.rm = TRUE)
center_temp   <- mean(timeseries_clean$temperature, na.rm = TRUE)

## Cumulative ORs (per 0.5 SD increase)
cp_runoff   <- crosspred(cb_runoff,   season_timeseries_model, cen = center_runoff, at = 0.5, cumul = TRUE)
cp_soil     <- crosspred(cb_soil_moisture, season_timeseries_model, cen = center_soil, at = 0.5, cumul = TRUE)
cp_temp     <- crosspred(cb_temperature, season_timeseries_model, cen = center_temp, at = 0.5, cumul = TRUE)
cp_precip   <- crosspred(cb_precipitation, season_timeseries_model, cen = center_precip, at = 0.5, cumul = TRUE)
cp_snow       <- crosspred(cb_snow_density, season_timeseries_model, cen = center_snow, at = 0.5, cumul = TRUE)
cp_wind     <- crosspred(cb_wind_speed, season_timeseries_model, cen = center_wind, at = 0.5, cumul = TRUE)
cp_bird     <- crosspred(cb_bird_abundance, season_timeseries_model, cen = center_bird, at = 0.5, cumul = TRUE)

## Slice (lag-specific)
cp_runoff_slice   <- crosspred(cb_runoff,   season_timeseries_model, cen = center_runoff, at = 0.5)
cp_soil_slice     <- crosspred(cb_soil_moisture, season_timeseries_model, cen = center_soil, at = 0.5)
cp_temp_slice     <- crosspred(cb_temperature, season_timeseries_model, cen = center_temp, at = 0.5)
cp_precip_slice   <- crosspred(cb_precipitation, season_timeseries_model, cen = center_precip, at = 0.5)
cp_snow_slice       <- crosspred(cb_snow_density, season_timeseries_model, cen = center_snow, at = 0.5)
cp_wind_slice     <- crosspred(cb_wind_speed, season_timeseries_model, cen = center_wind, at = 0.5)
cp_bird_slice     <- crosspred(cb_bird_abundance, season_timeseries_model, cen = center_bird, at = 0.5)

```


### Results from Timeseries Model with Astronomical Seasonal Fixed Effects


```{r, results astro season timeseries model}

cumulative_summary <- tibble(
  Predictor = c("Runoff", "Soil Moisture", "Temperature",
                "Precipitation", "Snow Density", "Wind Speed", "Total Bird Abundance"),
  OR = c(
    cp_runoff$allRRfit["0.5"],
    cp_soil$allRRfit["0.5"],
    cp_temp$allRRfit["0.5"],
    cp_precip$allRRfit["0.5"],
    cp_snow$allRRfit["0.5"],
    cp_wind$allRRfit["0.5"],
    cp_bird$allRRfit["0.5"]
  ),
  CI_low = c(
    cp_runoff$allRRlow["0.5"],
    cp_soil$allRRlow["0.5"],
    cp_temp$allRRlow["0.5"],
    cp_precip$allRRlow["0.5"],
    cp_snow$allRRlow["0.5"],
    cp_wind$allRRlow["0.5"],
    cp_bird$allRRlow["0.5"]
  ),
  CI_high = c(
    cp_runoff$allRRhigh["0.5"],
    cp_soil$allRRhigh["0.5"],
    cp_temp$allRRhigh["0.5"],
    cp_precip$allRRhigh["0.5"],
    cp_snow$allRRhigh["0.5"],
    cp_wind$allRRhigh["0.5"],
    cp_bird$allRRhigh["0.5"]
  )
) %>%
  mutate(`OR (95% CI)` = sprintf("%.2f (%.2f, %.2f)", OR, CI_low, CI_high)) %>%
  select(Predictor, `OR (95% CI)`) %>%
  gt() %>%
  tab_header(
    title = md("**Adjusted Timeseries Model: Cumulative Odds Ratios**"),
    subtitle = "Per 0.5 SD Increase, Adjusted for Season"
  ) %>%
  cols_label(`OR (95% CI)` = "Odds Ratio (95% Confidence Interval)") %>%
  tab_options(
    table.font.size = 14,
    heading.title.font.size = 16,
    heading.subtitle.font.size = 14,
    data_row.padding = px(6)
  )

cumulative_summary

plot_cumulative <- function(cp, title) {
  plot(cp, type = "cumulative", var = 0.5, ci = "area",
       col = "black", ci.arg = list(col = rgb(0,0,0,0.1), border = NA),
       lwd = 1.5, main = title, xlab = "", ylab = "", las = 1)
  abline(h = 1, lty = 2, col = "gray40")
}

plot_slices <- function(cp, title) {
  plot(cp, type = "slices", var = 0.5, ci = "area",
       col = "black", ci.arg = list(col = rgb(0,0,0,0.1), border = NA),
       lwd = 1.3, main = title, xlab = "", ylab = "", las = 1)
  abline(h = 1, lty = 2, col = "gray40")
}

png(file.path(figures_folder, "figure_cumulative_ORs_season_fixed_efffect.png"),
    width = 11, height = 7, units = "in", res = 300)

par(
  mfrow = c(2, 4),
  mar = c(4.2, 4.2, 2.5, 1),
  oma = c(3, 4, 0, 1),
  cex.main = 1.2,
  cex.lab = 1.2,
  cex.axis = 1.1,
  tcl = -0.3, mgp = c(2.5, 0.7, 0)
)


plot_cumulative(cp_runoff, "Runoff")
plot_cumulative(cp_soil,   "Soil Moisture")
plot_cumulative(cp_temp,   "Temperature")
plot_cumulative(cp_precip, "Precipitation")
plot_cumulative(cp_snow,     "Snow Density")
plot_cumulative(cp_wind,   "Wind Speed")
plot_cumulative(cp_bird,   "Total Bird Abundance")

mtext("Lag (weeks)", side = 1, outer = TRUE, line = 0, cex = 1.2)
mtext("Odds Ratio (95% CI)", side = 2, outer = TRUE, line = 0, cex = 1.2)

dev.off()

png(file.path(figures_folder, "figure_slice_ORs_astro_season_fixed_efffect.png"),
    width = 11, height = 7, units = "in", res = 300)

par(
  mfrow = c(2, 4),
  mar = c(4.2, 4.2, 2.5, 1),
  oma = c(3, 4, 0, 1),
  cex.main = 1.2,
  cex.lab = 1.2,
  cex.axis = 1.1,
  tcl = -0.3, mgp = c(2.5, 0.7, 0)
)


plot_slices(cp_runoff_slice, "Runoff")
plot_slices(cp_soil_slice,   "Soil Moisture")
plot_slices(cp_temp_slice,   "Temperature")
plot_slices(cp_precip_slice, "Precipitation")
plot_slices(cp_snow_slice,     "Snow Density")
plot_slices(cp_wind_slice,   "Wind Speed")
plot_slices(cp_bird_slice,   "Total Bird Abundance")

mtext("Lag (weeks)", side = 1, outer = TRUE, line = 0, cex = 1.2)
mtext("Odds Ratio (95% CI)", side = 2, outer = TRUE, line = 0, cex = 1.2)

dev.off()

# Lag-Specific Odds Ratios — Season-Adjusted Timeseries Model

extract_lag_ORs_wide <- function(cp_obj, varname, varval = "0.5", lags = 0:4) {
  lag_labels <- paste0("lag", lags)
  or   <- cp_obj$matRRfit[varval, lag_labels]
  low  <- cp_obj$matRRlow[varval, lag_labels]
  high <- cp_obj$matRRhigh[varval, lag_labels]

  tibble(
    Predictor = varname,
    !!!setNames(as.list(or),   paste0("OR_Lag", lags)),
    !!!setNames(as.list(low),  paste0("Low_Lag", lags)),
    !!!setNames(as.list(high), paste0("High_Lag", lags)),
    !!!setNames(as.list(sprintf("%.2f (%.2f, %.2f)", or, low, high)), paste0("Lag ", lags))
  )
}

table_lag_ORs <- bind_rows(
  extract_lag_ORs_wide(cp_runoff_slice, "Runoff"),
  extract_lag_ORs_wide(cp_soil_slice,   "Soil Moisture"),
  extract_lag_ORs_wide(cp_temp_slice,   "Temperature"),
  extract_lag_ORs_wide(cp_precip_slice, "Precipitation"),
  extract_lag_ORs_wide(cp_snow_slice,   "Snow Density"),
  extract_lag_ORs_wide(cp_wind_slice,   "Wind Speed"),
  extract_lag_ORs_wide(cp_bird_slice,   "Total Bird Abundance")
)

lags <- 0:4
highlight_matrix <- sapply(lags, function(i) {
  table_lag_ORs[[paste0("Low_Lag", i)]] > 1 | table_lag_ORs[[paste0("High_Lag", i)]] < 1
})
colnames(highlight_matrix) <- paste0("Lag ", lags)

# Build Flextable

ft <- table_lag_ORs %>%
  select(Predictor, paste0("Lag ", lags)) %>%
  flextable() %>%
  set_header_labels(Predictor = "Predictor") %>%
  add_header_row(
    values = c("Predictor", "Odds Ratio (95% CI)"),
    colwidths = c(1, length(lags))
  ) %>%
  fontsize(size = 11, part = "all") %>%
  autofit() %>%
  align(align = "center", part = "all") %>%
  theme_booktabs()

# Bold significant lag cells
for (i in lags) {
  ft <- bold(ft, i = which(highlight_matrix[, paste0("Lag ", i)]),
             j = paste0("Lag ", i))
}

# Export to Word
doc <- read_docx() %>%
  body_add_flextable(ft) %>%
  body_add_par(
    "Table: Lag-specific odds ratios from the adjusted time series model with DLNM crossbasis functions and astronomical season fixed effects. Estimates represent odds ratios (95% CI) per 0.5 standard deviation increase in each environmental predictor. Statistically significant lags are shown in bold.",
    style = "Normal"
  )

print(doc, target = file.path(tables_folder, "Table_LagSpecific_OR_Astronomical_SeasonAdjusted_Timeseries.docx"))

ft


```