---
title: "Join Data Sources (Daily Resolution)"
author: "Fintan Mooney"
date: "2026-05-12"
output: html_document
---

# NOTE: This script requires confidential HPAI spillover point-location data
# (not publicly available). The processed grid-cell level output is provided
# in data/ and downstream analyses can be run without this file.

# Joins USDA spillover events (by date_confirmed) to 10x10km grid cells covering
# Minnesota at DAILY resolution for Nov 15 2021 – Dec 31 2022!
# Brings in daily ERA5-Land met, linearly-interpolated daily eBird abundance,
# MPCA feedlots, and USGS landcover.
# Produces the base zone × date panel used to build the daily timeseries and
# time-stratified case-crossover datasets.

# Major revision (May 2026): all temporal operations moved from weekly to daily
# to address GeoHealth reviewer comments. Wild bird abundance is linearly
# interpolated within each zone between weekly eBird Status & Trends midpoints
# to produce a smooth daily series. The Nov–Dec 2021 leading window is required
# so that DLNM lag 0–28 windows for early-2022 cases stay fully observed.

```{r 0a-initiate, include=FALSE}
# 0a. Declare root directory, folder locations, and load packages + helpers
rm(list = ls())
project.folder = paste0(print(here::here()), '/')
source(paste0(project.folder, 'create_folder_structure.R'))
source(paste0(functions.folder, 'script_initiate.R'))
```

## 1. Load 10 km fishnet geometry

```{r 1a-load-fishnet}
# 1a. Single feature per zone — geometry only (one-to-many join key)
fishnet_geom <- st_read(paste0(spatial_data, "/fishnet_10km_timeseries.gpkg")) %>%
  distinct(zone_id, geom)
```

## 2. Build the zone × date scaffold (Nov 15 2021 – Dec 31 2022)

```{r 2a-build-scaffold}
# 2a. Full crossing of zones × days. The Nov–Dec 2021 leading window is
#     required so the DLNM lag 0–28 windows for early-2022 cases stay fully
#     observed (see 02_code/2a_data_prep/a_00_era5_land_gee.js (or a_00_era5_land_rgee.R)).
study_start <- as.Date("2021-11-15")
study_end   <- as.Date("2022-12-31")

date_seq <- seq(study_start, study_end, by = "day")

zone_ids <- fishnet_geom %>% st_drop_geometry() %>% pull(zone_id) %>% unique()

# Full crossing — one row per zone per day
daily_fishnet <- expand.grid(
  zone_id = zone_ids,
  date    = date_seq,
  stringsAsFactors = FALSE
) %>%
  mutate(
    year       = lubridate::year(date),
    month      = lubridate::month(date),
    iso_week   = lubridate::isoweek(date),
    iso_year   = lubridate::isoyear(date),
    dow        = lubridate::wday(date, label = TRUE, week_start = 1),
    doy        = lubridate::yday(date)
  ) %>%
  as.data.table()

setkey(daily_fishnet, zone_id, date)

cat("Daily fishnet scaffold:\n")
cat("  Zones:", length(zone_ids), "\n")
cat("  Days :", length(date_seq), "\n")
cat("  Rows :", nrow(daily_fishnet), "\n")

```

## 3. Load and join spillover events (by date_confirmed)

```{r 3a-spillover-data}

spillover_file <- file.path(spillover_data, "hpai-outbreaks-clean-placeholder-032124.csv")
if (!file.exists(spillover_file)) {
  stop("This script needs the confidential spillover data (hpai-outbreaks-clean-placeholder-032124.csv). See 01_data/1b_outcome_data/README.md for how to get it.")
}
spillover_raw <- read_csv(spillover_file)

# Filter to MN 2022 (case window — leading 2021 days are zero-outbreak by design)
spillover_mn <- spillover_raw %>%
  filter(state == "MN",
         dplyr::between(date_confirmed, as.Date("2022-01-01"), as.Date("2022-12-31")))

# If date_observed / date_sampled exist, attach for downstream sensitivity analysis;
# date_confirmed remains the primary stratum date for consistency with the submitted
# manuscript and prior USDA APHIS-based HPAI literature.
date_cols_present <- intersect(c("date_observed", "date_sampled", "date_analyzed"),
                               names(spillover_mn))
if (length(date_cols_present) > 0) {
  cat("Earlier-date columns present (kept for sensitivity):", paste(date_cols_present, collapse = ", "), "\n")
}

# Convert to sf and join to fishnet
spillover_sf <- spillover_mn %>%
  st_as_sf(coords = c("lon", "lat"), crs = 4326) %>%
  st_transform(crs = st_crs(fishnet_geom))

outbreaks_in_cells <- st_join(spillover_sf, fishnet_geom, join = st_within) %>%
  st_drop_geometry()

# Aggregate by zone × date
outbreak_daily <- outbreaks_in_cells %>%
  group_by(zone_id, date = date_confirmed) %>%
  summarise(
    outbreak_count = n(),
    outbreak_id    = paste(outbreak_id, collapse = ", "),
    .groups = "drop"
  ) %>%
  mutate(outbreak_binary = 1L) %>%
  as.data.table()

# Zone-level summaries (time-invariant tags)
any_outbreak_zones <- outbreak_daily[, .(any_outbreak = 1L), by = zone_id]
outbreak_dates_by_zone <- outbreak_daily[
  , .(outbreak_date = paste(sort(unique(as.character(date))), collapse = ", ")),
  by = zone_id
]

# Merge into daily scaffold
setkey(outbreak_daily, zone_id, date)
daily_fishnet <- merge(daily_fishnet, outbreak_daily,
                       by = c("zone_id", "date"), all.x = TRUE)
daily_fishnet <- merge(daily_fishnet, any_outbreak_zones, by = "zone_id", all.x = TRUE)
daily_fishnet <- merge(daily_fishnet, outbreak_dates_by_zone, by = "zone_id", all.x = TRUE)

daily_fishnet[, `:=`(
  outbreak_count  = fcoalesce(outbreak_count, 0L),
  outbreak_binary = fcoalesce(outbreak_binary, 0L),
  any_outbreak    = fcoalesce(any_outbreak, 0L),
  outbreak_id     = fcoalesce(outbreak_id, NA_character_),
  outbreak_date   = fcoalesce(outbreak_date, NA_character_)
)]

cat("\nSpillover join:\n")
cat("  Events in MN 2022:", nrow(spillover_mn), "\n")
cat("  Case zone-DAYS   :", uniqueN(outbreak_daily, by = c("zone_id", "date")), "\n")
cat("  Unique case zones:", uniqueN(outbreak_daily, by = "zone_id"), "\n")

```

## 4. Load DAILY ERA5-Land meteorological data

```{r 4a-load-met-daily}

# 2022 daily + Nov–Dec 2021 trailing daily (produced by 02_code/2a_data_prep/a_00_era5_land_gee.js (or a_00_era5_land_rgee.R))
met_files <- c(
  file.path(env_data, "mn_daily_era5land_2021_novdec.csv"),
  file.path(env_data, "mn_daily_era5land_2022.csv")
)
missing <- met_files[!file.exists(met_files)]
if (length(missing) > 0) {
  stop("Missing daily met file(s):\n  ",
       paste(missing, collapse = "\n  "),
       "\nRun 02_code/2a_data_prep/a_00_era5_land_gee.js (or a_00_era5_land_rgee.R) first.")
}

met_daily <- rbindlist(lapply(met_files, fread))

# Parse the date column (GEE exports as YYYYMMdd integers)
met_daily[, date := as.Date(as.character(date), format = "%Y%m%d")]

# Snow density: ERA5-Land returns ~99.9999 when no snow is present (default fill);
# this is a known artefact. Set to 0 for cells with no snow.
met_daily[, snow_density := fifelse(snow_density <= 100, 0, snow_density)]

# Rename runoff_sum → runoff for downstream code consistency
if ("runoff_sum" %in% names(met_daily)) setnames(met_daily, "runoff_sum", "runoff")

# Met columns to retain
met_cols <- c("temperature", "dewpoint", "relative_humidity", "wind_speed",
              "soil_moisture", "snow_density", "precipitation", "runoff",
              "evapotranspiration", "leaf_area_index", "shortwave_radiation",
              "longwave_radiation", "surface_pressure")
met_daily <- met_daily[, c("zone_id", "date", ..met_cols)]
setkey(met_daily, zone_id, date)

daily_fishnet <- merge(daily_fishnet, met_daily, by = c("zone_id", "date"), all.x = TRUE)

cat("\nMet join:\n")
cat("  Rows with any NA met var:",
    sum(daily_fishnet[, rowSums(is.na(.SD)), .SDcols = met_cols] > 0), "\n")

```

## 5. Load weekly eBird and linearly interpolate to daily within each zone

```{r 5a-bird-abundance-interpolation}

# eBird Status & Trends weekly outbreak-level estimates
bird_outbreak <- read_csv(file.path(bird_abundance_data,
                                    "mn-outbreak-weekly-bird-estimate-2022-012926.csv"))

# Species groupings used throughout the manuscript
anseriformes <- c(
  "mallar3", "cangoo", "wooduc", "rinduc", "comgol", "norsho",
  "buffle", "lessca", "rebmer", "gnwtea", "gadwal", "ambduc",
  "rudduc", "rethaw", "truswa", "commer"
)
predators <- c("buwtea", "grhowl", "amwpel", "baleag", "brdowl")

# Pivot wide and sum within group per (outbreak_id, mmwr_week)
bird_wide <- bird_outbreak %>%
  pivot_wider(
    id_cols = c(outbreak_id, mmwr_week, week, week_outbreak),
    names_from = species_name,
    values_from = c(abd_avg, abd_sum),
    names_glue = "{.value}_{species_name}"
  ) %>%
  mutate(
    total_bird_abundance = rowSums(across(starts_with("abd_avg_") &
                                          !matches("anseriformes|predators")), na.rm = TRUE),
    anseriformes = rowSums(across(all_of(paste0("abd_avg_", anseriformes))), na.rm = TRUE),
    predators    = rowSums(across(all_of(paste0("abd_avg_", predators))),    na.rm = TRUE)
  ) %>%
  select(outbreak_id, mmwr_week, total_bird_abundance, anseriformes, predators)

# Map outbreak_id → zone_id (so we can collapse multi-outbreak zones to one zone-week)
zone_outbreak_map <- outbreaks_in_cells %>%
  select(outbreak_id, zone_id) %>%
  distinct()

bird_zone_week <- bird_wide %>%
  inner_join(zone_outbreak_map, by = "outbreak_id") %>%
  group_by(zone_id, mmwr_week) %>%
  summarise(
    total_bird_abundance = mean(total_bird_abundance, na.rm = TRUE),
    anseriformes         = mean(anseriformes,         na.rm = TRUE),
    predators            = mean(predators,            na.rm = TRUE),
    .groups = "drop"
  ) %>%
  as.data.table()

# Negative-control zones (feedlot zones with no outbreak) — feedlot-level weekly bird series
bird_neg_path <- file.path(bird_abundance_data, "all-mn-feedlot-species-abundance-070125.csv")
bird_neg_raw  <- fread(bird_neg_path)

week_cols <- setdiff(names(bird_neg_raw), c("zone_id", "species"))
bird_neg_long <- melt(
  bird_neg_raw,
  id.vars = c("zone_id", "species"),
  measure.vars = week_cols,
  variable.name = "mmwr_week",
  value.name = "abundance"
)
bird_neg_long[, mmwr_week := as.integer(gsub("week_", "", mmwr_week))]
bird_neg_long[, group := fifelse(species %in% anseriformes, "anseriformes",
                          fifelse(species %in% predators, "predators", "other"))]

# Average across feedlot pixels within zone × week × species, then sum across species in group
bird_neg_zone_week <- bird_neg_long[
  , .(abundance = mean(abundance, na.rm = TRUE)),
  by = .(zone_id, mmwr_week, species, group)
]
bird_neg_group <- bird_neg_zone_week[
  , .(value = sum(abundance, na.rm = TRUE)),
  by = .(zone_id, mmwr_week, group)
]
bird_neg_wide <- dcast(bird_neg_group, zone_id + mmwr_week ~ group,
                      value.var = "value", fill = 0)
if (!"anseriformes" %in% names(bird_neg_wide)) bird_neg_wide[, anseriformes := 0]
if (!"predators"    %in% names(bird_neg_wide)) bird_neg_wide[, predators    := 0]

bird_neg_total <- bird_neg_zone_week[
  , .(total_bird_abundance = sum(abundance, na.rm = TRUE)),
  by = .(zone_id, mmwr_week)
]
bird_neg_wide <- merge(bird_neg_wide, bird_neg_total, by = c("zone_id", "mmwr_week"), all = TRUE)
bird_neg_wide[is.na(total_bird_abundance), total_bird_abundance := 0]
bird_neg_wide[is.na(anseriformes), anseriformes := 0]
bird_neg_wide[is.na(predators),    predators    := 0]

# Combine case-zone weekly bird and negative-control weekly bird into one zone-week series.
# Case zones get their outbreak-derived weekly values; negative-control zones get feedlot-derived.
case_zone_ids <- unique(bird_zone_week$zone_id)
bird_zone_week_all <- rbind(
  bird_zone_week[, .(zone_id, mmwr_week, total_bird_abundance, anseriformes, predators)],
  bird_neg_wide[!zone_id %in% case_zone_ids,
                .(zone_id, mmwr_week, total_bird_abundance, anseriformes, predators)]
)
setkey(bird_zone_week_all, zone_id, mmwr_week)

# Anchor each weekly value at the mid-point of its MMWR week within 2022
# (MMWR week 1 of 2022 starts Sun 2022-01-02; midpoint = +3 days)
mmwr_anchor <- function(week, year = 2022) {
  jan1 <- as.Date(paste0(year, "-01-01"))
  wday1 <- as.integer(format(jan1, "%w"))  # 0 = Sun
  # MMWR week 1 = week containing Jan 4; week starts on Sunday
  mmwr_start <- jan1 - wday1
  mmwr_start + 7 * (week - 1) + 3
}

bird_zone_week_all[, anchor_date := mmwr_anchor(mmwr_week, 2022)]

# Linear interpolation within each zone: anchor weekly values at midpoint, then
# interpolate to daily across the full study window. Use approx() with rule = 2
# so days before the first or after the last anchor receive the nearest weekly
# value (clamped endpoints). This produces a smooth daily series that exactly
# passes through the weekly midpoints.
interpolate_bird_daily <- function(anchors_df, target_dates) {
  vars <- c("total_bird_abundance", "anseriformes", "predators")
  out <- data.frame(date = target_dates)
  for (v in vars) {
    if (nrow(anchors_df) >= 2) {
      f <- stats::approxfun(anchors_df$anchor_date, anchors_df[[v]],
                            method = "linear", rule = 2)
      out[[v]] <- f(target_dates)
    } else if (nrow(anchors_df) == 1) {
      out[[v]] <- rep(anchors_df[[v]][1], length(target_dates))
    } else {
      out[[v]] <- NA_real_
    }
  }
  out
}

bird_daily_list <- lapply(split(bird_zone_week_all, bird_zone_week_all$zone_id), function(g) {
  anchors <- as.data.frame(g[order(anchor_date)])
  daily   <- interpolate_bird_daily(anchors, date_seq)
  daily$zone_id <- g$zone_id[1]
  daily
})
bird_daily <- rbindlist(bird_daily_list)
setcolorder(bird_daily, c("zone_id", "date", "total_bird_abundance",
                          "anseriformes", "predators"))
setkey(bird_daily, zone_id, date)

cat("\nBird interpolation:\n")
cat("  Zones with weekly bird data:", uniqueN(bird_zone_week_all, by = "zone_id"), "\n")
cat("  Daily bird rows produced  :", nrow(bird_daily), "\n")

# Merge into daily fishnet
daily_fishnet <- merge(daily_fishnet, bird_daily, by = c("zone_id", "date"), all.x = TRUE)

# Zones with no eBird coverage at all → zero by definition
daily_fishnet[is.na(total_bird_abundance), total_bird_abundance := 0]
daily_fishnet[is.na(anseriformes),         anseriformes         := 0]
daily_fishnet[is.na(predators),            predators            := 0]

# Tag zones with consistently zero abundance (for diagnostic mapping)
zone_zero_summary <- daily_fishnet[, .(total_abundance = sum(total_bird_abundance, na.rm = TRUE)),
                                   by = zone_id]
zone_ids_all_zero <- zone_zero_summary[total_abundance == 0, zone_id]
daily_fishnet[, bird_zero := fifelse(zone_id %in% zone_ids_all_zero, "No Birds", "Has Birds")]

```

## 6. Feedlot covariates (time-invariant)

```{r 6a-feedlots}

feedlots <- st_read(file.path(feedlot_data, "feedlots.shp")) %>%
  st_transform(crs = st_crs(fishnet_geom))

feedlots_clean <- feedlots %>%
  mutate(
    cafo_flag  = ifelse(!is.na(cafo_flag) & cafo_flag != "N", 1L, 0L),
    near_river = ifelse(!is.na(river_stre) & river_stre != "N", 1L, 0L)
  )

feedlots_in_cells <- st_join(feedlots_clean, fishnet_geom, join = st_within) %>%
  st_drop_geometry()

feedlot_summary <- feedlots_in_cells %>%
  group_by(zone_id) %>%
  summarise(
    feedlot_count = n(),
    au_count      = sum(au_count, na.rm = TRUE),
    birds_au      = sum(birds_au, na.rm = TRUE),
    cafo_flag     = max(cafo_flag, na.rm = TRUE),
    near_river    = max(near_river, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  as.data.table()

daily_fishnet <- merge(daily_fishnet, feedlot_summary, by = "zone_id", all.x = TRUE)
daily_fishnet[, `:=`(
  feedlot_count = fcoalesce(feedlot_count, 0L),
  au_count      = fcoalesce(au_count,      0),
  birds_au      = fcoalesce(birds_au,      0),
  cafo_flag     = fcoalesce(cafo_flag,     0L),
  near_river    = fcoalesce(near_river,    0L)
)]

```

## 7. Land cover (time-invariant)

```{r 7a-landcover}

land_cover <- st_read(file.path(spatial_data, "fishnet_landcover.gpkg"))

land_cover_clean <- land_cover %>%
  st_drop_geometry() %>%
  select(zone_id, land_cover_mode, land_cover_label, lat, lon)

daily_fishnet <- merge(daily_fishnet, land_cover_clean, by = "zone_id", all.x = TRUE)

# Re-level so coefficients are interpretable
daily_fishnet[, land_cover_label := relevel(
  factor(land_cover_label), ref = "Developed, Medium Intensity"
)]

daily_fishnet[, farm_possible := fifelse(
  land_cover_label %in% c("Cultivated Crops", "Pasture/Hay", "Emergent Herbaceous Wetlands"),
  1L, 0L
)]

```

## 8. Sort + final NA check

```{r 8a-final-checks}

setkey(daily_fishnet, zone_id, date)

na_summary <- daily_fishnet[, lapply(.SD, function(x) sum(is.na(x)))]
cat("\nNA counts by column (non-zero only):\n")
print(t(na_summary[, .SD, .SDcols = which(na_summary != 0)]))
cat("\n(Expected NAs only for outbreak_id / outbreak_date on non-case days.)\n")

```

## 9. Export daily zone × date panel

```{r 9a-export}

saveRDS(daily_fishnet, file.path(objects_folder, "mn_fishnet_daily_processed.rds"))

# Also write CSV for downstream non-R consumers (Python map script, etc.)
fwrite(daily_fishnet, file.path(processed_data, "mn_fishnet_daily_processed.csv"))

# Spatial export — single feature per zone with time-invariant covariates
fishnet_spatial <- fishnet_geom %>%
  left_join(
    daily_fishnet[, .(any_outbreak = max(any_outbreak),
                      feedlot_count = first(feedlot_count),
                      bird_zero = first(bird_zero),
                      land_cover_label = first(as.character(land_cover_label))),
                  by = zone_id],
    by = "zone_id"
  )
saveRDS(fishnet_spatial, file.path(objects_folder, "mn_fishnet_daily_processed_sf.rds"))
st_write(fishnet_spatial, file.path(processed_data, "mn_fishnet_daily_processed.gpkg"),
         delete_dsn = TRUE)

cat("\nExports written:\n")
cat("  ", file.path(objects_folder, "mn_fishnet_daily_processed.rds"), "\n")
cat("  ", file.path(processed_data, "mn_fishnet_daily_processed.csv"), "\n")
cat("  ", file.path(objects_folder, "mn_fishnet_daily_processed_sf.rds"), "\n")
cat("  ", file.path(processed_data, "mn_fishnet_daily_processed.gpkg"), "\n")

```
