---
title: "Create Daily Timeseries Dataframe + Independence Flagging"
author: "Fintan Mooney"
date: "2026-05-12"
output: html_document
---

# NOTE: This script requires confidential HPAI spillover point-location data
# (not publicly available). The processed grid-cell level output is provided
# in data/ and downstream analyses can be run without this file.

# Builds the DAILY zone × date timeseries panel and flags non-independent spillovers
# using pairwise distance × day-gap thresholds (200m / 500m / 1km × 28 days).
# Climate-anomaly computations have been moved to a standalone Rmd
# (02_code/2b_data_exploration/b_00_climate_anomaly_setup.Rmd) so the primary daily pipeline
# is no longer entangled with 25-year baselines.

# Major revision (May 2026): all temporal logic at day resolution. Day-gap
# threshold is 28 days (matches the prior 4-week threshold). Outputs:
#   - fishnet_timeseries_daily.RDS      — full daily panel + flags
#   - fishnet_timeseries_daily_modeling.RDS — model-ready (non-independent removed)
#   - spillover_summary.RDS             — sample-size cascade for the CONSORT figure

```{r 0a-initiate, include=FALSE}
# 0a. Declare root directory, folder locations, and load packages + helpers
rm(list = ls())
project.folder = paste0(print(here::here()), '/')
source(paste0(project.folder, 'create_folder_structure.R'))
source(paste0(functions.folder, 'script_initiate.R'))
```

## 1. Load daily fishnet

```{r 1a-load-daily-fishnet}

daily_fishnet <- readRDS(file.path(objects_folder, "mn_fishnet_daily_processed.rds"))
setDT(daily_fishnet)
setkey(daily_fishnet, zone_id, date)

cat("Daily fishnet:\n")
cat("  Rows :", nrow(daily_fishnet), "\n")
cat("  Zones:", uniqueN(daily_fishnet$zone_id), "\n")
cat("  Date range:", paste(range(daily_fishnet$date), collapse = " — "), "\n")

```

## 2. Build outbreak summary by day for independence checks

```{r 2a-outbreak-day-summary}

# 2022-only case window — leading 2021 days are control-exposure history, not cases
case_window_start <- as.Date("2022-01-01")
case_window_end   <- as.Date("2022-12-31")

zone_outbreaks <- daily_fishnet[
  outbreak_binary == 1 & date >= case_window_start & date <= case_window_end
] %>%
  .[order(zone_id, date)]

cat("\nOutbreak zone-days (in 2022 case window):", nrow(zone_outbreaks), "\n")
cat("Unique outbreak zones:", uniqueN(zone_outbreaks, by = "zone_id"), "\n")
cat("Total outbreak events :", sum(zone_outbreaks$outbreak_count), "\n")

```

## 3. Pairwise distance × day-gap between spillover events for independence flagging

```{r 3a-pairwise-distances}

# Reload raw spillover points (one row per event, with lat/lon)
spillover_file <- file.path(spillover_data, "hpai-outbreaks-clean-placeholder-032124.csv")
if (!file.exists(spillover_file)) {
  stop("This script needs the confidential spillover data (hpai-outbreaks-clean-placeholder-032124.csv).")
}
hpai_raw <- read_csv(spillover_file)

hpai_mn_2022 <- hpai_raw %>%
  filter(state == "MN",
         dplyr::between(date_confirmed, case_window_start, case_window_end))

# Get the same CRS used in the fishnet
fishnet_raw <- st_read(file.path(spatial_data, "fishnet_10km_timeseries.gpkg"))
target_crs <- st_crs(fishnet_raw)
fishnet_geom_sf <- fishnet_raw %>% distinct(zone_id, geom)

hpai_sf <- hpai_mn_2022 %>%
  st_as_sf(coords = c("lon", "lat"), crs = 4326) %>%
  st_transform(crs = target_crs) %>%
  st_join(fishnet_geom_sf, join = st_within)

# ---- Within-cell pairwise distance × day-gap ----
multi_outbreak_zones <- hpai_sf %>%
  st_drop_geometry() %>%
  count(zone_id) %>%
  filter(n > 1) %>%
  pull(zone_id)

pairwise_results <- list()
for (z in multi_outbreak_zones) {
  zone_pts <- hpai_sf %>% filter(zone_id == z) %>% arrange(date_confirmed)
  n <- nrow(zone_pts)
  if (n < 2) next
  for (i in 1:(n - 1)) {
    for (j in (i + 1):n) {
      dist_m <- as.numeric(st_distance(zone_pts[i, ], zone_pts[j, ]))
      days_apart <- as.numeric(zone_pts$date_confirmed[j] - zone_pts$date_confirmed[i])
      pairwise_results[[length(pairwise_results) + 1]] <- tibble(
        zone_id        = z,
        outbreak_id_1  = zone_pts$outbreak_id[i],
        outbreak_id_2  = zone_pts$outbreak_id[j],
        distance_km    = dist_m / 1000,
        days_apart     = days_apart,
        same_farm      = dist_m < 100
      )
    }
  }
}
pairwise_df <- bind_rows(pairwise_results) %>% arrange(distance_km, days_apart)

cat("\n--- Within-cell pairwise event summary ---\n")
cat("Pairs at <200m & <28d:", sum(pairwise_df$distance_km < 0.2 & pairwise_df$days_apart < 28 & !pairwise_df$same_farm), "\n")
cat("Pairs at <500m & <28d:", sum(pairwise_df$distance_km < 0.5 & pairwise_df$days_apart < 28 & !pairwise_df$same_farm), "\n")
cat("Pairs at <1km  & <28d:", sum(pairwise_df$distance_km < 1.0 & pairwise_df$days_apart < 28 & !pairwise_df$same_farm), "\n")
cat("Same-farm repeat pairs <28d:", sum(pairwise_df$same_farm & pairwise_df$days_apart < 28), "\n")

# ---- Cross-cell pairs within 1 km & 28 days ----
hpai_sf <- hpai_sf %>% arrange(date_confirmed)
n_evt <- nrow(hpai_sf)
cross_cell_results <- list()
for (i in 1:(n_evt - 1)) {
  for (j in (i + 1):n_evt) {
    if (hpai_sf$zone_id[i] == hpai_sf$zone_id[j]) next
    days_apart <- as.numeric(hpai_sf$date_confirmed[j] - hpai_sf$date_confirmed[i])
    if (days_apart > 28) next
    dist_m <- as.numeric(st_distance(hpai_sf[i, ], hpai_sf[j, ]))
    if (dist_m > 1000) next
    cross_cell_results[[length(cross_cell_results) + 1]] <- tibble(
      zone_id_1     = hpai_sf$zone_id[i],
      zone_id_2     = hpai_sf$zone_id[j],
      outbreak_id_1 = hpai_sf$outbreak_id[i],
      outbreak_id_2 = hpai_sf$outbreak_id[j],
      distance_km   = dist_m / 1000,
      days_apart    = days_apart
    )
  }
}
cross_cell_df <- if (length(cross_cell_results) > 0) bind_rows(cross_cell_results) else
  tibble(zone_id_1=integer(), zone_id_2=integer(), outbreak_id_1=integer(),
         outbreak_id_2=integer(), distance_km=numeric(), days_apart=numeric())

cat("\nCross-cell pairs within 1km & 28d:", nrow(cross_cell_df), "\n")

```

## 4. Flag the second outbreak in each non-independent pair (within-cell + cross-cell)

```{r 4a-flag-non-independent}

# IDs of the *later* outbreak in each non-independent pair, by distance threshold,
# combining same-farm repeats with proximity-based pairs.
build_flag_ids <- function(within_df, cross_df, dist_km, day_gap = 28) {
  within_ids <- within_df %>%
    filter(days_apart < day_gap, (distance_km < dist_km | same_farm)) %>%
    pull(outbreak_id_2) %>% unique()
  cross_ids <- cross_df %>%
    filter(distance_km < dist_km) %>%
    pull(outbreak_id_2) %>% unique()
  unique(c(within_ids, cross_ids))
}

ids_200m <- build_flag_ids(pairwise_df, cross_cell_df, 0.2)
ids_500m <- build_flag_ids(pairwise_df, cross_cell_df, 0.5)
ids_1km  <- build_flag_ids(pairwise_df, cross_cell_df, 1.0)

cat("\nFlagged event IDs (later event in each non-independent pair):\n")
cat("  <200m & <28d:", length(ids_200m), "\n")
cat("  <500m & <28d:", length(ids_500m), "\n")
cat("  <1km  & <28d:", length(ids_1km),  "\n")

# Annotate the daily panel — a zone-day flag is TRUE if any outbreak_id in that
# zone-day appears in the flag set. (Multi-event zone-days keep their stratum
# unless every event is flagged.)
contains_flagged <- function(id_str, flagged_ids) {
  if (is.na(id_str)) return(FALSE)
  ids <- as.numeric(trimws(unlist(strsplit(id_str, ","))))
  any(ids %in% flagged_ids)
}

# Per-row count of flagged outbreak_ids — used both for the season cascade and
# for outbreak_count adjustment in the modeling panel below.
count_flagged_ids <- function(id_str, flagged_ids) {
  if (is.na(id_str)) return(0L)
  ids <- as.numeric(trimws(unlist(strsplit(id_str, ","))))
  as.integer(sum(ids %in% flagged_ids))
}

daily_fishnet[, `:=`(
  flag_nearby_200m = vapply(outbreak_id, contains_flagged, logical(1), flagged_ids = ids_200m),
  flag_nearby_500m = vapply(outbreak_id, contains_flagged, logical(1), flagged_ids = ids_500m),
  flag_nearby_1km  = vapply(outbreak_id, contains_flagged, logical(1), flagged_ids = ids_1km)
)]

cat("\nflag_nearby_500m (primary):", sum(daily_fishnet$flag_nearby_500m), "zone-days\n")

```

## 5. Time covariates for downstream models

```{r 5a-time-covariates}

daily_fishnet[, `:=`(
  month         = lubridate::month(date),
  month_factor  = factor(lubridate::month(date)),
  warm_season   = as.integer(lubridate::month(date) %in% 5:9),
  spring        = as.integer(lubridate::month(date) %in% 3:5),
  spring_fall   = as.integer(lubridate::month(date) %in% c(3:5, 9:11)),
  season        = factor(
    fcase(
      lubridate::month(date) %in% c(12, 1, 2), "Winter",
      lubridate::month(date) %in% 3:5,         "Spring",
      lubridate::month(date) %in% 6:8,         "Summer",
      lubridate::month(date) %in% 9:11,        "Fall"
    ),
    levels = c("Winter", "Spring", "Summer", "Fall")
  )
)]

```

## 6. Export the daily timeseries (with flags) + the sample-size cascade

```{r 6a-export}

# Counts per season for the CONSORT figure. Independent events = raw events
# minus the number of flagged outbreak_ids on that zone-day (per-id, not
# all-or-nothing — a zone-day can hold >1 flagged event).
season_independent <- daily_fishnet[
  outbreak_binary == 1 & date >= case_window_start & date <= case_window_end
][, n_flagged_500m := vapply(outbreak_id, count_flagged_ids, integer(1),
                             flagged_ids = ids_500m)
][, .(total_events       = sum(outbreak_count),
      independent_events = sum(outbreak_count - n_flagged_500m)),
  by = season
][order(match(season, c("Winter", "Spring", "Summer", "Fall")))]

# Multi-event zone-day breakdown for the CONSORT figure
multi_event_zone_days <- daily_fishnet[
  outbreak_binary == 1 & outbreak_count > 1 &
    date >= case_window_start & date <= case_window_end,
  .N
]

spillover_summary <- list(
  total                  = sum(daily_fishnet[date >= case_window_start &
                                             date <= case_window_end]$outbreak_count),
  case_zone_days         = sum(daily_fishnet[date >= case_window_start &
                                             date <= case_window_end]$outbreak_binary),
  case_zones             = uniqueN(daily_fishnet[outbreak_binary == 1 &
                                                 date >= case_window_start &
                                                 date <= case_window_end]$zone_id),
  multi_event_zone_days  = multi_event_zone_days,
  non_independent_200m   = length(ids_200m),
  non_independent_500m   = length(ids_500m),
  non_independent_1km    = length(ids_1km),
  by_season              = season_independent
)

saveRDS(spillover_summary, file.path(objects_folder, "spillover_summary.RDS"))

cat("\n--- Spillover summary ---\n")
print(spillover_summary)

# Full daily timeseries with flags
saveRDS(daily_fishnet, file.path(objects_folder, "fishnet_timeseries_daily.RDS"))

# Modeling version: drop the later event of each non-independent pair (500m, 28d).
# Use `ids_500m` so the modeling panel's flag set is identical to the one used
# in the seasonal CONSORT cascade above (combines within-cell AND cross-cell
# pairs at the same threshold).
flagged_zone_days <- daily_fishnet[
  outbreak_binary == 1 & date >= case_window_start & date <= case_window_end
] %>%
  .[, n_flagged := vapply(outbreak_id, count_flagged_ids, integer(1),
                          flagged_ids = ids_500m)] %>%
  .[n_flagged > 0, .(zone_id, date, n_flagged)]

modeling <- copy(daily_fishnet)
modeling <- merge(modeling, flagged_zone_days, by = c("zone_id", "date"), all.x = TRUE)
modeling[, n_flagged := fcoalesce(n_flagged, 0L)]
modeling[, outbreak_count  := outbreak_count - n_flagged]
modeling[, outbreak_binary := as.integer(outbreak_count > 0)]
modeling[, n_flagged := NULL]

cat("\nModeling panel after removing non-independent later events:\n")
cat("  Total events:", sum(modeling$outbreak_count), "\n")
cat("  Case zone-days:", sum(modeling$outbreak_binary), "\n")

saveRDS(modeling, file.path(objects_folder, "fishnet_timeseries_daily_modeling.RDS"))

```

## 7. Pairwise diagnostic export (used by CONSORT figure)

```{r 7a-pairwise-export}

saveRDS(list(
  within_cell = pairwise_df,
  cross_cell  = cross_cell_df,
  ids_200m    = ids_200m,
  ids_500m    = ids_500m,
  ids_1km     = ids_1km
), file.path(objects_folder, "independence_screening_daily.RDS"))

```
