---
title: "Per-zone weekly bird abundance (multi-state, 7-state flyway)"
author: "Fintan Mooney"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

```{r initiate}
rm(list = ls())
# 0a.Declare root directory, folder location and load essential stuff
project.folder = paste0(print(here::here()), '/')
source(paste0(project.folder, 'create_folder_structure.R'))
source(paste0(functions.folder, 'script_initiate.R'))

# 0b.lubridate owns the date accessors under the project's conflicted setup.
conflict_prefer("year",  "lubridate", quiet = TRUE)
conflict_prefer("month", "lubridate", quiet = TRUE)

# 0c.Study region: the 7 flyway states with eBird abundance extracted. SD is
#    excluded throughout the multi-state pipeline (no bird data). These codes index
#    the spillover/ERA5 inputs and the fishnet `state` field alike.
study_states_abbr <- c("IA", "IL", "IN", "MI", "MN", "OH", "WI")
```

This is the multi-state analogue of the Minnesota `a_04_feedlot_bird_abundance.Rmd`.
The MN file extracts eBird relative-abundance rasters at feedlot points to build
per-zone bird abundance for control cells. For the multi-state pipeline we instead reuse
the **already-extracted** premises-level panel
(`study_spillover_events_casecrossover_2022_abundance_060326.csv`, the same input
`c_03`/`c_04` use): each outbreak premises carries its full weekly eBird series. Here
we spatially place each premises in its fishnet zone and collapse to a
**per-zone × ISO-week** bird-abundance table that the multi-state `a_04` consumes.

The study region is the **7 Mississippi/Central-flyway states** with eBird abundance
extracted: IA, IL, IN, MI, MN, OH, WI. (South Dakota has been removed from the
multi-state pipeline entirely — it had no eBird extraction.) Because the time-stratified
case-crossover compares a case zone
against control *days/weeks in the same zone*, bird abundance is only ever needed for
zones that hold an outbreak — which is exactly what a premises→zone join produces.

# 1. Load the premises-level weekly bird panel

```{r load-bird}
bird_raw <- fread(paste0(bird_abundance_data,
                         "study_spillover_events_casecrossover_2022_abundance_060326.csv"))

# 1a.Restrict to the 7 study states (drops SD if it ever appears) and keep ALL link
#    types. The independence split — independent (IND+INR) vs. all events (incl.
#    cross-farm CSLT and unsequenced NOSEQ) — is applied downstream in a_08, so bird
#    abundance must be built for *every* event zone, not just the independent ones.
keep_ids <- bird_raw[is_case_week == 1 & state %in% study_states_abbr, unique(outbreak_id)]
bird     <- bird_raw[outbreak_id %in% keep_ids]
bird[, week_start := as.Date(week_start)]

cat("Bird panel: ", length(keep_ids), " outbreaks (all links) across ",
    uniqueN(bird$state), " states; ", nrow(bird), " premises-week rows\n", sep = "")
print(bird[is_case_week == 1, .N, by = .(state, link)][order(state, link)])
```

# 2. Map each premises to its 7-state fishnet zone

```{r premises-zone}
# 2a.Multi-state fishnet (7 states; SD already removed). GEE writes EPSG:5070
#    coordinates but mislabels the CRS as WGS84, so we *set* (not transform) it.
fishnet <- st_read(paste0(env_data, "hpai_multistate_era5/fishnet_7state.geojson"),
                   quiet = TRUE)
fishnet <- fishnet[!st_is_empty(fishnet), ]
fishnet <- st_set_crs(fishnet, 5070)

# 2b.One point per premises → containing zone (st_within). Locations are identical
#    to the placeholder spillover CSV (verified), so the zone a premises lands in
#    here is the same case zone a_04 builds from date_confirmed.
prem <- unique(bird[, .(outbreak_id, state, lon, lat)])
prem_sf <- st_transform(st_as_sf(prem, coords = c("lon", "lat"), crs = 4326), 5070)
prem[, zone_id := fishnet$zone_id[as.integer(st_within(prem_sf, fishnet))]]

n_unmapped <- prem[is.na(zone_id), .N]
if (n_unmapped > 0) {
  # Premises on a cell boundary or just outside the fishnet — snap to nearest cell.
  miss <- which(is.na(prem$zone_id))
  nn   <- st_nearest_feature(prem_sf[miss, ], fishnet)
  prem[miss, zone_id := fishnet$zone_id[nn]]
  message(n_unmapped, " premises fell outside a cell; snapped via nearest-feature.")
}

cat("Premises mapped to a zone, by state:\n")
print(prem[, .(n_premises = .N, mapped = sum(!is.na(zone_id))), by = state][order(-n_premises)])
```

# 3. Collapse premises-weeks to per-zone × ISO-week abundance

```{r zone-week}
# 3a.Attach the zone to every premises-week row.
bird <- merge(bird, prem[, .(outbreak_id, zone_id)], by = "outbreak_id", all.x = TRUE)

# 3b.Derive the three aggregates the MN pipeline uses. anseri_abd/total_abd come
#    pre-summed in the panel; predators = the non-anseriform remainder (the raptor/
#    pelican detections: rethaw, grhowl, baleag, amwpel). Defining it as
#    total - anseri sidesteps the buwtea grouping ambiguity in the MN code.
bird[, predators := total_abd - anseri_abd]

# 3c.Average across any premises sharing a zone (10 km cells rarely hold >1), per
#    ISO-week. Keep week_start (Monday anchor) for the weekly→daily interpolation
#    a_04 does, and the case-window flags for diagnostics.
zone_week_bird <- bird[
  , .(total_bird_abundance = mean(total_abd,  na.rm = TRUE),
      anseriformes         = mean(anseri_abd, na.rm = TRUE),
      predators            = mean(predators,  na.rm = TRUE),
      n_premises           = uniqueN(outbreak_id)),
  by = .(zone_id, state, iso_year, iso_week, week_start)
]

# mean() over an all-NA week returns NaN — normalise to NA for clean downstream joins.
for (v in c("total_bird_abundance", "anseriformes", "predators")) {
  set(zone_week_bird, which(is.nan(zone_week_bird[[v]])), v, NA_real_)
}
setorder(zone_week_bird, zone_id, week_start)

cat("\nPer-zone weekly bird table: ", nrow(zone_week_bird), " zone-week rows across ",
    uniqueN(zone_week_bird$zone_id), " zones\n", sep = "")
```

# 4. Export

```{r export}
# Consumed by a_04_join_data.Rmd (joined by zone_id × ISO-week, then
# interpolated weekly→daily for the daily panel; used natively for the weekly panel).
fwrite(zone_week_bird,
       paste0(bird_abundance_data, "multistate-zone-weekly-bird-abundance-061626.csv"))
cat("Wrote multistate-zone-weekly-bird-abundance-061626.csv\n")
```
