---
title: "Create Time-Stratified Case-Crossover Dataframe (Daily)"
author: "Fintan Mooney"
date: "2026-05-12"
output: html_document
---

# NOTE: This script can be re-run from the publicly-available daily timeseries
# (fishnet_timeseries_daily.RDS) — it does not require the confidential raw
# spillover file directly.

# Builds the DAILY time-stratified case-crossover dataset for the HPAI
# outbreak analysis (May 2026 GeoHealth revision).
#
# Design (Lu et al. 2020, PLoS Medicine, doi:10.1371/journal.pmed.1003176;
# Gasparrini 2022, BMC Med Res Methodol 22:129):
#   - Stratum = zone_id × ISO year × calendar month × day-of-week
#   - Cases are zone-days with at least one independent HPAI detection
#     (after 500m × 28-day spatial-temporal independence screen)
#   - Controls are all other days in the same stratum that are NOT case-days
#   - Each case typically has 3–4 controls (other matching weekdays in the same month)
#
# Lagged exposures (lag 0–28 days) for DLNM crossbasis terms are constructed
# from the underlying daily zone timeseries.

```{r 0a-initiate, include=FALSE}
# 0a. Declare root directory, folder locations, and load packages + helpers
rm(list = ls())
project.folder = paste0(print(here::here()), '/')
source(paste0(project.folder, 'create_folder_structure.R'))
source(paste0(functions.folder, 'script_initiate.R'))
```

## 1. Load daily timeseries (modeling version — non-independent later events removed)

```{r 1a-load}

daily <- readRDS(file.path(objects_folder, "fishnet_timeseries_daily_modeling.RDS"))
setDT(daily)
setkey(daily, zone_id, date)

# Independence-screening details (for sensitivity variants at 1km / 200m)
screening <- readRDS(file.path(objects_folder, "independence_screening_daily.RDS"))

case_window_start <- as.Date("2022-01-01")
case_window_end   <- as.Date("2022-12-31")

```

## 2. CONSORT: starting counts before stratification

```{r 2a-consort-starting}

# Restrict to the 2022 case window for stratum construction; leading 2021 days
# stay in the panel only as lag exposure history.
case_year <- daily[date >= case_window_start & date <= case_window_end]

starting_events       <- sum(case_year$outbreak_count, na.rm = TRUE)
starting_case_zonedays<- sum(case_year$outbreak_binary, na.rm = TRUE)
starting_case_zones   <- uniqueN(case_year[outbreak_binary == 1]$zone_id)

cat("Starting (after independence screening):\n")
cat("  Independent events     :", starting_events, "\n")
cat("  Case zone-days         :", starting_case_zonedays, "\n")
cat("  Unique case zones      :", starting_case_zones, "\n")

```

## 3. Time-stratified case-crossover construction

```{r 3a-construct}

# Stratum key on every zone-day — calendar year × calendar month × DOW × zone
# (Lu et al. style; ensures controls are non-case days in the same zone,
#  the same calendar month, and share the case's weekday.)
case_year[, `:=`(
  cal_year  = lubridate::year(date),
  cal_month = lubridate::month(date),
  dow       = lubridate::wday(date, label = FALSE, week_start = 1)
)]
case_year[, stratum_id := sprintf("%d-%02d-%d-z%d", cal_year, cal_month, dow, zone_id)]

# Identify case days
case_days <- case_year[outbreak_binary == 1, .(zone_id, case_date = date, stratum_id)]

# Restrict to strata that contain at least one case day
active_strata <- unique(case_days$stratum_id)
cc_pool <- case_year[stratum_id %in% active_strata]

# Within each active stratum, mark case vs control
cc_pool[, case_control := fifelse(outbreak_binary == 1, "case", "control")]
cc_pool[, outbreak_binary := as.integer(outbreak_binary == 1)]

# Drop strata that have no controls (single-day month-DOW slot) — these
# contribute no information to the conditional likelihood.
controls_per_stratum <- cc_pool[case_control == "control", .N, by = stratum_id]
strata_with_controls <- controls_per_stratum[N >= 1]$stratum_id
strata_dropped_no_ctrl <- length(active_strata) - length(strata_with_controls)
cc_pool <- cc_pool[stratum_id %in% strata_with_controls]

cat("\nStratum construction:\n")
cat("  Active strata (≥1 case)        :", length(active_strata), "\n")
cat("  Strata dropped (no controls)   :", strata_dropped_no_ctrl, "\n")
cat("  Strata kept                    :", length(strata_with_controls), "\n")
cat("  Cases  (final)                 :", sum(cc_pool$outbreak_binary == 1), "\n")
cat("  Controls (final)               :", sum(cc_pool$outbreak_binary == 0), "\n")

cat("\nControls per case-stratum:\n")
print(cc_pool[case_control == "control", .N, by = stratum_id][, .N, by = N][order(N)])

```

## 4. Build lag 0–28 exposure history for each row (DLNM-ready)

```{r 4a-build-lags}

# Variables to lag — meteorological + bird
lag_vars <- c(
  "temperature", "dewpoint", "relative_humidity", "wind_speed",
  "soil_moisture", "snow_density", "precipitation", "runoff",
  "evapotranspiration", "leaf_area_index", "shortwave_radiation",
  "longwave_radiation", "surface_pressure",
  "total_bird_abundance", "anseriformes", "predators"
)

# Convert precipitation and runoff from m → mm (cosmetic; affects coefficient
# units only). NB: this mutates `daily` AND `cc_pool` in place via data.table
# reference semantics. Re-running this chunk without first re-loading `daily`
# would multiply by 1000 a second time. The stopifnot() guard below catches that.
#
# ERA5-Land daily precipitation in meters caps at ~0.2 m even on extreme days
# (200 mm/day); runoff is similarly bounded. If max > 1 we're already in mm.
stopifnot(
  "runoff already in mm — re-load `daily` (and re-build `cc_pool`) before re-running this chunk" =
    max(daily$runoff,        na.rm = TRUE) < 1,
  "precipitation already in mm — re-load `daily` (and re-build `cc_pool`) before re-running this chunk" =
    max(daily$precipitation, na.rm = TRUE) < 1
)

cc_pool[, runoff        := runoff        * 1000]
cc_pool[, precipitation := precipitation * 1000]
daily[,   runoff        := runoff        * 1000]
daily[,   precipitation := precipitation * 1000]

# For each variable and lag k, attach the value from (zone_id, date − k days).
# We pull from `daily` (the full 2021-Nov–2022-Dec panel) so early-year cases
# don't run off the start of 2022.
#
# Implementation: shift each variable in-zone using data.table::shift() across
# the densely-keyed source panel, then merge ALL lag columns in a single join.
# This replaces 16 × 29 = 464 per-(var, lag) merges with one merge — identical
# result because `daily` is a complete zone × date crossing with no gaps.
build_lag_matrix <- function(target_df, source_df, vars, max_lag = 28) {
  src <- copy(source_df)
  setkey(src, zone_id, date)
  lag_cols <- character(0)
  for (v in vars) {
    new_names <- paste0(v, "_Lag", 0:max_lag)
    src[, (new_names) := shift(.SD[[1]], n = 0:max_lag, type = "lag"),
        by = zone_id, .SDcols = v]
    lag_cols <- c(lag_cols, new_names)
  }
  out <- merge(
    target_df[, .(zone_id, date, stratum_id, outbreak_binary, case_control)],
    src[, c("zone_id", "date", ..lag_cols)],
    by = c("zone_id", "date"), all.x = TRUE
  )
  out
}

cat("Building lag 0–28 exposure matrix…\n")
lag_mat <- build_lag_matrix(cc_pool, daily, lag_vars, max_lag = 28)

# Re-attach all the non-lag columns (covariates, indicators)
non_lag_cols <- setdiff(names(cc_pool), names(lag_mat))
case_crossover_df <- merge(lag_mat, cc_pool[, c("zone_id", "date", ..non_lag_cols)],
                           by = c("zone_id", "date"))

cat("Case-crossover df: ", nrow(case_crossover_df), "rows × ", ncol(case_crossover_df), "cols\n")

# Check for any missing lag values (would indicate the 2021 leading window is insufficient)
lag0_cols <- grep("_Lag0$", names(case_crossover_df), value = TRUE)
na_summary <- case_crossover_df[, lapply(.SD, function(x) sum(is.na(x))), .SDcols = lag0_cols]
n_missing_lag0 <- sum(unlist(na_summary))
cat("Missing Lag0 values across vars:", n_missing_lag0, "\n")

deepest_lag_cols <- grep("_Lag28$", names(case_crossover_df), value = TRUE)
deep_na <- case_crossover_df[, lapply(.SD, function(x) sum(is.na(x))), .SDcols = deepest_lag_cols]
n_missing_lag28 <- sum(unlist(deep_na))
cat("Missing Lag28 values across vars:", n_missing_lag28, "\n")
if (n_missing_lag28 > 0) {
  warning("Some Lag28 exposures are NA — extend the GEE daily extraction back further than Nov 15 2021.")
}

```

## 5. Transform skewed variables (log1p) and z-scale

```{r 5a-transform}

skewed_vars <- c("total_bird_abundance", "runoff", "precipitation",
                 "anseriformes", "predators")
for (v in skewed_vars) {
  for (k in 0:28) {
    col <- paste0(v, "_Lag", k)
    if (col %in% names(case_crossover_df)) {
      case_crossover_df[[col]] <- log1p(case_crossover_df[[col]])
    }
  }
}

# Scale each lag column (z-scale across the full case-crossover pool so the
# 0.5-SD OR interpretation in the manuscript remains valid).
for (v in lag_vars) {
  for (k in 0:28) {
    col <- paste0(v, "_Lag", k)
    if (col %in% names(case_crossover_df)) {
      case_crossover_df[[col]] <- as.numeric(scale(case_crossover_df[[col]]))
    }
  }
}

```

## 6. CONSORT sample-size cascade (final numbers)

```{r 6a-consort-export}

cascade <- list(
  date_revised             = Sys.Date(),
  design                   = "Time-stratified case-crossover (zone × year × month × DOW), DLNM lag 0–28 d",
  # NB: `daily` here is the modeling panel (post-independence screening), so
  # `outbreak_count` is the independent event count. The raw event total lives
  # in spillover_summary$total (b_create_timeseries_df.Rmd).
  events_independent       = starting_events,
  case_zonedays_input      = starting_case_zonedays,
  case_zones_input         = starting_case_zones,
  strata_active            = length(active_strata),
  strata_dropped_no_ctrl   = strata_dropped_no_ctrl,
  strata_final             = length(strata_with_controls),
  cases_final              = sum(case_crossover_df$outbreak_binary == 1),
  controls_final           = sum(case_crossover_df$outbreak_binary == 0)
)
saveRDS(cascade, file.path(objects_folder, "case_crossover_cascade.RDS"))

cat("\n--- CONSORT cascade ---\n")
print(cascade)

```

## 7. Export primary case-crossover dataframe (500 m / 28 d screen applied upstream)

```{r 7a-export-primary}

saveRDS(case_crossover_df, file.path(objects_folder, "case_crossover_df.RDS"))
fwrite(case_crossover_df,
       file.path(processed_data, "case_crossover_df_daily_ind_500m.csv"))

```

## 8. Sensitivity variants: 1 km and 200 m spatial-independence thresholds

```{r 8a-sensitivity-variants}

count_flagged_ids <- function(id_str, flagged_ids) {
  if (is.na(id_str)) return(0L)
  ids <- as.numeric(trimws(unlist(strsplit(id_str, ","))))
  as.integer(sum(ids %in% flagged_ids))
}

build_sensitivity <- function(threshold_label, flag_ids) {
  # Re-derive the modeling outbreak counts using a different flag set
  flagged_zone_days <- daily[
    outbreak_binary == 1 & date >= case_window_start & date <= case_window_end
  ][, n_flagged := vapply(outbreak_id, count_flagged_ids, integer(1),
                          flagged_ids = flag_ids)
  ][n_flagged > 0, .(zone_id, date, n_flagged)]

  variant <- copy(daily)
  variant <- merge(variant, flagged_zone_days, by = c("zone_id", "date"), all.x = TRUE)
  variant[, n_flagged := fcoalesce(n_flagged, 0L)]
  variant[, outbreak_count  := outbreak_count - n_flagged]
  variant[, outbreak_binary := as.integer(outbreak_count > 0)]
  variant[, n_flagged := NULL]

  # Stratum construction
  v_window <- variant[date >= case_window_start & date <= case_window_end]
  v_window[, cal_year  := lubridate::year(date)]
  v_window[, cal_month := lubridate::month(date)]
  v_window[, dow       := lubridate::wday(date, week_start = 1)]
  v_window[, stratum_id := sprintf("%d-%02d-%d-z%d", cal_year, cal_month, dow, zone_id)]

  active <- unique(v_window[outbreak_binary == 1]$stratum_id)
  v_pool <- v_window[stratum_id %in% active]
  v_pool[, case_control := fifelse(outbreak_binary == 1, "case", "control")]

  ctrl_per <- v_pool[case_control == "control", .N, by = stratum_id]
  keep_strata <- ctrl_per[N >= 1]$stratum_id
  v_pool <- v_pool[stratum_id %in% keep_strata]

  lag_v <- build_lag_matrix(v_pool, variant, lag_vars, max_lag = 28)
  non_lag <- setdiff(names(v_pool), names(lag_v))
  out_df <- merge(lag_v, v_pool[, c("zone_id", "date", ..non_lag)],
                  by = c("zone_id", "date"))

  for (v in skewed_vars) for (k in 0:28) {
    col <- paste0(v, "_Lag", k)
    if (col %in% names(out_df)) out_df[[col]] <- log1p(out_df[[col]])
  }
  for (v in lag_vars) for (k in 0:28) {
    col <- paste0(v, "_Lag", k)
    if (col %in% names(out_df)) out_df[[col]] <- as.numeric(scale(out_df[[col]]))
  }

  cat(sprintf("\n[%s] cases: %d  controls: %d  strata: %d\n",
              threshold_label,
              sum(out_df$outbreak_binary == 1),
              sum(out_df$outbreak_binary == 0),
              length(keep_strata)))
  out_df
}

cc_1km  <- build_sensitivity("1 km", screening$ids_1km)
cc_200m <- build_sensitivity("200 m", screening$ids_200m)

saveRDS(cc_1km,  file.path(objects_folder, "case_crossover_df_1km.RDS"))
saveRDS(cc_200m, file.path(objects_folder, "case_crossover_df_200m.RDS"))
fwrite(cc_1km,  file.path(processed_data, "case_crossover_df_daily_ind_1km.csv"))
fwrite(cc_200m, file.path(processed_data, "case_crossover_df_daily_ind_200m.csv"))

```

## 9. Sensitivity variants: lag-window length (14, 42, 56 days)

```{r 9a-lag-sensitivity}

# Same case-crossover stratum structure, but with different lag matrix depth.
# Used by 02_code/2c_models/c_01_lag_sensitivity.Rmd.

build_lag_variant <- function(max_lag) {
  lv <- build_lag_matrix(cc_pool, daily, lag_vars, max_lag = max_lag)
  non_lag <- setdiff(names(cc_pool), names(lv))
  out <- merge(lv, cc_pool[, c("zone_id", "date", ..non_lag)], by = c("zone_id", "date"))
  for (v in skewed_vars) for (k in 0:max_lag) {
    col <- paste0(v, "_Lag", k); if (col %in% names(out)) out[[col]] <- log1p(out[[col]])
  }
  for (v in lag_vars) for (k in 0:max_lag) {
    col <- paste0(v, "_Lag", k); if (col %in% names(out)) out[[col]] <- as.numeric(scale(out[[col]]))
  }
  out
}

saveRDS(build_lag_variant(14), file.path(objects_folder, "case_crossover_df_14lag.RDS"))
saveRDS(build_lag_variant(42), file.path(objects_folder, "case_crossover_df_42lag.RDS"))
saveRDS(build_lag_variant(56), file.path(objects_folder, "case_crossover_df_56lag.RDS"))

```
