---
title: "Case-crossover dataframes — multi-state (unidirectional; daily + weekly)"
author: "Fintan Mooney"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

```{r initiate}
rm(list = ls())
# 0a.Declare root directory, folder location and load essential stuff
project.folder = paste0(print(here::here()), '/')
source(paste0(project.folder, 'create_folder_structure.R'))
source(paste0(functions.folder, 'script_initiate.R'))

# 0b.lubridate owns the date accessors under the project's conflicted setup.
conflict_prefer("year",  "lubridate", quiet = TRUE)
conflict_prefer("month", "lubridate", quiet = TRUE)
```

Builds the multi-state case-crossover dataframes with a **unidirectional** referent design:
each case is matched to up to four control periods drawn from the **four weeks immediately
before** the case, in the same 10 km cell. This is required because after an outbreak we
cannot assume the cell became virus-free, so post-event periods are not valid controls
(mirrors the manuscript and the flyway models `c_03`/`c_04`).

  * **daily**  — case day + control days at 7/14/21/28 d before (same weekday); DLNM lag 0–28 d.
  * **weekly** — case week + the 4 preceding weeks; DLNM lag 0–4 wk.
  * A **bidirectional** (time-stratified) daily df is written as a *secondary* comparison only.

Two case definitions per design: **all events** (`*_all`) and **independent** (no suffix,
primary — cross-farm CSLT/NOSEQ and spatio-temporal non-independent second events removed;
1 km / 200 m vary that spatial threshold). Control periods that themselves had an outbreak
in the same cell are excluded.

# 1. Load panels and the exposure variable set

```{r load}
daily  <- readRDS(paste0(objects_folder, "multistate_fishnet_timeseries_daily.RDS"))
weekly <- readRDS(paste0(objects_folder, "multistate_fishnet_timeseries_weekly.RDS"))
screening <- readRDS(paste0(objects_folder, "multistate_independence_screening.RDS"))
setDT(daily); setDT(weekly)
setkey(daily, zone_id, date)

case_window_start <- as.Date("2022-01-01")
case_window_end   <- as.Date("2022-12-31")

# Count how many of a zone-period's outbreak_ids are in a removal set.
count_flagged_ids <- function(id_str, flagged_ids) {
  if (is.na(id_str)) return(0L)
  as.integer(sum(as.numeric(trimws(unlist(strsplit(id_str, ",")))) %in% flagged_ids))
}

# Case definitions as outbreak_id removal sets: ALL_REMOVE keeps every event; ind_remove()
# drops cross-farm/unsequenced (link CSLT+NOSEQ) plus spatio-temporal non-independent events.
cross_farm_ids <- screening$cross_farm_ids
ALL_REMOVE     <- integer(0)
ind_remove     <- function(spatio_ids) union(cross_farm_ids, spatio_ids)

lag_vars <- intersect(c(
  "temperature", "dewpoint", "relative_humidity", "wind_speed",
  "skin_temperature", "soil_temperature", "soil_moisture", "soil_moisture_2",
  "snow_depth", "snow_cover", "snow_density",
  "precipitation", "snowfall", "runoff", "evapotranspiration",
  "leaf_area_index", "shortwave_radiation", "longwave_radiation", "surface_pressure",
  "total_bird_abundance", "anseriformes", "predators"), names(daily))
skewed_vars <- intersect(c("total_bird_abundance", "anseriformes", "predators",
                           "runoff", "precipitation"), lag_vars)

# log1p the skewed exposures, then z-scale every lag column (effects per +1 SD).
transform_lag_cols <- function(dt, max_lag) {
  for (v in skewed_vars) for (k in 0:max_lag) {
    col <- paste0(v, "_Lag", k); if (col %in% names(dt)) dt[[col]] <- log1p(dt[[col]])
  }
  for (v in lag_vars) for (k in 0:max_lag) {
    col <- paste0(v, "_Lag", k); if (col %in% names(dt)) dt[[col]] <- as.numeric(scale(dt[[col]]))
  }
  dt
}
```

# 2. Convert precipitation & runoff m → mm (guarded)

```{r units}
stopifnot(
  "runoff already in mm — reload panels before re-running" =
    max(daily$runoff,        na.rm = TRUE) < 1,
  "precipitation already in mm — reload panels before re-running" =
    max(daily$precipitation, na.rm = TRUE) < 1
)
daily[,  `:=`(runoff = runoff * 1000, precipitation = precipitation * 1000)]
weekly[, `:=`(runoff = runoff * 1000, precipitation = precipitation * 1000)]
```

# 3. Unidirectional builder (primary design, daily & weekly)

```{r build-uni}
# For each case (flag-removed outbreak in the case window), the matched set is the case
# period plus the periods 7/14/21/28 days before it in the same cell (up to 4 controls).
# Control periods that were themselves outbreaks (any event, same cell) are excluded, and
# strata are kept only if they retain the case and >=1 control. The DLNM lag matrix is built
# on the case zones' full series (so each period carries its own lag-0..max exposure history).
OFFSETS <- c(0, 7, 14, 21, 28)   # case (0) + 4 control periods, in days before

build_uni_cc <- function(panel, tcol, max_lag, flag_ids) {
  p <- copy(panel)
  p[, tkey := as.Date(get(tcol))]
  p[, is_outbreak_raw := as.integer(outbreak_binary == 1)]

  # 3a. Apply the case definition (remove flagged events → redefine cases).
  fl <- p[outbreak_binary == 1 & tkey >= case_window_start & tkey <= case_window_end
        ][, nf := vapply(outbreak_id, count_flagged_ids, integer(1), flagged_ids = flag_ids)
        ][nf > 0, .(zone_id, tkey, nf)]
  p <- merge(p, fl, by = c("zone_id", "tkey"), all.x = TRUE)
  p[, nf := fcoalesce(nf, 0L)]
  p[, outbreak_binary := as.integer(outbreak_count - nf > 0)]
  p[, nf := NULL]

  cases <- p[outbreak_binary == 1 & tkey >= case_window_start & tkey <= case_window_end,
             .(zone_id, ct = tkey)]
  cases[, stratum_id := paste0("z", zone_id, "_", format(ct, "%Y%m%d"))]

  # 3b. Lag matrix on case zones' full series.
  src <- p[zone_id %in% unique(cases$zone_id)]
  setorder(src, zone_id, tkey)
  for (v in lag_vars) {
    nm <- paste0(v, "_Lag", 0:max_lag)
    src[, (nm) := shift(.SD[[1]], n = 0:max_lag, type = "lag"), by = zone_id, .SDcols = v]
  }

  # 3c. Expand each case into its matched set, join exposures, drop outbreak controls.
  cases[, cid := .I]
  sets <- cases[, .(tkey = ct - OFFSETS, is_case = c(1L, 0L, 0L, 0L, 0L)),
                by = .(cid, zone_id, stratum_id)]
  matched <- merge(sets, src, by = c("zone_id", "tkey"), all.x = FALSE)
  matched <- matched[is_case == 1L | is_outbreak_raw == 0L]
  keep <- matched[, .(nc = sum(is_case == 1L), nk = sum(is_case == 0L)),
                  by = stratum_id][nc >= 1 & nk >= 1, stratum_id]
  matched <- matched[stratum_id %in% keep]

  matched[, `:=`(outbreak_binary = is_case,
                 case_control    = fifelse(is_case == 1L, "case", "control"))]
  setnames(matched, "tkey", tcol)
  matched <- transform_lag_cols(matched, max_lag)
  list(df = matched, n_keep = length(keep), n_cases = sum(matched$outbreak_binary == 1))
}
```

# 4. Bidirectional builder (secondary daily comparison only)

```{r build-bi}
# Time-stratified strata (zone × year × calendar-month × day-of-week): controls are the
# same-weekday days of the case's calendar month, before AND after. Retained only as a
# secondary sensitivity to the unidirectional primary.
build_daily_bi <- function(panel, flag_ids, max_lag = 28) {
  p <- copy(panel)
  fl <- p[outbreak_binary == 1 & date >= case_window_start & date <= case_window_end
        ][, nf := vapply(outbreak_id, count_flagged_ids, integer(1), flagged_ids = flag_ids)
        ][nf > 0, .(zone_id, date, nf)]
  p <- merge(p, fl, by = c("zone_id", "date"), all.x = TRUE)
  p[, nf := fcoalesce(nf, 0L)]
  p[, outbreak_binary := as.integer(outbreak_count - nf > 0)]; p[, nf := NULL]

  cy <- p[date >= case_window_start & date <= case_window_end]
  cy[, stratum_id := sprintf("%d-%02d-%d-z%s", lubridate::year(date), lubridate::month(date),
                             lubridate::wday(date, week_start = 1), zone_id)]
  active <- unique(cy[outbreak_binary == 1]$stratum_id)
  pool   <- cy[stratum_id %in% active]
  pool[, case_control := fifelse(outbreak_binary == 1, "case", "control")]
  keep   <- pool[case_control == "control", .N, by = stratum_id][N >= 1]$stratum_id
  pool   <- pool[stratum_id %in% keep]

  src <- copy(p[zone_id %in% unique(pool$zone_id)]); setorder(src, zone_id, date)
  lag_cols <- character(0)
  for (v in lag_vars) {
    nm <- paste0(v, "_Lag", 0:max_lag)
    src[, (nm) := shift(.SD[[1]], n = 0:max_lag, type = "lag"), by = zone_id, .SDcols = v]
    lag_cols <- c(lag_cols, nm)
  }
  lag_mat <- merge(pool[, .(zone_id, date, stratum_id, outbreak_binary, case_control)],
                   src[, c("zone_id", "date", ..lag_cols)], by = c("zone_id", "date"), all.x = TRUE)
  non_lag <- setdiff(names(pool), names(lag_mat))
  out <- merge(lag_mat, pool[, c("zone_id", "date", ..non_lag)], by = c("zone_id", "date"))
  out <- transform_lag_cols(out, max_lag)
  list(df = out, n_keep = length(keep), n_cases = sum(out$outbreak_binary == 1))
}
```

# 5. Daily dataframes — unidirectional primary (+ bidirectional secondary)

```{r daily}
cc_daily_all <- build_uni_cc(daily, "date", 28, ALL_REMOVE)
cc_daily     <- build_uni_cc(daily, "date", 28, ind_remove(screening$ids_500m))
cc_daily_1km <- build_uni_cc(daily, "date", 28, ind_remove(screening$ids_1km))
cc_daily_200 <- build_uni_cc(daily, "date", 28, ind_remove(screening$ids_200m))
cc_daily_bi  <- build_daily_bi(daily, ind_remove(screening$ids_500m))   # secondary

saveRDS(cc_daily_all$df, paste0(objects_folder, "case_crossover_df_multistate_daily_all.RDS"))
saveRDS(cc_daily$df,     paste0(objects_folder, "case_crossover_df_multistate_daily.RDS"))
saveRDS(cc_daily_1km$df, paste0(objects_folder, "case_crossover_df_multistate_daily_1km.RDS"))
saveRDS(cc_daily_200$df, paste0(objects_folder, "case_crossover_df_multistate_daily_200m.RDS"))
saveRDS(cc_daily_bi$df,  paste0(objects_folder, "case_crossover_df_multistate_daily_bidirectional.RDS"))
fwrite(cc_daily$df,      paste0(processed_data, "case_crossover_df_multistate_daily.csv"))
cat("Daily (unidirectional) — independent: ", cc_daily$n_cases, " cases / ",
    sum(cc_daily$df$outbreak_binary == 0), " controls, ", cc_daily$n_keep, " strata | all events: ",
    cc_daily_all$n_cases, " cases | bidirectional (secondary): ", cc_daily_bi$n_cases, " cases\n", sep = "")
```

# 6. Weekly dataframes — unidirectional

```{r weekly}
cc_weekly_all <- build_uni_cc(weekly, "week_start", 4, ALL_REMOVE)
cc_weekly     <- build_uni_cc(weekly, "week_start", 4, ind_remove(screening$ids_500m))
saveRDS(cc_weekly_all$df, paste0(objects_folder, "case_crossover_df_multistate_weekly_all.RDS"))
saveRDS(cc_weekly$df,     paste0(objects_folder, "case_crossover_df_multistate_weekly.RDS"))
fwrite(cc_weekly$df,      paste0(processed_data, "case_crossover_df_multistate_weekly.csv"))
cat("Weekly (unidirectional) — independent: ", cc_weekly$n_cases, " cases / ",
    sum(cc_weekly$df$outbreak_binary == 0), " controls, ", cc_weekly$n_keep, " strata | all events: ",
    cc_weekly_all$n_cases, " cases\n", sep = "")
```

# 7. CONSORT cascade (final counts)

```{r cascade}
sc <- function(x) list(cases = x$n_cases, controls = sum(x$df$outbreak_binary == 0), strata = x$n_keep)
cascade <- list(
  states        = c("IA", "IL", "IN", "MI", "MN", "OH", "WI"),
  daily_design  = "Unidirectional case-crossover (case + 4 prior control days, 7/14/21/28 d), DLNM lag 0-28 d",
  weekly_design = "Unidirectional case-crossover (case week + 4 prior control weeks), DLNM lag 0-4 wk",
  secondary     = "Daily bidirectional (time-stratified zone x year x month x DOW) retained as sensitivity",
  case_defs     = "all = every spillover event; independent = link IND+INR minus spatio-temporal 500 m",
  daily_independent         = sc(cc_daily),
  daily_all                 = sc(cc_daily_all),
  daily_independent_bidir   = sc(cc_daily_bi),
  weekly_independent        = sc(cc_weekly),
  weekly_all                = sc(cc_weekly_all)
)
saveRDS(cascade, paste0(objects_folder, "multistate_case_crossover_cascade.RDS"))
cat("Wrote unidirectional daily + weekly dfs (all + independent, + 1km/200m daily),",
    "bidirectional daily secondary, and cascade.\n")
```

# 8. No-spatial-restriction panels (the locked primary case definition)

```{r nospatial}
# The §5/§6 primary strips cross-farm CSLT/NOSEQ *and* the 500 m spatio-temporal events.
# The final analysis drops that spatial filter: CSLT exclusion is the principled,
# investigation-based criterion and the distance cut is an arbitrary proxy for the same
# thing that costs cases. These panels are what c_11 actually fits, for both the pooled
# flyway and the MN subset (Minnesota lives inside this grid, so MN and the flyway are
# finally on the same cells, same bird aggregation, same surface-runoff definition).
cc_daily_ns  <- build_uni_cc(daily,  "date",       28, cross_farm_ids)
cc_weekly_ns <- build_uni_cc(weekly, "week_start",  4, cross_farm_ids)

saveRDS(cc_daily_ns$df,  paste0(objects_folder, "case_crossover_df_multistate_daily_nospatial.RDS"))
saveRDS(cc_weekly_ns$df, paste0(objects_folder, "case_crossover_df_multistate_weekly_nospatial.RDS"))

cat("flyway pooled, no spatial restriction: ", cc_daily_ns$n_cases, " cases / ",
    cc_daily_ns$n_keep, " strata\n", sep = "")
print(setDT(cc_daily_ns$df)[outbreak_binary == 1, .(cases = .N), by = state][order(-cases)])
```
