---
title: "Join data sources — multi-state (7-state flyway, daily + weekly)"
author: "Fintan Mooney"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

```{r initiate}
rm(list = ls())
# 0a.Declare root directory, folder location and load essential stuff
project.folder = paste0(print(here::here()), '/')
source(paste0(project.folder, 'create_folder_structure.R'))
source(paste0(functions.folder, 'script_initiate.R'))

# 0b.lubridate owns the date accessors under the project's conflicted setup.
conflict_prefer("year",  "lubridate", quiet = TRUE)
conflict_prefer("month", "lubridate", quiet = TRUE)
```

Multi-state analogue of `a_01_join_data.Rmd`. The MN file loads the multi-state ERA5 pull
and **filters to Minnesota**, mapping multistate cells onto the legacy MN fishnet. This
file instead keeps the **7 flyway study states** (IA, IL, IN, MI, MN, OH, WI) and works
directly in the 7-state fishnet's own `zone_id` space — the multistate ERA5 CSVs already
carry those `zone_id`s, so no legacy remapping is needed. South Dakota has been removed
from the multi-state pipeline (no eBird extraction).

We build **two** panels, per the design:

  * **daily**  — daily ERA5 + bird abundance smoothed (weekly→daily) within each zone.
  * **weekly** — ERA5 aggregated to ISO-week (means for state vars, sums for fluxes) +
                 native weekly bird abundance.

`a_03_zone_bird_abundance.Rmd` must be run first (it writes the per-zone
weekly bird table this script consumes).

```{r study-region}
# 0c.Study region. These codes index the spillover/ERA5 inputs and the fishnet alike.
study_states_abbr <- c("IA", "IL", "IN", "MI", "MN", "OH", "WI")

# Study window: 2021 warm-up through the end of the 2022 case year. Cases are still defined
# within calendar 2022; the earlier start only buys exposure history.
# 2021-08-01, not 2021-11-15. The scaffold below is what defines the panel, so anything outside
# this window is dropped on the join no matter what ERA5 covers. The 90-day rolling baseline needs
# 90 days of history and the DLNM another 28 on top, so a case on 2022-02-08 needs met data back to
# 2021-10-13, and its February referents back to 2021-10-06. Starting at 2021-11-15 silently cost
# 8 cases (167 of 175) and 8 cells (153 of 161). The 2021 warm-up export covers 2021-08-01 on.
study_start <- as.Date("2021-08-01")
study_end   <- as.Date("2022-12-31")
date_seq    <- seq(study_start, study_end, by = "day")
```

# 1. Load the multi-state fishnet (7 states)

```{r fishnet}
# GEE writes EPSG:5070 coordinates but mislabels the CRS as WGS84 — *set*, don't
# transform (identical to c_04 and the MN a_04 multistate block). The 7-state
# fishnet already excludes South Dakota.
fishnet_geom <- st_read(paste0(env_data, "hpai_multistate_era5/fishnet_7state.geojson"),
                        quiet = TRUE)
fishnet_geom <- fishnet_geom[!st_is_empty(fishnet_geom), ]
fishnet_geom <- st_set_crs(fishnet_geom, 5070)
fishnet_geom <- fishnet_geom %>% distinct(zone_id, state, geometry)

cat("Fishnet zones by state:\n")
print(as.data.frame(table(fishnet_geom$state)))
```

# 2. Build the zone × date scaffold

```{r scaffold}
zone_state <- as.data.table(st_drop_geometry(fishnet_geom))   # zone_id ↔ state lookup
zone_ids   <- unique(zone_state$zone_id)

# CJ (data.table cross-join) is far lighter than expand.grid at ~4.9 M rows.
daily_fishnet <- CJ(zone_id = zone_ids, date = date_seq)
daily_fishnet <- merge(daily_fishnet, zone_state, by = "zone_id", all.x = TRUE)
daily_fishnet[, `:=`(
  year     = lubridate::year(date),
  month    = lubridate::month(date),
  iso_week = lubridate::isoweek(date),
  iso_year = lubridate::isoyear(date),
  dow      = lubridate::wday(date, label = TRUE, week_start = 1),
  doy      = lubridate::yday(date)
)]
setkey(daily_fishnet, zone_id, date)
cat("Scaffold: ", format(nrow(daily_fishnet), big.mark = ","), " zone-days\n", sep = "")
```

# 3. Load and join spillover events (daily, 7 states)

```{r spillover}
# 3a.Case set = ALL outbreaks in the multi-state bird panel (every link type,
#    7 states). The panel carries all events so a_08 can build both the all-events
#    and the independent-only case-crossover dataframes (a_08 re-derives each
#    outbreak's link from the bird panel and removes cross-farm/unsequenced events
#    via the same outbreak_id matching used for the spatio-temporal screen).
#    Defining cases from the bird panel guarantees the daily cases are *exactly* the
#    bird-data outbreaks (all carry valid coords + a 2022 date_confirmed).
bird_outbreak_ids <- fread(paste0(bird_abundance_data,
  "study_spillover_events_casecrossover_2022_abundance_060326.csv"))[
    is_case_week == 1 & state %in% study_states_abbr, unique(outbreak_id)]

spillover_file <- paste0(spillover_data, "hpai-outbreaks-clean-placeholder-032124.csv")
if (!file.exists(spillover_file)) stop("Missing ", spillover_file)
spillover_raw <- read_csv(spillover_file)

# 3b.date_confirmed is the case-defining date (USDA surveillance convention). Keep
#    the bird-panel outbreaks within the 2022 case year.
spillover_study <- spillover_raw %>%
  mutate(case_date = as.Date(date_confirmed)) %>%
  filter(outbreak_id %in% bird_outbreak_ids,
         !is.na(lon), !is.na(lat),
         dplyr::between(case_date, as.Date("2022-01-01"), as.Date("2022-12-31")))

spillover_sf <- spillover_study %>%
  st_as_sf(coords = c("lon", "lat"), crs = 4326) %>%
  st_transform(crs = st_crs(fishnet_geom))

outbreaks_in_cells <- st_join(spillover_sf, fishnet_geom, join = st_within) %>%
  st_drop_geometry()

outbreak_daily <- outbreaks_in_cells %>%
  filter(!is.na(zone_id)) %>%
  group_by(zone_id, date = case_date) %>%
  summarise(
    outbreak_count = n(),
    outbreak_id    = paste(outbreak_id, collapse = ", "),
    .groups = "drop"
  ) %>%
  mutate(outbreak_binary = 1L) %>%
  as.data.table()

any_outbreak_zones     <- outbreak_daily[, .(any_outbreak = 1L), by = zone_id]
outbreak_dates_by_zone <- outbreak_daily[
  , .(outbreak_date = paste(sort(unique(as.character(date))), collapse = ", ")),
  by = zone_id
]

setkey(outbreak_daily, zone_id, date)
daily_fishnet <- merge(daily_fishnet, outbreak_daily,
                       by = c("zone_id", "date"), all.x = TRUE)
daily_fishnet <- merge(daily_fishnet, any_outbreak_zones, by = "zone_id", all.x = TRUE)
daily_fishnet <- merge(daily_fishnet, outbreak_dates_by_zone, by = "zone_id", all.x = TRUE)
daily_fishnet[, `:=`(
  outbreak_count  = fcoalesce(outbreak_count, 0L),
  outbreak_binary = fcoalesce(outbreak_binary, 0L),
  any_outbreak    = fcoalesce(any_outbreak, 0L),
  outbreak_id     = fcoalesce(outbreak_id, NA_character_),
  outbreak_date   = fcoalesce(outbreak_date, NA_character_)
)]
cat("Cases (zone-days) joined: ", sum(daily_fishnet$outbreak_binary),
    " across ", uniqueN(daily_fishnet[outbreak_binary == 1]$zone_id), " zones\n", sep = "")
```

# 4. Load daily ERA5-Land met (7 states, native zone_id)

```{r met}
multistate_dir <- paste0(env_data, "hpai_multistate_era5/")

# 4a.Load the bi-weekly CSVs for the 7 study states only. The CSV `zone_id`
#    matches the fishnet `zone_id` directly (e.g. "-4,227"), so no remapping.
state_pat <- paste0("^era5_(", paste(study_states_abbr, collapse = "|"), ")_.*\\.csv$")
met_files <- list.files(multistate_dir, pattern = state_pat, full.names = TRUE)
if (length(met_files) == 0) stop("No multistate ERA5 CSVs found in ", multistate_dir)
cat("Loading ", length(met_files), " bi-weekly ERA5 CSVs (7 states)...\n", sep = "")

met_daily <- rbindlist(lapply(met_files, fread, colClasses = list(character = "zone_id")),
                       use.names = TRUE, fill = TRUE)
met_daily[, date := as.Date(as.character(date))]   # GEE format "YYYY-MM-DD"

# 4b.Rename raw ERA5 bands to the friendly names the rest of the pipeline expects.
raw_to_friendly <- c(
  "temperature_2m"                = "temperature",
  "dewpoint_temperature_2m"       = "dewpoint",
  "soil_temperature_level_1"      = "soil_temperature",
  "volumetric_soil_water_layer_1" = "soil_moisture",
  "volumetric_soil_water_layer_2" = "soil_moisture_2",
  "total_precipitation_hourly"    = "precipitation",
  "snowfall_hourly"               = "snowfall",
  "surface_runoff_hourly"         = "runoff",
  "total_evaporation_hourly"      = "evapotranspiration",
  "surface_solar_radiation_downwards_hourly"   = "shortwave_radiation",
  "surface_thermal_radiation_downwards_hourly" = "longwave_radiation",
  "leaf_area_index_low_vegetation" = "leaf_area_index"
)
for (raw in names(raw_to_friendly)) {
  if (raw %in% names(met_daily) && raw_to_friendly[[raw]] != raw) {
    setnames(met_daily, raw, raw_to_friendly[[raw]])
  }
}

# 4c.Derive wind_speed and relative_humidity (Magnus formula; ERA5-Land in K).
met_daily[, wind_speed := sqrt(u_component_of_wind_10m^2 + v_component_of_wind_10m^2)]
T_C  <- met_daily$temperature - 273.15
Td_C <- met_daily$dewpoint   - 273.15
met_daily[, relative_humidity := 100 *
            exp((17.625 * Td_C) / (243.04 + Td_C)) /
            exp((17.625 * T_C ) / (243.04 + T_C ))]
# snow_density ~100 is the no-snow sentinel → 0.
if ("snow_density" %in% names(met_daily)) {
  met_daily[, snow_density := fifelse(snow_density <= 100, 0, snow_density)]
}

# 4d.Keep the canonical met columns present, then join on zone_id × date.
met_cols <- c("temperature", "dewpoint", "relative_humidity", "wind_speed",
              "skin_temperature", "soil_temperature",
              "soil_moisture", "soil_moisture_2",
              "snow_depth", "snow_cover", "snow_density",
              "precipitation", "snowfall", "runoff",
              "evapotranspiration", "leaf_area_index", "shortwave_radiation",
              "longwave_radiation", "surface_pressure")
met_cols  <- intersect(met_cols, names(met_daily))
met_daily <- met_daily[, c("zone_id", "date", ..met_cols)]

# 4e.Drop duplicate zone-days. Cells on a state border are exported under both
#    adjacent states' CSVs (~106 zones, every date), with byte-identical values, so
#    keep-first is lossless. Without this, the join would double zone-days and
#    double-count summed fluxes (precip/runoff) in the weekly aggregation.
n_before  <- nrow(met_daily)
met_daily <- unique(met_daily, by = c("zone_id", "date"))
cat("Dropped ", format(n_before - nrow(met_daily), big.mark = ","),
    " duplicate border-cell zone-days.\n", sep = "")
setkey(met_daily, zone_id, date)

daily_fishnet <- merge(daily_fishnet, met_daily, by = c("zone_id", "date"), all.x = TRUE)
rm(met_daily); gc()
```

# 5. Per-zone weekly bird abundance → smoothed to daily

```{r bird}
# 5a.Read the per-zone weekly table from a_03_zone_bird_abundance.Rmd.
bird_file <- paste0(bird_abundance_data, "multistate-zone-weekly-bird-abundance-061626.csv")
if (!file.exists(bird_file)) {
  stop("Missing ", bird_file, " — run a_03_zone_bird_abundance.Rmd first.")
}
bird_zone_week <- fread(bird_file, colClasses = list(character = "zone_id"))
bird_zone_week[, week_start := as.Date(week_start)]

# 5b.Anchor each ISO-week at its midpoint (Thursday = week_start + 3) and linearly
#    interpolate within each zone to daily, matching the MN a_04 approach. Only
#    zones that actually hold bird data (case zones) are interpolated; every other
#    zone carries NA→0, since the case-crossover never references control zones.
bird_zone_week[, anchor_date := week_start + 3]
setorder(bird_zone_week, zone_id, anchor_date)

interpolate_bird_daily <- function(anchors_df, target_dates) {
  vars <- c("total_bird_abundance", "anseriformes", "predators")
  out  <- data.frame(date = target_dates)
  for (v in vars) {
    ok <- !is.na(anchors_df[[v]])
    if (sum(ok) >= 2) {
      f <- stats::approxfun(anchors_df$anchor_date[ok], anchors_df[[v]][ok],
                            method = "linear", rule = 2)
      out[[v]] <- f(target_dates)
    } else if (sum(ok) == 1) {
      out[[v]] <- rep(anchors_df[[v]][ok][1], length(target_dates))
    } else {
      out[[v]] <- NA_real_
    }
  }
  out
}

bird_daily <- rbindlist(lapply(
  split(bird_zone_week, bird_zone_week$zone_id),
  function(g) {
    d <- interpolate_bird_daily(as.data.frame(g[order(anchor_date)]), date_seq)
    d$zone_id <- g$zone_id[1]
    d
  }
))
setcolorder(bird_daily, c("zone_id", "date", "total_bird_abundance",
                          "anseriformes", "predators"))
setkey(bird_daily, zone_id, date)

daily_fishnet <- merge(daily_fishnet, bird_daily, by = c("zone_id", "date"), all.x = TRUE)
for (v in c("total_bird_abundance", "anseriformes", "predators")) {
  daily_fishnet[is.na(get(v)), (v) := 0]
}

# 5c.Flag zones that never carry bird abundance (all control zones), so downstream
#    code can exclude them from bird-based summaries.
zone_has_bird <- daily_fishnet[, .(tot = sum(total_bird_abundance, na.rm = TRUE)), by = zone_id]
no_bird_zones <- zone_has_bird[tot == 0, zone_id]
daily_fishnet[, bird_zero := fifelse(zone_id %in% no_bird_zones, "No Birds", "Has Birds")]
cat("Zones with bird abundance: ", uniqueN(zone_has_bird[tot > 0]$zone_id), "\n", sep = "")
```

# 6. Feedlot covariates (Minnesota only)

```{r feedlots}
# Feedlot density is a Minnesota Dept. of Agriculture product with no multi-state
# equivalent, so it is attached for MN zones only (per project decision). It is a
# zone-constant covariate and therefore drops out of the conditional-logistic
# case-crossover — it is retained here for descriptive/MN-subgroup use. The shape
# carries its own lat/long, so we spatially place feedlots in the 7-state fishnet
# rather than relying on the legacy MN zone_id.
feedlot_shp <- paste0(feedlot_data, "feedlots.shp")
if (file.exists(feedlot_shp)) {
  feedlots <- st_read(feedlot_shp, quiet = TRUE) %>%
    st_transform(crs = st_crs(fishnet_geom)) %>%
    mutate(
      cafo_flag  = ifelse(!is.na(cafo_flag) & cafo_flag != "N", 1L, 0L),
      near_river = ifelse(!is.na(river_stre) & river_stre != "N", 1L, 0L)
    )

  feedlots_in_cells <- st_join(feedlots, fishnet_geom, join = st_within) %>%
    st_drop_geometry() %>%
    filter(!is.na(zone_id))

  feedlot_summary <- feedlots_in_cells %>%
    group_by(zone_id) %>%
    summarise(
      feedlot_count = n(),
      au_count      = sum(au_count, na.rm = TRUE),
      birds_au      = sum(birds_au, na.rm = TRUE),
      cafo_flag     = max(cafo_flag, na.rm = TRUE),
      near_river    = max(near_river, na.rm = TRUE),
      .groups = "drop"
    ) %>%
    as.data.table()

  daily_fishnet <- merge(daily_fishnet, feedlot_summary, by = "zone_id", all.x = TRUE)
  rm(feedlots, feedlots_in_cells); gc()
} else {
  message("feedlots.shp not found — feedlot covariates left as 0.")
  daily_fishnet[, `:=`(feedlot_count = NA_integer_, au_count = NA_real_,
                       birds_au = NA_real_, cafo_flag = NA_integer_,
                       near_river = NA_integer_)]
}
daily_fishnet[, `:=`(
  feedlot_count = fcoalesce(feedlot_count, 0L),
  au_count      = fcoalesce(au_count,      0),
  birds_au      = fcoalesce(birds_au,      0),
  cafo_flag     = fcoalesce(cafo_flag,     0L),
  near_river    = fcoalesce(near_river,    0L)
)]
```

# 7. Land cover (DEFERRED — requires the multi-state NLCD/GEE pull)

```{r landcover}
# Land cover is not an exposure in any model (a case-crossover conditions every cell-level
# attribute out), so the NLCD join that used to sit here is gone. The two columns stay in the
# panel, empty, so downstream scripts that select them keep working.
daily_fishnet[, `:=`(land_cover_mode = NA, land_cover_label = NA_character_, farm_possible = NA)]
```

# 8. Build the weekly panel (ERA5 → ISO-week, native weekly birds)

```{r weekly}
# 8a.Aggregate the daily met to ISO-week per zone: means for state variables,
#    sums for fluxes (matching the c_04 weekly aggregation).
state_vars <- intersect(c("temperature", "dewpoint", "relative_humidity", "wind_speed",
                          "skin_temperature", "soil_temperature", "soil_moisture",
                          "soil_moisture_2", "snow_depth", "snow_cover", "snow_density",
                          "leaf_area_index", "shortwave_radiation", "longwave_radiation",
                          "surface_pressure"), names(daily_fishnet))
flux_vars  <- intersect(c("precipitation", "snowfall", "runoff", "evapotranspiration"),
                        names(daily_fishnet))

weekly_met <- daily_fishnet[, c(
  lapply(.SD[, state_vars, with = FALSE], mean, na.rm = TRUE),
  lapply(.SD[, flux_vars,  with = FALSE], sum,  na.rm = TRUE)
), by = .(zone_id, state, iso_year, iso_week)]

# 8b.Weekly outbreak flags (a zone-week is a case if any day that week was a case).
#    Carry the comma-joined outbreak_ids so a_08 can apply the link/independence
#    split to the weekly panel exactly as it does to the daily panel.
weekly_out <- daily_fishnet[, .(
  outbreak_count  = sum(outbreak_count),
  outbreak_binary = as.integer(sum(outbreak_binary) > 0),
  any_outbreak    = max(any_outbreak),
  outbreak_id     = paste(na.omit(outbreak_id), collapse = ", "),
  week_start      = min(date)
), by = .(zone_id, iso_year, iso_week)]
weekly_out[outbreak_id == "", outbreak_id := NA_character_]

weekly_fishnet <- merge(weekly_out, weekly_met,
                        by = c("zone_id", "iso_year", "iso_week"), all = TRUE)

# 8c.Native weekly bird abundance (no interpolation), joined by zone × ISO-week.
weekly_fishnet <- merge(
  weekly_fishnet,
  bird_zone_week[, .(zone_id, iso_year, iso_week,
                     total_bird_abundance, anseriformes, predators)],
  by = c("zone_id", "iso_year", "iso_week"), all.x = TRUE
)
for (v in c("total_bird_abundance", "anseriformes", "predators")) {
  weekly_fishnet[is.na(get(v)), (v) := 0]
}

# 8d.Carry the zone-constant feedlot covariates onto the weekly panel.
weekly_fishnet <- merge(
  weekly_fishnet,
  unique(daily_fishnet[, .(zone_id, feedlot_count, au_count, birds_au,
                           cafo_flag, near_river, bird_zero)]),
  by = "zone_id", all.x = TRUE
)
setkey(weekly_fishnet, zone_id, iso_year, iso_week)
cat("Weekly panel: ", format(nrow(weekly_fishnet), big.mark = ","), " zone-weeks\n", sep = "")
```

# 9. Export

```{r export}
setkey(daily_fishnet, zone_id, date)
saveRDS(daily_fishnet,  paste0(objects_folder, "multistate_fishnet_daily_processed.rds"))
saveRDS(weekly_fishnet, paste0(objects_folder, "multistate_fishnet_weekly_processed.rds"))
fwrite(daily_fishnet,   paste0(processed_data, "multistate_fishnet_daily_processed.csv"))
fwrite(weekly_fishnet,  paste0(processed_data, "multistate_fishnet_weekly_processed.csv"))

# Spatial summary for mapping (one row per zone).
fishnet_spatial <- fishnet_geom %>%
  left_join(
    daily_fishnet[, .(any_outbreak  = max(any_outbreak),
                      feedlot_count = first(feedlot_count),
                      bird_zero     = first(bird_zero)),
                  by = zone_id],
    by = "zone_id"
  )
saveRDS(fishnet_spatial, paste0(objects_folder, "multistate_fishnet_processed_sf.rds"))
cat("Wrote daily + weekly multi-state panels.\n")
```
