---
title: "create_casecrossover"
author: "Fintan Mooney"
date: "2025-10-15"
output: html_document
---

# This R markdown file creates the casecrossover dataframe for the projet on the association between meterological conditons, climate, and wild bird abundance, on risk of highly pathogenic avian influenza (HPAI) outbreaks in Minnesota farms during 2022

```{r setup, include=FALSE}
 
knitr::opts_chunk$set(echo = TRUE)
 
# ---- Package setup ----
packages <- c(
  "MatchIt", "readr", "sf", "survival", "dlnm", "gtsummary", "kableExtra",
  "tidyr", "tibble", "mgcv", "ggplot2", "car", "patchwork", "corrplot",
  "purrr", "ggspatial", "gt", "flextable", "officer", "conflicted",
  "dplyr", "data.table", "here", "stringr"
)

installed <- packages %in% rownames(installed.packages())
if (any(!installed)) install.packages(packages[!installed])

invisible(lapply(packages, library, character.only = TRUE))

# Resolve conflicts
conflict_prefer("select", "dplyr")
conflict_prefer("filter", "dplyr")

# Project folders (e.g., where the .Rproj file is)
base_folder <- here()
data_folder <- here("data")
env_folder <- here("data", "EarthEngine_MN")
objects_folder <-here("data", "objects")
```


## Read in the timeseries dataset

```{r, read in timeseries dataset}

# Timeseries
fishnet_timeseries_clean <- readRDS(file.path(objects_folder, "fishnet_timeseries.RDS"))

fishnet_timeseries_clean <- fishnet_timeseries_clean %>%
  arrange(zone_id, week) %>%
  # 1. Drop all columns that end with ".x"
  select(-ends_with(".x")) %>%
  # 2. Rename columns ending with ".y" to remove the suffix
  rename_with(~ str_remove(.x, "\\.y$"), ends_with(".y"))

# Spatial GPKG
gpkg_path <- file.path(data_folder, "Fishnet_Timeseries_10km.gpkg")
fishnet_weeks <- st_read(gpkg_path)

## Simplify
fishnet_weeks <- fishnet_weeks %>%
  dplyr::select(week, shape_length, shape_area, zone_id, geom)

```

## Map Cases in Casecrossover Model

```{r, casecrossover map}


# Map 
# Identify case zones
case_zone_ids <- fishnet_timeseries_clean %>%
  filter(any_outbreak == 1) %>%
  distinct(zone_id)

# Subset spatial features
case_zones_sf <- fishnet_weeks %>% filter(zone_id %in% case_zone_ids$zone_id)
control_zones_sf <- fishnet_weeks %>% filter(!(zone_id %in% case_zone_ids$zone_id))

# Define output paths
case_gpkg_out    <- file.path(data_folder, "Case_Zones.gpkg")
control_gpkg_out <- file.path(data_folder, "Control_Zones.gpkg")

# Export as GPKG
st_write(case_zones_sf, case_gpkg_out,    delete_layer = TRUE)
st_write(control_zones_sf, control_gpkg_out, delete_layer = TRUE)

# Reproject to EPSG:3857 (Web Mercator) for basemap compatibility
case_zones_sf_proj <- st_transform(case_zones_sf, 3857)
control_zones_sf_proj <- st_transform(control_zones_sf, 3857)

# Plot with basemap
ggplot() +
  # Add simple basemap
  annotation_map_tile(type = "cartolight", zoomin = 0) +

  # Control zones: outlines
  geom_sf(data = control_zones_sf_proj, fill = NA, color = "lightblue", linewidth = 0.05) +

  # Case zones: bold fill
  geom_sf(data = case_zones_sf_proj, fill = "#d73027", color = "red", linewidth = 0.1, alpha = 0.9) +

  # Scale bar
  annotation_scale(location = "bl", width_hint = 0.3, text_cex = 0.75, line_width = 0.5, style = "ticks") +

  # Theme and layout
  theme_void(base_size = 12) +
  theme(
    plot.title = element_text(size = 12, face = "bold", hjust = 0.5),
    plot.caption = element_text(size = 10, hjust = 0.5, face = "italic"),
    plot.background = element_rect(fill = "white", color = NA)
  ) +
  coord_sf(expand = FALSE) +
  labs(
    title = "HPAI Outbreak Case Zones in Minnesota (2022)"
  )
```

## Define Controls and Create Case-Crossover Dataset
```{r, define controls}

# ---- Define Cases and Controls ----

# Filter to zones with any outbreak
case_zones <- fishnet_timeseries_clean %>%
  filter(any_outbreak == 1) 

# Get cases: one row per zone_id and week with outbreak
cases <- case_zones %>%
  filter(outbreak_count > 0) %>%
  distinct(zone_id, case_week = week)

# Deduplicate case_zones for joining
case_zones_unique <- case_zones %>%
  distinct(zone_id, week, .keep_all = TRUE)

# Create control weeks (1-4 weeks before each case)
case_lags <- cases %>%
  crossing(lag = 1:4) %>%
  mutate(control_week = case_week - lag) %>%
  filter(control_week >= 1)

# Remove overlapping controls (control week = case week in same zone)
case_zone_weeks <- cases %>% select(zone_id, case_week)

case_lags_clean <- case_lags %>%
  anti_join(case_zone_weeks, by = c("zone_id", "control_week" = "case_week"))

cat("Controls removed (overlap with case weeks):", nrow(case_lags) - nrow(case_lags_clean), "\n")

### HERE WE REMOVED CONTROLS THAT OVERLAP
#### Specifically, some cells have outbreaks at two different farms within the same four week period; we do not delete them as cases, but we remove the overlapping controls. 

# Build controls dataset
controls <- case_lags_clean %>%
  left_join(case_zones_unique, by = c("zone_id", "control_week" = "week")) %>%
  mutate(
    case_control = "control",
    outbreak_binary = 0,
    stratum_id = paste(zone_id, case_week, sep = "_"),
    week = control_week
  ) %>%
  dplyr::select(-outbreak_count)

# Build cases dataset
case_data <- cases %>%
  left_join(case_zones_unique, by = c("zone_id", "case_week" = "week")) %>%
  mutate(
    case_control = "case",
    outbreak_binary = 1,
    stratum_id = paste(zone_id, case_week, sep = "_"),
    week = case_week,
    control_week = NA
  ) %>%
  dplyr::select(-outbreak_count)

# Combine cases and controls
case_crossover_df <- bind_rows(case_data, controls)

# ---- Summary checks ----
cat("\nCase-control distribution:\n")
table(case_crossover_df$case_control)

cat("\nTotal cases:", sum(case_crossover_df$outbreak_binary), "\n")

# Check controls per case
controls_per_case <- case_crossover_df %>%
  filter(case_control == "control") %>%
  count(stratum_id, name = "n_controls")

cat("\nControls per case:\n")
table(controls_per_case$n_controls)

colnames(case_crossover_df)



```

## Validate control selection for zones with multiple cases
```{r, validate multi-case zones}

# Check: how many zone-weeks have outbreaks?
case_zones %>%
  filter(outbreak_count > 0) %>%
  distinct(zone_id, week) %>%
  nrow()

# vs total outbreak events
print(sum(case_zones$outbreak_count))



# Zones with multiple cases
multi_case_zones <- cases %>%
  group_by(zone_id) %>%
  summarise(n_cases = n(), case_weeks = list(case_week), .groups = "drop") %>%
  filter(n_cases > 1)

cat("Zones with multiple cases:", nrow(multi_case_zones), "\n")
cat("Total cases in multi-case zones:", sum(multi_case_zones$n_cases), "\n")

# Check: Do any control weeks overlap with case weeks in the same zone?
case_zone_weeks <- cases %>% select(zone_id, case_week)

proposed_controls <- cases %>%
  crossing(lag = 1:4) %>%
  mutate(control_week = case_week - lag) %>%
  filter(control_week >= 1)

# Flag controls that are also case weeks in the same zone
overlap <- proposed_controls %>%
  inner_join(case_zone_weeks, by = c("zone_id", "control_week" = "case_week"))

cat("Control weeks that are ALSO case weeks in same zone:", nrow(overlap), "\n")

if (nrow(overlap) > 0) {
  print(overlap)
} else {
  cat("✓ No overlap — all control weeks are case-free in their respective zones.\n")
}

## WE DO NOT REMOVE CASES BECAUSE EVEN THOUGH THEY MAY HAVE HAPPENED IN THE SAME ZONE OVER THE FOUR WEEK PERIOD THEY WERE FROM DIFFERENT FARMS


# Check for duplicate zone-week combinations in controls
control_dupes <- case_crossover_df %>%
  filter(case_control == "control") %>%
  count(zone_id, week) %>%
  filter(n > 1)

cat("Control zone-weeks appearing more than once:", nrow(control_dupes), "\n")

if (nrow(control_dupes) > 0) {
  print(control_dupes)
  
  # Show which strata these belong to
  case_crossover_df %>%
    filter(case_control == "control") %>%
    inner_join(control_dupes, by = c("zone_id", "week")) %>%
    select(zone_id, week, stratum_id, case_week) %>%
    arrange(zone_id, week) %>%
    print(n = 30)
}

## WE DO NOT REMOVE CONTROLS ZONE-WEEKS THAT APPEAR MORE THAN ONE BECAUSE THEY COME FROM DIFFERENT STRATA AND THUS ARE SUITABLE FOR THE CONDITONAL LOGISTIC REGRESSION
```

# Check distribution of casecrossover variables

```{r, distribution caseccrossover}

# Move from m to mm

case_crossover_df <- case_crossover_df %>%
  mutate(
    runoff = runoff * 1000,
    precipitation = precipitation * 1000
  )


# Variables to inspect
vars_to_check <- c(
  "total_bird_abundance",   
  "leaf_area_index", "longwave_radiation", "runoff", "shortwave_radiation",
  "snow_density", "soil_moisture", "temperature",
  "evapotranspiration", "precipitation", "wind_speed", "relative_humidity", "anseriformes", "predators"
)

# Create histograms
plots <- lapply(vars_to_check, function(var) {
  ggplot(case_crossover_df, aes_string(x = var)) +
    geom_histogram(bins = 50, fill = "steelblue", color = "white") +
    labs(title = var) +
    theme_minimal()
})

# Display plots in a grid (adjust number of columns if needed)
wrap_plots(plots, ncol = )

library(e1071)
# Subset to variables
numeric_vars <- case_crossover_df %>%
  select(all_of(vars_to_check))

# Initialize result storage
shapiro_results <- lapply(names(numeric_vars), function(var) {
  x <- numeric_vars[[var]]
  x <- na.omit(x)
  
  # Only apply Shapiro if <= 5000 non-NA values
  shapiro_p <- if (length(x) <= 5000) {
    shapiro.test(x)$p.value
  } else {
    NA
  }
  
  data.frame(
    variable = var,
    skewness = skewness(x),
    shapiro_p = shapiro_p,
    n = length(x)
  )
})

# Combine into summary table
shapiro_df <- bind_rows(shapiro_results)

# Flag non-normality
shapiro_df <- shapiro_df %>%
  mutate(non_normal = ifelse(!is.na(shapiro_p) & shapiro_p < 0.05, TRUE, FALSE)) %>%
  arrange(desc(abs(skewness)))

# View top
print(shapiro_df)

### Need to transform runoff, total_bird_abundance, specific humidity, and precipitation, anseriformes and predators


```

## Create lagged variables

```{r, lag variables}

# Drop duplicates from original_df to ensure safe left_joins
case_zones_unique <- case_zones %>%
  distinct(zone_id, week, .keep_all = TRUE)


add_lagged_predictors <- function(df, original_df, vars, id_col = "zone_id", time_col = "week", max_lag = 4) {
  for (varname in vars) {
    for (lag in 0:max_lag) {
      lagged_col <- paste0(varname, "_Lag", lag)
      df <- df %>%
        left_join(
          original_df %>%
            dplyr::select(!!sym(id_col), !!sym(time_col), !!sym(varname)) %>%
            rename(!!lagged_col := !!sym(varname)) %>%
            mutate(!!time_col := !!sym(time_col) + lag),
          by = c(id_col, time_col)
        )
    }
  }
  return(df)
}


vars_to_lag <- c( "total_bird_abundance",
                 "leaf_area_index",
                 "longwave_radiation",
                 "runoff",
                 "shortwave_radiation",
                 "snow_density",
                 "soil_moisture",
                 "temperature",
                 "evapotranspiration",
                 "precipitation",
                 "wind_speed", 
                 "relative_humidity",
                 "anseriformes", "predators"
                 )

case_crossover_df <- add_lagged_predictors(
  df = case_crossover_df,
  original_df = case_zones_unique,
  vars = vars_to_lag,
  max_lag = 4
)

# Verify lagged variables were added correctly
cat("Columns in case_crossover_df:", ncol(case_crossover_df), "\n")
cat("Rows in case_crossover_df:", nrow(case_crossover_df), "\n")

# Check for missing values in lagged variables
case_crossover_df %>%
  select(ends_with("_Lag0")) %>%
  summarise(across(everything(), ~sum(is.na(.)))) %>%
  pivot_longer(everything(), names_to = "var", values_to = "n_missing") %>%
  filter(n_missing > 0)

## NOTHING IS MISSING
```




## Transform and Scale Variables
```{r, transform variables}

# Apply log1p only to skewed ones before scaling
for (var in c("total_bird_abundance", "runoff", "precipitation",  "anseriformes", "predators")) {
  for (lag in 0:4) {
    col <- paste0(var, "_Lag", lag)
    if (col %in% names(case_crossover_df)) {
      case_crossover_df[[col]] <- log1p(case_crossover_df[[col]])
    }
  }
}

```

```{r, scale}

# Filter to 2022 lag variables

vars_to_lag <- c("total_bird_abundance",
                 "leaf_area_index",
                 "longwave_radiation",
                 "runoff",
                 "shortwave_radiation",
                 "snow_density",
                 "soil_moisture",
                 "temperature",
                 "evapotranspiration",
                 "precipitation",
                 "wind_speed", 
                 "relative_humidity",
                 "anseriformes", "predators")

# Scale all lags (Lag0 to Lag4 for each variable)
for (var in vars_to_lag) {
  for (lag in 0:4) {
    lagged_col <- paste0(var, "_Lag", lag)
    case_crossover_df[[lagged_col]] <- scale(case_crossover_df[[lagged_col]])
  }
}
```

## Check that no control weeks have cases

```{r, validation check}

# Check for overlap in final dataset
case_weeks <- cases %>% select(zone_id, case_week)

overlap_check <- case_crossover_df %>%
  filter(case_control == "control") %>%
  inner_join(case_weeks, by = c("zone_id", "week" = "case_week"))

cat("Control weeks overlapping with case weeks:", nrow(overlap_check), "\n")  # should be 0
```

## Count number of cases and controls again

```{r, count number of cases and controls again}


# Cases and controls
cat("Total cases (zone-weeks):", sum(case_crossover_df$outbreak_binary == 1), "\n")
cat("Total controls (zone-weeks):", sum(case_crossover_df$outbreak_binary == 0), "\n")

# Total outbreak events (sum of outbreak_count for case rows)
total_outbreaks <- case_crossover_df %>%
  filter(outbreak_binary == 1) %>%
  left_join(
    case_zones_unique %>% select(zone_id, week, outbreak_count),
    by = c("zone_id", "week")
  ) %>%
  summarise(total = sum(outbreak_count, na.rm = TRUE)) %>%
  pull(total)

cat("Total outbreak events:", total_outbreaks, "\n")

# Summary table
cat("\nCase-control distribution:\n")
table(case_crossover_df$case_control)

# Controls per case
cat("\nControls per case:\n")
case_crossover_df %>%
  filter(case_control == "control") %>%
  count(stratum_id) %>%
  count(n, name = "n_cases") %>%
  rename(n_controls = n)

```

### Export casecrossover df

```{r, export casecrossover df}

saveRDS(case_crossover_df, file.path(objects_folder,"case_crossover_df.RDS"))


```