---
title: "create_timeseries_df"
author: "Fintan Mooney"
date: "2025-10-15"
output: html_document
---

# This R markdown file creates the timeseries dataframe for the projet on the association between meterological conditons, climate, and wild bird abundance, on risk of highly pathogenic avian influenza (HPAI) outbreaks in Minnesota farms during 2022

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
 
# ---- Package setup ----
packages <- c(
  "MatchIt", "readr", "sf", "survival", "stringr", "leaflet", "dlnm", "gtsummary", "kableExtra",
  "tidyr", "tibble", "mgcv", "ggplot2", "car", "patchwork", "corrplot",
  "purrr", "ggspatial", "gt", "flextable", "officer", "conflicted",
  "dplyr", "data.table", "here"
)

installed <- packages %in% rownames(installed.packages())
if (any(!installed)) install.packages(packages[!installed])

invisible(lapply(packages, library, character.only = TRUE))

# Resolve conflicts
conflict_prefer("select", "dplyr")
conflict_prefer("filter", "dplyr")
conflict_prefer("lag", "dplyr")

# Project folders (e.g., where the .Rproj file is)
base_folder <- here()
data_folder <- here("data")
env_folder <- here("data", "EarthEngine_MN")
objects_folder <-here("data", "objects")

```

## Bring in fishnet

```{r, bring in fisnet}
gpkg_path <- file.path(data_folder, "Fishnet_Timeseries_10km.gpkg")

# Read 
fishnet_weeks <- st_read(gpkg_path)

colnames(fishnet_weeks)

## Also bring in 

```


### Simplify fishnet timeseries 

```{r, simplify fishnet weeks}
fishnet_weeks <- fishnet_weeks %>%
  dplyr::select(week, shape_length, shape_area, zone_id, geom)
```

## Read in climate data

```{r, read in climate data}

data2022 <- read.csv(file.path(data_folder, "mn_weekly_era5land_2022.csv"))

data2022$week_start <- as.Date(as.character(data2022$week_start), format = "%Y%m%d")

# Create week index relative to 2022-01-01
data2022$week <- as.integer(floor(difftime(data2022$week_start, as.Date("2022-01-01"), units = "weeks")) + 1)
range(data2022$week)

colnames(data2022)
```

## Create 5-year climate periods

```{r, create baselines}

# Set folder and file list
all_years <- 1997:2022
files <- file.path(env_folder, paste0("mn_weekly_era5land_", all_years, ".csv"))

# Read all data with year column
all_data <- map2_dfr(files, all_years, ~ read_csv(.x) %>% mutate(year = .y))

# Convert week_start to Date
all_data <- all_data %>%
  mutate(week_start = as.Date(as.character(week_start), format = "%Y%m%d"))

data2022 <- all_data %>% dplyr::filter(year == 2022)
data2022$week <- as.integer(floor(difftime(data2022$week_start, as.Date("2022-01-01"), units = "weeks")) + 1)
range(data2022$week)

baseline_10yr <- all_data %>% filter(year %in% 2012:2021)  # 10-year: 2012–2021
baseline_15yr <- all_data %>% filter(year %in% 2007:2021)  # 15-year: 2007–2021
baseline_20yr <- all_data %>% filter(year %in% 2002:2021)  # 20-year: 2002–2021
baseline_25yr <- all_data %>% filter(year %in% 1997:2021)  # 25-year: 1997–2021


```

## Summarize climatology

```{r, summarize climatology}
vars <- c("leaf_area_index", "longwave_radiation", "runoff_sum", "shortwave_radiation",
          "snow_density", "soil_moisture", "relative_humidity", "temperature",
          "evapotranspiration", "precipitation", "wind_speed", "surface_pressure" )

baseline_10yr <- baseline_10yr %>%
  mutate(week = lubridate::isoweek(week_start)) %>%
  filter(week <= 52)

baseline_15yr <- baseline_15yr %>%
  mutate(week = lubridate::isoweek(week_start)) %>%
  filter(week <= 52)

baseline_20yr <- baseline_20yr %>%
  mutate(week = lubridate::isoweek(week_start)) %>%
  filter(week <= 52)

baseline_25yr <- baseline_25yr %>%
  mutate(week = lubridate::isoweek(week_start)) %>%
  filter(week <= 52)

# Function to summarize climatology
summarize_climatology <- function(df) {
  df %>%
    group_by(zone_id, week) %>%
    summarise(across(all_of(vars), list(mean = mean, sd = sd), .names = "{.col}_{.fn}"), .groups = "drop")
}

clim_10yr <- summarize_climatology(baseline_10yr)
clim_15yr <- summarize_climatology(baseline_15yr)
clim_20yr <- summarize_climatology(baseline_20yr)
clim_25yr <- summarize_climatology(baseline_25yr)
```

### Calculate z scores comparing 2022 to the 10yr, 15yr, 20yr, and 25yr means

```{r, join and calcualte z scores}

data2022_z <- data2022 %>%
  left_join(clim_10yr, by = c("zone_id", "week")) %>%
  left_join(clim_15yr, by = c("zone_id", "week"), suffix = c("_10yr", "_15yr")) %>%
  left_join(clim_20yr, by = c("zone_id", "week"), suffix = c("_15yr", "_20yr")) %>%
  left_join(clim_25yr, by = c("zone_id", "week"), suffix = c("_20yr", "_25yr"))

# Compute z-scores
for (v in vars) {
  for (yr in c(10, 15, 20, 25)) {
    m <- paste0(v, "_mean_", yr, "yr")
    s <- paste0(v, "_sd_", yr, "yr")
    z <- paste0(v, "_z_", yr, "yr")

    data2022_z[[z]] <- (data2022_z[[v]] - data2022_z[[m]]) / data2022_z[[s]]
  }
}



```

#### Look at distribution of z-scores

```{r, z-score distribution}

# Gather z-score columns into long format
z_cols <- grep("_z_(10yr|15yr|20yr|25yr)$", names(data2022_z), value = TRUE)

z_long <- data2022_z %>%
  select(all_of(z_cols)) %>%
  pivot_longer(cols = everything(), names_to = "variable", values_to = "z_score")

# Plot density by variable
ggplot(z_long, aes(x = z_score)) +
  geom_density(fill = "steelblue", alpha = 0.6) +
  facet_wrap(~ variable, scales = "free") +
  theme_minimal() +
  labs(title = "Distribution of Climate Z-Scores (2022 vs 10yr and 25yr)",
       x = "Z-score", y = "Density")

z_summary <- z_long %>%
  group_by(variable) %>%
  summarise(
    min = min(z_score, na.rm = TRUE),
    p01 = quantile(z_score, 0.01, na.rm = TRUE),
    p99 = quantile(z_score, 0.99, na.rm = TRUE),
    max = max(z_score, na.rm = TRUE),
    outlier_prop = mean(abs(z_score) > 4, na.rm = TRUE)
  )

print(z_summary)

```

#### There are outliers, so we opt to winsorize 

```{r, winsorize z-scores }

# Custom winsorization function with summary
winsorize_with_summary <- function(x, varname, limit = 4) {
  before_outliers <- sum(x < -limit | x > limit, na.rm = TRUE)
  x_winsorized <- pmin(pmax(x, -limit), limit)
  message(sprintf("%s: %d values winsorized (limit ±%s)", varname, before_outliers, limit))
  return(x_winsorized)
}

# Apply to all z-score columns with summary
for (v in z_cols) {
  data2022_z[[v]] <- winsorize_with_summary(data2022_z[[v]], v, limit = 4)
}


```

## Read in feedlot data (contains info on outbreaks)

```{r, add feedlot data}

fishnet_weeks <- data2022_z

# Define paths
first_path  <- file.path(data_folder, "feedlots_timeseries_10km_simple_first.shp")
second_path <- file.path(data_folder, "feedlots_timeseries_10km_simple_second.shp")

# Read shapefiles
feedlot_first  <- st_read(first_path)
feedlot_second <- st_read(second_path)

# Combine into one spatial object
feedlot_merged <- rbind(feedlot_first, feedlot_second)

# Standardize column names
feedlot_merged <- feedlot_merged %>%
  rename(
    outbreak_id     = outbreak_i,
    outbreak_dist   = Outbreak_D,
    outbreak_week   = Outbreak_W,
    outbreak_count  = Outbreak_C,
    close_outbreak  = Close_Outb,
    week            = Week
  ) %>%
  mutate(week = as.integer(week))

sum(feedlot_merged$outbreak_count)

```


## Check that outbreaks are independent at each feedlot/farm (meaning they >1 do not occur within 12 weeks)
```{r, check on distinct outbreaks}
#--- CHECKS
# Check feedlot identifiers in week 1
feedlot_merged %>%
  filter(week == 1) %>%
  summarise(
    n_rows = n(),
    n_unique_feedlot_id = n_distinct(feedlot_id),
    n_unique_name = n_distinct(name),
    n_unique_geometry = n_distinct(geometry)
  )

# Check if feedlot_id is unique per location
feedlot_merged %>%
  filter(week == 1) %>%
  st_drop_geometry() %>%
  group_by(feedlot_id) %>%
  summarise(n = n()) %>%
  filter(n > 1)

#-------

### Need to make sure outbreaks are independent (that is they do not happen in the same feedlot/farm within 12 weeks)

# ---- Filter to independent spillovers (12-week gap rule) ----
min_gap_weeks <- 12

# Get only the weeks where an outbreak occurred at each feedlot
feedlot_outbreaks <- feedlot_merged %>%
  st_drop_geometry() %>%
  filter(outbreak_count > 0) %>%
  arrange(feedlot_id, week) %>%
  group_by(feedlot_id) %>%
  mutate(
    detection_num = row_number(),
    weeks_since_last = week - lag(week),
    independent = detection_num == 1 | (!is.na(weeks_since_last) & weeks_since_last >= min_gap_weeks)
  ) %>%
  ungroup()

# Summary
cat("Total outbreak events:", sum(feedlot_outbreaks$outbreak_count), "\n")
cat("Unique feedlots with outbreaks:", n_distinct(feedlot_outbreaks$feedlot_id), "\n")
cat("Unique outbreak id:", n_distinct(feedlot_outbreaks$outbreak_id), "\n")
cat("Independent spillovers:", sum(feedlot_outbreaks$independent), "\n")
cat("Excluded (repeat <12 weeks):", sum(!feedlot_outbreaks$independent), "\n")

# Check feedlots with multiple outbreak weeks
feedlot_outbreaks %>%
  group_by(feedlot_id) %>%
  filter(n() > 1) %>%
  arrange(feedlot_id, week) %>%
  select(feedlot_id, zone_id, week, outbreak_count, detection_num, weeks_since_last, independent) %>%
  print(n = 20)

# The above confirms that all the outbreaks are independent 

# Which feedlot-weeks to exclude
excluded_feedlot_weeks <- feedlot_outbreaks %>%
  filter(!independent) %>%
  select(feedlot_id, zone_id, week)

cat("\nExcluded feedlot-weeks:\n")
print(excluded_feedlot_weeks, n = 20)


feedlot_outbreaks %>%
  group_by(outbreak_id) %>%
  filter(n() > 1) %>%
  arrange(outbreak_id) %>%
  select(outbreak_id, feedlot_id, zone_id, week, outbreak_count)

hpai_mn_2022 %>%
  group_by(outbreak_id) %>%
  filter(n() > 1) %>%
  arrange(outbreak_id, date_confirmed) %>%
  select(outbreak_id, premises, lat, lon, date_confirmed, type)
```

## Check with the raw outbreak data

```{r, check with the raw outbreak data}

# Load raw data 
hpai_outbreaks <- read_csv(file.path(data_folder, "hpai-outbreaks-clean-placeholder-032124.csv"))

# Filter to Minnesota 2022 
hpai_mn_2022 <- hpai_outbreaks %>%
  filter(
    state == "MN",
    date_confirmed >= as.Date("2022-01-01"),
    date_confirmed <= as.Date("2022-12-31")
  )

cat("Total MN outbreaks in 2022:", nrow(hpai_mn_2022), "\n")

# Check for premises with multiple detections
hpai_mn_2022 <- hpai_mn_2022 %>%
  mutate(week = as.numeric(floor(difftime(date_confirmed, as.Date("2022-01-01"), units = "weeks")) + 1))

# Are there repeat detections at same premises?
hpai_mn_2022 %>%
  group_by(lat, lon) %>%  # or use premises_id if available
  filter(n() > 1) %>%
  arrange(lat, lon, date_confirmed) %>%
  mutate(
    days_since_last = as.numeric(date_confirmed - lag(date_confirmed)),
    weeks_since_last = days_since_last / 7
  ) %>%
  select(lat, lon, date_confirmed, week, days_since_last, weeks_since_last) %>%
  print(n = 20)

## This confirms that there are repeat detection at same farm but they do not happen within 12-weeks of each other

```

# But wait one outbreak was not assigned to a feedlot
## Lets look what happened (must have been an issue with the join)
```{r, missing outbreak}
# Compare outbreak IDs between raw data and feedlot_merged
raw_ids <- hpai_mn_2022$outbreak_id
merged_ids <- feedlot_outbreaks %>%
  pull(outbreak_id) %>%
  str_split("_") %>%
  unlist() %>%
  as.numeric()

# Which outbreak ID is missing?
setdiff(raw_ids, merged_ids)

# Get the missing outbreak as sf
missing_sf <- hpai_mn_2022 %>%
  filter(outbreak_id == 309) %>%
  st_as_sf(coords = c("lon", "lat"), crs = 4326) %>%
  st_transform(st_crs(feedlot_merged))

# Buffer the missing outbreak by 20km and filter feedlots
missing_buffer <- st_buffer(missing_sf, dist = 20000)
feedlots_nearby <- feedlot_merged %>%
  st_filter(missing_buffer)

# Transform to WGS84 
feedlots_nearby_wgs84 <- feedlots_nearby %>%
  st_transform(4326)

# Get the missing outbreak
missing_outbreak <- hpai_mn_2022 %>%
  filter(outbreak_id == 309)

# Get all other outbreaks
other_outbreaks <- hpai_mn_2022 %>%
  filter(outbreak_id != 309)

# Transform case_zones to WGS84 for leaflet


# Split feedlots into those with and without outbreaks
feedlots_with_outbreak <- feedlots_nearby_wgs84 %>%
  filter(outbreak_count > 0)

feedlots_no_outbreak <- feedlots_nearby_wgs84 %>%
  filter(outbreak_count == 0 | is.na(outbreak_count))

# Create leaflet map
leaflet() %>%
  addProviderTiles("CartoDB.Positron") %>%
  addCircles(
    data = feedlots_no_outbreak,
    radius = 5,
    color = "orange",
    fillOpacity = 0.6,
    popup = ~paste("Outbreak Count:", outbreak_count),
    group = "Feedlots (No Outbreak)"
  ) %>%
  addCircles(
    data = feedlots_with_outbreak,
    radius = 10,
    color = "purple",
    fillOpacity = 0.9,
    popup = ~paste("Outbreak Count:", outbreak_count),
    group = "Feedlots (With Outbreak)"
  ) %>%
  addCircles(
    lng = other_outbreaks$lon,
    lat = other_outbreaks$lat,
    radius = 5,
    color = "green",
    fillOpacity = 0.6,
    label = paste("Outbreak ID:", other_outbreaks$outbreak_id),
    group = "Other Outbreaks"
  ) %>%
  addCircles(
    lng = missing_outbreak$lon,
    lat = missing_outbreak$lat,
    radius = 10,
    color = "red",
    fillOpacity = 0.8,
    popup = paste("MISSING - Outbreak ID:", missing_outbreak$outbreak_id, 
                  "<br>Date:", missing_outbreak$date_confirmed),
    group = "Missing Outbreak"
  ) %>%
  addLayersControl(
    overlayGroups = c("Case Zones", "Feedlots (No Outbreak)", "Feedlots (With Outbreak)", 
                      "Other Outbreaks", "Missing Outbreak"),
    options = layersControlOptions(collapsed = FALSE)
  ) %>%
  setView(
    lng = missing_outbreak$lon,
    lat = missing_outbreak$lat,
    zoom = 11
  )

## Lets find nearest feedlot 

# Get the missing outbreak as sf in the same CRS as feedlots
missing_sf <- hpai_mn_2022 %>%
  filter(outbreak_id == 309) %>%
  st_as_sf(coords = c("lon", "lat"), crs = 4326) %>%
  st_transform(st_crs(feedlot_merged))

# Find the nearest feedlot
nearest_idx <- st_nearest_feature(missing_sf, feedlot_merged)
nearest_feedlot <- feedlot_merged[nearest_idx, ]

# Check distance
distance_m <- st_distance(missing_sf, nearest_feedlot) %>% as.numeric()
cat("Distance to nearest feedlot:", round(distance_m), "meters\n")

# View the nearest feedlot details
nearest_feedlot %>%
  st_drop_geometry() %>%
  select(any_of(c("feedlot_id", "name", "zone_id", "Outbreak_C")))

colnames(feedlot_merged)
```

## Assing missing outbreak to nearest feedlot

```{r, assign missing outbreak}

# Get outbreak 309 details
outbreak_309 <- hpai_mn_2022 %>%
  filter(outbreak_id == 309) %>%
  select(outbreak_id, date_confirmed, week, lat, lon, premises, type)

print(outbreak_309)

# Get the nearest feedlot's feedlot_id
nearest_feedlot_id <- nearest_feedlot$feedlot_id[1]
nearest_zone_id <- nearest_feedlot$zone_id[1]

cat("Nearest feedlot_id:", nearest_feedlot_id, "\n")
cat("Zone_id:", nearest_zone_id, "\n")
cat("Outbreak 309 week:", outbreak_309$week, "\n")

# Update feedlot_merged to add outbreak 309 to the nearest feedlot for that week
feedlot_merged <- feedlot_merged %>%
  mutate(
    outbreak_id = case_when(
      feedlot_id == nearest_feedlot_id & week == outbreak_309$week ~ "309",
      TRUE ~ as.character(outbreak_id)
    ),
    outbreak_count = case_when(
      feedlot_id == nearest_feedlot_id & week == outbreak_309$week ~ 1,
      TRUE ~ as.numeric(outbreak_count)
    ),
    outbreak_week = case_when(
      feedlot_id == nearest_feedlot_id & week == outbreak_309$week ~ "35",
      TRUE ~ as.character(outbreak_week)
    ),
    outbreak_dist = case_when(
      feedlot_id == nearest_feedlot_id & week == outbreak_309$week ~ as.character(round(distance_m, 2)),
      TRUE ~ as.character(outbreak_dist)
    )
  )

# Verify the update
feedlot_merged %>%
  filter(feedlot_id == nearest_feedlot_id, week == outbreak_309$week) %>%
  st_drop_geometry() %>%
  select(feedlot_id, zone_id, week, outbreak_id, outbreak_count, outbreak_week)
```



## Clean feedlot data and rename 
```{r, add feedlot data}
# Get unique zones that contain feedlots
zones_with_feedlots <- feedlot_merged %>%
  st_drop_geometry() %>%
  distinct(zone_id)

cat("Zones with feedlots:", nrow(zones_with_feedlots), "\n")

# Aggregate feedlot data to zone-week level
feedlot_zone_week <- feedlot_merged %>%
  st_drop_geometry() %>%
  group_by(zone_id, week) %>%
  summarise(
    outbreak_binary = as.integer(any(outbreak_id == 1, na.rm = TRUE)),
    outbreak_dist   = min(outbreak_dist, na.rm = TRUE),
    outbreak_week   = min(outbreak_week, na.rm = TRUE),
    outbreak_count  = sum(outbreak_count, na.rm = TRUE),
    close_outbreak  = min(close_outbreak, na.rm = TRUE),
    .groups = "drop"
  )

# Filter fishnet_weeks to only zones with feedlots, then join
fishnet_weeks <- st_drop_geometry(fishnet_weeks)

fishnet_weeks <- fishnet_weeks %>%
  mutate(week = as.integer(week)) %>%
  inner_join(zones_with_feedlots, by = "zone_id") %>%  # Keep only zones with feedlots
  left_join(feedlot_zone_week, by = c("zone_id", "week"))

cat("Total zones in fishnet_weeks after filtering:", n_distinct(fishnet_weeks$zone_id), "\n")
cat("Total rows in fishnet_weeks:", nrow(fishnet_weeks), "\n")

# Continue with cleanup
fishnet_weeks <- fishnet_weeks %>%
  select(-any_of(c("outbreak_binary", "lat", "lon"))) %>%
  mutate(outbreak_binary = if_else(!is.na(outbreak_count) & outbreak_count > 0, 1L, 0L))

fishnet_timeseries <- fishnet_weeks

# Checks

# Check total outcomes and outcome_binary
cat("outbreaks:", sum(fishnet_timeseries$outbreak_binary), "\n")
cat("outbreak binary:", sum(fishnet_timeseries$outbreak_count), "\n")


# Check how many zones have cases vs total zones

fishnet_timeseries %>%
  group_by(zone_id) %>%
  summarise(has_case  = any(!is.na(outbreak_dist))) %>%
  count(has_case)

# Create indicator for zones that ever had a case
fishnet_timeseries <- fishnet_timeseries %>%
  group_by(zone_id) %>%
  mutate(any_outbreak = max(outbreak_binary, na.rm = TRUE)) %>%
  ungroup()

# Count case-weeks (zone-weeks in cells that had a case at any point)
case_weeks <- fishnet_timeseries %>%
  filter(any_outbreak == 1) %>%
  nrow()

# Count negative control-weeks (zone-weeks in cells that never had a case)
negative_control_weeks <- fishnet_timeseries %>%
  filter(any_outbreak == 0) %>%
  nrow()

# Summary
cat("Total zone-weeks:", nrow(fishnet_timeseries), "\n")
cat("Case-weeks (zones with any outbreak):", case_weeks, "\n")
cat("Negative control-weeks (zones never had outbreak):", negative_control_weeks, "\n")


```


# Read in bird data

```{r, read in full bird data}

# Load the data
csv_path <- file.path(data_folder, "all-mn-feedlot-species-abundance-070125.csv")
bird_abundance <- fread(csv_path)

# Quick check
bird_abundance[, .N]
head(bird_abundance)
str(bird_abundance)
unique(bird_abundance$species)



# Define species groups
anseriformes <- c(
  "mallar3", "cangoo", "wooduc", "rinduc", "comgol", "norsho",
  "buffle", "lessca", "rebmer", "gnwtea", "gadwal", "ambduc",
  "rudduc", "rethaw", "truswa", "commer"
)

predators <- c(
  "buwtea", "grhowl", "amwpel", "baleag", "brdowl"
)

# Melt from wide to long: assume all columns except first few are weeks
week_cols <- setdiff(names(bird_abundance), c("zone_id", "species"))
bird_long <- melt(
  bird_abundance,
  id.vars = c("zone_id", "species"),
  measure.vars = week_cols,
  variable.name = "week",
  value.name = "abundance"
)

# Add group labels
bird_long[, group := fifelse(species %in% anseriformes, "anseriformes", "predators")]
bird_long[, week := as.integer(gsub("week_", "", week))]

# Summarize: mean percent abundance per group per zone_id-week
bird_grouped <- bird_long[
  , .(
    group_sum = sum(abundance, na.rm = TRUE),
    species_n = uniqueN(species)
  ),
  by = .(zone_id, week, group)
]

bird_grouped[, group_avg := group_sum / species_n]

# Pivot to wide: one column per group
bird_wide <- dcast(
  bird_grouped,
  zone_id + week ~ group,
  value.var = "group_avg",
  fill = 0
)

# Compute total average abundance across all birds
bird_wide[, total_bird_abundance := anseriformes + predators]


# Convert to data.table (dropping geometry)
zone_week_grid <- unique(
  as.data.table(st_drop_geometry(fishnet_timeseries))[, .(zone_id, week)]
)
setkey(bird_wide, zone_id, week)
setkey(zone_week_grid, zone_id, week)

bird_wide_full <- merge(
  zone_week_grid,
  bird_wide,
  by = c("zone_id", "week"),
  all.x = TRUE
)

# Fill missing with 0
bird_wide_full[is.na(anseriformes), anseriformes := 0]
bird_wide_full[is.na(predators), predators := 0]
bird_wide_full[is.na(total_bird_abundance), total_bird_abundance := 0]

# Ensure both are data.tables
setDT(fishnet_timeseries)
setDT(bird_wide_full)

# Set keys
setkey(fishnet_timeseries, zone_id, week)
setkey(bird_wide_full, zone_id, week)

# Join: add bird abundance to fishnet
fishnet_with_birds <- merge(
  fishnet_timeseries,
  bird_wide_full,
  by = c("zone_id", "week"),
  all.x = TRUE
)

fishnet_with_birds <- fishnet_with_birds[!is.na(fishnet_with_birds$total_bird_abundance), ]

fishnet_timeseries <- fishnet_with_birds

# Summarize total bird abundance by zone_id
zone_zero_summary <- fishnet_with_birds[
  , .(total_abundance = sum(total_bird_abundance, na.rm = TRUE)), 
  by = zone_id
]

# Filter to zone_ids with all zero bird abundance
zone_ids_all_zero <- zone_zero_summary[total_abundance == 0, zone_id]

# Subset spatial layer to zones with all-zero birds
zones_all_zero_sf <- fishnet_timeseries[zone_id %in% zone_ids_all_zero]


# Tag all-zero zones in the fishnet data
fishnet_timeseries[, bird_zero := ifelse(zone_id %in% zone_ids_all_zero, "No Birds", "Has Birds")]

## Map if desired
# If not already an sf object, convert it
# fishnet_sf <- st_as_sf(fishnet_timeseries)
# 
# # Reproject to WGS 84 (for plotting or basemaps)
# fishnet_sf <- st_transform(fishnet_sf, 4326)
# 
# 
# ggplot(fishnet_sf) +
#   geom_sf(aes(fill = bird_zero), color = NA, alpha = 0.9) +
#   scale_fill_manual(
#     values = c("No Birds" = "red", "Has Birds" = "lightgray"),
#     name = "Bird Presence"
#   ) +
#   labs(
#     title = "Zones with No Bird Abundance (All Weeks = 0)"
#   ) +
#   theme_minimal()


```


# Create any extra variables

```{r, create extra variables}
fishnet_timeseries <- fishnet_timeseries %>%
  group_by(zone_id) %>%
  mutate(any_outbreak = as.integer(any(outbreak_binary > 0))) %>%
  ungroup()


fishnet_timeseries <- fishnet_timeseries %>%
  mutate(
    # Convert week index to month 
    month = lubridate::month(lubridate::ymd("2022-01-01") + lubridate::weeks(week - 1)),
    
    month_factor = as.factor(month),

    # Create warm season dummy: May (5) to September (9)
    warm_season = ifelse(month %in% 5:9, 1, 0),

    # Create spring dummy: March (3), April (4), May (5)
    spring = ifelse(month %in% 3:5, 1, 0),

    # Create spring/fall dummy: March-May or September-November
    spring_fall = ifelse(month %in% c(3:5, 9:11), 1, 0),
    

    # Create season factor
    season = case_when(
      month %in% c(12, 1, 2) ~ "Winter",
      month %in% 3:5         ~ "Spring",
      month %in% 6:8         ~ "Summer",
      month %in% 9:11        ~ "Fall"
    ),
    season = factor(season, levels = c("Winter", "Spring", "Summer", "Fall"))

  )



# Constants
p <- 1013.25  # pressure in hPa (standard sea-level pressure)

fishnet_timeseries <- fishnet_timeseries %>%
  mutate(
    # Convert temperature to Kelvin
    temp_k = temperature + 273.15,

  )

seasonal_outbreaks <- fishnet_timeseries %>%
  group_by(season) %>%
  summarise(outbreaks = sum(outbreak_count, na.rm = TRUE)) %>%
  arrange(match(season, c("Winter", "Spring", "Summer", "Fall")))

```

## Join in landcover

```{r, join landcover}

land_cover_path <- file.path(data_folder, "fishnet_landcover.gpkg")
land_cover <- st_read(land_cover_path)

land_cover_var <- c("zone_id", "land_cover_mode", "land_cover_label", "lat", "lon")

# Select only the needed columns from the land cover layer
land_cover_clean <- land_cover %>%
  select(all_of(land_cover_var))

# Perform the left join by zone_id
fishnet_timeseries <- fishnet_timeseries %>%
  left_join(land_cover_clean, by = "zone_id")

fishnet_timeseries <- fishnet_timeseries %>%
  arrange(zone_id, week)

unique(fishnet_timeseries$land_cover_label)
fishnet_timeseries$land_cover_label <- relevel(
  factor(fishnet_timeseries$land_cover_label),
  ref = "Developed, Medium Intensity"
)

fishnet_timeseries$farm_possible <- ifelse(
  fishnet_timeseries$land_cover_label %in% c(
    "Cultivated Crops",
    "Pasture/Hay",
    "Emergent Herbaceous Wetlands"
  ),
  1, 0
)
```

# Change any variable names 

```{r, change variable names}
 
fishnet_timeseries <- fishnet_timeseries %>%
  rename_with(~ gsub("runoff_sum", "runoff", .x), .cols = everything())

fishnet_weeks <- fishnet_weeks %>%
  rename_with(~ gsub("runoff_sum", "runoff", .x), .cols = everything())

```


## Export RDS

```{r, export rds}

saveRDS(fishnet_timeseries, file.path(objects_folder, "fishnet_timeseries.RDS"))
```