---
title: "HPAI spillover — manuscript exhibits"
subtitle: "Every figure and table in submission order, with a plain read of each"
format:
  html:
    toc: true
    toc-depth: 2
    toc-location: left
    embed-resources: true
    code-fold: true
    theme: flatly
execute:
  echo: false
  message: false
  warning: false
---

```{r setup}
#| include: false
knitr::opts_chunk$set(out.width = "100%", fig.align = "center")
project.folder <- paste0(here::here(), "/")
src <- readLines(paste0(project.folder, "create_folder_structure.R"))
eval(parse(text = paste(src[!grepl("^\\s*rm\\(list", src)], collapse = "\n")))
source(paste0(functions.folder, "script_initiate.R"))
source(paste0(functions.folder, "table_helpers.R"))
suppressMessages(library(knitr))
SPECS <- file.path(tables_folder, "table_specs")

# Nothing is refitted or rebuilt here. Figures come off disk and tables come from the display
# specs to_word() stashed when it wrote the Word files, so what is reviewed below is what gets
# submitted. If an exhibit is missing, re-knit the document that makes it, not this one.
missing <- character(0)
fig <- function(f) {
  p <- file.path(figures_main_folder, f)
  if (!file.exists(p)) { missing <<- c(missing, f)
    return(asis_output(sprintf("\n> **Not yet built:** `%s`\n", f))) }
  include_graphics(p)
}
tbl <- function(name, font_size = 12) {
  k <- show_spec(name, SPECS, font_size = font_size)
  if (is.null(k)) { missing <<- c(missing, paste0(name, ".RDS"))
    return(asis_output(sprintf("\n> **Not yet built:** `%s`\n", name))) }
  k
}
```

::: {.callout-note}
Panel: **175 spillover cases in 161 cells**, seven northern Mississippi Flyway states, 2022. Minnesota 85
cases in 73 cells. Two primary parameterisations everywhere: **absolute conditions** and the **90-day
shock** (departure from the preceding 90 days). Six exposures throughout: temperature, precipitation,
soil moisture, wind speed, runoff, Anseriformes. Rate ratios are per +0.5 SD; P(dir) is the posterior
probability of direction, reported continuously and never thresholded.

Source documents — figures: `d_02` → `d_15` (F1), `d_23` (F2), `d_17` (F3), `d_24` (F4), `d_08` (S1).
Tables: `d_18` (T1, T2), `d_20` (S1–S7b). Re-knit the source, then this.
:::

# Main text

## Figure 1. Study setting

```{r f1}
fig("figure2_study_setting.png")
```

- 161 case cells across the seven states; Minnesota's cluster sits in the west-central part of the state.
- Spillovers peak in weeks 13–18 (April–May) with a smaller wave in weeks 35–46 (September–November).
- Anseriformes abundance leads the spring wave by roughly two to three weeks in both panels. That lead is the
  lag structure the models keep finding.

## Figure 2. Cumulative rate ratios, both primary models

```{r f2}
fig("figure2_primary_forest.png")
```

- **Waterfowl rises with lag in every panel and both parameterisations.** Minnesota, 90-day shock, lag 22–28 d:
  2.42 (1.54, 3.78). Pooled: 1.47 (1.20, 1.80).
- **Soil moisture shock is protective in the pooled panel** — 0.81 (0.70, 0.94) at 0–7 d, 0.88 (0.80, 0.96) at
  8–14 d. Minnesota is the same direction but wider (0.84 (0.68, 1.03)), so this is a pooled result.
- Wind speed shock at 0–7 d, pooled, newly excludes 1 (1.18 (1.03, 1.35)). It did not before the panel grew;
  treat as provisional.
- Temperature, precipitation and runoff are null as main effects in every panel. Runoff is in for consistency
  with the interaction model, not for its estimate.
- The other-flyway waterfowl estimate attenuated to 1.21 (0.98, 1.48) when the eight early-season cases
  (8 Feb–11 Mar) were recovered by the 2021 warm-up: those cases precede the migration wave.

## Figure 3. Waterfowl cumulative curves, both lag windows

```{r f3}
fig("figure4_bird_cumulative_curves.png")
```

- Top row (0–28 d model): Minnesota crosses 1 only from about lag 24 — 2.32 (1.21, 4.42) at 24, 4.12
  (2.03, 8.37) at 28. Bottom row (0–56 d model): never crosses 1 from lag 7 onward (1.83 (1.39, 2.42)).
- The two windows agree on **that** waterfowl matters and disagree on **when** it starts, which is why both are
  shown rather than the tidier one.
- Predatory birds are flat throughout: the signal is waterfowl-specific.

## Figure 4. Meteorology in the wake of waterfowl

Three versions of the same estimates, so the choice can be made by eye. **Version 3 is the current
reference** (`figure4_interaction.png`); the other two are the earlier cuts.

### Version 3 — both windows plus the cumulative 0–14 d

```{r f4v3}
fig("figure4_interaction.png")
```

### Version 2 — the two pre-stated windows, 0–7 and 8–14 d

```{r f4v2}
fig("figure4_interaction_v2_two_windows.png")
```

### Version 1 — the 8–14 d window alone

```{r f4v1}
fig("figure4_interaction_v1_8-14only.png")
```

- **Cumulative 0–14 d, pooled, 90-day shock: temperature 2.09 (1.17, 3.68) at high waterfowl vs 0.74
  (0.42, 1.35) at low — ratio 2.77 (1.43, 5.45), P(dir) 0.999.** Other flyway states: ratio 7.95 (2.91, 22.43).
  This is the summary that matches the hypothesis as stated ("the two weeks prior") and cannot be
  window-picked.
- Decomposed: nothing separates at 0–7 d (pooled temperature ratio 1.12 (0.70, 1.79)); the split is in
  8–14 d (1.64 (1.12, 2.44) vs 0.66 (0.45, 0.98), ratio 2.48 (1.51, 3.99)).
- Minnesota precipitation shock is protective at high waterfowl — 0.53 (0.32, 0.87) over 0–14 d — and null at
  low (1.19).
- Runoff in Minnesota points the same way (1.58 (0.88, 2.81) at high waterfowl over 0–14 d) but does not clear 1.
- Under absolute conditions nothing clears 1 in any window: this is about *changes* in conditions, not levels.
- This explains the null main effects. Averaged over cell-days with and without prior waterfowl, the weather
  effect washes out.

## Table 1. Case-crossover sample

```{r t1}
tbl("Table1_case_crossover_sample")
```

- 175 cases in 161 cells. Minnesota 85/73, Wisconsin 29/29, Iowa 21/19, Michigan 16/16.
- Spring carries 56–81% of cases in every state; summer is 0–1 case per state.
- Repeat spillover within a cell is almost entirely Minnesota: 11 of its 73 case cells had two or more events,
  and the only three-event cell anywhere.
- About 23–25 referent cell-days per case under the one-month stratum with 7-day post-exclusion.

## Table 2. Minnesota conditions by season

```{r t2}
tbl("Table2_minnesota_by_season", font_size = 11)
```

- Spring (60 events in 52 cells): median temperature 4.0 °C, **not unusual** against the 25-year normal
  (−1.3 (−3.8, 1.3)); soil moisture 1.8 (0.1, 3.7) %vol **wetter** than normal; wind 0.4 (0.1, 0.7) m/s above.
- Winter case days were colder than normal (−2.4 (−3.7, −1.1) °C); fall case days warmer (1.3 (0.7, 1.9) °C).
- Anseriformes median 15.0 in spring versus 0.0 in winter and fall — the exposure contrast the paper rests on.
- Deviations are medians across cells with a credible interval from an AR(1) over days, so day-to-day
  autocorrelation is in the interval.

# Supplement

## Figure S1. Case and referent selection

```{r fs1}
fig("figure1_consort_flowchart.png")
```

- 176 eligible events → 175 case cell-days in 161 cells. The 2021 ERA5 warm-up removed the 90-day-baseline
  exclusion that previously cost eight cases.

## Table S1. Primary results, cumulative, lag-specific and P(direction)

```{r ts1}
tbl("TableS1_primary_results", font_size = 9)
```

- The numbers behind Figure 2, plus the per-day (lag-specific) contribution and P(dir) per window.
- Waterfowl P(dir) = 1.00 from lag 8–14 d onward in every panel.

## Table S2. Referent design and negative control

```{r ts2}
tbl("TableS2_referent_design", font_size = 10)
```

- The negative control (25-year normal temperature — cannot cause a 2022 outbreak) reads **2.99 (2.26, 3.95)**
  under the unidirectional design the original submission used, 1.24 (1.08, 1.43) with two-month strata, and
  1.13 (0.90, 1.42) under the selected one-month + 7-day design.
- Temperature tracks it: 2.60 (1.79, 3.76) unidirectional → 1.13 (0.89, 1.44) selected. What looked like a
  temperature effect was season.
- Justifies the time-stratified design. A design check, not a headline.

## Table S3. Lag structure

```{r ts3}
tbl("TableS3_lag_structure", font_size = 10)
```

- ΔWAIC sits on the specification row. The selected 0–28 d, ns(2) basis is the reference; every exposure is shown
  at all four windows so nothing is picked.

## Table S4. Prior on the crossbasis coefficients

```{r ts4}
tbl("TableS4_prior", font_size = 10)
```

- The prior is chosen on the implied rate-ratio range for a +0.5 SD contrast, not on WAIC — WAIC always prefers
  a tighter prior because shrinkage reduces effective complexity.

## Table S5. Spring versus outside spring

```{r ts5}
tbl("TableS5_season_interaction", font_size = 10)
```

- Waterfowl holds in both seasons; the meteorological terms do not separate by season. Residual seasonal
  confounding is not what the primary results are made of.

## Table S6. Leave-one-stratum-out cross-validation

```{r ts6}
tbl("TableS6_crossvalidation", font_size = 11)
```

- Minnesota, absolute conditions: +3.7 log-likelihood over an even spread across the stratum's days; the 90-day
  shock is below chance (−13.0, z = −2.04). Other panels sit below chance for both.
- Read as: these models identify associations; they do not predict *which day*. A doubling on one term moves
  little probability across ~23 candidate days.

## Table S7a. The interaction, by panel

```{r ts7a}
tbl("TableS7a_interaction", font_size = 9)
```

- The numbers behind Figure 4, both windows, with P(dir).
- One pre-stated specification: strata lag basis (independent 0–7 and 8–14 d bins), continuous modifier, primary
  prior, posterior sampling. Post hoc; say so in one clause.

## Table S7b. The interaction under alternative specifications

```{r ts7b}
tbl("TableS7b_interaction_robust", font_size = 9)
```

- The pooled temperature × waterfowl ratio at 8–14 d holds under all three lag bases and all three modifier
  definitions. This table is what makes the post hoc disclosure a clause rather than a paragraph.

# Build status

```{r status}
#| results: asis
if (length(missing)) cat("**Missing exhibits** — re-knit the document that produces each:\n\n",
                         paste0("- `", missing, "`", collapse = "\n"), "\n") else
  cat("**All exhibits present.** 4 main figures, 2 main tables, 1 supplement figure, 8 supplement tables.\n")
cat("\nBuilt ", format(Sys.time(), "%Y-%m-%d %H:%M"), ".\n", sep = "")
```
