---
title: "HPAI spillover — manuscript exhibits"
subtitle: "Every figure and table in submission order, with a plain read of each"
format:
  html:
    toc: true
    toc-depth: 2
    toc-location: left
    embed-resources: true
    code-fold: true
    theme: flatly
execute:
  echo: false
  message: false
  warning: false
---

```{r setup}
#| include: false
knitr::opts_chunk$set(out.width = "100%", fig.align = "center")
project.folder <- paste0(here::here(), "/")
src <- readLines(paste0(project.folder, "create_folder_structure.R"))
eval(parse(text = paste(src[!grepl("^\\s*rm\\(list", src)], collapse = "\n")))
source(paste0(functions.folder, "script_initiate.R"))
source(paste0(functions.folder, "table_helpers.R"))
suppressMessages(library(knitr))
SPECS <- file.path(tables_folder, "table_specs")

# Nothing is refitted or rebuilt here. Figures come off disk and tables come from the display
# specs to_word() stashed when it wrote the Word files, so what is reviewed below is what gets
# submitted. If an exhibit is missing, re-knit the document that makes it, not this one.
missing <- character(0)
fig <- function(f, w = "100%") {
  p <- file.path(figures_main_folder, f)
  if (!file.exists(p)) { missing <<- c(missing, f)
    return(asis_output(sprintf("\n> **Not yet built:** `%s`\n", f))) }
  include_graphics(p)
}
tbl <- function(name, font_size = 12) {
  k <- show_spec(name, SPECS, font_size = font_size)
  if (is.null(k)) { missing <<- c(missing, paste0(name, ".RDS"))
    return(asis_output(sprintf("\n> **Not yet built:** `%s`\n", name))) }
  k
}
```

::: {.callout-note}
**Primary analysis: Minnesota, 90-day shocks.** 85 spillover cases in 73 cells, 2022; time-stratified
case-crossover (one-month strata, 7-day post-event exclusion, ~23 referent cell-days per case); conditional
Poisson in INLA. Six exposures throughout: temperature, precipitation, soil moisture, wind speed, runoff,
Anseriformes, lag 0–28 d, ns(2) exposure–response and lag bases, N(0, 0.25²) priors. Rate ratios are per
+0.5 SD; P(dir) is the posterior probability of direction, reported continuously and never thresholded.

**Secondary, in the supplement:** absolute conditions as the parameterisation; the other six northern
Mississippi Flyway states (90 cases) and all seven states together (175 cases — one model on all strata, not a
pooling of estimates).

Source documents — figures: `d_02` → `d_15` (F1), `d_23` (F2, S2), `d_17` (F3, S4), `d_24` (F4, S3),
`d_08` (S1). Tables: `d_18` (T1, T2), `d_20` (S1–S7b). Re-knit the source, then this.
:::

# Main text

## Figure 1. Study setting

```{r f1}
fig("figure2_study_setting.png")
```

- 161 case cells across the seven states; Minnesota's cluster sits in the west-central part of the state.
- Spillovers peak in weeks 13–18 (April–May) with a smaller wave in weeks 35–46 (September–November).
- Anseriformes abundance leads the spring wave by roughly two to three weeks. That lead is the lag structure the
  models keep finding.

## Figure 2. Cumulative rate ratios — Minnesota, 90-day shocks

```{r f2}
fig("figure2_primary_forest.png")
```

- **Waterfowl rises with lag:** 1.25 (1.05, 1.48) at 8–14 d, 1.74 (1.37, 2.20) at 15–21 d, **2.42 (1.54, 3.78)
  at 22–28 d**. P(dir) = 1.00 from 8–14 d on.
- Every meteorological shock is null as a main effect in Minnesota. Soil moisture is the closest: 0.84
  (0.68, 1.03) at 0–7 d, 0.88 (0.77, 1.01) at 8–14 d — same direction as the all-states result (Figure S2),
  wider.
- Temperature 1.32 (0.88, 1.99) at 0–7 d, precipitation 0.90 (0.73, 1.10), wind 1.05 (0.87, 1.27), runoff 1.06
  (0.82, 1.37). Runoff is in for consistency with the interaction model, not for its estimate.

## Figure 3. Waterfowl lag–response — Minnesota, 90-day shock model

```{r f3}
fig("figure3_bird_lag_response.png")
```

- The **per-day** contribution of Anseriformes abundance, lag 0–28 d. It sits below 1 in the first week, crosses
  1 around lag 5 and climbs steadily to about 1.13 per +0.5 SD by lag 28, with the interval clear of 1 from
  roughly lag 8. The near-straight line is the ns(2) lag basis the primary model uses.
- This replaces the cumulative-from-lag-0 curve. A cumulative curve's starting point depends on where the fitted
  lag range starts, so the 0–28 and 0–56 d fits looked as if they disagreed about *when* the effect began; that
  was an artefact of the summary, not a finding. Figure S4 shows the same curve for all panels and the 0–56 d fit.
- Predatory birds sit on 1 throughout: the signal is waterfowl-specific.

## Figure 4. Meteorology in the wake of waterfowl — Minnesota, 90-day shocks

```{r f4}
fig("figure4_interaction.png")
```

- Post hoc interaction: the meteorological shock over lag **0–14 d** (weeks 1–2), modified by Anseriformes
  abundance over lag **15–28 d** (weeks 3–4), continuous, non-overlapping windows. Effects at +1 SD (high) and
  −1 SD (low) of the modifier. Lag windows across the top, as in Figure 2.
- **Precipitation shock is protective only where waterfowl were abundant:** cumulative 0–14 d, 0.48 (0.29, 0.79)
  at high waterfowl vs 1.33 (0.80, 2.15) at low; ratio 0.37 (0.16, 0.81), P(dir) 0.99. The split sits in the
  8–14 d bin; nothing separates at 0–7 d.
- **Runoff shock points the other way, suggestively:** 1.70 (0.97, 3.04) at high waterfowl vs 0.91 (0.52, 1.60);
  ratio 1.86 (0.81, 4.42), P(dir) 0.93.
- **Opposite signs on two water terms are not the pair splitting shared variance.** The precipitation and runoff
  shocks are only weakly correlated within stratum (r = 0.12 — in spring, runoff is snowmelt, not rain), and
  each result is unchanged when the other term is dropped: precipitation 0.49 (0.30, 0.78) without runoff,
  runoff 1.58 (0.89, 2.78) without precipitation (Table S7b).
- The reading that fits both: after waterfowl have arrived, **rain washes and dilutes** surface virus and keeps
  farm traffic down, while **snowmelt runoff moves** virus banked in frozen wetlands toward water near farms —
  the same "thaw" that shows up as the temperature interaction in the other states (Figure S3).
- **The effect needs the second week.** Restricting the weather term to lag 0–7 d (with birds 7–28 d) returns
  nothing — precipitation 0.99 (0.76, 1.27) — so the fortnight matters, not the first week alone.
- Soil moisture 0.68 (0.43, 1.07) at high vs 1.02 at low (P(dir) 0.90); temperature and wind do not separate.
  The Minnesota story is water, not temperature.

## Table 1. Case-crossover sample

```{r t1}
tbl("Table1_case_crossover_sample")
```

- 175 cases in 161 cells. Minnesota 85/73, Wisconsin 29/29, Iowa 21/19, Michigan 16/16.
- Spring carries 56–81% of cases in every state; summer is 0–1 case per state.
- Repeat spillover within a cell is almost entirely Minnesota: 11 of its 73 case cells had two or more events,
  and the only three-event cell anywhere.
- About 23–25 referent cell-days per case under the one-month stratum with 7-day post-exclusion.

## Table 2. Minnesota conditions by season

```{r t2}
tbl("Table2_minnesota_by_season", font_size = 11)
```

- Spring (60 events in 52 cells): median temperature 4.0 °C, **not unusual** against the 25-year normal
  (−1.3 (−3.8, 1.3)); soil moisture 1.8 (0.1, 3.7) %vol **wetter** than normal; wind 0.4 (0.1, 0.7) m/s above.
- Winter case days were colder than normal (−2.4 (−3.7, −1.1) °C); fall case days warmer (1.3 (0.7, 1.9) °C).
- Anseriformes median 15.0 in spring versus 0.0 in winter and fall — the exposure contrast the paper rests on.
- Deviations are medians across cells with a credible interval from an AR(1) over days, so day-to-day
  autocorrelation is in the interval.

# Supplement

## Figure S1. Case and referent selection

```{r fs1}
fig("figure1_consort_flowchart.png")
```

- 176 eligible events → 175 case cell-days in 161 cells. The 2021 ERA5 warm-up removed the 90-day-baseline
  exclusion that previously cost eight cases.

## Figure S2. Cumulative rate ratios — both parameterisations, all panels

```{r fs2}
fig("figureS2_primary_forest_all_panels.png")
```

- Waterfowl rises with lag in every panel and both parameterisations. All seven states, 90-day shock, lag
  22–28 d: 1.47 (1.20, 1.80); other flyway states 1.21 (0.98, 1.48).
- **Soil moisture shock is protective across all seven states** — 0.81 (0.70, 0.94) at 0–7 d, 0.88 (0.80, 0.96)
  at 8–14 d — where Minnesota alone has the same direction with a wider interval.
- Wind speed shock at 0–7 d, all states, 1.18 (1.03, 1.35). It did not exclude 1 before the panel grew; provisional.
- The other-flyway waterfowl estimate attenuated when the eight early-season cases (8 Feb–11 Mar) were recovered
  by the 2021 warm-up: those cases precede the migration wave.
- Under absolute conditions the meteorological terms are null everywhere; waterfowl holds.

## Figure S3. The interaction — both parameterisations, all panels

```{r fs3}
fig("figureS3_interaction_all_panels.png")
```

- **All seven states, 90-day shock, cumulative 0–14 d: a warm temperature shock raises spillover only where
  waterfowl were abundant** — 2.36 (1.33, 4.19) at high vs 0.57 (0.31, 1.03) at low; ratio 4.12 (2.05, 8.23),
  P(dir) 1.00. The other flyway states carry that result; Minnesota does not (Figure 4).
- Minnesota's precipitation interaction does not replicate in the other states (1.00 vs 0.89 at 8–14 d).
- Under absolute conditions nothing clears 1 in any panel or window: the interaction is about *changes* in
  conditions, not levels.

## Figure S4. Waterfowl lag–response — all panels, both lag ranges

```{r fs4}
fig("figureS4_bird_lag_response_all_panels.png")
```

- The per-day curve rises through lag 28 in every panel under the 0–28 d fit. The 0–56 d fit shows whether it
  keeps rising: it does not turn over in Minnesota.
- Predatory birds are flat in every panel.

## Table S1. Primary results, cumulative, lag-specific and P(direction)

```{r ts1}
tbl("TableS1_primary_results", font_size = 9)
```

- The numbers behind Figures 2 and S2 — both parameterisations, three panels, plus the per-day (lag-specific)
  contribution and P(dir) per window. 90-day shock first, as the primary.

## Table S2. Referent design and negative control

```{r ts2}
tbl("TableS2_referent_design", font_size = 9)
```

- The negative control (25-year normal temperature — cannot cause a 2022 outbreak) reads **2.82 (1.91, 4.15)**
  in Minnesota under the unidirectional design of the original submission, and **0.86 (0.63, 1.17)** under the
  selected one-month + 7-day design. All states: 2.99 → 1.13.
- Temperature tracks it: Minnesota 90-day shock 2.31 (1.52, 3.50) unidirectional → 1.18 (0.89, 1.56) selected.
  What looked like a temperature effect was season.
- A design check that justifies the time-stratified design; not a headline.

## Table S3. Lag structure (Minnesota)

```{r ts3}
tbl("TableS3_lag_structure", font_size = 10)
```

- ΔWAIC sits on the specification row. The selected 0–28 d, ns(2) basis is the reference; every exposure is shown
  at all four windows so nothing is picked.

## Table S4. Prior on the crossbasis coefficients (Minnesota)

```{r ts4}
tbl("TableS4_prior", font_size = 10)
```

- The prior is chosen on the implied rate-ratio range for a +0.5 SD contrast, not on WAIC — WAIC always prefers
  a tighter prior because shrinkage reduces effective complexity.

## Table S5. Spring versus outside spring (Minnesota)

```{r ts5}
tbl("TableS5_season_interaction", font_size = 10)
```

- Waterfowl holds in spring — 1.63 (1.23, 2.15) at 15–21 d, 2.24 (1.32, 3.80) at 22–28 d — and the
  meteorological terms do not separate by season. Residual seasonal confounding is not what the primary results
  are made of.

## Table S6. Leave-one-stratum-out cross-validation

```{r ts6}
tbl("TableS6_crossvalidation", font_size = 11)
```

- Minnesota, absolute conditions: +3.7 log-likelihood over an even spread across the stratum's days; the 90-day
  shock is below chance (−13.0, z = −2.04). Other panels sit below chance for both.
- Read as: these models identify associations; they do not predict *which day*. A doubling on one term moves
  little probability across ~23 candidate days. Worth stating plainly given the shock is the primary.

## Table S7a. The interaction, by panel

```{r ts7a}
tbl("TableS7a_interaction", font_size = 9)
```

- The numbers behind Figures 4 and S3, all three windows, with P(dir).
- One pre-stated specification: strata lag basis (independent 0–7 and 8–14 d bins), continuous modifier over
  lag 15–28 d, primary prior, posterior sampling. Post hoc; say so in one clause.

## Table S7b. The interaction under alternative specifications

```{r ts7b}
tbl("TableS7b_interaction_robust", font_size = 9)
```

- Minnesota's precipitation interaction and the all-states temperature interaction hold under all three lag
  bases, all three modifier definitions, and all four bird windows (21–28, 15–28, 14–28, 8–28 d), and each
  survives dropping the other water term.
- The one specification that returns nothing is weather 0–7 d with birds 7–28 d: the second week of weather
  carries the effect. This table is what makes the post hoc disclosure a clause rather than a paragraph.

# Build status

```{r status}
#| results: asis
if (length(missing)) cat("**Missing exhibits** — re-knit the document that produces each:\n\n",
                         paste0("- `", missing, "`", collapse = "\n"), "\n") else
  cat("**All exhibits present.** 4 main figures, 2 main tables, 4 supplement figures, 8 supplement tables.\n")
cat("\nBuilt ", format(Sys.time(), "%Y-%m-%d %H:%M"), ".\n", sep = "")
```
