---
title: "Table 1 — spillover cell characteristics (8-state grid)"
author: "Fintan Mooney"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.align = "center", dev = "ragg_png", dpi = 150)
```

```{r initiate}
rm(list = ls())
project.folder = paste0(print(here::here()), '/')
source(paste0(project.folder, 'create_folder_structure.R'))
source(paste0(functions.folder, 'script_initiate.R'))
source(paste0(functions.folder, 'inla_dlnm_helpers.R'))
```

Supersedes `d_00_table_1.Rmd`, which was built on the legacy MN fishnet (numeric `zone_id`,
3,334 cells). Everything now sits on the **8-state grid** (character `zone_id` like `-4,227`,
2,306 MN cells), so Table 1 describes the same cells the models actually fit.

Two things worth flagging up front:

* ERA5 ships `temperature` in **Kelvin**. Reporting it raw would print a mean of 279, so it
  gets converted here. `temperature_C` also exists in the refreshed exports.
* Only cells that had a spillover appear in Table 1. Non-spillover feedlot cells were in the
  old version, but they never enter the case-crossover risk set, so describing them invited a
  comparison the design does not make.

# 1. Spillover cells and their premises

```{r cells}
panel <- setDT(readRDS(paste0(objects_folder, multistate_ts_daily_rds)))[state == "Minnesota"]
scr   <- readRDS(paste0(objects_folder, multistate_independence_screening_rds))

# same case definition as the models: drop cross-farm CSLT/NOSEQ events
count_flagged <- function(s, ids) {
  if (is.na(s)) return(0L)
  as.integer(sum(as.numeric(trimws(unlist(strsplit(s, ",")))) %in% ids))
}
panel[, nf := vapply(outbreak_id, count_flagged, integer(1), ids = scr$cross_farm_ids)]
panel[, outbreak_count := pmax(outbreak_count - nf, 0L)][, nf := NULL]
panel[, outbreak_binary := as.integer(outbreak_count > 0)]

case_cells <- unique(panel[outbreak_binary == 1]$zone_id)
cells <- unique(panel[zone_id %in% case_cells,
                      .(zone_id, feedlot_count, au_count, birds_au, cafo_flag, land_cover_label)])

n_events <- sum(panel$outbreak_binary)
cat(sprintf("%d spillover events across %d cells\n", n_events, length(case_cells)))

num_row <- function(x, lab, dp = 0) data.table::data.table(
  Characteristic = lab,
  Value = sprintf(paste0("%.", dp, "f (%.", dp, "f–%.", dp, "f)"),
                  median(x, na.rm = TRUE), quantile(x, .25, na.rm = TRUE),
                  quantile(x, .75, na.rm = TRUE)))

tab_premises <- rbindlist(list(
  data.table(Characteristic = "Spillover events", Value = as.character(n_events)),
  data.table(Characteristic = "Cells with ≥1 event", Value = as.character(length(case_cells))),
  num_row(cells$feedlot_count, "Registered feedlots per cell"),
  num_row(cells$au_count,      "Animal units per cell"),
  num_row(cells$birds_au,      "Birds per cell"),
  data.table(Characteristic = "Cells with a CAFO",
             Value = sprintf("%d (%.0f%%)", sum(cells$cafo_flag, na.rm = TRUE),
                             100 * mean(cells$cafo_flag, na.rm = TRUE)))
))
print(tab_premises)
```

# 2. Land cover

```{r landcover}
# NLCD was only ever extracted onto the legacy MN fishnet, so the 8-state panel carries no
# land cover. The lookup carries it across by areal overlap between the two 10 km grids
# (see the landcover transfer script); 2,283 of 2,306 cells get a label.
lcm <- setDT(readRDS(paste0(objects_folder, "mn8_land_cover_lookup.RDS")))
cells <- merge(cells[, -"land_cover_label"], lcm[, .(zone_id, land_cover_label)],
               by = "zone_id", all.x = TRUE)
lc <- cells[!is.na(land_cover_label), .N, by = land_cover_label][order(-N)]
tab_lc <- if (nrow(lc)) {
  lc[, Value := sprintf("%d (%.0f%%)", N, 100 * N / sum(N))]
  lc[, .(Characteristic = land_cover_label, Value)]
} else {
  data.table::data.table(Characteristic = "Not available on the 8-state grid (NLCD pending)",
                         Value = "—")
}
print(tab_lc)
```

# 3. Events by meteorological season

```{r season-events}
# meteorological seasons: Dec-Feb, Mar-May, Jun-Aug, Sep-Nov (a_02 defines `season` this way)
ev <- panel[outbreak_binary == 1, .N, by = season][order(-N)]
ev[, Value := sprintf("%d (%.0f%%)", N, 100 * N / sum(N))]
tab_season <- ev[, .(Characteristic = as.character(season), Value)]
print(tab_season)
```

# 4. Meteorology by season, 2022 vs the 25-year normal

```{r met-season}
anom <- setDT(readRDS(paste0(objects_folder, "climate_anomaly_panel.RDS")))
anom <- anom[zone_id %in% case_cells]

# Kelvin -> degC for anything that is a temperature
MET <- c(temperature = "Temperature (°C)", runoff = "Surface runoff (mm)",
         soil_moisture = "Soil moisture (m³/m³)", precipitation = "Precipitation (mm)",
         wind_speed = "Wind speed (m/s)", snow_cover = "Snow cover (%)")
to_C   <- function(v, x) if (v == "temperature") x - 273.15 else x
to_mm  <- function(v, x) if (v %in% c("runoff", "precipitation")) x * 1000 else x

seas <- if ("season" %in% names(anom)) anom$season else
  cut(as.integer(strftime(anom$week_start, "%m")), c(0,2,5,8,11,12),
      labels = c("Winter","Spring","Summer","Fall","Winter"))
anom[, seas := seas]

rows <- rbindlist(lapply(names(MET), function(v) {
  mu <- paste0(v, "_mean_25yr"); zc <- paste0(v, "_z_25yr")
  if (!all(c(v, mu, zc) %in% names(anom))) return(NULL)
  anom[, .(Characteristic = MET[[v]],
           Season   = as.character(seas),
           Observed = to_mm(v, to_C(v, get(v))),
           Normal   = to_mm(v, to_C(v, get(mu))),
           Anomaly  = get(zc))][
    , .(Value = { o <- mean(Observed, na.rm = TRUE); n <- mean(Normal, na.rm = TRUE)
                   a <- mean(Anomaly, na.rm = TRUE)
                   # snow in summer is 0 everywhere, so its climatological sd is 0 and the
                   # z-score is undefined; report the levels and drop the anomaly
                   if (is.finite(a)) sprintf("%.1f vs %.1f (%+.2f SD)", o, n, a)
                   else              sprintf("%.1f vs %.1f (—)", o, n) }),
      by = .(Characteristic, Season)]
}))
tab_met <- dcast(rows, Characteristic ~ Season, value.var = "Value")
setcolorder(tab_met, c("Characteristic", intersect(c("Winter","Spring","Summer","Fall"), names(tab_met))))
print(tab_met)
```

# 5. Assemble and export

```{r export}
sec <- function(t, lab) rbindlist(list(data.table(Characteristic = lab, Value = ""), t), fill = TRUE)
tab1 <- rbindlist(list(sec(tab_premises, "Premises"), sec(tab_lc, "Land cover"),
                       sec(tab_season, "Events by season")), fill = TRUE)

ft <- flextable(tab1) |>
  bold(i = which(tab1$Value == ""), j = 1) |>
  autofit() |> theme_booktabs() |> align(j = 2, align = "right", part = "all")
ft_met <- flextable(tab_met) |> autofit() |> theme_booktabs() |>
  align(j = 2:ncol(tab_met), align = "center", part = "all")

# also stash the pieces so the exhibit register can render them inline
saveRDS(list(cells = tab1, meteorology = tab_met),
        paste0(objects_folder, "table1_parts.RDS"))

# bare tables, no caption or note; those go in the manuscript text
read_docx() |> body_add_flextable(ft) |>
  print(target = file.path(tables_main_folder, "Table1_spillover_cells.docx"))
read_docx() |> body_add_flextable(ft_met) |>
  print(target = file.path(tables_main_folder, "Table1_meteorology_by_season.docx"))
cat("wrote Table1_spillover_cells.docx and Table1_meteorology_by_season.docx\n")
```
