---
title: "Supporting Information tables"
author: "Fintan Mooney"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

```{r initiate}
rm(list = ls())
project.folder = paste0(print(here::here()), '/')
source(paste0(project.folder, 'create_folder_structure.R'))
source(paste0(functions.folder, 'script_initiate.R'))
source(paste0(functions.folder, 'inla_dlnm_helpers.R'))
```

Assembles the SI tables from the objects `c_11` saves. Nothing is refitted here, so this is
cheap to re-run after a spec change; it just fails loudly if `c_11` hasn't been run.

Each comparison is stacked one block per panel (rows = exposure, columns = the four weekly
windows) rather than laid out side by side. Side-by-side puts 12+ columns on a portrait page
and the numbers stop being readable.

```{r load}
need <- c(S2 = "si_referent_design.RDS", S3 = "si_case_definition.RDS",
          S4 = "si_weekly_model.RDS",     S5 = "si_specification.RDS",
          S7 = "si_feedlot_runoff.RDS")
paths <- paste0(objects_folder, need)
miss  <- need[!file.exists(paths)]
if (length(miss)) stop("run c_11 first, missing: ", paste(miss, collapse = ", "))
si <- lapply(setNames(paths, names(need)), function(p) setDT(readRDS(p)))
# older c_11 runs predate the flag; treat those as "everything identified"
if (!"shape_unidentified" %in% names(si$S2)) si$S2[, shape_unidentified := FALSE]

WIN <- names(INLA_LAG_WINDOWS)
ORD <- c("Surface Runoff","Soil Moisture","Temperature","Precipitation",
         "Wind Speed","Snow Cover","Anseriformes")

# the panel name rides in the table's own spanning header rather than a caption above it, so
# each block stays self-identifying with no surrounding text
blk_ft <- function(d, span) {
  d <- d[match(ORD, Predictor)][!is.na(Predictor)]
  flextable(d) |>
    add_header_row(values = c("", span), colwidths = c(1, ncol(d) - 1)) |>
    autofit() |> theme_booktabs() |>
    align(j = 2:ncol(d), align = "center", part = "all") |>
    fontsize(size = 8, part = "all")
}
stacked <- function(d, by, doc, lab = identity) {
  for (lev in unique(d[[by]])) {
    doc <- doc |>
      body_add_flextable(blk_ft(d[get(by) == lev, c("Predictor", ..WIN)], lab(lev))) |>
      body_add_par("", style = "Normal")
  }
  doc
}
```

# Table S2 — referent design

```{r s2}
print(dcast(si$S2, Predictor ~ design, value.var = "Lag 22-28 d"))
```

# Table S3 — case definition and spatial restriction

```{r s3}
print(unique(si$S3[, .(defn, n)]))
```

# Table S5 — model specification

```{r s5}
print(si$S5)
```

# Export

```{r export}
out <- function(doc, file) {
  print(doc, target = file.path(tables_sensitivity_folder, file)); cat("wrote", file, "\n")
}
out(stacked(si$S2, "design", read_docx()), "TableS2_referent_design.docx")

# carry n in the block header so the table doesn't need a line of prose under it
n_by <- unique(si$S3[, .(defn, n)])
out(stacked(si$S3, "defn", read_docx(),
            lab = function(x) sprintf("%s (n = %d)", x, n_by$n[match(x, n_by$defn)])),
    "TableS3_case_definition.docx")

# S4: weekly-resolution model, same layout as Table 2
out(read_docx() |> body_add_flextable(
      flextable(si$S4) |> autofit() |> theme_booktabs() |>
        align(j = 2:ncol(si$S4), align = "center", part = "all") |>
        fontsize(size = 8, part = "all")),
    "TableS4_weekly_model.docx")

# S7: runoff stratified by feedlot density, one block per stratum
n7 <- unique(si$S7[, .(Stratum, n)])
out(stacked(si$S7, "Stratum", read_docx(),
            lab = function(x) sprintf("%s (n = %d)", x, n7$n[match(x, n7$Stratum)])),
    "TableS7_feedlot_runoff.docx")

# merge the repeated Domain labels so S5 reads as grouped blocks
out(read_docx() |> body_add_flextable(
      flextable(si$S5) |> autofit() |> theme_booktabs() |>
        merge_v(j = "Domain") |> valign(j = "Domain", valign = "top") |>
        align(j = 2:ncol(si$S5), align = "center", part = "all") |>
        fontsize(size = 8, part = "all")),
    "TableS5_specification.docx")
```
