---
title: "hpai-mn-outbreak-mixture-analysis"
author: "Rishi Kowalski, ScM"
date: "2025-12-08"
---

-> Code for HPAI Outbreaks in MN analysis, bird mixture analysis.
-> 05/13/2025, rishi.kowalski@mssm.edu

-> Approach:
  - Assess cumulative impact of various bird mixtures on outbreak case/control cells at various lags via WQS
  - Geography: case and control cells 
  

```{r setup}
library(terra)
library(ebirdst)
library(dplyr)
library(sf)
library(gWQS)
library(lwqs)
library(tidyr)
library(ggplot2)
library(geosphere)
library(RColorBrewer)
library(cowplot)
library(parallel)
library(MMWRweek)
```



```{r read-in-bird-raster}
# Bird Top 15 Anseriforms in MN
# Initialize list of rasters - 15 anseriformes and 10 non-anseriform HPAI detections
# Top 15 Anseriforms identified in MN via iNaturalist List: https://www.inaturalist.org/guides/14807?taxon=6888

mn_anseri_15 <- c("mallar3", "cangoo", "wooduc", "buwtea", "commer", "rinduc", "comgol", "norsho", "buffle", "lessca", "rebmer", "gnwtea", "gadwal", "ambduc", "rudduc")


##### Abundance #####
#Anseriforms, n = 15
mallar3_abd_3km <- load_raster("mallar3" , product = "abundance", resolution = "3km")
cangoo_abd_3km <- load_raster("cangoo", product = "abundance", resolution = "3km")
wooduc_abd_3km <- load_raster("wooduc", product = "abundance", resolution = "3km")
buwtea_abd_3km <- load_raster("buwtea", product = "abundance", resolution = "3km")
commer_abd_3km <- load_raster("commer", product = "abundance", resolution = "3km")
rinduc_abd_3km <- load_raster("rinduc", product = "abundance", resolution = "3km")
comgol_abd_3km <- load_raster("comgol", product = "abundance", resolution = "3km")
norsho_abd_3km <- load_raster("norsho", product = "abundance", resolution = "3km")
buffle_abd_3km <- load_raster("buffle", product = "abundance", resolution = "3km")
lessca_abd_3km <- load_raster("lessca", product = "abundance", resolution = "3km")
rebmer_abd_3km <- load_raster("rebmer", product = "abundance", resolution = "3km")
gnwtea_abd_3km <- load_raster("gnwtea", product = "abundance", resolution = "3km")
gadwal_abd_3km <- load_raster("gadwal", product = "abundance", resolution = "3km")
ambduc_abd_3km <- load_raster("ambduc", product = "abundance", resolution = "3km")
rudduc_abd_3km <- load_raster("rudduc", product = "abundance", resolution = "3km")

#Non-Anseriform HPAI Detections, n = 6
rethaw_abd_3km <- load_raster("rethaw", product = "abundance", resolution = "3km")
truswa_abd_3km <- load_raster("truswa", product = "abundance", resolution = "3km")
grhowl_abd_3km <- load_raster("grhowl", product = "abundance", resolution = "3km")
amwpel_abd_3km <- load_raster("amwpel", product = "abundance", resolution = "3km")
baleag_abd_3km <- load_raster("baleag", product = "abundance", resolution = "3km")
brdowl_abd_3km <- load_raster("brdowl", product = "abundance", resolution = "3km")

spec_rast_list <- list(mallar3_abd_3km, cangoo_abd_3km, wooduc_abd_3km, buwtea_abd_3km, commer_abd_3km, rinduc_abd_3km, comgol_abd_3km, norsho_abd_3km, buffle_abd_3km, lessca_abd_3km, rebmer_abd_3km, gnwtea_abd_3km, gadwal_abd_3km, ambduc_abd_3km, rudduc_abd_3km, rethaw_abd_3km, truswa_abd_3km, grhowl_abd_3km, amwpel_abd_3km, baleag_abd_3km, brdowl_abd_3km)
names(spec_rast_list) <- c("mallar3", "cangoo", "wooduc", "buwtea", "commer", "rinduc", "comgol", "norsho", "buffle", "lessca", "rebmer", "gnwtea", "gadwal", "ambduc", "rudduc", "rethaw", "truswa", "grhowl", "amwpel", "baleag", "brdowl")
```

-> Data standardization
    - iterate over fishnet geometries, get abundance values for species of interest at point of farm location and aggregated (mean, max) for all cells w/in fishnet geometry.
    - Run once, dataset of feelots, outbreak status, and lag-specific bird abundance saved as: mn-case-ctl-cell-abd-farm-agg-121025.csv

```{r read-in-data}
#Read in Data
fishnet_weeks <- st_read("/Users/kowalr03/Library/CloudStorage/OneDrive-TheMountSinaiHospital/HPAI_data/hpai-data-sharing/Case_Crossover_Study_Design/casecontrol_sf.shp")

case_control_df <- read.csv("/Users/kowalr03/Library/CloudStorage/OneDrive-TheMountSinaiHospital/HPAI_data/hpai-data-sharing/Case_Crossover_Study_Design/casecontrol_df.csv", 
                            stringsAsFactors = F)

spec_name_com <- read.csv("//Users/kowalr03/Data/hpai/mn-ebird-names-lookup-050225.csv")
```


```{r data-standardization}
#Helper Function
# updated with farm point abd value, mean/max across grid for geometry, 12/10/25
# coimmented code has previous techniques -> random assignment, thresholding by areal weight, weighting values
extrHelper <- function(bird_rast, geo_cell, week_num, farm_pt) {
  #Extract data from raster, at entire geometry surrounding feedlot (multi cell), and cell of feedlot location
  extr_df_geo <- terra::extract(bird_rast, geo_cell, snap = "out", touches = T, weights = T, cells = T)
  extr_df_pt <- terra::extract(bird_rast, farm_pt, touches = T, weights = T, cells = T)
  
  #extr_df <- extr_df[extr_df$weight > .15, ] # previous
  #row_get <- sample(nrow(extr_df), 1)
  
  extr_df_geo_vals <- extr_df_geo[, -c(1, 54)] #remove id, weights for values only
  extr_df_pt <- extr_df_pt[, -1]
  
  wk_seq <- seq(week_num - 3, week_num) # define sequence
  wk_seq <- wk_seq[wk_seq > 0]
  
  vals_abd_geo <- extr_df_geo_vals[, wk_seq]
  vals_abd_pt <- extr_df_pt[, wk_seq]
  
  #vals_wei <- extr_df_geo_vals[, wk_seq] #* extr_df_geo$weight[wk_seq]
  if(length(wk_seq) == 1) {# adjust if only taking one
    #vals_wei <- matrix(vals_wei)
    vals_abd_geo  <- matrix(vals_abd_geo)
  }
  val_mean <- apply(vals_abd_geo, 2, mean, na.rm = T) # get mean/median
  val_max <- apply(vals_abd_geo, 2, max, na.rm = T)
  lag_mean <- c(rep(NA, 4 - length(val_mean)), val_mean) # structure lags if falling outside of 2022
  lag_max <- c(rep(NA, 4 - length(val_max)), val_max)
  lag_abd <- c(rep(NA, 4 - length(vals_abd_pt)), as.numeric(vals_abd_pt))
  
  spec_lag_df <- data.frame(lag = -3:0,  lag_abd_pt = lag_abd, lag_mean_grd = lag_mean, lag_max_grd = lag_max)
  return(spec_lag_df)
}

# Main Function
getCellBirdData <- function(case_cont_ind) {
  # Spatial wrangling - get total geometry and point of farm location
  #project to eBird CRS
  vct_geo <- vect(fishnet_weeks$geometry[case_cont_ind])
  vct_geo_prj <- project(vct_geo, spec_rast_list[[1]])
  
  vct_pt_df <- data.frame(longitude = fishnet_weeks$longitd[case_cont_ind], latitude = fishnet_weeks$latitud[case_cont_ind])
  vct_pt <- vect(vct_pt_df, geom = c("longitude", "latitude"), crs = crs(vct_geo))
  vct_pt_prj <- project(vct_pt, spec_rast_list[[1]])  # assigning in place didn't work for point but did for geom, weird
  
  #subset by week+lag
  ob_wk <- fishnet_weeks$Week[case_cont_ind]
  
  #structure lag data by species
  spec_lag_df <- do.call("rbind", lapply(spec_rast_list, extrHelper, geo_cell = vct_geo_prj, week_num = ob_wk, farm_pt = vct_pt_prj))
  spec_lag_df$species <- rep(names(spec_rast_list), each = 4)
  spec_lag_df$zone_id <- fishnet_weeks$Zone_ID[case_cont_ind]
  spec_lag_df$week <- fishnet_weeks$Week[case_cont_ind]
  spec_lag_df$n_outbreak <- fishnet_weeks$Cnt_Otb[case_cont_ind]
  row.names(spec_lag_df) <- NULL
  return(spec_lag_df[, c("zone_id", "week", "n_outbreak", "species", "lag", "lag_abd_pt", "lag_mean_grd", "lag_max_grd")])
}

# Implement
#set.seed(1)
case_ctl_cell_bird_lag <- do.call("rbind", parallel::mclapply(seq_len(nrow(case_control_df)), getCellBirdData, mc.cores = 6))
```

-> Analysis
   - WQS to assess bird mixture effect on outbreak, for various cumulative lags, for various combinations of birds of interest. 
   - American Black Duck, Red-Breasted Merganser removed from quantiling portion due to high missingness
   - Plotting examples included, but not all for brevity

```{r bird-wqs-analysis}
case_ctl_cell_bird_lag <- read.csv("mn-case-ctl-cell-abd-farm-agg-121025.csv")

#Add common names
case_ctl_cell_bird_lag <- case_ctl_cell_bird_lag %>%
                            mutate(common_name = spec_name_com$common_name[match(species, spec_name_com$species_code)])


# Structure data - cumulative lags for both spatial constructions 
# Pivot to wide format for each lag-sum
# 0-quantiling as suggested by CJ
zeroQuantVar <- function(bird_col) {
  bird_qt <- ifelse(bird_col <= 0, NA, bird_col)
  qtile <- quantile(bird_qt, na.rm = T)
  col_qtile <- cut(bird_qt, breaks = qtile, include.lowest = T)
  col_qtile <- ifelse(is.na(col_qtile), 0, col_qtile)  
  return(col_qtile)
}

##### Data Structuring - Farm Point Location #####
# include week of outbreak
abd_pt_lags_zero <- purrr::map(
                         0:-3, 
                          \(x) 
                            case_ctl_cell_bird_lag %>%
                            filter(lag >= x) %>%
                            select(-species) %>%
                            pivot_wider(., 
                                        id_cols = c("zone_id", "week", "n_outbreak"), 
                                        names_from = common_name, values_from = lag_abd_pt, values_fn = \(x) sum(x, na.rm = T)) %>%
                            mutate(any_outbreak = ifelse(n_outbreak >= 1, 1, 0)) %>%
                            select(-n_outbreak) %>%
                            relocate(zone_id, week, any_outbreak))


# don't include week of outbreak
abd_pt_lags_non_zero <- purrr::map(
                         -1:-3, 
                          \(x) 
                            case_ctl_cell_bird_lag %>%
                            filter(lag >= x & lag < 0) %>%
                            select(-species) %>%
                            pivot_wider(., 
                                        id_cols = c("zone_id", "week", "n_outbreak"), 
                                        names_from = common_name, values_from = lag_abd_pt, values_fn = \(x) sum(x, na.rm = T)) %>%
                            mutate(any_outbreak = ifelse(n_outbreak >= 1, 1, 0)) %>%
                            select(-n_outbreak) %>%
                            relocate(zone_id, week, any_outbreak))

# Zero-quantiling 
abd_pt_lags_zero_tile <- purrr::map(
                          abd_pt_lags_zero, 
                          \(x) x %>% select(-c(`Red-breasted Merganser`, `American Black Duck`)) %>% mutate(across(-c("zone_id", "week", "any_outbreak"), zeroQuantVar)))

abd_pt_lags_non_zero_tile <- purrr::map(
                          abd_pt_lags_non_zero, 
                          \(x) x %>% select(-c(`Red-breasted Merganser`, `American Black Duck`)) %>% mutate(across(-c("zone_id", "week", "any_outbreak"), zeroQuantVar)))

##### Data Structuring - Farm 3x3 Grid Cell Max #####
abd_max_lags_zero <- purrr::map(
                      0:-3, 
                      \(x) 
                        case_ctl_cell_bird_lag %>%
                        filter(lag >= x) %>%
                        select(-species) %>%
                        pivot_wider(., 
                                    id_cols = c("zone_id", "week", "n_outbreak"), 
                                    names_from = common_name, values_from = lag_max_grd, values_fn = \(x) sum(x, na.rm = T)) %>%
                        mutate(any_outbreak = ifelse(n_outbreak >= 1, 1, 0)) %>%
                        select(-n_outbreak) %>%
                        relocate(zone_id, week, any_outbreak))

abd_max_lags_non_zero <- purrr::map(
                         -1:-3, 
                          \(x) 
                            case_ctl_cell_bird_lag %>%
                            filter(lag >= x & lag < 0) %>%
                            select(-species) %>%
                            pivot_wider(., 
                                        id_cols = c("zone_id", "week", "n_outbreak"), 
                                        names_from = common_name, values_from = lag_max_grd, values_fn = \(x) sum(x, na.rm = T)) %>%
                            mutate(any_outbreak = ifelse(n_outbreak >= 1, 1, 0)) %>%
                            select(-n_outbreak) %>%
                            relocate(zone_id, week, any_outbreak))

# zero-quantile
abd_max_lags_zero_tile <- purrr::map(
                                abd_max_lags_zero,
                                \(x) x %>% select(-c(`American Black Duck`)) %>% mutate(across(-c("zone_id", "week", "any_outbreak"), zeroQuantVar)))


abd_max_lags_non_zero_tile <- purrr::map(
                                    abd_max_lags_non_zero,
                                    \(x) x %>% select(-c(`American Black Duck`)) %>% mutate(across(-c("zone_id", "week", "any_outbreak"), zeroQuantVar)))

##### Data Structuring - Farm 3x3 Grid Cell Mean #####
abd_mean_lags_zero <- purrr::map(
                      0:-3, 
                      \(x) 
                        case_ctl_cell_bird_lag %>%
                        filter(lag >= x) %>%
                        select(-species) %>%
                        pivot_wider(., 
                                    id_cols = c("zone_id", "week", "n_outbreak"), 
                                    names_from = common_name, values_from = lag_mean_grd, values_fn = \(x) sum(x, na.rm = T)) %>%
                        mutate(any_outbreak = ifelse(n_outbreak >= 1, 1, 0)) %>%
                        select(-n_outbreak) %>%
                        relocate(zone_id, week, any_outbreak))

abd_mean_lags_non_zero <- purrr::map(
                         -1:-3, 
                          \(x) 
                            case_ctl_cell_bird_lag %>%
                            filter(lag >= x & lag < 0) %>%
                            select(-species) %>%
                            pivot_wider(., 
                                        id_cols = c("zone_id", "week", "n_outbreak"), 
                                        names_from = common_name, values_from = lag_mean_grd, values_fn = \(x) sum(x, na.rm = T)) %>%
                            mutate(any_outbreak = ifelse(n_outbreak >= 1, 1, 0)) %>%
                            select(-n_outbreak) %>%
                            relocate(zone_id, week, any_outbreak))

# zero-quantile
abd_mean_lags_zero <- purrr::map(
                                abd_mean_lags_zero,
                                \(x) x %>% select(-c(`American Black Duck`)) %>% mutate(across(-c("zone_id", "week", "any_outbreak"), zeroQuantVar)))


abd_mean_lags_non_zero <- purrr::map(
                                    abd_max_lags_non_zero,
                                    \(x) x %>% select(-c(`American Black Duck`)) %>% mutate(across(-c("zone_id", "week", "any_outbreak"), zeroQuantVar)))


##### WQS Implementation #####
#WQS - lag 0 inclusive, farm point location
spec_mix_mod1 <- names(abd_pt_lags_zero_tile[[1]])[-c(1:3)]

wqs_pt_abd_inc_lag0 <- gwqs(any_outbreak ~ wqs, mix_name = spec_mix_mod1, data = abd_pt_lags_zero_tile[[1]], 
                        q = NULL, validation = 0.6, b = 50, b1_pos = TRUE, rh = 100, family = "binomial", seed = 1)
wqs_pt_abd_inc_lag1 <- gwqs(any_outbreak ~ wqs, mix_name = spec_mix_mod1, data = abd_pt_lags_zero_tile[[2]], 
                        q = NULL, validation = 0.6, b = 50, b1_pos = TRUE, rh = 100, family = "binomial", seed = 1)
wqs_pt_abd_inc_lag2 <- gwqs(any_outbreak ~ wqs, mix_name = spec_mix_mod1, data = abd_pt_lags_zero_tile[[3]], 
                        q = NULL, validation = 0.6, b = 50, b1_pos = TRUE, rh = 100, family = "binomial", seed = 1)
wqs_pt_abd_inc_lag3 <- gwqs(any_outbreak ~ wqs, mix_name = spec_mix_mod1, data = abd_pt_lags_zero_tile[[4]], 
                        q = NULL, validation = 0.6, b = 50, b1_pos = TRUE, rh = 100, family = "binomial", seed = 1)

#WQS - lag 1, farm point location
spec_mix_mod2 <- names(abd_pt_lags_non_zero[[1]])[-c(1:3)]

wqs_pt_abd_lag1 <- gwqs(any_outbreak ~ wqs, mix_name = spec_mix_mod2, data = abd_pt_lags_non_zero[[1]], 
                        q = NULL, validation = 0.6, b = 50, b1_pos = TRUE, rh = 100, family = "binomial", seed = 1)
wqs_pt_abd_lag2 <- gwqs(any_outbreak ~ wqs, mix_name = spec_mix_mod2, data = abd_pt_lags_non_zero[[2]], 
                        q = NULL, validation = 0.6, b = 50, b1_pos = TRUE, rh = 100, family = "binomial", seed = 1)
wqs_pt_abd_lag3 <- gwqs(any_outbreak ~ wqs, mix_name = spec_mix_mod2, data = abd_pt_lags_non_zero[[3]], 
                        q = NULL, validation = 0.6, b = 50, b1_pos = TRUE, rh = 100, family = "binomial", seed = 1)

#WQS - lag 0 inclusive, grid max
spec_mix_mod3 <- names(abd_max_lags_zero[[1]])[-c(1:3)]

wqs_grd_max_inc_lag0 <- gwqs(any_outbreak ~ wqs, mix_name = spec_mix_mod3, data = abd_max_lags_zero[[1]], 
                  q = NULL, validation = 0.6, b = 50, b1_pos = TRUE, rh = 100, family = "binomial", seed = 1)
wqs_grd_max_inc_lag1 <- gwqs(any_outbreak ~ wqs, mix_name = spec_mix_mod3, data = abd_max_lags_zero[[2]], 
                  q = NULL, validation = 0.6, b = 50, b1_pos = TRUE, rh = 100, family = "binomial", seed = 1)
wqs_grd_max_inc_lag2 <- gwqs(any_outbreak ~ wqs, mix_name = spec_mix_mod3, data = abd_max_lags_zero[[3]], 
                  q = NULL, validation = 0.6, b = 50, b1_pos = TRUE, rh = 100, family = "binomial", seed = 1)
wqs_grd_max_inc_lag3 <- gwqs(any_outbreak ~ wqs, mix_name = spec_mix_mod3, data = abd_max_lags_zero[[4]], 
                        q = NULL, validation = 0.6, b = 50, b1_pos = TRUE, rh = 100, family = "binomial", seed = 1)


#WQS - grid max
spec_mix_mod4 <- names(abd_max_lags_non_zero[[1]])[-c(1:3)]

wqs_grd_max_lag1 <- gwqs(any_outbreak ~ wqs, mix_name = spec_mix_mod4, data = abd_max_lags_non_zero[[1]], 
                  q = NULL, validation = 0.6, b = 50, b1_pos = TRUE, rh = 100, family = "binomial", seed = 1)
wqs_grd_max_lag2 <- gwqs(any_outbreak ~ wqs, mix_name = spec_mix_mod4, data = abd_max_lags_non_zero[[2]], 
                  q = NULL, validation = 0.6, b = 50, b1_pos = TRUE, rh = 100, family = "binomial", seed = 1)
wqs_grd_max_lag3 <- gwqs(any_outbreak ~ wqs, mix_name = spec_mix_mod4, data = abd_max_lags_non_zero[[3]], 
                        q = NULL, validation = 0.6, b = 50, b1_pos = TRUE, rh = 100, family = "binomial", seed = 1)

#WQS - lag 0 inclusive, grid mean
spec_mix_mod3 <- names(abd_mean_lags_zero[[1]])[-c(1:3)]

wqs_grd_mean_inc_lag0 <- gwqs(any_outbreak ~ wqs, mix_name = spec_mix_mod3, data = abd_mean_lags_zero[[1]], 
               q = NULL, validation = 0.6, b = 50, b1_pos = TRUE, rh = 100, family = "binomial", seed = 1)
wqs_grd_mean_inc_lag1 <- gwqs(any_outbreak ~ wqs, mix_name = spec_mix_mod3, data = abd_mean_lags_zero[[2]], 
               q = NULL, validation = 0.6, b = 50, b1_pos = TRUE, rh = 100, family = "binomial", seed = 1)
wqs_grd_mean_inc_lag2 <- gwqs(any_outbreak ~ wqs, mix_name = spec_mix_mod3, data = abd_mean_lags_zero[[3]], 
               q = NULL, validation = 0.6, b = 50, b1_pos = TRUE, rh = 100, family = "binomial", seed = 1)
wqs_grd_mean_inc_lag3 <- gwqs(any_outbreak ~ wqs, mix_name = spec_mix_mod3, data = abd_mean_lags_zero[[4]], 
                        q = NULL, validation = 0.6, b = 50, b1_pos = TRUE, rh = 100, family = "binomial", seed = 1)


#WQS - grid mean
spec_mix_mod6 <- names(abd_mean_lags_non_zero[[1]])[-c(1:3)]

wqs_grd_mean_lag1 <- gwqs(any_outbreak ~ wqs, mix_name = spec_mix_mod6, data = abd_mean_lags_non_zero[[1]], 
               q = NULL, validation = 0.6, b = 50, b1_pos = TRUE, rh = 100, family = "binomial", seed = 1)
wqs_grd_mean_lag2 <- gwqs(any_outbreak ~ wqs, mix_name = spec_mix_mod6, data = abd_mean_lags_non_zero[[2]], 
               q = NULL, validation = 0.6, b = 50, b1_pos = TRUE, rh = 100, family = "binomial", seed = 1)
wqs_grd_mean_lag3 <- gwqs(any_outbreak ~ wqs, mix_name = spec_mix_mod6, data = abd_mean_lags_non_zero[[3]], 
                        q = NULL, validation = 0.6, b = 10, b1_pos = TRUE, rh = 10, family = "binomial", seed = 1)

##### Plotting - Example #####
gwqs_boxplot(wqs_grd_mean_lag3) + labs(title = "All Species Qtile WQS, 3x3 Cell Grid Case-Control Mean, Cumulative Exposure Lags (3-1)", subtitle = Sys.Date(), x = "Species")
```



```{r code-archive, eval = F}
# Old code for testing bird raster aggregation
#Code for Testing
ex_vect <- vect(fishnet_weeks$geometry[1])
#ex_vect <- vect(fishnet_weeks$geometry[which(fishnet_weeks$Zone_ID == 72)])

ex_vect_p <- project(ex_vect, spec_rast_list[[13]])
ex_vect_p_buf_5k <- buffer(ex_vect_p, 5000)


bird_extr <- terra::extract(amwpel_abd_3km, ex_vect_p, snap = "out", cells = T, touches = T, weights = T)

bird_extr <- terra::extract(spec_rast_list[[13]], ex_vect_p, snap = "out", cells = T, touches = T, weights = T)


ext(ex_vect_p_buf_5k) <- ext(mallar3_abd_3km)
buf_crop <- crop(mallar3_abd_3km, ex_vect_p_buf_5k, mask = T, touches = T, snap = "out")
plot(buf_crop[[20]])
lines(buf_crop[[20]], add = T)
plot(ex_vect_p_buf_5k, add = T, lty = 2)
plot(ex_vect_p, add = T, lty = 1, lwd = 2)

ex_pt <- data.frame(lon = fishnet_weeks$longitd[1], lat = fishnet_weeks$latitud[1])
ex_pt_prj <- project(vect(ex_pt,  geom = c("lon", "lat"), crs = crs(vect(fishnet_weeks))), spec_rast_list[[13]])
```

