---
title: "season_model_tests"
author: "Fintan Mooney"
date: "2025-10-24"
output: html_document
---


```{r setup, include=FALSE}
 
knitr::opts_chunk$set(echo = TRUE)
 
# ---- Package setup ----
packages <- c(
  "MatchIt", "readr", "sf", "survival", "dlnm", "gtsummary", "kableExtra",
  "tidyr", "tibble", "mgcv", "ggplot2", "car", "patchwork", "corrplot",
  "purrr", "ggspatial", "gt", "flextable", "officer", "conflicted",
  "dplyr", "data.table", "here"
)

installed <- packages %in% rownames(installed.packages())
if (any(!installed)) install.packages(packages[!installed])

invisible(lapply(packages, library, character.only = TRUE))

# Resolve conflicts
conflict_prefer("select", "dplyr")
conflict_prefer("filter", "dplyr")

# Project folders (e.g., where the .Rproj file is)
base_folder <- here()
data_folder <- here("data")
env_folder <- here("data", "EarthEngine_MN")
objects_folder <-here("data", "objects")
figures_folder <-here("output", "figures")
tables_folder <-here("output", "tables")

list.files(here())

```

## Timeseries 

### Read in Timeseries 

```{r, import timeseries dataframe}

# Timeseries
fishnet_timeseries_clean <- readRDS(file.path(objects_folder, "fishnet_timeseries.RDS"))


```

### Prepare for modeling

```{r, prepare}
timeseries_clean <- fishnet_timeseries_clean %>%
  arrange(zone_id, week)

timeseries_clean$season <- factor(timeseries_clean$season,
                                  levels = c("Fall", "Spring", "Winter", "Summer"))

```


## Scale

```{r, scale}

# Variables to log1p transform
log_vars <- c("total_bird_abundance", "runoff", "precipitation")

# Apply log1p transformation only if column exists
timeseries_clean <- timeseries_clean %>%
  mutate(across(all_of(log_vars), ~ log1p(.), .names = "{.col}"))

# All variables to scale 
vars_to_scale <- c(
  "total_bird_abundance", "leaf_area_index", "longwave_radiation", "runoff",
  "shortwave_radiation", "snow_density", "soil_moisture", 
  "temperature", "evapotranspiration", "precipitation", "wind_speed",
  "relative_humidity"
)

# Apply z-score scaling
timeseries_clean <- timeseries_clean %>%
  mutate(across(all_of(vars_to_scale), scale, .names = "{.col}"))


```

### Crossbasis terms (timeseries)


#### Linear


```{r, crossbasis lin}

colnames(timeseries_clean)

# Crossbasis terms for selected variables using Lag0–Lag4
cb_lin_relative_humidity <- crossbasis(
  timeseries_clean$relative_humidity,
  lag = 4,
  argvar = list(fun = "lin"),
  arglag = list(fun = "ns", df = 2)
)

cb_lin_evap <- crossbasis(
  timeseries_clean$evapotranspiration,
  lag = 4,
  argvar = list(fun = "lin"),
  arglag = list(fun = "ns", df = 2)
)

cb_lin_leaf_area_index <- crossbasis(
  timeseries_clean$leaf_area_index,
  lag = 4,
  argvar = list(fun = "lin"),
  arglag = list(fun = "ns", df = 2)
)

cb_lin_runoff <- crossbasis(
  timeseries_clean$runoff,
  lag = 4,
  argvar = list(fun = "lin"),
  arglag = list(fun = "ns", df = 2)
)

cb_lin_shortwave_radiation <- crossbasis(
  timeseries_clean$shortwave_radiation,
  lag = 4,
  argvar = list(fun = "lin"),
  arglag = list(fun = "ns", df = 2)
)

cb_lin_longwave_radiation <- crossbasis(
  timeseries_clean$longwave_radiation,
  lag = 4,
  argvar = list(fun = "lin"),
  arglag = list(fun = "ns", df = 2)
)

cb_lin_snow_density <- crossbasis(
  timeseries_clean$snow_density,
  lag = 4,
  argvar = list(fun = "lin"),
  arglag = list(fun = "ns", df = 2)
)

cb_lin_soil_moisture <- crossbasis(
  timeseries_clean$soil_moisture,
  lag = 4,
  argvar = list(fun = "lin"),
  arglag = list(fun = "ns", df = 2)
)

cb_lin_temperature <- crossbasis(
  timeseries_clean$temperature,
  lag = 4,
  argvar = list(fun = "lin"),
  arglag = list(fun = "ns", df = 2)
)

cb_lin_precipitation <- crossbasis(
  timeseries_clean$precipitation,
  lag = 4,
  argvar = list(fun = "lin"),
  arglag = list(fun = "ns", df = 2)
)

cb_lin_wind_speed <- crossbasis(
  timeseries_clean$wind_speed,
  lag = 4,
  argvar = list(fun = "lin"),
  arglag = list(fun = "ns", df = 2)
)

cb_lin_bird_abundance <- crossbasis(
  timeseries_clean$total_bird_abundance,
  lag = 4,
  argvar = list(fun = "lin"),
  arglag = list(fun = "ns", df = 2)
)


```

#### Natural Spline (df = 3)

```{r, crossbasis ns df 3}


# Crossbasis terms for selected variables using Lag0–Lag4

cb_ns_relative_humidity <- crossbasis(
  timeseries_clean$relative_humidity,
  lag = 4,
  argvar = list(fun = "ns", df = 2),
  arglag = list(fun = "ns", df = 3)
)

cb_ns_evap <- crossbasis(
  timeseries_clean$evapotranspiration,
  lag = 4,
  argvar = list(fun = "ns", df = 2),
  arglag = list(fun = "ns", df = 3)
)

cb_ns_leaf_area_index <- crossbasis(
  timeseries_clean$leaf_area_index,
  lag = 4,
  argvar = list(fun = "ns", df = 2),
  arglag = list(fun = "ns", df = 3)
)

cb_ns_runoff <- crossbasis(
  timeseries_clean$runoff,
  lag = 4,
  argvar = list(fun = "ns", df = 2),
  arglag = list(fun = "ns", df = 3)
)

cb_ns_shortwave_radiation <- crossbasis(
  timeseries_clean$shortwave_radiation,
  lag = 4,
  argvar = list(fun = "ns", df = 2),
  arglag = list(fun = "ns", df = 3)
)

cb_ns_longwave_radiation <- crossbasis(
  timeseries_clean$longwave_radiation,
  lag = 4,
  argvar = list(fun = "ns", df = 2),
  arglag = list(fun = "ns", df = 3)
)

cb_ns_snow_density <- crossbasis(
  timeseries_clean$snow_density,
  lag = 4,
  argvar = list(fun = "ns", df = 2),
  arglag = list(fun = "ns", df = 3)
)

cb_ns_soil_moisture <- crossbasis(
  timeseries_clean$soil_moisture,
  lag = 4,
  argvar = list(fun = "ns", df = 2),
  arglag = list(fun = "ns", df = 3)
)

cb_ns_temperature <- crossbasis(
  timeseries_clean$temperature,
  lag = 4,
  argvar = list(fun = "ns", df = 2),
  arglag = list(fun = "ns", df = 3)
)

cb_ns_precipitation <- crossbasis(
  timeseries_clean$precipitation,
  lag = 4,
  argvar = list(fun = "ns", df = 2),
  arglag = list(fun = "ns", df = 3)
)

cb_ns_wind_speed <- crossbasis(
  timeseries_clean$wind_speed,
  lag = 4,
  argvar = list(fun = "ns", df = 2),
  arglag = list(fun = "ns", df = 3)
)

cb_ns_bird_abundance <- crossbasis(
  timeseries_clean$total_bird_abundance,
  lag = 4,
  argvar = list(fun = "ns", df = 2),
  arglag = list(fun = "ns", df = 3)
)


```

#### Natural Spline (df=2)

```{r, crossbasis ns df 2}


# Crossbasis terms for selected variables using Lag0–Lag4

cb_ns2_relative_humidity <- crossbasis(
  timeseries_clean$relative_humidity,
  lag = 4,
  argvar = list(fun = "ns", df = 2),
  arglag = list(fun = "ns", df = 2)
)

cb_ns2_evap <- crossbasis(
  timeseries_clean$evapotranspiration,
  lag = 4,
  argvar = list(fun = "ns", df = 2),
  arglag = list(fun = "ns", df = 2)
)

cb_ns2_leaf_area_index <- crossbasis(
  timeseries_clean$leaf_area_index,
  lag = 4,
  argvar = list(fun = "ns", df = 2),
  arglag = list(fun = "ns", df = 2)
)

cb_ns2_runoff <- crossbasis(
  timeseries_clean$runoff,
  lag = 4,
  argvar = list(fun = "ns", df = 2),
  arglag = list(fun = "ns", df = 2)
)

cb_ns2_shortwave_radiation <- crossbasis(
  timeseries_clean$shortwave_radiation,
  lag = 4,
  argvar = list(fun = "ns", df = 2),
  arglag = list(fun = "ns", df = 2)
)

cb_ns2_longwave_radiation <- crossbasis(
  timeseries_clean$longwave_radiation,
  lag = 4,
  argvar = list(fun = "ns", df = 2),
  arglag = list(fun = "ns", df = 2)
)

cb_ns2_snow_density <- crossbasis(
  timeseries_clean$snow_density,
  lag = 4,
  argvar = list(fun = "ns", df = 2),
  arglag = list(fun = "ns", df = 2)
)

cb_ns2_soil_moisture <- crossbasis(
  timeseries_clean$soil_moisture,
  lag = 4,
  argvar = list(fun = "ns", df = 2),
  arglag = list(fun = "ns", df = 2)
)

cb_ns2_temperature <- crossbasis(
  timeseries_clean$temperature,
  lag = 4,
  argvar = list(fun = "ns", df = 2),
  arglag = list(fun = "ns", df = 2)
)

cb_ns2_precipitation <- crossbasis(
  timeseries_clean$precipitation,
  lag = 4,
  argvar = list(fun = "ns", df = 2),
  arglag = list(fun = "ns", df = 2)
)

cb_ns2_wind_speed <- crossbasis(
  timeseries_clean$wind_speed,
  lag = 4,
  argvar = list(fun = "ns", df = 2),
  arglag = list(fun = "ns", df = 2)
)

cb_ns2_bird_abundance <- crossbasis(
  timeseries_clean$total_bird_abundance,
  lag = 4,
  argvar = list(fun = "ns", df = 2),
  arglag = list(fun = "ns", df = 2)
)


```


## Test Models with Seasonal Fixed Effects

```{r, model seasonal effects}

season_timeseries_model_lin <- bam(outbreak_binary ~
                   cb_lin_precipitation +
                   cb_lin_snow_density + 
                   cb_lin_runoff +
                   cb_lin_soil_moisture +
                   cb_lin_bird_abundance +
                   cb_lin_wind_speed +
                   cb_lin_temperature +
                     s(lat, lon, bs ="ds", k =200) +
                    season, # fixed effect
                    family = nb(),
                    data = timeseries_clean,
                    method = "fREML",
                    discrete = TRUE)

summary(season_timeseries_model_lin)

season_timeseries_model_ns <- bam(outbreak_binary ~
                   cb_ns_precipitation +
                   cb_ns_snow_density + 
                   cb_ns_runoff +
                   cb_ns_soil_moisture +
                   cb_ns_bird_abundance +
                   cb_ns_wind_speed +
                   cb_ns_temperature +
                     s(lat, lon, bs ="ds", k =200) +
                    season, # fixed effect
                    family = nb(),
                    data = timeseries_clean,
                    method = "fREML",
                    discrete = TRUE)

summary(season_timeseries_model_ns)

season_timeseries_model_ns_2 <- bam(outbreak_binary ~
                   cb_ns2_precipitation +
                   cb_ns2_snow_density + 
                   cb_ns2_runoff +
                   cb_ns2_soil_moisture +
                   cb_ns2_bird_abundance +
                   cb_ns2_wind_speed +
                   cb_ns2_temperature +
                     s(lat, lon, bs ="ds", k =200) +
                    season, # fixed effect
                    family = nb(),
                    data = timeseries_clean,
                    method = "fREML",
                    discrete = TRUE)

summary(season_timeseries_model_ns_2)


season_timeseries_model_binomial <- bam(outbreak_binary ~
                   cb_ns2_precipitation +
                   cb_ns2_snow_density + 
                   cb_ns2_runoff +
                   cb_ns2_soil_moisture +
                   cb_ns2_bird_abundance +
                   cb_ns2_wind_speed +
                   cb_ns2_temperature +
                     s(lat, lon, bs ="ds", k =200) +
                    season, # fixed effect
family = binomial(link = "logit"),
                    data = timeseries_clean,
                    method = "fREML",
                    discrete = TRUE)

summary(season_timeseries_model_binomial)

season_timeseries_model_quasi <- bam(outbreak_binary ~
                   cb_ns2_precipitation +
                   cb_ns2_snow_density + 
                   cb_ns2_runoff +
                   cb_ns2_soil_moisture +
                   cb_ns2_bird_abundance +
                   cb_ns2_wind_speed +
                   cb_ns2_temperature +
                     s(lat, lon, bs ="ds", k =200) +
                    season, # fixed effect
family = quasibinomial(link = "logit"),
                    data = timeseries_clean,
                    method = "fREML",
                    discrete = TRUE)

summary(season_timeseries_model_quasi)

season_timeseries_model_binomial_lin <- bam(outbreak_binary ~
                   cb_lin_precipitation +
                   cb_lin_snow_density + 
                   cb_lin_runoff +
                   cb_lin_soil_moisture +
                   cb_lin_bird_abundance +
                   cb_lin_wind_speed +
                   cb_lin_temperature +
                     s(lat, lon, bs ="ds", k =200) +
                    season, # fixed effect
family = binomial(link = "logit"),
                    data = timeseries_clean,
                    method = "fREML",
                    discrete = TRUE)

summary(season_timeseries_model_binomial)

season_timeseries_model_quasi_lin <- bam(outbreak_binary ~
                   cb_lin_precipitation +
                   cb_lin_snow_density + 
                   cb_lin_runoff +
                   cb_lin_soil_moisture +
                   cb_lin_bird_abundance +
                   cb_lin_wind_speed +
                   cb_lin_temperature +
                     s(lat, lon, bs ="ds", k =200) +
                    season, # fixed effect
family = quasibinomial(link = "logit"),
                    data = timeseries_clean,
                    method = "fREML",
                    discrete = TRUE)

summary(season_timeseries_model_quasi_lin)

AIC(season_timeseries_model_lin,season_timeseries_model_ns, season_timeseries_model_ns_2, season_timeseries_model_binomial,season_timeseries_model_binomial_lin,season_timeseries_model_quasi_lin)
season_timeseries_model_quasi$gcv.ubre

res_dev  <- deviance(season_timeseries_model_binomial)
df_resid <- df.residual(season_timeseries_model_binomial)
dispersion <- res_dev / df_resid
dispersion


```

# We choose the model that has non-linear (ns, df = 2) arglag and argvar crossbasis terms and is in the negative binomial family
## This choice is because we expect there to be a non-linear exposure-response relationship when modeling on the full dataset (timeseries) and because the outcome is overdispersed
### This is the season_timeseries_model_ns_2 above (any reductions in AIC between the other negative binomial models is negligible)