---
title: "casecrossover_model"
author: "Fintan Mooney"
date: "2025-10-07"
output: html_document
---
# This R markdown file uses a casecrossover study design to model the association between meterological conditons, climate, and wild bird abundance, on risk of highly pathogenic avian influenza (HPAI) outbreaks in Minnesota farms during 2022

```{r setup, include=FALSE}
 
knitr::opts_chunk$set(echo = TRUE)
 
# ---- Package setup ----
packages <- c(
  "MatchIt", "readr", "sf", "survival", "dlnm", "gtsummary", "kableExtra",
  "tidyr", "tibble", "mgcv", "ggplot2", "car", "patchwork", "corrplot",
  "purrr", "ggspatial", "gt", "flextable", "officer", "conflicted",
  "dplyr", "data.table", "here"
)

installed <- packages %in% rownames(installed.packages())
if (any(!installed)) install.packages(packages[!installed])

invisible(lapply(packages, library, character.only = TRUE))

# Resolve conflicts
conflict_prefer("select", "dplyr")
conflict_prefer("filter", "dplyr")

# Project folders (e.g., where the .Rproj file is)
base_folder <- here()
data_folder <- here("data")
env_folder <- here("data", "EarthEngine_MN")
objects_folder <-here("data", "objects")
figures_folder <-here("output", "figures")
tables_folder <-here("output", "tables")

list.files(here())

```

## Import casecrossover df

```{r, import casecrossover df}

case_crossover_df <- readRDS(file.path(objects_folder, "case_crossover_df.RDS"))

```


## Crossbasis Terms

```{r, finalized crossbasis}


# Finalized Crossbasis Terms for Climate and Bird Predictors

# Leaf Area Index
cb_leaf_area_index <- crossbasis(
  x = as.matrix(case_crossover_df[, paste0("leaf_area_index_Lag", 0:4)]),
  lag = c(0, 4),
  argvar = list(fun = "lin"),
  arglag = list(fun = "ns", df = 2)
)

# Longwave Radiation
cb_longwave_radiation <- crossbasis(
  x = as.matrix(case_crossover_df[, paste0("longwave_radiation_Lag", 0:4)]),
  lag = c(0, 4),
  argvar = list(fun = "lin"),
  arglag = list(fun = "ns", df = 2)
)

# Precipitation
cb_precipitation <- crossbasis(
  x = as.matrix(case_crossover_df[, paste0("precipitation_Lag", 0:4)]),
  lag = c(0, 4),
  argvar = list(fun = "lin"),
  arglag = list(fun = "ns", df = 2)
)

# Relative Humidity
cb_relative_humidity <- crossbasis(
  x = as.matrix(case_crossover_df[, paste0("relative_humidity_Lag", 0:4)]),
  lag = c(0, 4),
  argvar = list(fun = "lin"),
  arglag = list(fun = "ns", df = 2)
)

# Runoff
cb_runoff <- crossbasis(
  x = as.matrix(case_crossover_df[, paste0("runoff_Lag", 0:4)]),
  lag = c(0, 4),
  argvar = list(fun = "lin"),
  arglag = list(fun = "ns", df = 2)
)

# Shortwave Radiation
cb_shortwave_radiation <- crossbasis(
  x = as.matrix(case_crossover_df[, paste0("shortwave_radiation_Lag", 0:4)]),
  lag = c(0, 4),
  argvar = list(fun = "lin"),
  arglag = list(fun = "ns", df = 2)
)

# Snow Density
cb_snow_density <- crossbasis(
  x = as.matrix(case_crossover_df[, paste0("snow_density_Lag", 0:4)]),
  lag = c(0, 4),
  argvar = list(fun = "lin"),
  arglag = list(fun = "ns", df = 2)
)

# Soil Moisture
cb_soil_moisture <- crossbasis(
  x = as.matrix(case_crossover_df[, paste0("soil_moisture_Lag", 0:4)]),
  lag = c(0, 4),
  argvar = list(fun = "lin"),
  arglag = list(fun = "ns", df = 2)
)

# Temperature
cb_temperature <- crossbasis(
  x = as.matrix(case_crossover_df[, paste0("temperature_Lag", 0:4)]),
  lag = c(0, 4),
  argvar = list(fun = "lin"),
  arglag = list(fun = "ns", df = 2)
)


# Total Bird Abundance
cb_total_bird_abundance <- crossbasis(
  x = as.matrix(case_crossover_df[, paste0("total_bird_abundance_Lag", 0:4)]),
  lag = c(0, 4),
  argvar = list(fun = "lin"),
  arglag = list(fun = "ns", df = 2)
)

# Wind Speed
cb_wind_speed <- crossbasis(
  x = as.matrix(case_crossover_df[, paste0("wind_speed_Lag", 0:4)]),
  lag = c(0, 4),
  argvar = list(fun = "lin"),
  arglag = list(fun = "ns", df = 2)
)

# # # Anseriformes:
#  cb_anseri <- crossbasis(
#    x = as.matrix(case_crossover_df[, paste0("anseriformes_Lag", 0:8)]),
#    lag = c(0, 8),
#    argvar = list(fun = "lin"),
#    arglag = list(fun = "poly", degree = 2)
#  )
#  
# # # Predator:
#  cb_predators <- crossbasis(
#    x = as.matrix(case_crossover_df[, paste0("predators_Lag", 0:8)]),
#    lag = c(0, 8),
#    argvar = list(fun = "lin"),
#    arglag = list(fun = "poly", degree = 2)
#  )




```

### More flexible crossbasis terms

```{r, more flexible crossbasis terms}

# Flexible Crossbasis Terms for Climate and Bird Predictors (increased dfs and polynomial splines)

# Leaf Area Index: 
cb_flex_leaf_area_index <- crossbasis(
  x = as.matrix(case_crossover_df[, paste0("leaf_area_index_Lag", 0:4)]),
  lag = c(0, 4),
  argvar = list(fun = "lin"),
  arglag = list(fun = "ns", df = 2)
)

# Longwave Radiation: 
cb_flex_longwave_radiation <- crossbasis(
  x = as.matrix(case_crossover_df[, paste0("longwave_radiation_Lag", 0:4)]),
  lag = c(0, 4),
  argvar = list(fun = "lin"),
  arglag = list(fun = "ns", df = 2)
)

# Precipitation: 
cb_flex_precipitation <- crossbasis(
  x = as.matrix(case_crossover_df[, paste0("precipitation_Lag", 0:4)]),
  lag = c(0, 4),
  argvar = list(fun = "lin"),
  arglag = list(fun = "ns", df = 2)
)

# Relative Humidity: 
cb_flex_relative_humidity <- crossbasis(
  x = as.matrix(case_crossover_df[, paste0("relative_humidity_Lag", 0:4)]),
  lag = c(0, 4),
  argvar = list(fun = "lin"),
  arglag = list(fun = "poly", degree = 2)
)

# Runoff: 
cb_flex_runoff <- crossbasis(
  x = as.matrix(case_crossover_df[, paste0("runoff_Lag", 0:4)]),
  lag = c(0, 4),
  argvar = list(fun = "lin"),
  arglag = list(fun = "poly", degree = 2)
)

# Shortwave Radiation: 
cb_flex_shortwave_radiation <- crossbasis(
  x = as.matrix(case_crossover_df[, paste0("shortwave_radiation_Lag", 0:4)]),
  lag = c(0, 4),
  argvar = list(fun = "lin"),
  arglag = list(fun = "poly", degree = 4)
)

# Snow Density: 
cb_flex_snow_density <- crossbasis(
  x = as.matrix(case_crossover_df[, paste0("snow_density_Lag", 0:4)]),
  lag = c(0, 4),
  argvar = list(fun = "lin"),
  arglag = list(fun = "poly", degree = 3)
)

# Soil Moisture:
cb_flex_soil_moisture <- crossbasis(
  x = as.matrix(case_crossover_df[, paste0("soil_moisture_Lag", 0:4)]),
  lag = c(0, 4),
  argvar = list(fun = "lin"),
  arglag = list(fun = "ns", df = 2)
)

# Temperature:
cb_flex_temperature <- crossbasis(
  x = as.matrix(case_crossover_df[, paste0("temperature_Lag", 0:4)]),
  lag = c(0, 4),
  argvar = list(fun = "lin"),
  arglag = list(fun = "ns", df = 2)
)


# Total Bird Abundance: 
cb_flex_total_bird_abundance <- crossbasis(
  x = as.matrix(case_crossover_df[, paste0("total_bird_abundance_Lag", 0:4)]),
  lag = c(0, 4),
  argvar = list(fun = "lin"),
  arglag = list(fun = "ns", df = 2)
)

# Wind Speed: 
cb_flex_wind_speed <- crossbasis(
  x = as.matrix(case_crossover_df[, paste0("wind_speed_Lag", 0:4)]),
  lag = c(0, 4),
  argvar = list(fun = "lin"),
  arglag = list(fun = "poly", degree = 4)
)

# # Anseriformes:
# cb_flex_anseri <- crossbasis(
#   x = as.matrix(case_crossover_df[, paste0("anseriformes_Lag", 0:8)]),
#    lag = c(0, 8),
#    argvar = list(fun = "lin"),
#    arglag = list(fun = "poly", degree = 2)
#  )
#  
# # Predator: 
#  cb_flex_predators <- crossbasis(
#    x = as.matrix(case_crossover_df[, paste0("predators_Lag", 0:8)]),
#    lag = c(0, 8),
#    argvar = list(fun = "lin"),
#    arglag = list(fun = "poly", degree = 2)
#  )

```


## Case-Crossover Model

```{r, casecrossover model}

# ----------------------------
# Model Fitting
# ----------------------------
final_model <- clogit(
  outbreak_binary ~ 
    cb_runoff + cb_snow_density + cb_soil_moisture + cb_temperature +
    cb_precipitation + cb_wind_speed + cb_total_bird_abundance +
    strata(stratum_id),
  method = "efron",
  data = case_crossover_df
)

summary(final_model)
AIC(final_model)
saveRDS(final_model, file.path(objects_folder, "casecrossover_model.RDS"))

```

```{r, crosspred objects}
# ----------------------------
# Crosspred Objects
# ----------------------------
cp_runoff       <- crosspred(cb_runoff, final_model, cen = 0, at = 0.5, cumul = TRUE)
cp_snow         <- crosspred(cb_snow_density, final_model, cen = 0, at = 0.5, cumul = TRUE)
cp_soil         <- crosspred(cb_soil_moisture, final_model, cen = 0, at = 0.5, cumul = TRUE)
cp_temp         <- crosspred(cb_temperature, final_model, cen = 0, at = 0.5, cumul = TRUE)
cp_precip       <- crosspred(cb_precipitation, final_model, cen = 0, at = 0.5, cumul = TRUE)
cp_wind         <- crosspred(cb_wind_speed, final_model, cen = 0, at = 0.5, cumul = TRUE)
cp_bird         <- crosspred(cb_total_bird_abundance, final_model, cen = 0, at = 0.5, cumul = TRUE)

cp_slice_runoff <- crosspred(cb_runoff, final_model, cen = 0, at = 0.5)
cp_slice_snow   <- crosspred(cb_snow_density, final_model, cen = 0, at = 0.5)
cp_slice_soil   <- crosspred(cb_soil_moisture, final_model, cen = 0, at = 0.5)
cp_slice_temp   <- crosspred(cb_temperature, final_model, cen = 0, at = 0.5)
cp_slice_precip <- crosspred(cb_precipitation, final_model, cen = 0, at = 0.5)
cp_slice_wind   <- crosspred(cb_wind_speed, final_model, cen = 0, at = 0.5)
cp_slice_bird   <- crosspred(cb_total_bird_abundance, final_model, cen = 0, at = 0.5)

```

## Casecrossover Plots
```{r, plots casecrossover}
# ----------------------------
# Cumulative OR Summary Table
# ----------------------------
cumulative_summary <- tibble(
  Predictor = c("Runoff", "Snow Density", "Soil Moisture", "Temperature",
                "Precipitation", "Wind Speed", "Total Bird Abundance"),
      OR = c(
        cp_runoff$allRRfit["0.5"],
        cp_snow$allRRfit["0.5"],
        cp_soil$allRRfit["0.5"],
        cp_temp$allRRfit["0.5"],
        cp_precip$allRRfit["0.5"],
        cp_wind$allRRfit["0.5"],
        cp_bird$allRRfit["0.5"]
      ),
      CI_low = c(
        cp_runoff$allRRlow["0.5"],
        cp_snow$allRRlow["0.5"],
        cp_soil$allRRlow["0.5"],
        cp_temp$allRRlow["0.5"],
        cp_precip$allRRlow["0.5"],
        cp_wind$allRRlow["0.5"],
        cp_bird$allRRlow["0.5"]
      ),
      CI_high = c(
        cp_runoff$allRRhigh["0.5"],
        cp_snow$allRRhigh["0.5"],
        cp_soil$allRRhigh["0.5"],
        cp_temp$allRRhigh["0.5"],
        cp_precip$allRRhigh["0.5"],
        cp_wind$allRRhigh["0.5"],
        cp_bird$allRRhigh["0.5"]
      )
) %>%
  mutate(`OR (95% CI)` = sprintf("%.2f (%.2f, %.2f)", OR, CI_low, CI_high)) %>%
  select(Predictor, `OR (95% CI)`) %>%
  gt() %>%
  tab_header(
    title = md("**Case-Crossover Model: Cumulative Odds Ratios**"),
    subtitle = "Per 0.5 SD Increase"
  ) %>%
  cols_label(`OR (95% CI)` = "Odds Ratio (95% Confidence Interval)") %>%
  tab_options(
    table.font.size = 14,
    heading.title.font.size = 16,
    heading.subtitle.font.size = 14,
    data_row.padding = px(6)
  )

# ----------------------------
# Plotting Functions
# ----------------------------
plot_cumulative <- function(cp, title) {
  plot(cp, type = "cumulative", var = 0.5, ci = "area",
       col = "black", ci.arg = list(col = rgb(0,0,0,0.1), border = NA),
       lwd = 1.5, main = title, xlab = "", ylab = "", las = 1)
  abline(h = 1, lty = 2, col = "gray40")
}

plot_slices <- function(cp, title) {
  plot(cp, type = "slices", var = 0.5, ci = "area",
       col = "black", ci.arg = list(col = rgb(0,0,0,0.1), border = NA),
       lwd = 1.3, main = title, xlab = "", ylab = "", las = 1)
  abline(h = 1, lty = 2, col = "gray40")
}

# ----------------------------
# Save Cumulative Plot
# ----------------------------
png(file.path(figures_folder,"figure_cumulative_ORs_casecrossover.png"), width = 11, height = 7, units = "in", res = 300)

par(
  mfrow = c(2, 4),
  mar = c(4.2, 4.2, 2.5, 1),
  oma = c(3, 4, 0, 1),
  cex.main = 1.2,
  cex.lab = 1.2,
  cex.axis = 1.1,
  tcl = -0.3, mgp = c(2.5, 0.7, 0)
)

plot_cumulative(cp_runoff, "Runoff")
plot_cumulative(cp_snow, "Snow Density")
plot_cumulative(cp_soil, "Soil Moisture")
plot_cumulative(cp_temp, "Temperature")
plot_cumulative(cp_precip, "Precipitation")
plot_cumulative(cp_wind, "Wind Speed")
plot_cumulative(cp_bird, "Total Bird Abundance")

mtext("Lag (weeks)", side = 1, outer = TRUE, line = 0, cex = 1.2)
mtext("Odds Ratio (95% CI)", side = 2, outer = TRUE, line = 0, cex = 1.2)

dev.off()

# ----------------------------
# Save Slices Plot
# ----------------------------
png(file.path(figures_folder,"figure_slice_ORs_casecrossover.png"), width = 11, height = 7, units = "in", res = 300)

par(
  mfrow = c(2, 4),
  mar = c(4.2, 4.2, 2.5, 1),
  oma = c(3, 4, 0, 1),
  cex.main = 1.2,
  cex.lab = 1.2,
  cex.axis = 1.1,
  tcl = -0.3, mgp = c(2.5, 0.7, 0)
)

plot_slices(cp_slice_runoff, "Runoff")
plot_slices(cp_slice_snow, "Snow Density")
plot_slices(cp_slice_soil, "Soil Moisture")
plot_slices(cp_slice_temp, "Temperature")
plot_slices(cp_slice_precip, "Precipitation")
plot_slices(cp_slice_wind, "Wind Speed")
plot_slices(cp_slice_bird, "Total Bird Abundance")

mtext("Lag (weeks)", side = 1, outer = TRUE, line = 0, cex = 1.2)
mtext("Odds Ratio (95% CI)", side = 2, outer = TRUE, line = 0, cex = 1.2)

dev.off()


```


## Lag Specific Odds Ratio Table (Casecrossover Model)
```{r, lag specific odds ratio table}
# ----------------------------
# Extract Lag-Specific ORs (Wide)
# ----------------------------
extract_lag_ORs_wide <- function(cp_obj, varname, varval = "0.5", lags = 0:4) {
  lag_labels <- paste0("lag", lags)
  or    <- cp_obj$matRRfit[varval, lag_labels]
  low   <- cp_obj$matRRlow[varval, lag_labels]
  high  <- cp_obj$matRRhigh[varval, lag_labels]  # FIXED HERE
  tibble(
    Predictor = varname,
    !!!setNames(as.list(or),   paste0("OR_Lag", lags)),
    !!!setNames(as.list(low),  paste0("Low_Lag", lags)),
    !!!setNames(as.list(high), paste0("High_Lag", lags)),
    !!!setNames(as.list(sprintf("%.2f (%.2f, %.2f)", or, low, high)), paste0("Lag ", lags))
  )
}


# ----------------------------
# Create Lag-Specific Table
# ----------------------------
table_lag_ORs <- bind_rows(
  extract_lag_ORs_wide(cp_slice_runoff, "Runoff"),
  extract_lag_ORs_wide(cp_slice_snow, "Snow Density"),
  extract_lag_ORs_wide(cp_slice_soil, "Soil Moisture"),
  extract_lag_ORs_wide(cp_slice_temp, "Temperature"),
  extract_lag_ORs_wide(cp_slice_precip, "Precipitation"),
  extract_lag_ORs_wide(cp_slice_wind, "Wind Speed"),
  extract_lag_ORs_wide(cp_slice_bird, "Total Bird Abundance")
)



# Highlight significant rows
lags <- 0:4
highlight_matrix <- sapply(lags, function(i) {
  table_lag_ORs[[paste0("Low_Lag", i)]] > 1 | table_lag_ORs[[paste0("High_Lag", i)]] < 1
})
colnames(highlight_matrix) <- paste0("Lag ", lags)

# Export to Word
ft <- table_lag_ORs %>%
  select(Predictor, paste0("Lag ", lags)) %>%
  flextable() %>%
  set_header_labels(Predictor = "Predictor") %>%
  add_header_row(values = c("Predictor", "Odds Ratio (95% CI)"), colwidths = c(1, length(lags))) %>%
  fontsize(size = 11, part = "all") %>%
  autofit() %>%
  align(align = "center", part = "all") %>%
  theme_booktabs()

for (i in lags) {
  ft <- bold(ft, i = which(highlight_matrix[, paste0("Lag ", i)]), j = paste0("Lag ", i))
}

doc <- read_docx() %>%
  body_add_flextable(ft) %>%
  body_add_par(
    "Table: Lag-Specific Odds Ratios for Case-Crossover Study on the Association between Environmental Conditions and Highly Pathogenic Avian Influenza Outbreaks in Minnesota, 2022. Estimates are per 0.5 standard deviation increase in each predictor. Statistically significant odds ratios are shown in bold.",
    style = "Normal"
  )

print(doc, target = file.path(tables_folder, "Table_LagSpecific_OR_HPAI_Casecrossover.docx"))

ft

```

### Cumulative odds ratio table (0-4 lag)


```{r, cumulative odds ratios}
## Save lag specific so we can patch with cumulative

ft_lagspecific <- ft

# ----------------------------
# Extract Lag-Specific ORs (Wide)
# ----------------------------
extract_lag_ORs_wide <- function(cp_obj, varname, varval = "0.5", lags = 0:4) {
  lag_labels <- paste0("lag", lags)
  or    <- cp_obj$cumRRfit[varval, lag_labels]
  low   <- cp_obj$cumRRlow[varval, lag_labels]
  high  <- cp_obj$cumRRhigh[varval, lag_labels]  # FIXED HERE
  tibble(
    Predictor = varname,
    !!!setNames(as.list(or),   paste0("OR_Lag", lags)),
    !!!setNames(as.list(low),  paste0("Low_Lag", lags)),
    !!!setNames(as.list(high), paste0("High_Lag", lags)),
    !!!setNames(as.list(sprintf("%.2f (%.2f, %.2f)", or, low, high)), paste0("Lag ", lags))
  )
}


# ----------------------------
# Create Cumulative OR Table
# ----------------------------
table_lag_ORs_cum <- bind_rows(
  extract_lag_ORs_wide(cp_runoff, "Runoff"),
  extract_lag_ORs_wide(cp_snow, "Snow Density"),
  extract_lag_ORs_wide(cp_soil, "Soil Moisture"),
  extract_lag_ORs_wide(cp_temp, "Temperature"),
  extract_lag_ORs_wide(cp_precip, "Precipitation"),
  extract_lag_ORs_wide(cp_wind, "Wind Speed"),
  extract_lag_ORs_wide(cp_bird, "Total Bird Abundance")
)



# Highlight significant rows
lags <- 0:4
highlight_matrix <- sapply(lags, function(i) {
  table_lag_ORs_cum[[paste0("Low_Lag", i)]] > 1 | table_lag_ORs_cum[[paste0("High_Lag", i)]] < 1
})
colnames(highlight_matrix) <- paste0("Lag ", lags)

# Export to Word
ft_cum <- table_lag_ORs_cum %>%
  select(Predictor, paste0("Lag ", lags)) %>%
  flextable() %>%
  set_header_labels(Predictor = "Predictor") %>%
  add_header_row(values = c("Predictor", "Odds Ratio (95% CI)"), colwidths = c(1, length(lags))) %>%
  fontsize(size = 11, part = "all") %>%
  autofit() %>%
  align(align = "center", part = "all") %>%
  theme_booktabs()

for (i in lags) {
  ft_cum <- bold(ft_cum, i = which(highlight_matrix[, paste0("Lag ", i)]), j = paste0("Lag ", i))
}

doc <- read_docx() %>%
  body_add_flextable(ft_cum) %>%
  body_add_par(
    "Table: Cumulative Odds Ratios for Case-Crossover Study on the Association between Environmental Conditions and Highly Pathogenic Avian Influenza Outbreaks in Minnesota, 2022. Estimates are per 0.5 standard deviation increase in each predictor. Statistically significant odds ratios are shown in bold.",
    style = "Normal"
  )

print(doc, target = file.path(tables_folder, "Table_Cumulative_OR_HPAI_Casscrossover.docx"))

ft_cum

```

#### Combine cumulative and lag-specific odds ratio tables

```{r, combine and lagspecific odds ratio tables}

panel_label <- function(txt, size = 11) {
  fp <- fp_text(bold = TRUE, font.size = size)
  fpar(ftext(txt, fp))
}

doc <- read_docx() %>%
  # Panel A
  body_add_fpar(panel_label("A. Lag-specific odds ratios (per +0.5 SD)")) %>%
  body_add_flextable(ft_lagspecific) %>%
  body_add_par("", style = "Normal") %>%   # spacer

  # Panel B
  body_add_fpar(panel_label("B. Cumulative odds ratios (per +0.5 SD)")) %>%
  body_add_flextable(ft_cum)

print(doc, target = file.path(tables_folder, "Table_LagSpecific_and_Cumulative_ORs_Casecrossover.docx"))


```



### Cumulative odds ratio table (0-1 lag, 2-4 lag)

```{r, cumulative odds ratio lag sections}


# Cumulative Crosspreds by Lag Window (0–1 and 2–4)

# ---- Surface Runoff ----
cp_runoff_0_1 <- crosspred(cb_runoff, final_model, cen = 0, at = 0.5,
                           cumul = TRUE, from = 0, to = 1)
cp_runoff_2_4 <- crosspred(cb_runoff, final_model, cen = 0, at = 0.5,
                           cumul = TRUE, from = 2, to = 4)

# ---- Snow Density ----
cp_snow_0_1 <- crosspred(cb_snow_density, final_model, cen = 0, at = 0.5,
                         cumul = TRUE, from = 0, to = 1)
cp_snow_2_4 <- crosspred(cb_snow_density, final_model, cen = 0, at = 0.5,
                         cumul = TRUE, from = 2, to = 4)

# ---- Soil Moisture ----
cp_soil_0_1 <- crosspred(cb_soil_moisture, final_model, cen = 0, at = 0.5,
                         cumul = TRUE, from = 0, to = 1)
cp_soil_2_4 <- crosspred(cb_soil_moisture, final_model, cen = 0, at = 0.5,
                         cumul = TRUE, from = 2, to = 4)

# ---- Temperature ----
cp_temp_0_1 <- crosspred(cb_temperature, final_model, cen = 0, at = 0.5,
                         cumul = TRUE, from = 0, to = 1)
cp_temp_2_4 <- crosspred(cb_temperature, final_model, cen = 0, at = 0.5,
                         cumul = TRUE, from = 2, to = 4)

# ---- Precipitation ----
cp_precip_0_1 <- crosspred(cb_precipitation, final_model, cen = 0, at = 0.5,
                           cumul = TRUE, from = 0, to = 1)
cp_precip_2_4 <- crosspred(cb_precipitation, final_model, cen = 0, at = 0.5,
                           cumul = TRUE, from = 2, to = 4)

# ---- Wind Speed ----
cp_wind_0_1 <- crosspred(cb_wind_speed, final_model, cen = 0, at = 0.5,
                         cumul = TRUE, from = 0, to = 1)
cp_wind_2_4 <- crosspred(cb_wind_speed, final_model, cen = 0, at = 0.5,
                         cumul = TRUE, from = 2, to = 4)

# ---- Total Bird Abundance ----
cp_bird_0_1 <- crosspred(cb_total_bird_abundance, final_model, cen = 0, at = 0.5,
                         cumul = TRUE, from = 0, to = 1)
cp_bird_2_4 <- crosspred(cb_total_bird_abundance, final_model, cen = 0, at = 0.5,
                         cumul = TRUE, from = 2, to = 4)

# Table

table_cum_ORs <- tibble(
  Predictor = c("Surface Runoff", "Snow Density", "Soil Moisture",
                "Temperature", "Precipitation", "Wind Speed", "Total Bird Abundance"),

  `Lag 0–1` = c(
    sprintf("%.2f (%.2f, %.2f)", cp_runoff_0_1$cumRRfit["0.5", 1],
                                  cp_runoff_0_1$cumRRlow["0.5", 1],
                                  cp_runoff_0_1$cumRRhigh["0.5", 1]),
    sprintf("%.2f (%.2f, %.2f)", cp_snow_0_1$cumRRfit["0.5", 1],
                                  cp_snow_0_1$cumRRlow["0.5", 1],
                                  cp_snow_0_1$cumRRhigh["0.5", 1]),
    sprintf("%.2f (%.2f, %.2f)", cp_soil_0_1$cumRRfit["0.5", 1],
                                  cp_soil_0_1$cumRRlow["0.5", 1],
                                  cp_soil_0_1$cumRRhigh["0.5", 1]),
    sprintf("%.2f (%.2f, %.2f)", cp_temp_0_1$cumRRfit["0.5", 1],
                                  cp_temp_0_1$cumRRlow["0.5", 1],
                                  cp_temp_0_1$cumRRhigh["0.5", 1]),
    sprintf("%.2f (%.2f, %.2f)", cp_precip_0_1$cumRRfit["0.5", 1],
                                  cp_precip_0_1$cumRRlow["0.5", 1],
                                  cp_precip_0_1$cumRRhigh["0.5", 1]),
    sprintf("%.2f (%.2f, %.2f)", cp_wind_0_1$cumRRfit["0.5", 1],
                                  cp_wind_0_1$cumRRlow["0.5", 1],
                                  cp_wind_0_1$cumRRhigh["0.5", 1]),
    sprintf("%.2f (%.2f, %.2f)", cp_bird_0_1$cumRRfit["0.5", 1],
                                  cp_bird_0_1$cumRRlow["0.5", 1],
                                  cp_bird_0_1$cumRRhigh["0.5", 1])
  ),

  `Lag 2–4` = c(
    sprintf("%.2f (%.2f, %.2f)", cp_runoff_2_4$cumRRfit["0.5", 4],
                                  cp_runoff_2_4$cumRRlow["0.5", 4],
                                  cp_runoff_2_4$cumRRhigh["0.5", 4]),
    sprintf("%.2f (%.2f, %.2f)", cp_snow_2_4$cumRRfit["0.5", 4],
                                  cp_snow_2_4$cumRRlow["0.5", 4],
                                  cp_snow_2_4$cumRRhigh["0.5", 4]),
    sprintf("%.2f (%.2f, %.2f)", cp_soil_2_4$cumRRfit["0.5", 4],
                                  cp_soil_2_4$cumRRlow["0.5", 4],
                                  cp_soil_2_4$cumRRhigh["0.5", 4]),
    sprintf("%.2f (%.2f, %.2f)", cp_temp_2_4$cumRRfit["0.5", 4],
                                  cp_temp_2_4$cumRRlow["0.5", 4],
                                  cp_temp_2_4$cumRRhigh["0.5", 4]),
    sprintf("%.2f (%.2f, %.2f)", cp_precip_2_4$cumRRfit["0.5", 4],
                                  cp_precip_2_4$cumRRlow["0.5", 4],
                                  cp_precip_2_4$cumRRhigh["0.5", 4]),
    sprintf("%.2f (%.2f, %.2f)", cp_wind_2_4$cumRRfit["0.5", 4],
                                  cp_wind_2_4$cumRRlow["0.5", 4],
                                  cp_wind_2_4$cumRRhigh["0.5", 4]),
    sprintf("%.2f (%.2f, %.2f)", cp_bird_2_4$cumRRfit["0.5", 4],
                                  cp_bird_2_4$cumRRlow["0.5", 4],
                                  cp_bird_2_4$cumRRhigh["0.5", 4])
  )
)



# Highlight significant results


highlight_matrix <- sapply(c("Lag 0–1", "Lag 2–4"), function(w) {
  apply(table_cum_ORs, 1, function(x) {
    parts <- as.numeric(unlist(strsplit(gsub("[()]", "", x[[w]]), "[ ,]+")))
    low <- parts[2]; high <- parts[3]
    low > 1 | high < 1
  })
})
colnames(highlight_matrix) <- c("Lag 0–1", "Lag 2–4")


# Build Flextable


ft_cum <- table_cum_ORs %>%
  flextable() %>%
  set_header_labels(Predictor = "Predictor") %>%
  add_header_row(values = c("Predictor", "Cumulative Odds Ratio (95% CI)"), colwidths = c(1, 2)) %>%
  fontsize(size = 11, part = "all") %>%
  autofit() %>%
  align(align = "center", part = "all") %>%
  theme_booktabs()

for (w in c("Lag 0–1", "Lag 2–4")) {
  ft_cum <- bold(ft_cum, i = which(highlight_matrix[, w]), j = w)
}


# Export to Word


doc <- read_docx() %>%
  body_add_flextable(ft_cum) %>%
  body_add_par(
    "Table: Cumulative Odds Ratios for Case-Crossover Study on the Association between Environmental Conditions and Highly Pathogenic Avian Influenza Outbreaks in Minnesota (Fall + Spring 2022). Odds ratios represent cumulative effects over lag windows (0–1 weeks, 2–4 weeks) per 0.5 SD increase in each predictor. Statistically significant associations are shown in bold.",
    style = "Normal"
  )

print(doc, target = file.path(tables_folder, "Table_Cumulative_OR_HPAI_Minnesota.docx"))

ft_cum

```

## Case-Crossover Model with Flexible Crossbasis

```{r, casecrossover flex model}

# ----------------------------
# Model Fitting
# ----------------------------
final_model_flex <- clogit(
  outbreak_binary ~ 
    cb_flex_runoff + cb_flex_snow_density + cb_flex_soil_moisture + cb_flex_temperature +
    cb_flex_precipitation + cb_flex_wind_speed + cb_flex_total_bird_abundance +
    strata(stratum_id),
  method = "efron",
  data = case_crossover_df
)

summary(final_model_flex)
AIC(final_model_flex)
saveRDS(final_model_flex, file.path(objects_folder, "casecrossover_flexible_model.RDS"))

```

```{r, crosspred objects for flex model}
# ----------------------------
# Crosspred Objects
# ----------------------------
cp_flex_runoff       <- crosspred(cb_flex_runoff, final_model_flex, cen = 0, at = 0.5, cumul = TRUE)
cp_flex_snow         <- crosspred(cb_flex_snow_density, final_model_flex, cen = 0, at = 0.5, cumul = TRUE)
cp_flex_soil         <- crosspred(cb_flex_soil_moisture, final_model_flex, cen = 0, at = 0.5, cumul = TRUE)
cp_flex_temp         <- crosspred(cb_flex_temperature, final_model_flex, cen = 0, at = 0.5, cumul = TRUE)
cp_flex_precip       <- crosspred(cb_flex_precipitation, final_model_flex, cen = 0, at = 0.5, cumul = TRUE)
cp_flex_wind         <- crosspred(cb_flex_wind_speed, final_model_flex, cen = 0, at = 0.5, cumul = TRUE)
cp_flex_bird         <- crosspred(cb_flex_total_bird_abundance, final_model_flex, cen = 0, at = 0.5, cumul = TRUE)

cp_flex_slice_runoff <- crosspred(cb_flex_runoff, final_model_flex, cen = 0, at = 0.5)
cp_flex_slice_snow   <- crosspred(cb_flex_snow_density, final_model_flex, cen = 0, at = 0.5)
cp_flex_slice_soil   <- crosspred(cb_flex_soil_moisture, final_model_flex, cen = 0, at = 0.5)
cp_flex_slice_temp   <- crosspred(cb_flex_temperature, final_model_flex, cen = 0, at = 0.5)
cp_flex_slice_precip <- crosspred(cb_flex_precipitation, final_model_flex, cen = 0, at = 0.5)
cp_flex_slice_wind   <- crosspred(cb_flex_wind_speed, final_model_flex, cen = 0, at = 0.5)
cp_flex_slice_bird   <- crosspred(cb_flex_total_bird_abundance, final_model_flex, cen = 0, at = 0.5)

```

## Casecrossover Plots
```{r, plots casecrossover flex}
# ----------------------------
# Cumulative OR Summary Table
# ----------------------------
cumulative_summary <- tibble(
  Predictor = c("Surface Runoff", "Snow Density", "Soil Moisture", "Temperature",
                "Precipitation", "Wind Speed", "Total Bird Abundance"),
      OR = c(
        cp_flex_runoff$allRRfit["0.5"],
        cp_flex_snow$allRRfit["0.5"],
        cp_flex_soil$allRRfit["0.5"],
        cp_flex_temp$allRRfit["0.5"],
        cp_flex_precip$allRRfit["0.5"],
        cp_flex_wind$allRRfit["0.5"],
        cp_flex_bird$allRRfit["0.5"]
      ),
      CI_low = c(
        cp_flex_runoff$allRRlow["0.5"],
        cp_flex_snow$allRRlow["0.5"],
        cp_flex_soil$allRRlow["0.5"],
        cp_flex_temp$allRRlow["0.5"],
        cp_flex_precip$allRRlow["0.5"],
        cp_flex_wind$allRRlow["0.5"],
        cp_flex_bird$allRRlow["0.5"]
      ),
      CI_high = c(
        cp_flex_runoff$allRRhigh["0.5"],
        cp_flex_snow$allRRhigh["0.5"],
        cp_flex_soil$allRRhigh["0.5"],
        cp_flex_temp$allRRhigh["0.5"],
        cp_flex_precip$allRRhigh["0.5"],
        cp_flex_wind$allRRhigh["0.5"],
        cp_flex_bird$allRRhigh["0.5"]
      )
) %>%
  mutate(`OR (95% CI)` = sprintf("%.2f (%.2f, %.2f)", OR, CI_low, CI_high)) %>%
  select(Predictor, `OR (95% CI)`) %>%
  gt() %>%
  tab_header(
    title = md("**Case-Crossover Model: Cumulative Odds Ratios**"),
    subtitle = "Per 0.5 SD Increase"
  ) %>%
  cols_label(`OR (95% CI)` = "Odds Ratio (95% Confidence Interval)") %>%
  tab_options(
    table.font.size = 14,
    heading.title.font.size = 16,
    heading.subtitle.font.size = 14,
    data_row.padding = px(6)
  )

cumulative_summary

# ----------------------------
# Plotting Functions
# ----------------------------
plot_cumulative <- function(cp_flex_, title) {
  plot(cp_flex_, type = "cumulative", var = 0.5, ci = "area",
       col = "black", ci.arg = list(col = rgb(0,0,0,0.1), border = NA),
       lwd = 1.5, main = title, xlab = "", ylab = "", las = 1)
  abline(h = 1, lty = 2, col = "gray40")
}

plot_slices <- function(cp_flex_, title) {
  plot(cp_flex_, type = "slices", var = 0.5, ci = "area",
       col = "black", ci.arg = list(col = rgb(0,0,0,0.1), border = NA),
       lwd = 1.3, main = title, xlab = "", ylab = "", las = 1)
  abline(h = 1, lty = 2, col = "gray40")
}

# ----------------------------
# Save Cumulative Plot
# ----------------------------
png(file.path(figures_folder,"figure_cumulative_ORs_casecrossover_flexmodel.png"), width = 11, height = 7, units = "in", res = 300)

par(
  mfrow = c(2, 4),
  mar = c(4.2, 4.2, 2.5, 1),
  oma = c(3, 4, 0, 1),
  cex.main = 1.2,
  cex.lab = 1.2,
  cex.axis = 1.1,
  tcl = -0.3, mgp = c(2.5, 0.7, 0)
)

plot_cumulative(cp_flex_runoff, "Runoff")
plot_cumulative(cp_flex_snow, "Snow Density")
plot_cumulative(cp_flex_soil, "Soil Moisture")
plot_cumulative(cp_flex_temp, "Temperature")
plot_cumulative(cp_flex_precip, "Precipitation")
plot_cumulative(cp_flex_wind, "Wind Speed")
plot_cumulative(cp_flex_bird, "Total Bird Abundance")

mtext("Lag (weeks)", side = 1, outer = TRUE, line = 0, cex = 1.2)
mtext("Odds Ratio (95% CI)", side = 2, outer = TRUE, line = 0, cex = 1.2)

# ----------------------------
# Save Slices Plot
# ----------------------------
png(file.path(figures_folder,"figure_slice_ORs_casecrossover_flexmodel.png"), width = 11, height = 7, units = "in", res = 300)

par(
  mfrow = c(2, 4),
  mar = c(4.2, 4.2, 2.5, 1),
  oma = c(3, 4, 0, 1),
  cex.main = 1.2,
  cex.lab = 1.2,
  cex.axis = 1.1,
  tcl = -0.3, mgp = c(2.5, 0.7, 0)
)

plot_slices(cp_flex_slice_runoff, "Runoff")
plot_slices(cp_flex_slice_snow, "Snow Density")
plot_slices(cp_flex_slice_soil, "Soil Moisture")
plot_slices(cp_flex_slice_temp, "Temperature")
plot_slices(cp_flex_slice_precip, "Precipitation")
plot_slices(cp_flex_slice_wind, "Wind Speed")
plot_slices(cp_flex_slice_bird, "Total Bird Abundance")

mtext("Lag (weeks)", side = 1, outer = TRUE, line = 0, cex = 1.2)
mtext("Odds Ratio (95% CI)", side = 2, outer = TRUE, line = 0, cex = 1.2)

dev.off()

```


## Lag Specific Odds Ratio Table (Casecrossover Model)
```{r, lag specific odds ratio table for flex model}
# ----------------------------
# Extract Lag-Specific ORs (Wide)
# ----------------------------
extract_lag_ORs_wide <- function(cp_flex_obj, varname, varval = "0.5", lags = 0:4) {
  lag_labels <- paste0("lag", lags)
  or    <- cp_flex_obj$matRRfit[varval, lag_labels]
  low   <- cp_flex_obj$matRRlow[varval, lag_labels]
  high  <- cp_flex_obj$matRRhigh[varval, lag_labels]  
  tibble(
    Predictor = varname,
    !!!setNames(as.list(or),   paste0("OR_Lag", lags)),
    !!!setNames(as.list(low),  paste0("Low_Lag", lags)),
    !!!setNames(as.list(high), paste0("High_Lag", lags)),
    !!!setNames(as.list(sprintf("%.2f (%.2f, %.2f)", or, low, high)), paste0("Lag ", lags))
  )
}


# ----------------------------
# Create Lag-Specific Table
# ----------------------------
table_lag_ORs <- bind_rows(
  extract_lag_ORs_wide(cp_flex_slice_runoff, "Runoff"),
  extract_lag_ORs_wide(cp_flex_slice_snow, "Snow Density"),
  extract_lag_ORs_wide(cp_flex_slice_soil, "Soil Moisture"),
  extract_lag_ORs_wide(cp_flex_slice_temp, "Temperature"),
  extract_lag_ORs_wide(cp_flex_slice_precip, "Precipitation"),
  extract_lag_ORs_wide(cp_flex_slice_wind, "Wind Speed"),
  extract_lag_ORs_wide(cp_flex_slice_bird, "Total Bird Abundance")
)



# Highlight significant rows
lags <- 0:4
highlight_matrix <- sapply(lags, function(i) {
  table_lag_ORs[[paste0("Low_Lag", i)]] > 1 | table_lag_ORs[[paste0("High_Lag", i)]] < 1
})
colnames(highlight_matrix) <- paste0("Lag ", lags)

# Export to Word
ft <- table_lag_ORs %>%
  select(Predictor, paste0("Lag ", lags)) %>%
  flextable() %>%
  set_header_labels(Predictor = "Predictor") %>%
  add_header_row(values = c("Predictor", "Odds Ratio (95% CI)"), colwidths = c(1, length(lags))) %>%
  fontsize(size = 11, part = "all") %>%
  autofit() %>%
  align(align = "center", part = "all") %>%
  theme_booktabs()

for (i in lags) {
  ft <- bold(ft, i = which(highlight_matrix[, paste0("Lag ", i)]), j = paste0("Lag ", i))
}

doc <- read_docx() %>%
  body_add_flextable(ft) %>%
  body_add_par(
    "Table: Lag-Specific Odds Ratios for Case-Crossover Study on the Association between Environmental Conditions and Highly Pathogenic Avian Influenza Outbreaks in Minnesota, 2022. Estimates are per 0.5 standard deviation increase in each predictor. Statistically significant odds ratios are shown in bold.",
    style = "Normal"
  )

print(doc, target = file.path(tables_folder, "Table_LagSpecific_OR_HPAI_Minnesota_flexmodel.docx"))

ft

```



## Cumulative odds ratio table (0-1 lag, 2-4 lag)

```{r, cumulative odds ratio lag sections}


# Cumulative Crosspreds by Lag Window (0–1 and 2–4)

# ---- Surface Runoff ----
cp_runoff_0_1 <- crosspred(cb_flex_runoff, final_model_flex, cen = 0, at = 0.5,
                           cumul = TRUE, from = 0, to = 1)
cp_runoff_2_4 <- crosspred(cb_flex_runoff, final_model_flex, cen = 0, at = 0.5,
                           cumul = TRUE, from = 2, to = 4)

# ---- Snow Density ----
cp_snow_0_1 <- crosspred(cb_flex_snow_density, final_model_flex, cen = 0, at = 0.5,
                         cumul = TRUE, from = 0, to = 1)
cp_snow_2_4 <- crosspred(cb_flex_snow_density, final_model_flex, cen = 0, at = 0.5,
                         cumul = TRUE, from = 2, to = 4)

# ---- Soil Moisture ----
cp_soil_0_1 <- crosspred(cb_flex_soil_moisture, final_model_flex, cen = 0, at = 0.5,
                         cumul = TRUE, from = 0, to = 1)
cp_soil_2_4 <- crosspred(cb_flex_soil_moisture, final_model_flex, cen = 0, at = 0.5,
                         cumul = TRUE, from = 2, to = 4)

# ---- Temperature ----
cp_temp_0_1 <- crosspred(cb_flex_temperature, final_model_flex, cen = 0, at = 0.5,
                         cumul = TRUE, from = 0, to = 1)
cp_temp_2_4 <- crosspred(cb_flex_temperature, final_model_flex, cen = 0, at = 0.5,
                         cumul = TRUE, from = 2, to = 4)

# ---- Precipitation ----
cp_precip_0_1 <- crosspred(cb_flex_precipitation, final_model_flex, cen = 0, at = 0.5,
                           cumul = TRUE, from = 0, to = 1)
cp_precip_2_4 <- crosspred(cb_flex_precipitation, final_model_flex, cen = 0, at = 0.5,
                           cumul = TRUE, from = 2, to = 4)

# ---- Wind Speed ----
cp_wind_0_1 <- crosspred(cb_flex_wind_speed, final_model_flex, cen = 0, at = 0.5,
                         cumul = TRUE, from = 0, to = 1)
cp_wind_2_4 <- crosspred(cb_flex_wind_speed, final_model_flex, cen = 0, at = 0.5,
                         cumul = TRUE, from = 2, to = 4)

# ---- Total Bird Abundance ----
cp_bird_0_1 <- crosspred(cb_flex_total_bird_abundance, final_model_flex, cen = 0, at = 0.5,
                         cumul = TRUE, from = 0, to = 1)
cp_bird_2_4 <- crosspred(cb_flex_total_bird_abundance, final_model_flex, cen = 0, at = 0.5,
                         cumul = TRUE, from = 2, to = 4)

# Table

table_cum_ORs <- tibble(
  Predictor = c("Surface Runoff", "Snow Density", "Soil Moisture",
                "Temperature", "Precipitation", "Wind Speed", "Total Bird Abundance"),

  `Lag 0–1` = c(
    sprintf("%.2f (%.2f, %.2f)", cp_runoff_0_1$cumRRfit["0.5", 1],
                                  cp_runoff_0_1$cumRRlow["0.5", 1],
                                  cp_runoff_0_1$cumRRhigh["0.5", 1]),
    sprintf("%.2f (%.2f, %.2f)", cp_snow_0_1$cumRRfit["0.5", 1],
                                  cp_snow_0_1$cumRRlow["0.5", 1],
                                  cp_snow_0_1$cumRRhigh["0.5", 1]),
    sprintf("%.2f (%.2f, %.2f)", cp_soil_0_1$cumRRfit["0.5", 1],
                                  cp_soil_0_1$cumRRlow["0.5", 1],
                                  cp_soil_0_1$cumRRhigh["0.5", 1]),
    sprintf("%.2f (%.2f, %.2f)", cp_temp_0_1$cumRRfit["0.5", 1],
                                  cp_temp_0_1$cumRRlow["0.5", 1],
                                  cp_temp_0_1$cumRRhigh["0.5", 1]),
    sprintf("%.2f (%.2f, %.2f)", cp_precip_0_1$cumRRfit["0.5", 1],
                                  cp_precip_0_1$cumRRlow["0.5", 1],
                                  cp_precip_0_1$cumRRhigh["0.5", 1]),
    sprintf("%.2f (%.2f, %.2f)", cp_wind_0_1$cumRRfit["0.5", 1],
                                  cp_wind_0_1$cumRRlow["0.5", 1],
                                  cp_wind_0_1$cumRRhigh["0.5", 1]),
    sprintf("%.2f (%.2f, %.2f)", cp_bird_0_1$cumRRfit["0.5", 1],
                                  cp_bird_0_1$cumRRlow["0.5", 1],
                                  cp_bird_0_1$cumRRhigh["0.5", 1])
  ),

  `Lag 2–4` = c(
    sprintf("%.2f (%.2f, %.2f)", cp_runoff_2_4$cumRRfit["0.5", 4],
                                  cp_runoff_2_4$cumRRlow["0.5", 4],
                                  cp_runoff_2_4$cumRRhigh["0.5", 4]),
    sprintf("%.2f (%.2f, %.2f)", cp_snow_2_4$cumRRfit["0.5", 4],
                                  cp_snow_2_4$cumRRlow["0.5", 4],
                                  cp_snow_2_4$cumRRhigh["0.5", 4]),
    sprintf("%.2f (%.2f, %.2f)", cp_soil_2_4$cumRRfit["0.5", 4],
                                  cp_soil_2_4$cumRRlow["0.5", 4],
                                  cp_soil_2_4$cumRRhigh["0.5", 4]),
    sprintf("%.2f (%.2f, %.2f)", cp_temp_2_4$cumRRfit["0.5", 4],
                                  cp_temp_2_4$cumRRlow["0.5", 4],
                                  cp_temp_2_4$cumRRhigh["0.5", 4]),
    sprintf("%.2f (%.2f, %.2f)", cp_precip_2_4$cumRRfit["0.5", 4],
                                  cp_precip_2_4$cumRRlow["0.5", 4],
                                  cp_precip_2_4$cumRRhigh["0.5", 4]),
    sprintf("%.2f (%.2f, %.2f)", cp_wind_2_4$cumRRfit["0.5", 4],
                                  cp_wind_2_4$cumRRlow["0.5", 4],
                                  cp_wind_2_4$cumRRhigh["0.5", 4]),
    sprintf("%.2f (%.2f, %.2f)", cp_bird_2_4$cumRRfit["0.5", 4],
                                  cp_bird_2_4$cumRRlow["0.5", 4],
                                  cp_bird_2_4$cumRRhigh["0.5", 4])
  )
)



# Highlight significant results


highlight_matrix <- sapply(c("Lag 0–1", "Lag 2–4"), function(w) {
  apply(table_cum_ORs, 1, function(x) {
    parts <- as.numeric(unlist(strsplit(gsub("[()]", "", x[[w]]), "[ ,]+")))
    low <- parts[2]; high <- parts[3]
    low > 1 | high < 1
  })
})
colnames(highlight_matrix) <- c("Lag 0–1", "Lag 2–4")


# Build Flextable


ft_cum <- table_cum_ORs %>%
  flextable() %>%
  set_header_labels(Predictor = "Predictor") %>%
  add_header_row(values = c("Predictor", "Cumulative Odds Ratio (95% CI)"), colwidths = c(1, 2)) %>%
  fontsize(size = 11, part = "all") %>%
  autofit() %>%
  align(align = "center", part = "all") %>%
  theme_booktabs()

for (w in c("Lag 0–1", "Lag 2–4")) {
  ft_cum <- bold(ft_cum, i = which(highlight_matrix[, w]), j = w)
}


# Export to Word


doc <- read_docx() %>%
  body_add_flextable(ft_cum) %>%
  body_add_par(
    "Table: Cumulative Odds Ratios for Case-Crossover Study on the Association between Environmental Conditions and Highly Pathogenic Avian Influenza Outbreaks in Minnesota (Fall + Spring 2022). Odds ratios represent cumulative effects over lag windows (0–1 weeks, 2–4 weeks) per 0.5 SD increase in each predictor. Statistically significant associations are shown in bold.",
    style = "Normal"
  )

print(doc, target = file.path(tables_folder, "Table_Cumulative_OR_HPAI_Minnesota_flexmodel.docx"))

ft_cum

```

