---
title: "table_1"
author: "Fintan Mooney"
date: "2025-10-31"
output:
  word_document: default
  html_document: default
---
# This R markdown file creates a table 1 for the casecrossover study on the association between meterological conditons, climate, and wild bird abundance, on risk of highly pathogenic avian influenza (HPAI) outbreaks in Minnesota farms during 2022

```{r setup, include=FALSE}
 
knitr::opts_chunk$set(echo = TRUE)
 
# ---- Package setup ----
packages <- c(
  "MatchIt", "readr", "sf", "survival", "dlnm", "gtsummary", "kableExtra",
  "tidyr", "tibble", "mgcv", "ggplot2", "car", "patchwork", "corrplot",
  "purrr", "ggspatial", "gt", "flextable", "officer", "conflicted",
  "dplyr", "data.table", "here", "table1", "boot", "forcats", "gtsummary",
  "lmerTest", "emmeans", "broom.mixed", "lme4", "stringr", "clubSandwich"
)

installed <- packages %in% rownames(installed.packages())
if (any(!installed)) install.packages(packages[!installed])

invisible(lapply(packages, library, character.only = TRUE))

# Resolve conflicts
conflict_prefer("select", "dplyr")
conflict_prefer("filter", "dplyr")
conflicts_prefer(dplyr::first)
conflicts_prefer(lmerTest::lmer)


# Project folders (e.g., where the .Rproj file is)
base_folder <- here()
data_folder <- here("data")
env_folder <- here("data", "EarthEngine_MN")
objects_folder <-here("data", "objects")
figures_folder <-here("output", "figures")
tables_folder <-here("output", "tables")

list.files(here())

```

```{r, read in timeseries df}

fishnet_timeseries_clean <- readRDS(file.path(objects_folder, "fishnet_timeseries.RDS"))

```


```{r, list columns}
(colnames(fishnet_timeseries_clean))

```
## Subset to variables of interest

```{r, subset variables}

vars_subset <- c(
  # --- Absolute environmental variables ---
  "leaf_area_index",
  "longwave_radiation",
  "precipitation",
  "relative_humidity",
  "runoff",                 
  "shortwave_radiation",
  "snow_density",
  "soil_moisture",
  "temperature",
  "anseriformes",
  "predators",
  "wind_speed",

  # --- 25-year z-score anomalies ---
  "leaf_area_index_z_25yr",
  "longwave_radiation_z_25yr",
  "precipitation_z_25yr",
  "relative_humidity_z_25yr",
  "runoff_z_25yr",
  "shortwave_radiation_z_25yr",
  "snow_density_z_25yr",
  "soil_moisture_z_25yr",
  "temperature_z_25yr",
  "wind_speed_z_25yr",
  
    # --- 25-year mean  ---
  "leaf_area_index_mean_25yr",
  "longwave_radiation_mean_25yr",
  "precipitation_mean_25yr",
  "relative_humidity_mean_25yr",
  "runoff_mean_25yr",
  "shortwave_radiation_mean_25yr",
  "snow_density_mean_25yr",
  "soil_moisture_mean_25yr",
  "temperature_mean_25yr",
  "wind_speed_mean_25yr",

  # --- ID, time, outbreak, and context ---
  "zone_id",
  "week",
  "season",
  "total_bird_abundance",
  "any_outbreak",
  "land_cover_mode",
  "land_cover_label",
  "outbreak_count",
  "outbreak_binary"
)

fishnet_subset <- fishnet_timeseries_clean %>%
  select(any_of(vars_subset))  

# Check

glimpse(fishnet_subset)
colnames(fishnet_subset)

fishnet_subset %>%
  select(where(is.numeric)) %>% # Select only numeric columns
  summarise(across(everything(), list(mean = mean, sd = sd), na.rm = TRUE))
```

## Transform variables 

```{r, transform variables}

fishnet_t1 <- fishnet_subset %>%
  mutate(
    ## Temperature: Kelvin -> Celsius
    temperature_C = temperature - 273.15,

    ## Precipitation & Runoff: meters -> millimeters
    precipitation_mm = precipitation * 1000,
    runoff_mm        = runoff * 1000,

    ## Radiation: convert from W/m² to MJ/m² (divide by 1e6)
    longwave_radiation  = longwave_radiation  / 1e6,
    shortwave_radiation = shortwave_radiation / 1e6,

    ## Drop previous factor levels
    land_cover_label = factor(droplevels(land_cover_label)),
    
    ## Make outbreak count categorical
    
    #outbreak_count_factor = factor(outbreak_count),

    ## Grouping variable as labeled factor
    any_outbreak = factor(any_outbreak,
                          levels = c(0, 1),
                          labels = c("Cells with no outbreak", "Cells with an outbreak"))
  )

fishnet_t1$land_cover_label <- forcats::fct_infreq(fishnet_t1$land_cover_label)


# Check transformations
fishnet_t1 %>%
  summarise(
    across(
      c(temperature_C, precipitation_mm, runoff_mm,
        shortwave_radiation, longwave_radiation),
      list(mean = ~mean(.x, na.rm = TRUE), sd = ~sd(.x, na.rm = TRUE))
    )
  )

```

# Creeate Table 1

```{r, table 1}

# Categorical variable renderer: Percent only
my.render.cat <- function(x) {
  c("", sapply(stats.default(x), function(y) with(y,
    sprintf("%0.0f%%", PCT))))
}


pvalue <- function(x, ...) {
  # Construct vectors of data y, and groups (strata) g
  y <- unlist(x)
  g <- factor(rep(1:length(x), times = sapply(x, length)))
  
  if (is.numeric(y)) {
    # For numeric variables, perform a standard 2-sample t-test
    p <- t.test(y ~ g)$p.value
  } else {
    # For categorical variables, perform a chi-squared test of independence
    p <- chisq.test(table(y, g))$p.value
  }
  
  # Format the p-value (HTML-safe for <)
  formatted_p <- sub("<", "&lt;", format.pval(p, digits = 3, eps = 0.001))
  
  # Two blank strings before the value moves it down two rows
  c("", "", formatted_p)
}

rndr <- function(x, name, ...) {
  if (!is.numeric(x)) return(render.categorical.default(x))
  
  # Choose rendering style based on variable name
  if (name == "outbreak_count") {
    what <- "Sum (Max)"
  } else if (name %in% c(
    "temperature_C", "shortwave_radiation", "longwave_radiation",
    "relative_humidity", "wind_speed", "soil_moisture",
    "runoff_mm", "precipitation_mm", "snow_density",
    "leaf_area_index", "anseriformes", "predators"
  )) {
    what <- "Mean (SD) <br/>Median [Min, Max]"
  } else {
    what <- "Mean (SD)"
  }
  
  parse.abbrev.render.code(c("",what))(x)
}


# ============================================================
# Variable labels

label(fishnet_t1$outbreak_count)        <- "Outbreak Count"
label(fishnet_t1$temperature_C)        <- "Temperature (°C)"
label(fishnet_t1$shortwave_radiation)  <- "Shortwave Radiation (MJ/m²)"
label(fishnet_t1$longwave_radiation)   <- "Longwave Radiation (MJ/m²)"
label(fishnet_t1$relative_humidity)    <- "Relative Humidity (%)"
label(fishnet_t1$wind_speed)           <- "Wind Speed (m/s)"
label(fishnet_t1$soil_moisture)        <- "Soil Moisture (m³/m³)"
label(fishnet_t1$runoff_mm)            <- "Surface & Sub-Surface Runoff (mm)"
label(fishnet_t1$precipitation_mm)     <- "Precipitation (mm)"
label(fishnet_t1$snow_density)         <- "Snow Density (kg/m³)"
label(fishnet_t1$leaf_area_index)      <- "Leaf Area Index"
label(fishnet_t1$anseriformes)         <- "Anseriformes Abundance (%)"
label(fishnet_t1$predators)            <- "Predator Abundance (%)"
label(fishnet_t1$land_cover_label)     <- "Land Cover Type N (%)"

# Generate Table 1 

table_1 <- table1::table1(
  ~ outbreak_count + temperature_C + shortwave_radiation + longwave_radiation +
    relative_humidity + wind_speed +
    soil_moisture + runoff_mm + precipitation_mm + snow_density + leaf_area_index +
    anseriformes + predators + land_cover_label
  | any_outbreak,
  data = fishnet_t1,
  render=rndr,
  render.categorical = my.render.cat,
  render.missing = NULL,
  overall = FALSE,
  extra.col = list(`P-value` = pvalue)
)

table_1
```



# Table 1 but stratified by season


```{r, table 1 stratified by season}
# 
# # Subset to outbreak cells only
# fishnet_outbreak <- fishnet_t1 %>%
#   filter(any_outbreak == "Cells with an Outbreak") %>%
#   mutate(season = factor(droplevels(season)))
# 
# # Generate Table 1 stratified by season
# table_season <- table1::table1(
#   ~ outbreak_count + temperature_C + shortwave_radiation + longwave_radiation +
#     relative_humidity + wind_speed +
#     soil_moisture + runoff_mm + precipitation_mm + snow_density +
#     leaf_area_index + anseriformes + predators
#   | season,                    # stratify by season
#   data = fishnet_outbreak,
#   render = rndr,
#   render.categorical = my.render.cat,
#   render.missing = NULL,
#   overall = FALSE)
# 
# table_season
# 
# 



```




# Table 1 for CLimate Anomalies 


```{r, table 1 for climate anomalies}

# ============================================================
# 25-year Z-score anomaly variables
# ============================================================

anomaly_vars <- c(
  "leaf_area_index_z_25yr",
  "longwave_radiation_z_25yr",
  "precipitation_z_25yr",
  "relative_humidity_z_25yr",
  "runoff_z_25yr",
  "shortwave_radiation_z_25yr",
  "snow_density_z_25yr",
  "soil_moisture_z_25yr",
  "temperature_z_25yr",
  "wind_speed_z_25yr"
)

# ============================================================
# Variable labels (for clean table output)
# ============================================================

label(fishnet_t1$leaf_area_index_z_25yr)     <- "Leaf Area Index (Z-score)"
label(fishnet_t1$longwave_radiation_z_25yr)  <- "Longwave Radiation (Z-score)"
label(fishnet_t1$precipitation_z_25yr)       <- "Precipitation (Z-score)"
label(fishnet_t1$relative_humidity_z_25yr)   <- "Relative Humidity (Z-score)"
label(fishnet_t1$runoff_z_25yr)              <- "Surface & Sub-Surface Runoff (Z-score)"
label(fishnet_t1$shortwave_radiation_z_25yr) <- "Shortwave Radiation (Z-score)"
label(fishnet_t1$snow_density_z_25yr)        <- "Snow Density (Z-score)"
label(fishnet_t1$soil_moisture_z_25yr)       <- "Soil Moisture (Z-score)"
label(fishnet_t1$temperature_z_25yr)         <- "Temperature (Z-score)"
label(fishnet_t1$wind_speed_z_25yr)          <- "Wind Speed (Z-score)"


# ============================================================
# Renderer for Z-score anomalies
# ============================================================

rndr_anom <- function(x, name, ...) {
  if (!is.numeric(x)) return(render.categorical.default(x))
  
  # For anomalies, only show Mean (SD)
  what <- "Mean (SD)"
  parse.abbrev.render.code(c("", what))(x)
}



# Generate anomaly table

table_anom <- table1::table1(
  ~ outbreak_count + temperature_z_25yr + shortwave_radiation_z_25yr + longwave_radiation_z_25yr +
    relative_humidity_z_25yr + wind_speed_z_25yr +
    soil_moisture_z_25yr + runoff_z_25yr + precipitation_z_25yr + snow_density_z_25yr +
    leaf_area_index_z_25yr | any_outbreak,
  data = fishnet_t1,
  render = rndr_anom,
  render.categorical = my.render.cat,
  render.missing = NULL,
  overall = FALSE,
  extra.col = list(`P-value` = pvalue)
)

table_anom




```

## Primary Table 1 Export

```{r, primary table 1 export}


# Table 1: Outbreak vs No Outbreak

table_1_gt <- fishnet_t1 %>%
  select(
    outbreak_count,
    temperature_C, shortwave_radiation, longwave_radiation,
    relative_humidity, wind_speed, soil_moisture,
    runoff_mm, precipitation_mm, snow_density,
    leaf_area_index, anseriformes, predators,
    land_cover_label, any_outbreak
  ) %>%
  tbl_summary(
    by = any_outbreak,
    type = list(
      all_continuous() ~ "continuous2",
      land_cover_label ~ "categorical"
    ),
    statistic = list(
      all_continuous() ~ "{mean} [{min}, {max}]",
      outbreak_count ~ "{sum} ({max})",
      all_categorical() ~ "{n} ({p}%)"
    ),
    # >>> Force 2 decimals for continuous mean/min/max <<<
    digits = list(
      all_continuous() ~ c(mean = 1, min = 1, max = 1)
    ),
    label = list(
      outbreak_count ~ "Outbreak Count - N(%)",
      temperature_C ~ "Temperature (°C)",
      shortwave_radiation ~ "Shortwave Radiation (MJ/m²)",
      longwave_radiation ~ "Longwave Radiation (MJ/m²)",
      relative_humidity ~ "Relative Humidity (%)",
      wind_speed ~ "Wind Speed (m/s)",
      soil_moisture ~ "Soil Moisture (m³/m³)",
      runoff_mm ~ "Surface & Sub-Surface Runoff (mm)",
      precipitation_mm ~ "Precipitation (mm)",
      snow_density ~ "Snow Density (kg/m³)",
      leaf_area_index ~ "Leaf Area Index",
      anseriformes ~ "Anseriformes Abundance (N)",
      predators ~ "Predator Abundance (N)",
      land_cover_label ~ "Land Cover Type (%)"
    ),
    missing = "no"
  ) %>%
  add_p(
    test = list(
      all_continuous() ~ "t.test",
      all_categorical() ~ "chisq.test"
    )
  ) %>%
  modify_header(label ~ "**Variable**") %>%
  modify_caption(" ") %>%  
  bold_labels()


# ------------------------------------------------------------
# Convert to flextable, set font + alignment
# ------------------------------------------------------------
ft <- as_flex_table(table_1_gt) %>%
  font(fontname = "Arial", part = "all") %>%
  fontsize(size = 11, part = "all") %>%
  align(align = "left", j = 1, part = "all") %>%
  align(align = "center", j = 2:ncol(as_flex_table(table_1_gt)$body$dataset), part = "all") %>%
  # ---- Tighten layout ----
  set_table_properties(layout = "autofit", width = .8) %>%     # make table fit page width
  set_table_properties(layout = "fixed") %>%                   # tighter fixed layout
  width(width = 0.9) %>%                                       # shrink columns slightly
    line_spacing(space = 0.8, part = "all") %>%                  # tighten line spacing

  autofit()


# ------------------------------------------------------------
# Export to Word
# ------------------------------------------------------------
save_as_docx(ft, path = file.path(tables_folder, "Table1_Primary.docx"))

```

## Season Stratification Table 1 Export

```{r, season strat table 1 export}

# ------------------------------------------------------------
# Subset: Only outbreak cells
# ------------------------------------------------------------
fishnet_outbreak <- fishnet_t1 %>%
  filter(any_outbreak == "Cells with an outbreak") %>%
  mutate(season = droplevels(factor(season)))

# ------------------------------------------------------------
# Table 2: Outbreak Cells Stratified by Season
# ------------------------------------------------------------
table_season_gt <- fishnet_outbreak %>%
  select(
    outbreak_count,
    temperature_C, shortwave_radiation, longwave_radiation,
    relative_humidity, wind_speed, soil_moisture,
    runoff_mm, precipitation_mm, snow_density,
    leaf_area_index, anseriformes, predators,
    season
  ) %>%
 tbl_summary(
    by = season,
    type = list(
      all_continuous() ~ "continuous2"    ),
    statistic = list(
      all_continuous() ~ "{mean} [{min}, {max}]",
      outbreak_count ~ "{sum} ({max})",
      all_categorical() ~ "{n} ({p}%)"
    ),
    # >>> Force 2 decimals for continuous mean/min/max <<<
    digits = list(
      all_continuous() ~ c(mean = 1, min = 1, max = 1)
    ),
    label = list(
      outbreak_count ~ "Outbreak Count - N(%)",
      temperature_C ~ "Temperature (°C)",
      shortwave_radiation ~ "Shortwave Radiation (MJ/m²)",
      longwave_radiation ~ "Longwave Radiation (MJ/m²)",
      relative_humidity ~ "Relative Humidity (%)",
      wind_speed ~ "Wind Speed (m/s)",
      soil_moisture ~ "Soil Moisture (m³/m³)",
      runoff_mm ~ "Surface & Sub-Surface Runoff (mm)",
      precipitation_mm ~ "Precipitation (mm)",
      snow_density ~ "Snow Density (kg/m³)",
      leaf_area_index ~ "Leaf Area Index",
      anseriformes ~ "Anseriformes Abundance (N)",
      predators ~ "Predator Abundance (N)"    
      ),
    missing = "no"
  ) %>%
  modify_header(label ~ "**Variable**") %>%
  modify_caption(" ") %>%
  bold_labels()

# ------------------------------------------------------------
# Convert to flextable (Arial 11 pt, same styling)
# ------------------------------------------------------------
ft_season <- as_flex_table(table_season_gt) %>%
  font(fontname = "Arial", part = "all") %>%
  fontsize(size = 11, part = "all") %>%
  align(align = "left", j = 1, part = "all") %>%
  align(align = "center", j = 2:ncol(as_flex_table(table_season_gt)$body$dataset), part = "all") %>%
  set_table_properties(layout = "autofit", width = .8) %>%
  set_table_properties(layout = "fixed") %>%
  width(width = 0.9) %>%
  line_spacing(space = 0.8, part = "all") %>%
  autofit()

# ------------------------------------------------------------
# Export to Word
# ------------------------------------------------------------
save_as_docx(ft_season, path = file.path(tables_folder, "Table1_Season.docx"))




```

# Season stratifications comped to 25-year mean

```{r, season stratifications comped to 25-year mean}

#  Transform the 25-yr mean columns to match current-variable units

fishnet_outbreak <- fishnet_outbreak %>%
  mutate(

    temperature_mean_25yr_C            = temperature_mean_25yr - 273.15,
    precipitation_mean_25yr_mm         = precipitation_mean_25yr * 1000,
    runoff_mean_25yr_mm                = runoff_mean_25yr * 1000,
    longwave_radiation_mean_25yr_MJm2  = longwave_radiation_mean_25yr / 1e6,
    shortwave_radiation_mean_25yr_MJm2 = shortwave_radiation_mean_25yr / 1e6

  )


# Map current vars -> transformed 25-yr counterparts

compare_map <- c(
  temperature_C        = "temperature_mean_25yr_C",
  shortwave_radiation  = "shortwave_radiation_mean_25yr_MJm2",
  longwave_radiation   = "longwave_radiation_mean_25yr_MJm2",
  relative_humidity    = "relative_humidity_mean_25yr",
  wind_speed           = "wind_speed_mean_25yr",
  soil_moisture        = "soil_moisture_mean_25yr",
  runoff_mm            = "runoff_mean_25yr_mm",
  precipitation_mm     = "precipitation_mean_25yr_mm",
  snow_density         = "snow_density_mean_25yr",
  leaf_area_index      = "leaf_area_index_mean_25yr"
)

# Nice labels for the table
var_labels <- c(
  temperature_C       = "Temperature (°C)",
  shortwave_radiation = "Shortwave Radiation (MJ/m²)",
  longwave_radiation  = "Longwave Radiation (MJ/m²)",
  relative_humidity   = "Relative Humidity (%)",
  wind_speed          = "Wind Speed (m/s)",
  soil_moisture       = "Soil Moisture (m³/m³)",
  runoff_mm           = "Surface & Sub-Surface Runoff (mm)",
  precipitation_mm    = "Precipitation (mm)",
  snow_density        = "Snow Density (kg/m³)",
  leaf_area_index     = "Leaf Area Index"
)


# Build a tidy season-comparison table:
compare_df <- lapply(names(compare_map), function(cur_var) {
  ref_var <- compare_map[[cur_var]]

  fishnet_outbreak %>%
    group_by(season) %>%
    summarise(
      current_mean = mean(.data[[cur_var]], na.rm = TRUE),
      ref25_mean   = mean(.data[[ref_var]], na.rm = TRUE),
      .groups = "drop"
    ) %>%
    mutate(
      variable   = cur_var,
      delta_mean = current_mean - ref25_mean
    )
}) %>%
  bind_rows() %>%
  mutate(
    Variable = var_labels[variable],
    # format to 1 decimal
    Current_Mean = sprintf("%.1f", current_mean),
    Mean_25yr    = sprintf("%.1f", ref25_mean),
    Delta        = sprintf("%.1f", delta_mean)
  ) %>%
  select(Variable, season, Current_Mean, Mean_25yr, Delta) %>%
  arrange(match(Variable, unname(var_labels[names(compare_map)])),
          season)


# Wider by season 
compare_wide <- compare_df %>%
  pivot_longer(cols = c(Current_Mean, Mean_25yr, Delta),
               names_to = "stat", values_to = "value") %>%
  unite("Season_Stat", season, stat, sep = " — ") %>%
  pivot_wider(names_from = Season_Stat, values_from = value) %>%
  arrange(match(Variable, unname(var_labels[names(compare_map)])))

ft_compare_wide <- flextable(compare_wide) %>%
  font(fontname = "Arial", part = "all") %>%
  fontsize(size = 11, part = "all") %>%
  align(align = "center", part = "all") %>%
  align(j = "Variable", align = "left", part = "all") %>%
  autofit()


# 5) Export to Word 
save_as_docx(
  "Table 2b. Outbreak Cells: Current vs 25-yr Means by Season" = ft_compare_wide,
  path = file.path(tables_folder, "Table2b_Season_vs_25yr.docx")
)


```

# Season stratification with model comparing absolute and 25yr mean

## Prepare model

```{r, prepare for modeling difference between absolute and 25yr mean}


# Convert 25-yr means to match absolute-variable units

fishnet_outbreak <- fishnet_outbreak %>%
  mutate(
    # Kelvin -> °C
    temperature_mean_25yr_C = temperature_mean_25yr - 273.15,

    # meters -> mm
    precipitation_mean_25yr_mm = precipitation_mean_25yr * 1000,
    runoff_mean_25yr_mm        = runoff_mean_25yr        * 1000,

    # W/m^2 -> MJ/m^2  (per your note: divide by 1e6)
    longwave_radiation_mean_25yr_MJm2  = longwave_radiation_mean_25yr  / 1e6,
    shortwave_radiation_mean_25yr_MJm2 = shortwave_radiation_mean_25yr / 1e6
  )


# Variable lists / mapping
vars_abs <- c(
  "temperature_C",
  "shortwave_radiation",
  "longwave_radiation",
  "relative_humidity",
  "wind_speed",
  "soil_moisture",
  "runoff_mm",
  "precipitation_mm",
  "snow_density",
  "leaf_area_index"
)

means_map <- c(
  temperature_C       = "temperature_mean_25yr_C",
  shortwave_radiation = "shortwave_radiation_mean_25yr_MJm2",
  longwave_radiation  = "longwave_radiation_mean_25yr_MJm2",
  relative_humidity   = "relative_humidity_mean_25yr",
  wind_speed          = "wind_speed_mean_25yr",
  soil_moisture       = "soil_moisture_mean_25yr",
  runoff_mm           = "runoff_mean_25yr_mm",
  precipitation_mm    = "precipitation_mean_25yr_mm",
  snow_density        = "snow_density_mean_25yr",
  leaf_area_index     = "leaf_area_index_mean_25yr"
)

missing_means <- setdiff(unname(means_map), names(fishnet_outbreak))
if (length(missing_means)) {
  stop("Missing 25yr mean columns after conversion: ",
       paste(missing_means, collapse = ", "))
}

nice_var <- function(x) x %>% str_replace_all("_", " ") %>% str_to_title()


```

## Now model...

```{r, modeling difference between absolute and 25yr mean}

emm_options(lmer.df = "asymptotic")


fit_anomaly_once <- function(dat, abs_name, mean_name) {
  stopifnot("week" %in% names(dat))
  dd <- dat %>%
    transmute(
      zone_id, season, week,
      .abs    = .data[[abs_name]],
      .mean25 = .data[[mean_name]],
      .anom   = .abs - .mean25
    )

  fit <- lmerTest::lmer(
    .anom ~ season + (1|zone_id) + (1|week),
    data = dd, REML = TRUE,
    control = lme4::lmerControl(check.conv.singular = "ignore")
  )

  emm <- emmeans::emmeans(fit, ~ season)
  broom::tidy(emm, conf.int = TRUE) %>%
    transmute(
      variable = abs_name,
      season   = as.character(season),
      estimate, conf.low, conf.high, p.value
    )
}

emm_all <- purrr::map_dfr(
  vars_abs,
  ~fit_anomaly_once(fishnet_outbreak, .x, means_map[[.x]])
)

emm_all <- emm_all %>%
  group_by(season) %>%
  mutate(p.adj = p.adjust(p.value, method = "BH")) %>%
  ungroup() %>%
  mutate(
    sig = case_when(
      p.adj < 0.001 ~ "***",
      p.adj < 0.01  ~ "**",
      p.adj < 0.05  ~ "*",
      TRUE          ~ ""
    ),
    delta_ci = sprintf("%+.2f [%.2f, %.2f]%s", estimate, conf.low, conf.high, sig)
  )

delta_wide <- emm_all %>%
  select(variable, season, delta_ci) %>%
  pivot_wider(names_from = season, values_from = delta_ci) %>%
  mutate(Row = "Δ vs 25-yr (95% CI)") %>%
  relocate(Row)

```

## Produce results

```{r, produce results from model on difference between absolute and 25yr mean}


label_map <- c(
  temperature_C       = "Temperature (°C)",
  shortwave_radiation = "Shortwave Radiation (MJ/m²)",
  longwave_radiation  = "Longwave Radiation (MJ/m²)",
  relative_humidity   = "Relative Humidity (%)",
  wind_speed          = "Wind Speed (m/s)",
  soil_moisture       = "Soil Moisture (m³/m³)",
  runoff_mm           = "Surface & Sub-Surface Runoff (mm)",
  precipitation_mm    = "Precipitation (mm)",
  snow_density        = "Snow Density (kg/m³)",
  leaf_area_index     = "Leaf Area Index"
)

means_long <- purrr::map_dfr(vars_abs, function(v) {
  fishnet_outbreak %>%
    transmute(
      variable = v,
      season   = as.character(season),
      value    = .data[[v]]
    ) %>%
    group_by(variable, season) %>%
    summarise(mean_2022 = mean(value, na.rm = TRUE), .groups = "drop")
})

means_wide <- means_long %>%
  mutate(mean_fmt = sprintf("%.2f", mean_2022)) %>%
  select(variable, season, mean_fmt) %>%
  pivot_wider(names_from = season, values_from = mean_fmt) %>%
  mutate(Row = "Mean (2022)") %>%
  relocate(Row)

tbl_mat <- bind_rows(means_wide, delta_wide) %>%
  arrange(factor(variable, levels = vars_abs), match(Row, c("Mean (2022)", "Δ vs 25-yr (95% CI)"))) %>%
  mutate(Variable = ifelse(variable %in% names(label_map), label_map[variable], variable)) %>%
  select(Variable, Row, Winter, Spring, Summer, Fall)

ft_tbl <- flextable(tbl_mat) |>
  set_header_labels(
    Variable = "Variable",
    Row = "",
    Winter = "Winter",
    Spring = "Spring",
    Summer = "Summer",
    Fall = "Fall"
  ) |>
  add_footer_row(values = "Row 1: mean absolute 2022 value; Row 2: Δ vs 25-yr mean [95% CI], BH-adjusted (* p<0.05, ** p<0.01, *** p<0.001)", colwidths = 6) |>
  bold(i = ~ Row == "Mean (2022)", bold = TRUE, part = "body") |>
  color(i = ~ Row == "Δ vs 25-yr (95% CI)", color = "#666666", part = "body") |>
  align(j = 2:6, align = "center", part = "all") |>
  align(j = 1, align = "left", part = "all") |>
  font(fontname = "Arial", part = "all") |>
  fontsize(size = 11, part = "all") |>
  line_spacing(space = 0.8, part = "all") |>
  autofit()

# preview in viewer
ft_tbl

# export
save_as_docx(ft_tbl, path = file.path(tables_folder, "Table_Season_w25yr_comp.docx"))


```


# Quick Statistics 

## How many case and control cells?

```{r, quick statistics}

vars <- c(
  "leaf_area_index","longwave_radiation","precipitation","relative_humidity",
  "runoff","shortwave_radiation","snow_density","soil_moisture","temperature",
  "anseriformes","predators","wind_speed",
  "any_outbreak",          # binary (0/1)
  "outbreak_count"         # numeric count per row
)

zone_summary <- fishnet_subset %>%
  group_by(zone_id) %>%
  summarise(
    # Means & SDs for all listed variables
    across(all_of(vars),
           list(mean = ~mean(.x, na.rm = TRUE),
                sd   = ~sd(.x,   na.rm = TRUE)),
           .names = "{.col}_{.fn}"),

    # Zone-level outbreak summaries
    any_outbreak_any    = as.integer(any(any_outbreak == 1, na.rm = TRUE)), # ever had 1?
    outbreak_count_total = sum(outbreak_count, na.rm = TRUE),               # total across time

    n_obs = n(),
    .groups = "drop"
  )

zone_summary

zone_outbreak_counts <- zone_summary %>%
  count(any_outbreak_any, name = "n_zone_ids") %>%
  mutate(any_outbreak_any = ifelse(any_outbreak_any == 1, "has_outbreak", "no_outbreak"))

zone_outbreak_counts

## 77 have outbreak and 2,065 do not
```
