---
title: "Bird Order Sensitivity Analysis"
author: "Fintan Mooney"
date: "2025-10-15"
output: html_document
---

# Sensitivity analysis replacing total bird abundance with order-level abundance (e.g., Anseriformes, Charadriiformes)... testing whether specific bird orders drive HPAI outbreak risk differently!

```{r include=FALSE}
rm(list=ls())
project.folder = paste0(print(here::here()),'/')
source(paste0(project.folder,'create_folder_structure.R'))
source(paste0(functions.folder,'script_initiate.R'))
```

## Read in casecrossover dataset that has the 56-day lag window

```{r, read in casecrossover dataset with 56-day lag window}
# 56-day (8-week) lag window is appropriate for bird-order analysis because
# Anseriformes migration windows are longer than the 28-day primary lag span.
case_crossover_df_56lag <- readRDS(file.path(objects_folder, "case_crossover_df_56lag.RDS"))
```

## Crossbasis

## Analyzing cumulative and lag specific odds ratios from case-crossover with 8 lags 

```{r, 8 lag test}

## Crossbasis for 8-lag

# Finalized Crossbasis Terms for Climate and Bird Predictors

# Leaf Area Index
cb_leaf_area_index <- crossbasis(
  x = as.matrix(case_crossover_df_56lag[, paste0("leaf_area_index_Lag", 0:56)]),
  lag = c(0, 56),
  argvar = list(fun = "lin"),
  arglag = list(fun = "ns", df = 6)
)

# Longwave Radiation
cb_longwave_radiation <- crossbasis(
  x = as.matrix(case_crossover_df_56lag[, paste0("longwave_radiation_Lag", 0:56)]),
  lag = c(0, 56),
  argvar = list(fun = "lin"),
  arglag = list(fun = "ns", df = 6)
)

# Precipitation
cb_precipitation <- crossbasis(
  x = as.matrix(case_crossover_df_56lag[, paste0("precipitation_Lag", 0:56)]),
  lag = c(0, 56),
  argvar = list(fun = "lin"),
  arglag = list(fun = "ns", df = 6)
)

# Relative Humidity
cb_relative_humidity <- crossbasis(
  x = as.matrix(case_crossover_df_56lag[, paste0("relative_humidity_Lag", 0:56)]),
  lag = c(0, 56),
  argvar = list(fun = "lin"),
  arglag = list(fun = "ns", df = 6)
)

# Runoff
cb_runoff <- crossbasis(
  x = as.matrix(case_crossover_df_56lag[, paste0("runoff_Lag", 0:56)]),
  lag = c(0, 56),
  argvar = list(fun = "lin"),
  arglag = list(fun = "ns", df = 6)
)

# Shortwave Radiation
cb_shortwave_radiation <- crossbasis(
  x = as.matrix(case_crossover_df_56lag[, paste0("shortwave_radiation_Lag", 0:56)]),
  lag = c(0, 56),
  argvar = list(fun = "lin"),
  arglag = list(fun = "ns", df = 6)
)

# Snow Density
cb_snow_density <- crossbasis(
  x = as.matrix(case_crossover_df_56lag[, paste0("snow_density_Lag", 0:56)]),
  lag = c(0, 56),
  argvar = list(fun = "lin"),
  arglag = list(fun = "ns", df = 6)
)

# Soil Moisture
cb_soil_moisture <- crossbasis(
  x = as.matrix(case_crossover_df_56lag[, paste0("soil_moisture_Lag", 0:56)]),
  lag = c(0, 56),
  argvar = list(fun = "lin"),
  arglag = list(fun = "ns", df = 6)
)

# Temperature
cb_temperature <- crossbasis(
  x = as.matrix(case_crossover_df_56lag[, paste0("temperature_Lag", 0:56)]),
  lag = c(0, 56),
  argvar = list(fun = "lin"),
  arglag = list(fun = "ns", df = 6)
)


# Total Bird Abundance
cb_total_bird_abundance <- crossbasis(
  x = as.matrix(case_crossover_df_56lag[, paste0("total_bird_abundance_Lag", 0:56)]),
  lag = c(0, 56),
  argvar = list(fun = "lin"),
  arglag = list(fun = "ns", df = 6)
)

# Wind Speed
cb_wind_speed <- crossbasis(
  x = as.matrix(case_crossover_df_56lag[, paste0("wind_speed_Lag", 0:56)]),
  lag = c(0, 56),
  argvar = list(fun = "lin"),
  arglag = list(fun = "ns", df = 6)
)



## BIRD ORDER CROSSBASIS - WHAT WE ARE TESTING

### Poly

#### Anseriformes:
cb_anseri <- crossbasis(
    x = as.matrix(case_crossover_df_56lag[, paste0("anseriformes_Lag", 0:56)]),
    lag = c(0, 56),
    argvar = list(fun = "lin"),
  arglag = list(fun = "poly", degree = 2)
  )
 
#### Predator:
cb_predators <- crossbasis(
x = as.matrix(case_crossover_df_56lag[, paste0("predators_Lag", 0:56)]),
  lag = c(0, 56),
  argvar = list(fun = "lin"),
  arglag = list(fun = "poly", degree = 2)
)

## Natural Spline

#### Anseriformes:
cb_anseri_ns <- crossbasis(
    x = as.matrix(case_crossover_df_56lag[, paste0("anseriformes_Lag", 0:56)]),
    lag = c(0, 56),
    argvar = list(fun = "lin"),
  arglag = list(fun = "ns", df = 6)
  )
 
#### Predator:
cb_predators_ns <- crossbasis(
x = as.matrix(case_crossover_df_56lag[, paste0("predators_Lag", 0:56)]),
  lag = c(0, 56),
  argvar = list(fun = "lin"),
  arglag = list(fun = "ns", df = 6)
)


```


# Fit Models

```{r, fit models}

## Adjusted Models
model_8lag_poly <- clogit(
  outbreak_binary ~ 
    cb_anseri + cb_predators + cb_soil_moisture + cb_temperature +
    cb_precipitation + cb_wind_speed + cb_total_bird_abundance +
    strata(stratum_id),
  method = "efron",
  data = case_crossover_df_56lag
)

model_8lag_ns <- clogit(
  outbreak_binary ~ 
    cb_anseri_ns + cb_predators_ns + cb_soil_moisture + cb_temperature +
    cb_precipitation + cb_wind_speed + cb_total_bird_abundance +
    strata(stratum_id),
  method = "efron",
  data = case_crossover_df_56lag
)

AIC(model_8lag_poly, model_8lag_ns)

## NS has improved fit, slightly 

## Crude Models

model_8lag_crude_poly <- clogit(
  outbreak_binary ~ 
    cb_anseri + cb_predators + 
    strata(stratum_id),
  method = "efron",
  data = case_crossover_df_56lag
)

model_8lag_crude_ns <- clogit(
  outbreak_binary ~ 
    cb_anseri_ns + cb_predators_ns + 
    strata(stratum_id),
  method = "efron",
  data = case_crossover_df_56lag
)

AIC(model_8lag_crude_poly, model_8lag_crude_ns)

## Polynomial wins
```




## Report Results of the sensitivity analysis on stratification of total bird abundance by predators and anseriforms

```{r, crosspred objects}

# --- Crude model
cp_anseri_crude    <- crosspred(cb_anseri,    model_8lag_crude_poly,    cen = 0, at = 0.5, cumul = TRUE)
cp_predator_crude  <- crosspred(cb_predators, model_8lag_crude_poly,    cen = 0, at = 0.5, cumul = TRUE)

cp_slice_anseri_crude    <- crosspred(cb_anseri,    model_8lag_crude_poly,    cen = 0, at = 0.5)
cp_slice_predator_crude  <- crosspred(cb_predators, model_8lag_crude_poly,    cen = 0, at = 0.5)

# --- Adjusted model
cp_anseri_adj      <- crosspred(cb_anseri,    model_8lag_crude_poly, cen = 0, at = 0.5, cumul = TRUE)
cp_predator_adj    <- crosspred(cb_predators, model_8lag_crude_poly, cen = 0, at = 0.5, cumul = TRUE)

cp_slice_anseri_adj    <- crosspred(cb_anseri,    model_8lag_crude_poly, cen = 0, at = 0.5)
cp_slice_predator_adj  <- crosspred(cb_predators, model_8lag_crude_poly, cen = 0, at = 0.5)


```

```{r, cumulative OR summary table}
# -- Cumulative OR Summary Table --

cumulative_summary <- tibble(
  Predictor = c("Anseriformes (Crude)", "Predators (Crude)",
                "Anseriformes (Adjusted)", "Predators (Adjusted)"),
  OR = c(
    cp_anseri_crude$allRRfit["0.5"],
    cp_predator_crude$allRRfit["0.5"],
    cp_anseri_adj$allRRfit["0.5"],
    cp_predator_adj$allRRfit["0.5"]
  ),
  CI_low = c(
    cp_anseri_crude$allRRlow["0.5"],
    cp_predator_crude$allRRlow["0.5"],
    cp_anseri_adj$allRRlow["0.5"],
    cp_predator_adj$allRRlow["0.5"]
  ),
  CI_high = c(
    cp_anseri_crude$allRRhigh["0.5"],
    cp_predator_crude$allRRhigh["0.5"],
    cp_anseri_adj$allRRhigh["0.5"],
    cp_predator_adj$allRRhigh["0.5"]
  )
) %>%
  mutate(`OR (95% CI)` = sprintf("%.2f (%.2f, %.2f)", OR, CI_low, CI_high)) %>%
  select(Predictor, `OR (95% CI)`) %>%
  gt() %>%
  tab_header(
    title = md("**Case-Crossover Models (8-lag): Cumulative Odds Ratios**"),
    subtitle = "Per 0.5 SD Increase (Crude vs Adjusted)"
  ) %>%
  cols_label(`OR (95% CI)` = "Odds Ratio (95% Confidence Interval)") %>%
  tab_options(
    table.font.size = 14,
    heading.title.font.size = 16,
    heading.subtitle.font.size = 14,
    data_row.padding = px(6)
  )

cumulative_summary

```

```{r, cumlative and slices plots}

# -- Cumulative and Slices Plots (Crude + Adjusted) --

plot_cumulative <- function(cp, title) {
  plot(cp, type = "cumulative", var = 0.5, ci = "area",
       col = "black", ci.arg = list(col = rgb(0,0,0,0.1), border = NA),
       lwd = 1.7, main = title, xlab = "", ylab = "", las = 1)
  abline(h = 1, lty = 2, col = "gray40", lwd = 1)
}

plot_slices <- function(cp, title) {
  plot(cp, type = "slices", var = 0.5, ci = "area",
       col = "black", ci.arg = list(col = rgb(0,0,0,0.1), border = NA),
       lwd = 1.3, main = title, xlab = "", ylab = "", las = 1)
  abline(h = 1, lty = 2, col = "gray40", lwd = 1)
}

# ----------------------------
# Save Cumulative Plot
# ----------------------------
png(file.path(figures_sensitivity_folder, "figure_cumulative_ORs_anseri_predators.png"),
    width = 9, height = 6, units = "in", res = 300)

par(
  mfrow = c(2, 2),
  mar = c(4.2, 4.2, 2.5, 1),
  oma = c(3, 4, 0, 1),   
  cex.main = 1.2,
  cex.lab = 1.2,
  cex.axis = 1.1,
  tcl = -0.3, mgp = c(2.5, 0.7, 0)
)

plot_cumulative(cp_anseri_crude,   "Anseriformes (Crude)")
plot_cumulative(cp_predator_crude, "Predators (Crude)")
plot_cumulative(cp_anseri_adj,     "Anseriformes (Adjusted)")
plot_cumulative(cp_predator_adj,   "Predators (Adjusted)")

mtext("Lag (days)", side = 1, outer = TRUE, line = 0, cex = 1.2)
mtext("Odds Ratio (95% CI)", side = 2, outer = TRUE, line = 0, cex = 1.2)

dev.off()

# ----------------------------
# Save Slices Plot
# ----------------------------
png(file.path(figures_sensitivity_folder, "figure_slice_ORs_anseri_predators.png"),
    width = 9, height = 6, units = "in", res = 300)

par(
  mfrow = c(2, 2),
  mar = c(4.2, 4.2, 2.5, 1),
  oma = c(3, 4, 0, 1),   
  cex.main = 1.2,
  cex.lab = 1.2,
  cex.axis = 1.1,
  tcl = -0.3, mgp = c(2.5, 0.7, 0)
)

plot_slices(cp_slice_anseri_crude,   "Anseriformes (Crude)")
plot_slices(cp_slice_predator_crude, "Predators (Crude)")
plot_slices(cp_slice_anseri_adj,     "Anseriformes (Adjusted)")
plot_slices(cp_slice_predator_adj,   "Predators (Adjusted)")

mtext("Lag (days)", side = 1, outer = TRUE, line = 0, cex = 1.2)
mtext("Odds Ratio (95% CI)", side = 2, outer = TRUE, line = 0, cex = 1.2)

dev.off()


```

## Just the unadjusted plots now

```{r, unadjusted plots}

plot_cumulative <- function(cp, title, y_min = 0, y_max = 7) {
  plot(cp, type = "cumulative", var = 0.5, ci = "area",
       col = "black", ci.arg = list(col = rgb(0,0,0,0.1), border = NA),
       lwd = 1.7, main = title, xlab = "", ylab = "", las = 1,
       ylim = c(y_min, y_max), yaxt = "n")
  axis(2, at = seq(y_min, y_max, by = 1), las = 1)
  abline(h = 1, lty = 2, col = "gray40", lwd = 1)
}

plot_slices <- function(cp, title, y_min = 0, y_max = 7) {
  plot(cp, type = "slices", var = 0.5, ci = "area",
       col = "black", ci.arg = list(col = rgb(0,0,0,0.1), border = NA),
       lwd = 1.3, main = title, xlab = "", ylab = "", las = 1,
       ylim = c(y_min, y_max), yaxt = "n")
  axis(2, at = seq(y_min, y_max, by = 1), las = 1)
  abline(h = 1, lty = 2, col = "gray40", lwd = 1)
}


png(file.path(figures_sensitivity_folder, "figure_unadjusted_slice_cumulative_anseri_predators.png"),
    width = 10, height = 8, units = "in", res = 300)

par(
  mfrow = c(2, 2),            # 2 rows (Anseriformes, Predators) x 2 cols (Slice, Cumulative)
  mar = c(4.2, 4.2, 2.5, 1),
  oma = c(3, 4, 0, 1),
  cex.main = 1.2,
  cex.lab = 1.2,
  cex.axis = 1.1,
  tcl = -0.3, mgp = c(2.5, 0.7, 0)
)

# Row 1: Anseriformes (UNADJUSTED)
plot_slices(   cp_slice_anseri_crude,   "Anseriformes — Lag Specific")
plot_cumulative(cp_anseri_crude,        "Anseriformes — Cumulative")

# Row 2: Predators (UNADJUSTED)
plot_slices(   cp_slice_predator_crude, "Predators — Lag-Specific")
plot_cumulative(cp_predator_crude,      "Predators — Cumulative")

mtext("Lag (days)", side = 1, outer = TRUE, line = 0, cex = 1.2)
mtext("Odds Ratio (95% CI)", side = 2, outer = TRUE, line = 0, cex = 1.2)

dev.off()

```

```{r, lag specific odds ratio table}

# -- Lag-Specific Odds Ratio Table (0–8 weeks) --

extract_lag_ORs_wide <- function(cp_obj, varname, varval = "0.5", lags = 0:8) {
  lag_labels <- paste0("lag", lags)
  or   <- cp_obj$matRRfit[varval, lag_labels]
  low  <- cp_obj$matRRlow[varval, lag_labels]
  high <- cp_obj$matRRhigh[varval, lag_labels]
  tibble(
    Predictor = varname,
    !!!setNames(as.list(or),   paste0("OR_Lag", lags)),
    !!!setNames(as.list(low),  paste0("Low_Lag", lags)),
    !!!setNames(as.list(high), paste0("High_Lag", lags)),
    !!!setNames(as.list(sprintf("%.2f (%.2f, %.2f)", or, low, high)), paste0("Lag ", lags))
  )
}

# Combine Anseriformes + Predators (Crude + Adjusted)
table_lag_ORs <- bind_rows(
  extract_lag_ORs_wide(cp_slice_anseri_crude,   "Anseriformes (Crude)"),
  extract_lag_ORs_wide(cp_slice_predator_crude, "Predators (Crude)"),
  extract_lag_ORs_wide(cp_slice_anseri_adj,     "Anseriformes (Adjusted)"),
  extract_lag_ORs_wide(cp_slice_predator_adj,   "Predators (Adjusted)")
)

# Identify significant lags (CI entirely above/below 1)
lags <- 0:8
highlight_matrix <- sapply(lags, function(i) {
  table_lag_ORs[[paste0("Low_Lag", i)]] > 1 | table_lag_ORs[[paste0("High_Lag", i)]] < 1
})
colnames(highlight_matrix) <- paste0("Lag ", lags)

# Build flextable for export
ft <- table_lag_ORs %>%
  select(Predictor, paste0("Lag ", lags)) %>%
  flextable() %>%
  set_header_labels(Predictor = "Predictor") %>%
  add_header_row(values = c("Predictor", "Odds Ratio (95% CI)"),
                 colwidths = c(1, length(lags))) %>%
  fontsize(size = 11, part = "all") %>%
  autofit() %>%
  align(align = "center", part = "all") %>%
  theme_booktabs()

for (i in lags) {
  ft <- bold(ft, i = which(highlight_matrix[, paste0("Lag ", i)]),
             j = paste0("Lag ", i))
}

# Export Word table
doc <- read_docx() %>%
  body_add_flextable(ft) %>%
  body_add_par(
    "Table: Lag-Specific Odds Ratios for Anseriformes and Predator Bird Abundance in 8-week Case-Crossover Models (2022). Crude models include only the respective abundance; adjusted models include environmental covariates. Odds ratios are per 0.5 SD increase; statistically significant values are shown in bold.",
    style = "Normal"
  )

print(doc, target = file.path(tables_sensitivity_folder,
  "Table_LagSpecific_OR_Anseri_Predators_8lag.docx"))

ft

```

## We still get very 
