---
title: "Penalized conditional logistic LASSO across all candidate exposures"
author: "Fintan Mooney"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

```{r initiate}
rm(list = ls())
# 0a.Declare root directory, folder location and load essential stuff
project.folder = paste0(print(here::here()), '/')
source(paste0(project.folder, 'create_folder_structure.R'))
source(paste0(functions.folder, 'script_initiate.R'))
```

# 1. Load case-crossover dataset

```{r load}
case_crossover_df <- readRDS(paste0(objects_folder, "case_crossover_df.RDS"))
setDT(case_crossover_df)
```

# 2. Build crossbases (primary spec: met = ns(df=4); bird = weekly strata)

```{r crossbasis-met}
cb_temperature <- crossbasis(
  as.matrix(case_crossover_df[, paste0("temperature_Lag", 0:28), with = FALSE]),
  lag = c(0, 28), argvar = list(fun = "lin"), arglag = list(fun = "ns", df = 4)
)
cb_dewpoint <- crossbasis(
  as.matrix(case_crossover_df[, paste0("dewpoint_Lag", 0:28), with = FALSE]),
  lag = c(0, 28), argvar = list(fun = "lin"), arglag = list(fun = "ns", df = 4)
)
cb_relative_humidity <- crossbasis(
  as.matrix(case_crossover_df[, paste0("relative_humidity_Lag", 0:28), with = FALSE]),
  lag = c(0, 28), argvar = list(fun = "lin"), arglag = list(fun = "ns", df = 4)
)
cb_wind_speed <- crossbasis(
  as.matrix(case_crossover_df[, paste0("wind_speed_Lag", 0:28), with = FALSE]),
  lag = c(0, 28), argvar = list(fun = "lin"), arglag = list(fun = "ns", df = 4)
)
cb_soil_moisture <- crossbasis(
  as.matrix(case_crossover_df[, paste0("soil_moisture_Lag", 0:28), with = FALSE]),
  lag = c(0, 28), argvar = list(fun = "lin"), arglag = list(fun = "ns", df = 4)
)
cb_snow_density <- crossbasis(
  as.matrix(case_crossover_df[, paste0("snow_density_Lag", 0:28), with = FALSE]),
  lag = c(0, 28), argvar = list(fun = "lin"), arglag = list(fun = "ns", df = 4)
)
cb_precipitation <- crossbasis(
  as.matrix(case_crossover_df[, paste0("precipitation_Lag", 0:28), with = FALSE]),
  lag = c(0, 28), argvar = list(fun = "lin"), arglag = list(fun = "ns", df = 4)
)
cb_runoff <- crossbasis(
  as.matrix(case_crossover_df[, paste0("runoff_Lag", 0:28), with = FALSE]),
  lag = c(0, 28), argvar = list(fun = "lin"), arglag = list(fun = "ns", df = 4)
)
cb_evapotranspiration <- crossbasis(
  as.matrix(case_crossover_df[, paste0("evapotranspiration_Lag", 0:28), with = FALSE]),
  lag = c(0, 28), argvar = list(fun = "lin"), arglag = list(fun = "ns", df = 4)
)
cb_leaf_area_index <- crossbasis(
  as.matrix(case_crossover_df[, paste0("leaf_area_index_Lag", 0:28), with = FALSE]),
  lag = c(0, 28), argvar = list(fun = "lin"), arglag = list(fun = "ns", df = 4)
)
cb_shortwave_radiation <- crossbasis(
  as.matrix(case_crossover_df[, paste0("shortwave_radiation_Lag", 0:28), with = FALSE]),
  lag = c(0, 28), argvar = list(fun = "lin"), arglag = list(fun = "ns", df = 4)
)
cb_longwave_radiation <- crossbasis(
  as.matrix(case_crossover_df[, paste0("longwave_radiation_Lag", 0:28), with = FALSE]),
  lag = c(0, 28), argvar = list(fun = "lin"), arglag = list(fun = "ns", df = 4)
)
cb_surface_pressure <- crossbasis(
  as.matrix(case_crossover_df[, paste0("surface_pressure_Lag", 0:28), with = FALSE]),
  lag = c(0, 28), argvar = list(fun = "lin"), arglag = list(fun = "ns", df = 4)
)
```

```{r crossbasis-bird}
cb_total_bird_abundance <- crossbasis(
  as.matrix(case_crossover_df[, paste0("total_bird_abundance_Lag", 0:28), with = FALSE]),
  lag = c(0, 28), argvar = list(fun = "lin"), arglag = list(fun = "strata", breaks = c(7, 14, 21))
)
cb_anseriformes <- crossbasis(
  as.matrix(case_crossover_df[, paste0("anseriformes_Lag", 0:28), with = FALSE]),
  lag = c(0, 28), argvar = list(fun = "lin"), arglag = list(fun = "strata", breaks = c(7, 14, 21))
)
cb_predators <- crossbasis(
  as.matrix(case_crossover_df[, paste0("predators_Lag", 0:28), with = FALSE]),
  lag = c(0, 28), argvar = list(fun = "lin"), arglag = list(fun = "strata", breaks = c(7, 14, 21))
)
```

# 3. Lag-0 correlation matrix across all 16 candidates

```{r correlation}
vars_lag0 <- c(
  "temperature", "dewpoint", "relative_humidity", "wind_speed",
  "soil_moisture", "snow_density", "precipitation", "runoff",
  "evapotranspiration", "leaf_area_index", "shortwave_radiation",
  "longwave_radiation", "surface_pressure",
  "total_bird_abundance", "anseriformes", "predators"
)
cor_matrix <- cor(
  case_crossover_df[, paste0(vars_lag0, "_Lag0"), with = FALSE],
  use = "complete.obs"
)
colnames(cor_matrix) <- vars_lag0
rownames(cor_matrix) <- vars_lag0
cor_matrix <- round(cor_matrix, 2)

png(paste0(figures_sensitivity_folder, "figure_correlation_matrix_lag0.png"),
    width = 10, height = 10, units = "in", res = 300)
corrplot(
  cor_matrix, method = "color", type = "upper",
  addCoef.col = "black", tl.col = "black", tl.srt = 45,
  number.cex = 0.6,
  col = colorRampPalette(c("purple4", "white", "darkgreen"))(200),
  mar = c(0, 0, 2, 0)
)
title("Correlation Matrix of Lag-0 Predictors", line = -1)
dev.off()

print(cor_matrix)
```

# 4. Assemble LASSO design matrix

```{r build-X}
X_temperature          <- as.matrix(unclass(cb_temperature));          colnames(X_temperature)          <- paste0("cb_temperature_l",          seq_len(ncol(X_temperature)))
X_dewpoint             <- as.matrix(unclass(cb_dewpoint));             colnames(X_dewpoint)             <- paste0("cb_dewpoint_l",             seq_len(ncol(X_dewpoint)))
X_relative_humidity    <- as.matrix(unclass(cb_relative_humidity));    colnames(X_relative_humidity)    <- paste0("cb_relative_humidity_l",    seq_len(ncol(X_relative_humidity)))
X_wind_speed           <- as.matrix(unclass(cb_wind_speed));           colnames(X_wind_speed)           <- paste0("cb_wind_speed_l",           seq_len(ncol(X_wind_speed)))
X_soil_moisture        <- as.matrix(unclass(cb_soil_moisture));        colnames(X_soil_moisture)        <- paste0("cb_soil_moisture_l",        seq_len(ncol(X_soil_moisture)))
X_snow_density         <- as.matrix(unclass(cb_snow_density));         colnames(X_snow_density)         <- paste0("cb_snow_density_l",         seq_len(ncol(X_snow_density)))
X_precipitation        <- as.matrix(unclass(cb_precipitation));        colnames(X_precipitation)        <- paste0("cb_precipitation_l",        seq_len(ncol(X_precipitation)))
X_runoff               <- as.matrix(unclass(cb_runoff));               colnames(X_runoff)               <- paste0("cb_runoff_l",               seq_len(ncol(X_runoff)))
X_evapotranspiration   <- as.matrix(unclass(cb_evapotranspiration));   colnames(X_evapotranspiration)   <- paste0("cb_evapotranspiration_l",   seq_len(ncol(X_evapotranspiration)))
X_leaf_area_index      <- as.matrix(unclass(cb_leaf_area_index));      colnames(X_leaf_area_index)      <- paste0("cb_leaf_area_index_l",      seq_len(ncol(X_leaf_area_index)))
X_shortwave_radiation  <- as.matrix(unclass(cb_shortwave_radiation));  colnames(X_shortwave_radiation)  <- paste0("cb_shortwave_radiation_l",  seq_len(ncol(X_shortwave_radiation)))
X_longwave_radiation   <- as.matrix(unclass(cb_longwave_radiation));   colnames(X_longwave_radiation)   <- paste0("cb_longwave_radiation_l",   seq_len(ncol(X_longwave_radiation)))
X_surface_pressure     <- as.matrix(unclass(cb_surface_pressure));     colnames(X_surface_pressure)     <- paste0("cb_surface_pressure_l",     seq_len(ncol(X_surface_pressure)))
X_total_bird_abundance <- as.matrix(unclass(cb_total_bird_abundance)); colnames(X_total_bird_abundance) <- paste0("cb_total_bird_abundance_l", seq_len(ncol(X_total_bird_abundance)))
X_anseriformes         <- as.matrix(unclass(cb_anseriformes));         colnames(X_anseriformes)         <- paste0("cb_anseriformes_l",         seq_len(ncol(X_anseriformes)))
X_predators            <- as.matrix(unclass(cb_predators));            colnames(X_predators)            <- paste0("cb_predators_l",            seq_len(ncol(X_predators)))

X <- cbind(
  X_temperature, X_dewpoint, X_relative_humidity, X_wind_speed,
  X_soil_moisture, X_snow_density, X_precipitation, X_runoff,
  X_evapotranspiration, X_leaf_area_index,
  X_shortwave_radiation, X_longwave_radiation, X_surface_pressure,
  X_total_bird_abundance, X_anseriformes, X_predators
)

y      <- case_crossover_df$outbreak_binary
strata <- case_crossover_df$stratum_id

# 4a.Keep only strata with at least one case and one control
keep_strata <- names(which(tapply(y, strata, function(z) length(unique(z))) == 2))
sel    <- strata %in% keep_strata
X      <- X[sel, , drop = FALSE]
y      <- y[sel]
strata <- strata[sel]
```

# 5. Cross-validated penalized conditional logistic regression

```{r lasso}
set.seed(2026)
fit   <- clogitL1(x = X, y = y, strata = strata,
                  numLambda = 50, minLambdaRatio = 0.001, alpha = 1)
cvfit <- cv.clogitL1(fit, numFolds = 10)

png(paste0(figures_sensitivity_folder, "figure_clogitL1_model_stability.png"),
    width = 8, height = 6, units = "in", res = 300)
plot(cvfit)
title("Cross-validated Penalized Conditional Logistic Regression (Model Stability)", line = 1)
dev.off()

plot(cvfit)
title("Cross-validated Penalized Conditional Logistic Regression (Model Stability)", line = 1)

best_lambda <- cvfit$minCV_lambda
idx_best    <- which.min(abs(fit$lambda - best_lambda))
beta_best   <- fit$beta[idx_best, ]
names(beta_best) <- colnames(X)
```

# 6. Parent-variable summary of selected coefficients

```{r summary}
parent_of <- function(v) fcase(
  grepl("^cb_temperature_",          v), "Temperature",
  grepl("^cb_dewpoint_",             v), "Dewpoint",
  grepl("^cb_relative_humidity_",    v), "Relative Humidity",
  grepl("^cb_wind_speed_",           v), "Wind Speed",
  grepl("^cb_soil_moisture_",        v), "Soil Moisture",
  grepl("^cb_snow_density_",         v), "Snow Density",
  grepl("^cb_precipitation_",        v), "Precipitation",
  grepl("^cb_runoff_",               v), "Runoff",
  grepl("^cb_evapotranspiration_",   v), "Evapotranspiration",
  grepl("^cb_leaf_area_index_",      v), "Leaf Area Index",
  grepl("^cb_shortwave_radiation_",  v), "Shortwave Radiation",
  grepl("^cb_longwave_radiation_",   v), "Longwave Radiation",
  grepl("^cb_surface_pressure_",     v), "Surface Pressure",
  grepl("^cb_total_bird_abundance_", v), "Total Bird Abundance",
  grepl("^cb_anseriformes_",         v), "Anseriformes",
  grepl("^cb_predators_",            v), "Predators"
)

all_cols    <- data.table(variable = colnames(X), parent_var = parent_of(colnames(X)))
block_sizes <- all_cols[, .(n_basis_total = .N), by = parent_var]

coef_dt <- data.table(variable = names(beta_best), beta = beta_best,
                      parent_var = parent_of(names(beta_best)))

summary_vars <- coef_dt[beta != 0,
  .(n_active      = .N,
    n_positive    = sum(beta > 0),
    n_negative    = sum(beta < 0),
    mean_beta     = mean(beta),
    mean_abs_beta = mean(abs(beta)),
    max_abs_beta  = max(abs(beta)),
    sum_abs_beta  = sum(abs(beta))),
  by = parent_var]

summary_vars <- block_sizes[summary_vars, on = "parent_var"]
summary_vars[is.na(n_active),      n_active      := 0L]
summary_vars[is.na(n_positive),    n_positive    := 0L]
summary_vars[is.na(n_negative),    n_negative    := 0L]
summary_vars[is.na(mean_beta),     mean_beta     := 0]
summary_vars[is.na(mean_abs_beta), mean_abs_beta := 0]
summary_vars[is.na(max_abs_beta),  max_abs_beta  := 0]
summary_vars[is.na(sum_abs_beta),  sum_abs_beta  := 0]
summary_vars[, prop_active := n_active / n_basis_total]

# 6a.Append zeroed-out parents so the table is exhaustive
all_parents <- unique(all_cols$parent_var)
missing_parents <- setdiff(all_parents, summary_vars$parent_var)
if (length(missing_parents) > 0) {
  pad <- block_sizes[parent_var %in% missing_parents, .(parent_var, n_basis_total)]
  pad[, `:=`(n_active = 0L, n_positive = 0L, n_negative = 0L,
             mean_beta = 0, mean_abs_beta = 0, max_abs_beta = 0,
             sum_abs_beta = 0, prop_active = 0)]
  summary_vars <- rbind(summary_vars, pad, fill = TRUE)
}

summary_vars <- summary_vars[order(-sum_abs_beta)]
summary_vars_export <- copy(summary_vars)
summary_vars_export[, `:=`(
  mean_beta     = round(mean_beta,     3),
  mean_abs_beta = round(mean_abs_beta, 3),
  max_abs_beta  = round(max_abs_beta,  3),
  sum_abs_beta  = round(sum_abs_beta,  3),
  prop_active   = round(prop_active,   3)
)]

print(summary_vars_export)
```

# 7. Export LASSO selection table

```{r export-table}
ft_stability <- summary_vars_export[, .(parent_var, n_active, prop_active,
                                        mean_beta, mean_abs_beta, sum_abs_beta)] %>%
  regulartable() %>%
  set_header_labels(
    parent_var    = "Variable",
    n_active      = "Active Terms (β ≠ 0)",
    prop_active   = "Proportion Active",
    mean_beta     = "Mean β",
    mean_abs_beta = "Mean |β|",
    sum_abs_beta  = "Σ |β|"
  ) %>%
  align(part = "header", align = "center") %>%
  align(j = c("n_active", "prop_active", "mean_beta", "mean_abs_beta", "sum_abs_beta"),
        align = "right", part = "body") %>%
  autofit()

doc <- read_docx() %>%
  body_add_par("Table Sx. Variable-level LASSO selection summary for crossbasis predictors", style = "Normal") %>%
  body_add_par("Summary of non-zero penalized conditional logistic regression coefficients at the optimal lambda across all 16 candidate exposures. Strata: zone x year x calendar month x day-of-week.",
               style = "Normal") %>%
  body_add_flextable(ft_stability)
print(doc, target = paste0(tables_sensitivity_folder, "Table_ClogitL1_Model_Stability.docx"))

saveRDS(list(fit = fit, cvfit = cvfit, beta_best = beta_best,
             summary_vars = summary_vars_export),
        paste0(objects_folder, "lasso_screen.RDS"))
```

# 8. Reduced clogit using LASSO-active parents

```{r reduced-fit}
parent_to_cb <- c(
  "Temperature"          = "cb_temperature",
  "Dewpoint"             = "cb_dewpoint",
  "Relative Humidity"    = "cb_relative_humidity",
  "Wind Speed"           = "cb_wind_speed",
  "Soil Moisture"        = "cb_soil_moisture",
  "Snow Density"         = "cb_snow_density",
  "Precipitation"        = "cb_precipitation",
  "Runoff"               = "cb_runoff",
  "Evapotranspiration"   = "cb_evapotranspiration",
  "Leaf Area Index"      = "cb_leaf_area_index",
  "Shortwave Radiation"  = "cb_shortwave_radiation",
  "Longwave Radiation"   = "cb_longwave_radiation",
  "Surface Pressure"     = "cb_surface_pressure",
  "Total Bird Abundance" = "cb_total_bird_abundance",
  "Anseriformes"         = "cb_anseriformes",
  "Predators"            = "cb_predators"
)
active_parents <- summary_vars[n_active > 0, parent_var]
active_cb      <- parent_to_cb[active_parents]
print(active_cb)

reduced_form  <- as.formula(paste0("outbreak_binary ~ ",
                                   paste(active_cb, collapse = " + "),
                                   " + strata(stratum_id)"))
reduced_model <- clogit(reduced_form, data = case_crossover_df, method = "efron")
saveRDS(reduced_model, paste0(models_folder, "casecrossover_model_lasso_reduced.RDS"))

print(summary(reduced_model))
```

# 9. Cumulative ORs for reduced-model predictors (per 0.5 SD, lag 0–28 d)

```{r reduced-results}
cp_list <- list()
for (cb_name in active_cb) {
  cp_list[[cb_name]] <- eval(parse(text = sprintf(
    "crosspred(%s, reduced_model, cen = 0, at = 0.5, cumul = TRUE)",
    cb_name
  )))
}
names(cp_list) <- active_parents

reduced_tbl <- rbindlist(lapply(names(cp_list), function(label) {
  cp <- cp_list[[label]]
  n  <- ncol(cp$cumRRfit)
  data.table(
    predictor = label,
    OR_CI = sprintf("%.2f (%.2f, %.2f)",
                    cp$cumRRfit["0.5",  n],
                    cp$cumRRlow["0.5",  n],
                    cp$cumRRhigh["0.5", n])
  )
}))
print(reduced_tbl)

ft_reduced <- reduced_tbl %>%
  flextable() %>%
  set_header_labels(predictor = "Predictor",
                    OR_CI     = "Cumulative OR (95% CI), lag 0–28 d") %>%
  fontsize(size = 11, part = "all") %>%
  autofit() %>%
  align(align = "center", part = "all") %>%
  theme_booktabs()

doc <- read_docx() %>%
  body_add_flextable(ft_reduced) %>%
  body_add_par(
    "Cumulative odds ratios from the LASSO-reduced case-crossover model, retaining parent variables with at least one non-zero penalized coefficient at the optimal lambda. Per 0.5 SD increase.",
    style = "Normal"
  )
print(doc, target = paste0(tables_sensitivity_folder, "Table_LASSO_Reduced_Cumulative_OR.docx"))
```

# 10. AIC comparison vs full primary model

```{r aic-compare}
full_model <- clogit(
  outbreak_binary ~
    cb_runoff + cb_snow_density + cb_soil_moisture + cb_temperature +
    cb_precipitation + cb_wind_speed + cb_total_bird_abundance +
    strata(stratum_id),
  data = case_crossover_df, method = "efron"
)
aic_compare <- AIC(reduced_model, full_model)
print(aic_compare)
```
