---
title: "Andersen–Gill recurrent-event Cox sensitivity (daily zone × day panel)"
author: "Fintan Mooney"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

```{r initiate}
rm(list = ls())
# 0a.Declare root directory, folder location and load essential stuff
project.folder = paste0(print(here::here()), '/')
source(paste0(project.folder, 'create_folder_structure.R'))
source(paste0(functions.folder, 'script_initiate.R'))
```

# 1. Load daily timeseries (modeling version)

```{r load}
daily <- readRDS(paste0(objects_folder, "fishnet_timeseries_daily_modeling.RDS"))
setDT(daily)
setkey(daily, zone_id, date)

case_window_start <- as.Date("2022-01-01")
case_window_end   <- as.Date("2022-12-31")

daily <- daily[date >= case_window_start & date <= case_window_end]
```

# 2. Transform exposures (match case-crossover scale)

```{r transform}
stopifnot(
  "runoff already in mm — reload `daily` before re-running" =
    max(daily$runoff,        na.rm = TRUE) < 1,
  "precipitation already in mm — reload `daily` before re-running" =
    max(daily$precipitation, na.rm = TRUE) < 1
)
daily[, runoff        := runoff        * 1000]
daily[, precipitation := precipitation * 1000]

daily[, anseriformes := log1p(anseriformes)]
daily[, runoff               := log1p(runoff)]
daily[, precipitation        := log1p(precipitation)]
daily[, anseriformes         := log1p(anseriformes)]
daily[, predators            := log1p(predators)]

daily[, temperature          := as.numeric(scale(temperature))]
daily[, dewpoint             := as.numeric(scale(dewpoint))]
daily[, relative_humidity    := as.numeric(scale(relative_humidity))]
daily[, wind_speed           := as.numeric(scale(wind_speed))]
daily[, soil_moisture        := as.numeric(scale(soil_moisture))]
daily[, precipitation        := as.numeric(scale(precipitation))]
daily[, runoff               := as.numeric(scale(runoff))]
daily[, evapotranspiration   := as.numeric(scale(evapotranspiration))]
daily[, leaf_area_index      := as.numeric(scale(leaf_area_index))]
daily[, shortwave_radiation  := as.numeric(scale(shortwave_radiation))]
daily[, longwave_radiation   := as.numeric(scale(longwave_radiation))]
daily[, surface_pressure     := as.numeric(scale(surface_pressure))]
daily[, anseriformes := as.numeric(scale(anseriformes))]
daily[, anseriformes         := as.numeric(scale(anseriformes))]
daily[, predators            := as.numeric(scale(predators))]
```

# 3. Build (start, stop, event) interval panel

```{r intervals}
setorder(daily, zone_id, date)
daily[, `:=`(
  tstart = seq_len(.N) - 1L,
  tstop  = seq_len(.N),
  event  = as.integer(outbreak_binary)
), by = zone_id]

case_zones <- daily[event == 1, unique(zone_id)]
ag_panel   <- daily[zone_id %in% case_zones]
ag_panel[, stratum_month := lubridate::month(date)]
```

# 4. Build crossbasis terms (group = zone_id so lag does not cross zones)

```{r crossbasis}
cb_runoff <- crossbasis(
  ag_panel$runoff, lag = c(0, 14),
  argvar = list(fun = "lin"),
  arglag = list(fun = "ns", df = 2),
  group  = ag_panel$zone_id
)
cb_soil_moisture <- crossbasis(
  ag_panel$soil_moisture, lag = c(0, 14),
  argvar = list(fun = "lin"),
  arglag = list(fun = "ns", df = 2),
  group  = ag_panel$zone_id
)
cb_temperature <- crossbasis(
  ag_panel$temperature, lag = c(0, 14),
  argvar = list(fun = "lin"),
  arglag = list(fun = "ns", df = 2),
  group  = ag_panel$zone_id
)
cb_precipitation <- crossbasis(
  ag_panel$precipitation, lag = c(0, 14),
  argvar = list(fun = "lin"),
  arglag = list(fun = "ns", df = 2),
  group  = ag_panel$zone_id
)
cb_wind_speed <- crossbasis(
  ag_panel$wind_speed, lag = c(0, 14),
  argvar = list(fun = "lin"),
  arglag = list(fun = "ns", df = 2),
  group  = ag_panel$zone_id
)
cb_anseriformes <- crossbasis(
  ag_panel$anseriformes, lag = c(0, 14),
  argvar = list(fun = "lin"),
  arglag = list(fun = "strata", breaks = c(7)),
  group  = ag_panel$zone_id
)
```

# 5. Fit Andersen–Gill recurrent-event Cox model

```{r fit}
ag_model <- coxph(
  Surv(tstart, tstop, event) ~
    cb_runoff + cb_soil_moisture + cb_temperature +
    cb_precipitation + cb_wind_speed + cb_anseriformes +
    strata(stratum_month),
  data    = ag_panel,
  ties    = "efron",
  cluster = zone_id
)
saveRDS(ag_model, paste0(models_folder, "casecrossover_model_cox_ag.RDS"))

print(summary(ag_model))
```

# 6. Cumulative hazard ratios (lag 0–14 d)

```{r cumulative-hr}
cp_runoff_ag <- crosspred(cb_runoff,               ag_model, cen = 0, at = 0.5, cumul = TRUE)
cp_soil_ag   <- crosspred(cb_soil_moisture,        ag_model, cen = 0, at = 0.5, cumul = TRUE)
cp_temp_ag   <- crosspred(cb_temperature,          ag_model, cen = 0, at = 0.5, cumul = TRUE)
cp_precip_ag <- crosspred(cb_precipitation,        ag_model, cen = 0, at = 0.5, cumul = TRUE)
cp_wind_ag   <- crosspred(cb_wind_speed,           ag_model, cen = 0, at = 0.5, cumul = TRUE)
cp_bird_ag   <- crosspred(cb_anseriformes, ag_model, cen = 0, at = 0.5, cumul = TRUE)

n <- ncol(cp_runoff_ag$cumRRfit)
hr_runoff <- sprintf("%.2f (%.2f, %.2f)", cp_runoff_ag$cumRRfit["0.5", n], cp_runoff_ag$cumRRlow["0.5", n], cp_runoff_ag$cumRRhigh["0.5", n])
hr_soil   <- sprintf("%.2f (%.2f, %.2f)", cp_soil_ag$cumRRfit["0.5", n],   cp_soil_ag$cumRRlow["0.5", n],   cp_soil_ag$cumRRhigh["0.5", n])
hr_temp   <- sprintf("%.2f (%.2f, %.2f)", cp_temp_ag$cumRRfit["0.5", n],   cp_temp_ag$cumRRlow["0.5", n],   cp_temp_ag$cumRRhigh["0.5", n])
hr_precip <- sprintf("%.2f (%.2f, %.2f)", cp_precip_ag$cumRRfit["0.5", n], cp_precip_ag$cumRRlow["0.5", n], cp_precip_ag$cumRRhigh["0.5", n])
hr_wind   <- sprintf("%.2f (%.2f, %.2f)", cp_wind_ag$cumRRfit["0.5", n],   cp_wind_ag$cumRRlow["0.5", n],   cp_wind_ag$cumRRhigh["0.5", n])
hr_bird   <- sprintf("%.2f (%.2f, %.2f)", cp_bird_ag$cumRRfit["0.5", n],   cp_bird_ag$cumRRlow["0.5", n],   cp_bird_ag$cumRRhigh["0.5", n])

cum_hr <- tibble(
  predictor = c("Runoff", "Soil Moisture", "Temperature",
                "Precipitation", "Wind Speed", "Anseriformes"),
  HR_CI     = c(hr_runoff, hr_soil, hr_temp, hr_precip, hr_wind, hr_bird)
)

ft <- cum_hr %>%
  flextable() %>%
  set_header_labels(predictor = "Predictor",
                    HR_CI     = "Cumulative HR (95% CI), lag 0–14 d") %>%
  fontsize(size = 11, part = "all") %>%
  autofit() %>%
  align(align = "center", part = "all") %>%
  theme_booktabs()

doc <- read_docx() %>%
  body_add_flextable(ft) %>%
  body_add_par(
    "Cumulative hazard ratios from Andersen-Gill recurrent-event Cox model on the daily zone × day panel, per 0.5 SD increase in each predictor. Baseline hazard stratified by calendar month; cluster-robust SEs clustered on zone.",
    style = "Normal"
  )
print(doc, target = paste0(tables_sensitivity_folder, "Table_Cox_AG_Cumulative_HR.docx"))

print(cum_hr)
```

# 7. Side-by-side with the time-stratified case-crossover

```{r compare}
primary_model     <- readRDS(paste0(models_folder, "casecrossover_model.RDS"))
case_crossover_df <- readRDS(paste0(objects_folder, "case_crossover_df.RDS"))

# Crosspreds for the primary model need cb_ names matching its formula terms;
# wrap in local() so these reassignments don't clobber the AG-sized globals.
cc_cp <- local({
  cb_runoff <- crossbasis(
    as.matrix(case_crossover_df[, paste0("runoff_Lag", 0:28), with = FALSE]),
    lag = c(0, 28), argvar = list(fun = "lin"), arglag = list(fun = "ns", df = 3)
  )
  cb_soil_moisture <- crossbasis(
    as.matrix(case_crossover_df[, paste0("soil_moisture_Lag", 0:28), with = FALSE]),
    lag = c(0, 28), argvar = list(fun = "lin"), arglag = list(fun = "ns", df = 3)
  )
  cb_temperature <- crossbasis(
    as.matrix(case_crossover_df[, paste0("temperature_Lag", 0:28), with = FALSE]),
    lag = c(0, 28), argvar = list(fun = "lin"), arglag = list(fun = "ns", df = 3)
  )
  cb_precipitation <- crossbasis(
    as.matrix(case_crossover_df[, paste0("precipitation_Lag", 0:28), with = FALSE]),
    lag = c(0, 28), argvar = list(fun = "lin"), arglag = list(fun = "ns", df = 3)
  )
  cb_wind_speed <- crossbasis(
    as.matrix(case_crossover_df[, paste0("wind_speed_Lag", 0:28), with = FALSE]),
    lag = c(0, 28), argvar = list(fun = "lin"), arglag = list(fun = "ns", df = 3)
  )
  cb_anseriformes <- crossbasis(
    as.matrix(case_crossover_df[, paste0("anseriformes_Lag", 0:28), with = FALSE]),
    lag = c(0, 28), argvar = list(fun = "lin"), arglag = list(fun = "strata", breaks = c(7, 14, 21))
  )
  list(
    runoff = crosspred(cb_runoff,               primary_model, cen = 0, at = 0.5, cumul = TRUE),
    soil   = crosspred(cb_soil_moisture,        primary_model, cen = 0, at = 0.5, cumul = TRUE),
    temp   = crosspred(cb_temperature,          primary_model, cen = 0, at = 0.5, cumul = TRUE),
    precip = crosspred(cb_precipitation,        primary_model, cen = 0, at = 0.5, cumul = TRUE),
    wind   = crosspred(cb_wind_speed,           primary_model, cen = 0, at = 0.5, cumul = TRUE),
    bird   = crosspred(cb_anseriformes, primary_model, cen = 0, at = 0.5, cumul = TRUE)
  )
})
cp_runoff_cc <- cc_cp$runoff
cp_soil_cc   <- cc_cp$soil
cp_temp_cc   <- cc_cp$temp
cp_precip_cc <- cc_cp$precip
cp_wind_cc   <- cc_cp$wind
cp_bird_cc   <- cc_cp$bird

n_cc <- ncol(cp_runoff_cc$cumRRfit)
or_runoff_cc <- sprintf("%.2f (%.2f, %.2f)", cp_runoff_cc$cumRRfit["0.5", n_cc], cp_runoff_cc$cumRRlow["0.5", n_cc], cp_runoff_cc$cumRRhigh["0.5", n_cc])
or_soil_cc   <- sprintf("%.2f (%.2f, %.2f)", cp_soil_cc$cumRRfit["0.5", n_cc],   cp_soil_cc$cumRRlow["0.5", n_cc],   cp_soil_cc$cumRRhigh["0.5", n_cc])
or_temp_cc   <- sprintf("%.2f (%.2f, %.2f)", cp_temp_cc$cumRRfit["0.5", n_cc],   cp_temp_cc$cumRRlow["0.5", n_cc],   cp_temp_cc$cumRRhigh["0.5", n_cc])
or_precip_cc <- sprintf("%.2f (%.2f, %.2f)", cp_precip_cc$cumRRfit["0.5", n_cc], cp_precip_cc$cumRRlow["0.5", n_cc], cp_precip_cc$cumRRhigh["0.5", n_cc])
or_wind_cc   <- sprintf("%.2f (%.2f, %.2f)", cp_wind_cc$cumRRfit["0.5", n_cc],   cp_wind_cc$cumRRlow["0.5", n_cc],   cp_wind_cc$cumRRhigh["0.5", n_cc])
or_bird_cc   <- sprintf("%.2f (%.2f, %.2f)", cp_bird_cc$cumRRfit["0.5", n_cc],   cp_bird_cc$cumRRlow["0.5", n_cc],   cp_bird_cc$cumRRhigh["0.5", n_cc])

comparison <- tibble(
  Predictor                     = c("Runoff", "Soil Moisture", "Temperature",
                                    "Precipitation", "Wind Speed", "Anseriformes"),
  `Case-crossover OR (95% CI)`  = c(or_runoff_cc, or_soil_cc, or_temp_cc, or_precip_cc, or_wind_cc, or_bird_cc),
  `Cox AG HR (95% CI)`          = c(hr_runoff, hr_soil, hr_temp, hr_precip, hr_wind, hr_bird)
)

ft_compare <- comparison %>%
  flextable() %>%
  fontsize(size = 11, part = "all") %>%
  autofit() %>%
  align(align = "center", part = "all") %>%
  theme_booktabs()

doc <- read_docx() %>%
  body_add_flextable(ft_compare) %>%
  body_add_par(
    "Cumulative effect estimates (per 0.5 SD) for the time-stratified case-crossover model (lag 0–28 d) and the Andersen-Gill Cox sensitivity model (lag 0–14 d, reduced for identifiability).",
    style = "Normal"
  )
print(doc, target = paste0(tables_sensitivity_folder, "Table_Cox_AG_vs_CaseCrossover.docx"))

print(comparison)
```

# 8. Lag-response slice figure

```{r plot, fig.width = 12, fig.height = 7}
png(paste0(figures_sensitivity_folder, "figure_cox_ag_slices.png"),
    width = 12, height = 7, units = "in", res = 300)

par(mfrow = c(2, 3), mar = c(4.2, 4.2, 2.5, 1), oma = c(3, 4, 0, 1),
    cex.main = 1.0, cex.lab = 1.0, cex.axis = 0.9)

plot(cp_runoff_ag, type = "slices", var = 0.5, ci = "area", col = "black", ci.arg = list(col = rgb(0,0,0,0.15), border = NA), lwd = 1.3, main = "Runoff", xlab = "", ylab = "", las = 1); abline(h = 1, lty = 2, col = "gray40")
plot(cp_soil_ag,   type = "slices", var = 0.5, ci = "area", col = "black", ci.arg = list(col = rgb(0,0,0,0.15), border = NA), lwd = 1.3, main = "Soil Moisture", xlab = "", ylab = "", las = 1); abline(h = 1, lty = 2, col = "gray40")
plot(cp_temp_ag,   type = "slices", var = 0.5, ci = "area", col = "black", ci.arg = list(col = rgb(0,0,0,0.15), border = NA), lwd = 1.3, main = "Temperature", xlab = "", ylab = "", las = 1); abline(h = 1, lty = 2, col = "gray40")
plot(cp_precip_ag, type = "slices", var = 0.5, ci = "area", col = "black", ci.arg = list(col = rgb(0,0,0,0.15), border = NA), lwd = 1.3, main = "Precipitation", xlab = "", ylab = "", las = 1); abline(h = 1, lty = 2, col = "gray40")
plot(cp_wind_ag,   type = "slices", var = 0.5, ci = "area", col = "black", ci.arg = list(col = rgb(0,0,0,0.15), border = NA), lwd = 1.3, main = "Wind Speed", xlab = "", ylab = "", las = 1); abline(h = 1, lty = 2, col = "gray40")
plot(cp_bird_ag,   type = "slices", var = 0.5, ci = "area", col = "black", ci.arg = list(col = rgb(0,0,0,0.15), border = NA), lwd = 1.3, main = "Anseriformes", xlab = "", ylab = "", las = 1); abline(h = 1, lty = 2, col = "gray40")

mtext("Lag (days)",            side = 1, outer = TRUE, line = 0, cex = 1.0)
mtext("Hazard Ratio (95% CI)", side = 2, outer = TRUE, line = 0, cex = 1.0)
dev.off()
```

# 9. Proportional hazards diagnostic

```{r ph-test}
zph_test <- cox.zph(ag_model, transform = "identity")
print(zph_test)

png(paste0(figures_sensitivity_folder, "figure_cox_ag_ph_diagnostic.png"),
    width = 12, height = 9, units = "in", res = 300)
par(mfrow = c(3, 3), mar = c(4, 4, 2.5, 1))
plot(zph_test)
dev.off()

ph_summary <- as.data.frame(zph_test$table) %>%
  tibble::rownames_to_column("term") %>%
  mutate(across(c(chisq, df, p), ~ signif(.x, 3)))

ft_ph <- ph_summary %>%
  flextable() %>%
  fontsize(size = 11, part = "all") %>%
  autofit() %>%
  align(align = "center", part = "all") %>%
  theme_booktabs()

doc <- read_docx() %>%
  body_add_flextable(ft_ph) %>%
  body_add_par(
    "Schoenfeld residual test for the proportional hazards assumption. Low p-values indicate a time-varying effect (PH violation).",
    style = "Normal"
  )
print(doc, target = paste0(tables_sensitivity_folder, "Table_Cox_AG_PH_Test.docx"))
```
