---
title: "Bayesian inference of cumulative effects (flat and weakly informative priors)"
author: "Fintan Mooney"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

```{r initiate}
rm(list = ls())
# 0a.Declare root directory, folder location and load essential stuff
project.folder = paste0(print(here::here()), '/')
source(paste0(project.folder, 'create_folder_structure.R'))
source(paste0(functions.folder, 'script_initiate.R'))
library(MASS)
```

# 1. Load case-crossover dataframe

```{r load}
case_crossover_df <- readRDS(paste0(objects_folder, "case_crossover_df.RDS"))
setDT(case_crossover_df)
```

# 2. Build crossbases (primary spec: lag 0–28 d, ns(df=3) met, strata weekly bird)

```{r crossbases}
cb_runoff <- crossbasis(
  as.matrix(case_crossover_df[, paste0("runoff_Lag", 0:28), with = FALSE]),
  lag = c(0, 28), argvar = list(fun = "lin"), arglag = list(fun = "ns", df = 3)
)
cb_soil_moisture <- crossbasis(
  as.matrix(case_crossover_df[, paste0("soil_moisture_Lag", 0:28), with = FALSE]),
  lag = c(0, 28), argvar = list(fun = "lin"), arglag = list(fun = "ns", df = 3)
)
cb_temperature <- crossbasis(
  as.matrix(case_crossover_df[, paste0("temperature_Lag", 0:28), with = FALSE]),
  lag = c(0, 28), argvar = list(fun = "lin"), arglag = list(fun = "ns", df = 3)
)
cb_precipitation <- crossbasis(
  as.matrix(case_crossover_df[, paste0("precipitation_Lag", 0:28), with = FALSE]),
  lag = c(0, 28), argvar = list(fun = "lin"), arglag = list(fun = "ns", df = 3)
)
cb_wind_speed <- crossbasis(
  as.matrix(case_crossover_df[, paste0("wind_speed_Lag", 0:28), with = FALSE]),
  lag = c(0, 28), argvar = list(fun = "lin"), arglag = list(fun = "ns", df = 3)
)
cb_anseriformes <- crossbasis(
  as.matrix(case_crossover_df[, paste0("anseriformes_Lag", 0:28), with = FALSE]),
  lag = c(0, 28), argvar = list(fun = "lin"),
  arglag = list(fun = "strata", breaks = c(7, 14, 21))
)
```

# 3. Frequentist conditional logistic fit (the likelihood)

```{r fit}
freq_model <- clogit(
  outbreak_binary ~
    cb_runoff + cb_soil_moisture + cb_temperature +
    cb_precipitation + cb_wind_speed + cb_anseriformes +
    strata(stratum_id),
  method = "efron", data = case_crossover_df
)
saveRDS(freq_model, paste0(models_folder, "casecrossover_model_for_bayes.RDS"))

beta_hat <- coef(freq_model)
V_hat    <- vcov(freq_model)
```

# 4. Posterior under flat prior — asymptotic MVN approximation

```{r flat-posterior}
# Under a flat (improper uniform) prior on the coefficient vector, the
# asymptotic posterior is N(beta_hat, V_hat). This is the standard "Bayesian
# flat prior" interpretation of the maximum partial likelihood and is
# asymptotically valid for moderate sample sizes.
set.seed(2026)
post_flat <- MASS::mvrnorm(10000, beta_hat, V_hat)
colnames(post_flat) <- names(beta_hat)
```

# 5. Posterior under weakly informative N(0, 0.3) prior on each coefficient

```{r info-posterior}
# Weakly informative prior: each crossbasis coefficient ~ N(0, 0.3) independent.
# Per-coefficient 95% prior CrI: [-0.6, 0.6] on the log-OR scale → modest
# regularization toward zero. Prevents extreme estimates from over-fit basis
# columns (e.g. the lag-42 d and lag-56 d explosions in the lag-window sensitivity)
# without anchoring strongly to any directional hypothesis.
prior_mean <- rep(0, length(beta_hat))
prior_sd   <- rep(0.3, length(beta_hat))
V_prior    <- diag(prior_sd^2, nrow = length(beta_hat))

# Conjugate MVN-MVN update: V_post^-1 = V_prior^-1 + V_hat^-1
V_post  <- solve(solve(V_prior) + solve(V_hat))
mu_post <- as.vector(V_post %*% (solve(V_prior) %*% prior_mean + solve(V_hat) %*% beta_hat))

set.seed(2026)
post_info <- MASS::mvrnorm(10000, mu_post, V_post)
colnames(post_info) <- names(beta_hat)
```

# 6. Build cumulative-effect contrast vectors

```{r contrasts}
lag_seq        <- 0:28
lag_basis_met  <- do.call(dlnm::onebasis,
                          c(list(x = lag_seq), attr(cb_runoff,      "arglag")))
lag_basis_bird <- do.call(dlnm::onebasis,
                          c(list(x = lag_seq), attr(cb_anseriformes, "arglag")))

# +0.5 SD constant exposure across lag 0-28 → linear predictor contribution =
# 0.5 * sum_k arglag(k) %*% beta = (0.5 * colSums(arglag_basis)) %*% beta
contrast_met  <- 0.5 * colSums(lag_basis_met)
contrast_bird <- 0.5 * colSums(lag_basis_bird)
```

# 7. Posterior summaries per predictor (cumulative OR per +0.5 SD, lag 0–28 d)

```{r posterior-summaries}
summarize_predictor <- function(post_samples, prefix, contrast) {
  cols   <- grep(paste0("^", prefix, "v"), colnames(post_samples))
  log_or <- as.numeric(post_samples[, cols, drop = FALSE] %*% contrast)
  or_s   <- exp(log_or)
  list(
    median = median(or_s),
    cr_lo  = unname(quantile(or_s, 0.025)),
    cr_hi  = unname(quantile(or_s, 0.975)),
    p_pos  = mean(log_or > 0)
  )
}

predictor_specs <- list(
  "Runoff"        = list(prefix = "cb_runoff",        contrast = contrast_met),
  "Soil Moisture" = list(prefix = "cb_soil_moisture", contrast = contrast_met),
  "Temperature"   = list(prefix = "cb_temperature",   contrast = contrast_met),
  "Precipitation" = list(prefix = "cb_precipitation", contrast = contrast_met),
  "Wind Speed"    = list(prefix = "cb_wind_speed",    contrast = contrast_met),
  "Anseriformes"  = list(prefix = "cb_anseriformes",  contrast = contrast_bird)
)

bayes_tbl <- rbindlist(lapply(names(predictor_specs), function(name) {
  spec <- predictor_specs[[name]]
  flat <- summarize_predictor(post_flat, spec$prefix, spec$contrast)
  info <- summarize_predictor(post_info, spec$prefix, spec$contrast)
  data.table(
    Predictor                     = name,
    `OR (flat, 95% CrI)`          = sprintf("%.2f (%.2f, %.2f)",
                                            flat$median, flat$cr_lo, flat$cr_hi),
    `P(effect > 0) | flat`        = sprintf("%.3f", flat$p_pos),
    `OR (informative, 95% CrI)`   = sprintf("%.2f (%.2f, %.2f)",
                                            info$median, info$cr_lo, info$cr_hi),
    `P(effect > 0) | informative` = sprintf("%.3f", info$p_pos)
  )
}))

print(bayes_tbl)
```

# 8. Export comparison table

```{r export-table}
ft_bayes <- bayes_tbl %>%
  flextable() %>%
  fontsize(size = 10, part = "all") %>%
  autofit() %>%
  align(align = "center", part = "all") %>%
  theme_booktabs()

doc <- read_docx() %>%
  body_add_flextable(ft_bayes) %>%
  body_add_par(
    "Bayesian posterior summaries for cumulative effects (per +0.5 SD, lag 0-28 d) from the time-stratified daily case-crossover DLNM. Flat prior: asymptotic MVN approximation of the conditional logistic partial likelihood under an improper uniform prior. Informative prior: each crossbasis coefficient ~ N(0, 0.3), weakly regularizing toward zero (prevents extreme estimates from over-fit basis columns without anchoring to any directional hypothesis). Median odds ratio with 95% credible interval; P(effect > 0) is the posterior probability that the cumulative log-OR exceeds zero. P(effect > 0) > 0.95 corresponds to a 95% credible interval that excludes the null on one side.",
    style = "Normal"
  )
print(doc, target = paste0(tables_sensitivity_folder, "Table_Bayesian_Cumulative_OR.docx"))
```

# 9. Posterior densities of cumulative log-OR

```{r posterior-densities, fig.width = 10, fig.height = 6}
draw_density_panel <- function(post_samples, contrast_lookup, prior_label) {
  par(mfrow = c(2, 3), mar = c(4, 4, 2.5, 1), oma = c(1, 1, 2, 0),
      cex.main = 1.0, cex.lab = 0.95, cex.axis = 0.85)
  for (name in names(predictor_specs)) {
    spec <- predictor_specs[[name]]
    cols <- grep(paste0("^", spec$prefix, "v"), colnames(post_samples))
    log_or <- as.numeric(post_samples[, cols, drop = FALSE] %*% spec$contrast)
    dens <- density(log_or)
    plot(dens, main = name, xlab = "Cumulative log-OR (per +0.5 SD)",
         ylab = "Posterior density", las = 1)
    abline(v = 0, lty = 2, col = "gray40")
    p_pos <- mean(log_or > 0)
    legend("topright", legend = sprintf("P(>0) = %.2f", p_pos), bty = "n", cex = 0.9)
  }
  mtext(prior_label, side = 3, outer = TRUE, line = 0.5, cex = 1.1)
}

png(paste0(figures_sensitivity_folder, "figure_bayesian_posteriors_flat.png"),
    width = 12, height = 7, units = "in", res = 300)
draw_density_panel(post_flat, predictor_specs, "Posterior under flat prior")
dev.off()

png(paste0(figures_sensitivity_folder, "figure_bayesian_posteriors_info.png"),
    width = 12, height = 7, units = "in", res = 300)
draw_density_panel(post_info, predictor_specs, "Posterior under weakly informative N(0, 0.3) prior")
dev.off()
```

# 10. Save results bundle

```{r save}
saveRDS(list(
  freq_model = freq_model,
  beta_hat   = beta_hat,
  V_hat      = V_hat,
  post_flat  = post_flat,
  post_info  = post_info,
  bayes_tbl  = bayes_tbl
), paste0(objects_folder, "bayesian_inference.RDS"))
```
