---
title: "Daily vs Weekly DLNM comparison — SPRING-OUTBREAK SUBSET"
author: "Fintan Mooney"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

```{r initiate}
rm(list = ls())
project.folder = paste0(print(here::here()), '/')
source(paste0(project.folder, 'create_folder_structure.R'))
source(paste0(functions.folder, 'script_initiate.R'))
```

# Scope

Spring case-crossover models, restricted to strata whose case occurred in
March–May 2022 (matches the existing `spring` indicator in the daily panel).
Models are *refit* on the subset rather than just filtered post-hoc, so the
crossbases and conditional likelihood reflect only spring case-control
contrasts. Daily and weekly use the same primary specs as c_22.

The 1:4 unidirectional weekly spring model is underpowered (~59 case-weeks
with 14 coefficients to estimate from ns(df=2) × 6 met + ns(df=2)
anseriformes), and several predictors hit near-separation. The full-year
weekly model is the identifiable design at weekly resolution. The daily
spring fit (60 cases × ~4 controls across 29 daily lag points) is the
primary spring sensitivity.

# 1. Load full data and identify spring strata

```{r load}
# Daily primary (snow_cover, ns(df=3) at lag 0-28 d)
daily_df <- readRDS(paste0(objects_folder, "case_crossover_df.RDS"))
setDT(daily_df)

# Weekly primary (snow_cover, ns(df=2) at lag 0-4 wk, CSLT excluded)
weekly_bundle <- readRDS(paste0(objects_folder, "weekly_primary_ab.RDS"))
weekly_cc_A   <- copy(weekly_bundle$cc_A); setDT(weekly_cc_A)

# --- Identify spring strata ---
# Daily: keep strata where the case row is in spring (months 3-5)
daily_spring_strata <- daily_df[case_control == "case" & spring == 1, unique(stratum_id)]
daily_spring <- daily_df[stratum_id %in% daily_spring_strata]
cat("Daily — spring strata: ", length(daily_spring_strata),
    " | rows: ",  nrow(daily_spring),
    " | cases: ", sum(daily_spring$outbreak_binary == 1), "\n", sep = "")

# Weekly: define spring as case-week's month in 3-5
weekly_cc_A[, case_month := lubridate::month(week_start)]
weekly_spring_strata <- weekly_cc_A[outbreak_binary == 1 & case_month %in% 3:5,
                                     unique(stratum_id)]
weekly_spring <- weekly_cc_A[stratum_id %in% weekly_spring_strata]
cat("Weekly — spring strata: ", length(weekly_spring_strata),
    " | rows: ",   nrow(weekly_spring),
    " | cases: ",  sum(weekly_spring$outbreak_binary == 1), "\n", sep = "")
```

# 2. Refit both models on the spring subsets

```{r refit-spring}
DAILY_MAX <- 28
WEEK_MAX  <- 4

# --- Daily crossbases (lin × ns(df=3); anseriformes strata(7,14,21)) ---
mk_cb_d <- function(varname, df_lag = 3, argvar = list(fun = "lin")) {
  X <- as.matrix(daily_spring[, paste0(varname, "_Lag", 0:DAILY_MAX), with = FALSE])
  crossbasis(X, lag = c(0, DAILY_MAX),
             argvar = argvar,
             arglag = list(fun = "ns", df = df_lag))
}
cb_runoff        <- mk_cb_d("runoff")
cb_soil_moisture <- mk_cb_d("soil_moisture")
cb_temperature   <- mk_cb_d("temperature")
cb_precipitation <- mk_cb_d("precipitation")
cb_wind_speed    <- mk_cb_d("wind_speed")
cb_snow_cover    <- mk_cb_d("snow_cover")
cb_anseriformes  <- crossbasis(
  as.matrix(daily_spring[, paste0("anseriformes_Lag", 0:DAILY_MAX), with = FALSE]),
  lag = c(0, DAILY_MAX), argvar = list(fun = "lin"),
  arglag = list(fun = "strata", breaks = c(7, 14, 21)))

daily_model <- clogit(
  outbreak_binary ~ cb_runoff + cb_soil_moisture + cb_temperature +
                    cb_precipitation + cb_wind_speed + cb_snow_cover +
                    cb_anseriformes + strata(stratum_id),
  data = daily_spring, method = "efron")
cat("Daily spring model: ", length(coef(daily_model)), " coefs | AIC ",
    round(AIC(daily_model), 2), "\n", sep = "")

# --- Weekly crossbases (lin × ns(df=2)) ---
mk_cb_w <- function(varname, df_lag = 2, argvar = list(fun = "lin")) {
  X <- as.matrix(weekly_spring[, paste0(varname, "_Lag", 0:WEEK_MAX), with = FALSE])
  crossbasis(X, lag = c(0, WEEK_MAX),
             argvar = argvar,
             arglag = list(fun = "ns", df = df_lag))
}
cb_runoff_A         <- mk_cb_w("runoff")
cb_soil_moisture_A  <- mk_cb_w("soil_moisture")
cb_temperature_A    <- mk_cb_w("temperature")
cb_precipitation_A  <- mk_cb_w("precipitation")
cb_wind_speed_A     <- mk_cb_w("wind_speed")
cb_snow_cover_A     <- mk_cb_w("snow_cover")
cb_anseriformes_A   <- mk_cb_w("anseriformes")

weekly_model_A <- clogit(
  outbreak_binary ~ cb_runoff_A + cb_soil_moisture_A + cb_temperature_A +
                    cb_precipitation_A + cb_wind_speed_A + cb_snow_cover_A +
                    cb_anseriformes_A + strata(stratum_id),
  data = weekly_spring, method = "efron")
cat("Weekly spring model: ", length(coef(weekly_model_A)), " coefs | AIC ",
    round(AIC(weekly_model_A), 2), "\n", sep = "")
```


# 3. Extract lag-specific + cumulative ORs from each model via manual contrast vectors

```{r extract-lag-response}
extract_lag_response <- function(model, cb_obj, prefix, lag_max) {
  beta <- coef(model); V <- vcov(model)
  cols <- grep(paste0("^", prefix, "v"), names(beta))
  b    <- beta[cols]; Vsub <- V[cols, cols]

  lag_seq   <- 0:lag_max
  lag_basis <- do.call(dlnm::onebasis,
                       c(list(x = lag_seq), attr(cb_obj, "arglag")))

  # Lag-specific (per-lag) — OR if +0.5 SD on that single lag day/week
  lag_specific <- data.table(
    lag = lag_seq,
    OR = NA_real_, OR_lo = NA_real_, OR_hi = NA_real_
  )
  for (i in seq_along(lag_seq)) {
    contrast <- 0.5 * lag_basis[i, ]
    lo <- as.numeric(contrast %*% b)
    se <- sqrt(as.numeric(t(contrast) %*% Vsub %*% contrast))
    lag_specific[i, `:=`(OR = exp(lo),
                         OR_lo = exp(lo - 1.96 * se),
                         OR_hi = exp(lo + 1.96 * se))]
  }

  # Cumulative — OR if +0.5 SD across lags 0..k
  cumulative <- data.table(
    lag = lag_seq,
    OR = NA_real_, OR_lo = NA_real_, OR_hi = NA_real_
  )
  for (i in seq_along(lag_seq)) {
    contrast <- 0.5 * colSums(lag_basis[1:i, , drop = FALSE])
    lo <- as.numeric(contrast %*% b)
    se <- sqrt(as.numeric(t(contrast) %*% Vsub %*% contrast))
    cumulative[i, `:=`(OR = exp(lo),
                       OR_lo = exp(lo - 1.96 * se),
                       OR_hi = exp(lo + 1.96 * se))]
  }
  list(lag_specific = lag_specific, cumulative = cumulative)
}

predictor_map <- list(
  list(name = "Runoff",        daily_cb = cb_runoff,        daily_prefix = "cb_runoff",
       weekly_cb = cb_runoff_A,         weekly_prefix = "cb_runoff_A"),
  list(name = "Soil Moisture", daily_cb = cb_soil_moisture, daily_prefix = "cb_soil_moisture",
       weekly_cb = cb_soil_moisture_A,  weekly_prefix = "cb_soil_moisture_A"),
  list(name = "Temperature",   daily_cb = cb_temperature,   daily_prefix = "cb_temperature",
       weekly_cb = cb_temperature_A,    weekly_prefix = "cb_temperature_A"),
  list(name = "Precipitation", daily_cb = cb_precipitation, daily_prefix = "cb_precipitation",
       weekly_cb = cb_precipitation_A,  weekly_prefix = "cb_precipitation_A"),
  list(name = "Wind Speed",    daily_cb = cb_wind_speed,    daily_prefix = "cb_wind_speed",
       weekly_cb = cb_wind_speed_A,     weekly_prefix = "cb_wind_speed_A"),
  list(name = "Snow Cover",    daily_cb = cb_snow_cover,    daily_prefix = "cb_snow_cover",
       weekly_cb = cb_snow_cover_A,     weekly_prefix = "cb_snow_cover_A"),
  list(name = "Anseriformes",  daily_cb = cb_anseriformes,  daily_prefix = "cb_anseriformes",
       weekly_cb = cb_anseriformes_A,   weekly_prefix = "cb_anseriformes_A")
)

build_plot_data <- function() {
  rbindlist(lapply(predictor_map, function(p) {
    d_res <- extract_lag_response(daily_model,    p$daily_cb,  p$daily_prefix,  DAILY_MAX)
    w_res <- extract_lag_response(weekly_model_A, p$weekly_cb, p$weekly_prefix, WEEK_MAX)

    # Convert weekly lag to days for unified x-axis
    w_res$lag_specific[, lag_days := lag * 7]
    w_res$cumulative[,   lag_days := lag * 7]
    d_res$lag_specific[, lag_days := lag]
    d_res$cumulative[,   lag_days := lag]

    rbind(
      d_res$lag_specific[, .(Predictor = p$name, Resolution = "Daily",  Type = "Lag-specific",
                              lag_days, OR, OR_lo, OR_hi)],
      w_res$lag_specific[, .(Predictor = p$name, Resolution = "Weekly", Type = "Lag-specific",
                              lag_days, OR, OR_lo, OR_hi)],
      d_res$cumulative[,   .(Predictor = p$name, Resolution = "Daily",  Type = "Cumulative",
                              lag_days, OR, OR_lo, OR_hi)],
      w_res$cumulative[,   .(Predictor = p$name, Resolution = "Weekly", Type = "Cumulative",
                              lag_days, OR, OR_lo, OR_hi)]
    )
  }))
}

plot_data <- build_plot_data()
plot_data[, Predictor := factor(Predictor,
  levels = c("Runoff", "Soil Moisture", "Temperature",
             "Precipitation", "Wind Speed", "Snow Cover", "Anseriformes"))]
```

# 4. Figure 1 — Lag-specific OR side-by-side comparison

```{r fig1-lag-specific, fig.width = 14, fig.height = 8}
# Filter to daily-only — weekly spring fit is unstable (near-separation; see Scope)
fig1_data <- plot_data[Type == "Lag-specific" & Resolution == "Daily"]

p1 <- ggplot(fig1_data, aes(x = lag_days, y = OR,
                            color = Resolution, fill = Resolution)) +
  geom_hline(yintercept = 1, linetype = "dashed", color = "gray40") +
  geom_ribbon(aes(ymin = OR_lo, ymax = OR_hi), alpha = 0.18, color = NA) +
  geom_line(linewidth = 0.9) +
  geom_point(data = fig1_data[Resolution == "Weekly"], size = 2) +
  scale_color_manual(values = c("Daily" = "black", "Weekly" = "firebrick")) +
  scale_fill_manual( values = c("Daily" = "black", "Weekly" = "firebrick")) +
  coord_cartesian(ylim = c(0, 4)) +
  scale_x_continuous(breaks = c(0, 7, 14, 21, 28),
                     labels = c("0", "7", "14", "21", "28")) +
  facet_wrap(~ Predictor, scales = "fixed", ncol = 3) +
  labs(
    x        = "Lag (days)",
    y        = "Odds Ratio (95% CI), per +0.5 SD on that lag day / week",
    color    = "Resolution",
    fill     = "Resolution",
    title    = "Lag-specific (per-lag) OR — daily DLNM (spring strata only)",
    subtitle = "Daily case-crossover lag 0–28 d, ns(df=3); CSLT excluded; strata limited to case-months Mar–May"
  ) +
  theme_minimal(base_size = 12) +
  theme(panel.grid.minor = element_blank(),
        legend.position  = "top",
        strip.text       = element_text(face = "bold"))

ggsave(paste0(figures_sensitivity_folder, "figure_daily_vs_weekly_lag_specific_spring.png"),
       p1, width = 14, height = 8, units = "in", dpi = 300)

print(p1)
```

# 5. Figure 2 — Cumulative OR side-by-side comparison

```{r fig2-cumulative, fig.width = 14, fig.height = 8}
fig2_data <- plot_data[Type == "Cumulative" & Resolution == "Daily"]

p2 <- ggplot(fig2_data, aes(x = lag_days, y = OR,
                            color = Resolution, fill = Resolution)) +
  geom_hline(yintercept = 1, linetype = "dashed", color = "gray40") +
  geom_ribbon(aes(ymin = OR_lo, ymax = OR_hi), alpha = 0.18, color = NA) +
  geom_line(linewidth = 0.9) +
  geom_point(data = fig2_data[Resolution == "Weekly"], size = 2) +
  scale_color_manual(values = c("Daily" = "black", "Weekly" = "firebrick")) +
  scale_fill_manual( values = c("Daily" = "black", "Weekly" = "firebrick")) +
  coord_cartesian(ylim = c(0, 4)) +
  scale_x_continuous(breaks = c(0, 7, 14, 21, 28),
                     labels = c("0", "7", "14", "21", "28")) +
  facet_wrap(~ Predictor, scales = "fixed", ncol = 3) +
  labs(
    x        = "Cumulative through lag (days)",
    y        = "Cumulative OR (95% CI), per +0.5 SD across lags 0…k",
    color    = "Resolution",
    fill     = "Resolution",
    title    = "Cumulative OR — daily DLNM (spring strata only)",
    subtitle = "Daily case-crossover lag 0–28 d, ns(df=3); CSLT excluded; strata limited to case-months Mar–May"
  ) +
  theme_minimal(base_size = 12) +
  theme(panel.grid.minor = element_blank(),
        legend.position  = "top",
        strip.text       = element_text(face = "bold"))

ggsave(paste0(figures_sensitivity_folder, "figure_daily_vs_weekly_cumulative_spring.png"),
       p2, width = 14, height = 8, units = "in", dpi = 300)

print(p2)
```

# 6. Save underlying data

```{r save}
saveRDS(list(
  plot_data    = plot_data,
  daily_model  = daily_model,
  weekly_model = weekly_model_A
), paste0(objects_folder, "daily_vs_weekly_comparison_spring.RDS"))

# Also export the merged plot data as CSV
fwrite(plot_data, paste0(tables_sensitivity_folder, "Table_Daily_vs_Weekly_LagResponse_spring.csv"))
```
