---
title: "Primary Case-Crossover DLNM Model (Daily)"
author: "Fintan Mooney"
date: "2026-05-12"
output: html_document
---
# Primary DAILY time-stratified case-crossover DLNM model for HPAI outbreak risk
# in MN poultry farms (May 2026 GeoHealth revision). Fits distributed lag
# nonlinear models for meteorological, environmental, and wild bird abundance
# predictors over lag 0–28 days, stratified by zone × year × month × day-of-week.

```{r include=FALSE}
rm(list=ls())
project.folder = paste0(print(here::here()),'/')
source(paste0(project.folder,'create_folder_structure.R'))
source(paste0(functions.folder,'script_initiate.R'))
```

## Import casecrossover df (daily, time-stratified)

```{r, import casecrossover df}

case_crossover_df <- readRDS(file.path(objects_folder, "case_crossover_df.RDS"))

```

## DLNM specification (constants)

```{r, dlnm config}

# Lag dimension: 0–28 days, natural cubic spline with 4 df (Lu et al. 2020).
# Exposure dimension: linear (per 0.5 SD), matching the submitted manuscript's
# coefficient interpretation. The "flexible" variant raises arglag df to 5 and
# allows polynomial argvar for selected predictors (sensitivity).
MAX_LAG <- 28
ARGLAG_PRIMARY <- list(fun = "ns", df = 4)
ARGVAR_LIN     <- list(fun = "lin")

make_cb <- function(varname, df_lag = 4, argvar = ARGVAR_LIN) {
  X <- as.matrix(case_crossover_df[, paste0(varname, "_Lag", 0:MAX_LAG), with = FALSE])
  crossbasis(X, lag = c(0, MAX_LAG),
             argvar = argvar,
             arglag = list(fun = "ns", df = df_lag))
}

```


## Crossbasis Terms (lag 0–28 days, ns df=4)

```{r, finalized crossbasis}

cb_leaf_area_index      <- make_cb("leaf_area_index")
cb_longwave_radiation   <- make_cb("longwave_radiation")
cb_precipitation        <- make_cb("precipitation")
cb_relative_humidity    <- make_cb("relative_humidity")
cb_runoff               <- make_cb("runoff")
cb_shortwave_radiation  <- make_cb("shortwave_radiation")
cb_snow_density         <- make_cb("snow_density")
cb_soil_moisture        <- make_cb("soil_moisture")
cb_temperature          <- make_cb("temperature")
cb_total_bird_abundance <- make_cb("total_bird_abundance")
cb_wind_speed           <- make_cb("wind_speed")

```

### Flexible crossbasis (higher df / polynomial argvar — sensitivity only)

```{r, more flexible crossbasis terms}

cb_flex_leaf_area_index      <- make_cb("leaf_area_index",      df_lag = 5)
cb_flex_longwave_radiation   <- make_cb("longwave_radiation",   df_lag = 5)
cb_flex_precipitation        <- make_cb("precipitation",        df_lag = 5)
cb_flex_relative_humidity    <- make_cb("relative_humidity",    df_lag = 5)
cb_flex_runoff               <- make_cb("runoff",               df_lag = 5)
cb_flex_shortwave_radiation  <- make_cb("shortwave_radiation",  df_lag = 5)
cb_flex_snow_density         <- make_cb("snow_density",         df_lag = 5)
cb_flex_soil_moisture        <- make_cb("soil_moisture",        df_lag = 5)
cb_flex_temperature          <- make_cb("temperature",          df_lag = 5)
cb_flex_total_bird_abundance <- make_cb("total_bird_abundance", df_lag = 5)
cb_flex_wind_speed           <- make_cb("wind_speed",            df_lag = 5)

```


## Case-Crossover Model

```{r, casecrossover model}

# ----------------------------
# Model Fitting
# ----------------------------
final_model <- clogit(
  outbreak_binary ~ 
    cb_runoff + cb_snow_density + cb_soil_moisture + cb_temperature +
    cb_precipitation + cb_wind_speed + cb_total_bird_abundance +
    strata(stratum_id),
  method = "efron",
  data = case_crossover_df
)

summary(final_model)
AIC(final_model)
saveRDS(final_model, file.path(models_folder, "casecrossover_model.RDS"))

```

```{r, crosspred objects}
# ----------------------------
# Crosspred Objects
# ----------------------------
cp_runoff       <- crosspred(cb_runoff, final_model, cen = 0, at = 0.5, cumul = TRUE)
cp_snow         <- crosspred(cb_snow_density, final_model, cen = 0, at = 0.5, cumul = TRUE)
cp_soil         <- crosspred(cb_soil_moisture, final_model, cen = 0, at = 0.5, cumul = TRUE)
cp_temp         <- crosspred(cb_temperature, final_model, cen = 0, at = 0.5, cumul = TRUE)
cp_precip       <- crosspred(cb_precipitation, final_model, cen = 0, at = 0.5, cumul = TRUE)
cp_wind         <- crosspred(cb_wind_speed, final_model, cen = 0, at = 0.5, cumul = TRUE)
cp_bird         <- crosspred(cb_total_bird_abundance, final_model, cen = 0, at = 0.5, cumul = TRUE)

cp_slice_runoff <- crosspred(cb_runoff, final_model, cen = 0, at = 0.5)
cp_slice_snow   <- crosspred(cb_snow_density, final_model, cen = 0, at = 0.5)
cp_slice_soil   <- crosspred(cb_soil_moisture, final_model, cen = 0, at = 0.5)
cp_slice_temp   <- crosspred(cb_temperature, final_model, cen = 0, at = 0.5)
cp_slice_precip <- crosspred(cb_precipitation, final_model, cen = 0, at = 0.5)
cp_slice_wind   <- crosspred(cb_wind_speed, final_model, cen = 0, at = 0.5)
cp_slice_bird   <- crosspred(cb_total_bird_abundance, final_model, cen = 0, at = 0.5)

```

## Casecrossover Plots
```{r, plots casecrossover}
# ----------------------------
# Cumulative OR Summary Table
# ----------------------------
cumulative_summary <- tibble(
  Predictor = c("Runoff", "Snow Density", "Soil Moisture", "Temperature",
                "Precipitation", "Wind Speed", "Total Bird Abundance"),
      OR = c(
        cp_runoff$allRRfit["0.5"],
        cp_snow$allRRfit["0.5"],
        cp_soil$allRRfit["0.5"],
        cp_temp$allRRfit["0.5"],
        cp_precip$allRRfit["0.5"],
        cp_wind$allRRfit["0.5"],
        cp_bird$allRRfit["0.5"]
      ),
      CI_low = c(
        cp_runoff$allRRlow["0.5"],
        cp_snow$allRRlow["0.5"],
        cp_soil$allRRlow["0.5"],
        cp_temp$allRRlow["0.5"],
        cp_precip$allRRlow["0.5"],
        cp_wind$allRRlow["0.5"],
        cp_bird$allRRlow["0.5"]
      ),
      CI_high = c(
        cp_runoff$allRRhigh["0.5"],
        cp_snow$allRRhigh["0.5"],
        cp_soil$allRRhigh["0.5"],
        cp_temp$allRRhigh["0.5"],
        cp_precip$allRRhigh["0.5"],
        cp_wind$allRRhigh["0.5"],
        cp_bird$allRRhigh["0.5"]
      )
) %>%
  mutate(`OR (95% CI)` = sprintf("%.2f (%.2f, %.2f)", OR, CI_low, CI_high)) %>%
  select(Predictor, `OR (95% CI)`) %>%
  gt() %>%
  tab_header(
    title = md("**Case-Crossover Model: Cumulative Odds Ratios**"),
    subtitle = "Per 0.5 SD Increase"
  ) %>%
  cols_label(`OR (95% CI)` = "Odds Ratio (95% Confidence Interval)") %>%
  tab_options(
    table.font.size = 14,
    heading.title.font.size = 16,
    heading.subtitle.font.size = 14,
    data_row.padding = px(6)
  )

# ----------------------------
# Plotting Functions
# ----------------------------
plot_cumulative <- function(cp, title) {
  plot(cp, type = "cumulative", var = 0.5, ci = "area",
       col = "black", ci.arg = list(col = rgb(0,0,0,0.1), border = NA),
       lwd = 1.5, main = title, xlab = "", ylab = "", las = 1)
  abline(h = 1, lty = 2, col = "gray40")
}

plot_slices <- function(cp, title) {
  plot(cp, type = "slices", var = 0.5, ci = "area",
       col = "black", ci.arg = list(col = rgb(0,0,0,0.1), border = NA),
       lwd = 1.3, main = title, xlab = "", ylab = "", las = 1)
  abline(h = 1, lty = 2, col = "gray40")
}

# ----------------------------
# Save Cumulative Plot
# ----------------------------
png(file.path(figures_folder,"figure_cumulative_ORs_casecrossover.png"), width = 11, height = 7, units = "in", res = 300)

par(
  mfrow = c(2, 4),
  mar = c(4.2, 4.2, 2.5, 1),
  oma = c(3, 4, 0, 1),
  cex.main = 1.2,
  cex.lab = 1.2,
  cex.axis = 1.1,
  tcl = -0.3, mgp = c(2.5, 0.7, 0)
)

plot_cumulative(cp_runoff, "Runoff")
plot_cumulative(cp_snow, "Snow Density")
plot_cumulative(cp_soil, "Soil Moisture")
plot_cumulative(cp_temp, "Temperature")
plot_cumulative(cp_precip, "Precipitation")
plot_cumulative(cp_wind, "Wind Speed")
plot_cumulative(cp_bird, "Total Bird Abundance")

mtext("Lag (days)", side = 1, outer = TRUE, line = 0, cex = 1.2)
mtext("Odds Ratio (95% CI)", side = 2, outer = TRUE, line = 0, cex = 1.2)

dev.off()

# ----------------------------
# Save Slices Plot
# ----------------------------
png(file.path(figures_folder,"figure_slice_ORs_casecrossover.png"), width = 11, height = 7, units = "in", res = 300)

par(
  mfrow = c(2, 4),
  mar = c(4.2, 4.2, 2.5, 1),
  oma = c(3, 4, 0, 1),
  cex.main = 1.2,
  cex.lab = 1.2,
  cex.axis = 1.1,
  tcl = -0.3, mgp = c(2.5, 0.7, 0)
)

plot_slices(cp_slice_runoff, "Runoff")
plot_slices(cp_slice_snow, "Snow Density")
plot_slices(cp_slice_soil, "Soil Moisture")
plot_slices(cp_slice_temp, "Temperature")
plot_slices(cp_slice_precip, "Precipitation")
plot_slices(cp_slice_wind, "Wind Speed")
plot_slices(cp_slice_bird, "Total Bird Abundance")

mtext("Lag (days)", side = 1, outer = TRUE, line = 0, cex = 1.2)
mtext("Odds Ratio (95% CI)", side = 2, outer = TRUE, line = 0, cex = 1.2)

dev.off()


```


## Lag Specific Odds Ratio Table (Daily — weekly anchor lags)

```{r, lag specific odds ratio table}
# Display lag-specific ORs at weekly anchor points (0, 7, 14, 21, 28 days).
LAG_ANCHORS <- c(0, 7, 14, 21, 28)

extract_lag_ORs_wide <- function(cp_obj, varname, varval = "0.5", lags = LAG_ANCHORS) {
  lag_labels <- paste0("lag", lags)
  or    <- cp_obj$matRRfit[varval, lag_labels]
  low   <- cp_obj$matRRlow[varval, lag_labels]
  high  <- cp_obj$matRRhigh[varval, lag_labels]
  col_labels <- paste0("Lag ", lags, " d")
  tibble(
    Predictor = varname,
    !!!setNames(as.list(or),   paste0("OR_Lag", lags)),
    !!!setNames(as.list(low),  paste0("Low_Lag", lags)),
    !!!setNames(as.list(high), paste0("High_Lag", lags)),
    !!!setNames(as.list(sprintf("%.2f (%.2f, %.2f)", or, low, high)), col_labels)
  )
}

table_lag_ORs <- bind_rows(
  extract_lag_ORs_wide(cp_slice_runoff, "Runoff"),
  extract_lag_ORs_wide(cp_slice_snow,   "Snow Density"),
  extract_lag_ORs_wide(cp_slice_soil,   "Soil Moisture"),
  extract_lag_ORs_wide(cp_slice_temp,   "Temperature"),
  extract_lag_ORs_wide(cp_slice_precip, "Precipitation"),
  extract_lag_ORs_wide(cp_slice_wind,   "Wind Speed"),
  extract_lag_ORs_wide(cp_slice_bird,   "Total Bird Abundance")
)

# Highlight significant rows
col_labels <- paste0("Lag ", LAG_ANCHORS, " d")
highlight_matrix <- sapply(LAG_ANCHORS, function(i) {
  table_lag_ORs[[paste0("Low_Lag", i)]] > 1 | table_lag_ORs[[paste0("High_Lag", i)]] < 1
})
colnames(highlight_matrix) <- col_labels

ft <- table_lag_ORs %>%
  select(Predictor, all_of(col_labels)) %>%
  flextable() %>%
  set_header_labels(Predictor = "Predictor") %>%
  add_header_row(values = c("Predictor", "Odds Ratio (95% CI)"),
                 colwidths = c(1, length(col_labels))) %>%
  fontsize(size = 11, part = "all") %>%
  autofit() %>%
  align(align = "center", part = "all") %>%
  theme_booktabs()

for (k in seq_along(LAG_ANCHORS)) {
  ft <- bold(ft, i = which(highlight_matrix[, k]), j = col_labels[k])
}

doc <- read_docx() %>%
  body_add_flextable(ft) %>%
  body_add_par(
    "Table: Lag-Specific Odds Ratios at Weekly Anchor Lags (0, 7, 14, 21, 28 days) for the Daily Time-Stratified Case-Crossover Model of HPAI Outbreaks in Minnesota, 2022. Estimates are per 0.5 standard deviation increase in each predictor. Statistically significant odds ratios are shown in bold.",
    style = "Normal"
  )

print(doc, target = file.path(tables_folder, "Table_LagSpecific_OR_HPAI_Casecrossover_daily.docx"))

ft
```

### Cumulative odds ratio table (cumulative through anchor lags)

```{r, cumulative odds ratios}
ft_lagspecific <- ft

# Cumulative OR through each weekly anchor lag (0, 7, 14, 21, 28).
extract_cum_ORs_wide <- function(cp_obj, varname, varval = "0.5", lags = LAG_ANCHORS) {
  lag_labels <- paste0("lag", lags)
  or    <- cp_obj$cumRRfit[varval, lag_labels]
  low   <- cp_obj$cumRRlow[varval, lag_labels]
  high  <- cp_obj$cumRRhigh[varval, lag_labels]
  col_labels <- paste0("Cum ≤ ", lags, " d")
  tibble(
    Predictor = varname,
    !!!setNames(as.list(or),   paste0("OR_Lag", lags)),
    !!!setNames(as.list(low),  paste0("Low_Lag", lags)),
    !!!setNames(as.list(high), paste0("High_Lag", lags)),
    !!!setNames(as.list(sprintf("%.2f (%.2f, %.2f)", or, low, high)), col_labels)
  )
}

table_lag_ORs_cum <- bind_rows(
  extract_cum_ORs_wide(cp_runoff, "Runoff"),
  extract_cum_ORs_wide(cp_snow,   "Snow Density"),
  extract_cum_ORs_wide(cp_soil,   "Soil Moisture"),
  extract_cum_ORs_wide(cp_temp,   "Temperature"),
  extract_cum_ORs_wide(cp_precip, "Precipitation"),
  extract_cum_ORs_wide(cp_wind,   "Wind Speed"),
  extract_cum_ORs_wide(cp_bird,   "Total Bird Abundance")
)

col_labels_cum <- paste0("Cum ≤ ", LAG_ANCHORS, " d")
highlight_matrix <- sapply(LAG_ANCHORS, function(i) {
  table_lag_ORs_cum[[paste0("Low_Lag", i)]] > 1 | table_lag_ORs_cum[[paste0("High_Lag", i)]] < 1
})
colnames(highlight_matrix) <- col_labels_cum

ft_cum <- table_lag_ORs_cum %>%
  select(Predictor, all_of(col_labels_cum)) %>%
  flextable() %>%
  set_header_labels(Predictor = "Predictor") %>%
  add_header_row(values = c("Predictor", "Cumulative Odds Ratio (95% CI)"),
                 colwidths = c(1, length(col_labels_cum))) %>%
  fontsize(size = 11, part = "all") %>%
  autofit() %>%
  align(align = "center", part = "all") %>%
  theme_booktabs()

for (k in seq_along(LAG_ANCHORS)) {
  ft_cum <- bold(ft_cum, i = which(highlight_matrix[, k]), j = col_labels_cum[k])
}

doc <- read_docx() %>%
  body_add_flextable(ft_cum) %>%
  body_add_par(
    "Table: Cumulative Odds Ratios through weekly anchor lags (≤0, ≤7, ≤14, ≤21, ≤28 days) for the daily time-stratified case-crossover model of HPAI outbreaks in Minnesota, 2022. Estimates are per 0.5 SD increase in each predictor. Statistically significant ORs in bold.",
    style = "Normal"
  )

print(doc, target = file.path(tables_folder, "Table_Cumulative_OR_HPAI_Casscrossover_daily.docx"))

ft_cum
```

#### Combine cumulative and lag-specific odds ratio tables

```{r, combine and lagspecific odds ratio tables}

panel_label <- function(txt, size = 11) {
  fp <- fp_text(bold = TRUE, font.size = size)
  fpar(ftext(txt, fp))
}

doc <- read_docx() %>%
  # Panel A
  body_add_fpar(panel_label("A. Lag-specific odds ratios (per +0.5 SD)")) %>%
  body_add_flextable(ft_lagspecific) %>%
  body_add_par("", style = "Normal") %>%   # spacer

  # Panel B
  body_add_fpar(panel_label("B. Cumulative odds ratios (per +0.5 SD)")) %>%
  body_add_flextable(ft_cum)

print(doc, target = file.path(tables_folder, "Table_LagSpecific_and_Cumulative_ORs_Casecrossover.docx"))


```



### Cumulative odds ratio by weekly lag window (0–7d, 8–14d, 15–21d, 22–28d)

```{r, cumulative odds ratio lag sections}

# Weekly lag windows expressed in days. Each window is fitted via a fresh
# crosspred over the specified [from, to] daily lag range.
lag_windows <- list(
  "Lag 0–7 d"   = c(0, 7),
  "Lag 8–14 d"  = c(8, 14),
  "Lag 15–21 d" = c(15, 21),
  "Lag 22–28 d" = c(22, 28)
)

cb_list <- list(
  "Surface Runoff"       = cb_runoff,
  "Snow Density"         = cb_snow_density,
  "Soil Moisture"        = cb_soil_moisture,
  "Temperature"          = cb_temperature,
  "Precipitation"        = cb_precipitation,
  "Wind Speed"           = cb_wind_speed,
  "Total Bird Abundance" = cb_total_bird_abundance
)

window_cum_OR <- function(cb, model, lo, hi) {
  cp <- crosspred(cb, model, cen = 0, at = 0.5,
                  cumul = TRUE, from = lo, to = hi)
  idx <- hi - lo + 1
  sprintf("%.2f (%.2f, %.2f)",
          cp$cumRRfit["0.5", idx],
          cp$cumRRlow["0.5", idx],
          cp$cumRRhigh["0.5", idx])
}

table_cum_ORs <- tibble(
  Predictor = names(cb_list),
  !!!setNames(
    lapply(lag_windows, function(w) {
      vapply(cb_list, function(cb) window_cum_OR(cb, final_model, w[1], w[2]),
             character(1))
    }),
    names(lag_windows)
  )
)

# Highlight significant cells
parse_ci <- function(s) {
  parts <- as.numeric(unlist(strsplit(gsub("[()]", "", s), "[ ,]+")))
  c(low = parts[2], high = parts[3])
}
highlight_matrix <- sapply(names(lag_windows), function(w) {
  vapply(table_cum_ORs[[w]], function(s) {
    ci <- parse_ci(s); ci["low"] > 1 | ci["high"] < 1
  }, logical(1))
})
colnames(highlight_matrix) <- names(lag_windows)

ft_cum <- table_cum_ORs %>%
  flextable() %>%
  set_header_labels(Predictor = "Predictor") %>%
  add_header_row(values = c("Predictor", "Cumulative Odds Ratio (95% CI)"),
                 colwidths = c(1, length(lag_windows))) %>%
  fontsize(size = 11, part = "all") %>%
  autofit() %>%
  align(align = "center", part = "all") %>%
  theme_booktabs()

for (w in names(lag_windows)) {
  ft_cum <- bold(ft_cum, i = which(highlight_matrix[, w]), j = w)
}

doc <- read_docx() %>%
  body_add_flextable(ft_cum) %>%
  body_add_par(
    "Table: Cumulative Odds Ratios by Weekly Lag Window for the Daily Time-Stratified Case-Crossover Model of HPAI Outbreaks in Minnesota, 2022. Each cell reports the cumulative OR over the specified daily lag window per 0.5 SD increase in each predictor. Statistically significant associations are shown in bold.",
    style = "Normal"
  )

print(doc, target = file.path(tables_folder, "Table_Cumulative_OR_HPAI_Minnesota_daily.docx"))

ft_cum
```

## Case-Crossover Model with Flexible Crossbasis

```{r, casecrossover flex model}

# ----------------------------
# Model Fitting
# ----------------------------
final_model_flex <- clogit(
  outbreak_binary ~ 
    cb_flex_runoff + cb_flex_snow_density + cb_flex_soil_moisture + cb_flex_temperature +
    cb_flex_precipitation + cb_flex_wind_speed + cb_flex_total_bird_abundance +
    strata(stratum_id),
  method = "efron",
  data = case_crossover_df
)

summary(final_model_flex)
AIC(final_model_flex)
saveRDS(final_model_flex, file.path(models_folder, "casecrossover_flexible_model.RDS"))

```

```{r, crosspred objects for flex model}
# ----------------------------
# Crosspred Objects
# ----------------------------
cp_flex_runoff       <- crosspred(cb_flex_runoff, final_model_flex, cen = 0, at = 0.5, cumul = TRUE)
cp_flex_snow         <- crosspred(cb_flex_snow_density, final_model_flex, cen = 0, at = 0.5, cumul = TRUE)
cp_flex_soil         <- crosspred(cb_flex_soil_moisture, final_model_flex, cen = 0, at = 0.5, cumul = TRUE)
cp_flex_temp         <- crosspred(cb_flex_temperature, final_model_flex, cen = 0, at = 0.5, cumul = TRUE)
cp_flex_precip       <- crosspred(cb_flex_precipitation, final_model_flex, cen = 0, at = 0.5, cumul = TRUE)
cp_flex_wind         <- crosspred(cb_flex_wind_speed, final_model_flex, cen = 0, at = 0.5, cumul = TRUE)
cp_flex_bird         <- crosspred(cb_flex_total_bird_abundance, final_model_flex, cen = 0, at = 0.5, cumul = TRUE)

cp_flex_slice_runoff <- crosspred(cb_flex_runoff, final_model_flex, cen = 0, at = 0.5)
cp_flex_slice_snow   <- crosspred(cb_flex_snow_density, final_model_flex, cen = 0, at = 0.5)
cp_flex_slice_soil   <- crosspred(cb_flex_soil_moisture, final_model_flex, cen = 0, at = 0.5)
cp_flex_slice_temp   <- crosspred(cb_flex_temperature, final_model_flex, cen = 0, at = 0.5)
cp_flex_slice_precip <- crosspred(cb_flex_precipitation, final_model_flex, cen = 0, at = 0.5)
cp_flex_slice_wind   <- crosspred(cb_flex_wind_speed, final_model_flex, cen = 0, at = 0.5)
cp_flex_slice_bird   <- crosspred(cb_flex_total_bird_abundance, final_model_flex, cen = 0, at = 0.5)

```

## Casecrossover Plots
```{r, plots casecrossover flex}
# ----------------------------
# Cumulative OR Summary Table
# ----------------------------
cumulative_summary <- tibble(
  Predictor = c("Surface Runoff", "Snow Density", "Soil Moisture", "Temperature",
                "Precipitation", "Wind Speed", "Total Bird Abundance"),
      OR = c(
        cp_flex_runoff$allRRfit["0.5"],
        cp_flex_snow$allRRfit["0.5"],
        cp_flex_soil$allRRfit["0.5"],
        cp_flex_temp$allRRfit["0.5"],
        cp_flex_precip$allRRfit["0.5"],
        cp_flex_wind$allRRfit["0.5"],
        cp_flex_bird$allRRfit["0.5"]
      ),
      CI_low = c(
        cp_flex_runoff$allRRlow["0.5"],
        cp_flex_snow$allRRlow["0.5"],
        cp_flex_soil$allRRlow["0.5"],
        cp_flex_temp$allRRlow["0.5"],
        cp_flex_precip$allRRlow["0.5"],
        cp_flex_wind$allRRlow["0.5"],
        cp_flex_bird$allRRlow["0.5"]
      ),
      CI_high = c(
        cp_flex_runoff$allRRhigh["0.5"],
        cp_flex_snow$allRRhigh["0.5"],
        cp_flex_soil$allRRhigh["0.5"],
        cp_flex_temp$allRRhigh["0.5"],
        cp_flex_precip$allRRhigh["0.5"],
        cp_flex_wind$allRRhigh["0.5"],
        cp_flex_bird$allRRhigh["0.5"]
      )
) %>%
  mutate(`OR (95% CI)` = sprintf("%.2f (%.2f, %.2f)", OR, CI_low, CI_high)) %>%
  select(Predictor, `OR (95% CI)`) %>%
  gt() %>%
  tab_header(
    title = md("**Case-Crossover Model: Cumulative Odds Ratios**"),
    subtitle = "Per 0.5 SD Increase"
  ) %>%
  cols_label(`OR (95% CI)` = "Odds Ratio (95% Confidence Interval)") %>%
  tab_options(
    table.font.size = 14,
    heading.title.font.size = 16,
    heading.subtitle.font.size = 14,
    data_row.padding = px(6)
  )

cumulative_summary

# ----------------------------
# Plotting Functions
# ----------------------------
plot_cumulative <- function(cp_flex_, title) {
  plot(cp_flex_, type = "cumulative", var = 0.5, ci = "area",
       col = "black", ci.arg = list(col = rgb(0,0,0,0.1), border = NA),
       lwd = 1.5, main = title, xlab = "", ylab = "", las = 1)
  abline(h = 1, lty = 2, col = "gray40")
}

plot_slices <- function(cp_flex_, title) {
  plot(cp_flex_, type = "slices", var = 0.5, ci = "area",
       col = "black", ci.arg = list(col = rgb(0,0,0,0.1), border = NA),
       lwd = 1.3, main = title, xlab = "", ylab = "", las = 1)
  abline(h = 1, lty = 2, col = "gray40")
}

# ----------------------------
# Save Cumulative Plot
# ----------------------------
png(file.path(figures_folder,"figure_cumulative_ORs_casecrossover_flexmodel.png"), width = 11, height = 7, units = "in", res = 300)

par(
  mfrow = c(2, 4),
  mar = c(4.2, 4.2, 2.5, 1),
  oma = c(3, 4, 0, 1),
  cex.main = 1.2,
  cex.lab = 1.2,
  cex.axis = 1.1,
  tcl = -0.3, mgp = c(2.5, 0.7, 0)
)

plot_cumulative(cp_flex_runoff, "Runoff")
plot_cumulative(cp_flex_snow, "Snow Density")
plot_cumulative(cp_flex_soil, "Soil Moisture")
plot_cumulative(cp_flex_temp, "Temperature")
plot_cumulative(cp_flex_precip, "Precipitation")
plot_cumulative(cp_flex_wind, "Wind Speed")
plot_cumulative(cp_flex_bird, "Total Bird Abundance")

mtext("Lag (days)", side = 1, outer = TRUE, line = 0, cex = 1.2)
mtext("Odds Ratio (95% CI)", side = 2, outer = TRUE, line = 0, cex = 1.2)

dev.off()

# ----------------------------
# Save Slices Plot
# ----------------------------
png(file.path(figures_folder,"figure_slice_ORs_casecrossover_flexmodel.png"), width = 11, height = 7, units = "in", res = 300)

par(
  mfrow = c(2, 4),
  mar = c(4.2, 4.2, 2.5, 1),
  oma = c(3, 4, 0, 1),
  cex.main = 1.2,
  cex.lab = 1.2,
  cex.axis = 1.1,
  tcl = -0.3, mgp = c(2.5, 0.7, 0)
)

plot_slices(cp_flex_slice_runoff, "Runoff")
plot_slices(cp_flex_slice_snow, "Snow Density")
plot_slices(cp_flex_slice_soil, "Soil Moisture")
plot_slices(cp_flex_slice_temp, "Temperature")
plot_slices(cp_flex_slice_precip, "Precipitation")
plot_slices(cp_flex_slice_wind, "Wind Speed")
plot_slices(cp_flex_slice_bird, "Total Bird Abundance")

mtext("Lag (days)", side = 1, outer = TRUE, line = 0, cex = 1.2)
mtext("Odds Ratio (95% CI)", side = 2, outer = TRUE, line = 0, cex = 1.2)

dev.off()

```


## Lag-Specific Table (Flex Model — weekly anchor lags)

```{r, lag specific odds ratio table for flex model}
extract_lag_ORs_wide <- function(cp_flex_obj, varname, varval = "0.5", lags = LAG_ANCHORS) {
  lag_labels <- paste0("lag", lags)
  or    <- cp_flex_obj$matRRfit[varval, lag_labels]
  low   <- cp_flex_obj$matRRlow[varval, lag_labels]
  high  <- cp_flex_obj$matRRhigh[varval, lag_labels]
  col_labels <- paste0("Lag ", lags, " d")
  tibble(
    Predictor = varname,
    !!!setNames(as.list(or),   paste0("OR_Lag", lags)),
    !!!setNames(as.list(low),  paste0("Low_Lag", lags)),
    !!!setNames(as.list(high), paste0("High_Lag", lags)),
    !!!setNames(as.list(sprintf("%.2f (%.2f, %.2f)", or, low, high)), col_labels)
  )
}

table_lag_ORs <- bind_rows(
  extract_lag_ORs_wide(cp_flex_slice_runoff, "Runoff"),
  extract_lag_ORs_wide(cp_flex_slice_snow,   "Snow Density"),
  extract_lag_ORs_wide(cp_flex_slice_soil,   "Soil Moisture"),
  extract_lag_ORs_wide(cp_flex_slice_temp,   "Temperature"),
  extract_lag_ORs_wide(cp_flex_slice_precip, "Precipitation"),
  extract_lag_ORs_wide(cp_flex_slice_wind,   "Wind Speed"),
  extract_lag_ORs_wide(cp_flex_slice_bird,   "Total Bird Abundance")
)

col_labels <- paste0("Lag ", LAG_ANCHORS, " d")
highlight_matrix <- sapply(LAG_ANCHORS, function(i) {
  table_lag_ORs[[paste0("Low_Lag", i)]] > 1 | table_lag_ORs[[paste0("High_Lag", i)]] < 1
})
colnames(highlight_matrix) <- col_labels

ft <- table_lag_ORs %>%
  select(Predictor, all_of(col_labels)) %>%
  flextable() %>%
  set_header_labels(Predictor = "Predictor") %>%
  add_header_row(values = c("Predictor", "Odds Ratio (95% CI)"),
                 colwidths = c(1, length(col_labels))) %>%
  fontsize(size = 11, part = "all") %>%
  autofit() %>%
  align(align = "center", part = "all") %>%
  theme_booktabs()

for (k in seq_along(LAG_ANCHORS)) {
  ft <- bold(ft, i = which(highlight_matrix[, k]), j = col_labels[k])
}

doc <- read_docx() %>%
  body_add_flextable(ft) %>%
  body_add_par(
    "Table: Lag-Specific Odds Ratios at Weekly Anchor Lags (Flexible DLNM) for HPAI Outbreaks in Minnesota, 2022. Estimates are per 0.5 SD increase in each predictor. Statistically significant odds ratios are shown in bold.",
    style = "Normal"
  )

print(doc, target = file.path(tables_folder, "Table_LagSpecific_OR_HPAI_Minnesota_flexmodel_daily.docx"))

ft
```



## Cumulative odds ratio by weekly lag window (Flex Model)

```{r, cumulative odds ratio lag sections}

cb_flex_list <- list(
  "Surface Runoff"       = cb_flex_runoff,
  "Snow Density"         = cb_flex_snow_density,
  "Soil Moisture"        = cb_flex_soil_moisture,
  "Temperature"          = cb_flex_temperature,
  "Precipitation"        = cb_flex_precipitation,
  "Wind Speed"           = cb_flex_wind_speed,
  "Total Bird Abundance" = cb_flex_total_bird_abundance
)

table_cum_ORs <- tibble(
  Predictor = names(cb_flex_list),
  !!!setNames(
    lapply(lag_windows, function(w) {
      vapply(cb_flex_list, function(cb) window_cum_OR(cb, final_model_flex, w[1], w[2]),
             character(1))
    }),
    names(lag_windows)
  )
)

highlight_matrix <- sapply(names(lag_windows), function(w) {
  vapply(table_cum_ORs[[w]], function(s) {
    ci <- parse_ci(s); ci["low"] > 1 | ci["high"] < 1
  }, logical(1))
})
colnames(highlight_matrix) <- names(lag_windows)

ft_cum <- table_cum_ORs %>%
  flextable() %>%
  set_header_labels(Predictor = "Predictor") %>%
  add_header_row(values = c("Predictor", "Cumulative Odds Ratio (95% CI)"),
                 colwidths = c(1, length(lag_windows))) %>%
  fontsize(size = 11, part = "all") %>%
  autofit() %>%
  align(align = "center", part = "all") %>%
  theme_booktabs()

for (w in names(lag_windows)) {
  ft_cum <- bold(ft_cum, i = which(highlight_matrix[, w]), j = w)
}

doc <- read_docx() %>%
  body_add_flextable(ft_cum) %>%
  body_add_par(
    "Table: Cumulative Odds Ratios by Weekly Lag Window (Flexible DLNM) for HPAI Outbreaks in Minnesota, 2022. Each cell reports the cumulative OR over the specified daily lag window per 0.5 SD increase. Significant ORs in bold.",
    style = "Normal"
  )

print(doc, target = file.path(tables_folder, "Table_Cumulative_OR_HPAI_Minnesota_flexmodel_daily.docx"))

ft_cum
```

