---
title: "Primary case-crossover DLNM model (daily)"
author: "Fintan Mooney"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

```{r initiate}
rm(list = ls())
# 0a.Declare root directory, folder location and load essential stuff
project.folder = paste0(print(here::here()), '/')
source(paste0(project.folder, 'create_folder_structure.R'))
source(paste0(functions.folder, 'script_initiate.R'))
```

# 1. Load case-crossover dataframe

```{r, import casecrossover df}

case_crossover_df <- readRDS(file.path(objects_folder, "case_crossover_df.RDS"))

```

## DLNM specification (constants)

```{r, dlnm config}

# Lag dimension: 0–14 days, natural cubic spline with 4 df (Lu et al. 2020).
# Exposure dimension: linear (per 0.5 SD), matching the submitted manuscript's
# coefficient interpretation. The "flexible" variant raises arglag df to 5 and
# allows polynomial argvar for selected predictors (sensitivity).
MAX_LAG <- 28
ARGLAG_PRIMARY <- list(fun = "ns", df = 3)
ARGVAR_LIN     <- list(fun = "lin")

make_cb <- function(varname, df_lag = 3, argvar = ARGVAR_LIN) {
  X <- as.matrix(case_crossover_df[, paste0(varname, "_Lag", 0:MAX_LAG), with = FALSE])
  crossbasis(X, lag = c(0, MAX_LAG),
             argvar = argvar,
             arglag = list(fun = "ns", df = df_lag))
}

```


## Crossbasis Terms (lag 0–28 days, ns df=4 for met; strata weekly for bird)

```{r, finalized crossbasis}

cb_leaf_area_index      <- make_cb("leaf_area_index")
cb_longwave_radiation   <- make_cb("longwave_radiation")
cb_precipitation        <- make_cb("precipitation")
cb_relative_humidity    <- make_cb("relative_humidity")
cb_runoff               <- make_cb("runoff")
cb_shortwave_radiation  <- make_cb("shortwave_radiation")
cb_snow_cover           <- make_cb("snow_cover")
cb_soil_moisture        <- make_cb("soil_moisture")
cb_temperature          <- make_cb("temperature")
cb_wind_speed           <- make_cb("wind_speed")

# Bird abundance: weekly eBird linearly interpolated to daily, so daily values
# are nearly collinear within each 7-day window. Use a piecewise-constant lag
# basis with weekly breaks instead of a smooth spline.
cb_anseriformes <- crossbasis(
  as.matrix(case_crossover_df[, paste0("anseriformes_Lag", 0:MAX_LAG), with = FALSE]),
  lag    = c(0, MAX_LAG),
  argvar = ARGVAR_LIN,
  arglag = list(fun = "strata", breaks = c(7, 14, 21))
)

```

### Flexible crossbasis (higher df / polynomial argvar — sensitivity only)

```{r, more flexible crossbasis terms}

cb_flex_leaf_area_index      <- make_cb("leaf_area_index",      df_lag = 5)
cb_flex_longwave_radiation   <- make_cb("longwave_radiation",   df_lag = 5)
cb_flex_precipitation        <- make_cb("precipitation",        df_lag = 5)
cb_flex_relative_humidity    <- make_cb("relative_humidity",    df_lag = 5)
cb_flex_runoff               <- make_cb("runoff",               df_lag = 5)
cb_flex_shortwave_radiation  <- make_cb("shortwave_radiation",  df_lag = 5)
cb_flex_snow_cover           <- make_cb("snow_cover",           df_lag = 5)
cb_flex_soil_moisture        <- make_cb("soil_moisture",        df_lag = 5)
cb_flex_temperature          <- make_cb("temperature",          df_lag = 5)
cb_flex_wind_speed           <- make_cb("wind_speed",           df_lag = 5)

cb_flex_anseriformes <- crossbasis(
  as.matrix(case_crossover_df[, paste0("anseriformes_Lag", 0:MAX_LAG), with = FALSE]),
  lag    = c(0, MAX_LAG),
  argvar = ARGVAR_LIN,
  arglag = list(fun = "strata", breaks = c(7, 14, 21))
)

```


## Case-Crossover Model

```{r, casecrossover model}

# ----------------------------
# Model Fitting
# ----------------------------
final_model <- clogit(
  outbreak_binary ~
    cb_runoff + cb_soil_moisture + cb_temperature +
    cb_precipitation + cb_wind_speed + cb_snow_cover + cb_anseriformes +
    strata(stratum_id),
  method = "efron",
  data = case_crossover_df
)

summary(final_model)
AIC(final_model)
saveRDS(final_model, file.path(models_folder, "casecrossover_model.RDS"))

```

```{r, crosspred objects}
# ----------------------------
# Crosspred Objects
# ----------------------------
cp_runoff       <- crosspred(cb_runoff, final_model, cen = 0, at = 0.5, cumul = TRUE)
cp_soil         <- crosspred(cb_soil_moisture, final_model, cen = 0, at = 0.5, cumul = TRUE)
cp_temp         <- crosspred(cb_temperature, final_model, cen = 0, at = 0.5, cumul = TRUE)
cp_precip       <- crosspred(cb_precipitation, final_model, cen = 0, at = 0.5, cumul = TRUE)
cp_wind         <- crosspred(cb_wind_speed, final_model, cen = 0, at = 0.5, cumul = TRUE)
cp_snow         <- crosspred(cb_snow_cover, final_model, cen = 0, at = 0.5, cumul = TRUE)
cp_bird         <- crosspred(cb_anseriformes, final_model, cen = 0, at = 0.5, cumul = TRUE)

cp_slice_runoff <- crosspred(cb_runoff, final_model, cen = 0, at = 0.5)
cp_slice_soil   <- crosspred(cb_soil_moisture, final_model, cen = 0, at = 0.5)
cp_slice_temp   <- crosspred(cb_temperature, final_model, cen = 0, at = 0.5)
cp_slice_precip <- crosspred(cb_precipitation, final_model, cen = 0, at = 0.5)
cp_slice_wind   <- crosspred(cb_wind_speed, final_model, cen = 0, at = 0.5)
cp_slice_snow   <- crosspred(cb_snow_cover, final_model, cen = 0, at = 0.5)
cp_slice_bird   <- crosspred(cb_anseriformes, final_model, cen = 0, at = 0.5)

```

## Casecrossover Plots
```{r, plots casecrossover}
# ----------------------------
# Cumulative OR Summary Table
# ----------------------------
cumulative_summary <- tibble(
  Predictor = c("Runoff", "Soil Moisture", "Temperature",
                "Precipitation", "Wind Speed", "Anseriformes"),
      OR = c(
        cp_runoff$allRRfit["0.5"],
        cp_soil$allRRfit["0.5"],
        cp_temp$allRRfit["0.5"],
        cp_precip$allRRfit["0.5"],
        cp_wind$allRRfit["0.5"],
        cp_bird$allRRfit["0.5"]
      ),
      CI_low = c(
        cp_runoff$allRRlow["0.5"],
        cp_soil$allRRlow["0.5"],
        cp_temp$allRRlow["0.5"],
        cp_precip$allRRlow["0.5"],
        cp_wind$allRRlow["0.5"],
        cp_bird$allRRlow["0.5"]
      ),
      CI_high = c(
        cp_runoff$allRRhigh["0.5"],
        cp_soil$allRRhigh["0.5"],
        cp_temp$allRRhigh["0.5"],
        cp_precip$allRRhigh["0.5"],
        cp_wind$allRRhigh["0.5"],
        cp_bird$allRRhigh["0.5"]
      )
) %>%
  mutate(`OR (95% CI)` = sprintf("%.2f (%.2f, %.2f)", OR, CI_low, CI_high)) %>%
  select(Predictor, `OR (95% CI)`) %>%
  gt() %>%
  tab_header(
    title = md("**Case-Crossover Model: Cumulative Odds Ratios**"),
    subtitle = "Per 0.5 SD Increase"
  ) %>%
  cols_label(`OR (95% CI)` = "Odds Ratio (95% Confidence Interval)") %>%
  tab_options(
    table.font.size = 14,
    heading.title.font.size = 16,
    heading.subtitle.font.size = 14,
    data_row.padding = px(6)
  )

# ----------------------------
# Plotting Functions
# ----------------------------
plot_cumulative <- function(cp, title) {
  plot(cp, type = "cumulative", var = 0.5, ci = "area",
       col = "black", ci.arg = list(col = rgb(0,0,0,0.1), border = NA),
       lwd = 1.5, main = title, xlab = "", ylab = "", las = 1)
  abline(h = 1, lty = 2, col = "gray40")
}

plot_slices <- function(cp, title) {
  plot(cp, type = "slices", var = 0.5, ci = "area",
       col = "black", ci.arg = list(col = rgb(0,0,0,0.1), border = NA),
       lwd = 1.3, main = title, xlab = "", ylab = "", las = 1)
  abline(h = 1, lty = 2, col = "gray40")
}

# ----------------------------
# Save Cumulative Plot
# ----------------------------
png(file.path(figures_main_folder,"figure_cumulative_ORs_casecrossover.png"), width = 11, height = 7, units = "in", res = 300)

par(
  mfrow = c(2, 3),
  mar = c(4.2, 4.2, 2.5, 1),
  oma = c(3, 4, 0, 1),
  cex.main = 1.2,
  cex.lab = 1.2,
  cex.axis = 1.1,
  tcl = -0.3, mgp = c(2.5, 0.7, 0)
)

plot_cumulative(cp_runoff, "Runoff")
plot_cumulative(cp_soil, "Soil Moisture")
plot_cumulative(cp_temp, "Temperature")
plot_cumulative(cp_precip, "Precipitation")
plot_cumulative(cp_wind, "Wind Speed")
plot_cumulative(cp_bird, "Anseriformes")

mtext("Lag (days)", side = 1, outer = TRUE, line = 0, cex = 1.2)
mtext("Odds Ratio (95% CI)", side = 2, outer = TRUE, line = 0, cex = 1.2)

dev.off()

# ----------------------------
# Save Slices Plot
# ----------------------------
png(file.path(figures_main_folder,"figure_slice_ORs_casecrossover.png"), width = 11, height = 7, units = "in", res = 300)

par(
  mfrow = c(2, 3),
  mar = c(4.2, 4.2, 2.5, 1),
  oma = c(3, 4, 0, 1),
  cex.main = 1.2,
  cex.lab = 1.2,
  cex.axis = 1.1,
  tcl = -0.3, mgp = c(2.5, 0.7, 0)
)

plot_slices(cp_slice_runoff, "Runoff")
plot_slices(cp_slice_soil, "Soil Moisture")
plot_slices(cp_slice_temp, "Temperature")
plot_slices(cp_slice_precip, "Precipitation")
plot_slices(cp_slice_wind, "Wind Speed")
plot_slices(cp_slice_bird, "Anseriformes")

mtext("Lag (days)", side = 1, outer = TRUE, line = 0, cex = 1.2)
mtext("Odds Ratio (95% CI)", side = 2, outer = TRUE, line = 0, cex = 1.2)

dev.off()


```


## Lag Specific Odds Ratio Table (Daily — weekly anchor lags)

```{r, lag specific odds ratio table}
# Display lag-specific ORs at weekly anchor points (0, 7, 14 days).
LAG_ANCHORS <- c(0, 7, 14, 21, 28)

extract_lag_ORs_wide <- function(cp_obj, varname, varval = "0.5", lags = LAG_ANCHORS) {
  lag_labels <- paste0("lag", lags)
  or    <- cp_obj$matRRfit[varval, lag_labels]
  low   <- cp_obj$matRRlow[varval, lag_labels]
  high  <- cp_obj$matRRhigh[varval, lag_labels]
  col_labels <- paste0("Lag ", lags, " d")
  tibble(
    Predictor = varname,
    !!!setNames(as.list(or),   paste0("OR_Lag", lags)),
    !!!setNames(as.list(low),  paste0("Low_Lag", lags)),
    !!!setNames(as.list(high), paste0("High_Lag", lags)),
    !!!setNames(as.list(sprintf("%.2f (%.2f, %.2f)", or, low, high)), col_labels)
  )
}

table_lag_ORs <- bind_rows(
  extract_lag_ORs_wide(cp_slice_runoff, "Runoff"),
  extract_lag_ORs_wide(cp_slice_soil,   "Soil Moisture"),
  extract_lag_ORs_wide(cp_slice_temp,   "Temperature"),
  extract_lag_ORs_wide(cp_slice_precip, "Precipitation"),
  extract_lag_ORs_wide(cp_slice_wind,   "Wind Speed"),
  extract_lag_ORs_wide(cp_slice_bird,   "Anseriformes")
)

# Highlight significant rows
col_labels <- paste0("Lag ", LAG_ANCHORS, " d")
highlight_matrix <- sapply(LAG_ANCHORS, function(i) {
  table_lag_ORs[[paste0("Low_Lag", i)]] > 1 | table_lag_ORs[[paste0("High_Lag", i)]] < 1
})
colnames(highlight_matrix) <- col_labels

ft <- table_lag_ORs %>%
  select(Predictor, all_of(col_labels)) %>%
  flextable() %>%
  set_header_labels(Predictor = "Predictor") %>%
  add_header_row(values = c("Predictor", "Odds Ratio (95% CI)"),
                 colwidths = c(1, length(col_labels))) %>%
  fontsize(size = 11, part = "all") %>%
  autofit() %>%
  align(align = "center", part = "all") %>%
  theme_booktabs()

for (k in seq_along(LAG_ANCHORS)) {
  ft <- bold(ft, i = which(highlight_matrix[, k]), j = col_labels[k])
}

doc <- read_docx() %>%
  body_add_flextable(ft) %>%
  body_add_par(
    "Table: Lag-Specific Odds Ratios at Weekly Anchor Lags (0, 7, 14 days) for the Daily Time-Stratified Case-Crossover Model of HPAI Outbreaks in Minnesota, 2022. Estimates are per 0.5 standard deviation increase in each predictor. Statistically significant odds ratios are shown in bold.",
    style = "Normal"
  )

print(doc, target = file.path(tables_main_folder, "Table_LagSpecific_OR_HPAI_Casecrossover_daily.docx"))

ft
```

### Cumulative odds ratio table (cumulative through anchor lags)

```{r, cumulative odds ratios}
ft_lagspecific <- ft

# Cumulative OR through each weekly anchor lag (0, 7, 14).
extract_cum_ORs_wide <- function(cp_obj, varname, varval = "0.5", lags = LAG_ANCHORS) {
  lag_labels <- paste0("lag", lags)
  or    <- cp_obj$cumRRfit[varval, lag_labels]
  low   <- cp_obj$cumRRlow[varval, lag_labels]
  high  <- cp_obj$cumRRhigh[varval, lag_labels]
  col_labels <- paste0("Cum ≤ ", lags, " d")
  tibble(
    Predictor = varname,
    !!!setNames(as.list(or),   paste0("OR_Lag", lags)),
    !!!setNames(as.list(low),  paste0("Low_Lag", lags)),
    !!!setNames(as.list(high), paste0("High_Lag", lags)),
    !!!setNames(as.list(sprintf("%.2f (%.2f, %.2f)", or, low, high)), col_labels)
  )
}

table_lag_ORs_cum <- bind_rows(
  extract_cum_ORs_wide(cp_runoff, "Runoff"),
  extract_cum_ORs_wide(cp_soil,   "Soil Moisture"),
  extract_cum_ORs_wide(cp_temp,   "Temperature"),
  extract_cum_ORs_wide(cp_precip, "Precipitation"),
  extract_cum_ORs_wide(cp_wind,   "Wind Speed"),
  extract_cum_ORs_wide(cp_bird,   "Anseriformes")
)

col_labels_cum <- paste0("Cum ≤ ", LAG_ANCHORS, " d")
highlight_matrix <- sapply(LAG_ANCHORS, function(i) {
  table_lag_ORs_cum[[paste0("Low_Lag", i)]] > 1 | table_lag_ORs_cum[[paste0("High_Lag", i)]] < 1
})
colnames(highlight_matrix) <- col_labels_cum

ft_cum <- table_lag_ORs_cum %>%
  select(Predictor, all_of(col_labels_cum)) %>%
  flextable() %>%
  set_header_labels(Predictor = "Predictor") %>%
  add_header_row(values = c("Predictor", "Cumulative Odds Ratio (95% CI)"),
                 colwidths = c(1, length(col_labels_cum))) %>%
  fontsize(size = 11, part = "all") %>%
  autofit() %>%
  align(align = "center", part = "all") %>%
  theme_booktabs()

for (k in seq_along(LAG_ANCHORS)) {
  ft_cum <- bold(ft_cum, i = which(highlight_matrix[, k]), j = col_labels_cum[k])
}

doc <- read_docx() %>%
  body_add_flextable(ft_cum) %>%
  body_add_par(
    "Table: Cumulative Odds Ratios through weekly anchor lags (≤0, ≤7, ≤14 days) for the daily time-stratified case-crossover model of HPAI outbreaks in Minnesota, 2022. Estimates are per 0.5 SD increase in each predictor. Statistically significant ORs in bold.",
    style = "Normal"
  )

print(doc, target = file.path(tables_main_folder, "Table_Cumulative_OR_HPAI_Casscrossover_daily.docx"))

ft_cum
```

#### Combine cumulative and lag-specific odds ratio tables

```{r, combine and lagspecific odds ratio tables}

panel_label <- function(txt, size = 11) {
  fp <- fp_text(bold = TRUE, font.size = size)
  fpar(ftext(txt, fp))
}

doc <- read_docx() %>%
  # Panel A
  body_add_fpar(panel_label("A. Lag-specific odds ratios (per +0.5 SD)")) %>%
  body_add_flextable(ft_lagspecific) %>%
  body_add_par("", style = "Normal") %>%   # spacer

  # Panel B
  body_add_fpar(panel_label("B. Cumulative odds ratios (per +0.5 SD)")) %>%
  body_add_flextable(ft_cum)

print(doc, target = file.path(tables_main_folder, "Table_LagSpecific_and_Cumulative_ORs_Casecrossover.docx"))


```



### Cumulative odds ratio by weekly lag window (0–7d, 8–14d, 15–21d, 22–28d)

```{r, cumulative odds ratio lag sections}

# Weekly lag windows expressed in days. Each window is fitted via a fresh
# crosspred over the specified [from, to] daily lag range.
lag_windows <- list(
  "Lag 0–7 d"   = c(0, 7),
  "Lag 8–14 d"  = c(8, 14),
  "Lag 15–21 d" = c(15, 21),
  "Lag 22–28 d" = c(22, 28)
)

# Met variables share the same ns(df=3) lag basis; bird abundance uses a
# separate strata lag basis (weekly breaks), so its contrast vectors are
# different.
lag_seq        <- 0:28
lag_basis_met  <- do.call(dlnm::onebasis, c(list(x = lag_seq), attr(cb_runoff, "arglag")))
lag_basis_bird <- do.call(dlnm::onebasis, c(list(x = lag_seq), attr(cb_anseriformes, "arglag")))

contrast_0_7        <- 0.5 * colSums(lag_basis_met[lag_seq  >=  0 & lag_seq <=  7, , drop = FALSE])
contrast_8_14       <- 0.5 * colSums(lag_basis_met[lag_seq  >=  8 & lag_seq <= 14, , drop = FALSE])
contrast_15_21      <- 0.5 * colSums(lag_basis_met[lag_seq  >= 15 & lag_seq <= 21, , drop = FALSE])
contrast_22_28      <- 0.5 * colSums(lag_basis_met[lag_seq  >= 22 & lag_seq <= 28, , drop = FALSE])

contrast_bird_0_7   <- 0.5 * colSums(lag_basis_bird[lag_seq >=  0 & lag_seq <=  7, , drop = FALSE])
contrast_bird_8_14  <- 0.5 * colSums(lag_basis_bird[lag_seq >=  8 & lag_seq <= 14, , drop = FALSE])
contrast_bird_15_21 <- 0.5 * colSums(lag_basis_bird[lag_seq >= 15 & lag_seq <= 21, , drop = FALSE])
contrast_bird_22_28 <- 0.5 * colSums(lag_basis_bird[lag_seq >= 22 & lag_seq <= 28, , drop = FALSE])

# Coefficients and vcov submatrix per crossbasis
idx_runoff <- grep("^cb_runoffv",               names(coef(final_model))); b_runoff <- coef(final_model)[idx_runoff]; V_runoff <- vcov(final_model)[idx_runoff, idx_runoff]
idx_soil   <- grep("^cb_soil_moisturev",        names(coef(final_model))); b_soil   <- coef(final_model)[idx_soil];   V_soil   <- vcov(final_model)[idx_soil,   idx_soil]
idx_temp   <- grep("^cb_temperaturev",          names(coef(final_model))); b_temp   <- coef(final_model)[idx_temp];   V_temp   <- vcov(final_model)[idx_temp,   idx_temp]
idx_precip <- grep("^cb_precipitationv",        names(coef(final_model))); b_precip <- coef(final_model)[idx_precip]; V_precip <- vcov(final_model)[idx_precip, idx_precip]
idx_wind   <- grep("^cb_wind_speedv",           names(coef(final_model))); b_wind   <- coef(final_model)[idx_wind];   V_wind   <- vcov(final_model)[idx_wind,   idx_wind]
idx_snow   <- grep("^cb_snow_coverv",           names(coef(final_model))); b_snow   <- coef(final_model)[idx_snow];   V_snow   <- vcov(final_model)[idx_snow,   idx_snow]
idx_bird   <- grep("^cb_anseriformesv", names(coef(final_model))); b_bird   <- coef(final_model)[idx_bird];   V_bird   <- vcov(final_model)[idx_bird,   idx_bird]

# Cumulative log-OR and SE for each (predictor, window)
lo_runoff_0_7   <- sum(contrast_0_7   * b_runoff); se_runoff_0_7   <- sqrt(as.numeric(t(contrast_0_7)   %*% V_runoff %*% contrast_0_7))
lo_runoff_8_14  <- sum(contrast_8_14  * b_runoff); se_runoff_8_14  <- sqrt(as.numeric(t(contrast_8_14)  %*% V_runoff %*% contrast_8_14))
lo_runoff_15_21 <- sum(contrast_15_21 * b_runoff); se_runoff_15_21 <- sqrt(as.numeric(t(contrast_15_21) %*% V_runoff %*% contrast_15_21))
lo_runoff_22_28 <- sum(contrast_22_28 * b_runoff); se_runoff_22_28 <- sqrt(as.numeric(t(contrast_22_28) %*% V_runoff %*% contrast_22_28))


lo_soil_0_7   <- sum(contrast_0_7   * b_soil); se_soil_0_7   <- sqrt(as.numeric(t(contrast_0_7)   %*% V_soil %*% contrast_0_7))
lo_soil_8_14  <- sum(contrast_8_14  * b_soil); se_soil_8_14  <- sqrt(as.numeric(t(contrast_8_14)  %*% V_soil %*% contrast_8_14))
lo_soil_15_21 <- sum(contrast_15_21 * b_soil); se_soil_15_21 <- sqrt(as.numeric(t(contrast_15_21) %*% V_soil %*% contrast_15_21))
lo_soil_22_28 <- sum(contrast_22_28 * b_soil); se_soil_22_28 <- sqrt(as.numeric(t(contrast_22_28) %*% V_soil %*% contrast_22_28))

lo_temp_0_7   <- sum(contrast_0_7   * b_temp); se_temp_0_7   <- sqrt(as.numeric(t(contrast_0_7)   %*% V_temp %*% contrast_0_7))
lo_temp_8_14  <- sum(contrast_8_14  * b_temp); se_temp_8_14  <- sqrt(as.numeric(t(contrast_8_14)  %*% V_temp %*% contrast_8_14))
lo_temp_15_21 <- sum(contrast_15_21 * b_temp); se_temp_15_21 <- sqrt(as.numeric(t(contrast_15_21) %*% V_temp %*% contrast_15_21))
lo_temp_22_28 <- sum(contrast_22_28 * b_temp); se_temp_22_28 <- sqrt(as.numeric(t(contrast_22_28) %*% V_temp %*% contrast_22_28))

lo_precip_0_7   <- sum(contrast_0_7   * b_precip); se_precip_0_7   <- sqrt(as.numeric(t(contrast_0_7)   %*% V_precip %*% contrast_0_7))
lo_precip_8_14  <- sum(contrast_8_14  * b_precip); se_precip_8_14  <- sqrt(as.numeric(t(contrast_8_14)  %*% V_precip %*% contrast_8_14))
lo_precip_15_21 <- sum(contrast_15_21 * b_precip); se_precip_15_21 <- sqrt(as.numeric(t(contrast_15_21) %*% V_precip %*% contrast_15_21))
lo_precip_22_28 <- sum(contrast_22_28 * b_precip); se_precip_22_28 <- sqrt(as.numeric(t(contrast_22_28) %*% V_precip %*% contrast_22_28))

lo_wind_0_7   <- sum(contrast_0_7   * b_wind); se_wind_0_7   <- sqrt(as.numeric(t(contrast_0_7)   %*% V_wind %*% contrast_0_7))
lo_wind_8_14  <- sum(contrast_8_14  * b_wind); se_wind_8_14  <- sqrt(as.numeric(t(contrast_8_14)  %*% V_wind %*% contrast_8_14))
lo_wind_15_21 <- sum(contrast_15_21 * b_wind); se_wind_15_21 <- sqrt(as.numeric(t(contrast_15_21) %*% V_wind %*% contrast_15_21))
lo_wind_22_28 <- sum(contrast_22_28 * b_wind); se_wind_22_28 <- sqrt(as.numeric(t(contrast_22_28) %*% V_wind %*% contrast_22_28))

lo_snow_0_7   <- sum(contrast_0_7   * b_snow); se_snow_0_7   <- sqrt(as.numeric(t(contrast_0_7)   %*% V_snow %*% contrast_0_7))
lo_snow_8_14  <- sum(contrast_8_14  * b_snow); se_snow_8_14  <- sqrt(as.numeric(t(contrast_8_14)  %*% V_snow %*% contrast_8_14))
lo_snow_15_21 <- sum(contrast_15_21 * b_snow); se_snow_15_21 <- sqrt(as.numeric(t(contrast_15_21) %*% V_snow %*% contrast_15_21))
lo_snow_22_28 <- sum(contrast_22_28 * b_snow); se_snow_22_28 <- sqrt(as.numeric(t(contrast_22_28) %*% V_snow %*% contrast_22_28))

lo_bird_0_7   <- sum(contrast_bird_0_7   * b_bird); se_bird_0_7   <- sqrt(as.numeric(t(contrast_bird_0_7)   %*% V_bird %*% contrast_bird_0_7))
lo_bird_8_14  <- sum(contrast_bird_8_14  * b_bird); se_bird_8_14  <- sqrt(as.numeric(t(contrast_bird_8_14)  %*% V_bird %*% contrast_bird_8_14))
lo_bird_15_21 <- sum(contrast_bird_15_21 * b_bird); se_bird_15_21 <- sqrt(as.numeric(t(contrast_bird_15_21) %*% V_bird %*% contrast_bird_15_21))
lo_bird_22_28 <- sum(contrast_bird_22_28 * b_bird); se_bird_22_28 <- sqrt(as.numeric(t(contrast_bird_22_28) %*% V_bird %*% contrast_bird_22_28))

table_cum_ORs <- tibble(
  Predictor = c("Surface Runoff", "Soil Moisture",
                "Temperature", "Precipitation", "Wind Speed", "Snow Cover", "Anseriformes"),

  `Lag 0–7 d` = c(
    sprintf("%.2f (%.2f, %.2f)", exp(lo_runoff_0_7), exp(lo_runoff_0_7 - 1.96 * se_runoff_0_7), exp(lo_runoff_0_7 + 1.96 * se_runoff_0_7)),
    sprintf("%.2f (%.2f, %.2f)", exp(lo_soil_0_7),   exp(lo_soil_0_7   - 1.96 * se_soil_0_7),   exp(lo_soil_0_7   + 1.96 * se_soil_0_7)),
    sprintf("%.2f (%.2f, %.2f)", exp(lo_temp_0_7),   exp(lo_temp_0_7   - 1.96 * se_temp_0_7),   exp(lo_temp_0_7   + 1.96 * se_temp_0_7)),
    sprintf("%.2f (%.2f, %.2f)", exp(lo_precip_0_7), exp(lo_precip_0_7 - 1.96 * se_precip_0_7), exp(lo_precip_0_7 + 1.96 * se_precip_0_7)),
    sprintf("%.2f (%.2f, %.2f)", exp(lo_wind_0_7),   exp(lo_wind_0_7   - 1.96 * se_wind_0_7),   exp(lo_wind_0_7   + 1.96 * se_wind_0_7)),
    sprintf("%.2f (%.2f, %.2f)", exp(lo_snow_0_7),   exp(lo_snow_0_7   - 1.96 * se_snow_0_7),   exp(lo_snow_0_7   + 1.96 * se_snow_0_7)),
    sprintf("%.2f (%.2f, %.2f)", exp(lo_bird_0_7),   exp(lo_bird_0_7   - 1.96 * se_bird_0_7),   exp(lo_bird_0_7   + 1.96 * se_bird_0_7))
  ),

  `Lag 8–14 d` = c(
    sprintf("%.2f (%.2f, %.2f)", exp(lo_runoff_8_14), exp(lo_runoff_8_14 - 1.96 * se_runoff_8_14), exp(lo_runoff_8_14 + 1.96 * se_runoff_8_14)),
    sprintf("%.2f (%.2f, %.2f)", exp(lo_soil_8_14),   exp(lo_soil_8_14   - 1.96 * se_soil_8_14),   exp(lo_soil_8_14   + 1.96 * se_soil_8_14)),
    sprintf("%.2f (%.2f, %.2f)", exp(lo_temp_8_14),   exp(lo_temp_8_14   - 1.96 * se_temp_8_14),   exp(lo_temp_8_14   + 1.96 * se_temp_8_14)),
    sprintf("%.2f (%.2f, %.2f)", exp(lo_precip_8_14), exp(lo_precip_8_14 - 1.96 * se_precip_8_14), exp(lo_precip_8_14 + 1.96 * se_precip_8_14)),
    sprintf("%.2f (%.2f, %.2f)", exp(lo_wind_8_14),   exp(lo_wind_8_14   - 1.96 * se_wind_8_14),   exp(lo_wind_8_14   + 1.96 * se_wind_8_14)),
    sprintf("%.2f (%.2f, %.2f)", exp(lo_snow_8_14),   exp(lo_snow_8_14   - 1.96 * se_snow_8_14),   exp(lo_snow_8_14   + 1.96 * se_snow_8_14)),
    sprintf("%.2f (%.2f, %.2f)", exp(lo_bird_8_14),   exp(lo_bird_8_14   - 1.96 * se_bird_8_14),   exp(lo_bird_8_14   + 1.96 * se_bird_8_14))
  ),

  `Lag 15–21 d` = c(
    sprintf("%.2f (%.2f, %.2f)", exp(lo_runoff_15_21), exp(lo_runoff_15_21 - 1.96 * se_runoff_15_21), exp(lo_runoff_15_21 + 1.96 * se_runoff_15_21)),
    sprintf("%.2f (%.2f, %.2f)", exp(lo_soil_15_21),   exp(lo_soil_15_21   - 1.96 * se_soil_15_21),   exp(lo_soil_15_21   + 1.96 * se_soil_15_21)),
    sprintf("%.2f (%.2f, %.2f)", exp(lo_temp_15_21),   exp(lo_temp_15_21   - 1.96 * se_temp_15_21),   exp(lo_temp_15_21   + 1.96 * se_temp_15_21)),
    sprintf("%.2f (%.2f, %.2f)", exp(lo_precip_15_21), exp(lo_precip_15_21 - 1.96 * se_precip_15_21), exp(lo_precip_15_21 + 1.96 * se_precip_15_21)),
    sprintf("%.2f (%.2f, %.2f)", exp(lo_wind_15_21),   exp(lo_wind_15_21   - 1.96 * se_wind_15_21),   exp(lo_wind_15_21   + 1.96 * se_wind_15_21)),
    sprintf("%.2f (%.2f, %.2f)", exp(lo_snow_15_21),   exp(lo_snow_15_21   - 1.96 * se_snow_15_21),   exp(lo_snow_15_21   + 1.96 * se_snow_15_21)),
    sprintf("%.2f (%.2f, %.2f)", exp(lo_bird_15_21),   exp(lo_bird_15_21   - 1.96 * se_bird_15_21),   exp(lo_bird_15_21   + 1.96 * se_bird_15_21))
  ),

  `Lag 22–28 d` = c(
    sprintf("%.2f (%.2f, %.2f)", exp(lo_runoff_22_28), exp(lo_runoff_22_28 - 1.96 * se_runoff_22_28), exp(lo_runoff_22_28 + 1.96 * se_runoff_22_28)),
    sprintf("%.2f (%.2f, %.2f)", exp(lo_soil_22_28),   exp(lo_soil_22_28   - 1.96 * se_soil_22_28),   exp(lo_soil_22_28   + 1.96 * se_soil_22_28)),
    sprintf("%.2f (%.2f, %.2f)", exp(lo_temp_22_28),   exp(lo_temp_22_28   - 1.96 * se_temp_22_28),   exp(lo_temp_22_28   + 1.96 * se_temp_22_28)),
    sprintf("%.2f (%.2f, %.2f)", exp(lo_precip_22_28), exp(lo_precip_22_28 - 1.96 * se_precip_22_28), exp(lo_precip_22_28 + 1.96 * se_precip_22_28)),
    sprintf("%.2f (%.2f, %.2f)", exp(lo_wind_22_28),   exp(lo_wind_22_28   - 1.96 * se_wind_22_28),   exp(lo_wind_22_28   + 1.96 * se_wind_22_28)),
    sprintf("%.2f (%.2f, %.2f)", exp(lo_snow_22_28),   exp(lo_snow_22_28   - 1.96 * se_snow_22_28),   exp(lo_snow_22_28   + 1.96 * se_snow_22_28)),
    sprintf("%.2f (%.2f, %.2f)", exp(lo_bird_22_28),   exp(lo_bird_22_28   - 1.96 * se_bird_22_28),   exp(lo_bird_22_28   + 1.96 * se_bird_22_28))
  )
)

# Highlight significant cells
parse_ci <- function(s) {
  parts <- as.numeric(unlist(strsplit(gsub("[()]", "", s), "[ ,]+")))
  c(low = parts[2], high = parts[3])
}
highlight_matrix <- sapply(names(lag_windows), function(w) {
  vapply(table_cum_ORs[[w]], function(s) {
    ci <- parse_ci(s); ci["low"] > 1 | ci["high"] < 1
  }, logical(1))
})
colnames(highlight_matrix) <- names(lag_windows)

ft_cum <- table_cum_ORs %>%
  flextable() %>%
  set_header_labels(Predictor = "Predictor") %>%
  add_header_row(values = c("Predictor", "Cumulative Odds Ratio (95% CI)"),
                 colwidths = c(1, length(lag_windows))) %>%
  fontsize(size = 11, part = "all") %>%
  autofit() %>%
  align(align = "center", part = "all") %>%
  theme_booktabs()

for (w in names(lag_windows)) {
  ft_cum <- bold(ft_cum, i = which(highlight_matrix[, w]), j = w)
}

doc <- read_docx() %>%
  body_add_flextable(ft_cum) %>%
  body_add_par(
    "Table: Cumulative Odds Ratios by Weekly Lag Window for the Daily Time-Stratified Case-Crossover Model of HPAI Outbreaks in Minnesota, 2022. Each cell reports the cumulative OR over the specified daily lag window per 0.5 SD increase in each predictor. Statistically significant associations are shown in bold.",
    style = "Normal"
  )

print(doc, target = file.path(tables_main_folder, "Table_Cumulative_OR_HPAI_Minnesota_daily.docx"))

ft_cum
```

## Case-Crossover Model with Flexible Crossbasis

```{r, casecrossover flex model}

# ----------------------------
# Model Fitting
# ----------------------------
final_model_flex <- clogit(
  outbreak_binary ~
    cb_flex_runoff + cb_flex_soil_moisture + cb_flex_temperature +
    cb_flex_precipitation + cb_flex_wind_speed + cb_flex_snow_cover +
    cb_flex_anseriformes +
    strata(stratum_id),
  method = "efron",
  data = case_crossover_df
)

summary(final_model_flex)
AIC(final_model_flex)
saveRDS(final_model_flex, file.path(models_folder, "casecrossover_flexible_model.RDS"))

```

```{r, crosspred objects for flex model}
# ----------------------------
# Crosspred Objects
# ----------------------------
cp_flex_runoff       <- crosspred(cb_flex_runoff, final_model_flex, cen = 0, at = 0.5, cumul = TRUE)
cp_flex_soil         <- crosspred(cb_flex_soil_moisture, final_model_flex, cen = 0, at = 0.5, cumul = TRUE)
cp_flex_temp         <- crosspred(cb_flex_temperature, final_model_flex, cen = 0, at = 0.5, cumul = TRUE)
cp_flex_precip       <- crosspred(cb_flex_precipitation, final_model_flex, cen = 0, at = 0.5, cumul = TRUE)
cp_flex_wind         <- crosspred(cb_flex_wind_speed, final_model_flex, cen = 0, at = 0.5, cumul = TRUE)
cp_flex_bird         <- crosspred(cb_flex_anseriformes, final_model_flex, cen = 0, at = 0.5, cumul = TRUE)

cp_flex_slice_runoff <- crosspred(cb_flex_runoff, final_model_flex, cen = 0, at = 0.5)
cp_flex_slice_soil   <- crosspred(cb_flex_soil_moisture, final_model_flex, cen = 0, at = 0.5)
cp_flex_slice_temp   <- crosspred(cb_flex_temperature, final_model_flex, cen = 0, at = 0.5)
cp_flex_slice_precip <- crosspred(cb_flex_precipitation, final_model_flex, cen = 0, at = 0.5)
cp_flex_slice_wind   <- crosspred(cb_flex_wind_speed, final_model_flex, cen = 0, at = 0.5)
cp_flex_slice_bird   <- crosspred(cb_flex_anseriformes, final_model_flex, cen = 0, at = 0.5)

```

## Casecrossover Plots
```{r, plots casecrossover flex}
# ----------------------------
# Cumulative OR Summary Table
# ----------------------------
cumulative_summary <- tibble(
  Predictor = c("Surface Runoff", "Soil Moisture", "Temperature",
                "Precipitation", "Wind Speed", "Anseriformes"),
      OR = c(
        cp_flex_runoff$allRRfit["0.5"],
        cp_flex_soil$allRRfit["0.5"],
        cp_flex_temp$allRRfit["0.5"],
        cp_flex_precip$allRRfit["0.5"],
        cp_flex_wind$allRRfit["0.5"],
        cp_flex_bird$allRRfit["0.5"]
      ),
      CI_low = c(
        cp_flex_runoff$allRRlow["0.5"],
        cp_flex_soil$allRRlow["0.5"],
        cp_flex_temp$allRRlow["0.5"],
        cp_flex_precip$allRRlow["0.5"],
        cp_flex_wind$allRRlow["0.5"],
        cp_flex_bird$allRRlow["0.5"]
      ),
      CI_high = c(
        cp_flex_runoff$allRRhigh["0.5"],
        cp_flex_soil$allRRhigh["0.5"],
        cp_flex_temp$allRRhigh["0.5"],
        cp_flex_precip$allRRhigh["0.5"],
        cp_flex_wind$allRRhigh["0.5"],
        cp_flex_bird$allRRhigh["0.5"]
      )
) %>%
  mutate(`OR (95% CI)` = sprintf("%.2f (%.2f, %.2f)", OR, CI_low, CI_high)) %>%
  select(Predictor, `OR (95% CI)`) %>%
  gt() %>%
  tab_header(
    title = md("**Case-Crossover Model: Cumulative Odds Ratios**"),
    subtitle = "Per 0.5 SD Increase"
  ) %>%
  cols_label(`OR (95% CI)` = "Odds Ratio (95% Confidence Interval)") %>%
  tab_options(
    table.font.size = 14,
    heading.title.font.size = 16,
    heading.subtitle.font.size = 14,
    data_row.padding = px(6)
  )

cumulative_summary

# ----------------------------
# Plotting Functions
# ----------------------------
plot_cumulative <- function(cp_flex_, title) {
  plot(cp_flex_, type = "cumulative", var = 0.5, ci = "area",
       col = "black", ci.arg = list(col = rgb(0,0,0,0.1), border = NA),
       lwd = 1.5, main = title, xlab = "", ylab = "", las = 1)
  abline(h = 1, lty = 2, col = "gray40")
}

plot_slices <- function(cp_flex_, title) {
  plot(cp_flex_, type = "slices", var = 0.5, ci = "area",
       col = "black", ci.arg = list(col = rgb(0,0,0,0.1), border = NA),
       lwd = 1.3, main = title, xlab = "", ylab = "", las = 1)
  abline(h = 1, lty = 2, col = "gray40")
}

# ----------------------------
# Save Cumulative Plot
# ----------------------------
png(file.path(figures_main_folder,"figure_cumulative_ORs_casecrossover_flexmodel.png"), width = 11, height = 7, units = "in", res = 300)

par(
  mfrow = c(2, 3),
  mar = c(4.2, 4.2, 2.5, 1),
  oma = c(3, 4, 0, 1),
  cex.main = 1.2,
  cex.lab = 1.2,
  cex.axis = 1.1,
  tcl = -0.3, mgp = c(2.5, 0.7, 0)
)

plot_cumulative(cp_flex_runoff, "Runoff")
plot_cumulative(cp_flex_soil, "Soil Moisture")
plot_cumulative(cp_flex_temp, "Temperature")
plot_cumulative(cp_flex_precip, "Precipitation")
plot_cumulative(cp_flex_wind, "Wind Speed")
plot_cumulative(cp_flex_bird, "Anseriformes")

mtext("Lag (days)", side = 1, outer = TRUE, line = 0, cex = 1.2)
mtext("Odds Ratio (95% CI)", side = 2, outer = TRUE, line = 0, cex = 1.2)

dev.off()

# ----------------------------
# Save Slices Plot
# ----------------------------
png(file.path(figures_main_folder,"figure_slice_ORs_casecrossover_flexmodel.png"), width = 11, height = 7, units = "in", res = 300)

par(
  mfrow = c(2, 3),
  mar = c(4.2, 4.2, 2.5, 1),
  oma = c(3, 4, 0, 1),
  cex.main = 1.2,
  cex.lab = 1.2,
  cex.axis = 1.1,
  tcl = -0.3, mgp = c(2.5, 0.7, 0)
)

plot_slices(cp_flex_slice_runoff, "Runoff")
plot_slices(cp_flex_slice_soil, "Soil Moisture")
plot_slices(cp_flex_slice_temp, "Temperature")
plot_slices(cp_flex_slice_precip, "Precipitation")
plot_slices(cp_flex_slice_wind, "Wind Speed")
plot_slices(cp_flex_slice_bird, "Anseriformes")

mtext("Lag (days)", side = 1, outer = TRUE, line = 0, cex = 1.2)
mtext("Odds Ratio (95% CI)", side = 2, outer = TRUE, line = 0, cex = 1.2)

dev.off()

```


## Lag-Specific Table (Flex Model — weekly anchor lags)

```{r, lag specific odds ratio table for flex model}
extract_lag_ORs_wide <- function(cp_flex_obj, varname, varval = "0.5", lags = LAG_ANCHORS) {
  lag_labels <- paste0("lag", lags)
  or    <- cp_flex_obj$matRRfit[varval, lag_labels]
  low   <- cp_flex_obj$matRRlow[varval, lag_labels]
  high  <- cp_flex_obj$matRRhigh[varval, lag_labels]
  col_labels <- paste0("Lag ", lags, " d")
  tibble(
    Predictor = varname,
    !!!setNames(as.list(or),   paste0("OR_Lag", lags)),
    !!!setNames(as.list(low),  paste0("Low_Lag", lags)),
    !!!setNames(as.list(high), paste0("High_Lag", lags)),
    !!!setNames(as.list(sprintf("%.2f (%.2f, %.2f)", or, low, high)), col_labels)
  )
}

table_lag_ORs <- bind_rows(
  extract_lag_ORs_wide(cp_flex_slice_runoff, "Runoff"),
  extract_lag_ORs_wide(cp_flex_slice_soil,   "Soil Moisture"),
  extract_lag_ORs_wide(cp_flex_slice_temp,   "Temperature"),
  extract_lag_ORs_wide(cp_flex_slice_precip, "Precipitation"),
  extract_lag_ORs_wide(cp_flex_slice_wind,   "Wind Speed"),
  extract_lag_ORs_wide(cp_flex_slice_bird,   "Anseriformes")
)

col_labels <- paste0("Lag ", LAG_ANCHORS, " d")
highlight_matrix <- sapply(LAG_ANCHORS, function(i) {
  table_lag_ORs[[paste0("Low_Lag", i)]] > 1 | table_lag_ORs[[paste0("High_Lag", i)]] < 1
})
colnames(highlight_matrix) <- col_labels

ft <- table_lag_ORs %>%
  select(Predictor, all_of(col_labels)) %>%
  flextable() %>%
  set_header_labels(Predictor = "Predictor") %>%
  add_header_row(values = c("Predictor", "Odds Ratio (95% CI)"),
                 colwidths = c(1, length(col_labels))) %>%
  fontsize(size = 11, part = "all") %>%
  autofit() %>%
  align(align = "center", part = "all") %>%
  theme_booktabs()

for (k in seq_along(LAG_ANCHORS)) {
  ft <- bold(ft, i = which(highlight_matrix[, k]), j = col_labels[k])
}

doc <- read_docx() %>%
  body_add_flextable(ft) %>%
  body_add_par(
    "Table: Lag-Specific Odds Ratios at Weekly Anchor Lags (Flexible DLNM) for HPAI Outbreaks in Minnesota, 2022. Estimates are per 0.5 SD increase in each predictor. Statistically significant odds ratios are shown in bold.",
    style = "Normal"
  )

print(doc, target = file.path(tables_main_folder, "Table_LagSpecific_OR_HPAI_Minnesota_flexmodel_daily.docx"))

ft
```



## Cumulative odds ratio by weekly lag window (Flex Model)

```{r, cumulative odds ratio lag sections flex}

# Flex model uses arglag ns df=5 so the lag basis is rebuilt from cb_flex_runoff
lag_seq        <- 0:28
lag_basis_flex      <- do.call(dlnm::onebasis, c(list(x = lag_seq), attr(cb_flex_runoff, "arglag")))
lag_basis_flex_bird <- do.call(dlnm::onebasis, c(list(x = lag_seq), attr(cb_flex_anseriformes, "arglag")))

contrast_0_7        <- 0.5 * colSums(lag_basis_flex[lag_seq      >=  0 & lag_seq <=  7, , drop = FALSE])
contrast_8_14       <- 0.5 * colSums(lag_basis_flex[lag_seq      >=  8 & lag_seq <= 14, , drop = FALSE])
contrast_15_21      <- 0.5 * colSums(lag_basis_flex[lag_seq      >= 15 & lag_seq <= 21, , drop = FALSE])
contrast_22_28      <- 0.5 * colSums(lag_basis_flex[lag_seq      >= 22 & lag_seq <= 28, , drop = FALSE])

contrast_bird_0_7   <- 0.5 * colSums(lag_basis_flex_bird[lag_seq >=  0 & lag_seq <=  7, , drop = FALSE])
contrast_bird_8_14  <- 0.5 * colSums(lag_basis_flex_bird[lag_seq >=  8 & lag_seq <= 14, , drop = FALSE])
contrast_bird_15_21 <- 0.5 * colSums(lag_basis_flex_bird[lag_seq >= 15 & lag_seq <= 21, , drop = FALSE])
contrast_bird_22_28 <- 0.5 * colSums(lag_basis_flex_bird[lag_seq >= 22 & lag_seq <= 28, , drop = FALSE])

idx_runoff <- grep("^cb_flex_runoffv",               names(coef(final_model_flex))); b_runoff <- coef(final_model_flex)[idx_runoff]; V_runoff <- vcov(final_model_flex)[idx_runoff, idx_runoff]
idx_soil   <- grep("^cb_flex_soil_moisturev",        names(coef(final_model_flex))); b_soil   <- coef(final_model_flex)[idx_soil];   V_soil   <- vcov(final_model_flex)[idx_soil,   idx_soil]
idx_temp   <- grep("^cb_flex_temperaturev",          names(coef(final_model_flex))); b_temp   <- coef(final_model_flex)[idx_temp];   V_temp   <- vcov(final_model_flex)[idx_temp,   idx_temp]
idx_precip <- grep("^cb_flex_precipitationv",        names(coef(final_model_flex))); b_precip <- coef(final_model_flex)[idx_precip]; V_precip <- vcov(final_model_flex)[idx_precip, idx_precip]
idx_wind   <- grep("^cb_flex_wind_speedv",           names(coef(final_model_flex))); b_wind   <- coef(final_model_flex)[idx_wind];   V_wind   <- vcov(final_model_flex)[idx_wind,   idx_wind]
idx_snow   <- grep("^cb_flex_snow_coverv",           names(coef(final_model_flex))); b_snow   <- coef(final_model_flex)[idx_snow];   V_snow   <- vcov(final_model_flex)[idx_snow,   idx_snow]
idx_bird   <- grep("^cb_flex_anseriformesv", names(coef(final_model_flex))); b_bird   <- coef(final_model_flex)[idx_bird];   V_bird   <- vcov(final_model_flex)[idx_bird,   idx_bird]

lo_runoff_0_7   <- sum(contrast_0_7   * b_runoff); se_runoff_0_7   <- sqrt(as.numeric(t(contrast_0_7)   %*% V_runoff %*% contrast_0_7))
lo_runoff_8_14  <- sum(contrast_8_14  * b_runoff); se_runoff_8_14  <- sqrt(as.numeric(t(contrast_8_14)  %*% V_runoff %*% contrast_8_14))
lo_runoff_15_21 <- sum(contrast_15_21 * b_runoff); se_runoff_15_21 <- sqrt(as.numeric(t(contrast_15_21) %*% V_runoff %*% contrast_15_21))
lo_runoff_22_28 <- sum(contrast_22_28 * b_runoff); se_runoff_22_28 <- sqrt(as.numeric(t(contrast_22_28) %*% V_runoff %*% contrast_22_28))


lo_soil_0_7   <- sum(contrast_0_7   * b_soil); se_soil_0_7   <- sqrt(as.numeric(t(contrast_0_7)   %*% V_soil %*% contrast_0_7))
lo_soil_8_14  <- sum(contrast_8_14  * b_soil); se_soil_8_14  <- sqrt(as.numeric(t(contrast_8_14)  %*% V_soil %*% contrast_8_14))
lo_soil_15_21 <- sum(contrast_15_21 * b_soil); se_soil_15_21 <- sqrt(as.numeric(t(contrast_15_21) %*% V_soil %*% contrast_15_21))
lo_soil_22_28 <- sum(contrast_22_28 * b_soil); se_soil_22_28 <- sqrt(as.numeric(t(contrast_22_28) %*% V_soil %*% contrast_22_28))

lo_temp_0_7   <- sum(contrast_0_7   * b_temp); se_temp_0_7   <- sqrt(as.numeric(t(contrast_0_7)   %*% V_temp %*% contrast_0_7))
lo_temp_8_14  <- sum(contrast_8_14  * b_temp); se_temp_8_14  <- sqrt(as.numeric(t(contrast_8_14)  %*% V_temp %*% contrast_8_14))
lo_temp_15_21 <- sum(contrast_15_21 * b_temp); se_temp_15_21 <- sqrt(as.numeric(t(contrast_15_21) %*% V_temp %*% contrast_15_21))
lo_temp_22_28 <- sum(contrast_22_28 * b_temp); se_temp_22_28 <- sqrt(as.numeric(t(contrast_22_28) %*% V_temp %*% contrast_22_28))

lo_precip_0_7   <- sum(contrast_0_7   * b_precip); se_precip_0_7   <- sqrt(as.numeric(t(contrast_0_7)   %*% V_precip %*% contrast_0_7))
lo_precip_8_14  <- sum(contrast_8_14  * b_precip); se_precip_8_14  <- sqrt(as.numeric(t(contrast_8_14)  %*% V_precip %*% contrast_8_14))
lo_precip_15_21 <- sum(contrast_15_21 * b_precip); se_precip_15_21 <- sqrt(as.numeric(t(contrast_15_21) %*% V_precip %*% contrast_15_21))
lo_precip_22_28 <- sum(contrast_22_28 * b_precip); se_precip_22_28 <- sqrt(as.numeric(t(contrast_22_28) %*% V_precip %*% contrast_22_28))

lo_wind_0_7   <- sum(contrast_0_7   * b_wind); se_wind_0_7   <- sqrt(as.numeric(t(contrast_0_7)   %*% V_wind %*% contrast_0_7))
lo_wind_8_14  <- sum(contrast_8_14  * b_wind); se_wind_8_14  <- sqrt(as.numeric(t(contrast_8_14)  %*% V_wind %*% contrast_8_14))
lo_wind_15_21 <- sum(contrast_15_21 * b_wind); se_wind_15_21 <- sqrt(as.numeric(t(contrast_15_21) %*% V_wind %*% contrast_15_21))
lo_wind_22_28 <- sum(contrast_22_28 * b_wind); se_wind_22_28 <- sqrt(as.numeric(t(contrast_22_28) %*% V_wind %*% contrast_22_28))

lo_snow_0_7   <- sum(contrast_0_7   * b_snow); se_snow_0_7   <- sqrt(as.numeric(t(contrast_0_7)   %*% V_snow %*% contrast_0_7))
lo_snow_8_14  <- sum(contrast_8_14  * b_snow); se_snow_8_14  <- sqrt(as.numeric(t(contrast_8_14)  %*% V_snow %*% contrast_8_14))
lo_snow_15_21 <- sum(contrast_15_21 * b_snow); se_snow_15_21 <- sqrt(as.numeric(t(contrast_15_21) %*% V_snow %*% contrast_15_21))
lo_snow_22_28 <- sum(contrast_22_28 * b_snow); se_snow_22_28 <- sqrt(as.numeric(t(contrast_22_28) %*% V_snow %*% contrast_22_28))

lo_bird_0_7   <- sum(contrast_bird_0_7   * b_bird); se_bird_0_7   <- sqrt(as.numeric(t(contrast_bird_0_7)   %*% V_bird %*% contrast_bird_0_7))
lo_bird_8_14  <- sum(contrast_bird_8_14  * b_bird); se_bird_8_14  <- sqrt(as.numeric(t(contrast_bird_8_14)  %*% V_bird %*% contrast_bird_8_14))
lo_bird_15_21 <- sum(contrast_bird_15_21 * b_bird); se_bird_15_21 <- sqrt(as.numeric(t(contrast_bird_15_21) %*% V_bird %*% contrast_bird_15_21))
lo_bird_22_28 <- sum(contrast_bird_22_28 * b_bird); se_bird_22_28 <- sqrt(as.numeric(t(contrast_bird_22_28) %*% V_bird %*% contrast_bird_22_28))

table_cum_ORs <- tibble(
  Predictor = c("Surface Runoff", "Soil Moisture",
                "Temperature", "Precipitation", "Wind Speed", "Snow Cover", "Anseriformes"),

  `Lag 0–7 d` = c(
    sprintf("%.2f (%.2f, %.2f)", exp(lo_runoff_0_7), exp(lo_runoff_0_7 - 1.96 * se_runoff_0_7), exp(lo_runoff_0_7 + 1.96 * se_runoff_0_7)),
    sprintf("%.2f (%.2f, %.2f)", exp(lo_soil_0_7),   exp(lo_soil_0_7   - 1.96 * se_soil_0_7),   exp(lo_soil_0_7   + 1.96 * se_soil_0_7)),
    sprintf("%.2f (%.2f, %.2f)", exp(lo_temp_0_7),   exp(lo_temp_0_7   - 1.96 * se_temp_0_7),   exp(lo_temp_0_7   + 1.96 * se_temp_0_7)),
    sprintf("%.2f (%.2f, %.2f)", exp(lo_precip_0_7), exp(lo_precip_0_7 - 1.96 * se_precip_0_7), exp(lo_precip_0_7 + 1.96 * se_precip_0_7)),
    sprintf("%.2f (%.2f, %.2f)", exp(lo_wind_0_7),   exp(lo_wind_0_7   - 1.96 * se_wind_0_7),   exp(lo_wind_0_7   + 1.96 * se_wind_0_7)),
    sprintf("%.2f (%.2f, %.2f)", exp(lo_snow_0_7),   exp(lo_snow_0_7   - 1.96 * se_snow_0_7),   exp(lo_snow_0_7   + 1.96 * se_snow_0_7)),
    sprintf("%.2f (%.2f, %.2f)", exp(lo_bird_0_7),   exp(lo_bird_0_7   - 1.96 * se_bird_0_7),   exp(lo_bird_0_7   + 1.96 * se_bird_0_7))
  ),

  `Lag 8–14 d` = c(
    sprintf("%.2f (%.2f, %.2f)", exp(lo_runoff_8_14), exp(lo_runoff_8_14 - 1.96 * se_runoff_8_14), exp(lo_runoff_8_14 + 1.96 * se_runoff_8_14)),
    sprintf("%.2f (%.2f, %.2f)", exp(lo_soil_8_14),   exp(lo_soil_8_14   - 1.96 * se_soil_8_14),   exp(lo_soil_8_14   + 1.96 * se_soil_8_14)),
    sprintf("%.2f (%.2f, %.2f)", exp(lo_temp_8_14),   exp(lo_temp_8_14   - 1.96 * se_temp_8_14),   exp(lo_temp_8_14   + 1.96 * se_temp_8_14)),
    sprintf("%.2f (%.2f, %.2f)", exp(lo_precip_8_14), exp(lo_precip_8_14 - 1.96 * se_precip_8_14), exp(lo_precip_8_14 + 1.96 * se_precip_8_14)),
    sprintf("%.2f (%.2f, %.2f)", exp(lo_wind_8_14),   exp(lo_wind_8_14   - 1.96 * se_wind_8_14),   exp(lo_wind_8_14   + 1.96 * se_wind_8_14)),
    sprintf("%.2f (%.2f, %.2f)", exp(lo_snow_8_14),   exp(lo_snow_8_14   - 1.96 * se_snow_8_14),   exp(lo_snow_8_14   + 1.96 * se_snow_8_14)),
    sprintf("%.2f (%.2f, %.2f)", exp(lo_bird_8_14),   exp(lo_bird_8_14   - 1.96 * se_bird_8_14),   exp(lo_bird_8_14   + 1.96 * se_bird_8_14))
  ),

  `Lag 15–21 d` = c(
    sprintf("%.2f (%.2f, %.2f)", exp(lo_runoff_15_21), exp(lo_runoff_15_21 - 1.96 * se_runoff_15_21), exp(lo_runoff_15_21 + 1.96 * se_runoff_15_21)),
    sprintf("%.2f (%.2f, %.2f)", exp(lo_soil_15_21),   exp(lo_soil_15_21   - 1.96 * se_soil_15_21),   exp(lo_soil_15_21   + 1.96 * se_soil_15_21)),
    sprintf("%.2f (%.2f, %.2f)", exp(lo_temp_15_21),   exp(lo_temp_15_21   - 1.96 * se_temp_15_21),   exp(lo_temp_15_21   + 1.96 * se_temp_15_21)),
    sprintf("%.2f (%.2f, %.2f)", exp(lo_precip_15_21), exp(lo_precip_15_21 - 1.96 * se_precip_15_21), exp(lo_precip_15_21 + 1.96 * se_precip_15_21)),
    sprintf("%.2f (%.2f, %.2f)", exp(lo_wind_15_21),   exp(lo_wind_15_21   - 1.96 * se_wind_15_21),   exp(lo_wind_15_21   + 1.96 * se_wind_15_21)),
    sprintf("%.2f (%.2f, %.2f)", exp(lo_snow_15_21),   exp(lo_snow_15_21   - 1.96 * se_snow_15_21),   exp(lo_snow_15_21   + 1.96 * se_snow_15_21)),
    sprintf("%.2f (%.2f, %.2f)", exp(lo_bird_15_21),   exp(lo_bird_15_21   - 1.96 * se_bird_15_21),   exp(lo_bird_15_21   + 1.96 * se_bird_15_21))
  ),

  `Lag 22–28 d` = c(
    sprintf("%.2f (%.2f, %.2f)", exp(lo_runoff_22_28), exp(lo_runoff_22_28 - 1.96 * se_runoff_22_28), exp(lo_runoff_22_28 + 1.96 * se_runoff_22_28)),
    sprintf("%.2f (%.2f, %.2f)", exp(lo_soil_22_28),   exp(lo_soil_22_28   - 1.96 * se_soil_22_28),   exp(lo_soil_22_28   + 1.96 * se_soil_22_28)),
    sprintf("%.2f (%.2f, %.2f)", exp(lo_temp_22_28),   exp(lo_temp_22_28   - 1.96 * se_temp_22_28),   exp(lo_temp_22_28   + 1.96 * se_temp_22_28)),
    sprintf("%.2f (%.2f, %.2f)", exp(lo_precip_22_28), exp(lo_precip_22_28 - 1.96 * se_precip_22_28), exp(lo_precip_22_28 + 1.96 * se_precip_22_28)),
    sprintf("%.2f (%.2f, %.2f)", exp(lo_wind_22_28),   exp(lo_wind_22_28   - 1.96 * se_wind_22_28),   exp(lo_wind_22_28   + 1.96 * se_wind_22_28)),
    sprintf("%.2f (%.2f, %.2f)", exp(lo_snow_22_28),   exp(lo_snow_22_28   - 1.96 * se_snow_22_28),   exp(lo_snow_22_28   + 1.96 * se_snow_22_28)),
    sprintf("%.2f (%.2f, %.2f)", exp(lo_bird_22_28),   exp(lo_bird_22_28   - 1.96 * se_bird_22_28),   exp(lo_bird_22_28   + 1.96 * se_bird_22_28))
  )
)

highlight_matrix <- sapply(names(lag_windows), function(w) {
  vapply(table_cum_ORs[[w]], function(s) {
    ci <- parse_ci(s); ci["low"] > 1 | ci["high"] < 1
  }, logical(1))
})
colnames(highlight_matrix) <- names(lag_windows)

ft_cum <- table_cum_ORs %>%
  flextable() %>%
  set_header_labels(Predictor = "Predictor") %>%
  add_header_row(values = c("Predictor", "Cumulative Odds Ratio (95% CI)"),
                 colwidths = c(1, length(lag_windows))) %>%
  fontsize(size = 11, part = "all") %>%
  autofit() %>%
  align(align = "center", part = "all") %>%
  theme_booktabs()

for (w in names(lag_windows)) {
  ft_cum <- bold(ft_cum, i = which(highlight_matrix[, w]), j = w)
}

doc <- read_docx() %>%
  body_add_flextable(ft_cum) %>%
  body_add_par(
    "Table: Cumulative Odds Ratios by Weekly Lag Window (Flexible DLNM) for HPAI Outbreaks in Minnesota, 2022. Each cell reports the cumulative OR over the specified daily lag window per 0.5 SD increase. Significant ORs in bold.",
    style = "Normal"
  )

print(doc, target = file.path(tables_main_folder, "Table_Cumulative_OR_HPAI_Minnesota_flexmodel_daily.docx"))

ft_cum
```

