---
title: "Lag-Window Sensitivity Analysis (Daily)"
author: "Fintan Mooney"
date: "2026-05-12"
output: html_document
---

# Sensitivity testing alternative DLNM lag windows for the daily time-stratified
# case-crossover model. Primary analysis uses lag 0–28 days; here we re-fit at
# 0–14, 0–42, and 0–56 days to check robustness.
# Replaces the prior 2/6/8-week lag sensitivity from the original submission.

```{r include=FALSE}
rm(list=ls())
project.folder = paste0(print(here::here()),'/')
source(paste0(project.folder,'create_folder_structure.R'))
source(paste0(functions.folder,'script_initiate.R'))
```

## Load pre-built lag-window variants (built by c_create_casscrossover_df.Rmd)

```{r load}

cc_14 <- readRDS(file.path(objects_folder, "case_crossover_df_14lag.RDS"))
cc_28 <- readRDS(file.path(objects_folder, "case_crossover_df.RDS"))
cc_42 <- readRDS(file.path(objects_folder, "case_crossover_df_42lag.RDS"))
cc_56 <- readRDS(file.path(objects_folder, "case_crossover_df_56lag.RDS"))

```

## Fit DLNM at each lag window

```{r fit each}

predictors_lag <- c("runoff", "snow_density", "soil_moisture", "temperature",
                    "precipitation", "wind_speed", "total_bird_abundance")

# Helper: build a crossbasis matrix and a clogit model for one (df, lag) pair.
fit_dlnm <- function(df, max_lag) {
  cb_list <- lapply(predictors_lag, function(v) {
    X <- as.matrix(df[, paste0(v, "_Lag", 0:max_lag), with = FALSE])
    crossbasis(X, lag = c(0, max_lag),
               argvar = list(fun = "lin"),
               arglag = list(fun = "ns", df = max(2, round(max_lag / 7))))
  })
  names(cb_list) <- predictors_lag
  # Build the model formula dynamically
  for (v in predictors_lag) assign(paste0("cb_", v), cb_list[[v]], envir = environment())
  rhs <- paste(paste0("cb_", predictors_lag), collapse = " + ")
  fml <- as.formula(paste("outbreak_binary ~", rhs, "+ strata(stratum_id)"))
  m <- clogit(fml, method = "efron", data = df)
  list(model = m, cb = cb_list)
}

res_14 <- fit_dlnm(cc_14, 14)
res_28 <- fit_dlnm(cc_28, 28)
res_42 <- fit_dlnm(cc_42, 42)
res_56 <- fit_dlnm(cc_56, 56)

saveRDS(res_14, file.path(models_folder, "casecrossover_model_14lag.RDS"))
saveRDS(res_28, file.path(models_folder, "casecrossover_model_28lag.RDS"))
saveRDS(res_42, file.path(models_folder, "casecrossover_model_42lag.RDS"))
saveRDS(res_56, file.path(models_folder, "casecrossover_model_56lag.RDS"))

```

## Cumulative OR comparison across lag windows

```{r cumulative comparison}

# Cumulative OR over the full lag window. Use by-position matrix access on
# cumRRfit/Low/High (the last column = cumulative through the deepest lag)
# rather than dimnames, which differ across DLNM versions.
extract_cum <- function(res, label) {
  out <- lapply(names(res$cb), function(v) {
    cp <- crosspred(res$cb[[v]], res$model, cen = 0, at = 0.5, cumul = TRUE)
    n_cols <- ncol(cp$cumRRfit)
    tibble(
      window    = label,
      predictor = v,
      OR        = cp$cumRRfit["0.5",  n_cols],
      lo        = cp$cumRRlow["0.5",  n_cols],
      hi        = cp$cumRRhigh["0.5", n_cols]
    )
  })
  bind_rows(out)
}

comparison <- bind_rows(
  extract_cum(res_14, "0–14 d"),
  extract_cum(res_28, "0–28 d (primary)"),
  extract_cum(res_42, "0–42 d"),
  extract_cum(res_56, "0–56 d")
) %>%
  mutate(OR_CI = sprintf("%.2f (%.2f, %.2f)", OR, lo, hi))

print(comparison %>% select(window, predictor, OR_CI))

# Wide table for Word export
tab_wide <- comparison %>%
  select(predictor, window, OR_CI) %>%
  pivot_wider(names_from = window, values_from = OR_CI)

ft <- tab_wide %>%
  flextable() %>%
  set_header_labels(predictor = "Predictor") %>%
  fontsize(size = 11, part = "all") %>%
  autofit() %>%
  align(align = "center", part = "all") %>%
  theme_booktabs()

doc <- read_docx() %>%
  body_add_flextable(ft) %>%
  body_add_par(
    "Table: Cumulative odds ratios across DLNM lag windows (0–14, 0–28 [primary], 0–42, 0–56 days) for the daily time-stratified case-crossover model. Estimates are per 0.5 SD increase in each predictor.",
    style = "Normal"
  )
print(doc, target = file.path(tables_folder, "Table_LagWindow_Sensitivity_daily.docx"))

```

## Plot lag-response slices across lag windows (primary vs sensitivity)

```{r plot lag response, fig.width = 12, fig.height = 8}

# Reproduce the lag-response slice for each predictor at each lag window.
png(file.path(figures_folder, "figure_lag_sensitivity_daily.png"),
    width = 13, height = 9, units = "in", res = 300)

par(mfrow = c(length(predictors_lag), 4),
    mar = c(3.5, 4, 2.5, 1), oma = c(3, 3, 0, 0),
    cex.main = 0.95, cex.lab = 1.0, cex.axis = 0.9)

for (v in predictors_lag) {
  for (res_lab in list(c("res_14","0–14 d"), c("res_28","0–28 d (primary)"),
                       c("res_42","0–42 d"), c("res_56","0–56 d"))) {
    res <- get(res_lab[[1]])
    cp <- crosspred(res$cb[[v]], res$model, cen = 0, at = 0.5)
    plot(cp, type = "slices", var = 0.5, ci = "area",
         col = "black", ci.arg = list(col = rgb(0,0,0,0.15), border = NA),
         lwd = 1.3, main = paste(v, "—", res_lab[[2]]), xlab = "", ylab = "", las = 1)
    abline(h = 1, lty = 2, col = "gray40")
  }
}
mtext("Lag (days)", side = 1, outer = TRUE, line = 1, cex = 1.0)
mtext("Odds Ratio (95% CI)", side = 2, outer = TRUE, line = 1, cex = 1.0)

dev.off()

```
