---
title: "Multi-state weekly case-crossover DLNM — flyway climate + bird abundance"
author: "Fintan Mooney"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

```{r initiate}
rm(list = ls())
# 0a. Declare root directory, folder location and load essential stuff
project.folder = paste0(print(here::here()), '/')
source(paste0(project.folder, 'create_folder_structure.R'))
source(paste0(functions.folder, 'script_initiate.R'))

# 0b. lubridate owns the date accessors under the project's conflicted setup.
conflict_prefer("year",  "lubridate", quiet = TRUE)
conflict_prefer("month", "lubridate", quiet = TRUE)
```

The weekly bird-only model ([c_03](c_03_casecrossover_flyway_birds.Rmd)) established the
premises-level case-crossover for the flyway states. This script adds the **meteorological**
exposures from the multi-state ERA5 pull, giving the full climate + bird weekly model that
mirrors the MN weekly design ([c_01](c_01_casecrossover_weekly.Rmd)) but pooled across states.

Climate is attached by mapping each outbreak premises to its 10 km ERA5 fishnet zone and
aggregating that zone's daily series to ISO weeks. Premises in states without an ERA5 pull
yet (OH/IL/IN until their exports land) get NA climate and drop from the climate+bird model
automatically — they remain covered by the c_03 bird-only arm. Both referent designs:

  * **unidirectional**  — 1:4 matching, controls at the four weeks before each case week.
  * **time-stratified** — controls are the other weeks in the case's calendar month.

# 1. Load bird panel and apply the independence filter (mirrors c_03)

```{r load-bird}
era5_dir <- paste0(env_data, "hpai_multistate_era5/")

cc <- fread(paste0(bird_abundance_data,
                   "study_spillover_events_casecrossover_2022_abundance_060326.csv"))

# IND + INR only (drop CSLT / NOSEQ); link is constant within an outbreak.
keep_ids <- cc[is_case_week == 1 & link %in% c("IND", "INR"), unique(outbreak_id)]
cc <- cc[outbreak_id %in% keep_ids]
cc[, week_start := as.Date(week_start)]
setorder(cc, outbreak_id, week_start)

cat("Outbreaks (IND+INR): ", length(keep_ids),
    " across ", uniqueN(cc$state), " states\n", sep = "")
```

# 2. Map each premises to its ERA5 fishnet zone

```{r premises-zone}
# 2a. Multi-state fishnet. GEE writes EPSG:5070 coordinates but mislabels the CRS
#     as WGS84, so we *set* (not transform) the CRS, matching a_01_join_data.
fishnet <- st_read(paste0(era5_dir, "fishnet_8state.geojson"), quiet = TRUE)
fishnet <- fishnet[!st_is_empty(fishnet), ]
fishnet <- st_set_crs(fishnet, 5070)

# 2b. One point per premises → containing zone. Premises in states absent from the
#     fishnet (OH/IL/IN until re-pull) return NA and carry no climate.
prem <- unique(cc[is_case_week == 1, .(outbreak_id, state, lon, lat)])
prem_sf <- st_transform(st_as_sf(prem, coords = c("lon", "lat"), crs = 4326), 5070)
prem[, zone_id := fishnet$zone_id[as.integer(st_within(prem_sf, fishnet))]]

cat("Premises mapped to an ERA5 zone, by state:\n")
print(prem[, .(n = .N, with_climate = sum(!is.na(zone_id))), by = state][order(-n)])

cc <- merge(cc, prem[, .(outbreak_id, zone_id)], by = "outbreak_id", all.x = TRUE)
prem_zones <- prem[!is.na(zone_id), unique(zone_id)]
```

# 3. Load multi-state ERA5, derive vars, aggregate to ISO-week per zone

```{r era5-weekly}
# 3a. Read only the columns needed for the 6 primary climate exposures (matching
#     c_01). Restrict to premises zones to keep memory modest.
era5_cols <- c("zone_id", "state", "date",
               "temperature_2m", "volumetric_soil_water_layer_1", "snow_cover",
               "u_component_of_wind_10m", "v_component_of_wind_10m",
               "total_precipitation_hourly", "surface_runoff_hourly")
era5_files <- list.files(era5_dir, pattern = "^era5_[A-Z]{2}_.*\\.csv$",
                         full.names = TRUE)
era5 <- rbindlist(lapply(era5_files, fread, select = era5_cols),
                  use.names = TRUE, fill = TRUE)
era5[, date := as.Date(date)]
era5 <- era5[zone_id %in% prem_zones]            # join on zone_id across all states

# 3b. Derived fields + friendly names (a_01 logic). wind_speed from u/v; the rest
#     are direct renames. Temperature kept in K — the model z-scales, so the unit
#     is immaterial to the per-0.5-SD effect.
era5[, wind_speed := sqrt(u_component_of_wind_10m^2 + v_component_of_wind_10m^2)]
setnames(era5,
         c("temperature_2m", "volumetric_soil_water_layer_1",
           "total_precipitation_hourly", "surface_runoff_hourly"),
         c("temperature", "soil_moisture", "precipitation", "runoff"))

# 3c. Aggregate to ISO-week per zone: means for state variables, sums for fluxes
#     (matching the c_01 weekly aggregation).
era5[, `:=`(iso_year = lubridate::isoyear(date),
            iso_week = lubridate::isoweek(date))]
clim_weekly <- era5[, .(
  temperature   = mean(temperature,   na.rm = TRUE),
  soil_moisture = mean(soil_moisture, na.rm = TRUE),
  snow_cover    = mean(snow_cover,    na.rm = TRUE),
  wind_speed    = mean(wind_speed,    na.rm = TRUE),
  precipitation = sum(precipitation,  na.rm = TRUE),
  runoff        = sum(runoff,         na.rm = TRUE)
), by = .(zone_id, iso_year, iso_week)]

# 3d. Same skew handling as c_01: m→mm then log1p for fluxes; snow % → fraction.
clim_weekly[, runoff        := log1p(runoff        * 1000)]
clim_weekly[, precipitation := log1p(precipitation * 1000)]
clim_weekly[, snow_cover    := snow_cover / 100]
```

# 4. Join climate to premises-weeks; transform and z-scale all exposures

```{r join-scale}
# 4a. Attach weekly climate by zone × ISO-week. The bird study CSV already carries
#     iso_year / iso_week, so the join is exact.
cc <- merge(cc, clim_weekly, by = c("zone_id", "iso_year", "iso_week"), all.x = TRUE)

# 4b. Exposure set: 6 climate (from ERA5) + anseriformes (from the bird CSV).
climate_vars  <- c("temperature", "soil_moisture", "snow_cover",
                   "wind_speed", "precipitation", "runoff")
exposure_vars <- c(climate_vars, "anseri_abd")
name_map <- c(temperature = "Temperature", soil_moisture = "Soil Moisture",
              snow_cover = "Snow Cover", wind_speed = "Wind Speed",
              precipitation = "Precipitation", runoff = "Runoff",
              anseri_abd = "Anseriformes")

# 4c. log1p the right-skewed exposures (bird; fluxes already logged above), then
#     z-scale every exposure across all premises-weeks → effects per +0.5 SD.
cc[, anseri_abd := log1p(anseri_abd)]
setorder(cc, outbreak_id, week_start)
for (v in exposure_vars) {
  cc[, (v) := as.numeric(scale(get(v)))]
}

# 4d. Weekly lag-0..4 columns per exposure, shifted within outbreak.
MAX_LAG_WK <- 4
for (v in exposure_vars) {
  cc[, (paste0(v, "_Lag", 0:MAX_LAG_WK)) :=
       shift(.SD[[1]], n = 0:MAX_LAG_WK, type = "lag"),
     by = outbreak_id, .SDcols = v]
}
```

# 5. Assemble referent designs and the per-exposure stratum filter

```{r designs}
# 5a. unidirectional (1:4) and time-stratified (case calendar month), as in c_03.
design_uni <- cc[lag_weeks_from_case %in% c(0, -1, -2, -3, -4)]
design_uni[, stratum_id := paste0("uni_", outbreak_id)]

case_month <- cc[is_case_week == 1, .(outbreak_id,
                                      c_year  = year(week_start),
                                      c_month = month(week_start))]
design_ts <- merge(cc, case_month, by = "outbreak_id")
design_ts <- design_ts[year(week_start) == c_year & month(week_start) == c_month]
design_ts[, stratum_id := paste0("ts_", outbreak_id)]

# 5b. Keep rows whose every-exposure lag-0..4 is observed (this is what restricts
#     the climate+bird model to premises with ERA5), then strata holding a case
#     AND >=1 control.
prep_strata <- function(dd, vars) {
  lag_cols <- unlist(lapply(vars, function(v) paste0(v, "_Lag", 0:MAX_LAG_WK)))
  dd <- dd[complete.cases(dd[, ..lag_cols])]
  ok <- dd[, .(nc = sum(is_case_week == 1L), nk = sum(is_case_week == 0L)),
           by = stratum_id][nc >= 1L & nk >= 1L, stratum_id]
  dd[stratum_id %in% ok]
}

climate_states <- design_uni[!is.na(temperature) & is_case_week == 1,
                             sort(unique(state))]
cat("States entering the climate+bird model: ",
    paste(climate_states, collapse = ", "), "\n", sep = "")
```

# 6. Fit the joint climate + bird DLNM (both designs)

```{r fit}
# 6a. Weekly crossbasis (linear exposure-response, ns df=2 lag over 0–4 wk) — one
#     per exposure. Named cb_<var> so coefficients are grep-able per predictor.
make_cb <- function(dd, v) crossbasis(
  as.matrix(dd[, paste0(v, "_Lag", 0:MAX_LAG_WK), with = FALSE]),
  lag = c(0, MAX_LAG_WK), argvar = list(fun = "lin"),
  arglag = list(fun = "ns", df = 2))

# 6b. Fit one conditional-logistic DLNM with all 7 exposures jointly, then extract
#     each predictor's cumulative (whole 0–4-wk window) and per-lag OR at +0.5 SD.
fit_joint <- function(dd, vars, design) {
  dd  <- prep_strata(dd, vars)
  cbs <- setNames(lapply(vars, function(v) make_cb(dd, v)), paste0("cb_", vars))
  list2env(cbs, environment())                      # expose cb_* to the formula
  form <- as.formula(paste("is_case_week ~",
                           paste(names(cbs), collapse = " + "),
                           "+ strata(stratum_id)"))
  m <- clogit(form, method = "efron", data = dd)

  beta <- coef(m); V <- vcov(m)
  or_rows <- rbindlist(lapply(seq_along(vars), function(i) {
    cb  <- cbs[[i]]
    ix  <- grep(paste0("^cb_", vars[i], "v"), names(beta))
    b   <- beta[ix]; Vsub <- V[ix, ix, drop = FALSE]
    lb  <- do.call(dlnm::onebasis, c(list(x = 0:MAX_LAG_WK), attr(cb, "arglag")))
    or_ci <- function(contrast) {
      lo <- as.numeric(contrast %*% b)
      se <- sqrt(as.numeric(t(contrast) %*% Vsub %*% contrast))
      list(OR = exp(lo), lcl = exp(lo - 1.96 * se), ucl = exp(lo + 1.96 * se))
    }
    cum <- or_ci(0.5 * colSums(lb))
    per <- rbindlist(lapply(0:MAX_LAG_WK, function(k) {
      v <- or_ci(0.5 * lb[k + 1, ]); data.table(lag_wk = k, OR = v$OR,
                                                 lcl = v$lcl, ucl = v$ucl)
    }))
    data.table(design = design, exposure = vars[i], label = name_map[vars[i]],
               cum_OR = cum$OR, cum_lcl = cum$lcl, cum_ucl = cum$ucl,
               per_lag = list(per))
  }))
  list(model = m, n_strata = uniqueN(dd$stratum_id), ors = or_rows)
}

fit_uni <- fit_joint(design_uni, exposure_vars, "unidirectional")
fit_ts  <- fit_joint(design_ts,  exposure_vars, "time-stratified")

saveRDS(fit_uni$model, paste0(models_folder, "flyway_weekly_climate_bird_unidirectional.RDS"))
saveRDS(fit_ts$model,  paste0(models_folder, "flyway_weekly_climate_bird_timestrat.RDS"))

cum_all <- rbindlist(list(fit_uni$ors, fit_ts$ors))
fmt_or  <- function(OR, lcl, ucl) sprintf("%.2f (%.2f, %.2f)", OR, lcl, ucl)
cat("\nN strata — unidirectional: ", fit_uni$n_strata,
    " | time-stratified: ", fit_ts$n_strata, "\n", sep = "")
print(cum_all[, .(design, label,
                  `OR (95% CI)` = fmt_or(cum_OR, cum_lcl, cum_ucl))])
```

# 7. Table — cumulative ORs, unidirectional vs. time-stratified

```{r table}
cum_all[, OR_CI := fmt_or(cum_OR, cum_lcl, cum_ucl)]
tbl <- dcast(cum_all, label ~ design, value.var = "OR_CI")
setcolorder(tbl, c("label", "unidirectional", "time-stratified"))
setnames(tbl, "label", "Exposure (per +0.5 SD)")

ft <- flextable(tbl) |>
  fontsize(size = 11, part = "all") |> autofit() |>
  align(align = "center", part = "all") |> theme_booktabs()
read_docx() |>
  body_add_flextable(ft) |>
  body_add_par(
    sprintf(paste0("Cumulative odds ratios (per +0.5 SD, lag 0–4 weeks) from the ",
                   "joint weekly climate + bird DLNM, pooled across %d flyway states ",
                   "(%s; IND+INR events). unidirectional = 1:4 matching; ",
                   "time-stratified = case calendar-month controls. N strata = %d."),
            length(climate_states), paste(climate_states, collapse = ", "),
            fit_uni$n_strata),
    style = "Normal") |>
  print(target = paste0(tables_main_folder, "Table_Flyway_Weekly_ClimateBird_OR.docx"))
```

# 8. Figure — cumulative OR forest (both designs)

```{r figure}
forest <- copy(cum_all)
ord <- forest[design == "unidirectional"][order(cum_OR), label]
forest[, label := factor(label, levels = ord)]

fig <- ggplot(forest, aes(cum_OR, label, colour = design)) +
  geom_vline(xintercept = 1, linetype = "dashed", colour = "grey50") +
  geom_errorbarh(aes(xmin = cum_lcl, xmax = cum_ucl),
                 height = 0.25, linewidth = 0.5,
                 position = position_dodge(width = 0.5)) +
  geom_point(size = 2, position = position_dodge(width = 0.5)) +
  scale_colour_manual(values = c(unidirectional = "#32135C",
                                 `time-stratified` = "#C9405A"), name = NULL) +
  labs(x = "Cumulative OR (per +0.5 SD, lag 0–4 wk)", y = NULL,
       title = "Weekly climate + bird and HPAI spillover (flyway, pooled)") +
  theme_spark() +
  theme(axis.text.x = element_text(angle = 0, hjust = 0.5))
ggsave(paste0(figures_main_folder, "figure_flyway_weekly_climate_bird_forest.png"),
       fig, width = 8, height = 5, dpi = 300, bg = "white")
```

# 9. Save result bundle

```{r save}
saveRDS(list(
  cum_all        = cum_all,
  n_strata_uni   = fit_uni$n_strata,
  n_strata_ts    = fit_ts$n_strata,
  climate_states = climate_states,
  designs        = list(unidirectional = design_uni, time_stratified = design_ts)
), paste0(objects_folder, "flyway_weekly_climate_bird.RDS"))

cat("Done — flyway weekly climate+bird case-crossover models, table, and figure written.\n")
```
