---
title: "LASSO Variable Selection"
author: "Fintan Mooney"
date: "2025-10-31"
output: html_document
---

# LASSO penalized conditional logistic regression for variable selection... identifying the most parsimonious set of predictors for the case-crossover HPAI model!

```{r include=FALSE}
rm(list=ls())
project.folder = paste0(print(here::here()),'/')
source(paste0(project.folder,'create_folder_structure.R'))
source(paste0(functions.folder,'script_initiate.R'))
```

## Import casecrossover df

```{r, import casecrossover df}

case_crossover_df <- readRDS(file.path(objects_folder, "case_crossover_df.RDS"))

```


## Crossbasis Terms

```{r, finalized crossbasis}


# Finalized Crossbasis Terms for Climate and Bird Predictors

# Leaf Area Index
cb_leaf_area_index <- crossbasis(
  x = as.matrix(case_crossover_df[, paste0("leaf_area_index_Lag", 0:28)]),
  lag = c(0, 28),
  argvar = list(fun = "lin"),
  arglag = list(fun = "ns", df = 4)
)

# Longwave Radiation
cb_longwave_radiation <- crossbasis(
  x = as.matrix(case_crossover_df[, paste0("longwave_radiation_Lag", 0:28)]),
  lag = c(0, 28),
  argvar = list(fun = "lin"),
  arglag = list(fun = "ns", df = 4)
)

# Precipitation
cb_precipitation <- crossbasis(
  x = as.matrix(case_crossover_df[, paste0("precipitation_Lag", 0:28)]),
  lag = c(0, 28),
  argvar = list(fun = "lin"),
  arglag = list(fun = "ns", df = 4)
)

# Relative Humidity
cb_relative_humidity <- crossbasis(
  x = as.matrix(case_crossover_df[, paste0("relative_humidity_Lag", 0:28)]),
  lag = c(0, 28),
  argvar = list(fun = "lin"),
  arglag = list(fun = "ns", df = 4)
)

# Runoff
cb_runoff <- crossbasis(
  x = as.matrix(case_crossover_df[, paste0("runoff_Lag", 0:28)]),
  lag = c(0, 28),
  argvar = list(fun = "lin"),
  arglag = list(fun = "ns", df = 4)
)

# Shortwave Radiation
cb_shortwave_radiation <- crossbasis(
  x = as.matrix(case_crossover_df[, paste0("shortwave_radiation_Lag", 0:28)]),
  lag = c(0, 28),
  argvar = list(fun = "lin"),
  arglag = list(fun = "ns", df = 4)
)

# Snow Density
cb_snow_density <- crossbasis(
  x = as.matrix(case_crossover_df[, paste0("snow_density_Lag", 0:28)]),
  lag = c(0, 28),
  argvar = list(fun = "lin"),
  arglag = list(fun = "ns", df = 4)
)

# Soil Moisture
cb_soil_moisture <- crossbasis(
  x = as.matrix(case_crossover_df[, paste0("soil_moisture_Lag", 0:28)]),
  lag = c(0, 28),
  argvar = list(fun = "lin"),
  arglag = list(fun = "ns", df = 4)
)

# Temperature
cb_temperature <- crossbasis(
  x = as.matrix(case_crossover_df[, paste0("temperature_Lag", 0:28)]),
  lag = c(0, 28),
  argvar = list(fun = "lin"),
  arglag = list(fun = "ns", df = 4)
)


# Total Bird Abundance
cb_total_bird_abundance <- crossbasis(
  x = as.matrix(case_crossover_df[, paste0("total_bird_abundance_Lag", 0:28)]),
  lag = c(0, 28),
  argvar = list(fun = "lin"),
  arglag = list(fun = "ns", df = 4)
)

# Wind Speed
cb_wind_speed <- crossbasis(
  x = as.matrix(case_crossover_df[, paste0("wind_speed_Lag", 0:28)]),
  lag = c(0, 28),
  argvar = list(fun = "lin"),
  arglag = list(fun = "ns", df = 4)
)

# # # Anseriformes:
#  cb_anseri <- crossbasis(
#    x = as.matrix(case_crossover_df[, paste0("anseriformes_Lag", 0:8)]),
#    lag = c(0, 8),
#    argvar = list(fun = "lin"),
#    arglag = list(fun = "poly", degree = 2)
#  )
#  
# # # Predator:
#  cb_predators <- crossbasis(
#    x = as.matrix(case_crossover_df[, paste0("predators_Lag", 0:8)]),
#    lag = c(0, 8),
#    argvar = list(fun = "lin"),
#    arglag = list(fun = "poly", degree = 2)
#  )




```


## Lets take a quick look at multi-colinearity 

```{r, correlation of variables}


# Correlation Matrix of Predictors at Lag 0

# Include all environmental + bird variables currently used
vars <- c(
  "leaf_area_index",
  "longwave_radiation",
  "shortwave_radiation",
  "precipitation",
  "relative_humidity",
  "soil_moisture",
  "snow_density",
  "temperature",
  "runoff",
  "wind_speed",
  "total_bird_abundance"
)

cor_matrix <- cor(
  case_crossover_df[, paste0(vars, "_Lag0")],
  use = "complete.obs"
)

cor_matrix <- round(cor_matrix, 2)

print(cor_matrix)


# Correlation Heatmap

# Remove "_Lag0" from variable names
colnames(cor_matrix) <- gsub("_Lag0$", "", colnames(cor_matrix))
rownames(cor_matrix) <- gsub("_Lag0$", "", rownames(cor_matrix))


png(
  file.path(figures_folder, "figure_correlation_matrix_lag0.png"),
  width = 8, height = 8, units = "in", res = 300
)

corrplot(
  cor_matrix,
  method = "color",
  type = "upper",
  addCoef.col = "black",
  tl.col = "black",
  tl.srt = 45,
  number.cex = 0.7,
  col = colorRampPalette(c("purple4", "white", "darkgreen"))(200),
  mar = c(0, 0, 2, 0)
)
title("Correlation Matrix of Lag-0 Environmental & Bird Predictors", line = -1)

dev.off()


```

## Cross-validated Penalized Conditional Logistic Regression (LASSO) for Model Stability

```{r, leave one case out validation}


## Rename each crossbasis block ----
rename_cb <- function(cb, prefix) {
  cb_mat <- as.matrix(unclass(cb))
  if (is.null(colnames(cb_mat))) {
    colnames(cb_mat) <- paste0("l", seq_len(ncol(cb_mat)))
  }
  colnames(cb_mat) <- paste0(prefix, "_", colnames(cb_mat))
  cb_mat
}

## Build X matrix using crossbasis variables 
X <- cbind(
  rename_cb(cb_runoff,               "cb_runoff"),
  rename_cb(cb_snow_density,         "cb_snow_density"),
  rename_cb(cb_soil_moisture,        "cb_soil_moisture"),
  rename_cb(cb_temperature,          "cb_temperature"),
  rename_cb(cb_precipitation,        "cb_precipitation"),
  rename_cb(cb_wind_speed,           "cb_wind_speed"),
  rename_cb(cb_total_bird_abundance, "cb_total_bird_abundance")
)

## Define outcome and strata 
y <- case_crossover_df$outbreak_binary
strata <- case_crossover_df$stratum_id

## Keep only strata with both case/control
keep_strata <- names(which(tapply(y, strata, function(z) length(unique(z))) == 2))
sel <- strata %in% keep_strata

X <- X[sel, , drop = FALSE]
y <- y[sel]
strata <- strata[sel]

## Fit penalised model and cross-validate 
fit <- clogitL1(
  x = X,
  y = y,
  strata = strata,
  numLambda = 50,
  minLambdaRatio = 0.001,
  alpha = 1         # LASSO penalty
)

cvfit <- cv.clogitL1(fit, numFolds = 10)
plot(cvfit)

## Extract coefficients at best lambda 
best_lambda <- cvfit$minCV_lambda
idx_best <- which.min(abs(fit$lambda - best_lambda))
beta_best <- fit$beta[idx_best, ]
names(beta_best) <- colnames(X)

### CREATE TABLE OF RESULTS

# Total number of basis terms per parent variable

all_cols <- tibble(variable = colnames(X)) |>
  mutate(
    parent_var = case_when(
      grepl("^cb_runoff",               variable) ~ "Runoff",
      grepl("^cb_snow_density",         variable) ~ "Snow Density",
      grepl("^cb_soil_moisture",        variable) ~ "Soil Moisture",
      grepl("^cb_temperature",          variable) ~ "Temperature",
      grepl("^cb_precipitation",        variable) ~ "Precipitation",
      grepl("^cb_wind_speed",           variable) ~ "Wind Speed",
      grepl("^cb_total_bird_abundance", variable) ~ "Total Bird Abundance",
      TRUE ~ "Unknown"
    )
  )

block_sizes <- all_cols |>
  group_by(parent_var) |>
  summarise(
    n_basis_total = n(),
    .groups = "drop"
  )

# 
# Summarize LASSO coefficients
# 
# All parent variables (including those fully zeroed out)
all_parents <- all_cols %>%
  distinct(parent_var)

# Summarize LASSO coefficients (only non-zero)
coef_table <- data.frame(variable = names(beta_best), beta = beta_best) %>%
  mutate(
    parent_var = case_when(
      grepl("^cb_runoff",               variable) ~ "Runoff",
      grepl("^cb_snow_density",         variable) ~ "Snow Density",
      grepl("^cb_soil_moisture",        variable) ~ "Soil Moisture",
      grepl("^cb_temperature",          variable) ~ "Temperature",
      grepl("^cb_precipitation",        variable) ~ "Precipitation",
      grepl("^cb_wind_speed",           variable) ~ "Wind Speed",
      grepl("^cb_total_bird_abundance", variable) ~ "Total Bird Abundance",
      TRUE ~ "Unknown"
    )
  )

summary_vars <- coef_table %>%
  filter(beta != 0) %>%
  group_by(parent_var) %>%
  summarise(
    n_active      = n(),
    n_positive    = sum(beta > 0),
    n_negative    = sum(beta < 0),
    mean_beta     = mean(beta),
    mean_abs_beta = mean(abs(beta)),
    max_abs_beta  = max(abs(beta)),
    sum_abs_beta  = sum(abs(beta)),
    .groups = "drop"
  ) %>%
  right_join(block_sizes, by = "parent_var") %>%
  mutate(
    across(c(n_active, n_positive, n_negative), ~ replace_na(.x, 0L)),
    across(c(mean_beta, mean_abs_beta, max_abs_beta, sum_abs_beta), ~ replace_na(.x, 0)),
    prop_active = n_active / n_basis_total
  ) %>%
  arrange(desc(sum_abs_beta))

summary_vars_export <- summary_vars |>
  mutate(
    mean_beta     = round(mean_beta,     3),
    mean_abs_beta = round(mean_abs_beta, 3),
    max_abs_beta  = round(max_abs_beta,  3),
    sum_abs_beta  = round(sum_abs_beta,  3),
    prop_active   = round(prop_active,   3)
  )


# Save Plot
png(
  file.path(figures_folder, "figure_clogitL1_model_stability.png"),
  width = 7, height = 5, units = "in", res = 300
)
plot(cvfit)
title("Cross-validated Penalized Conditional Logistic Regression (Model Stability)", line = 1)
dev.off()

# Build table to export to word
ft_stability <- summary_vars_export |>
  select(
    parent_var,
    n_active,
    prop_active,
    mean_beta,
    mean_abs_beta,
    sum_abs_beta
  ) |>
  regulartable() |>
  set_header_labels(
    parent_var   = "Variable",
    n_active     = "Active Terms (β ≠ 0)",
    prop_active  = "Proportion Active",
    mean_beta    = "Mean β",
    mean_abs_beta= "Mean |β|",
    sum_abs_beta = "Σ |β|"
  ) |>
  align(part = "header", align = "center") |>
  align(
    j = c("n_active", "prop_active", "mean_beta", "mean_abs_beta", "sum_abs_beta"),
    align = "right", part = "body"
  ) |>
  autofit()


# Export
doc <- read_docx()

doc <- body_add_par(
  doc,
  "Table Sx. Variable-level LASSO selection summary for crossbasis predictors",
  style = "Normal"
)

doc <- body_add_par(
  doc,
  "Summary of non-zero penalized conditional logistic regression coefficients at the optimal λ, grouped by parent environmental and bird abundance variables.",
  style = "Normal"
)

doc <- body_add_flextable(doc, ft_stability)

print(
  doc,
  target = file.path(tables_folder, "Table_ClogitL1_Model_Stability.docx")
)


## Temperature and wind speed have smallest effect

# pick one representative temp basis column
tc <- grep("^cb_temperature_", colnames(X), value = TRUE)[1]

within_sd <- tapply(X[, tc], strata, sd, na.rm = TRUE)
summary(within_sd)
mean(within_sd == 0, na.rm = TRUE)

temp_cols <- grep("^cb_temperature_", colnames(X), value = TRUE)
soil_cols <- grep("^cb_soil_moisture_", colnames(X), value = TRUE)
prec_cols <- grep("^cb_precipitation_", colnames(X), value = TRUE)

cor(X[, temp_cols[1]], X[, soil_cols[1]])  # crude check
cor(X[, temp_cols[1]], X[, prec_cols[1]])

```

## Temperature and wind speed have smallest effect, so lets fit a reduced model without them

```{r, reduced model from cross validation}

reduced_model <- clogit(
  outbreak_binary ~
    cb_precipitation +
    cb_soil_moisture +
    cb_total_bird_abundance +
    cb_runoff + 
    cb_snow_density +
    strata(stratum_id),
  data = case_crossover_df,
  method = "efron"
)


```

## Results 

```{r, report reduced model results}


# Crosspred Objects (Reduced Model)

cp_runoff <- crosspred(cb_runoff, reduced_model, cen = 0, at = 0.5, cumul = TRUE)
cp_snow   <- crosspred(cb_snow_density, reduced_model, cen = 0, at = 0.5, cumul = TRUE)
cp_soil   <- crosspred(cb_soil_moisture, reduced_model, cen = 0, at = 0.5, cumul = TRUE)
cp_precip <- crosspred(cb_precipitation, reduced_model, cen = 0, at = 0.5, cumul = TRUE)
cp_bird   <- crosspred(cb_total_bird_abundance, reduced_model, cen = 0, at = 0.5, cumul = TRUE)

cp_slice_runoff <- crosspred(cb_runoff, reduced_model, cen = 0, at = 0.5)
cp_slice_snow   <- crosspred(cb_snow_density, reduced_model, cen = 0, at = 0.5)
cp_slice_soil   <- crosspred(cb_soil_moisture, reduced_model, cen = 0, at = 0.5)
cp_slice_precip <- crosspred(cb_precipitation, reduced_model, cen = 0, at = 0.5)
cp_slice_bird   <- crosspred(cb_total_bird_abundance, reduced_model, cen = 0, at = 0.5)

# Cumulative OR Summary Table

cumulative_summary <- tibble(
  Predictor = c("Runoff", "Snow Density", "Soil Moisture", 
                "Precipitation", "Total Bird Abundance"),
  OR = c(
    cp_runoff$allRRfit["0.5"],
    cp_snow$allRRfit["0.5"],
    cp_soil$allRRfit["0.5"],
    cp_precip$allRRfit["0.5"],
    cp_bird$allRRfit["0.5"]
  ),
  CI_low = c(
    cp_runoff$allRRlow["0.5"],
    cp_snow$allRRlow["0.5"],
    cp_soil$allRRlow["0.5"],
    cp_precip$allRRlow["0.5"],
    cp_bird$allRRlow["0.5"]
  ),
  CI_high = c(
    cp_runoff$allRRhigh["0.5"],
    cp_snow$allRRhigh["0.5"],
    cp_soil$allRRhigh["0.5"],
    cp_precip$allRRhigh["0.5"],
    cp_bird$allRRhigh["0.5"]
  )
) %>%
  mutate(`OR (95% CI)` = sprintf("%.2f (%.2f, %.2f)", OR, CI_low, CI_high)) %>%
  select(Predictor, `OR (95% CI)`) %>%
  gt() %>%
  tab_header(
    title = md("**Case-Crossover Model: Cumulative Odds Ratios**"),
    subtitle = "Per 0.5 SD Increase"
  ) %>%
  cols_label(`OR (95% CI)` = "Odds Ratio (95% Confidence Interval)") %>%
  tab_options(
    table.font.size = 14,
    heading.title.font.size = 16,
    heading.subtitle.font.size = 14,
    data_row.padding = px(6)
  )

# Plotting Functions

plot_cumulative <- function(cp, title) {
  plot(cp, type = "cumulative", var = 0.5, ci = "area",
       col = "black", ci.arg = list(col = rgb(0,0,0,0.1), border = NA),
       lwd = 1.5, main = title, xlab = "", ylab = "", las = 1)
  abline(h = 1, lty = 2, col = "gray40")
}

plot_slices <- function(cp, title) {
  plot(cp, type = "slices", var = 0.5, ci = "area",
       col = "black", ci.arg = list(col = rgb(0,0,0,0.1), border = NA),
       lwd = 1.3, main = title, xlab = "", ylab = "", las = 1)
  abline(h = 1, lty = 2, col = "gray40")
}


# Save Cumulative Plot

png(file.path(figures_folder, "figure_cumulative_ORs_casecrossover_reduced.png"), 
    width = 11, height = 7, units = "in", res = 300)

par(
  mfrow = c(2, 3),
  mar = c(4.2, 4.2, 2.5, 1),
  oma = c(1, 1, 0, 1),
  cex.main = 1.2,
  cex.lab = 1.2,
  cex.axis = 1.1,
  tcl = -0.3, mgp = c(2.5, 0.7, 0)
)

plot_cumulative(cp_runoff, "Runoff")
plot_cumulative(cp_snow, "Snow Density")
plot_cumulative(cp_soil, "Soil Moisture")
plot_cumulative(cp_precip, "Precipitation")
plot_cumulative(cp_bird, "Total Bird Abundance")

mtext("Lag (days)", side = 1, outer = TRUE, line = -1, cex = 1.2)
mtext("Odds Ratio (95% CI)", side = 2, outer = TRUE, line = -1, cex = 1.2)

dev.off()


# Save Slices Plot

png(file.path(figures_folder, "figure_slice_ORs_casecrossover_reduced.png"), 
    width = 11, height = 7, units = "in", res = 300)

par(
  mfrow = c(2, 3),
  mar = c(4.2, 4.2, 2.5, 1),
  oma = c(1, 1, 0, 1),
  cex.main = 1.2,
  cex.lab = 1.2,
  cex.axis = 1.1,
  tcl = -0.3, mgp = c(2.5, 0.7, 0)
)

plot_slices(cp_slice_runoff, "Runoff")
plot_slices(cp_slice_snow, "Snow Density")
plot_slices(cp_slice_soil, "Soil Moisture")
plot_slices(cp_slice_precip, "Precipitation")
plot_slices(cp_slice_bird, "Total Bird Abundance")

mtext("Lag (days)", side = 1, outer = TRUE, line = -1, cex = 1.2)
mtext("Odds Ratio (95% CI)", side = 2, outer = TRUE, line = -1, cex = 1.2)

dev.off()


# -- Lag-Specific Odds Ratio Table (Reduced Case-Crossover Model) --

# Helper Function

extract_lag_ORs_wide <- function(cp_obj, varname, varval = "0.5", lags = 0:4) {
  lag_labels <- paste0("lag", lags)
  or   <- cp_obj$matRRfit[varval, lag_labels]
  low  <- cp_obj$matRRlow[varval, lag_labels]
  high <- cp_obj$matRRhigh[varval, lag_labels]
  
  tibble(
    Predictor = varname,
    !!!setNames(as.list(or),   paste0("OR_Lag", lags)),
    !!!setNames(as.list(low),  paste0("Low_Lag", lags)),
    !!!setNames(as.list(high), paste0("High_Lag", lags)),
    !!!setNames(as.list(sprintf("%.2f (%.2f, %.2f)", or, low, high)), paste0("Lag ", lags))
  )
}


# Create Lag-Specific Table

table_lag_ORs <- bind_rows(
  extract_lag_ORs_wide(cp_slice_runoff, "Runoff"),
  extract_lag_ORs_wide(cp_slice_snow, "Snow Density"),
  extract_lag_ORs_wide(cp_slice_soil, "Soil Moisture"),
  extract_lag_ORs_wide(cp_slice_precip, "Precipitation"),
  extract_lag_ORs_wide(cp_slice_bird, "Total Bird Abundance")
)


# Identify Significant Lags

lags <- c(0, 7, 14, 21, 28)  # weekly anchor lags (daily resolution)
highlight_matrix <- sapply(lags, function(i) {
  table_lag_ORs[[paste0("Low_Lag", i)]] > 1 | table_lag_ORs[[paste0("High_Lag", i)]] < 1
})
colnames(highlight_matrix) <- paste0("Lag ", lags)


# Build Flextable

ft <- table_lag_ORs %>%
  select(Predictor, paste0("Lag ", lags)) %>%
  flextable() %>%
  set_header_labels(Predictor = "Predictor") %>%
  add_header_row(values = c("Predictor", "Odds Ratio (95% CI)"), colwidths = c(1, length(lags))) %>%
  fontsize(size = 11, part = "all") %>%
  autofit() %>%
  align(align = "center", part = "all") %>%
  theme_booktabs()

# Bold statistically significant ORs
for (i in lags) {
  ft <- bold(ft, i = which(highlight_matrix[, paste0("Lag ", i)]), j = paste0("Lag ", i))
}


# Export to Word

doc <- read_docx() %>%
  body_add_flextable(ft) %>%
  body_add_par(
    "Table: Lag-specific odds ratios for the reduced case-crossover model assessing the association between environmental conditions and Highly Pathogenic Avian Influenza outbreaks in Minnesota, 2022. Estimates are per 0.5 SD increase in each predictor. Statistically significant odds ratios are shown in bold.",
    style = "Normal"
  )

print(doc, target = file.path(tables_folder, "Table_LagSpecific_OR_CaseCrossover_Reduced.docx"))

ft




```

## Test AIC of reduced model against the full model

```{r, test AIC of reduced model against the full model}


final_model <- clogit(
  outbreak_binary ~ 
    cb_runoff + cb_snow_density + cb_soil_moisture + cb_temperature +
    cb_precipitation + cb_wind_speed + cb_total_bird_abundance +
    strata(stratum_id),
  method = "efron",
  data = case_crossover_df
)

AIC(reduced_model, final_model)



```


