---
title: "Multi-state DLNM specification diagnostics (independent events)"
author: "Fintan Mooney"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

```{r initiate}
rm(list = ls())
# 0a. Declare root directory, folder location and load essential stuff
project.folder = paste0(print(here::here()), '/')
source(paste0(project.folder, 'create_folder_structure.R'))
source(paste0(functions.folder, 'script_initiate.R'))
conflict_prefer("year",  "lubridate", quiet = TRUE)
conflict_prefer("month", "lubridate", quiet = TRUE)
```

Diagnostics to decide the DLNM lag-basis specification for the gridded multi-state
case-crossover models (`c_05` daily, `c_06` weekly), independent events only. Three checks:

1. **Events-per-parameter (EPV)** across candidate lag df — is the daily model over-parameterised?
2. **Collinearity** among the 7 exposures (correlation matrix + VIF) — what is driving the wide daily CIs?
3. **AIC** of the daily/weekly joint models across lag df — does the extra flexibility earn its parameters?

```{r load}
daily  <- setDT(readRDS(paste0(objects_folder, "case_crossover_df_multistate_daily.RDS")))
weekly <- setDT(readRDS(paste0(objects_folder, "case_crossover_df_multistate_weekly.RDS")))

predictors <- c("runoff", "soil_moisture", "temperature", "precipitation",
                "wind_speed", "snow_cover", "anseriformes")
pred_labels <- c(runoff = "Runoff", soil_moisture = "Soil Moisture",
                 temperature = "Temperature", precipitation = "Precipitation",
                 wind_speed = "Wind Speed", snow_cover = "Snow Cover",
                 anseriformes = "Anseriformes")
```

# 1. Fit one spec, return params / EPV / AIC

```{r fit-spec}
# Build the joint clogit DLNM at a given met-lag df and report fit metrics. In the daily
# model the bird term keeps its weekly-strata basis (4 params) regardless of met df; in the
# weekly model every term shares ns(df=k).
fit_spec <- function(df, res, met_df) {
  max_lag <- if (res == "daily") 28 else 4
  cbs <- setNames(lapply(predictors, function(v) {
    X  <- as.matrix(df[, paste0(v, "_Lag", 0:max_lag), with = FALSE])
    al <- if (res == "daily" && v == "anseriformes") list(fun = "strata", breaks = c(7, 14, 21))
          else list(fun = "ns", df = met_df)
    crossbasis(X, lag = c(0, max_lag), argvar = list(fun = "lin"), arglag = al)
  }), paste0("cb_", predictors))
  list2env(cbs, environment())
  m <- clogit(as.formula(paste("outbreak_binary ~", paste(names(cbs), collapse = " + "),
                               "+ strata(stratum_id)")), method = "efron", data = df)
  n_cases <- sum(df$outbreak_binary == 1); k <- length(coef(m))
  data.table(resolution = res, met_lag_df = met_df, n_params = k, n_cases = n_cases,
             EPV = round(n_cases / k, 1), AIC = round(AIC(m), 1))
}
```

# 2. EPV + AIC across candidate lag df

```{r epv-aic}
daily_specs  <- rbindlist(lapply(c(2, 3, 4), function(k) fit_spec(daily,  "daily",  k)))
weekly_specs <- rbindlist(lapply(c(1, 2, 3), function(k) fit_spec(weekly, "weekly", k)))

cat("DAILY (current primary = met ns df=3; bird strata adds 4 params):\n")
print(daily_specs)
cat("\nWEEKLY (current primary = ns df=2):\n")
print(weekly_specs)
cat("\nGuide: EPV >= 10 conventional; dAIC < 2 = no meaningful gain from the extra df.\n")
```

# 3. Collinearity among the 7 exposures (contemporaneous, lag 0)

```{r collinearity}
# Correlation matrix and variance-inflation factors from the lag-0 (contemporaneous)
# z-scaled exposures. VIF_j = j-th diagonal of the inverse correlation matrix (standardised
# predictors). Exposures are highly autocorrelated across lags, so lag 0 is a fair proxy
# for the cross-exposure collinearity entering the joint model.
lag0  <- paste0(predictors, "_Lag0")
X0    <- as.matrix(daily[, ..lag0]); colnames(X0) <- predictors
R     <- cor(X0, use = "complete.obs")
VIF   <- diag(solve(R))

cat("Correlation matrix (lag 0):\n")
print(round(R, 2))

vif_tbl <- data.table(Predictor = pred_labels[predictors], VIF = round(VIF, 2))[order(-VIF)]
cat("\nVariance inflation factors (lag 0):\n")
print(vif_tbl)
cat("\nGuide: VIF > 5 = notable collinearity; VIF > 10 = severe.\n")
```

# 4. Read-out

```{r summary}
best_daily  <- daily_specs[which.min(AIC)]
hi_vif      <- vif_tbl[VIF > 5]
cat(sprintf("Daily: current df=3 gives EPV %.1f (params %d). df=2 gives EPV %.1f.\n",
            daily_specs[met_lag_df == 3, EPV], daily_specs[met_lag_df == 3, n_params],
            daily_specs[met_lag_df == 2, EPV]))
cat(sprintf("Lowest-AIC daily spec: met df=%d (AIC %.1f). dAIC(df2 - df3) = %.1f.\n",
            best_daily$met_lag_df, best_daily$AIC,
            daily_specs[met_lag_df == 2, AIC] - daily_specs[met_lag_df == 3, AIC]))
cat(sprintf("Exposures with VIF > 5: %s\n",
            if (nrow(hi_vif)) paste(hi_vif$Predictor, collapse = ", ") else "none"))
```
