---
title: "Multi-state case-crossover — seasonal-anomaly vs raw (unidirectional)"
author: "Fintan Mooney"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

```{r initiate}
rm(list = ls())
project.folder = paste0(print(here::here()), '/')
source(paste0(project.folder, 'create_folder_structure.R'))
source(paste0(functions.folder, 'script_initiate.R'))
conflict_prefer("year",  "lubridate", quiet = TRUE)
conflict_prefer("month", "lubridate", quiet = TRUE)
library(splines)
```

Directly tests Reviewer #1's seasonal-confounding critique of the unidirectional design.
Each exposure is **deseasonalized** — for every 10 km cell we fit a smooth spline of date
(`ns(date, df=6)`) across the study period and take the residual, i.e. the short-term
**anomaly** (departure from that cell's own seasonal cycle). We then rebuild the same
**unidirectional** case-crossover (case + 4 prior control periods) on anomalies and compare
per-lag ORs to the raw-value model. Signals that survive deseasonalization are genuine
short-term departures; signals that collapse were seasonal trend.

Note: the multi-state ERA5 pull is 2021–2022 only, so there is no multi-year climatology
(that exists for MN alone). Deseasonalizing *within* the study period via a per-cell spline
is the defensible single-season approach; `df` is a tunable smoothness knob (6 ≈ bi-monthly
knots — removes the seasonal cycle while preserving ≤4-week variation, the referent window).

# 1. Load panels and settings

```{r load}
daily  <- setDT(readRDS(paste0(objects_folder, "multistate_fishnet_timeseries_daily.RDS")))
weekly <- setDT(readRDS(paste0(objects_folder, "multistate_fishnet_timeseries_weekly.RDS")))
screening <- readRDS(paste0(objects_folder, "multistate_independence_screening.RDS"))

case_window_start <- as.Date("2022-01-01"); case_window_end <- as.Date("2022-12-31")
DF_SEASON <- 6                                   # spline df for the per-cell seasonal fit
OFFSETS   <- c(0, 7, 14, 21, 28)                 # unidirectional: case + 4 prior periods (days)

predictors  <- c("runoff", "soil_moisture", "temperature", "precipitation",
                 "wind_speed", "snow_cover", "anseriformes")
pred_labels <- c(runoff = "Runoff", soil_moisture = "Soil Moisture", temperature = "Temperature",
                 precipitation = "Precipitation", wind_speed = "Wind Speed",
                 snow_cover = "Snow Cover", anseriformes = "Anseriformes")
skewed <- c("runoff", "precipitation", "anseriformes")

remove_ids <- union(screening$cross_farm_ids, screening$ids_500m)   # independent def
count_flagged_ids <- function(id_str, ids) {
  if (is.na(id_str)) return(0L)
  as.integer(sum(as.numeric(trimws(unlist(strsplit(id_str, ",")))) %in% ids))
}

# m -> mm before any log1p (panels store runoff/precip in metres).
daily[,  `:=`(runoff = runoff * 1000, precipitation = precipitation * 1000)]
weekly[, `:=`(runoff = runoff * 1000, precipitation = precipitation * 1000)]
```

# 2. Prepare raw and anomaly exposure panels

```{r prep}
# Raw: log1p the skewed exposures (matches c_05/c_06). Anomaly: log1p, then subtract the
# per-cell seasonal spline (residual = short-term anomaly). Both are z-scaled later.
deseason <- function(y, tnum, df) as.numeric(resid(lm(y ~ ns(tnum, df = df), na.action = na.exclude)))

prep_panel <- function(panel, tcol, anomaly) {
  p <- copy(panel)
  p[, tnum := as.numeric(as.Date(get(tcol)))]
  for (v in predictors) {
    if (v %in% skewed) p[, (v) := log1p(get(v))]
    if (anomaly)       p[, (v) := deseason(get(v), tnum, DF_SEASON), by = zone_id]
  }
  p
}
```

# 3. Unidirectional builder (z-scale only; exposures already transformed)

```{r builder}
build_uni <- function(panel, tcol, max_lag) {
  p <- copy(panel); p[, tkey := as.Date(get(tcol))]
  p[, is_outbreak_raw := as.integer(outbreak_binary == 1)]
  fl <- p[outbreak_binary == 1 & tkey >= case_window_start & tkey <= case_window_end
        ][, nf := vapply(outbreak_id, count_flagged_ids, integer(1), ids = remove_ids)
        ][nf > 0, .(zone_id, tkey, nf)]
  p <- merge(p, fl, by = c("zone_id", "tkey"), all.x = TRUE)
  p[, nf := fcoalesce(nf, 0L)]; p[, outbreak_binary := as.integer(outbreak_count - nf > 0)]

  cases <- p[outbreak_binary == 1 & tkey >= case_window_start & tkey <= case_window_end,
             .(zone_id, ct = tkey)]
  cases[, `:=`(stratum_id = paste0("z", zone_id, "_", format(ct, "%Y%m%d")), cid = .I)]

  src <- p[zone_id %in% unique(cases$zone_id)]; setorder(src, zone_id, tkey)
  for (v in predictors) src[, (paste0(v, "_Lag", 0:max_lag)) :=
                             shift(.SD[[1]], n = 0:max_lag, type = "lag"), by = zone_id, .SDcols = v]

  sets <- cases[, .(tkey = ct - OFFSETS, is_case = c(1L, 0L, 0L, 0L, 0L)),
                by = .(cid, zone_id, stratum_id)]
  m <- merge(sets, src, by = c("zone_id", "tkey"), all.x = FALSE)
  m <- m[is_case == 1L | is_outbreak_raw == 0L]
  keep <- m[, .(nc = sum(is_case == 1L), nk = sum(is_case == 0L)), by = stratum_id][nc >= 1 & nk >= 1, stratum_id]
  m <- m[stratum_id %in% keep]
  m[, outbreak_binary := is_case]
  for (v in predictors) for (k in 0:max_lag) {
    col <- paste0(v, "_Lag", k); m[[col]] <- as.numeric(scale(m[[col]]))
  }
  m[]
}
```

# 4. Fit DLNM + per-lag OR extraction

```{r fit-extract}
fit_dlnm <- function(df, res) {
  setDT(df); max_lag <- if (res == "daily") 28 else 4
  arglag_met  <- if (res == "daily") list(fun = "ns", df = 3) else list(fun = "ns", df = 2)
  arglag_bird <- if (res == "daily") list(fun = "strata", breaks = c(7, 14, 21)) else list(fun = "ns", df = 2)
  cbs <- setNames(lapply(predictors, function(v) {
    X  <- as.matrix(df[, paste0(v, "_Lag", 0:max_lag), with = FALSE])
    crossbasis(X, lag = c(0, max_lag), argvar = list(fun = "lin"),
               arglag = if (v == "anseriformes") arglag_bird else arglag_met)
  }), paste0("cb_", predictors))
  list2env(cbs, environment())
  m <- clogit(as.formula(paste("outbreak_binary ~", paste(names(cbs), collapse = " + "),
                               "+ strata(stratum_id)")), method = "efron", data = df)
  list(model = m, cbs = cbs, max_lag = max_lag)
}

perlag <- function(fit, lags) {
  beta <- coef(fit$model); V <- vcov(fit$model); ll <- 0:fit$max_lag
  rbindlist(lapply(predictors, function(v) {
    ix <- grep(paste0("^cb_", v, "v"), names(beta)); b <- beta[ix]; Vs <- V[ix, ix, drop = FALSE]
    lb <- do.call(dlnm::onebasis, c(list(x = ll), attr(fit$cbs[[paste0("cb_", v)]], "arglag")))
    rbindlist(lapply(lags, function(L) {
      ct <- 0.5 * lb[L + 1, ]; lo <- sum(ct * b); se <- sqrt(as.numeric(t(ct) %*% Vs %*% ct))
      data.table(label = pred_labels[[v]], lag = L,
                 OR = exp(lo), lcl = exp(lo - 1.96 * se), ucl = exp(lo + 1.96 * se))
    }))
  }))
}

run_one <- function(panel, tcol, res, anomaly) {
  df  <- build_uni(prep_panel(panel, tcol, anomaly), tcol, if (res == "daily") 28 else 4)
  fit <- fit_dlnm(df, res)
  list(df = df, fit = fit)
}

wk_raw  <- run_one(weekly, "week_start", "weekly", FALSE)
wk_anom <- run_one(weekly, "week_start", "weekly", TRUE)
dy_raw  <- run_one(daily,  "date",       "daily",  FALSE)
dy_anom <- run_one(daily,  "date",       "daily",  TRUE)

saveRDS(wk_anom$fit$model, paste0(models_folder, "casecrossover_multistate_weekly_anomaly.RDS"))
saveRDS(dy_anom$fit$model, paste0(models_folder, "casecrossover_multistate_daily_anomaly.RDS"))
```

# 5. Weekly comparison — raw vs anomaly (per-lag, the manuscript metric)

```{r weekly-compare}
star <- function(lo, hi) ifelse(lo > 1 | hi < 1, "*", "")
mk <- function(x, tag) { x[, cell := sprintf("%.2f (%.2f,%.2f)%s", OR, lcl, ucl, star(lcl, ucl))]; x[, model := tag]; x }
wk <- rbindlist(list(mk(perlag(wk_raw$fit, 0:4), "raw"), mk(perlag(wk_anom$fit, 0:4), "anomaly")))
wk_wide <- dcast(wk[lag %in% c(0, 4)], label + lag ~ model, value.var = "cell")
setorder(wk_wide, label, lag)
cat("WEEKLY per-lag OR (per +0.5 SD), raw vs seasonal-anomaly; * = CI excludes 1\n")
print(wk_wide)
```

# 6. Daily comparison — raw vs anomaly

```{r daily-compare}
dy <- rbindlist(list(mk(perlag(dy_raw$fit, c(0,7,14,21,28)), "raw"),
                     mk(perlag(dy_anom$fit, c(0,7,14,21,28)), "anomaly")))
dy_wide <- dcast(dy[lag %in% c(0, 7, 28)], label + lag ~ model, value.var = "cell")
setorder(dy_wide, label, lag)
cat("\nDAILY per-lag OR (per +0.5 SD), raw vs seasonal-anomaly; * = CI excludes 1\n")
print(dy_wide)
```

# 7. Figure — weekly per-lag curves, raw vs anomaly

```{r figure, fig.width = 10, fig.height = 7}
allwk <- rbindlist(list(perlag(wk_raw$fit, 0:4)[, model := "Raw value"],
                        perlag(wk_anom$fit, 0:4)[, model := "Seasonal anomaly"]))
allwk[, label := factor(label, levels = pred_labels[predictors])]
fig <- ggplot(allwk, aes(lag, OR, colour = model, fill = model)) +
  geom_hline(yintercept = 1, linetype = "dashed", colour = "grey40") +
  geom_ribbon(aes(ymin = lcl, ymax = ucl), alpha = 0.12, colour = NA) +
  geom_line(linewidth = 0.7) + geom_point(size = 1.6) +
  facet_wrap(~ label, ncol = 3, scales = "free_y") +
  scale_colour_manual(values = c("Raw value" = "#B2182B", "Seasonal anomaly" = "#2166AC"), name = NULL) +
  scale_fill_manual(values = c("Raw value" = "#B2182B", "Seasonal anomaly" = "#2166AC"), guide = "none") +
  labs(title = "Weekly per-lag OR — raw vs seasonal anomaly (unidirectional, 7-state)",
       subtitle = "If the signal collapses under deseasonalization, it was seasonal trend",
       x = "Lag (weeks)", y = "OR (95% CI) per +0.5 SD") +
  theme_minimal(base_family = "Avenir") +
  theme(legend.position = "top", plot.title = element_text(family = "Avenir Heavy"),
        strip.text = element_text(family = "Avenir Heavy"), panel.grid.minor = element_blank())
ggsave(paste0(figures_main_folder, "figure_multistate_weekly_raw_vs_anomaly.png"),
       fig, width = 10, height = 7, dpi = 300, bg = "white")
cat("\nWrote raw-vs-anomaly weekly figure + anomaly models.\n")
```

# 8. Sensitivity — robustness to the deseasonalizing smoothness (df = 4 / 6 / 8)

```{r df-sensitivity}
# The seasonal-anomaly result should not hinge on the spline df. Re-deseasonalize the weekly
# panel at df = 4 (looser, ~quarterly knots), 6 (primary), and 8 (tighter) and compare the
# per-lag ORs at each signal's key lag. If climate stays null and birds stay significant
# across all three, the collapse is a robust finding, not an artifact of over/under-smoothing.
key <- data.table(
  label = c("Runoff", "Soil Moisture", "Wind Speed", "Temperature", "Temperature",
            "Snow Cover", "Precipitation", "Anseriformes", "Anseriformes"),
  lag   = c(0L, 4L, 4L, 0L, 4L, 4L, 4L, 4L, 0L))
orstar <- function(pl) pl[, .(label, lag, v = sprintf("%.2f%s", OR, ifelse(lcl > 1 | ucl < 1, "*", "")))]

sens <- merge(key, orstar(perlag(wk_raw$fit, 0:4))[, .(label, lag, raw = v)],
              by = c("label", "lag"), all.x = TRUE)
for (d in c(4, 6, 8)) {
  DF_SEASON <- d
  fit_d <- fit_dlnm(build_uni(prep_panel(weekly, "week_start", TRUE), "week_start", 4), "weekly")
  pl <- orstar(perlag(fit_d, 0:4)); setnames(pl, "v", paste0("anom_df", d))
  sens <- merge(sens, pl, by = c("label", "lag"), all.x = TRUE)
}
DF_SEASON <- 6
setorder(sens, label, lag)
cat("WEEKLY key per-lag OR (per +0.5 SD) — raw vs seasonal anomaly at df 4/6/8; * = CI excludes 1\n")
print(sens)
```
