---
title: "Descriptive Statistics - Table 1"
author: "Fintan Mooney"
date: "2025-10-31"
output:
  word_document: default
  html_document: default
---
# Table 1 for the HPAI case-crossover study — descriptive stats for case and control zones

```{r include=FALSE}
rm(list=ls())
project.folder = paste0(print(here::here()),'/')
source(paste0(project.folder,'create_folder_structure.R'))
source(paste0(functions.folder,'script_initiate.R'))
```

```{r, read in timeseries df}

fishnet_timeseries<- readRDS(file.path(objects_folder, "fishnet_timeseries.RDS"))

```

## Filter to remove cells not studied (did not include a spillover or a feedlot)

```{r, filter to remove cells not studied}

fishnet_timeseries_clean <- fishnet_timeseries %>%
  filter(any_outbreak == 1 | feedlot_count > 0)

```


```{r, list columns}
(colnames(fishnet_timeseries_clean))

```
## Subset to variables of interest

```{r, subset variables}

vars_subset <- c(
  # --- Absolute environmental variables ---
  "leaf_area_index",
  "longwave_radiation",
  "precipitation",
  "relative_humidity",
  "runoff",                 
  "shortwave_radiation",
  "snow_density",
  "soil_moisture",
  "temperature",
  "anseriformes",
  "predators",
  "wind_speed",

  # --- 25-year z-score anomalies ---
  "leaf_area_index_z_25yr",
  "longwave_radiation_z_25yr",
  "precipitation_z_25yr",
  "relative_humidity_z_25yr",
  "runoff_z_25yr",
  "shortwave_radiation_z_25yr",
  "snow_density_z_25yr",
  "soil_moisture_z_25yr",
  "temperature_z_25yr",
  "wind_speed_z_25yr",
  
    # --- 25-year mean  ---
  "leaf_area_index_mean_25yr",
  "longwave_radiation_mean_25yr",
  "precipitation_mean_25yr",
  "relative_humidity_mean_25yr",
  "runoff_mean_25yr",
  "shortwave_radiation_mean_25yr",
  "snow_density_mean_25yr",
  "soil_moisture_mean_25yr",
  "temperature_mean_25yr",
  "wind_speed_mean_25yr",

  # --- ID, time, outbreak, and context ---
  "zone_id",
  "week",
  "season",
  "total_bird_abundance",
  "any_outbreak",
  "land_cover_mode",
  "land_cover_label",
  "outbreak_count",
  "outbreak_binary",
  "feedlot_count",
  "au_count",
  "birds_au"   
)

fishnet_subset <- fishnet_timeseries_clean %>%
  select(any_of(vars_subset))  

# Check

glimpse(fishnet_subset)
colnames(fishnet_subset)

fishnet_subset %>%
  select(where(is.numeric)) %>% # Select only numeric columns
  summarise(across(everything(), list(mean = mean, sd = sd), na.rm = TRUE))
```

## Transform variables 

```{r, transform variables}

fishnet_t1 <- fishnet_subset %>%
  mutate(
    ## Temperature: Kelvin -> Celsius
    temperature_C = temperature - 273.15,

    ## Precipitation & Runoff: meters -> millimeters
    precipitation_mm = precipitation * 1000,
    runoff_mm        = runoff * 1000,

    ## Radiation: convert from W/m² to MJ/m² (divide by 1e6)
    longwave_radiation  = longwave_radiation  / 1e6,
    shortwave_radiation = shortwave_radiation / 1e6,

    ## Drop previous factor levels
    land_cover_label = factor(droplevels(land_cover_label)),
    
    ## Make outbreak count categorical
    
    #outbreak_count_factor = factor(outbreak_count),

    ## Grouping variable as labeled factor
    any_outbreak = factor(any_outbreak,
                          levels = c(0, 1),
                          labels = c("Cells with no outbreak", "Cells with an outbreak"))
  )

fishnet_t1$land_cover_label <- forcats::fct_infreq(fishnet_t1$land_cover_label)


# Check transformations
fishnet_t1 %>%
  summarise(
    across(
      c(temperature_C, precipitation_mm, runoff_mm,
        shortwave_radiation, longwave_radiation),
      list(mean = ~mean(.x, na.rm = TRUE), sd = ~sd(.x, na.rm = TRUE))
    )
  )

```

# Creeate Table 1

```{r, table 1}

# Categorical variable renderer: Percent only
my.render.cat <- function(x) {
  c("", sapply(stats.default(x), function(y) with(y,
    sprintf("%0.0f%%", PCT))))
}


pvalue <- function(x, ...) {
  # Construct vectors of data y, and groups (strata) g
  y <- unlist(x)
  g <- factor(rep(1:length(x), times = sapply(x, length)))
  
  if (is.numeric(y)) {
    # For numeric variables, perform a standard 2-sample t-test
    p <- t.test(y ~ g)$p.value
  } else {
    # For categorical variables, perform a chi-squared test of independence
    p <- chisq.test(table(y, g))$p.value
  }
  
  # Format the p-value (HTML-safe for <)
  formatted_p <- sub("<", "&lt;", format.pval(p, digits = 3, eps = 0.001))
  
  # Two blank strings before the value moves it down two rows
  c("", "", formatted_p)
}

rndr <- function(x, name, ...) {
  if (!is.numeric(x)) return(render.categorical.default(x))
  
  # Choose rendering style based on variable name
  if (name == "outbreak_count") {
    what <- "Sum (Max)"
  } else if (name %in% c(
    "temperature_C", "shortwave_radiation", "longwave_radiation",
    "relative_humidity", "wind_speed", "soil_moisture",
    "runoff_mm", "precipitation_mm", "snow_density",
    "leaf_area_index", "anseriformes", "predators",
  "au_count",
  "birds_au"   
  )) {
    what <- "Mean (SD) <br/>Median [Min, Max]"
  } else {
    what <- "Mean (SD)"
  }
  
  parse.abbrev.render.code(c("",what))(x)
}


# ============================================================
# Variable labels

label(fishnet_t1$outbreak_count)        <- "Spillover Count"
label(fishnet_t1$temperature_C)        <- "Temperature (°C)"
label(fishnet_t1$shortwave_radiation)  <- "Shortwave Radiation (MJ/m²)"
label(fishnet_t1$longwave_radiation)   <- "Longwave Radiation (MJ/m²)"
label(fishnet_t1$relative_humidity)    <- "Relative Humidity (%)"
label(fishnet_t1$wind_speed)           <- "Wind Speed (m/s)"
label(fishnet_t1$soil_moisture)        <- "Soil Moisture (m³/m³)"
label(fishnet_t1$runoff_mm)            <- "Surface & Sub-Surface Runoff (mm)"
label(fishnet_t1$precipitation_mm)     <- "Precipitation (mm)"
label(fishnet_t1$snow_density)         <- "Snow Density (kg/m³)"
label(fishnet_t1$leaf_area_index)      <- "Leaf Area Index"
label(fishnet_t1$anseriformes)         <- "Anseriformes Abundance (%)"
label(fishnet_t1$predators)            <- "Predator Abundance (%)"
label(fishnet_t1$au_count)              <- "Animal Unit Count"
label(fishnet_t1$birds_au)            <- "Bird Unit Count"
label(fishnet_t1$land_cover_label)     <- "Land Cover Type N (%)"

# Generate Table 1 

table_1 <- table1::table1(
  ~ outbreak_count + temperature_C + shortwave_radiation + longwave_radiation +
    relative_humidity + wind_speed +
    soil_moisture + runoff_mm + precipitation_mm + snow_density + leaf_area_index +
    anseriformes + predators + au_count +  birds_au + land_cover_label
  | any_outbreak,
  data = fishnet_t1,
  render=rndr,
  render.categorical = my.render.cat,
  render.missing = NULL,
  overall = FALSE,
  extra.col = list(`P-value` = pvalue)
)

table_1
```



# Table 1 but stratified by season


```{r, table 1 stratified by season}
# 
# # Subset to outbreak cells only
# fishnet_outbreak <- fishnet_t1 %>%
#   filter(any_outbreak == "Cells with an Outbreak") %>%
#   mutate(season = factor(droplevels(season)))
# 
# # Generate Table 1 stratified by season
# table_season <- table1::table1(
#   ~ outbreak_count + temperature_C + shortwave_radiation + longwave_radiation +
#     relative_humidity + wind_speed +
#     soil_moisture + runoff_mm + precipitation_mm + snow_density +
#     leaf_area_index + anseriformes + predators
#   | season,                    # stratify by season
#   data = fishnet_outbreak,
#   render = rndr,
#   render.categorical = my.render.cat,
#   render.missing = NULL,
#   overall = FALSE)
# 
# table_season
# 
# 



```




# Table 1 for CLimate Anomalies 


```{r, table 1 for climate anomalies}

# ============================================================
# 25-year Z-score anomaly variables
# ============================================================

anomaly_vars <- c(
  "leaf_area_index_z_25yr",
  "longwave_radiation_z_25yr",
  "precipitation_z_25yr",
  "relative_humidity_z_25yr",
  "runoff_z_25yr",
  "shortwave_radiation_z_25yr",
  "snow_density_z_25yr",
  "soil_moisture_z_25yr",
  "temperature_z_25yr",
  "wind_speed_z_25yr"
)

# ============================================================
# Variable labels (for clean table output)
# ============================================================

label(fishnet_t1$leaf_area_index_z_25yr)     <- "Leaf Area Index (Z-score)"
label(fishnet_t1$longwave_radiation_z_25yr)  <- "Longwave Radiation (Z-score)"
label(fishnet_t1$precipitation_z_25yr)       <- "Precipitation (Z-score)"
label(fishnet_t1$relative_humidity_z_25yr)   <- "Relative Humidity (Z-score)"
label(fishnet_t1$runoff_z_25yr)              <- "Surface & Sub-Surface Runoff (Z-score)"
label(fishnet_t1$shortwave_radiation_z_25yr) <- "Shortwave Radiation (Z-score)"
label(fishnet_t1$snow_density_z_25yr)        <- "Snow Density (Z-score)"
label(fishnet_t1$soil_moisture_z_25yr)       <- "Soil Moisture (Z-score)"
label(fishnet_t1$temperature_z_25yr)         <- "Temperature (Z-score)"
label(fishnet_t1$wind_speed_z_25yr)          <- "Wind Speed (Z-score)"


# ============================================================
# Renderer for Z-score anomalies
# ============================================================

rndr_anom <- function(x, name, ...) {
  if (!is.numeric(x)) return(render.categorical.default(x))
  
  # For anomalies, only show Mean (SD)
  what <- "Mean (SD)"
  parse.abbrev.render.code(c("", what))(x)
}



# Generate anomaly table

table_anom <- table1::table1(
  ~ outbreak_count + temperature_z_25yr + shortwave_radiation_z_25yr + longwave_radiation_z_25yr +
    relative_humidity_z_25yr + wind_speed_z_25yr +
    soil_moisture_z_25yr + runoff_z_25yr + precipitation_z_25yr + snow_density_z_25yr +
    leaf_area_index_z_25yr | any_outbreak,
  data = fishnet_t1,
  render = rndr_anom,
  render.categorical = my.render.cat,
  render.missing = NULL,
  overall = FALSE,
  extra.col = list(`P-value` = pvalue)
)

table_anom




```

## Primary Table 1 Export

```{r, primary table 1 export}


# Read spillover summary computed in create_timeseries_df.Rmd
spillover_summary <- readRDS(file.path(objects_folder, "spillover_summary.RDS"))

total_spillovers <- spillover_summary$total
n_flag_200m      <- spillover_summary$non_independent_200m
n_flag_500m      <- spillover_summary$non_independent_500m
n_flag_1km       <- spillover_summary$non_independent_1km
n_independent    <- total_spillovers - n_flag_500m

# Relevel so outbreak cells come first
fishnet_t1 <- fishnet_t1 %>%
  mutate(
    any_outbreak = fct_relevel(any_outbreak, "Cells with an outbreak"),
    land_cover_label = forcats::fct_lump_prop(land_cover_label, prop = 0.01, other_level = "Other")
  )

# Reorder land cover by frequency in outbreak cells
outbreak_lc_order <- fishnet_t1 %>%
  filter(any_outbreak == "Cells with an outbreak") %>%
  count(land_cover_label, sort = TRUE) %>%
  pull(land_cover_label)

fishnet_t1 <- fishnet_t1 %>%
  mutate(land_cover_label = factor(land_cover_label, levels = c(as.character(outbreak_lc_order), 
                                                                 setdiff(levels(land_cover_label), outbreak_lc_order))))

# Table 1: Outbreak vs No Outbreak
table_1_gt <- fishnet_t1 %>%
  select(
    feedlot_count, au_count, birds_au,
    temperature_C, shortwave_radiation, longwave_radiation,
    relative_humidity, wind_speed, soil_moisture,
    runoff_mm, precipitation_mm, snow_density,
    leaf_area_index, anseriformes, predators,
    land_cover_label, any_outbreak
  ) %>%
  tbl_summary(
    by = any_outbreak,
    type = list(
      all_continuous() ~ "continuous2",
      land_cover_label ~ "categorical"
    ),
    statistic = list(
      all_continuous() ~ "{mean} [{min}, {max}]",
      all_categorical() ~ "{n} ({p}%)"
    ),
    digits = list(
      all_continuous() ~ c(mean = 1, min = 1, max = 1),
      c(feedlot_count, au_count, birds_au) ~ c(mean = 0, min = 0, max = 0)
    ),
    label = list(
      feedlot_count ~ "Feedlot Count (N)",
      au_count ~ "Total Animal Units (AU)",
      birds_au ~ "Poultry Animal Units (AU)",
      temperature_C ~ "Temperature (°C)",
      shortwave_radiation ~ "Shortwave Radiation (MJ/m²)",
      longwave_radiation ~ "Longwave Radiation (MJ/m²)",
      relative_humidity ~ "Relative Humidity (%)",
      wind_speed ~ "Wind Speed (m/s)",
      soil_moisture ~ "Soil Moisture (m³/m³)",
      runoff_mm ~ "Surface & Sub-Surface Runoff (mm)",
      precipitation_mm ~ "Precipitation (mm)",
      snow_density ~ "Snow Density (kg/m³)",
      leaf_area_index ~ "Leaf Area Index",
      anseriformes ~ "Anseriformes Relative Abundance",
      predators ~ "Predator Relative Abundance",
      land_cover_label ~ "Land Cover Type (%)"
    ),
    missing = "no"
  ) %>%
  add_p(
    test = list(
      all_continuous() ~ "t.test",
      all_categorical() ~ "chisq.test"
    )
  ) %>%
  # Add spillover rows at the top
  modify_table_body(~ .x %>%
    add_row(
      variable = "spillover_events",
      var_type = "categorical",
      var_label = "Spillover Events",
      row_type = "label",
      label = "Spillover Events",
      stat_1 = "", stat_2 = "",
      .before = 1
    ) %>%
    add_row(
      variable = "spillover_events",
      var_type = "categorical",
      row_type = "level",
      label = "    Total",
      stat_1 = paste0(as.character(total_spillovers), " (102 cell-weeks)"),
      stat_2 = "—",
      .after = 1
    ) %>%
    add_row(
      variable = "spillover_events",
      var_type = "categorical",
      row_type = "level",
      label = "    Excluding within 200m & 4 weeks of prior event",
      stat_1 = paste0(as.character(total_spillovers - n_flag_200m), " (101 cell-weeks)"),
      stat_2 = "—",
      .after = 2
    ) %>%
    add_row(
      variable = "spillover_events",
      var_type = "categorical",
      row_type = "level",
      label = "    Excluding within 500m & 4 weeks of prior event",
      stat_1 = paste0(as.character(total_spillovers - n_flag_500m), " (101 cell-weeks)"),
      stat_2 = "—",
      .after = 3
    ) %>%
    add_row(
      variable = "spillover_events",
      var_type = "categorical",
      row_type = "level",
      label = "    Excluding within 1km & 4 weeks of prior event",
      stat_1 = paste0(as.character(total_spillovers - n_flag_1km), " (95 cell-weeks)"),
      stat_2 = "—",
      .after = 4
    )
  ) %>%
  modify_header(label ~ "**Variable**") %>%
  modify_caption(" ") %>%
  bold_labels()

# Convert to flextable
ft <- as_flex_table(table_1_gt) %>%
  font(fontname = "Calibri", part = "all") %>%
  fontsize(size = 11, part = "all") %>%
  align(align = "left", j = 1, part = "all") %>%
  align(align = "center", j = 2:ncol(as_flex_table(table_1_gt)$body$dataset), part = "all") %>%
  set_table_properties(layout = "autofit", width = .8) %>%
  set_table_properties(layout = "fixed") %>%
  width(width = 0.9) %>%
  line_spacing(space = 0.8, part = "all") %>%
  autofit() %>%
  add_footer_lines("AU = animal unit (1,000 lbs of live animal weight). Source: Minnesota Pollution Control Agency feedlot registry.")

ft

save_as_docx(ft, path = file.path(tables_folder, "Table1_Primary.docx"))

```

## Season Stratification Table 1 Export

```{r, season strat table 1 export}

# Compute season spillover counts from the modeling dataframe
fishnet_timeseries_modeling <- readRDS(file.path(objects_folder, "fishnet_timeseries_modeling.RDS"))
fishnet_timeseries_raw <- readRDS(file.path(objects_folder, "fishnet_timeseries.RDS"))

# Independent events per season (from modeling df)
season_indep_df <- fishnet_timeseries_modeling %>%
  filter(outbreak_binary == 1) %>%
  group_by(season) %>%
  summarise(
    independent_events = sum(outbreak_count, na.rm = TRUE),
    .groups = "drop"
  )

# Total events per season (from raw df)
season_total_df <- fishnet_timeseries_raw %>%
  filter(outbreak_binary == 1) %>%
  group_by(season) %>%
  summarise(
    total_events = sum(outbreak_count, na.rm = TRUE),
    .groups = "drop"
  )

# Combine
season_spill <- season_indep_df %>%
  left_join(season_total_df, by = "season")

# Subset: Only outbreak cells
fishnet_outbreak <- fishnet_t1 %>%
  filter(any_outbreak == "Cells with an outbreak") %>%
  mutate(season = droplevels(factor(season)))

# Get season levels in order
season_levels <- levels(fishnet_outbreak$season)

# Build season stat values
season_total_vec <- setNames(as.character(season_spill$total_events), season_spill$season)
season_indep_vec <- setNames(as.character(season_spill$independent_events), season_spill$season)

# Table 2: Outbreak Cells Stratified by Season
table_season_gt <- fishnet_outbreak %>%
  select(
    temperature_C, shortwave_radiation, longwave_radiation,
    relative_humidity, wind_speed, soil_moisture,
    runoff_mm, precipitation_mm, snow_density,
    leaf_area_index, anseriformes, predators,
    season
  ) %>%
  tbl_summary(
    by = season,
    type = list(
      all_continuous() ~ "continuous2"
    ),
    statistic = list(
      all_continuous() ~ "{mean} [{min}, {max}]",
      all_categorical() ~ "{n} ({p}%)"
    ),
    digits = list(
      all_continuous() ~ c(mean = 1, min = 1, max = 1),
      c(soil_moisture, leaf_area_index, runoff_mm) ~ c(mean = 2, min = 2, max = 2),
      c(anseriformes, predators) ~ c(mean = 1, min = 1, max = 1)    ),
    label = list(
      temperature_C ~ "Temperature (\u00B0C)",
      shortwave_radiation ~ "Shortwave Radiation (MJ/m\u00B2)",
      longwave_radiation ~ "Longwave Radiation (MJ/m\u00B2)",
      relative_humidity ~ "Relative Humidity (%)",
      wind_speed ~ "Wind Speed (m/s)",
      soil_moisture ~ "Soil Moisture (m\u00B3/m\u00B3)",
      runoff_mm ~ "Surface & Sub-Surface Runoff (mm)",
      precipitation_mm ~ "Precipitation (mm)",
      snow_density ~ "Snow Density (kg/m\u00B3)",
      leaf_area_index ~ "Leaf Area Index",
      anseriformes ~ "Anseriformes Relative Abundance",
      predators ~ "Predator Relative Abundance"
    ),
    missing = "no"
  ) %>%
  modify_table_body(~ {
    stat_cols <- grep("^stat_", names(.x), value = TRUE)
    
    total_row <- tibble(
      variable = "spillover_events", var_type = "categorical",
      var_label = "Spillover Events", row_type = "label",
      label = "Spillover Events"
    )
    events_row <- tibble(
      variable = "spillover_events", var_type = "categorical",
      row_type = "level", label = "    Total Events"
    )
    indep_row <- tibble(
      variable = "spillover_events", var_type = "categorical",
      row_type = "level", label = "    Independent Events (>500m or >4 weeks)"
    )
    
    for (i in seq_along(stat_cols)) {
      s <- season_levels[i]
      total_row[[stat_cols[i]]] <- ""
      events_row[[stat_cols[i]]] <- ifelse(s %in% names(season_total_vec), season_total_vec[s], "0")
      indep_row[[stat_cols[i]]] <- ifelse(s %in% names(season_indep_vec), season_indep_vec[s], "0")
    }
    
    bind_rows(total_row, events_row, indep_row, .x)
  }) %>%
  modify_header(label ~ "**Variable**") %>%
  modify_caption(" ") %>%
  bold_labels()

# Convert to flextable
ft_season <- as_flex_table(table_season_gt) %>%
  font(fontname = "Calibri", part = "all") %>%
  fontsize(size = 10, part = "all") %>%
  align(align = "left", j = 1, part = "all") %>%
  align(align = "center", j = 2:ncol(as_flex_table(table_season_gt)$body$dataset), part = "all") %>%
  set_table_properties(layout = "autofit", width = .8) %>%
  set_table_properties(layout = "fixed") %>%
  width(width = 0.9) %>%
  line_spacing(space = 0.8, part = "all") %>%
  autofit()

save_as_docx(ft_season, path = file.path(tables_folder, "Table1_Season.docx"))
```


# Season stratifications compared to 25-year mean
```{r, season stratifications comped to 25-year mean}

# Transform the 25-yr mean columns to match current-variable units
fishnet_outbreak <- fishnet_outbreak %>%
  mutate(
    temperature_mean_25yr_C            = temperature_mean_25yr - 273.15,
    precipitation_mean_25yr_mm         = precipitation_mean_25yr * 1000,
    runoff_mean_25yr_mm                = runoff_mean_25yr * 1000,
    longwave_radiation_mean_25yr_MJm2  = longwave_radiation_mean_25yr / 1e6,
    shortwave_radiation_mean_25yr_MJm2 = shortwave_radiation_mean_25yr / 1e6
  )

# Map current vars -> transformed 25-yr counterparts
compare_map <- c(
  temperature_C        = "temperature_mean_25yr_C",
  shortwave_radiation  = "shortwave_radiation_mean_25yr_MJm2",
  longwave_radiation   = "longwave_radiation_mean_25yr_MJm2",
  relative_humidity    = "relative_humidity_mean_25yr",
  wind_speed           = "wind_speed_mean_25yr",
  soil_moisture        = "soil_moisture_mean_25yr",
  runoff_mm            = "runoff_mean_25yr_mm",
  precipitation_mm     = "precipitation_mean_25yr_mm",
  snow_density         = "snow_density_mean_25yr",
  leaf_area_index      = "leaf_area_index_mean_25yr"
)

var_labels <- c(
  temperature_C       = "Temperature (\u00B0C)",
  shortwave_radiation = "Shortwave Radiation (MJ/m\u00B2)",
  longwave_radiation  = "Longwave Radiation (MJ/m\u00B2)",
  relative_humidity   = "Relative Humidity (%)",
  wind_speed          = "Wind Speed (m/s)",
  soil_moisture       = "Soil Moisture (m\u00B3/m\u00B3)",
  runoff_mm           = "Surface & Sub-Surface Runoff (mm)",
  precipitation_mm    = "Precipitation (mm)",
  snow_density        = "Snow Density (kg/m\u00B3)",
  leaf_area_index     = "Leaf Area Index"
)

compare_df <- lapply(names(compare_map), function(cur_var) {
  ref_var <- compare_map[[cur_var]]
  fishnet_outbreak %>%
    group_by(season) %>%
    summarise(
      current_mean = mean(.data[[cur_var]], na.rm = TRUE),
      ref25_mean   = mean(.data[[ref_var]], na.rm = TRUE),
      .groups = "drop"
    ) %>%
    mutate(
      variable   = cur_var,
      delta_mean = current_mean - ref25_mean
    )
}) %>%
  bind_rows() %>%
  mutate(
    Variable = var_labels[variable],
    Current_Mean = sprintf("%.1f", current_mean),
    Mean_25yr    = sprintf("%.1f", ref25_mean),
    Delta        = sprintf("%.1f", delta_mean)
  ) %>%
  select(Variable, season, Current_Mean, Mean_25yr, Delta) %>%
  arrange(match(Variable, unname(var_labels[names(compare_map)])), season)

compare_wide <- compare_df %>%
  pivot_longer(cols = c(Current_Mean, Mean_25yr, Delta),
               names_to = "stat", values_to = "value") %>%
  unite("Season_Stat", season, stat, sep = " \u2014 ") %>%
  pivot_wider(names_from = Season_Stat, values_from = value) %>%
  arrange(match(Variable, unname(var_labels[names(compare_map)])))

ft_compare_wide <- flextable(compare_wide) %>%
  font(fontname = "Calibri", part = "all") %>%
  fontsize(size = 10, part = "all") %>%
  align(align = "center", part = "all") %>%
  align(j = "Variable", align = "left", part = "all") %>%
  autofit()

save_as_docx(
  "Table 2b. Outbreak Cells: Current vs 25-yr Means by Season" = ft_compare_wide,
  path = file.path(tables_folder, "Table2b_Season_vs_25yr.docx")
)
```

# Season stratification with model comparing absolute and 25yr mean

## Prepare model
```{r, prepare for modeling difference between absolute and 25yr mean}

fishnet_outbreak <- fishnet_outbreak %>%
  mutate(
    temperature_mean_25yr_C            = temperature_mean_25yr - 273.15,
    precipitation_mean_25yr_mm         = precipitation_mean_25yr * 1000,
    runoff_mean_25yr_mm                = runoff_mean_25yr        * 1000,
    longwave_radiation_mean_25yr_MJm2  = longwave_radiation_mean_25yr  / 1e6,
    shortwave_radiation_mean_25yr_MJm2 = shortwave_radiation_mean_25yr / 1e6
  )

vars_abs <- c(
  "temperature_C", "shortwave_radiation", "longwave_radiation",
  "relative_humidity", "wind_speed", "soil_moisture",
  "runoff_mm", "precipitation_mm", "snow_density", "leaf_area_index"
)

means_map <- c(
  temperature_C       = "temperature_mean_25yr_C",
  shortwave_radiation = "shortwave_radiation_mean_25yr_MJm2",
  longwave_radiation  = "longwave_radiation_mean_25yr_MJm2",
  relative_humidity   = "relative_humidity_mean_25yr",
  wind_speed          = "wind_speed_mean_25yr",
  soil_moisture       = "soil_moisture_mean_25yr",
  runoff_mm           = "runoff_mean_25yr_mm",
  precipitation_mm    = "precipitation_mean_25yr_mm",
  snow_density        = "snow_density_mean_25yr",
  leaf_area_index     = "leaf_area_index_mean_25yr"
)

missing_means <- setdiff(unname(means_map), names(fishnet_outbreak))
if (length(missing_means)) {
  stop("Missing 25yr mean columns after conversion: ",
       paste(missing_means, collapse = ", "))
}
```

## Now model...
```{r, modeling difference between absolute and 25yr mean}

emm_options(lmer.df = "asymptotic")

fit_anomaly_once <- function(dat, abs_name, mean_name) {
  stopifnot("week" %in% names(dat))
  dd <- dat %>%
    transmute(
      zone_id, season, week,
      .abs    = .data[[abs_name]],
      .mean25 = .data[[mean_name]],
      .anom   = .abs - .mean25
    )

  fit <- lmerTest::lmer(
    .anom ~ season + (1|zone_id) + (1|week),
    data = dd, REML = TRUE,
    control = lme4::lmerControl(check.conv.singular = "ignore")
  )

  emm <- emmeans::emmeans(fit, ~ season)
  broom::tidy(emm, conf.int = TRUE) %>%
    transmute(
      variable = abs_name,
      season   = as.character(season),
      estimate, conf.low, conf.high, p.value
    )
}

emm_all <- purrr::map_dfr(
  vars_abs,
  ~fit_anomaly_once(fishnet_outbreak, .x, means_map[[.x]])
)

emm_all <- emm_all %>%
  group_by(season) %>%
  mutate(p.adj = p.adjust(p.value, method = "BH")) %>%
  ungroup() %>%
  mutate(
    is_sig = p.adj < 0.05,
    delta_ci = sprintf("%+.2f [%.2f, %.2f]", estimate, conf.low, conf.high)
  )

delta_wide <- emm_all %>%
  select(variable, season, delta_ci, is_sig) %>%
  pivot_wider(names_from = season, values_from = c(delta_ci, is_sig)) %>%
  mutate(Row = "Mean difference from 25yr climatology (95% CI)") %>%
  relocate(Row)
```

## Produce results
```{r, produce results from model on difference between absolute and 25yr mean}

label_map <- c(
  temperature_C       = "Temperature (\u00B0C)",
  shortwave_radiation = "Shortwave Radiation (MJ/m\u00B2)",
  longwave_radiation  = "Longwave Radiation (MJ/m\u00B2)",
  relative_humidity   = "Relative Humidity (%)",
  wind_speed          = "Wind Speed (m/s)",
  soil_moisture       = "Soil Moisture (m\u00B3/m\u00B3)",
  runoff_mm           = "Surface & Sub-Surface Runoff (mm)",
  precipitation_mm    = "Precipitation (mm)",
  snow_density        = "Snow Density (kg/m\u00B3)",
  leaf_area_index     = "Leaf Area Index"
)

means_long <- purrr::map_dfr(vars_abs, function(v) {
  fishnet_outbreak %>%
    transmute(
      variable = v,
      season   = as.character(season),
      value    = .data[[v]]
    ) %>%
    group_by(variable, season) %>%
    summarise(mean_2022 = mean(value, na.rm = TRUE), .groups = "drop")
})

means_wide <- means_long %>%
  mutate(mean_fmt = sprintf("%.1f", mean_2022)) %>%
  select(variable, season, mean_fmt) %>%
  pivot_wider(names_from = season, values_from = mean_fmt) %>%
  mutate(Row = "Mean [Min, Max]") %>%
  relocate(Row)

# Spillover rows
spillover_total_row <- tibble(
  variable = "spillover", Row = "Spillover Event Count",
  Winter = season_total_vec["Winter"], Spring = season_total_vec["Spring"],
  Summer = season_total_vec["Summer"], Fall = season_total_vec["Fall"]
)

spillover_indep_row <- tibble(
  variable = "spillover", Row = "Independent Events (>500m or >4 weeks)",
  Winter = season_indep_vec["Winter"], Spring = season_indep_vec["Spring"],
  Summer = season_indep_vec["Summer"], Fall = season_indep_vec["Fall"]
)

# Extract delta text columns only (drop is_sig columns for the table text)
delta_text <- delta_wide %>%
  select(variable, Row, 
         Winter = delta_ci_Winter, Spring = delta_ci_Spring, 
         Summer = delta_ci_Summer, Fall = delta_ci_Fall)

# Build the display table
tbl_mat <- bind_rows(
  spillover_total_row, spillover_indep_row,
  means_wide, delta_text
) %>%
  arrange(
    factor(variable, levels = c("spillover", vars_abs)),
    match(Row, c("Spillover Event Count", "Independent Events (>500m or >4 weeks)",
                 "Mean [Min, Max]", "Mean difference from 25yr climatology (95% CI)"))
  ) %>%
  mutate(Variable = case_when(
    variable == "spillover" & Row == "Spillover Event Count" ~ "Spillover Events",
    variable == "spillover" ~ "",
    variable %in% names(label_map) ~ label_map[variable],
    TRUE ~ variable
  )) %>%
  select(Variable, Row, Winter, Spring, Summer, Fall)

# Build significance index matrix for bolding
# For each delta row, check if the corresponding is_sig is TRUE
sig_rows <- delta_wide %>%
  arrange(factor(variable, levels = vars_abs))

# Find row indices in tbl_mat that are delta rows
delta_row_indices <- which(tbl_mat$Row == "Mean difference from 25yr climatology (95% CI)")

ft_tbl <- flextable(tbl_mat) |>
  set_header_labels(
    Variable = "Variable",
    Row = "",
    Winter = "Winter",
    Spring = "Spring",
    Summer = "Summer",
    Fall = "Fall"
  ) |>
  add_footer_row(
    values = "Mean differences [95% CIs] between 2022 conditions and 25-year climatological means (1997-2021) were computed from linear mixed-effects models with a fixed effect for season, random intercepts for cell and week, estimated via REML. BH-adjusted significant differences (p < 0.05) are shown in bold. Seasons follow meteorological definitions: winter (Dec-Feb), spring (Mar-May), summer (Jun-Aug), fall (Sep-Nov).",
    colwidths = 6
  ) |>
  bold(i = ~ Row %in% c("Spillover Event Count", "Independent Events (>500m or >4 weeks)", "Mean [Min, Max]"), bold = TRUE, part = "body") |>
  color(i = ~ Row == "Mean difference from 25yr climatology (95% CI)", color = "#555555", part = "body") |>
  align(j = 2:6, align = "center", part = "all") |>
  align(j = 1, align = "left", part = "all") |>
  font(fontname = "Calibri", part = "all") |>
  fontsize(size = 10, part = "all") |>
  line_spacing(space = 0.8, part = "all") |>
  autofit()

# Bold significant cells individually
for (k in seq_along(delta_row_indices)) {
  row_idx <- delta_row_indices[k]
  var_name <- vars_abs[k]
  
  # Get significance flags for this variable
  sig_row <- sig_rows %>% filter(variable == var_name)
  
  if (nrow(sig_row) == 1) {
    if (isTRUE(sig_row$is_sig_Winter))  ft_tbl <- bold(ft_tbl, i = row_idx, j = "Winter", bold = TRUE)
    if (isTRUE(sig_row$is_sig_Spring))  ft_tbl <- bold(ft_tbl, i = row_idx, j = "Spring", bold = TRUE)
    if (isTRUE(sig_row$is_sig_Summer))  ft_tbl <- bold(ft_tbl, i = row_idx, j = "Summer", bold = TRUE)
    if (isTRUE(sig_row$is_sig_Fall))    ft_tbl <- bold(ft_tbl, i = row_idx, j = "Fall", bold = TRUE)
  }
}

ft_tbl

save_as_docx(ft_tbl, path = file.path(tables_folder, "Table_Season_w25yr_comp.docx"))
```


# Quick Statistics 

## How many case and control cells?

```{r, quick statistics}

vars <- c(
  "leaf_area_index","longwave_radiation","precipitation","relative_humidity",
  "runoff","shortwave_radiation","snow_density","soil_moisture","temperature",
  "anseriformes","predators","wind_speed",
  "any_outbreak",          # binary (0/1)
  "outbreak_count"         # numeric count per row
)

zone_summary <- fishnet_subset %>%
  group_by(zone_id) %>%
  summarise(
    # Means & SDs for all listed variables
    across(all_of(vars),
           list(mean = ~mean(.x, na.rm = TRUE),
                sd   = ~sd(.x,   na.rm = TRUE)),
           .names = "{.col}_{.fn}"),

    # Zone-level outbreak summaries
    any_outbreak_any    = as.integer(any(any_outbreak == 1, na.rm = TRUE)), # ever had 1?
    outbreak_count_total = sum(outbreak_count, na.rm = TRUE),               # total across time

    n_obs = n(),
    .groups = "drop"
  )

zone_summary

zone_outbreak_counts <- zone_summary %>%
  count(any_outbreak_any, name = "n_zone_ids") %>%
  mutate(any_outbreak_any = ifelse(any_outbreak_any == 1, "has_outbreak", "no_outbreak"))

zone_outbreak_counts

## 76 have outbreak and 2,070 do not

# Overall
mean(fishnet_t1$snow_density == 100, na.rm = TRUE)

# By season (or whatever grouping you're using)
fishnet_t1 %>%
  group_by(season) %>%
  summarise(
    n = n(),
    pct_100 = mean(snow_density == 100, na.rm = TRUE),
    min = min(snow_density, na.rm = TRUE),
    max = max(snow_density, na.rm = TRUE)
  )
```
## Other checks

```{r, other checks}


# counts at each object
cat("fishnet_timeseries:", nrow(fishnet_timeseries), "\n")
cat("fishnet_timeseries_clean:", nrow(fishnet_timeseries_clean), "\n")
cat("fishnet_subset:", nrow(fishnet_subset), "\n")
cat("fishnet_t1:", nrow(fishnet_t1), "\n")

# confirm what the 'unit' is
cat("unique zone_id:", dplyr::n_distinct(fishnet_t1$zone_id), "\n")
cat("unique zone_id-week:", dplyr::n_distinct(paste(fishnet_t1$zone_id, fishnet_t1$week)), "\n")

# full season distribution INCLUDING missing
table(fishnet_t1$season, useNA = "ifany")

# how many season are NA?
sum(is.na(fishnet_t1$season))

# if you expect only 2 seasons, see what the "other" values are
sort(unique(fishnet_t1$season))

vars_in_table1 <- c(
  "outbreak_count", "temperature_C", "shortwave_radiation", "longwave_radiation",
  "relative_humidity", "wind_speed", "soil_moisture", "runoff_mm", "precipitation_mm",
  "snow_density", "leaf_area_index", "anseriformes", "predators", "au_count", "birds_au",
  "land_cover_label", "any_outbreak", "season", "zone_id", "week"
)

cc_n <- fishnet_t1 |>
  dplyr::select(dplyr::all_of(vars_in_table1)) |>
  stats::complete.cases() |>
  sum()

cat("Complete-case rows for table vars:", cc_n, "\n")

na_counts <- sapply(fishnet_t1[, vars_in_table1], function(x) sum(is.na(x)))
sort(na_counts, decreasing = TRUE)
```