---
title: "Join Data Sources"
author: "Fintan Mooney"
date: "2026-02-09"
output: html_document
---

# NOTE: This script requires confidential HPAI spillover point-location data
# (not publicly available). The processed grid-cell level output is provided
# in data/ and downstream analyses can be run without this file.

# Joins USDA spillover event data to 10x10km grid cells covering Minnesota for 2022!
# Also brings in eBird abundance, MPCA feedlot data, USGS landcover, and ERA5 met data...
# This produces the base dataframe used to build the timeseries and case-crossover datasets.

```{r include=FALSE}
rm(list=ls())
project.folder = paste0(print(here::here()),'/')
source(paste0(project.folder,'create_folder_structure.R'))
source(paste0(functions.folder,'script_initiate.R'))
```

## Load in Weekly Fishnet 


```{r load weekly fishnet}

weekly_fishnet <- st_read(paste0(spatial_data, "/fishnet_10km_timeseries.gpkg"))

ggplot() +
  geom_sf(data = weekly_fishnet,fill = NA, color = "blue", linewidth = 0.2)
```

## Load Spillover Data

### Filter spillover data

```{r, spillover data}
spillover_file <- file.path(spillover_data, "hpai-outbreaks-clean-placeholder-032124.csv")
if (!file.exists(spillover_file)) {
  stop("This script needs the confidential spillover data (hpai-outbreaks-clean-placeholder-032124.csv). See data/confidential/README.md for how to get it.")
}
spillover_data <- read_csv(spillover_file)

start_date <- as.Date("2022-01-01")
end_date <- as.Date("2023-01-01")

spillover_data_mn <- spillover_data %>% filter(state == "MN") %>%
                     filter(dplyr::between(date_confirmed, start_date, end_date))
                      
# Create week column in spillover_data_mn
spillover_data_mn <- spillover_data_mn %>%
  mutate(week = lubridate::week(date_confirmed))

# Convert to sf then transform to match fishnet
spillover_sf <- spillover_data_mn %>%
  st_as_sf(coords = c("lon", "lat"), crs = 4326) %>%
  st_transform(crs = st_crs(weekly_fishnet))

```
### Join spillover data into weekly_fishnet

```{r, bring spillover data in}

# Spatial join to get zone_id for each outbreak point
outbreaks_in_cells <- st_join(spillover_sf, weekly_fishnet %>% distinct(zone_id, geom), join = st_within)

# Aggregate by zone_id and week 
outbreak_summary <- outbreaks_in_cells %>%
  st_drop_geometry() %>%
  group_by(zone_id, week) %>%
  summarise(
    outbreak_count = n(),
    outbreak_id = paste(outbreak_id, collapse = ", "),
    .groups = "drop"
  ) %>%
  mutate(outbreak_binary = 1L)

outbreak_weeks_by_zone <- outbreak_summary %>%
  group_by(zone_id) %>%
  summarise(outbreak_week = paste(sort(unique(week)), collapse = ", "), .groups = "drop")

any_outbreak_zones <- outbreak_summary %>%
  distinct(zone_id) %>%
  mutate(any_outbreak = 1L)

# Join outbreak info to weekly_fishnet
weekly_fishnet <- weekly_fishnet %>%
  left_join(outbreak_summary, by = c("zone_id", "week")) %>%
  left_join(outbreak_weeks_by_zone, by = "zone_id") %>%
  left_join(any_outbreak_zones, by = "zone_id") %>%
  mutate(
    outbreak_count = replace_na(outbreak_count, 0),
    outbreak_binary = replace_na(outbreak_binary, 0L),
    outbreak_id = replace_na(outbreak_id, NA_character_),
    outbreak_week = replace_na(outbreak_week, NA_character_),
    any_outbreak = replace_na(any_outbreak, 0L)
  )


```



## Load bird abundance data for outbreaks
```{r, load bird abundance data}
bird_data <- read_csv(file.path(bird_abundance_data,"mn-outbreak-weekly-bird-estimate-2022-012926.csv"))
print(unique(bird_data$species_name))
print(colnames(bird_data))

# Pivot bird data to wide format
bird_wide <- bird_data %>%
  pivot_wider(
    id_cols = c(outbreak_id, mmwr_week, week, week_outbreak),
    names_from = species_name,
    values_from = c(abd_avg, abd_sum),
    names_glue = "{.value}_{species_name}"
  )
print(colnames(bird_wide))

# Define species groups
anseriformes <- c(
  "mallar3", "cangoo", "wooduc", "rinduc", "comgol", "norsho",
  "buffle", "lessca", "rebmer", "gnwtea", "gadwal", "ambduc",
  "rudduc", "rethaw", "truswa", "commer"
)
predators <- c(
  "buwtea", "grhowl", "amwpel", "baleag", "brdowl"
)

# abd_avg (mean relative abundance across pixels) summed across species within each group
bird_wide <- bird_wide %>%
  mutate(
    week_date = week,
    week = mmwr_week,
    total_bird_abundance = rowSums(across(starts_with("abd_avg_") & !matches("anseriformes|predators")), na.rm = TRUE),
    anseriformes = rowSums(across(all_of(paste0("abd_avg_", anseriformes))), na.rm = TRUE),
    predators = rowSums(across(all_of(paste0("abd_avg_", predators))), na.rm = TRUE)
  ) %>%
  select(outbreak_id, week_date, week, week_outbreak, total_bird_abundance, anseriformes, predators)


# Check

bird_data %>%
  filter(species_name == "mallar3") %>%
  summarise(min_avg = min(abd_avg, na.rm=TRUE), max_avg = max(abd_avg, na.rm=TRUE), mean_avg = mean(abd_avg, na.rm=TRUE),
            min_sum = min(abd_sum, na.rm=TRUE), max_sum = max(abd_sum, na.rm=TRUE), mean_sum = mean(abd_sum, na.rm=TRUE))
```

### Merge bird data for case cells into the weekly fishnet
```{r, merge spillover data and fishnet}
# Get zone_id <-> outbreak_id mapping
zone_outbreak_map <- outbreaks_in_cells %>%
  st_drop_geometry() %>%
  select(outbreak_id, zone_id) %>%
  distinct()

# For zones with multiple outbreaks, average bird data across outbreak_ids per zone-week
bird_by_zone <- bird_wide %>%
  left_join(zone_outbreak_map, by = "outbreak_id") %>%
  group_by(zone_id, week) %>%
  summarise(
    total_bird_abundance = mean(total_bird_abundance, na.rm = TRUE),
    anseriformes = mean(anseriformes, na.rm = TRUE),
    predators = mean(predators, na.rm = TRUE),
    .groups = "drop"
  )

# Join bird data to weekly_fishnet — only for outbreak zones, all 52 weeks
weekly_fishnet <- weekly_fishnet %>%
  left_join(bird_by_zone, by = c("zone_id", "week"))
```

## Load feedlot data

```{r, load feedlot data}

feedlots <- st_read(file.path(feedlot_data, "feedlots.shp")) %>%
  st_transform(crs = st_crs(weekly_fishnet))

# Clean
feedlots_clean <- feedlots %>%
  mutate(
    cafo_flag = ifelse(!is.na(cafo_flag) & cafo_flag != "N", 1L, 0L),
    near_river = ifelse(!is.na(river_stre) & river_stre != "N", 1L, 0L)
  )

```

### Join to weekly_fishnet
```{r, join feedlot}
# Spatial join feedlots to fishnet cells
feedlots_in_cells <- st_join(feedlots_clean, weekly_fishnet %>% distinct(zone_id, geom), join = st_within)

# Aggregate by zone_id
feedlot_summary <- feedlots_in_cells %>%
  st_drop_geometry() %>%
  group_by(zone_id) %>%
  summarise(
    feedlot_count = n(),
    au_count = sum(au_count, na.rm = TRUE),
    birds_au = sum(birds_au, na.rm = TRUE),
    cafo_flag = max(cafo_flag, na.rm = TRUE),
    near_river = max(near_river, na.rm = TRUE),
    .groups = "drop"
  )

# Join to weekly_fishnet
weekly_fishnet <- weekly_fishnet %>%
  left_join(feedlot_summary, by = "zone_id") %>%
  mutate(
    feedlot_count = replace_na(feedlot_count, 0),
    au_count = replace_na(au_count, 0),
    birds_au = replace_na(birds_au, 0),
    cafo_flag = replace_na(cafo_flag, 0L),
    near_river = replace_na(near_river, 0L)
  )

```

## Load Meterological Data

```{r, load meterological data}

data2022 <- read_csv(file.path(env_data, "mn_weekly_era5land_2022.csv"))

data2022$week_start <- as.Date(as.character(data2022$week_start), format = "%Y%m%d")

# Create week index relative to 2022-01-01
data2022$week <- as.integer(floor(difftime(data2022$week_start, as.Date("2022-01-01"), units = "weeks")) + 1)
range(data2022$week)

colnames(data2022)

# Remove the lower bound for snow density of 99.99998 and make it 0 (this means no snow)
data2022 <- data2022 %>%
  mutate(snow_density = ifelse(snow_density <= 100, 0, snow_density))

```

### Join to weekly fishnet 

```{r, join to weekly fishnet }

weekly_fishnet <- weekly_fishnet %>%
  left_join(data2022, by = c("zone_id", "week"))


```


## Drop geometry to conduct joins

```{r, drop geometry}
# Save geometry for later mapping
fishnet_geom <- weekly_fishnet %>% select(zone_id) %>% distinct()

# Drop geometry before data.table operations
weekly_fishnet <- st_drop_geometry(weekly_fishnet)


```


## Load landcover

```{r, load landcover data}

land_cover_path <- file.path(spatial_data, "fishnet_landcover.gpkg")
land_cover <- st_read(land_cover_path)

land_cover_var <- c("zone_id", "land_cover_mode", "land_cover_label", "lat", "lon")

# Select only the needed columns from the land cover layer
land_cover_clean <- land_cover %>%
  select(all_of(land_cover_var))

```

### Join in landcover

```{r, join landcover}

# Perform the left join by zone_id
weekly_fishnet <- weekly_fishnet %>%
  left_join(land_cover_clean, by = "zone_id")

weekly_fishnet <- weekly_fishnet %>%
  arrange(zone_id, week)

unique(weekly_fishnet$land_cover_label)
weekly_fishnet$land_cover_label <- relevel(
  factor(weekly_fishnet$land_cover_label),
  ref = "Developed, Medium Intensity"
)

weekly_fishnet$farm_possible <- ifelse(
  weekly_fishnet$land_cover_label %in% c(
    "Cultivated Crops",
    "Pasture/Hay",
    "Emergent Herbaceous Wetlands"
  ),
  1, 0
)
```

## Load in bird abundance data for zones with feedlots but without spillover events (negative controls)
```{r, bird abundance negative controls}
# Load the data
csv_path <- file.path(bird_abundance_data, "all-mn-feedlot-species-abundance-070125.csv")
bird_abundance <- fread(csv_path)

# Define species groups
anseriformes <- c(
  "mallar3", "cangoo", "wooduc", "rinduc", "comgol", "norsho",
  "buffle", "lessca", "rebmer", "gnwtea", "gadwal", "ambduc",
  "rudduc", "rethaw", "truswa", "commer"
)
predators <- c(
  "buwtea", "grhowl", "amwpel", "baleag", "brdowl"
)

# Melt to long format
week_cols <- setdiff(names(bird_abundance), c("zone_id", "species"))
bird_long <- melt(
  bird_abundance,
  id.vars = c("zone_id", "species"),
  measure.vars = week_cols,
  variable.name = "week",
  value.name = "abundance"
)
bird_long[, group := fifelse(species %in% anseriformes, "anseriformes", "predators")]
bird_long[, week := as.integer(gsub("week_", "", week))]

# Step 1: Average abundance across feedlot points within each zone-week-species
# (parallels outbreak code which averages across outbreak points per zone)
bird_zone_avg <- bird_long[
  , .(abundance = mean(abundance, na.rm = TRUE)),
  by = .(zone_id, week, species)
]

# Re-assign groups after averaging
bird_zone_avg[, group := fifelse(species %in% anseriformes, "anseriformes", "predators")]

# Step 2: Sum across species within each group per zone-week
bird_grouped <- bird_zone_avg[
  , .(group_total = sum(abundance, na.rm = TRUE)),
  by = .(zone_id, week, group)
]

# Total — sum across ALL species per zone-week
bird_total <- bird_zone_avg[
  , .(total_bird_abundance = sum(abundance, na.rm = TRUE)),
  by = .(zone_id, week)
]

# Pivot groups wide
bird_wide_neg <- dcast(
  bird_grouped,
  zone_id + week ~ group,
  value.var = "group_total",
  fill = 0
)

# Merge total
bird_wide_neg <- merge(bird_wide_neg, bird_total, by = c("zone_id", "week"), all.x = TRUE)

# Only get zone-week combos for negative controls (no outbreak in the cell)
negative_zone_weeks <- as.data.table(st_drop_geometry(weekly_fishnet))[
  any_outbreak == 0, .(zone_id, week)
] %>% unique()

setkey(bird_wide_neg, zone_id, week)
setkey(negative_zone_weeks, zone_id, week)

bird_negative <- merge(
  negative_zone_weeks,
  bird_wide_neg,
  by = c("zone_id", "week"),
  all.x = TRUE
)

# Fill missing with 0
bird_negative[is.na(anseriformes), anseriformes := 0]
bird_negative[is.na(predators), predators := 0]
bird_negative[is.na(total_bird_abundance), total_bird_abundance := 0]
```

### Join bird data for negative controls into weekly fishnet
```{r, join bird data for negative controls}
# Join back to weekly_fishnet — only updates non-outbreak cells
# First rename to avoid collision with outbreak bird columns
setnames(bird_negative, 
  c("total_bird_abundance", "anseriformes", "predators"),
  c("neg_total_bird_abundance", "neg_anseriformes", "neg_predators")
)

setDT(weekly_fishnet)
setkey(weekly_fishnet, zone_id, week)
setkey(bird_negative, zone_id, week)

weekly_fishnet <- merge(
  weekly_fishnet,
  bird_negative,
  by = c("zone_id", "week"),
  all.x = TRUE
)

# Coalesce: use outbreak bird data for cases, negative control bird data for controls
weekly_fishnet[, `:=`(
  total_bird_abundance = fifelse(!is.na(total_bird_abundance), total_bird_abundance, neg_total_bird_abundance),
  anseriformes = fifelse(!is.na(anseriformes), anseriformes, neg_anseriformes),
  predators = fifelse(!is.na(predators), predators, neg_predators)
)]

# Drop temporary columns
weekly_fishnet[, c("neg_total_bird_abundance", "neg_anseriformes", "neg_predators") := NULL]

# Tag zones with zero bird abundance
zone_zero_summary <- weekly_fishnet[, .(total_abundance = sum(total_bird_abundance, na.rm = TRUE)), by = zone_id]
zone_ids_all_zero <- zone_zero_summary[total_abundance == 0, zone_id]
weekly_fishnet[, bird_zero := fifelse(zone_id %in% zone_ids_all_zero, "No Birds", "Has Birds")]
```

### Map bird data

```{r, map bird data}

weekly_fishnet_sf <- fishnet_geom %>%
  left_join(weekly_fishnet, by = "zone_id")


# Reproject to WGS 84 (for plotting or basemaps)

weekly_fishnet_sf <- st_transform(
weekly_fishnet_sf, 4326)

ggplot(
weekly_fishnet_sf) +
   geom_sf(aes(fill = bird_zero), color = NA, alpha = 0.9) +
   scale_fill_manual(
     values = c("No Birds" = "red", "Has Birds" = "lightgray"),
     name = "Bird Presence"
   ) +
   labs(
     title = "Zones with No Bird Abundance or Feedlots (All Weeks = 0)"
   ) +   theme_minimal()



```

# Check for any NAs in the columns

```{r, check for any NAs in the columns}

weekly_fishnet %>%
  summarise(across(everything(), ~sum(is.na(.))))

### Should only have NA for outbreak_id and outbreak_week
```

# Export dataframe

```{r, export dataframe}

saveRDS(weekly_fishnet, file.path(objects_folder, "mn_fishnet_processed.rds"))

export_csv <- file.path(processed_data,"mn_fishnet_processed.csv")
write.csv(weekly_fishnet,export_csv, row.names= FALSE) 

```


# Export spatial dataframe

```{r, export spatial dataframe}

saveRDS(weekly_fishnet_sf, file.path(objects_folder, "mn_fishnet_processed_sf.rds"))

export_sf <- file.path(processed_data,"mn_fishnet_processed.gpkg")
st_write(weekly_fishnet_sf,export_sf, delete_dsn = TRUE) 

```
```

