---
title: "Create Timeseries Dataframe"
author: "Fintan Mooney"
date: "2025-10-15"
output: html_document
---

# NOTE: This script requires confidential HPAI spillover point-location data
# (not publicly available). The processed grid-cell level output is provided
# in data/ and downstream analyses can be run without this file.

# Builds the timeseries panel dataframe for analyzing met conditions, climate anomalies,
# and wild bird abundance against HPAI outbreak risk on MN farms in 2022!
# Computes multi-year climatologies, z-scores, and flags non-independent spillovers...

```{r include=FALSE}
rm(list=ls())
project.folder = paste0(print(here::here()),'/')
source(paste0(project.folder,'create_folder_structure.R'))
source(paste0(functions.folder,'script_initiate.R'))
```


# Bring in base dataframe

```{r, bring in base dataframe}

## Bring in the weekly fishnet with data

weekly_fishnet <- readRDS(file.path(objects_folder, "mn_fishnet_processed.rds"))


```


# Read in climate data

```{r, read in climate data}
# Load all historical climate data for baselines
all_years <- 1997:2022
files <- file.path(env_data, paste0("mn_weekly_era5land_", all_years, ".csv"))
all_data <- map2_dfr(files, all_years, ~ read_csv(.x) %>% mutate(year = .y))
all_data <- all_data %>%
  mutate(
    week_start = as.Date(as.character(week_start), format = "%Y%m%d"),
    snow_density = ifelse(snow_density <= 100, 0, snow_density) # remove lower bound for snow density, default value for no snow
  )

# Create baseline periods
baseline_10yr <- all_data %>% filter(year %in% 2012:2021)
baseline_15yr <- all_data %>% filter(year %in% 2007:2021)
baseline_20yr <- all_data %>% filter(year %in% 2002:2021)
baseline_25yr <- all_data %>% filter(year %in% 1997:2021)

# Climate variables
vars <- c("leaf_area_index", "longwave_radiation", "runoff_sum", "shortwave_radiation",
          "snow_density", "soil_moisture", "relative_humidity", "temperature",
          "evapotranspiration", "precipitation", "wind_speed", "surface_pressure")

# Add ISO week and filter
add_week <- function(df) df %>% mutate(week = lubridate::isoweek(week_start)) %>% filter(week <= 52)
baseline_10yr <- add_week(baseline_10yr)
baseline_15yr <- add_week(baseline_15yr)
baseline_20yr <- add_week(baseline_20yr)
baseline_25yr <- add_week(baseline_25yr)

```

## Create 5-year climatologies 

```{r, create climatolgy}
# Summarize climatology
summarize_climatology <- function(df) {
  df %>%
    group_by(zone_id, week) %>%
    summarise(across(all_of(vars), list(mean = mean, sd = sd), .names = "{.col}_{.fn}"), .groups = "drop")
}

# Function to add year suffix to climatology columns
add_suffix <- function(df, yr_label) {
  clim_cols <- grep("_(mean|sd)$", names(df), value = TRUE)
  new_names <- paste0(clim_cols, "_", yr_label)
  setNames(df, ifelse(names(df) %in% clim_cols, 
                       paste0(names(df), "_", yr_label), 
                       names(df)))
}

clim_10yr <- add_suffix(summarize_climatology(baseline_10yr), "10yr")
clim_15yr <- add_suffix(summarize_climatology(baseline_15yr), "15yr")
clim_20yr <- add_suffix(summarize_climatology(baseline_20yr), "20yr")
clim_25yr <- add_suffix(summarize_climatology(baseline_25yr), "25yr")

# Join 
fishnet_timeseries <- weekly_fishnet %>%
  left_join(clim_10yr, by = c("zone_id", "week")) %>%
  left_join(clim_15yr, by = c("zone_id", "week")) %>%
  left_join(clim_20yr, by = c("zone_id", "week")) %>%
  left_join(clim_25yr, by = c("zone_id", "week"))

# Compute z-scores
for (v in vars) {
  for (yr in c(10, 15, 20, 25)) {
    m <- paste0(v, "_mean_", yr, "yr")
    s <- paste0(v, "_sd_", yr, "yr")
    z <- paste0(v, "_z_", yr, "yr")
    fishnet_timeseries[[z]] <- (fishnet_timeseries[[v]] - fishnet_timeseries[[m]]) / fishnet_timeseries[[s]]
  }
}


```

## We have extreme outliers so need to winsorize

```{r, winsorize}
# Winsorize z-scores
z_cols <- grep("_z_(10yr|15yr|20yr|25yr)$", names(fishnet_timeseries), value = TRUE)

winsorize_with_summary <- function(x, varname, limit = 4) {
  before_outliers <- sum(x < -limit | x > limit, na.rm = TRUE)
  x_winsorized <- pmin(pmax(x, -limit), limit)
  message(sprintf("%s: %d values winsorized (limit ±%s)", varname, before_outliers, limit))
  return(x_winsorized)
}

for (v in z_cols) {
  fishnet_timeseries[[v]] <- winsorize_with_summary(fishnet_timeseries[[v]], v, limit = 4)
}

```

# Check that outbreaks are independent at each zone (>4 weeks apart)
```{r, check on distinct outbreaks}

min_gap_weeks <- 4

# Get zone-weeks with outbreaks
zone_outbreaks <- fishnet_timeseries %>%
  filter(outbreak_binary == 1) %>%
  arrange(zone_id, week) %>%
  group_by(zone_id) %>%
  mutate(
    detection_num = row_number(),
    weeks_since_last = week - lag(week),
    independent = detection_num == 1 | (!is.na(weeks_since_last) & weeks_since_last >= min_gap_weeks)
  ) %>%
  ungroup()

# Summary
cat("Total outbreak zone-weeks:", nrow(zone_outbreaks), "\n")
cat("Unique zones with outbreaks:", n_distinct(zone_outbreaks$zone_id), "\n")
cat("Independent spillovers:", sum(zone_outbreaks$independent), "\n")
cat("Excluded (repeat <12 weeks):", sum(!zone_outbreaks$independent), "\n")

# Inspect zones with multiple outbreak weeks
zone_outbreaks %>%
  group_by(zone_id) %>%
  filter(n() > 1) %>%
  arrange(zone_id, week) %>%
  select(zone_id, week, outbreak_count, outbreak_id, detection_num, weeks_since_last, independent) %>%
  print(n = 20)

# Excluded zone-weeks
excluded_zone_weeks <- zone_outbreaks %>%
  filter(!independent) %>%
  select(zone_id, week)

cat("\nExcluded zone-weeks:\n")
print(excluded_zone_weeks, n = 20)
```

- There are 19 zones that have more than one spillover event within 4 weeks; however, we need to check with the original farm-outbreak dataset to see if these outbreaks actually came from the same farm

## Check with raw outbreak data 

```{r, check with the raw outbreak data}

# Load raw data
spillover_file <- file.path(spillover_data, "hpai-outbreaks-clean-placeholder-032124.csv")
if (!file.exists(spillover_file)) {
  stop("This script needs the confidential spillover data (hpai-outbreaks-clean-placeholder-032124.csv). See data/confidential/README.md for how to get it.")
}
hpai_outbreaks <- read_csv(spillover_file)

# Filter to Minnesota 2022
hpai_mn_2022 <- hpai_outbreaks %>%
  filter(
    state == "MN",
    date_confirmed >= as.Date("2022-01-01"),
    date_confirmed <= as.Date("2022-12-31")
  ) %>%
  mutate(week = as.numeric(floor(difftime(date_confirmed, as.Date("2022-01-01"), units = "weeks")) + 1))

cat("Total MN outbreaks in 2022:", nrow(hpai_mn_2022), "\n")

# Check repeat detections at same premises
hpai_mn_2022 %>%
  group_by(lat, lon) %>%
  filter(n() > 1) %>%
  arrange(lat, lon, date_confirmed) %>%
  mutate(
    days_since_last = as.numeric(date_confirmed - lag(date_confirmed)),
    weeks_since_last = days_since_last / 7
  ) %>%
  select(lat, lon, date_confirmed, week, days_since_last, weeks_since_last) %>%
  print()

```

## Check distance between outbreaks at same cell

```{r, look at distance between outbreaks at same cell}

# Get the CRS from the original fishnet timeseries
fishnet_raw <- st_read(file.path(spatial_data, "fishnet_10km_timeseries.gpkg"))
target_crs <- st_crs(fishnet_raw)
print(target_crs)

weekly_fishnet_sf_path <- file.path(processed_data, "mn_fishnet_processed.gpkg")

weekly_fishnet_sf <- st_read(weekly_fishnet_sf_path) %>%
  st_transform(crs = target_crs)

hpai_mn_sf <- hpai_mn_2022 %>%
  st_as_sf(coords = c("lon", "lat"), crs = 4326) %>%
  st_transform(crs = target_crs)

# Assign zone_id
hpai_mn_sf <- st_join(hpai_mn_sf, weekly_fishnet_sf %>% distinct(zone_id, geom), join = st_within)

# For zones with multiple outbreaks, compute pairwise distances and time gaps
multi_outbreak_zones <- hpai_mn_sf %>%
  st_drop_geometry() %>%
  count(zone_id) %>%
  filter(n > 1) %>%
  pull(zone_id)

# Pairwise comparison within each zone
pairwise_results <- list()

for (z in multi_outbreak_zones) {
  zone_pts <- hpai_mn_sf %>% filter(zone_id == z) %>% arrange(date_confirmed)
  n <- nrow(zone_pts)
  if (n < 2) next
  
  for (i in 1:(n - 1)) {
    for (j in (i + 1):n) {
      dist_m <- as.numeric(st_distance(zone_pts[i, ], zone_pts[j, ]))
      days_apart <- as.numeric(zone_pts$date_confirmed[j] - zone_pts$date_confirmed[i])
      
      pairwise_results[[length(pairwise_results) + 1]] <- tibble(
        zone_id = z,
        outbreak_id_1 = zone_pts$outbreak_id[i],
        outbreak_id_2 = zone_pts$outbreak_id[j],
        distance_km = dist_m / 1000,
        days_apart = days_apart,
        weeks_apart = days_apart / 7,
        same_farm = dist_m < 100  # <100m = essentially same location
      )
    }
  }
}

pairwise_df <- bind_rows(pairwise_results) %>%
  arrange(distance_km, days_apart)

print(pairwise_df, n = 40)

cat("\nDistance summary for within-zone outbreak pairs:\n")
summary(pairwise_df$distance_km)


# Look at non-same-farm pairs that are close and within 4 weeks
pairwise_df %>%
  filter(!same_farm, weeks_apart < 4) %>%
  mutate(
    exclude_200m = distance_km < 0.2,
    exclude_500m = distance_km < 0.5,
    exclude_1km  = distance_km < 1.0
  ) %>%
  select(zone_id, outbreak_id_1, outbreak_id_2, distance_km, weeks_apart, 
         exclude_200m, exclude_500m, exclude_1km) %>%
  print(n = 40)

# Count
cat("\nExcluded under <200m & <4k:", 
    sum(pairwise_df$distance_km < 0.2 & pairwise_df$weeks_apart < 4 & !pairwise_df$same_farm), "\n")
cat("Excluded under <500m & <4wk:", 
    sum(pairwise_df$distance_km < 0.5 & pairwise_df$weeks_apart < 4 & !pairwise_df$same_farm), "\n")
cat("Excluded under <1km & <4wk:", 
    sum(pairwise_df$distance_km < 1.0 & pairwise_df$weeks_apart < 4 & !pairwise_df$same_farm), "\n")

pairwise_df %>%
  filter(!same_farm) %>%
  mutate(dist_bin = cut(distance_km, breaks = c(0, 0.2, 0.5, 1, 2, 5, 11))) %>%
  group_by(dist_bin) %>%
  summarise(
    n_pairs = n(),
    pct_within_4wk = mean(weeks_apart < 4) * 100,
    median_weeks_apart = median(weeks_apart)
  )


```

## Check for non-independent outbreaks across different cells
```{r, cross-cell pairwise distances}

# All pairwise distances between outbreaks in DIFFERENT zones, within 4 weeks
cross_cell_results <- list()

# Sort by date for consistent ordering
hpai_mn_sf <- hpai_mn_sf %>% arrange(date_confirmed)

n_outbreaks <- nrow(hpai_mn_sf)

for (i in 1:(n_outbreaks - 1)) {
  for (j in (i + 1):n_outbreaks) {
    # Only compare outbreaks in different zones
    if (hpai_mn_sf$zone_id[i] == hpai_mn_sf$zone_id[j]) next
    
    days_apart <- as.numeric(hpai_mn_sf$date_confirmed[j] - hpai_mn_sf$date_confirmed[i])
    
    # Only check pairs within 4 weeks to save computation
    if (days_apart > 28) next
    
    dist_m <- as.numeric(st_distance(hpai_mn_sf[i, ], hpai_mn_sf[j, ]))
    
    # Only keep if within 1km (max threshold we care about)
    if (dist_m > 1000) next
    
    cross_cell_results[[length(cross_cell_results) + 1]] <- tibble(
      zone_id_1 = hpai_mn_sf$zone_id[i],
      zone_id_2 = hpai_mn_sf$zone_id[j],
      outbreak_id_1 = hpai_mn_sf$outbreak_id[i],
      outbreak_id_2 = hpai_mn_sf$outbreak_id[j],
      distance_km = dist_m / 1000,
      days_apart = days_apart,
      weeks_apart = days_apart / 7
    )
  }
}

if (length(cross_cell_results) > 0) {
  cross_cell_df <- bind_rows(cross_cell_results) %>%
    arrange(distance_km, days_apart)
  
  cat("Cross-cell pairs within 1km & 4 weeks:\n")
  print(cross_cell_df)
  
  cat("\nCross-cell pairs by threshold:\n")
  cat("  <200m & <4 wks:", sum(cross_cell_df$distance_km < 0.2), "\n")
  cat("  <500m & <4 wks:", sum(cross_cell_df$distance_km < 0.5), "\n")
  cat("  <1km & <4 wks:",  sum(cross_cell_df$distance_km < 1.0), "\n")
} else {
  cat("No cross-cell outbreak pairs within 1km & 4 weeks.\n")
}
```



## We are going to flag the second outbreaks that occur in the same cells as another outbreak within 4 weeks and less than 200m, 500m, and 1km.

```{r, flag second outbreaks by rules}

# Get outbreak_ids flagged under each threshold (always the later outbreak)
flag_ids <- pairwise_df %>%
  filter(weeks_apart < 4) %>%
  mutate(
    flag_200m = distance_km < 0.2,
    flag_500m = distance_km < 0.5,
    flag_1km  = distance_km < 1.0
  ) %>%
  summarise(
    ids_200m = list(unique(outbreak_id_2[flag_200m])),
    ids_500m = list(unique(outbreak_id_2[flag_500m])),
    ids_1km  = list(unique(outbreak_id_2[flag_1km]))
  )

ids_200m <- flag_ids$ids_200m[[1]]
ids_500m <- flag_ids$ids_500m[[1]]
ids_1km  <- flag_ids$ids_1km[[1]]
# Add cross-cell non-independent outbreak to 1km flags
ids_1km <- union(ids_1km, 268)

# Helper: check if any flagged id is in a comma-separated string
contains_flagged <- function(id_str, flagged_ids) {
  if (is.na(id_str)) return(FALSE)
  ids <- as.numeric(trimws(unlist(strsplit(id_str, ","))))
  any(ids %in% flagged_ids)
}

fishnet_timeseries <- fishnet_timeseries %>%
  mutate(
    flag_nearby_200m = sapply(outbreak_id, contains_flagged, ids_200m),
    flag_nearby_500m = sapply(outbreak_id, contains_flagged, ids_500m),
    flag_nearby_1km  = sapply(outbreak_id, contains_flagged, ids_1km)
  )

table(fishnet_timeseries$flag_nearby_200m)
table(fishnet_timeseries$flag_nearby_500m)
table(fishnet_timeseries$flag_nearby_1km)

```



# More checks
```{r, more checks}


# Check total outcomes and outcome_binary
cat("outbreak binary:", sum(fishnet_timeseries$outbreak_binary), "\n")
cat("outbreak count:", sum(fishnet_timeseries$outbreak_count), "\n")

# Check outbreaks that are within 4 weeks and in a distance threshold
cat("outbreak is after an outbreak within 4 weeks and less than 200m away:", sum(fishnet_timeseries$flag_nearby_200m), "\n")
cat("outbreak is after an outbreak within 4 weeks and less than 500m away:", sum(fishnet_timeseries$flag_nearby_500m), "\n")
cat("outbreak is after an outbreak within 4 weeks and less than 1km away:", sum(fishnet_timeseries$flag_nearby_1km), "\n")

# Same-farm repeat outbreaks within 4 weeks
same_farm_4wk <- pairwise_df %>%
  filter(same_farm, weeks_apart < 4) %>%
  pull(outbreak_id_2) %>%
  unique()

# Combined flags (same farm + distance threshold)
ids_same_farm   <- same_farm_4wk
ids_200m_all    <- union(ids_same_farm, ids_200m)
ids_500m_all    <- union(ids_same_farm, ids_500m)
ids_1km_all     <- union(ids_same_farm, ids_1km)

# Full breakdown using outbreak events (not zone-weeks)
outbreak_events <- hpai_mn_sf %>%
  st_drop_geometry() %>%
  select(outbreak_id, zone_id, date_confirmed)

# Total outbreak events
total_events <- sum(fishnet_timeseries$outbreak_count)
cat("Total outbreak events:", total_events, "\n")
cat("Outbreak zone-weeks:", sum(fishnet_timeseries$outbreak_binary), "\n")

# Non-independent events at each distance threshold (<4 wks)
cat("\n--- Non-independent outbreak events ---\n")
cat("Non-independent (<200m & <4 wks):", length(ids_200m), "\n")
cat("Non-independent (<500m & <4 wks):", length(ids_500m), "\n")
cat("Non-independent (<1km & <4 wks):", length(ids_1km), "\n")

# Independent events
cat("\n--- Independent events ---\n")
cat("Independent (200m threshold):", total_events - length(ids_200m), "\n")
cat("Independent (500m threshold):", total_events - length(ids_500m), "\n")
cat("Independent (1km threshold):", total_events - length(ids_1km), "\n")

# Check how many zones have cases vs total zones
fishnet_timeseries %>%
  group_by(zone_id) %>%
  reframe(has_case = any(any_outbreak == 1)) %>%
  count(has_case)

# Count case-weeks and control-weeks
case_weeks <- fishnet_timeseries %>% filter(any_outbreak == 1) %>% nrow()
negative_control_weeks <- fishnet_timeseries %>% filter(any_outbreak == 0) %>% nrow()

cat("\nTotal zone-weeks:", nrow(fishnet_timeseries), "\n")
cat("Case-weeks (zones with any outbreak):", case_weeks, "\n")
cat("Negative control-weeks (zones never had outbreak):", negative_control_weeks, "\n")

```



# Create any extra variables

```{r, create extra variables}
fishnet_timeseries <- fishnet_timeseries %>%
  group_by(zone_id) %>%
  mutate(any_outbreak = as.integer(any(outbreak_binary > 0))) %>%
  ungroup()


fishnet_timeseries <- fishnet_timeseries %>%
  mutate(
    # Convert week index to month 
    month = lubridate::month(lubridate::ymd("2022-01-01") + lubridate::weeks(week - 1)),
    
    month_factor = as.factor(month),

    # Create warm season dummy: May (5) to September (9)
    warm_season = ifelse(month %in% 5:9, 1, 0),

    # Create spring dummy: March (3), April (4), May (5)
    spring = ifelse(month %in% 3:5, 1, 0),

    # Create spring/fall dummy: March-May or September-November
    spring_fall = ifelse(month %in% c(3:5, 9:11), 1, 0),
    

    # Create season factor
    season = case_when(
      month %in% c(12, 1, 2) ~ "Winter",
      month %in% 3:5         ~ "Spring",
      month %in% 6:8         ~ "Summer",
      month %in% 9:11        ~ "Fall"
    ),
    season = factor(season, levels = c("Winter", "Spring", "Summer", "Fall"))

  )



# Constants
p <- 1013.25  # pressure in hPa (standard sea-level pressure)

fishnet_timeseries <- fishnet_timeseries %>%
  mutate(
    # Convert temperature to Kelvin
    temp_k = temperature + 273.15,

  )

seasonal_outbreaks <- fishnet_timeseries %>%
  group_by(season) %>%
  summarise(outbreaks = sum(outbreak_count, na.rm = TRUE)) %>%
  arrange(match(season, c("Winter", "Spring", "Summer", "Fall")))

```

## Export summary of independent outbreaks (needed season column)
```{r, export summary of independent outreaks}


# Export Summary

# Per-season breakdown
season_independent <- fishnet_timeseries %>%
  filter(outbreak_binary == 1) %>%
  mutate(
    independent_count = ifelse(flag_nearby_500m, outbreak_count - 1L, outbreak_count)
  ) %>%
  group_by(season) %>%
  summarise(
    total_events = sum(outbreak_count, na.rm = TRUE),
    independent_events = sum(independent_count, na.rm = TRUE),
    .groups = "drop"
  )

spillover_summary <- list(
  total = sum(fishnet_timeseries$outbreak_count, na.rm = TRUE),
  non_independent_200m = length(ids_200m),
  non_independent_500m = length(ids_500m),
  non_independent_1km  = length(ids_1km),
  by_season = season_independent
)

saveRDS(spillover_summary, file.path(objects_folder, "spillover_summary.RDS"))

```


# Change any variable names 

```{r, change variable names}
 
fishnet_timeseries <- fishnet_timeseries %>%
  rename_with(~ gsub("runoff_sum", "runoff", .x), .cols = everything())

weekly_fishnet <- weekly_fishnet %>%
  rename_with(~ gsub("runoff_sum", "runoff", .x), .cols = everything())

```

# Note on using the dataframe
# 
# weekly_fishnet (or fishnet_timeseries) is a zone_id x week panel (173,368 rows).
#
# Key outcome columns:
#   - outbreak_binary: 1 if outbreak in that zone-week, 0 otherwise
#   - outbreak_count: number of outbreaks in that zone-week
#   - any_outbreak: 1 if the zone ever had an outbreak (any week), 0 otherwise
#
# Independence flags (for excluding potential lateral spread):
#   - flag_nearby_1km:  TRUE if zone-week contains a "second" outbreak <1km & <12wks from a prior outbreak in same zone
#   - flag_nearby_500m: TRUE if <500m & <12wks
#   - flag_nearby_200m: TRUE if <200m & <12wks
#
# Analysis filtering:
#   Primary analysis:      filter(!flag_nearby_1km | outbreak_binary == 0)   — removes 9 zone-weeks
#   Sensitivity 1:         filter(!flag_nearby_500m | outbreak_binary == 0)  — removes 3 zone-weeks
#   Sensitivity 2:         no exclusion, use all data
#
# Bird abundance columns:
#   - total_bird_abundance, anseriformes, predators: 
#       For outbreak zones (any_outbreak == 1): derived from outbreak-specific eBird estimates (averaged if >1 outbreak per zone-week)
#       For non-outbreak zones (any_outbreak == 0): derived from zone-level eBird estimates for negative controls
#
# Climate columns:
#   - Raw 2022 values: temperature, precipitation, wind_speed, etc.
#   - Z-scores: {variable}_z_{10/15/20/25}yr — 2022 values compared to baseline climatology (winsorized at ±4)
#
# Feedlot columns:
#   - feedlot_count, au_count, birds_au: summed per zone
#   - cafo_flag: 1 if any CAFO in zone
#   - near_river: 1 if any feedlot near river in zone
#
# Land cover columns:
#   - land_cover_label: majority land cover type per zone (reference: "Developed, Medium Intensity")
#   - farm_possible: 1 if majority land cover is Cultivated Crops, Pasture/Hay, or Emergent Herbaceous Wetlands
#
# Other:
#   - outbreak_week: comma-separated weeks when outbreaks occurred in that zone (zone-level, same for all weeks)
#   - outbreak_id: comma-separated outbreak IDs for that zone-week (NA if no outbreak)
#   - bird_zero: "No Birds" / "Has Birds" — zones with zero bird abundance across all weeks


## Export RDS

```{r, export rds}

saveRDS(fishnet_timeseries, file.path(objects_folder, "fishnet_timeseries.RDS"))
```

# Export version of timeseries excluding the second outbreaks that occur within 4 weeks and within 500m of a prior outbreak
## This dataframe will be used in the modeling of the climate anomalies and the model that controls for astrological season

```{r, export modeling timeseries dataframe}

# Identify the LATER outbreak in each non-independent pair (within 500m and 4 weeks, including same farm)
non_independent_pairs <- pairwise_df %>%
  filter(distance_km < 0.5, weeks_apart < 4)

# The proceeding outbreak 
non_independent_outbreaks <- non_independent_pairs %>%
  pull(outbreak_id_2) %>%
  unique()

# Count how many non-independent outbreak IDs fall in each zone-week
excluded_zone_weeks <- fishnet_timeseries %>%
  filter(outbreak_binary == 1) %>%
  rowwise() %>%
  mutate(
    n_flagged = sum(as.numeric(trimws(unlist(strsplit(outbreak_id, ",")))) %in% non_independent_outbreaks)
  ) %>%
  ungroup() %>%
  filter(n_flagged > 0) %>%
  select(zone_id, week, n_flagged)

cat("Zone-weeks with flagged outbreaks:", nrow(excluded_zone_weeks), "\n")
print(excluded_zone_weeks)

# Subtract the actual number of non-independent events per zone-week
fishnet_timeseries_modeling <- fishnet_timeseries %>%
  left_join(excluded_zone_weeks, by = c("zone_id", "week")) %>%
  mutate(
    n_flagged = ifelse(is.na(n_flagged), 0L, n_flagged),
    outbreak_count = outbreak_count - n_flagged,
    outbreak_binary = ifelse(outbreak_count > 0, 1L, 0L)
  ) %>%
  select(-n_flagged)

# Verify
cat("Original outbreak events:", sum(fishnet_timeseries$outbreak_count), "\n")
cat("Modeling outbreak events:", sum(fishnet_timeseries_modeling$outbreak_count), "\n")
cat("Removed:", sum(fishnet_timeseries$outbreak_count) - sum(fishnet_timeseries_modeling$outbreak_count), "\n")
cat("Outbreak binary zone-weeks:", sum(fishnet_timeseries_modeling$outbreak_binary), "\n")

saveRDS(fishnet_timeseries_modeling, file.path(objects_folder, "fishnet_timeseries_modeling.RDS"))


```