---
title: "Create Case-Crossover Dataframe"
author: "Fintan Mooney"
date: "2025-10-15"
output: html_document
---

# NOTE: This script requires confidential HPAI spillover point-location data
# (not publicly available). The processed grid-cell level output is provided
# in data/ and downstream analyses can be run without this file.

# Creates the case-crossover dataframe for the HPAI outbreak analysis!
# Defines case and control windows, adds lagged predictors, and handles
# non-independent spillover exclusions at multiple distance thresholds...

```{r include=FALSE}
rm(list=ls())
project.folder = paste0(print(here::here()),'/')
source(paste0(project.folder,'create_folder_structure.R'))
source(paste0(functions.folder,'script_initiate.R'))
```


## Read in the timeseries dataset

```{r, read in timeseries dataset}

# Timeseries
fishnet_timeseries_clean <- readRDS(file.path(objects_folder, "fishnet_timeseries.RDS"))

fishnet_timeseries_clean <- fishnet_timeseries_clean %>%
  arrange(zone_id, week) %>%
  # 1. Drop all columns that end with ".x"
  select(-ends_with(".x")) %>%
  # 2. Rename columns ending with ".y" to remove the suffix
  rename_with(~ str_remove(.x, "\\.y$"), ends_with(".y"))

# Spatial GPKG
weekly_fishnet_sf <- st_read(paste0(spatial_data, "/fishnet_10km_timeseries.gpkg"))


```

## Map Cases in Casecrossover Model

```{r, casecrossover map}


# Map 
# Identify case zones
case_zone_ids <- fishnet_timeseries_clean %>%
  filter(any_outbreak == 1) %>%
  distinct(zone_id)

# Identify negative control zones (has feedlots but no outbreak)
neg_control_zone_ids <- fishnet_timeseries_clean %>%
  filter(any_outbreak == 0, feedlot_count > 0) %>%
  distinct(zone_id)

# Subset spatial features (first week only to remove timeseries)
case_zones_sf <- weekly_fishnet_sf %>%
  filter(zone_id %in% case_zone_ids$zone_id) %>%
  group_by(zone_id) %>%
  slice(1) %>%
  ungroup()

control_zones_sf <- weekly_fishnet_sf %>%
  filter(!(zone_id %in% case_zone_ids$zone_id)) %>%
  group_by(zone_id) %>%
  slice(1) %>%
  ungroup()

neg_control_zones_sf <- weekly_fishnet_sf %>%
  filter(zone_id %in% neg_control_zone_ids$zone_id) %>%
  group_by(zone_id) %>%
  slice(1) %>%
  ungroup()

# Define output paths
case_gpkg_out        <- file.path(processed_data, "Case_Zones.gpkg")
control_gpkg_out     <- file.path(processed_data, "Control_Zones.gpkg")
neg_control_gpkg_out <- file.path(processed_data, "Neg_Control_Zones.gpkg")

# Export as GPKG
st_write(case_zones_sf, case_gpkg_out, delete_layer = TRUE)
st_write(control_zones_sf, control_gpkg_out, delete_layer = TRUE)
st_write(neg_control_zones_sf, neg_control_gpkg_out, delete_layer = TRUE)

# Reproject to EPSG:3857 (Web Mercator) for basemap compatibility
case_zones_sf_proj <- st_transform(case_zones_sf, 3857)
control_zones_sf_proj <- st_transform(control_zones_sf, 3857)
neg_control_zones_sf_proj <- st_transform(neg_control_zones_sf, 3857)

# Plot with basemap
ggplot() +
  # Add simple basemap
  annotation_map_tile(type = "cartolight", zoomin = 0) +

  # Negative Control zones: outlines and fill
  geom_sf(data = neg_control_zones_sf_proj, fill = "lightblue", color = "lightblue", linewidth = 0.05) +

  # Case zones: bold fill
  geom_sf(data = case_zones_sf_proj, fill = "#d73027", color = "red", linewidth = 0.1, alpha = 0.9) +

  # Control zones: outlines (without feedlot or case)
  geom_sf(data = control_zones_sf_proj, fill = NA, color = "lightgrey", linewidth = 0.05) +

  # Scale bar
  annotation_scale(location = "bl", width_hint = 0.3, text_cex = 0.75, line_width = 0.5, style = "ticks") +

  # Theme and layout
  theme_void(base_size = 12) +
  theme(
    plot.title = element_text(size = 12, face = "bold", hjust = 0.5),
    plot.caption = element_text(size = 10, hjust = 0.5, face = "italic"),
    plot.background = element_rect(fill = "white", color = NA)
  ) +
  coord_sf(expand = FALSE) +
  labs(
    title = "HPAI Outbreak Case Zones in Minnesota (2022)"
  )
```

## Define Controls and Create Case-Crossover Dataset
```{r, define controls}

# ---- Define Cases and Controls ----

# Filter to zones with any outbreak
case_zones <- fishnet_timeseries_clean %>%
  filter(any_outbreak == 1) 

# Get cases: one row per zone_id and week with outbreak
cases <- case_zones %>%
  filter(outbreak_count > 0) %>%
  distinct(zone_id, case_week = week)

# Deduplicate case_zones for joining
case_zones_unique <- case_zones %>%
  distinct(zone_id, week, .keep_all = TRUE)

# Create control weeks (1-4 weeks before each case)
case_lags <- cases %>%
  crossing(lag = 1:4) %>%
  mutate(control_week = case_week - lag) %>%
  filter(control_week >= 1)

# Remove overlapping controls (control week = case week in same zone)
case_zone_weeks <- cases %>% select(zone_id, case_week)

case_lags_clean <- case_lags %>%
  anti_join(case_zone_weeks, by = c("zone_id", "control_week" = "case_week"))

cat("Controls removed (overlap with case weeks):", nrow(case_lags) - nrow(case_lags_clean), "\n")

### HERE WE REMOVED CONTROLS THAT OVERLAP
#### Specifically, some cells have outbreaks at two different farms within the same four week period; we do not delete them as cases, but we remove the overlapping controls. 

# Build controls dataset
controls <- case_lags_clean %>%
  left_join(case_zones_unique, by = c("zone_id", "control_week" = "week")) %>%
  mutate(
    case_control = "control",
    outbreak_binary = 0,
    stratum_id = paste(zone_id, case_week, sep = "_"),
    week = control_week
  ) %>%
  dplyr::select(-outbreak_count)

# Build cases dataset
case_data <- cases %>%
  left_join(case_zones_unique, by = c("zone_id", "case_week" = "week")) %>%
  mutate(
    case_control = "case",
    outbreak_binary = 1,
    stratum_id = paste(zone_id, case_week, sep = "_"),
    week = case_week,
    control_week = NA
  ) %>%
  dplyr::select(-outbreak_count)

# Combine cases and controls
case_crossover_df <- bind_rows(case_data, controls)

# ---- Summary checks ----
cat("\nCase-control distribution:\n")
table(case_crossover_df$case_control)

cat("\nTotal cases:", sum(case_crossover_df$outbreak_binary), "\n")

# Check controls per case
controls_per_case <- case_crossover_df %>%
  filter(case_control == "control") %>%
  count(stratum_id, name = "n_controls")

cat("\nControls per case:\n")
table(controls_per_case$n_controls)

colnames(case_crossover_df)



```

## Validate control selection for zones with multiple cases
```{r, validate multi-case zones}

# Check: how many zone-weeks have outbreaks?
case_zones %>%
  filter(outbreak_count > 0) %>%
  distinct(zone_id, week) %>%
  nrow()

# vs total outbreak events
print(sum(case_zones$outbreak_count))



# Zones with multiple cases
multi_case_zones <- cases %>%
  group_by(zone_id) %>%
  summarise(n_cases = n(), case_weeks = list(case_week), .groups = "drop") %>%
  filter(n_cases > 1)

cat("Zones with multiple cases:", nrow(multi_case_zones), "\n")
cat("Total cases in multi-case zones:", sum(multi_case_zones$n_cases), "\n")

# Check: Do any control weeks overlap with case weeks in the same zone?
case_zone_weeks <- cases %>% select(zone_id, case_week)

proposed_controls <- cases %>%
  crossing(lag = 1:4) %>%
  mutate(control_week = case_week - lag) %>%
  filter(control_week >= 1)

# Flag controls that are also case weeks in the same zone
overlap <- proposed_controls %>%
  inner_join(case_zone_weeks, by = c("zone_id", "control_week" = "case_week"))

cat("Control weeks that are ALSO case weeks in same zone:", nrow(overlap), "\n")

if (nrow(overlap) > 0) {
  print(overlap)
} else {
  cat("✓ No overlap — all control weeks are case-free in their respective zones.\n")
}

## WE DO NOT REMOVE CASES BECAUSE EVEN THOUGH THEY MAY HAVE HAPPENED IN THE SAME ZONE OVER THE FOUR WEEK PERIOD THEY WERE FROM DIFFERENT FARMS


# Check for duplicate zone-week combinations in controls
control_dupes <- case_crossover_df %>%
  filter(case_control == "control") %>%
  count(zone_id, week) %>%
  filter(n > 1)

cat("Control zone-weeks appearing more than once:", nrow(control_dupes), "\n")

if (nrow(control_dupes) > 0) {
  print(control_dupes)
  
  # Show which strata these belong to
  case_crossover_df %>%
    filter(case_control == "control") %>%
    inner_join(control_dupes, by = c("zone_id", "week")) %>%
    select(zone_id, week, stratum_id, case_week) %>%
    arrange(zone_id, week) %>%
    print(n = 30)
}

## WE DO NOT REMOVE CONTROLS ZONE-WEEKS THAT APPEAR MORE THAN ONE BECAUSE THEY COME FROM DIFFERENT STRATA AND THUS ARE SUITABLE FOR THE CONDITONAL LOGISTIC REGRESSION
```

# Check distribution of casecrossover variables

```{r, distribution caseccrossover}

# Move from m to mm

case_crossover_df <- case_crossover_df %>%
  mutate(
    runoff = runoff * 1000,
    precipitation = precipitation * 1000
  )


# Variables to inspect
vars_to_check <- c(
  "total_bird_abundance",   
  "leaf_area_index", "longwave_radiation", "runoff", "shortwave_radiation",
  "snow_density", "soil_moisture", "temperature",
  "evapotranspiration", "precipitation", "wind_speed", "relative_humidity", "anseriformes", "predators"
)

# Create histograms
plots <- lapply(vars_to_check, function(var) {
  ggplot(case_crossover_df, aes_string(x = var)) +
    geom_histogram(bins = 50, fill = "steelblue", color = "white") +
    labs(title = var) +
    theme_minimal()
})

# Display plots in a grid (adjust number of columns if needed)
wrap_plots(plots, ncol = )

library(e1071)
# Subset to variables
numeric_vars <- case_crossover_df %>%
  select(all_of(vars_to_check))

# Initialize result storage
shapiro_results <- lapply(names(numeric_vars), function(var) {
  x <- numeric_vars[[var]]
  x <- na.omit(x)
  
  # Only apply Shapiro if <= 5000 non-NA values
  shapiro_p <- if (length(x) <= 5000) {
    shapiro.test(x)$p.value
  } else {
    NA
  }
  
  data.frame(
    variable = var,
    skewness = skewness(x),
    shapiro_p = shapiro_p,
    n = length(x)
  )
})

# Combine into summary table
shapiro_df <- bind_rows(shapiro_results)

# Flag non-normality
shapiro_df <- shapiro_df %>%
  mutate(non_normal = ifelse(!is.na(shapiro_p) & shapiro_p < 0.05, TRUE, FALSE)) %>%
  arrange(desc(abs(skewness)))

# View top
print(shapiro_df)

### Need to transform runoff, total_bird_abundance, specific humidity, and precipitation, anseriformes and predators


```

## Create lagged variables

```{r, lag variables}

# Drop duplicates from original_df to ensure safe left_joins
case_zones_unique <- case_zones %>%
  distinct(zone_id, week, .keep_all = TRUE)


add_lagged_predictors <- function(df, original_df, vars, id_col = "zone_id", time_col = "week", max_lag = 4) {
  for (varname in vars) {
    for (lag in 0:max_lag) {
      lagged_col <- paste0(varname, "_Lag", lag)
      df <- df %>%
        left_join(
          original_df %>%
            dplyr::select(!!sym(id_col), !!sym(time_col), !!sym(varname)) %>%
            rename(!!lagged_col := !!sym(varname)) %>%
            mutate(!!time_col := !!sym(time_col) + lag),
          by = c(id_col, time_col)
        )
    }
  }
  return(df)
}


vars_to_lag <- c( "total_bird_abundance",
                 "leaf_area_index",
                 "longwave_radiation",
                 "runoff",
                 "shortwave_radiation",
                 "snow_density",
                 "soil_moisture",
                 "temperature",
                 "evapotranspiration",
                 "precipitation",
                 "wind_speed", 
                 "relative_humidity",
                 "anseriformes", "predators"
                 )

case_crossover_df <- add_lagged_predictors(
  df = case_crossover_df,
  original_df = case_zones_unique,
  vars = vars_to_lag,
  max_lag = 4
)

# Verify lagged variables were added correctly
cat("Columns in case_crossover_df:", ncol(case_crossover_df), "\n")
cat("Rows in case_crossover_df:", nrow(case_crossover_df), "\n")

# Check for missing values in lagged variables
case_crossover_df %>%
  select(ends_with("_Lag0")) %>%
  summarise(across(everything(), ~sum(is.na(.)))) %>%
  pivot_longer(everything(), names_to = "var", values_to = "n_missing") %>%
  filter(n_missing > 0)

## NOTHING IS MISSING
```




## Transform and Scale Variables
```{r, transform variables}

# Apply log1p only to skewed ones before scaling
for (var in c("total_bird_abundance", "runoff", "precipitation",  "anseriformes", "predators")) {
  for (lag in 0:4) {
    col <- paste0(var, "_Lag", lag)
    if (col %in% names(case_crossover_df)) {
      case_crossover_df[[col]] <- log1p(case_crossover_df[[col]])
    }
  }
}

```

```{r, scale}

# Filter to 2022 lag variables

vars_to_lag <- c("total_bird_abundance",
                 "leaf_area_index",
                 "longwave_radiation",
                 "runoff",
                 "shortwave_radiation",
                 "snow_density",
                 "soil_moisture",
                 "temperature",
                 "evapotranspiration",
                 "precipitation",
                 "wind_speed", 
                 "relative_humidity",
                 "anseriformes", "predators")

# Scale all lags (Lag0 to Lag4 for each variable)
for (var in vars_to_lag) {
  for (lag in 0:4) {
    lagged_col <- paste0(var, "_Lag", lag)
    case_crossover_df[[lagged_col]] <- scale(case_crossover_df[[lagged_col]])
  }
}
```

## Check that no control weeks have cases

```{r, validation check}

# Check for overlap in final dataset
case_weeks <- cases %>% select(zone_id, case_week)

overlap_check <- case_crossover_df %>%
  filter(case_control == "control") %>%
  inner_join(case_weeks, by = c("zone_id", "week" = "case_week"))

cat("Control weeks overlapping with case weeks:", nrow(overlap_check), "\n")  # should be 0
```

## Count number of cases and controls again

```{r, count number of cases and controls again}


# Cases and controls
cat("Total cases (zone-weeks):", sum(case_crossover_df$outbreak_binary == 1), "\n")
cat("Total controls (zone-weeks):", sum(case_crossover_df$outbreak_binary == 0), "\n")

# Total outbreak events (sum of outbreak_count for case rows)
total_outbreaks <- case_crossover_df %>%
  filter(outbreak_binary == 1) %>%
  left_join(
    case_zones_unique %>% select(zone_id, week, outbreak_count),
    by = c("zone_id", "week")
  ) %>%
  summarise(total = sum(outbreak_count, na.rm = TRUE)) %>%
  pull(total)

cat("Total outbreak events:", total_outbreaks, "\n")

# Summary table
cat("\nCase-control distribution:\n")
table(case_crossover_df$case_control)

# Controls per case
cat("\nControls per case:\n")
case_crossover_df %>%
  filter(case_control == "control") %>%
  count(stratum_id) %>%
  count(n, name = "n_cases") %>%
  rename(n_controls = n)

```

# Only remove strata where the sole outbreak is non-independent (outbreak_count == 1)
```{r, remove non-independent}

# Get strata that contain a flagged zone-week
flagged_strata <- case_crossover_df %>%
  filter(flag_nearby_500m == TRUE, case_control == "case") %>%
  distinct(stratum_id, zone_id, week)

# Join outbreak_count to determine if the zone-week has other independent outbreaks
flagged_with_count <- flagged_strata %>%
  left_join(
    case_zones_unique %>% select(zone_id, week, outbreak_count),
    by = c("zone_id", "week")
  )

cat("Flagged strata and their outbreak counts:\n")
print(flagged_with_count)

# Only remove strata where the sole outbreak is the non-independent one
strata_to_remove <- flagged_with_count %>%
  filter(outbreak_count == 1) %>%
  distinct(stratum_id)

cat("\nStrata to remove (outbreak_count == 1):", nrow(strata_to_remove), "\n")
cat("Strata retained (outbreak_count > 1):", nrow(flagged_with_count) - nrow(strata_to_remove), "\n")

# Remove only those strata
case_crossover_df_ex <- case_crossover_df %>%
  filter(!(stratum_id %in% strata_to_remove$stratum_id))

cat("\nStrata removed:", nrow(strata_to_remove), "\n")
cat("Rows removed:", nrow(case_crossover_df) - nrow(case_crossover_df_ex), "\n")

# Cases and controls
cat("Total cases (zone-weeks):", sum(case_crossover_df_ex$outbreak_binary == 1), "\n")
cat("Total controls (zone-weeks):", sum(case_crossover_df_ex$outbreak_binary == 0), "\n")

# Summary table
cat("\nCase-control distribution after excluding non-independent:\n")
table(case_crossover_df_ex$case_control)

# Controls per case
cat("\nControls per case after excluding non-independent:\n")
case_crossover_df_ex %>%
  filter(case_control == "control") %>%
  count(stratum_id) %>%
  count(n, name = "n_cases") %>%
  rename(n_controls = n)
```

# 1km sensitivity
```{r, remove non-independent 1km}

# Get strata that contain a flagged zone-week (1km)
flagged_strata_1km <- case_crossover_df %>%
  filter(flag_nearby_1km == TRUE, case_control == "case") %>%
  distinct(stratum_id, zone_id, week)

# Join outbreak_count
flagged_with_count_1km <- flagged_strata_1km %>%
  left_join(
    case_zones_unique %>% select(zone_id, week, outbreak_count),
    by = c("zone_id", "week")
  )

cat("Flagged strata (1km) and their outbreak counts:\n")
print(flagged_with_count_1km)

# Only remove strata where outbreak_count == 1
strata_to_remove_1km <- flagged_with_count_1km %>%
  filter(outbreak_count == 1) %>%
  distinct(stratum_id)

cat("\nStrata to remove (1km, outbreak_count == 1):", nrow(strata_to_remove_1km), "\n")
cat("Strata retained (1km, outbreak_count > 1):", nrow(flagged_with_count_1km) - nrow(strata_to_remove_1km), "\n")

# Remove only those strata
case_crossover_df_ex_1km <- case_crossover_df %>%
  filter(!(stratum_id %in% strata_to_remove_1km$stratum_id))

cat("\nStrata removed:", nrow(strata_to_remove_1km), "\n")
cat("Rows removed:", nrow(case_crossover_df) - nrow(case_crossover_df_ex_1km), "\n")

cat("Total cases (zone-weeks):", sum(case_crossover_df_ex_1km$outbreak_binary == 1), "\n")
cat("Total controls (zone-weeks):", sum(case_crossover_df_ex_1km$outbreak_binary == 0), "\n")

cat("\nCase-control distribution after excluding non-independent (1km):\n")
table(case_crossover_df_ex_1km$case_control)

cat("\nControls per case after excluding non-independent (1km):\n")
case_crossover_df_ex_1km %>%
  filter(case_control == "control") %>%
  count(stratum_id) %>%
  count(n, name = "n_cases") %>%
  rename(n_controls = n)
```

# 200m sensitivity
```{r, 200m sensitivity}
flagged_strata_200m <- case_crossover_df %>%
  filter(flag_nearby_200m == TRUE, case_control == "case") %>%
  distinct(stratum_id, zone_id, week)

flagged_with_count_200m <- flagged_strata_200m %>%
  left_join(
    case_zones_unique %>% select(zone_id, week, outbreak_count),
    by = c("zone_id", "week")
  )

cat("Flagged strata (200m) and their outbreak counts:\n")
print(flagged_with_count_200m)

strata_to_remove_200m <- flagged_with_count_200m %>%
  filter(outbreak_count == 1) %>%
  distinct(stratum_id)

case_crossover_df_ex_200m <- case_crossover_df %>%
  filter(!(stratum_id %in% strata_to_remove_200m$stratum_id))

cat("Strata removed (200m):", nrow(strata_to_remove_200m), "\n")
cat("Total cases:", sum(case_crossover_df_ex_200m$outbreak_binary == 1), "\n")
```

# Export casecrossover df with exclusions (this is our main casecrossover dataframe)

```{r, export casecrossover df}

saveRDS(case_crossover_df_ex, file.path(objects_folder,"case_crossover_df.RDS"))
write.csv(case_crossover_df_ex, file.path(processed_data, "case_crossover_df_ind_500m.csv"), row.names = FALSE)

```

### Export full casecrossover df (no exclusions)

```{r, export casecrossover df}

saveRDS(case_crossover_df, file.path(objects_folder,"full_case_crossover_df.RDS"))
```


```{r, export casecrossover df 1km sensitivity}
saveRDS(case_crossover_df_ex_1km, file.path(objects_folder, "case_crossover_df_1km.RDS"))
write.csv(case_crossover_df_ex_1km, file.path(processed_data, "case_crossover_df_ind_1km.csv"), row.names = FALSE)
```


```{r, export casecrossover df 200m sensitivity}
saveRDS(case_crossover_df_ex_200m, file.path(objects_folder, "case_crossover_df_200m.RDS"))
write.csv(case_crossover_df_ex_200m, file.path(processed_data, "case_crossover_df_ind_200m.csv"), row.names = FALSE)
```

