---
title: "Climate anomaly baselines"
author: "Fintan Mooney"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

```{r initiate}
rm(list = ls())
# 0a.Declare root directory, folder location and load essential stuff
project.folder = paste0(print(here::here()), '/')
source(paste0(project.folder, 'create_folder_structure.R'))
source(paste0(functions.folder, 'script_initiate.R'))
```

# 1. Load the GEE weekly climatology and the comparison-year panel

This panel is on the **8-state grid**, not the legacy MN fishnet. `zone_id` is a
character grid coordinate like `"-4,227"`, and everything joined in below has to
be on the same grid or the joins match nothing.

The climatology used to be built here, by reading 26 per-year CSVs and taking
`group_by(zone_id, week) %>% summarise(mean, sd)`.
`a_00_era5_land_gee_mn8_climatology.js` does that reduction server-side now, so
this reads three files instead of 26 and ~520k rows instead of 4.5M.

Weeks are indexed `week_idx` 0-51, meaning "Jan 1 + 7i", and that is the join key
throughout. The old code joined on `lubridate::isoweek(week_start)`, which drifts:
`isoweek("2010-01-01")` is 53, so `filter(week <= 52)` silently dropped the first
week of 1999, 2005, 2010, 2016 and 2021, and `isoweek()` maps `week_idx` onto a
different calendar week in every year. `week` is still carried below, but purely
as a label for downstream scripts, never as a join key.

```{r load}
# 1a.baseline label -> GEE climatology file. add a row to add a baseline.
# 15yr and 20yr are gone, this script built them and nothing ever read them.
# 10yr stays because b_03 and s_06 both want _z_10yr.
clim_files <- c("25yr" = "mn8_climatology_1997_2021.csv",
                "10yr" = "mn8_climatology_2012_2021.csv")

# 1b.the comparison year, i.e. the numerator in z = (x - mean) / sd.
# zone_id here is the 8-state grid coordinate ("-4,227"), a character key, NOT the
# old numeric MN fishnet id. forced to character everywhere below so a type
# mismatch can't quietly hand back an empty join.
panel <- read_csv(paste0(env_data, "mn8_weekly_era5land_2022.csv")) %>%
  mutate(
    zone_id    = as.character(zone_id),
    year       = 2022L,
    week_start = as.Date(as.character(week_start), format = "%Y%m%d"),
    # kept for downstream joins and the (1|week) random intercept in d_00
    week       = lubridate::isoweek(week_start)
  )

# 1c.snow_density is already de-sentinelled in GEE (it parks near 100 kg/m3 with
# no snow). redoing it here is a no-op; doing it ONLY here would put the panel and
# the baseline on different scales.

# 1d.vars come from whatever the climatology actually holds, so adding a band in
# the GEE config flows through without anyone touching this file.
vars <- read_csv(paste0(env_data, clim_files[[1]]), n_max = 0) %>%
  names() %>%
  grep("_mean$", ., value = TRUE) %>%
  sub("_mean$", "", .)

message("Anomaly variables (", length(vars), "): ", paste(vars, collapse = ", "))
stopifnot(all(vars %in% names(panel)))
```

# 2. Join the baselines and compute z-scores

```{r climatology}
# 2a.each climatology chips in <var>_mean_<label> and <var>_sd_<label>.
# lat/lon get dropped, they already ride along on the panel.
anomaly_panel <- panel

for (lbl in names(clim_files)) {
  clim <- read_csv(paste0(env_data, clim_files[[lbl]])) %>%
    mutate(zone_id = as.character(zone_id)) %>%
    select(-any_of(c("lat", "lon", "state"))) %>%
    rename_with(~ paste0(.x, "_", lbl), -c(zone_id, week_idx))

  anomaly_panel <- left_join(anomaly_panel, clim, by = c("zone_id", "week_idx"))
}

# 2b.every panel row should have found a baseline row. if this trips, the two
# sides were built on different week definitions.
stopifnot(!anyNA(anomaly_panel[[paste0(vars[1], "_mean_", names(clim_files)[1])]]))

for (v in vars) {
  for (lbl in names(clim_files)) {
    anomaly_panel[[paste0(v, "_z_", lbl)]] <-
      (anomaly_panel[[v]] - anomaly_panel[[paste0(v, "_mean_", lbl)]]) /
        anomaly_panel[[paste0(v, "_sd_", lbl)]]
  }
}
```

# 3. Winsorize z-scores at ±4

```{r winsorize}
z_cols <- grep("_z_(10yr|25yr)$", names(anomaly_panel), value = TRUE)

winsorize <- function(x, limit = 4) pmin(pmax(x, -limit), limit)

for (v in z_cols) {
  anomaly_panel[[v]] <- winsorize(anomaly_panel[[v]], limit = 4)
}

# 3a.b_03, s_06 and d_00 all want `runoff`, not `runoff_sum`. on the 8-state
# files this is already a no-op since the column ships as `runoff` (surface).
# kept as a safety net in case legacy runoff_sum columns come back through.
anomaly_panel <- anomaly_panel %>%
  rename_with(~ gsub("runoff_sum", "runoff", .x), .cols = everything())
```

# 4. Attach outcome and spatial covariates

```{r outcomes}
# panel is on the 8-state grid, so outcomes have to come off the 8-state
# timeseries too. fishnet_timeseries_daily.RDS is the legacy MN fishnet with
# numeric zone_ids 1..3334, and joining that here matches nothing and hands back a
# table of zeros that looks completely reasonable. hence the guard at 4b.
daily <- readRDS(paste0(objects_folder, "multistate_fishnet_timeseries_daily.RDS"))
setDT(daily)
daily <- daily[state == "Minnesota"]
daily[, zone_id := as.character(zone_id)]

# 4a.bin outbreak days onto the same Jan-1-anchored grid the weather uses, so a
# late-December case gets late-December weather. the old isoweek grouping put
# Dec 26-31 in isoweek 52, which in this panel is the week of Jan 1. Dec 31 lands
# in a 53rd stub week, so clamp it into week 51 rather than binning the day.
week_idx_of <- function(d) {
  pmin(as.integer(floor(as.numeric(d - as.Date("2022-01-01")) / 7)), 51L)
}

weekly_outbreak <- daily[date >= as.Date("2022-01-01") & date <= as.Date("2022-12-31"),
                          .(outbreak_binary = as.integer(sum(outbreak_binary) > 0),
                            outbreak_count  = sum(outbreak_count, na.rm = TRUE),
                            season          = first(as.character(season)),
                            month           = first(month)),
                          by = .(zone_id, week_idx = week_idx_of(date))]

zone_static <- daily[, .(feedlot_count    = first(feedlot_count),
                         land_cover_label = first(as.character(land_cover_label))),
                     by = zone_id]

# 4b.the join that used to fail silently. if the two sides drift apart again
# (grid change, zone_id type change) this kills the run instead of quietly
# handing back an all-zero outcome column.
matched <- length(intersect(unique(anomaly_panel$zone_id), unique(weekly_outbreak$zone_id)))
if (matched == 0) {
  stop("Outcome join matched 0 zone_ids. anomaly_panel has e.g. '",
       anomaly_panel$zone_id[1], "', outcomes have e.g. '",
       weekly_outbreak$zone_id[1], "'. Are both on the 8-state grid?")
}
message("Outcome join: ", matched, " zone_ids matched (expect 2306).")

anomaly_panel <- anomaly_panel %>%
  left_join(weekly_outbreak, by = c("zone_id", "week_idx")) %>%
  left_join(zone_static,     by = "zone_id") %>%
  mutate(
    outbreak_binary = tidyr::replace_na(outbreak_binary, 0L),
    outbreak_count  = tidyr::replace_na(outbreak_count,  0L),
    season          = factor(season, levels = c("Winter", "Spring", "Summer", "Fall"))
  )
```

# 5. Export

```{r export}
saveRDS(anomaly_panel, paste0(objects_folder, climate_anomaly_panel_rds))
```
