---
title: "Multi-state case-crossover DLNM — Mississippi-flyway bird abundance & HPAI spillover"
author: "Fintan Mooney"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

```{r initiate}
rm(list = ls())
# 0a. Declare root directory, folder location and load essential stuff
project.folder = paste0(print(here::here()), '/')
source(paste0(project.folder, 'create_folder_structure.R'))
source(paste0(functions.folder, 'script_initiate.R'))

# 0b. Resolve the year()/month() ambiguity (lubridate vs data.table) for the
#     project's conflicted setup; we use lubridate's date accessors below.
conflict_prefer("year",  "lubridate", quiet = TRUE)
conflict_prefer("month", "lubridate", quiet = TRUE)
```

This script extends the Minnesota weekly case-crossover (`c_01_casecrossover_weekly.Rmd`)
to the Mississippi-flyway states for which eBird relative-abundance was extracted
(MN, IA, WI, MI, OH, IL, IN). The input is already in spillover-event case-crossover form:
each outbreak premises carries its full surrounding weekly series, with `is_case_week`
flagging the outbreak week and `lag_weeks_from_case` the offset. The exposure here is
bird abundance only (no meteorology), so the two referent designs are fit on the bird
crossbasis alone. We retain the manuscript's two matching schemes but name them plainly:

  * **unidirectional**  — 1:4 matching, controls at the four weeks before each case week.
  * **time-stratified** — controls are the other weeks in the case's calendar month (2022).

# 1. Load spillover-event panel and apply the independence filter

```{r load-filter}
# 1a. Read the study-region case-crossover panel (eBird abundance already joined).
cc <- fread(paste0(bird_abundance_data,
                   "study_spillover_events_casecrossover_2022_abundance_060326.csv"))

# 1b. Keep only independent (IND) and INR-linked outbreaks, dropping contiguous-
#     spread (CSLT) and unsequenced (NOSEQ) events. This matches the manuscript's
#     independence criterion and the flyway EDA panel (b_06). `link` is constant
#     within an outbreak, so we can filter on the whole premises series at once.
keep_ids <- cc[is_case_week == 1 & link %in% c("IND", "INR"), unique(outbreak_id)]
cc <- cc[outbreak_id %in% keep_ids]

# Derive the modeled-state set from the data so counts/labels never go stale as
# the input CSV grows (states are added upstream as bird abundance is extracted).
study_states   <- sort(unique(cc[is_case_week == 1, state]))
n_study_states <- length(study_states)

# 1c. Order each premises' weekly series so within-outbreak lags (shift) are correct.
cc[, week_start := as.Date(week_start)]
setorder(cc, outbreak_id, week_start)

cat("Outbreaks (IND+INR): ", length(keep_ids),
    " across ", uniqueN(cc$state), " states\n", sep = "")
print(cc[is_case_week == 1, .N, by = state][order(-N)])
```

```{r exposure-vars}
# 1d. Define the exposures. anseri_abd (anseriformes aggregate) is primary and
#     mirrors the MN "anseriformes" predictor; total_abd (all 15 species) is a
#     secondary aggregate; the 15 individual species drive a univariate screen.
species_codes <- c("mallar3", "cangoo", "wooduc", "truswa", "hoomer", "buwtea",
                   "comgol", "commer", "buffle", "norsho", "rethaw", "grhowl",
                   "baleag", "amwpel", "snogoo")
agg_vars      <- c("anseri_abd", "total_abd")
exposure_vars <- c(agg_vars, species_codes)

# 1e. Common-name map for tables/figures. The project lookup covers most codes;
#     hoomer and snogoo are not in it, so add them explicitly.
spec_lookup <- fread(paste0(bird_abundance_data, "mn-ebird-names-lookup-050225.csv"))
name_map <- setNames(spec_lookup$common_name, spec_lookup$species_code)
name_map["hoomer"] <- "Hooded Merganser"
name_map["snogoo"] <- "Snow Goose"
name_map["anseri_abd"] <- "Anseriformes (aggregate)"
name_map["total_abd"]  <- "All species (aggregate)"

# Waterfowl vs. other HPAI-detected birds (raptors + pelican) for forest grouping.
non_anseri_spp <- c("rethaw", "grhowl", "baleag", "amwpel")
```

# 2. Transform exposures and build the weekly lag matrix (lag 0–4 weeks)

```{r scale-and-lag}
# 2a. log1p the right-skewed abundances, then z-scale each across all weekly rows
#     of the retained outbreaks (a stable space-time SD for the study region).
#     Effects are reported per +0.5 SD, matching the manuscript's scale.
for (v in exposure_vars) {
  cc[, (v) := as.numeric(scale(log1p(get(v))))]
}

# 2b. For each exposure, create lag-0..4-week columns by shifting within outbreak.
#     The series is consecutive weeks per premises, so shift() gives true weekly
#     lags. Cells reaching into late-2021 (pre-eBird-join) become NA and are
#     dropped per-stratum below.
MAX_LAG_WK <- 4
for (v in exposure_vars) {
  cc[, (paste0(v, "_Lag", 0:MAX_LAG_WK)) :=
       shift(.SD[[1]], n = 0:MAX_LAG_WK, type = "lag"),
     by = outbreak_id, .SDcols = v]
}
```

# 3. Assemble the two referent designs

```{r build-designs}
# 3a. unidirectional (1:4): case week + the four weeks immediately before it.
design_uni <- cc[lag_weeks_from_case %in% c(0, -1, -2, -3, -4)]
design_uni[, stratum_id := paste0("uni_", outbreak_id)]

# 3b. time-stratified: all weeks sharing the case week's calendar year + month.
case_month <- cc[is_case_week == 1,
                 .(outbreak_id,
                   c_year  = year(week_start),
                   c_month = month(week_start))]
design_ts <- merge(cc, case_month, by = "outbreak_id")
design_ts <- design_ts[year(week_start) == c_year & month(week_start) == c_month]
design_ts[, stratum_id := paste0("ts_", outbreak_id)]

# 3c. Per-exposure stratum prep: keep rows whose lag-0..4 are all observed, then
#     retain only strata that still hold a case AND >=1 control (clogit needs the
#     contrast). Returned fresh per exposure since the NA pattern differs by var.
prep_strata <- function(dd, var) {
  lag_cols <- paste0(var, "_Lag", 0:MAX_LAG_WK)
  dd <- dd[complete.cases(dd[, ..lag_cols])]
  ok <- dd[, .(nc = sum(is_case_week == 1L), nk = sum(is_case_week == 0L)),
           by = stratum_id][nc >= 1L & nk >= 1L, stratum_id]
  dd[stratum_id %in% ok]
}

cat("unidirectional rows: ", nrow(design_uni),
    " | time-stratified rows: ", nrow(design_ts), "\n", sep = "")
```

# 4. Core fit + extraction helper (single-exposure clogit DLNM)

```{r fit-helper}
# 4a. Weekly crossbasis: linear exposure-response, natural-spline lag (df = 2)
#     over 0–4 weeks — identical specification to the MN weekly model.
make_cb_weekly <- function(dd, var) {
  crossbasis(
    as.matrix(dd[, paste0(var, "_Lag", 0:MAX_LAG_WK), with = FALSE]),
    lag = c(0, MAX_LAG_WK),
    argvar = list(fun = "lin"),
    arglag = list(fun = "ns", df = 2)
  )
}

# 4b. Fit one conditional logistic DLNM and extract cumulative + per-lag ORs at
#     +0.5 SD. Cumulative contrast sums the lag basis (whole 0–4-wk window); the
#     per-lag contrast evaluates a single lag week. Wrapped in tryCatch so an
#     under-powered stratum set (e.g. a small state) degrades to NA, not an error.
fit_cc <- function(dd, var, design, scope = "pooled") {
  dd <- prep_strata(dd, var)
  out <- tryCatch({
    cb <- make_cb_weekly(dd, var)
    m  <- clogit(is_case_week ~ cb + strata(stratum_id),
                 method = "efron", data = dd)
    beta <- coef(m); V <- vcov(m); ix <- grep("^cb", names(beta))
    b <- beta[ix]; Vsub <- V[ix, ix, drop = FALSE]
    lag_basis <- do.call(dlnm::onebasis,
                         c(list(x = 0:MAX_LAG_WK), attr(cb, "arglag")))

    or_ci <- function(contrast) {
      lo <- as.numeric(contrast %*% b)
      se <- sqrt(as.numeric(t(contrast) %*% Vsub %*% contrast))
      c(OR = exp(lo), lcl = exp(lo - 1.96 * se), ucl = exp(lo + 1.96 * se))
    }
    cum <- or_ci(0.5 * colSums(lag_basis))
    per_lag <- rbindlist(lapply(0:MAX_LAG_WK, function(k) {
      v <- or_ci(0.5 * lag_basis[k + 1, ])
      data.table(lag_wk = k, OR = v["OR"], lcl = v["lcl"], ucl = v["ucl"])
    }))
    list(model = m,
         cumulative = data.table(OR = cum["OR"], lcl = cum["lcl"], ucl = cum["ucl"]),
         per_lag = per_lag,
         n_strata = uniqueN(dd$stratum_id))
  }, error = function(e) list(model = NULL,
                              cumulative = data.table(OR = NA_real_, lcl = NA_real_, ucl = NA_real_),
                              per_lag = NULL, n_strata = uniqueN(dd$stratum_id)))

  meta <- data.table(exposure = var, label = unname(name_map[var]),
                     design = design, scope = scope, n_strata = out$n_strata)
  out$cum_row <- cbind(meta, out$cumulative)
  if (!is.null(out$per_lag)) out$per_lag <- cbind(meta, out$per_lag)
  out
}

# 4c. Tidy OR (95% CI) string helper for tables.
fmt_or <- function(OR, lcl, ucl) ifelse(is.na(OR), "—",
  sprintf("%.2f (%.2f, %.2f)", OR, lcl, ucl))

# 4d. Write one styled flextable + caption to a .docx. All section-8 tables share
#     the same booktabs styling and read_docx → caption → print flow, so they
#     differ only in the data, font size, footnote, and target path.
write_docx_table <- function(tbl, caption, target, font_size = 11) {
  ft <- flextable(tbl) |>
    fontsize(size = font_size, part = "all") |> autofit() |>
    align(align = "center", part = "all") |> theme_booktabs()
  read_docx() |>
    body_add_flextable(ft) |>
    body_add_par(caption, style = "Normal") |>
    print(target = target)
}
```

# 5. Primary + secondary aggregates (anseriformes, total) — both designs

```{r fit-aggregates}
designs <- list(unidirectional = design_uni, `time-stratified` = design_ts)

agg_fits <- list()
for (d in names(designs)) {
  for (v in agg_vars) {
    agg_fits[[paste(d, v)]] <- fit_cc(designs[[d]], v, d)
  }
}

# 5a. Save the two primary anseriformes models for downstream reuse.
saveRDS(agg_fits[["unidirectional anseri_abd"]]$model,
        paste0(models_folder, "flyway_anseri_unidirectional.RDS"))
saveRDS(agg_fits[["time-stratified anseri_abd"]]$model,
        paste0(models_folder, "flyway_anseri_timestrat.RDS"))

agg_cum   <- rbindlist(lapply(agg_fits, `[[`, "cum_row"))
agg_perlag <- rbindlist(lapply(agg_fits, `[[`, "per_lag"))
print(agg_cum[, .(design, label, n_strata,
                  `OR (95% CI)` = fmt_or(OR, lcl, ucl))])
```

# 6. Species-specific univariate screen — both designs

```{r fit-species}
# Each species fit on its own (a single 15-species clogit would be badly
# collinear on ~165 strata); these are an exploratory screen, not adjusted
# effects. Cumulative OR per +0.5 SD over the 0–4-week window.
species_fits <- list()
for (d in names(designs)) {
  for (v in species_codes) {
    species_fits[[paste(d, v)]] <- fit_cc(designs[[d]], v, d)
  }
}
species_cum <- rbindlist(lapply(species_fits, `[[`, "cum_row"))
species_cum[, group := fifelse(exposure %in% non_anseri_spp,
                               "Other HPAI-detected", "Waterfowl (Anseriformes)")]
```

# 7. State sensitivity — pooled vs. per-state anseriformes (unidirectional)

```{r fit-states}
# Pooled is the primary estimate; per-state shown where n supports it. Small
# states (OH, IL, IN; n < 15) can't stand alone, so the per-state set is the
# larger states (MN/IA/WI/MI) — derived by the n >= 15 threshold, not hardcoded.
pooled_label <- sprintf("Pooled (%d states)", n_study_states)
state_n     <- design_uni[is_case_week == 1, .N, by = state]
state_keep  <- state_n[N >= 15, state]                 # MN, IA, WI, MI
state_fits  <- list(fit_cc(design_uni, "anseri_abd", "unidirectional", pooled_label))
for (s in state_keep) {
  state_fits[[length(state_fits) + 1]] <-
    fit_cc(design_uni[state == s], "anseri_abd", "unidirectional", s)
}
state_cum <- rbindlist(lapply(state_fits, `[[`, "cum_row"))
state_cum[, scope := factor(scope, levels = rev(c(pooled_label, state_keep)))]
print(state_cum[, .(scope, n_strata, `OR (95% CI)` = fmt_or(OR, lcl, ucl))])
```

# 8. Tables (Word) — main and sensitivity

```{r tables}
# 8a. MAIN — aggregate cumulative ORs, unidirectional vs. time-stratified.
agg_cum[, OR_CI := fmt_or(OR, lcl, ucl)]
tbl_agg <- dcast(agg_cum, label ~ design, value.var = "OR_CI")
setcolorder(tbl_agg, c("label", "unidirectional", "time-stratified"))
setnames(tbl_agg, "label", "Exposure (per +0.5 SD)")
write_docx_table(
  tbl_agg,
  sprintf(paste0("Cumulative odds ratios (per +0.5 SD, lag 0–4 weeks) for bird ",
                 "abundance and HPAI spillover across %d Mississippi-flyway states ",
                 "(%s; IND+INR events only). unidirectional = ",
                 "1:4 matching (controls at weeks -1 to -4); time-stratified = controls ",
                 "in the case's calendar month. N strata (unidirectional) = %d."),
          n_study_states, paste(study_states, collapse = ", "),
          agg_fits[["unidirectional anseri_abd"]]$n_strata),
  paste0(tables_main_folder, "Table_Flyway_Bird_Cumulative_OR.docx"))

# 8b. MAIN — anseriformes per-lag-week ORs.
anseri_lag <- agg_perlag[exposure == "anseri_abd"]
anseri_lag[, OR_CI := fmt_or(OR, lcl, ucl)]
tbl_lag <- dcast(anseri_lag, design ~ lag_wk, value.var = "OR_CI")
setnames(tbl_lag, c("0", "1", "2", "3", "4"),
         paste0("Lag ", 0:4, " wk"))
setnames(tbl_lag, "design", "Referent design")
write_docx_table(
  tbl_lag,
  "Per-lag-week odds ratios (per +0.5 SD) for anseriformes abundance, weekly DLNM (lag 0–4 wk, ns df = 2).",
  paste0(tables_main_folder, "Table_Flyway_Anseri_PerLagWeek.docx"))

# 8c. SENSITIVITY — species cumulative ORs, both designs.
species_cum[, OR_CI := fmt_or(OR, lcl, ucl)]
tbl_spp <- dcast(species_cum, label + group ~ design, value.var = "OR_CI")
setorder(tbl_spp, group, label)
setnames(tbl_spp, c("label", "group"), c("Species", "Group"))
write_docx_table(
  tbl_spp,
  "Univariate species-specific cumulative ORs (per +0.5 SD, lag 0–4 wk). Exploratory screen; species abundances are spatially correlated, so estimates are unadjusted for co-occurring species.",
  paste0(tables_sensitivity_folder, "Table_Flyway_Species_Cumulative_OR.docx"),
  font_size = 10)

# 8d. SENSITIVITY — pooled vs. per-state anseriformes.
state_tbl <- copy(state_cum)
state_tbl[, OR_CI := fmt_or(OR, lcl, ucl)]
state_tbl <- state_tbl[order(-as.integer(scope)),
                       .(State = as.character(scope), `N strata` = n_strata,
                         `OR (95% CI)` = OR_CI)]
write_docx_table(
  state_tbl,
  "Pooled vs. per-state cumulative ORs for anseriformes abundance (per +0.5 SD, unidirectional design). Per-state estimates limited to states with >=15 outbreaks.",
  paste0(tables_sensitivity_folder, "Table_Flyway_State_Cumulative_OR.docx"))
```

# 9. Figures (SPARK theme, Avenir, 300 DPI)

```{r figures}
# 9. All figures export at the SPARK-standard 300 DPI on a white background; this
#    wrapper fixes those so each call differs only in plot, path, and dimensions.
save_fig <- function(plot, target, width, height) {
  ggsave(target, plot, width = width, height = height, dpi = 300, bg = "white")
}

# 9a. MAIN — anseriformes lag-response curve, both designs. One palette feeds both
#     the line/point colour and the ribbon fill so the two designs match across them.
design_pal <- c(unidirectional = "#32135C", `time-stratified` = "#C9405A")
fig_lag <- ggplot(agg_perlag[exposure == "anseri_abd"],
                  aes(lag_wk, OR, colour = design, fill = design)) +
  geom_hline(yintercept = 1, linetype = "dashed", colour = "grey50") +
  geom_ribbon(aes(ymin = lcl, ymax = ucl), alpha = 0.15, colour = NA) +
  geom_line(linewidth = 0.7) +
  geom_point(size = 1.8) +
  scale_colour_manual(values = design_pal, name = NULL) +
  scale_fill_manual(values = design_pal, name = NULL) +
  labs(x = "Lag (weeks)", y = "Odds ratio (per +0.5 SD)",
       title = "Anseriformes abundance and HPAI spillover, six-state flyway") +
  theme_spark()
save_fig(fig_lag, paste0(figures_main_folder, "figure_flyway_anseri_lag_response.png"),
         width = 7, height = 5)

# 9b. SENSITIVITY — species cumulative-OR forest (unidirectional), grouped.
spp_forest <- species_cum[design == "unidirectional"][order(group, OR)]
spp_forest[, label := factor(label, levels = label)]
fig_spp <- ggplot(spp_forest, aes(OR, label, colour = group)) +
  geom_vline(xintercept = 1, linetype = "dashed", colour = "grey50") +
  geom_errorbarh(aes(xmin = lcl, xmax = ucl), height = 0.25, linewidth = 0.5) +
  geom_point(size = 2) +
  scale_colour_manual(values = c("Waterfowl (Anseriformes)" = "#2D6FA3",
                                 "Other HPAI-detected" = "#C9405A"), name = NULL) +
  labs(x = "Cumulative OR (per +0.5 SD, lag 0–4 wk)", y = NULL,
       title = "Species-specific abundance and HPAI spillover") +
  theme_spark() +
  theme(axis.text.x = element_text(angle = 0, hjust = 0.5))
save_fig(fig_spp, paste0(figures_sensitivity_folder, "figure_flyway_species_forest.png"),
         width = 8, height = 6)

# 9c. SENSITIVITY — pooled vs. per-state anseriformes forest.
fig_state <- ggplot(state_cum, aes(OR, scope)) +
  geom_vline(xintercept = 1, linetype = "dashed", colour = "grey50") +
  geom_errorbarh(aes(xmin = lcl, xmax = ucl), height = 0.2, linewidth = 0.5,
                 colour = "#32135C") +
  geom_point(size = 2.4, colour = "#32135C") +
  labs(x = "Cumulative OR (per +0.5 SD, lag 0–4 wk)", y = NULL,
       title = "Anseriformes abundance: pooled vs. per-state") +
  theme_spark() +
  theme(axis.text.x = element_text(angle = 0, hjust = 0.5))
save_fig(fig_state, paste0(figures_sensitivity_folder, "figure_flyway_state_forest.png"),
         width = 7, height = 4.5)
```

# 10. Save result bundle

```{r save-bundle}
saveRDS(list(
  agg_cum     = agg_cum,
  agg_perlag  = agg_perlag,
  species_cum = species_cum,
  state_cum   = state_cum,
  n_outbreaks = length(keep_ids),
  designs     = list(unidirectional = design_uni, time_stratified = design_ts)
), paste0(objects_folder, "flyway_bird_casecrossover.RDS"))

cat("Done — flyway bird case-crossover models, tables, and figures written.\n")
```
