---
title: "Cross-validation by held-out stratum"
author: "Fintan Mooney"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

```{r initiate}
rm(list = ls())
project.folder = paste0(print(here::here()), '/')
source(paste0(project.folder, 'create_folder_structure.R'))
source(paste0(functions.folder, 'script_initiate.R'))
stopifnot(requireNamespace("INLA", quietly = TRUE))
library(INLA); library(Matrix)
INLA::inla.setOption(num.threads = 1)
source(paste0(functions.folder, 'inla_dlnm_helpers.R'))
```

INLA's CPO is leave-one-*observation*-out and is invalid here: every stratum carries its own
free intercept, so removing one row changes its own stratum's fit and the approximation breaks.
INLA agrees — it flags all 2,604 observations as CPO failures.

The design does support cross-validation at the level it was built on: **hold out whole matched
sets**. For a held-out stratum the natural question is the one the case-crossover actually asks —
given the exposures on each day in this set, which day was the case? The conditional
log-likelihood answers it, and the stratum intercept cancels out of that quantity, so nothing
has to be estimated for the held-out set. Higher is better; a set of *m* days scores
`log(1/m)` under chance.

```{r cv-machinery}
K_FOLDS <- 5
SEED    <- 20260827

PRIM <- setDT(readRDS(paste0(objects_folder, "case_crossover_df_sym_g14.RDS")))[
  state == "Minnesota"]

# Fix the temperature threshold on the full panel. Left to itself, assemble_cc_dlnm derives it
# per-panel, so a training subset would get a different basis from its own test set.
TEMP_AV <- list(temperature = list(fun = "thr", side = "h",
                                   thr.value = temp_threshold_z(PRIM)))

# Linear predictor from the crossbasis and weekly terms only. The stratum intercept is
# deliberately excluded: it cancels in the conditional likelihood.
eta_fixed <- function(fit, obj) {
  cbn <- unlist(obj$cb_cols)
  eta <- rep(0, nrow(obj$data))
  if (length(cbn)) {
    b <- fit$summary.fixed[match(cbn, rownames(fit$summary.fixed)), "mean"]
    eta <- eta + as.matrix(obj$data[, cbn, drop = FALSE]) %*% b
  }
  for (v in obj$weekly_vars) {
    w <- obj$weekly[[v]]
    bb <- utils::tail(fit$summary.random[[w$idx]], length(w$knots))$mean
    eta <- eta + w$Z %*% bb
  }
  as.numeric(eta)
}

# conditional log-likelihood of the observed case day within each held-out stratum
cond_ll <- function(eta, y, stratum) {
  d <- data.table(eta, y, stratum)
  d[, .(ll = eta[y == 1][1] - (max(eta) + log(sum(exp(eta - max(eta)))))), by = stratum]$ll
}

cv_score <- function(label, panel = PRIM, ...) {
  set.seed(SEED)
  st   <- unique(panel$stratum_id)
  fold <- setNames(sample(rep_len(1:K_FOLDS, length(st))), st)
  ll <- unlist(lapply(1:K_FOLDS, function(k) {
    tr <- panel[stratum_id %in% names(fold)[fold != k]]
    te <- panel[stratum_id %in% names(fold)[fold == k]]
    o_tr <- do.call(assemble_cc_dlnm, c(list(tr), list(argvar_by_var = TEMP_AV), list(...)))
    o_te <- do.call(assemble_cc_dlnm, c(list(te), list(argvar_by_var = TEMP_AV), list(...)))
    f_tr <- fit_cc_inla_dlnm(o_tr)
    cond_ll(eta_fixed(f_tr, o_te), o_te$data$outbreak_binary, te$stratum_id)
  }))
  # keep the per-stratum values: mean differences between specifications are paired, and
  # without the pairing there is no way to tell a real gap from fold noise
  data.table(Specification = label, stratum = seq_along(ll), ll = ll)
}
```

```{r cv-run}
SKEW  <- c("runoff", "precipitation")
NOLOG <- setDT(readRDS(paste0(objects_folder, "case_crossover_df_sym_g14_nolog.RDS")))[
  state == "Minnesota"]

cv_raw <- rbindlist(list(
  cv_score("Primary"),
  cv_score("No bird term",        weekly_vars = character(0)),
  cv_score("+ predators",         weekly_vars = c("anseriformes", "predators")),
  cv_score("Lag window 14 d",     max_lag = 14, bird_knots = c(0, 7, 14)),
  cv_score("Lag window 56 d",     max_lag = 56, bird_knots = seq(0, 56, 7)),
  cv_score("No log, scale only",  panel = NOLOG)
))
cv <- cv_raw[, .(n_strata = .N,
                 `Mean conditional log-likelihood` = round(mean(ll), 4),
                 Total = round(sum(ll), 1)), by = Specification]
# chance level: a stratum of m days scores log(1/m)
chance <- log(1 / (nrow(PRIM) / uniqueN(PRIM$stratum_id)))
cv[, `vs chance` := round(`Mean conditional log-likelihood` - chance, 4)]
print(cv)
cat(sprintf("\nchance level for a %.0f-day matched set: %.4f\n",
            nrow(PRIM) / uniqueN(PRIM$stratum_id), chance))
# paired comparison against the primary: same held-out strata, so difference them per stratum
W <- dcast(cv_raw, stratum ~ Specification, value.var = "ll")
ref <- "Primary"
pair <- rbindlist(lapply(setdiff(names(W), c("stratum", ref)), function(nm) {
  d <- W[[nm]] - W[[ref]]
  data.table(Specification = nm, `Mean difference vs primary` = round(mean(d), 4),
             SE = round(sd(d) / sqrt(length(d)), 4),
             `z` = round(mean(d) / (sd(d) / sqrt(length(d))), 2))
}))
print(pair)
saveRDS(cv, paste0(objects_folder, "si_crossvalidation.RDS"))
saveRDS(pair, paste0(objects_folder, "si_crossvalidation_paired.RDS"))
```
